### 2016-04-16 paper/paper.tex
added first draft of the research paper including all resource files

%#############################
%# Copyright 2016 Otmar Ertl #
%#############################

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode

% This is a simple template for a LaTeX document using the "article" class.
% See "book", "report", "letter" for other types of document.

\documentclass[11pt]{article} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{natbib}

%%% Examples of Article customizations
% These packages are optional, depending whether you want the features they provide.
% See the LaTeX Companion or other references for full information.

%%% PAGE DIMENSIONS
\usepackage{geometry} % to change the page dimensions
\geometry
{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=2in} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

\usepackage{graphicx} % support the \includegraphics command and options
\usepackage{svg}

\usepackage{hyperref}
\usepackage{algpseudocode,algorithm,algorithmicx}
\usepackage{float}
\usepackage{url}

% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
% These packages are all incorporated in the memoir class to one degree or another...

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

%%% SECTION TITLE APPEARANCE
\usepackage{sectsty}
\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
% (This matches ConTeXt defaults)

%%% ToC (table of contents) APPEARANCE
\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!

% ALGORITHMS
\newcommand*{\algoref}[1]{\hyperref[#1]{Algorithm~\ref*{#1}}}

% SYMBOLS
\newcommand*{\symDefine}[2]{\newcommand{{#1}}{{#2}}}

\symDefine{\symPrecision}{p}
\symDefine{\symRegRange}{q}
\symDefine{\symRegVal}{k}
\symDefine{\symRegValVariate}{K}
\symDefine{\symNumReg}{m}
\symDefine{\symDataItem}{D}
\symDefine{\symBitRepA}{a}
\symDefine{\symBitRepB}{b}
\symDefine{\symIndexI}{i}
\symDefine{\symIndexK}{k}
\symDefine{\symIndexL}{l}
\symDefine{\symIndexJ}{j}
\symDefine{\symCount}{c}
\symDefine{\symAlpha}{\alpha}
\symDefine{\symBeta}{\beta}
\symDefine{\symA}{a}
\symDefine{\symB}{b}
\symDefine{\symE}{e}
\symDefine{\symX}{x}
\symDefine{\symXEstimate}{\hat{\symX}}
\symDefine{\symXNormalized}{\symX'}
\symDefine{\symN}{n}
\symDefine{\symS}{s}
\symDefine{\symY}{y}
\symDefine{\symZ}{z}
\symDefine{\symCardinality}{n}
\symDefine{\symCardinalityEstimate}{\hat{\symCardinality}}
\symDefine{\symCardinalityOriginalEstimate}{\symCardinalityEstimate_\text{raw}}
\symDefine{\symError}{\varepsilon}
\symDefine{\symStopDelta}{\delta}
\symDefine{\symStopEpsilon}{\varepsilon}
\symDefine{\symBigO}{\mathcal{O}}
\symDefine{\symProbability}{P}
\symDefine{\symProbabilityMass}{\rho}
\symDefine{\symRegProbability}{\gamma}
\symDefine{\symNu}{\nu}
\symDefine{\symKappa}{\kappa}
\symDefine{\symLikelihood}{\mathcal{L}}
\symDefine{\symPoissonRate}{\lambda}
\symDefine{\symPoissonRateEstimate}{\hat{\symPoissonRate}}
\symDefine{\symFunc}{f}
\symDefine{\symFuncPrime}{g}
\symDefine{\symHelper}{h}
\symDefine{\symHelperApprox}{\tilde{\symHelper}}
\symDefine{\symSetA}{A}
\symDefine{\symSetB}{B}
\symDefine{\symSetS}{S}
\symDefine{\symSetX}{X}
\symDefine{\symSetASuffix}{a}
\symDefine{\symSetBSuffix}{b}
\symDefine{\symSetXSuffix}{x}
\symDefine{\symCountMatrix}{\mathbf{c}}

%%% END Article customizations

\title{Maximum likelihood estimation of cardinalities\\ from HyperLogLog sketches}
\author{Otmar Ertl}
%\date{} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed

\begin{document}
\maketitle
\begin{abstract}
This paper presents a new method to estimate the cardinality of a multiset recorded by HyperLogLog sketches. The maximum likelihood method is applied to the probability mass function under the Poisson model. A numerical algorithm is presented that is able to calculate the estimate efficiently. Simulations finally show that this new estimation procedure works well and is inherently unbiased over a wide range of cardinalities.
\end{abstract}

\section{Introduction}
Counting the number of distinct elements in a data stream or large datasets is a common problem in big data processing. Often there are parallel streams  or data is spread over a cluster, which makes this task even more challenging.
In principle, finding the number of distinct elements $\symCardinality$ with a maximum relative error $\symError$  in a data stream requires $\Omega(\symCardinality)$ space \cite{Alon1999}. However, probabilistic algorithms that achieve the requested accuracy only with high probability are able to drastically reduce space requirements. Many different probabilistic algorithms have been developed over the past two decades \cite{Metwally2008,Ting2014}. An algorithm with an optimal space complexity of $\Omega(\symError^{-2}+\log \symCardinality)$ \cite{Alon1999, Indyk2003} was finally presented \cite{Kane2010}. This algorithm, however, is not very efficient in practice \cite{Ting2014}.

Currently, the best algorithm that also works for distributed setups is the near-optimal HyperLogLog algorithm \cite{Flajolet2007} with space complexity $\Omega(\symError^{-2} \log\log\symCardinality +\log \symCardinality)$. The originally proposed estimation method has some problems to guarantee the same estimation error over the entire range of cardinalities. It was proposed to correct the estimate by empirical means \cite{Heule2013,Rhodes2015,Sanfilippo2014}.

In case the data is not distributed and results do not need to be aggregated further, there are even more efficient estimation algorithms available. On the one hand there is the self-learning bitmap \cite{Chen2011, Chen2015}, and on the other hand there is the HyperLogLog algorithm extended by a historic inverse probability estimator that is continuously updated together with the HyperLogLog sketch \cite{Ting2014}. Both achieve the same estimation error using less space. However, the estimated cardinality depends on the insertion order of elements and hence cannot be used in a distributed environment.

%In contrast, HyperLogLog sketches (without historic inverse probability estimator) built from different sets of elements can be merged and the result is identical to a HyperLogLog sketch obtained by inserting all elements of the union set. Therefore, the HyperLogLog sketch inherently supports cardinality estimation of unions. However, there are also applications which require the cardinality estimation of intersections.

\section{The HyperLogLog algorithm}
The HyperLogLog algorithm uses a sketching data structure that consists of $\symNumReg$ registers. For performance reasons the number of registers is chosen to be a power of 2: $\symNumReg = 2^\symPrecision$. $\symPrecision$ is the precision that directly influences the relative error scaling like $1/\sqrt{\symNumReg}$. The number of bits per register defines the value range that can be represented. Initially all registers are set equal to 0.

\subsection{Data element  insertion}
In order to insert a data element into a HyperLogLog data structure a hash value is calculated. The first $\symPrecision$ bits of the hash value are used to select one of the $2^\symPrecision$ registers. For another $\symRegRange$ bits, the position of the first 1-bit is determined. If this position exceeds the value of the selected register, the register value is replaced. \algoref{alg:insert} shows the update procedure for inserting a data element into the HyperLogLog sketch.

A HyperLogLog sketch can be characterized by the parameter pair $(\symPrecision, \symRegRange)$. The first parameter controls the relative error while the second defines the possible value range of a register. A register can take all values starting from 0 to $\symRegRange+1$, inclusively.

The number of bits of the consumed hash value $\symPrecision+\symRegRange$ limit the maximum cardinality that can be tracked. Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent  , which will reduce estimation accuracy drastically.

At any time a $(\symPrecision, \symRegRange)$-HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be compressed into a $(\symPrecision', \symRegRange')$-HyperLogLog data structure, if $\symPrecision'\leq\symPrecision$ and $\symPrecision'+\symRegRange' \leq\symPrecision + \symRegRange$ is satisfied. This transformation is lossless in a sense that the resulting HyperLogLog sketch is the same as if all elements would have been recorded by a $(\symPrecision', \symRegRange')$-HyperLogLog sketch right from the beginning.

Note that a $(\symPrecision, 0)$-HyperLogLog sketch corresponds to a bit array as used by linear probabilistic counting \cite{Whang1990}. In this case each register value is represented by a single bit. Hence, linear probabilistic counting can be regarded as a special case of the HyperLogLog algorithm, where $\symRegRange = 0$.

\begin{algorithm}
\caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers $\vec{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg$) initialized to 0 that can take all integers in $[0, \symRegRange+1]$.}
\label{alg:insert}
\begin{algorithmic}
\Procedure {InsertElement}{\symDataItem, $\vec{\symRegVal}$}
\State $\langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision,\symBitRepB_1,\ldots,\symBitRepB_\symRegRange\rangle_2 \gets$ $(\symPrecision + \symRegRange)$-bit hash value of $\symDataItem$
\State $\symIndexJ \gets 1+ \langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision\rangle_2$
\State $\symRegVal' = \min(\{\symS\mid \symS \in [1, \symRegRange]  \wedge  \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
\State $\symRegVal_\symIndexJ \gets\max(\symRegVal_\symIndexJ, \symRegVal')$
\EndProcedure
\end{algorithmic}
\end{algorithm}

\subsection{Original cardinality estimation method}
The cardinality estimate is based on the idea that approximately $\symNumReg  2^{\symRegVal_\symIndexJ}$ distinct element insertions are needed until the value of register $\symIndexJ$ reaches $\symRegVal_\symIndexJ$. Given that, a cardinality estimate can be obtained by calculating the average over the values $\lbrace \symNumReg 2^{\symRegVal_1},\ldots,\symNumReg2^{\symRegVal_\symNumReg}\rbrace$.

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. It is used for the \emph{raw} estimate of the HyperLogLog algorithm and is given by
\begin{equation}
\label{equ:intermediate_estimate}
\symCardinalityOriginalEstimate = \frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
\end{equation}
where $\symAlpha_\symNumReg$  is a correction factor that is fixed for a given number of registers $\symNumReg$ and which was derived to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{-1}
\end{equation}
For $\symPrecision \geq 4$ theses values can be approximated as follows
\begin{equation}
\symAlpha_{16}\approx 0.673,
\,
\symAlpha_{32}\approx 0.697,
\,
\symAlpha_{64}\approx 0.709,
\,
\symAlpha_\symNumReg\approx 0.7213/(1+1.079/\symNumReg)
\,\text{for}\,\symNumReg\geq 64 
\end{equation}

\autoref{fig:raw_estimate} shows the distribution of the relative error of the \emph{raw} estimate for $10000$ randomly generated HyperLogLog sketches over the number of distinct element insertions. Obviously, the \emph{raw} estimate is biased for small and large cardinalities and fails to return accurate estimates. As a consequence, to cover the entire range of cardinalities, a small and a large correction have been proposed, respectively.

For small cardinalities the same estimator as used by the linear probabilistic counting algorithm \cite{Whang1990} which only incorporates the number of registers equal to 0 denoted by $\symCount_0$:
\begin{equation}
\label{equ:small_range_estimate}
\symCardinalityEstimate_\text{small} = \symNumReg \log\left(
\frac{\symNumReg}{\symCount_0}
\right)
\end{equation}
The corresponding relative estimation as depicted in \autoref{fig:small_range_estimate} shows that this estimator is convenient for small cardinalities.
It was proposed to use the small-range estimator as long as $\symCardinalityOriginalEstimate \leq \frac{5}{2}\symNumReg$ where the factor $\frac{5}{2}$ was empirically determined.

For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$ for which a lot of registers already are already in a saturated state meaning that they have already reached the maximum possible value $\symRegRange + 1$, the \emph{raw} estimate underestimates the cardinalities. For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$ which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
\begin{equation}
\label{equ:large_range_estimate}
\symCardinalityEstimate_\text{large}
=
-2^{32}\log\left(1-\frac{\symCardinalityOriginalEstimate}{2^{32}}\right).
\end{equation}

The entire original estimation algorithm as presented in \cite{Flajolet2007} including small and large range corrections is summarized by \algoref{alg:estimate_original}. 
\begin{algorithm}
\caption{Original procedure for estimating the cardinality from a  HyperLogLog data structure using 32-bit hash values ($\symPrecision+\symRegRange = 32$) for insertion of data items \cite{Flajolet2007}.}
\label{alg:estimate_original}
\begin{algorithmic}
\Function {EstimateCardinality}{$\vec{\symRegVal}$}
\State $\symNumReg \gets \dim \vec{\symRegVal}$
\State $\symCardinalityOriginalEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegVal_\symIndexJ}\right)^{-1}$
\Comment \emph{raw} estimate \eqref{equ:intermediate_estimate}
\If{$\symCardinalityOriginalEstimate\leq \frac{5}{2}\symNumReg$}
\State $\symCount_0 = \left|\{\symIndexJ |\symRegVal_\symIndexJ=0\}\right|$
\If{$\symCount_0\neq0$}
\State \Return $\symNumReg \log(\symNumReg / \symCount_0)$ \Comment small range correction \eqref{equ:small_range_estimate}
\Else
\State \Return $\symCardinalityOriginalEstimate$
\EndIf
\ElsIf{$\symCardinalityOriginalEstimate\leq \frac{1}{30}2^{32}$}
\State \Return $\symCardinalityOriginalEstimate$
\Else
\State\Return $-2^{32}\log(1-\symCardinalityOriginalEstimate/2^{32})$\Comment large range correction \eqref{equ:large_range_estimate}
\EndIf
\EndFunction
\end{algorithmic}
\end{algorithm}
The relative estimation error for the original method is shown in \autoref{fig:original_estimate}. Unfortunatley, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityOriginalEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap. Therefore, the estimation error is much larger near the transition region. To reduce the estimation error for cardinalities close to this region, it was proposed to correct $\symCardinalityOriginalEstimate$ for bias. The empirically determined bias correction data can be either stored as set of interpolation points \cite{Heule2013}, as lookup table \cite{Rhodes2015}, or as best-fitting polynomial \cite{Sanfilippo2014}.

\begin{figure}
\centering
\includesvg[width=1\textwidth]{raw_estimate}
\caption{The distribution of the relative estimation error over the cardinality as obtained by the raw estimation method \cite{Flajolet2007}. 10000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$ have been evaluated.}
\label{fig:raw_estimate}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{small_range_estimate}
\caption{The distribution of the relative estimation error for the small range correction.}
\label{fig:small_range_estimate}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{original_estimate}
\caption{The distribution of the relative estimation error over the cardinality as obtained by the original estimation method \cite{Flajolet2007}. 10000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$ have been evaluated.}
\label{fig:original_estimate}
\end{figure}

The large range correction is not satisfying either. The large range correction does not reduce the estimation error. It even increases the estimation error. However, instead of underestimating the cardinalities, they are now overestimated. A simple approach to avoid the large range correction at all is the use of larger hash values $(\symPrecision+\symRegRange>32)$ which shift the biased region of $\symCardinalityOriginalEstimate$ to higher cardinalities. However, in case $\symRegRange \geq 30$ 6-bit registers as used in \cite{Heule2013} are required to allow representing all values in range $[0, \symRegRange+1]$. If 64-bit hash values are used as proposed in \cite{Heule2013} $(\symPrecision+\symRegRange=64)$, a correction for large cardinalities is not required. Cardinalities in the order of $2^{64}$ for which saturated register become a problem never occur in practice.

%\begin{algorithm}
%\caption{Calculate the sufficient statistic $\vec{\symCount}=(\symCount_0, \ldots,\symCount_{\symRegRange+1})$ for given register values $\vec{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg)$.}
%\begin{algorithmic}
%\Function {CountRegisterValues}{$\vec{\symRegVal}$}
%\State $\vec{\symCount}\gets\vec{0}$
%\ForAll{$\symRegVal'\in\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$}
%\State $\symCount_{\symRegVal'} \gets \symCount_{\symRegVal'} + 1$
%\EndFor
%\State\Return $\vec{\symCount}$
%\EndFunction
%\end{algorithmic}
%\end{algorithm}
%

\section{Cardinality estimation}
\label{sec:cardinality_estimation}
In the following we describe a new cardinality estimation method that does not require special handling for small and large cardinalities and which also does not require any empirically determined constants or corrections. The method is based on the maximum likelihood method.

The maximum likelihood method was already applied to HyperLogLog sketches in \cite{Clifford2012}. However, there, a different element insertion algorithm was assumed that works with time complexity $\symBigO(\symNumReg)$, because each register calculates its own hash value from the inserted element and updates its current value accordingly. In comparison, \algoref{alg:insert} uses only a single hash value that defines which register needs to be updated. As a consequence, the register values are dependent, which makes the joint probability distribution of all register values more complex.

In the following we use the Poisson approximation in order to simplify the model. Despite this approximation the maximum likelihood estimate will give us high quality estimates for the entire cardinality range as simulation results will show later. Furthermore, we will demonstrate that consequent application of the maximum likelihood method to the Poisson model directly reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values. The history of the HyperLogLog algorithm shows that the \emph{raw} estimate \eqref{alg:estimate_original} was first found after several attempts using the geometric mean \cite{Durand2003, Flajolet2007}.

\subsection{Joint probability distribution of register values}

Under the assumption of an ideal hash function the probability mass function for the register values $\vec{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
\begin{equation}
\label{equ:multinomialProbabilityMass}
\symProbabilityMass(\vec{\symRegVal}\vert\symCardinality)
=
\sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots\symCardinality_\symNumReg}
\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexJ=1}^\symNumReg \symRegProbability_{\symCardinality_\symIndexJ\symRegVal_\symIndexJ},
\end{equation}
where $\symCardinality$ is the cardinality. The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities. $\symRegProbability_{\symNu\symKappa}$ is the probability that a register is equal to $\symKappa$ after inserting $\symNu$ distinct elements
\begin{equation}
\symRegProbability_{\symNu\symKappa} 
:=
\begin{cases}
1 & \symNu=0 \wedge \symKappa = 0\\
0 & \symNu=0 \wedge 1\leq\symKappa\leq\symRegRange+1\\
0 & \symNu\geq1 \wedge \symKappa = 0\\
\left(1-\frac{1}{2^\symKappa}\right)^\symNu - \left(1-\frac{1}{2^{\symKappa-1}}\right)^\symNu & \symNu \geq 1 \wedge 1\leq\symKappa\leq\symRegRange\\
1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symNu & \symNu\geq 1 \wedge \symKappa = \symRegRange +1 \\
\end{cases}
\end{equation}
Note that the order of register values $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ is not important for the estimation of the cardinality. More formally, the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$ is a sufficient statistic for $\symCardinality$.
Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace = 0^{\symCount_0}1^{\symCount_1}\cdots\symRegRange^{\symCount_{\symRegRange}}(\symRegRange+1)^{\symCount_{\symRegRange+1}}$ where $\symCount_\symIndexJ$ is the multiplicity of value $\symIndexJ$. As a consequence, the multiplicity vector $\vec{\symCount} := (\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is also a sufficient statistic for the cardinality. In addition, this vector also contains all the information about the HyperLogLog sketch that is required for  cardinality estimation: $\symRegRange=\dim(\vec{\symCount})-2$ and  $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$.

\subsection{Poisson approximation}
Unfortunately, estimating the cardinality for given register values $\vec{\symRegVal}$ by maximizing the likelihood $\symLikelihood(n \vert \vec{\symRegVal}) = \symProbabilityMass(\vec{\symRegVal}\vert\symCardinality)$ is very expensive. Therefore, as proposed in \cite{Flajolet2007} a Poisson model is used instead, which assumes that the cardinality itself is distributed according to a Poisson distribution
\begin{equation}
\symCardinality \sim \text{Poisson}(\symPoissonRate).
\end{equation}
Under the Poisson model the register values are distributed according to
\begin{align}
\symProbabilityMass(\vec{\symRegVal}\vert\symPoissonRate) 
&= 
\sum_{\symCardinality=0}^\infty \symProbabilityMass(\vec{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\\
&= 
\sum_{\symCardinality_1=0}^\infty
\cdots
\sum_{\symCardinality_\symNumReg=0}^\infty
\prod_{\symIndexJ=1}^\symNumReg
\symRegProbability_
{\symCardinality_\symIndexJ\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexJ}}{\symCardinality_\symIndexJ!\symNumReg^{\symCardinality_\symIndexJ}}
\\
&= 
\prod_{\symIndexJ=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability_
{\symCardinality\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\\
&= 
\prod_{\symRegVal=0}^{\symRegRange+1} \left[
\sum_{\symCardinality=0}^\infty
\symRegProbability_
{\symCardinality\symRegVal}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\right]^{\symCount_\symRegVal}
\\
&=
e^{-\symCount_0\frac{\symPoissonRate}{\symNumReg}}
\cdot
\left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\symCount_\symRegVal}\right)
\cdot
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\symCount_{\symRegRange+1}}
\end{align}
This factorization shows that all register values are independent and identically distributed under the Poisson model. The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is
\begin{equation}
\label{equ:register_value_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
=
\begin{cases}
0 & \symRegVal < 0 \\
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}} & 0\leq \symRegVal \leq \symRegRange \\
1 & \symRegVal > \symRegRange
\end{cases}
\end{equation}

\subsection{Maximum likelihood estimation}
The key idea of this paper is to find an estimate $\symPoissonRateEstimate$ for the Poisson parameter $\symPoissonRate$ via maximum likelihood estimation and use it directly as the cardinality estimate 
\begin{equation}
\symCardinalityEstimate_\text{ML} := \symPoissonRateEstimate
\end{equation}

As supported by simulations that will be presented later, this approximation works well for the entire range of cardinalities. In order to get an estimate for $\symPoissonRateEstimate$, maximum-likelihood estimation is used which works out quite well under the Poisson model. The log-likelihood and its derivative are given by
\begin{equation}
\label{equ:log_likelihood_single}
\log \mathcal{\symLikelihood}(\symPoissonRate\vert\vec{\symRegVal}) = 
-\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCount_\symRegVal}{2^\symRegVal}+ 
\sum_{\symRegVal=1}^{\symRegRange}\symCount_\symRegVal \log\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)
+
\symCount_{\symRegRange+1} \log\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegRange}}}\right)
\end{equation}
and
\begin{equation}
\frac{d}{d\symPoissonRate}\log \mathcal{\symLikelihood}(\symPoissonRate\vert\vec{\symRegVal}) 
=
-\frac{1}{\symPoissonRate}\left(
\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}}
+\symCount_{\symRegRange+1}\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}}
\right).
\end{equation}
As a consequence, the maximum-likelihood estimate for the Poisson parameter is given by 
\begin{equation}
\symPoissonRateEstimate = \symNumReg\symXEstimate,
\end{equation}
if $\symXEstimate$ is the root of the function
\begin{equation}
\label{equ:funcdef}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\frac{\frac{\symX}{2^\symRegVal}}{1-e^{\frac{\symX}{2^\symRegVal}}}
+\symCount_{\symRegRange+1}\frac{\frac{\symX}{2^\symRegRange}}{1-e^{\frac{\symX}{2^\symRegRange}}}.
\end{equation}
which can also be written as
\begin{equation}
\label{equ:func}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\left(\frac{\symX}{2^\symRegVal}\right)
+
\symCount_{\symRegRange+1}\symHelper\left(\frac{\symX}{2^\symRegRange}\right)
-
\left(\symNumReg-\symCount_0\right),
\end{equation}
if, the function $\symHelper(\symX)$ is defined as
\begin{equation}
\label{equ:helper}
\symHelper(\symX):=1-\frac{\symX}{e^{\symX}-1}.
\end{equation}
$\symHelper(\symX)$ is strictly increasing and concave as can be seen in \autoref{fig:helper_function}. For non-negative values $\symX$ the function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
\begin{figure}
\centering
\includesvg[width=0.45\textwidth]{helper}
\caption{The function $\symHelper(\symX)$.}
\label{fig:helper_function}
\end{figure}
Since the function $\symFunc(x)$ is also strictly increasing, it is obvious that there exists a unique root $\symXEstimate$ for which $\symFunc(\symXEstimate)=0$. The function is non-positive at 0 since $\symFunc(0)=\symCount_0-\symNumReg\leq 0$ and, in case $\symCount_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing. $\symCount_{\symRegRange}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $(\symRegRange+1)$, for which the maximum likelihood estimate would be infinity.

It is easy to see that the $\symPoissonRateEstimate$ remains equal or is larger after updating the HyperLogLog data structure as defined by \algoref{alg:insert}. An update potentially changes the cardinality vector $(\symCount_0,\ldots,\symCount_{\symRegRange+1})$ to $(\symCount_0,\ldots,\symCount_\symIndexI-1,\ldots,\symCount_\symIndexJ+1,\ldots,\symCount_{\symRegRange+1})$ where $\symIndexI < \symIndexJ$. Writing \eqref{equ:func} as 
\begin{multline}
\symFunc(\symX)
:=
\symCount_0 \symX
+
\symCount_1 \left(
\symHelper\left(\frac{x}{2^1}\right)
+\frac{\symX}{2^1}-1
\right)
+
\symCount_2 \left(
\symHelper\left(\frac{x}{2^2}\right)
+\frac{\symX}{2^2}-1
\right)
+
\ldots
\\
\dots
+
\symCount_\symRegRange
\left(
\symHelper\left(\frac{\symX}{2^\symRegRange}\right)
+\frac{\symX}{2^\symRegRange}-1
\right)
+
\symCount_{\symRegRange+1}
\left(
\symHelper\left(\frac{\symX}{2^{\symRegRange}}\right)
-1
\right).
\end{multline}
shows that the coefficient of $\symCount_\symIndexI$ is larger than the coefficient of $\symCount_\symIndexJ$ in case $\symIndexI < \symIndexJ$. Keeping $\symX$ fixed during an update decreases $\symFunc(\symX)$. As a consequence, since $\symFunc(\symX)$ is increasing the new root and hence the estimate must be larger than before the update.

Note that \eqref{equ:funcdef} can be solved analytically for the special case $\symRegRange=0$ which corresponds to the already mentioned linear probabilistic counting algorithm. In this case, the maximum likelihood method under the Poisson model directly leads to the estimator presented in \cite{Whang1990} and which was also used for small range estimation \eqref{equ:small_range_estimate}. Despite the assumption of a Poisson model, it is a very good approximation of the optimal martingale estimator presented in \cite{Ting2014}. Due to this fact we could expect that maximum likelihood estimation under the Poisson model also works very well for the more  general HyperLogLog case.

\subsection{Inequalities for the maximum likelihood estimate}
In the following lower and upper bounds for $\symXEstimate$ are derived.
Applying Jensen's inequality on $\symHelper$ in \eqref{equ:func} gives an upper bound for $\symFunc(\symX)$:
\begin{equation}
\symFunc(\symX)
\leq
\symX
\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
+
\left(\symNumReg-\symCount_0\right)\cdot
\symHelper\left(
\symX\cdot
\left(
\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
\right)\right)
-
\left(\symNumReg-\symCount_0\right).
\end{equation}
The left-hand side is zero, if $\symXEstimate$ is inserted. Resolution for $\symXEstimate$ finally gives the lower bound
\begin{equation}
\label{equ:strong_lower_bound}
\symXEstimate\geq \frac{\symNumReg-\symCount_0}{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
}
\log\left(
1
+
\frac{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
}
{
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
}
\right).
\end{equation}
This bound can be weakened using $\log(1+x) \geq \frac{2x}{x+2}$ for $x\geq0$ which results in
\begin{equation}
\label{equ:weak_lower_bound}
\symXEstimate
\geq
\frac{\symNumReg-\symCount_0}
{\symCount_0+\frac{3}{2}\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCount_\symRegVal}{2^\symRegVal} + \frac{\symCount_{\symRegRange+1}}{2^\symRegRange}}.
\end{equation}
Using the monotonicity of $\symHelper$ the lower bound
\begin{equation}
\symFunc(\symX)
\geq
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
+
\symCount_{\symRegRange+1}\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
-
\left(\symNumReg-\symCount_0\right),
\end{equation}
for $\symFunc$ can be found, where $\symRegVal'_\text{max} := \min(\symRegVal_\text{max}, \symRegRange)$ and
$\symRegVal_\text{max} := \max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$.
Again, inserting $\symXEstimate$ and transformation gives
\begin{equation}
\label{equ:strong_upper_bound}
\symXEstimate
\leq
2^{\symRegVal'_\text{max}}
\log\left(
1+
\frac{\symNumReg-\symCount_0}
{
2^{\symRegVal'_\text{max}}
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
}
\right)
\end{equation}
as upper bound which can be weakened using $\log(1+\symX)\leq \symX$ for $\symX\geq 0$
\begin{equation}
\label{equ:weak_upper_bound}
\symXEstimate
\leq
\frac{\symNumReg-\symCount_0}
{\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}}.
\end{equation}
Note, if the HyperLogLog sketch is in the intermediate range, where $\symCount_0=\symCount_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the \emph{raw} estimate \eqref{equ:intermediate_estimate}. Hence, the maximum likelihood method directly leads to the harmonic mean that is used for the \emph{raw} estimate. If $\symCount_0>0$ or $\symCount_{\symRegRange+1}>0$ the bounds do no longer follow the harmonic mean and this is also the reason why the \emph{raw} estimate fails for small and large cardinalities.

\subsection{Computation of the maximum likelihood estimate}
Since $\symFunc$ is concave and increasing, both, Newton-Raphson iteration and the secant method, converge to the root, if the function is negative for the starting points. In the following the secant method is chosen to derive the cardinality estimation algorithm presented below. Even though it has the disadvantage of slower convergence, a single iteration is simpler to calculate as it does not require the evaluation of the first derivative as shown in \autoref{fig:original_estimate}.

An iteration step of the secant method can be written as
\begin{equation}
\symX_{\symIndexI} = 
\symX_{\symIndexI-1} -
\left(\symX_{\symIndexI-1}-\symX_{\symIndexI-2}\right)
\frac{\symFunc(\symX_{\symIndexI-1})}{\symFunc(\symX_{\symIndexI-1}) - \symFunc(\symX_{\symIndexI-2})}
\end{equation}
If $\symX_0 = 0$ with $\symFunc(\symX_0)=-\left(\symNumReg-\symCount_0\right)$, and $\symX_1$ is equal to one of the derived lower bounds \eqref{equ:strong_lower_bound} or \eqref{equ:weak_lower_bound}, the sequence $\lbrace\symX_\symIndexI\rbrace$ is montone increasing. Using the definitions
\begin{equation}
\Delta\symX_\symIndexI := \symX_\symIndexI-\symX_{\symIndexI-1}
\end{equation}
and
\begin{equation}
\label{equ:funcprime}
\symFuncPrime(\symX):=\symFunc(\symX) + \left(\symNumReg-\symCount_0\right)
=\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\left(\frac{\symX}{2^\symRegVal}\right)
+
\symCount_{\symRegRange+1}\symHelper\left(\frac{\symX}{2^\symRegRange}\right)
\end{equation}
the iteration scheme can also be written as
\begin{gather}
\label{equ:secant_delta}
\Delta\symX_{\symIndexI} = \Delta\symX_{\symIndexI-1}
\frac{\left(\symNumReg-\symCount_0\right)-\symFuncPrime(\symX_{\symIndexI-1})}{\symFuncPrime(\symX_{\symIndexI-1}) - \symFuncPrime(\symX_{\symIndexI-2})}\\
\symX_{\symIndexI} = \symX_{\symIndexI-1} + \Delta\symX_{\symIndexI}
\end{gather}
The iteration can be stopped, if $\Delta\symX_{\symIndexI} \leq \symStopDelta\cdot \symX_{\symIndexI}$. Since the expected statistical errror for the HyperLogLog data structure scales according to $\frac{1}{\sqrt{\symNumReg}}$ \cite{Flajolet2007}, it makes sense to choose $\symStopDelta = \frac{\symStopEpsilon}{\sqrt{\symNumReg}}$ with some constant $\symStopEpsilon$. For all following results we have used $\symStopEpsilon = 10^{-2}$.

\subsection{The new cardinality estimation algorithm}
In order to get a fast cardinality estimation algorithm, it is crucial to minimize the evaluation costs for \eqref{equ:funcprime}. A couple of optimizations allow significant reduction of the computational effort:
\begin{itemize}
\item Only a fraction of all count values $\symCount_\symRegVal$ is non-zero. If we denote $\symRegVal_\text{min}:=\min\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$ and $\symRegVal_\text{max}:=\max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$,  it is sufficient to loop over all indices in the range $[\symRegVal_\text{min}, \symRegVal_\text{max}]$.
\item It is sufficient to calculate the coefficient of the linear term $\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}$ once and reuse it for any further iterations of the secant method.
\item Many programming languages allow the efficient multiplication and division by any integral power of 2 using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\lbrace\frac{\symX}{2^{\symRegVal'_\text{max}}},\frac{\symX}{2^{\symRegVal'_\text{max}-1}}\ldots,\frac{\symX}{2^{\symRegVal_\text{min}}}\rbrace$ where $\symRegVal'_\text{max} := \min(\symRegVal_\text{max}, \symRegRange)$. This series corresponds to geometric series with ratio 2. A straightforward calculation using \eqref{equ:helper} is very expensive because of the exponential function. However, if we know $\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
\begin{equation}
\label{equ:helper_recursion1}
\symHelper\left(2\symX\right) = \frac{\symX+2\symHelper\left(\symX\right)\left(1-\symHelper\left(\symX\right)\right)}{\symX+2\left(1-\symHelper\left(\symX\right)\right)}
\end{equation}
or
\begin{equation}
\label{equ:helper_recursion2}
\symHelper\left(\frac{\symX}{2^{\symRegVal}}\right) = \frac{\frac{\symX}{2^{\symRegVal+2}}+\symHelper\left(\frac{\symX}{2^{\symRegVal+1}}\right)\left(1-\symHelper\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}{\frac{\symX}{2^{\symRegVal+2}}+\left(1-\symHelper\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}
\end{equation}
Note, this recursive formula is stable in a sense that the relative error of $\symHelper\left(2\symX\right)$ is smaller than that of $\symHelper\left(\symX\right)$ as shown in \autoref{app:helper_stable}.

\item If $\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is smaller than 0.5, the function $\symHelper(\symX)$ can be well approximated by a Taylor series around $\symX=0$
\begin{equation}
\symHelper(\symX)
=
\frac{\symX}{2} - \frac{\symX^2}{12} +\frac{\symX^4}{720}-\frac{\symX^6}{30240} + \symBigO(\symX^{8})
\end{equation}
which can be optimized for numerical evaluation using Estrin's scheme and $\symX' := \frac{\symX}{2}$ and $\symX'' := \symX' \symX'$
\begin{equation}
\label{equ:taylor}
\symHelper(\symX)
=
\symX' - \symX''/3 + \left(\symX'' \symX''\right)\left(1/45-\symX''/472.5\right)
+ \symBigO(\symX^{8})
\end{equation}
In fact, $\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)\leq 0.5$ is almost always fulfilled as long as registers are not saturated. Using \eqref{equ:strong_upper_bound} it is straightforward to see that $\frac{\symX}{2^{\symRegVal'_\text{max}}} \leq \log 2 \approx 0.693$, if $\symCount_{\symRegRange+1}=0$.
In case $\frac{\symX}{2^{\symRegVal'_\text{max}}} > 0.5$, the value $\frac{\symX}{2^{\symE+1}}$ is taken instead, where $\symE$ is the exponent of the floating point representation of $\symX$, $\symE = 1+\lfloor\log_2(\symX)\rfloor$. By definition, $\frac{\symX}{2^{\symE+1}}\leq 0.5$ which allows using the Taylor approximation. $\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is finally obtained after $\symE + 1 -\symRegVal'_\text{max}$ iterations using \eqref{equ:helper_recursion2}.
\end{itemize}

All these optimizations together finally give the new cardinality estimation algorithm presented in \algoref{alg:estimate_ml}.

\begin{algorithm}
\caption{Cardinality estimation}
\ContinuedFloat
\label{alg:estimate_ml}
\begin{algorithmic}
\Function {EstimateCardinality}{$\vec{\symCount}$}
\State $\symRegRange\gets\dim(\vec{\symCount})-2$
\State $\symRegVal_\text{min} \gets \min\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$
\If{$\symRegVal_\text{min} > \symRegRange$}
\State\Return $\infty$
\EndIf
\State $\symRegVal'_\text{min} \gets \max(\symRegVal_\text{min}, 1)$
\State $\symRegVal_\text{max} \gets \max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$
\State $\symRegVal'_\text{max} \gets \min(\symRegVal_\text{max}, \symRegRange)$
\State $\symZ \gets 0$
\State $\symNumReg'\gets \symCount_{\symRegRange+1}$
\State $\symY \gets 2^{-\symRegVal'_\text{max}}$
\For{$\symRegVal \gets \symRegVal'_\text{max},\symRegVal'_\text{min}$}
\State $\symZ \gets \symZ + \symCount_\symRegVal\cdot\symY$
\Comment here $\symY = 2^{-\symRegVal}$
\State $\symY \gets 2\symY$
\State $\symNumReg'\gets\symNumReg' + \symCount_\symRegVal$
\EndFor
\State $\symNumReg \gets \symNumReg' + \symCount_0$ 
\State $\symCount' \gets \symCount_{\symRegRange+1}$
\If{$\symRegRange\geq 1$}
\State $\symCount' \gets \symCount' + \symCount_{\symRegVal'_\text{max}}$
\EndIf
\State $\symFuncPrime_\text{prev}\gets 0$
\State $\symA\gets \symZ + \symCount_0$
\State $\symB\gets \symZ + 
\symCount_{\symRegRange+1}\cdot 2^{-\symRegRange}$
\If{$\symB \leq 1.5\cdot\symA$}
\State $\symX \gets \symNumReg'/(0.5\cdot \symB+\symA)$ \Comment weak lower bound \eqref{equ:weak_lower_bound}
\Else
\State $\symX \gets \symNumReg'/b \cdot \log(1+\symB/\symA)$ \Comment strong lower bound \eqref{equ:strong_lower_bound}
\EndIf
\algstore{myalg}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Cardinality estimation (continued)}
\begin{algorithmic}
\algrestore{myalg}
\State $\Delta\symX \gets \symX$
\While{$\Delta\symX > \symX\cdot\symError$} \Comment secant method iteration, $\symStopEpsilon = 10^{-2}$
\State $\symE \gets 1+\lfloor\log_2(\symX)\rfloor$
\State $\symXNormalized \gets
\symX \cdot 2^{-\max(\symRegVal'_\text{max}+1, \symE+2)}$
\Comment $\symXNormalized \in [0, 0.25]$
\State $\symX''\gets \symXNormalized\cdot\symXNormalized$
\State $\symHelper \gets
\symX' - \symX''/3 + \left(\symX''\cdot \symX''\right)\cdot\left(1/45-\symX''/472.5\right)$
\Comment Taylor approximation \eqref{equ:taylor}
\For{$\symRegVal\gets \symE,\symRegVal'_\text{max}$}
\State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
\Comment calculate $\symHelper\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion2}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
\State $\symXNormalized \gets 2\symXNormalized$
\EndFor
\State $\symFuncPrime \gets \symCount'\cdot\symHelper$
\Comment compare \eqref{equ:funcprime}
\For{$\symRegVal\gets(\symRegVal'_\text{max}-1),\symRegVal'_\text{min}$}
\State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
\Comment calculate $\symHelper\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion2}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
\State $\symFuncPrime\gets \symFuncPrime + \symCount_\symRegVal\cdot\symHelper$
\State $\symXNormalized \gets 2\symXNormalized$
\EndFor
\State $\symFuncPrime\gets \symFuncPrime + \symX\cdot\symA$
\If{$\symFuncPrime > \symFuncPrime_\text{prev} \wedge \symNumReg' \geq\symFuncPrime$}
\State $\Delta\symX \gets \Delta\symX \cdot \frac{\symNumReg' - \symFuncPrime}{\symFuncPrime - \symFuncPrime_\text{prev}}$
\Comment see \eqref{equ:secant_delta}
\Else
\State $\Delta\symX \gets 0$
\EndIf
\State $\symX \gets \symX + \Delta\symX$
\State $\symFuncPrime_\text{prev}\gets \symFuncPrime$
\EndWhile
\State \Return $\symNumReg\cdot\symX$
\EndFunction
\end{algorithmic}
\end{algorithm}

\section{Results}

\subsection{Estimation error}
\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_8_24}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
\label{fig:estimation_error_8_24}
\end{figure}
\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_20}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:estimation_error_12_20}
\end{figure}
\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_14}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
\label{fig:estimation_error_12_14}
\end{figure}
\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_16_16}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
\label{fig:estimation_error_16_16}
\end{figure}
\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_20_12}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 20$ and $\symRegRange=12$.}
\label{fig:estimation_error_20_12}
\end{figure}
\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_52}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
\label{fig:estimation_error_12_52}
\end{figure}

\autoref{fig:estimation_error_8_24}
\autoref{fig:estimation_error_12_14}
\autoref{fig:estimation_error_12_20}
\autoref{fig:estimation_error_12_52}
\autoref{fig:estimation_error_16_16}
\autoref{fig:estimation_error_20_12}

\subsection{Performance}
\autoref{fig:avg_exec_time}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{avg_exec_time}
\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
\label{fig:avg_exec_time}
\end{figure}

\section{Cardinality estimation of set intersections and differences}
Assume two different sets $\symSetS_1$ and $\symSetS_2$ that have been recorded by two HyperLogLog sketches with same parameters $(\symPrecision,\symRegRange)$. Given the corresponding register values $\vec{\symRegVal}_1$ and
$\vec{\symRegVal}_2$ we attempt to estimate the cardinalities of the pair-wise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$. Motivated by the good results we have obtained for cardinality estimation of a single HyperLogLog sketch, we want to get these estimates using the maximum likelihood method applied to the joint probability distribution of $\vec{\symRegVal}_1$ and $\vec{\symRegVal}_2$.

Under the Poisson model the register values are independent and identically distributed. Therefore, we first consider a single register and derive the joint probability distribution for its values in both sketches.

The HyperLogLog sketch that represents $\symSetS_1$ could have also been obtained by constructing two HyperLogLog sketches from sets $\symSetA$ and $\symSetX$ and by merging both by taking for each register the maximum value of both sketches. Analogously, the HyperLogLog sketch for $\symSetS_2$ could have been obtained from sketches for $\symSetB$ and $\symSetX$. Let us consider the register values $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ at a certain position of the HyperLogLog sketches for $\symSetA$, $\symSetB$, and $\symSetX$, respectively. The corresponding values in sketches for $\symSetS_1$ and $\symSetS_2$ are given by
\begin{equation}
\symRegValVariate_1 = \max\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right)
,\quad
\symRegValVariate_2 = \max\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right)
\end{equation}
Their joint cumulative probability function is given as
\begin{align}
\symProbability\left(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
\right)
&=
\symProbability\left(
\max\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_1
\wedge
\max\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_2
\right)
\nonumber\\
&=
\symProbability
\left(
\symRegValVariate_\symSetASuffix \leq \symRegVal_1
\wedge
\symRegValVariate_\symSetBSuffix \leq \symRegVal_2
\wedge
\symRegValVariate_\symSetXSuffix \leq \min\left(\symRegVal_1, \symRegVal_2\right)
\right)
\nonumber\\
&=
\symProbability
\left(
\symRegValVariate_\symSetASuffix \leq \symRegVal_1
\right)
\symProbability
\left(
\symRegValVariate_\symSetBSuffix \leq \symRegVal_2
\right)
\symProbability
\left(
\symRegValVariate_\symSetXSuffix \leq \min\left(\symRegVal_1, \symRegVal_2\right)
\right)
\end{align}
Here the last transformation used the independence of $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$, because by definition, the sets $\symSetA$, $\symSetB$, and $\symSetX$ are disjoint. Furthermore, under the Poisson model $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ obey 
\eqref{equ:register_value_distribution}. If we assume that elements are added to $\symSetA$, $\symSetB$, and $\symSetX$ at rates $\symPoissonRate_\symSetXSuffix$, $\symPoissonRate_\symSetASuffix$, and $\symPoissonRate_\symSetBSuffix$, respectively, the probability that a certain register has a value less than or equal to $\symRegVal_1$ in the first HyperLogLog sketch and simultaneously a value less than or equal to $\symRegVal_2$ in the second one can be written as
\begin{equation}
\symProbability(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
)
=\begin{cases}
0 & \symRegVal_1 < 0 \vee \symRegVal_2 < 0
\\
e^{
-
\frac{\symPoissonRate_{\symSetASuffix}}{\symNumReg 2^{\symRegVal_1}}
-
\frac{\symPoissonRate_{\symSetBSuffix}}{\symNumReg 2^{\symRegVal_2}}
-
\frac{\symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\min(\symRegVal_1, \symRegVal_2)}}
}
& 0\leq\symRegVal_1 \leq \symRegRange \wedge 0\leq\symRegVal_2\leq\symRegRange
\\
e^{
-
\frac{\symPoissonRate_{\symSetBSuffix} + \symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\symRegVal_2}}
}
& 0\leq\symRegVal_2 \leq \symRegRange < \symRegVal_1
\\
e^{
-
\frac{\symPoissonRate_{\symSetASuffix} + \symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\symRegVal_1}}
}
&  0\leq\symRegVal_1 \leq \symRegRange < \symRegVal_2
\\
1
&
\symRegRange < \symRegVal_1=\symRegVal_2
\end{cases}
\end{equation}

The joint probability mass function for both register values can be calculated using
\begin{multline}
\symProbabilityMass(\symRegVal_1,\symRegVal_2)
=
\symProbability(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
)
-
\symProbability(
\symRegValVariate_1 \leq \symRegVal_1-1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
)
\\
-\symProbability(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2-1
)
+\symProbability(
\symRegValVariate_1 \leq \symRegVal_1-1
\wedge
\symRegValVariate_2 \leq \symRegVal_2-1
)
\end{multline}
which finally gives
\begin{multline}
\label{equ:register_value_joint_pmf}
\symProbabilityMass(\symRegVal_1,\symRegVal_2)
=
\\
\begin{cases}
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg }
-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\symRegVal_2}}
}
\left(
1-
e^{
-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\symRegVal_2}}
}
\right)
&
0 = \symRegVal_1 < \symRegVal_2 \leq \symRegRange
\\
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg}
}
\left(
1-
e^{
-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\symRegRange}}
}
\right)
&
0 = \symRegVal_1 < \symRegVal_2 = \symRegRange + 1
\\
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^{\symRegVal_1}}
-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\symRegVal_2}}
}
\left(
1-
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_1}}
}
\right)
\left(
1-
e^{
-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\symRegVal_2}}
}
\right)
&
1 \leq \symRegVal_1 < \symRegVal_2 \leq \symRegRange
\\
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^{\symRegVal_1}}
}
\left(
1-
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_1}}
}
\right)
\left(
1-
e^{
-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\symRegRange}}
}
\right)
&
1 \leq \symRegVal_1 < \symRegVal_2 = \symRegRange + 1
\\
e^{-\frac{\symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg }
-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\symRegVal_1}}
}
\left(
1-
e^{
-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\symRegVal_1}}
}
\right)
&
0 = \symRegVal_2 < \symRegVal_1 \leq \symRegRange
\\
e^{-\frac{\symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg}
}
\left(
1-
e^{
-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\symRegRange}}
}
\right)
&
0 = \symRegVal_2 < \symRegVal_1 = \symRegRange + 1
\\
e^{-\frac{\symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^{\symRegVal_2}}
-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\symRegVal_1}}
}
\left(
1-
e^{-\frac{\symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_2}}
}
\right)
\left(
1-
e^{
-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\symRegVal_1}}
}
\right)
&
1 \leq \symRegVal_2 < \symRegVal_1 \leq \symRegRange
\\
e^{-\frac{\symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^{\symRegVal_2}}
}
\left(
1-
e^{-\frac{\symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_2}}
}
\right)
\left(
1-
e^{
-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\symRegRange}}
}
\right)
&
1 \leq \symRegVal_2 < \symRegVal_1 = \symRegRange + 1
\\
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg}
}
&
0 = \symRegVal_1 = \symRegVal_2
\\
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
\left(
1
-
e^{-\frac{\symPoissonRate_\symSetASuffix +  \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
-
e^{-\frac{\symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
+
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
\right)
&
1 \leq \symRegVal_1 = \symRegVal_2 = \symRegVal\leq \symRegRange
\\
1
-
e^{-\frac{\symPoissonRate_\symSetASuffix +  \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegRange}
}
-
e^{-\frac{\symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegRange}
}
+
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegRange}
}
&
\symRegVal_1 = \symRegVal_2 = \symRegRange + 1
\end{cases}
\end{multline}

The logarithm of the joint probability mass function can be written using Iverson bracket notation ($\left[\text{true}\right]:=1$, $\left[\text{false}\right]:=0$) as
\begin{align}
\label{equ:joint_log_pmf_single_register}
\log(\symProbabilityMass(\symRegVal_1,\symRegVal_2))
=&
-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\symRegVal_1}}
\left[\symRegVal_1\leq\symRegRange\right]
-
\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\symRegVal_2}}
\left[\symRegVal_2\leq\symRegRange\right]
-
\frac{\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1,\symRegVal_2\right)}}
\left[\symRegVal_1\leq\symRegRange\vee\symRegVal_2\leq\symRegRange\right]
\nonumber\\
&
+
\log\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_1}}}\right)
\left[1\leq\symRegVal_1<\symRegVal_2\right]
\nonumber\\
&
+
\log\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}\right)
\left[\symRegVal_2<\symRegVal_1\right]
\nonumber\\
&
+
\log\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_2}}}\right)
\left[1\leq\symRegVal_2<\symRegVal_1\right]
\nonumber\\
&
+
\log\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal_2, \symRegRange\right)}}}\right)
\left[\symRegVal_1<\symRegVal_2\right]
\nonumber\\
&
+
\log\left(
1
-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
-
e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
+
e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
\right)
\left[1\leq\symRegVal_1=\symRegVal_2\right]
\end{align}

Since under the Poisson model, the values for different registers are independent we are now able to write the joint probability mass function for the joint state of both HyperLogLog sketches.
\begin{equation}
\symProbabilityMass(\vec{\symRegVal}_1,\vec{\symRegVal}_2)
=
\prod_{\symRegVal_1 = 0}^{\symRegRange+1}
\prod_{\symRegVal_2 = 0}^{\symRegRange+1}
\symProbabilityMass(\symRegVal_1,\symRegVal_2)
^{\symCount_{\symRegVal_1\symRegVal_2}}
\end{equation}
Here we have used the multiplicity matrix $\symCountMatrix = \left(\symCount_{\symRegVal_1\symRegVal_2}\right)_{\symRegVal_1\symRegVal_2\in[0,\symRegRange+1]}$ defined by
\begin{equation}
\symCount_{\symRegVal_1\symRegVal_2}
:=
\left|\left\lbrace
\left(\symIndexI,\symIndexJ\right) \vert
\symRegVal_{\symIndexI,1} = \symRegVal_1
\wedge
\symRegVal_{\symIndexJ,2} = \symRegVal_2
\right\rbrace\right|
\end{equation}
As in \autoref{sec:cardinality_estimation} where we found that the multiplicity vector $\vec{\symCount}$ of a sketch is a sufficient statistic for the cardinality, the multiplicity matrix of two sketches is a sufficient statistic for the cardinalities of $\symSetX$, $\symSetA$, and $\symSetB$.

In order to get the maximum likelihood estimates $\symPoissonRateEstimate_\symSetASuffix$,
 $\symPoissonRateEstimate_\symSetBSuffix$, and  $\symPoissonRateEstimate_\symSetXSuffix$ we need to maximize the log-likelihood function given by
\begin{equation}
\log \symLikelihood(
\symPoissonRate_\symSetASuffix,
\symPoissonRate_\symSetBSuffix,
\symPoissonRate_\symSetXSuffix
\vert
\vec{\symRegVal}_1,
\vec{\symRegVal}_2
)
=
\sum_{\symRegVal_1 = 0}^{\symRegRange+1}
\sum_{\symRegVal_2 = 0}^{\symRegRange+1}
\symCount_{\symRegVal_1\symRegVal_2}
\log(\symProbabilityMass(\symRegVal_1,\symRegVal_2))
\end{equation}
Insertion of \eqref{equ:joint_log_pmf_single_register} results in
\begin{align}
\label{equ:log_likelihood_pair}
\log \symLikelihood(
\symPoissonRate_\symSetASuffix,
\symPoissonRate_\symSetBSuffix,
\symPoissonRate_\symSetXSuffix
\vert
\vec{\symRegVal}_1,
\vec{\symRegVal}_2
)
=
&
-
\frac{\symPoissonRate_\symSetASuffix}{\symNumReg}
\sum_{\symRegVal=0}^{\symRegRange}
\frac{
  \symCount^\leftarrow_\symRegVal+
  \symCount_{\symRegVal\symRegVal}+
  \symCount^\rightarrow_\symRegVal
}{2^{\symRegVal}}
-
\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg}
\sum_{\symRegVal=0}^{\symRegRange}
\frac{
  \symCount^\uparrow_\symRegVal+
  \symCount_{\symRegVal\symRegVal}+
  \symCount^\downarrow_\symRegVal
}{2^{\symRegVal}}
\nonumber\\
&
-
\frac{\symPoissonRate_\symSetXSuffix}{\symNumReg}
\sum_{\symRegVal=0}^{\symRegRange}
\frac{
  \symCount^\rightarrow_\symRegVal+
  \symCount_{\symRegVal\symRegVal}+
  \symCount^\downarrow_\symRegVal
}{2^{\symRegVal}}
\nonumber\\
&
+
\sum_{\symRegVal=1}^{\symRegRange+1}
\log\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
\symCount_\symRegVal^\rightarrow
+
\log\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
\symCount_\symRegVal^\leftarrow
\nonumber\\
&
+
\sum_{\symRegVal=1}^{\symRegRange+1}
\log\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
\symCount_\symRegVal^\downarrow
+
\log\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
\symCount_\symRegVal^\uparrow
\nonumber\\
&
+
\sum_{\symRegVal=1}^{\symRegRange+1}
\log\left(
1
-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
-
e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
+
e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
\right)
\symCount_{\symRegVal\symRegVal}
\end{align}
which is the two HyperLogLog case analog of \eqref{equ:log_likelihood_single}.
The constants $\symCount^\uparrow_\symRegVal:=\sum_{\symIndexI=0}^{\symRegVal-1} \symCount_{\symIndexI\symRegVal}$, $\symCount^\rightarrow_\symRegVal:=\sum_{\symIndexI=\symRegVal+1}^{\symRegRange+1} \symCount_{\symRegVal\symIndexI}$, $\symCount^\downarrow_\symRegVal:=\sum_{\symIndexI=\symRegVal+1}^{\symRegRange+1} \symCount_{\symIndexI\symRegVal}$, and $\symCount^\leftarrow_\symRegVal:=\sum_{\symIndexI=0}^{\symRegVal-1} \symCount_{\symRegVal\symIndexI}$ correspond to sums within the multiplicity matrix, which are obtained by aggregating all elements that are in up, right, down, and left directions relative to the $\symRegVal$-th diagonal entry  $\symCount_{\symRegVal\symRegVal}$, respectively.

The log-likelihood function \eqref{equ:log_likelihood_pair} does not always have a strict global maximum point. For example, in case $\symCountMatrix$ is a strict lower triangular matrix which corresponds to the case that each register of the first HyperLogLog sketch is larger than the corresponding value in the second HyperLogLog sketch, the function can be rewritten as sum of two functions, one dependent on $\symPoissonRate_\symSetASuffix$ and the other dependent on $(\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix)$. 
The maximum is obtained, if $\symPoissonRate_\symSetASuffix = \symPoissonRateEstimate_1$ and $\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix = \symPoissonRateEstimate_2$. Here $\symPoissonRateEstimate_1$ and $\symPoissonRateEstimate_2$ are the cardinality estimates for the first and second HyperLogLog sketch, respectively. Similarly, if all register values of the first are smaller than those of the second HyperLogLog sketch, the maximum is obtained when 
$\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix = \symPoissonRateEstimate_1$ and $\symPoissonRate_\symSetBSuffix = \symPoissonRateEstimate_2$.

If we know that the two HyperLogLog sketches have been filled by elements taken from disjoint sets we can assume $\symPoissonRate_\symSetXSuffix=0$. In this case \eqref{equ:log_likelihood_pair} is separable into the sum of two log-likelihood functions that depend on $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$, respectively, and that follow \eqref{equ:log_likelihood_single}. Hence, the joint maximum likelihood estimation of $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$ gives the same results as estimating them independently by maximizing \eqref{equ:log_likelihood_single}.

The first derivatives are
\begin{align}
\frac{\partial\log \symLikelihood}{\partial\symPoissonRate_\symSetASuffix}
(
\symPoissonRate_\symSetASuffix,
\symPoissonRate_\symSetBSuffix,
\symPoissonRate_\symSetXSuffix
\vert
\vec{\symRegVal}_1,
\vec{\symRegVal}_2
)
=
&
-
\frac{1}{\symNumReg}
\sum_{\symRegVal=0}^{\symRegRange}
\frac{
  \symCount^\leftarrow_\symRegVal+
  \symCount_{\symRegVal\symRegVal}+
  \symCount^\rightarrow_\symRegVal
}{2^{\symRegVal}}
+
\frac{1}{\symNumReg}
\sum_{\symRegVal=1}^{\symRegRange+1}
\frac{1}{e^{\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}-1}
\frac{\symCount^\rightarrow_\symRegVal}{2^{\min\left(\symRegVal,\symRegRange\right)}}
\nonumber\\
&
+
\frac{1}{\symNumReg}\sum_{\symRegVal=1}^{\symRegRange+1}
\frac{1}{e^{\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}-1}
\frac{\symCount^\leftarrow_\symRegVal}{2^{\min\left(\symRegVal,\symRegRange\right)}}
\nonumber\\
&
+
\frac{1}{\symNumReg}
\sum_{\symRegVal=1}^{\symRegRange+1}
\frac{
e^{\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
-
1
}{
e^{\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
-
e^{\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
-
e^{\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
+
1
}
\frac{\symCount_{\symRegVal\symRegVal}
}{2^{\min\left(\symRegVal,\symRegRange\right)}}
\end{align}

\begin{align}
\frac{\partial\log \symLikelihood}{\partial\symPoissonRate_\symSetXSuffix}
(
\symPoissonRate_\symSetASuffix,
\symPoissonRate_\symSetBSuffix,
\symPoissonRate_\symSetXSuffix
\vert
\vec{\symRegVal}_1,
\vec{\symRegVal}_2
)
=
&
-
\frac{1}{\symNumReg}
\sum_{\symRegVal=0}^{\symRegRange}
\frac{
  \symCount^\rightarrow_\symRegVal+
  \symCount_{\symRegVal\symRegVal}+
  \symCount^\downarrow_\symRegVal
}{2^{\symRegVal}}
+
\frac{1}{\symNumReg}
\sum_{\symRegVal=1}^{\symRegRange+1}
\frac{1}{e^{\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}-1}
\frac{\symCount^\rightarrow_\symRegVal}{2^{\min\left(\symRegVal,\symRegRange\right)}}
\nonumber\\
&
+
\frac{1}{\symNumReg}\sum_{\symRegVal=1}^{\symRegRange+1}
\frac{1}{e^{\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}-1}
\frac{\symCount^\downarrow_\symRegVal}{2^{\min\left(\symRegVal,\symRegRange\right)}}
\nonumber\\
&
+
\frac{1}{\symNumReg}
\sum_{\symRegVal=1}^{\symRegRange+1}
\frac{
e^{\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
+
e^{\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
-
1
}{
e^{\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
-
e^{\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
-
e^{\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
+
1
}
\frac{\symCount_{\symRegVal\symRegVal}
}{2^{\min\left(\symRegVal,\symRegRange\right)}}
\end{align}

\section{Conclusion and future work}

\appendix

\section{Numerical stability of recursion formula for $\symHelper(\symX)$}
\label{app:helper_stable}
In order to investigate the error propagation of a single recursion step using \eqref{equ:helper_recursion1} we define $\symHelper_1 := \symHelper(\symX)$ and $\symHelper_2 := \symHelper(2\symX)$. The recursion formula simplifies to
\begin{equation}
\label{equ:h2}
\symHelper_2 = \frac{\symX+2\symHelper_1(1-\symHelper_1)}{\symX+2(1-\symHelper_1)}.
\end{equation}
If $\symHelper_1$ is approximated by $\tilde{\symHelper}_1 = \symHelper_1\left(1+\symError_1\right)$ with relative error $\symError_1$, the recursion formula will give an approximation for $\symHelper_2$
\begin{equation}
\label{equ:h2_approx}
\tilde{\symHelper}_2 = 
\frac{\symX+2\tilde{\symHelper}_1(1-\tilde{\symHelper}_1)}{\symX+2(1-\tilde{\symHelper_1})}
\end{equation}
The corresponding relative error $\symError_2$ is given by
\begin{equation}
\label{equ:h2_relative_error}
\symError_2 = \frac{\tilde{\symHelper}_2}{\symHelper_2}-1
\end{equation}
Putting \eqref{equ:h2} and \eqref{equ:h2_approx} into \eqref{equ:h2_relative_error} and using first-order approximations 
\begin{equation}
\frac{\symX+2\tilde{\symHelper}_1(1-\tilde{\symHelper}_1)}
{\symX+2\symHelper_1(1-\symHelper_1)}
=
1
+
\symError_1
\frac{2\symHelper_1\left(1-2\symHelper_1\right)}
{\symX + 2\symHelper_1\left(1-\symHelper_1\right)}
+\symBigO(\symError_1^2)
\end{equation}
and
\begin{equation}
\frac{\symX+2(1-\symHelper_1)}
{\symX+2(1-\tilde{\symHelper}_1)}
=
1
+
\symError_1
\frac{2\symHelper_1}{x+2\left(1-\symHelper_1\right)}
+\symBigO(\symError_1^2)
\end{equation}
we obtain
\begin{equation}
\symError_2
=
\symError_1
\left(
\frac{2\symHelper_1\left(1-2\symHelper_1\right)}
{\symX + 2\symHelper_1\left(1-\symHelper_1\right)}
+
\frac{2\symHelper_1}{x+2\left(1-\symHelper_1\right)}
\right)
+\symBigO(\symError_1^2)
\end{equation}
By numerical means it can be easily shown that
\begin{equation}
0\leq\frac{2\symHelper(\symX)\left(1-2\symHelper(\symX)\right)}
{\symX + 2\symHelper(\symX)\left(1-\symHelper(\symX)\right)}
+
\frac{2\symHelper(\symX)}{x+2\left(1-\symHelper(\symX)\right)} \leq \input{max_errror_propagation_factor.txt}
\end{equation}
holds for all $\symX \geq 0$. Therefore,
\begin{equation}
|\symError_2|\leq \input{max_errror_propagation_factor.txt}\  |\symError_1| + \symBigO(\symError_1^2)
\end{equation}
which means that the relative error is decreasing in each recursion step and the recursive calculation of $\symHelper$ is numerically stable.

\section{Error caused by approximation of $\symHelper(\symX)$}
According to \eqref{equ:func} the exact estimate $\symXEstimate$ fulfills 
\begin{equation}
\label{equ:max_likelihood_2}
\symXEstimate\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\left(\frac{\symXEstimate}{2^\symRegVal}\right)
+
\symCount_{\symRegRange+1}\symHelper\left(\frac{\symXEstimate}{2^\symRegRange}\right)
=
\symNumReg-\symCount_0
\end{equation}
If $\symHelper$ is not calculated exactly but approximated by an approximation $\symHelperApprox$ with maximum relative error $\symError_\symHelper\ll 1$
\begin{equation}
\label{equ:abs_error_helper}
\left|\symHelperApprox(\symX) - \symHelper(\symX) \right|  \leq \symError_\symHelper\symHelper(\symX)
\end{equation}
the solution of the equation will be off by a relative error $\symError_\symX$:
\begin{equation}
\label{equ:ml_equation}
\symXEstimate\left(1+\symError_\symX\right)\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelperApprox\left(\frac{\symXEstimate\left(1+\symError_\symX\right)}{2^\symRegVal}\right)
+
\symCount_{\symRegRange+1}
\symHelperApprox\left(
\frac{\symXEstimate\left(1+\symError_\symX\right)}{2^\symRegRange}\right)
=
\symNumReg-\symCount_0
\end{equation}
Due to \eqref{equ:abs_error_helper} there exists some $\symAlpha \in [-\symError_\symHelper, \symError_\symHelper]$ for which
\begin{multline}
\symXEstimate\left(1+\symError_\symX\right)\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\left(
1+\symAlpha
\right)
\left(
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal
\symHelper\left(\frac{\symXEstimate\left(1+\symError_\symX\right)
}{2^\symRegVal}\right)
+
\symCount_{\symRegRange+1}
\symHelper\left(
\frac{\symXEstimate\left(1+\symError_\symX\right)}{2^\symRegRange}\right)
\right)
=
\symNumReg-\symCount_0
\end{multline}
For $\symX \geq 0$ $\symHelper'(\symX)\in[0,0.5]$. Hence, there exists a $\symBeta\in[0,0.5]$ for which
\begin{multline}
\label{equ:appendix3}
\symXEstimate\left(1+\symError_\symX\right)\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\left(
1+\symAlpha
\right)
\left(
\sum_{\symRegVal=1}^\symRegRange 
\symCount_\symRegVal
\symHelper\left(\frac{\symXEstimate
}{2^\symRegVal}\right)
+
\frac{\symCount_\symRegVal}{2^\symRegVal}
\symXEstimate\symError_\symX
\symBeta
\right)
+
\\
+
\left(
1+\symAlpha
\right)
\left(
\symCount_{\symRegRange+1}
\symHelper\left(
\frac{\symXEstimate}{2^\symRegRange}\right)
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
\symXEstimate\symError_\symX\symBeta
\right)
=
\symNumReg-\symCount_0
\end{multline}
Subtracting \eqref{equ:max_likelihood_2} multiplied by $\left(1+\symAlpha\right)$ from \eqref{equ:appendix3} and resolving 
$\symError_\symX$ gives
\begin{equation}
\symError_\symX
=
\symAlpha
\frac{
\symXEstimate
\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
-\left(\symNumReg-\symCount_0\right)
}
{
\symXEstimate
\left(
\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
+
\left(1+\symAlpha\right)
\symBeta
\left(
\sum_{\symRegVal=1}^\symRegRange 
\frac{\symCount_\symRegVal
}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}
}{2^\symRegRange}
\right)
\right)
}
\end{equation}
Using $\left|\symAlpha\right|\leq\symError_\symHelper$, $\symBeta\geq0$, and \eqref{equ:weak_upper_bound} the absolute value of the relative error
can be bounded by
\begin{equation}
\left|\symError_\symX\right| 
\leq
\left|\symError_\symHelper\right| 
\frac{
\left(\symNumReg-\symCount_0\right)-
\symXEstimate
\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
}
{
\symXEstimate
\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
}
\end{equation}
Furthermore, using \eqref{equ:weak_lower_bound} we finally get
\begin{equation}
\left|\symError_\symX\right| 
\leq
\left|\symError_\symHelper\right|
\frac{
\frac{1}{2}
\sum_{\symRegVal=1}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}
}{2^\symRegRange}
}
{
\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
}
\leq
\left|\symError_\symHelper\right|
\left(
\frac{1}{2}
+
\frac{
\frac{\symCount_{\symRegRange+1}
}{2^\symRegRange}
}
{
\sum_{\symRegVal=1}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
}
\right)
\leq
\left|\symError_\symHelper\right|
\left(
\frac{1}{2}
+
\frac{
\symCount_{\symRegRange+1}
}
{
\symNumReg-\symCount_{\symRegRange+1}
}
\right)
\end{equation}
Hence, as long as most registers are not in the saturated state ($\symCount_{\symRegRange+1}\ll\symNumReg$), the relative error $\symError_\symX$ of the calculated estimate using the approximation $\symHelperApprox(\symX)$ for $\symHelper(\symX)$ has the same order of magnitude as $\symError_\symHelper$.

% TODO is it possible to prove 
%$\left|\symError_\symX\right| 
%\leq
%\left|\symError_\symHelper\right|
%\frac{1}{2}
%\frac{
%\symNumReg
%}
%{
%\symNumReg-\symCount_{\symRegRange+1}
%}
%$}

\bibliographystyle{unsrt}
\bibliography{bibliography.bib}

\end{document}

### 2016-04-16 c++/benchmark.cpp

@@ -1,0 +1,218 @@
+//#############################
+//# Copyright 2016 Otmar Ertl #
+//#############################
+
+#include "cardinalityestimation.hpp"
+#include "hyperloglog.hpp"
+
+#include <iostream>
+#include <fstream>
+#include <iostream>
+#include <algorithm>
+#include <string>
+#include <iomanip>
+#include <cassert>
+#include <cstdlib>
+#include <sstream>
+#include <memory>
+#include <chrono>
+
+using namespace std;
+
+const int minP = 4;
+const int maxP = 22;
+const int maxPplusQ = 64;
+
+
+const int numLoops = 100;
+
+const int numData = 1000;
+
+const string seedFileName = "../data/hll/seed.dat";
+const string cardFileName = "../data/cardinalities.dat";
+const string dataPathName = "../data/hll/";
+
+const int p1 = 12;
+const int q1 = 52;
+const string resultsFileName1 = "../data/benchmark_results_12_52.dat";
+const int p2 = 12;
+const int q2 = 20;
+const string resultsFileName2 = "../data/benchmark_results_12_20.dat";
+
+vector<int> readCounts(string s) {
+	stringstream stream(s);
+	int i;
+	vector<int> v;
+	while(stream >> i) {
+		v.push_back(i);
+	}
+	v.shrink_to_fit();
+	return v;
+}
+
+string getFileName(const string& seed, int p) {
+	stringstream s;
+	s << dataPathName;
+	s << seed;
+	s << "_" << setfill('0') << setw(2) << dec << p;
+	s << ".dat";
+	return s.str();
+}
+
+void run(const int p, const int q, const string& resultsFileName) {
+	
+	
+	mt19937_64 rng(0);
+
+	vector<string> seeds;
+	{
+		ifstream seedFile(seedFileName);
+		
+		string line;
+		while (getline(seedFile, line))
+		{
+			seeds.push_back(line);
+		}
+	}
+	
+	cout << "seedSize "  << seeds.size() << endl;
+	
+	assert(numData <= seeds.size());
+	
+	vector<long> cardinalities;
+	{
+		ifstream cardFile(cardFileName);
+		
+		string line;
+		while (getline(cardFile, line))
+		{
+			cardinalities.push_back(stol(line));
+		}
+	}
+	
+	cout << "cardinalitySize "  << cardinalities.size() << endl;
+	
+	
[…]

### 2016-04-16 c++/cardinalityestimation.hpp

@@ -1,0 +1,603 @@
+//#############################
+//# Copyright 2016 Otmar Ertl #
+//#############################
+
+#ifndef _CARDINALITY_ESTIMATION_HPP_
+#define _CARDINALITY_ESTIMATION_HPP_
+
+#include <vector>
+#include <cassert>
+#include <cmath>
+#include <limits>
+#include <algorithm>
+#include <iostream>
+#include <cstdlib>
+
+#include <gsl/gsl_vector.h>
+#include <gsl/gsl_multimin.h>
+
+
+int getPFromNumberRegisters(int m) {
+	int p;
+	std::frexp(m, &p);
+	p -= 1;
+	assert(m == size_t(1) << p);
+	return p;
+}
+
+int getPFromCounts(const std::vector<int>& c) {
+	int m = std::accumulate(c.begin(), c.end(), 0);
+	return getPFromNumberRegisters(m);
+}
+
+double maxLikelihoodEstimate(const std::vector<int>& c, int& outerLoopIterationsCount, int& innerLoop1IterationsCount, int& innerLoop2IterationsCount, int& logEvaluationCount) {
+
+	const double eps = 1e-2;
+	
+	outerLoopIterationsCount = 0;
+	innerLoop1IterationsCount = 0;
+	innerLoop2IterationsCount = 0;
+	logEvaluationCount = 0;
+	
+	int q = c.size() - 2;
+
+	int kMin;
+	for(kMin=0; c[kMin]==0; ++kMin);
+	int kMinPrime = std::max(1, kMin);
+	
+	int kMax;
+	for(kMax=q+1; c[kMax]==0; --kMax);
+	int kMaxPrime = std::min(q, kMax);
+
+	if (kMin > q) return std::numeric_limits<double>::infinity();
+	
+	double z = 0.;
+	int mPrime = c[q+1];
+	double y = ldexp(1., -kMaxPrime);
+	for (int k = kMaxPrime; k >= kMinPrime; --k) {
+		z += y*c[k];
+		y += y;
+		mPrime += c[k];
+	}
+	int m = mPrime + c[0];
+	
+	int cPrime = c[q+1];
+	if (q >= 1) {
+		cPrime += c[kMaxPrime];
+	}
+
+    double gPrevious = 0;
+    
+    double x;
+	double a = z + c[0];
+    {
+		double b = z + ldexp(c[q+1], -q);
+		if (b <= 1.5*a) { // TODO constant 1.5
[…]

### 2016-04-16 c++/definecardinalities.cpp

@@ -1,0 +1,30 @@
+//#############################
+//# Copyright 2016 Otmar Ertl #
+//#############################
+
+#include <iostream>
+#include <algorithm>
+
+using namespace std;
+
+const double factor = 1.01;
+const long maxCardinality = 50000000000;
+
+int main(int argc, char* argv[])
+{
+	vector<long> cardinalities;
+	
+	long cardinality = maxCardinality;
+	while(cardinality >= 0) {
+		cardinalities.push_back(cardinality);
+		cardinality = min(lround(cardinality/factor), cardinality-1);
+	}
+	
+	sort(cardinalities.begin(), cardinalities.end());
+	auto last = std::unique(cardinalities.begin(), cardinalities.end());
+	cardinalities.erase(last, cardinalities.end()); 
+	
+	for(long c : cardinalities) {
+		cout << c << endl;
+	}
+}

### 2016-04-16 c++/hyperloglog.hpp

@@ -1,0 +1,209 @@
+//#############################
+//# Copyright 2016 Otmar Ertl #
+//#############################
+
+#ifndef _HYPERLOGLOG_HPP_
+#define _HYPERLOGLOG_HPP_
+
+#include <vector>
+#include <cstdint>
+#include <cassert>
+#include <algorithm>
+#include <random>
+#include <iostream>
+
+class HyperLogLog {
+	
+	unsigned char p;
+	unsigned char q;
+	
+	std::vector<unsigned char> registers;
+	
+	std::vector<int> counts;
+	
+	unsigned char minCount;
+	uint_fast64_t registerValueFilter;
+	
+	HyperLogLog(unsigned char p_, unsigned char q_, std::vector<unsigned char> registers_) : p(p_), q(q_), registers(registers_), counts(q_+2), minCount(0) {
+		assert((size_t(1) << p_ ) == registers.size());
+		for (unsigned char r : registers) {
+			assert(r < q+2);
+			counts[r] += 1;
+		}
+		while(counts[minCount]>0) {
+			minCount += 1;
+		}
+		registerValueFilter = (~uint_fast64_t(0)) << minCount;
+	} 
+
+public:
+	
+	HyperLogLog(unsigned char p_, unsigned char q_) : p(p_), q(q_), registers(std::size_t(1) << p_), counts(q_+2), minCount(0), registerValueFilter(~uint_fast64_t(0)) {
+		counts[0] = registers.size();
+	}
+
+	static HyperLogLog createFromCounts(const std::vector<int>& counts, std::mt19937_64& rng) {
+		
+		assert(counts.size() >= 2);
+		unsigned char q = counts.size() - 2;
+		size_t sum = 0;
+		for (int c : counts) {
+			sum += c;
+		}
+		int p;
+		std::frexp(sum, &p);
+		p -= 1;
+		assert(sum == size_t(1) << p);
+
+		std::vector<unsigned char> registers(sum);
+		
+		size_t idx = 0;
+		for (size_t j = 0; j < counts.size(); ++j) {
+			for (int k = 0; k < counts[j]; ++k) {
+				registers[idx] = j;
+				idx += 1;
+			}
+		}
+		
+		std::shuffle(registers.begin(), registers.end(), rng);
+		
+		return HyperLogLog(p, q, registers);
+	}
+	
+	static HyperLogLog merge(const HyperLogLog& hll1, const HyperLogLog& hll2) {
[…]

### 2016-04-16 c++/hyperloglog_data_generator.cpp

@@ -1,0 +1,106 @@
+//#############################
+//# Copyright 2016 Otmar Ertl #
+//#############################
+
+#include "hyperloglog.hpp"
+
+#include <fstream>
+#include <random>
+#include <iostream>
+#include <algorithm>
+#include <string>
+#include <iomanip>
+#include <chrono>
+#include <sstream>
+#include <array>
+#include <memory>
+
+using namespace std;
+
+const int maxP = 22;
+const int minP = 4;
+
+const string outputFolder = "../data/hll/";
+const string seedFileName = "../data/hll/seed.dat";
+const string cardFileName = "../data/cardinalities.dat";
+
+string getFileName(const uint_fast64_t seed, int q) {
+	stringstream s;
+	s << outputFolder;
+	s << hex <<  setfill('0') << setw(16) << seed;
+	s << "_" << setw(2) << dec << q;
+	s << ".dat";
+	return s.str();
+}
+
+int main(int argc, char* argv[])
+{
+	// read cardinalities
+	ifstream cardFile(cardFileName);
+	vector<long> cardinalities;
+	string line;
+	while (getline(cardFile, line))
+	{
+		cardinalities.push_back(stol(line));
+	}
+	
+	if (cardinalities.empty()) {
+		return 1;
+	}
+	
+	mt19937_64 rng(chrono::system_clock::now().time_since_epoch().count());
+	
+	#pragma omp parallel
+	while(true) {
+		
+		uint_fast64_t seed;
+		#pragma omp critical
+		{
+			seed = rng();
+		}
+		
+		mt19937_64 rng2(seed);
+		
+		long lastCardinality = 0;
+	
+		HyperLogLog hll(maxP, 64-maxP);
+		
+		vector<vector<vector<int>>> data(maxP-minP+1);
+		
+		for(size_t cardinalityCounter = 0; cardinalityCounter < cardinalities.size(); cardinalityCounter+=1) {
+			long cardinality = cardinalities[cardinalityCounter];
+			for (long e = lastCardinality; e < cardinality; ++e) {
+				hll.add(rng2());
+			}
+			HyperLogLog hll2 = hll;
+			data[maxP-minP].push_back(hll2.getCounts());
+			for (int p = maxP-1; p >= minP; --p) {
+				hll2 = hll2.reduce(p, 64-p);
+				data[p-minP].push_back(hll2.getCounts());
+			}
+			
+			lastCardinality = cardinality;
+		}
+		
[…]

### 2016-04-16 c++/intersection_calculation.cpp

@@ -1,0 +1,287 @@
+//#############################
+//# Copyright 2016 Otmar Ertl #
+//#############################
+
+#include "cardinalityestimation.hpp"
+#include "hyperloglog.hpp"
+
+#include <iostream>
+#include <fstream>
+#include <iostream>
+#include <algorithm>
+#include <string>
+#include <iomanip>
+#include <cassert>
+#include <cstdlib>
+#include <sstream>
+#include <gsl/gsl_statistics.h>
+
+using namespace std;
+
+const int minP = 4;
+const int maxP = 22;
+const int maxPplusQ = 64;
+
+const double pm3s = (1.-erf(3./sqrt(2.)))/2.;
+const double pp3s = (1.+erf(3./sqrt(2.)))/2.;
+const double pm2s = (1.-erf(2./sqrt(2.)))/2.;
+const double pp2s = (1.+erf(2./sqrt(2.)))/2.;
+const double pm1s = (1.-erf(1./sqrt(2.)))/2.;
+const double pp1s = (1.+erf(1./sqrt(2.)))/2.;
+const double median = 0.5;
+
+
+const string intersectDataPathName = "../data/hll_joint/";
+const string jointCardFileName = intersectDataPathName + "joint_cardinalities.dat";
+const string resultsFileName = intersectDataPathName + "results.dat";
+
+
+template <typename T> void printToFile(const string& fileName, const T* data, size_t numSketches, bool append) {
+	ofstream file(fileName, (append)?(ofstream::out | ofstream::app):ofstream::out);
+	for (size_t s = 0; s < numSketches; ++s) {
+		if (s>0) {
+			file << " ";
+		}
+		file << data[s];
+	}
+	file << endl;
+}
+
+vector<int> readCounts(string s) {
+	stringstream stream(s);
+	int i;
+	vector<int> v;
+	while(stream >> i) {
+		v.push_back(i);
+	}
+	v.shrink_to_fit();
+	return v;
+}
+
+int main(int argc, char* argv[])
+{
+	
+	vector<string> jointCardFileNames;
+	{
+		ifstream jointCardFileNamesFile(jointCardFileName);
+		
+		string line;
+		while (getline(jointCardFileNamesFile, line))
+		{
[…]

### 2016-04-16 c++/joint_hyperloglog_data_generator.cpp

@@ -1,0 +1,208 @@
+//#############################
+//# Copyright 2016 Otmar Ertl #
+//#############################
+
+#include "cardinalityestimation.hpp"
+#include "hyperloglog.hpp"
+
+#include <iostream>
+#include <fstream>
+#include <iostream>
+#include <algorithm>
+#include <string>
+#include <iomanip>
+#include <cassert>
+#include <cstdlib>
+#include <sstream>
+#include <functional>
+#include <chrono>
+
+using namespace std;
+
+const int minP = 4;
+const int maxP = 22;
+const int maxPplusQ = 64;
+
+const string seedFileName = "../data/hll/seed.dat";
+const string cardFileName = "../data/cardinalities.dat";
+const string dataPathName = "../data/hll/";
+const string intersectDataPathName = "../data/hll_joint/";
+const string jointCardFileName = intersectDataPathName + "joint_cardinalities.dat";
+
+template <typename T> void printToFile(const string& fileName, const T* data, size_t numSketches, bool append) {
+	ofstream file(fileName, (append)?(ofstream::out | ofstream::app):ofstream::out);
+	for (size_t s = 0; s < numSketches; ++s) {
+		if (s>0) {
+			file << " ";
+		}
+		file << data[s];
+	}
+	file << endl;
+}
+
+vector<int> readCounts(string s) {
+	stringstream stream(s);
+	int i;
+	vector<int> v;
+	while(stream >> i) {
+		v.push_back(i);
+	}
+	v.shrink_to_fit();
+	return v;
+}
+
+string getFileName(const string& seed, int p) {
+	stringstream s;
+	s << dataPathName;
+	s << seed;
+	s << "_" << setfill('0') << setw(2) << dec << p;
+	s << ".dat";
+	return s.str();
+}
+
+
+
+string getOutputFileName(const long cardA, long cardB, long cardX, int p, int q) {
+	stringstream s;
+	s << setfill('0') << setw(11) << cardA;
+	s << "_";
+	s << setfill('0') << setw(11) << cardB;
+	s << "_";
+	s << setfill('0') << setw(11) << cardX;
+	s << "_";
+	s << setfill('0') << setw(2) << p;
+	s << "_";
+	s << setfill('0') << setw(2) << q;
+	s << ".dat";
+	return s.str();
[…]

### 2016-04-16 c++/simulator.cpp

@@ -1,0 +1,204 @@
+//#############################
+//# Copyright 2016 Otmar Ertl #
+//#############################
+
+#include "cardinalityestimation.hpp"
+#include "hyperloglog.hpp"
+
+#include <iostream>
+#include <fstream>
+#include <iostream>
+#include <algorithm>
+#include <string>
+#include <iomanip>
+#include <cassert>
+#include <cstdlib>
+#include <sstream>
+
+using namespace std;
+
+const int minP = 4;
+const int maxP = 22;
+const int maxPplusQ = 64;
+
+const string seedFileName = "../data/hll/seed.dat";
+const string cardFileName = "../data/cardinalities.dat";
+const string dataPathName = "../data/hll/";
+
+template <typename T> void printToFile(const string& fileName, const T* data, size_t numSketches, size_t numCardinalities) {
+	ofstream file(fileName);
+	for (size_t c = 0; c < numCardinalities; ++c) {
+		for (size_t s = 0; s < numSketches; ++s) {
+			if (s>0) {
+				file << " ";
+			}
+			file << data[c*numSketches+s];
+		}
+		file << '\n';
+	}
+}
+
+class parameter {
+	int p;
+	int q;
+	
+public:
+	parameter(int p_, int q_) : p(p_), q(q_) {
+		assert(p <= maxP);
+		assert(p >= minP);
+		assert(q >= 0);
+		assert(p+q <= maxPplusQ);
+	}
+	
+	int getP() const {return p;};
+	int getQ() const {return q;};
+};
+
+
+string getFilePrefix(const parameter& par) {
+	stringstream s;
+	s << "../data/";
+	s << par.getP();
+	s << "_";
+	s << par.getQ();
+	s << "_";
+	return s.str();
+}
+
+vector<parameter> readParameters() {
+	vector<parameter> parameters;
+	ifstream f("param.txt");
+	string line;
+	while (getline(f, line)) {
+		istringstream iss(line);
+		int precision;
+		int maxRegisterValue;
+		iss >> precision;
+		iss >> maxRegisterValue;
+		parameters.push_back(parameter(precision, maxRegisterValue));
+	}
+	return parameters;
+}
+
+vector<int> readCounts(string s) {
+	stringstream stream(s);
+	int i;
+	vector<int> v;
+	while(stream >> i) {
+		v.push_back(i);
+	}
+	v.shrink_to_fit();
+	return v;
+}
+
[…]

### 2016-04-16 python/benchmark-charts.py

@@ -1,0 +1,127 @@
+#############################
+# Copyright 2016 Otmar Ertl #
+#############################
+
+from math import expm1, cosh, tanh, sinh, log
+import matplotlib.pyplot as plt
+import csv
+import sys
+import numpy
+
+numPoints = 100000
+xMin = 1e-3
+xMax = 110
+filename_12_20 = '../data/benchmark_results_12_20.dat'
+filename_12_52 = '../data/benchmark_results_12_52.dat'
+
+def readData(filename):
+	csv.field_size_limit(sys.maxsize)
+	with open(filename , 'r') as file:
+		reader = csv.reader(file, skipinitialspace=True,delimiter=' ')
+		return numpy.array([[float(e) for e in r] for r in reader])
+		
+
+plt.rc('text', usetex=True)
+plt.rc('font', family='serif')
+
+
+data_12_20 = readData(filename_12_20)
+data_12_52 = readData(filename_12_52)
+
+cardinalities_12_20 = data_12_20[:,0]
+avg_exec_time_nanos_12_20 = data_12_20[:,1]
+avg_outer_loop_count_12_20 = data_12_20[:,2]
+avg_inner_loop1_count_12_20 = data_12_20[:,3]
+avg_inner_loop2_count_12_20 = data_12_20[:,4]
+avg_log_eval_count_12_20 = data_12_20[:,5]
+
+cardinalities_12_52 = data_12_52[:,0]
+avg_exec_time_nanos_12_52 = data_12_52[:,1]
+avg_outer_loop_count_12_52 = data_12_52[:,2]
+avg_inner_loop1_count_12_52 = data_12_52[:,3]
+avg_inner_loop2_count_12_52 = data_12_52[:,4]
+avg_log_eval_count_12_52 = data_12_52[:,5]
+
+xMin = 1
+xMax = min(cardinalities_12_20[-1], cardinalities_12_52[-1])
+
+# exec times
+fig_time, ax_time = plt.subplots(subplot_kw=dict(xlim=[xMin,xMax], ylim=[0,800], xscale='log', yscale='linear'))
+fig_time.set_size_inches(10, 5)
+ax_time.yaxis.grid(b=True, which='major', color='gray', linestyle='--')
+ax_time.xaxis.grid(b=True, which='major', color='gray', linestyle='--')
+ax_time.set_xlabel("cardinality")
+ax_time.set_ylabel("time (ns)")
+
[…]

### 2016-04-16 python/distribution_charts.py

@@ -1,0 +1,80 @@
+#############################
+# Copyright 2016 Otmar Ertl #
+#############################
+
+from numpy import ones, array
+from math import expm1, cosh, tanh, sinh, log, erf, sqrt
+import matplotlib.pyplot as plt
+import csv
+import sys
+import numpy
+
+dataDir = '../data/'
+
+def readData(filename):
+	csv.field_size_limit(sys.maxsize)
+	with open(dataDir + filename , 'r') as file:
+		reader = csv.reader(file, skipinitialspace=True,delimiter=' ')
+		return numpy.array([[float(e) for e in r] for r in reader][1:])
+
+def createFigure(p, q, dataSource, figureName):
+	
+	filename = str(p) + '_' + str(q)
+		
+	m = pow(2, p)
+
+	cardinalities = readData('cardinalities.dat').flatten()
+	data = readData(filename + '_' + dataSource + '.dat')
+
+	plt.rc('text', usetex=True)
+	plt.rc('font', family='serif')
+
+	yMax = 6./pow(2., p*0.5)
+	fig1, ax1 = plt.subplots(subplot_kw=dict(xlim=[cardinalities[0],cardinalities[-1]], ylim=[-yMax,yMax], xscale='log', yscale='linear'))
+
+	fig1.set_size_inches(10, 5)
+	ax1.yaxis.grid(b=True, which='major', color='gray', linestyle='--')
+	ax1.xaxis.grid(b=True, which='major', color='gray', linestyle='--')
+	ax1.set_xlabel("cardinality")
+	ax1.set_ylabel("relative error")
+
+
+	pm3s = (1.-erf(3./sqrt(2.)))/2.
+	pp3s = (1.+erf(3./sqrt(2.)))/2.
+	pm2s = (1.-erf(2./sqrt(2.)))/2.
+	pp2s = (1.+erf(2./sqrt(2.)))/2.
+	pm1s = (1.-erf(1./sqrt(2.)))/2.
+	pp1s = (1.+erf(1./sqrt(2.)))/2.
+
+	datam3s = numpy.divide(numpy.percentile(data, pm3s*100., axis=1), cardinalities)-1.
+	datap3s = numpy.divide(numpy.percentile(data, pp3s*100., axis=1), cardinalities)-1.
+	datam2s = numpy.divide(numpy.percentile(data, pm2s*100., axis=1), cardinalities)-1.
+	datap2s = numpy.divide(numpy.percentile(data, pp2s*100., axis=1), cardinalities)-1.
[…]

### 2016-04-16 python/helper-function.py

@@ -1,0 +1,46 @@
+#############################
+# Copyright 2016 Otmar Ertl #
+#############################
+
+from numpy import ones, array
+from math import expm1, cosh, tanh, sinh, log
+import matplotlib.pyplot as plt
+
+numPoints = 100000
+xMin = 1e-3
+xMax = 110
+
+def helper(x):
+    if x == 0:
+        return 0
+    else:
+        return 1 - x/expm1(x)
+
+def helperApprox(x):
+    if x < 1e-2:
+        x2 = x*x
+        return 0.5*x*(1. - x/6.*(1.-x2/60.*(1.-x2/42.)))
+    elif x > 45:
+        return 1
+    else:
+        x_half = x*0.5
+        q_half = helperApprox(x_half)
+        t = 2*(1-q_half)
+        return (x_half+q_half*t)/(x_half+t)
+
+
+iter = range(0,numPoints)
+xValues = list([xMin + i*((xMax - xMin)/(numPoints-1)) for i in iter])
+yValues = list([helperApprox(xValues[i]) for i in iter])
+
+plt.rc('text', usetex=True)
+plt.rc('font', family='serif')
+
+fig1, ax1 = plt.subplots(subplot_kw=dict(xlim=[0,20], ylim=[0,1.2]))
+fig1.set_size_inches(6, 4)
+ax1.plot(xValues, yValues, color='black')
+ax1.grid(b=True, which='major', color='black', linestyle='--')
+ax1.set_xlabel(r"$x$")
+ax1.set_ylabel(r"$h(x)$")
+fig1.savefig('../paper/helper.svg', format='svg', dpi=600)
+

### 2016-04-16 python/recursion-stability.py

@@ -1,0 +1,22 @@
+#############################
+# Copyright 2016 Otmar Ertl #
+#############################
+
+from math import expm1, ceil
+from scipy.optimize import minimize
+
+def h(x):
+    if x == 0:
+        return 0
+    else:
+        return 1 - x/expm1(x)
+        
+def f(x):
+	return -((2*h(x)*(1-2*h(x)))/(x+2*h(x)*(1-h(x)))) - 2*h(x)/(x+2*(1-h(x)))
+	
+	
+res = minimize(f, 1)
+
+out = open('../paper/max_errror_propagation_factor.txt', 'w')
+out.write(str(ceil(-res.fun*1000)/1000.))
+

### 2016-04-25 paper/paper.tex
some minor changes

@@
 The number of bits of the consumed hash value $\symPrecision+\symRegRange$ limit the maximum cardinality that can be tracked.
-Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent  , which will reduce estimation accuracy drastically.
+Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent, which will reduce estimation accuracy drastically.
 
@@
 
-For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$ for which a lot of registers already are already in a saturated state meaning that they have already reached the maximum possible value $\symRegRange + 1$, the \emph{raw} estimate underestimates the cardinalities.
+For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state meaning that they have reached the maximum possible value $\symRegRange + 1$, the \emph{raw} estimate underestimates the cardinalities.
 For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$ which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
@@
 
-The entire original estimation algorithm as presented in \cite{Flajolet2007} including small and large range corrections is summarized by \algoref{alg:estimate_original}. 
+The original estimation algorithm as presented in \cite{Flajolet2007} including small and large range corrections is summarized by \algoref{alg:estimate_original}. 
 \begin{algorithm}
@@
 
-The large range correction is not satisfying either.
-The large range correction does not reduce the estimation error.
-It even increases the estimation error.
+The large range correction is not satisfying either as it does not reduce the estimation error.
+Quite the contrary, it even increases the estimation error.
 However, instead of underestimating the cardinalities, they are now overestimated.
 A simple approach to avoid the large range correction at all is the use of larger hash values $(\symPrecision+\symRegRange>32)$ which shift the biased region of $\symCardinalityOriginalEstimate$ to higher cardinalities.
-However, in case $\symRegRange \geq 30$ 6-bit registers as used in \cite{Heule2013} are required to allow representing all values in range $[0, \symRegRange+1]$.
+However, in case $\symRegRange \geq 30$ 6-bit registers as used in \cite{Heule2013} are required to allow representing all values in the range $[0, \symRegRange+1]$.
 If 64-bit hash values are used as proposed in \cite{Heule2013} $(\symPrecision+\symRegRange=64)$, a correction for large cardinalities is not required.
@@
 \label{sec:cardinality_estimation}
-In the following we describe a new cardinality estimation method that does not require special handling for small and large cardinalities and which also does not require any empirically determined constants or corrections.
+In the following we describe a new cardinality estimation method that does not require special handling for small or large cardinalities and which also does not require any empirically determined constants or corrections.
 The method is based on the maximum likelihood method.
@@
 In comparison, \algoref{alg:insert} uses only a single hash value that defines which register needs to be updated.
-As a consequence, the register values are dependent, which makes the joint probability distribution of all register values more complex.
+As a consequence, the register values are statistically dependent, which makes the joint probability distribution of all register values more complex.
 
@@
 Furthermore, we will demonstrate that consequent application of the maximum likelihood method to the Poisson model directly reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values.
-The history of the HyperLogLog algorithm shows that the \emph{raw} estimate \eqref{alg:estimate_original} was first found after several attempts using the geometric mean \cite{Durand2003, Flajolet2007}.
+The history of the HyperLogLog algorithm shows that the \emph{raw} estimate \eqref{alg:estimate_original} was first found after several attempts using the geometric mean \cite{Flajolet2007, Durand2003}.
 
@@
 
-Under the assumption of an ideal hash function the probability mass function for the register values $\vec{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
+Under the assumption of an ideal hash function, the probability mass function for the register values $\vec{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
 \begin{equation}
@@
 \end{equation}
-which can also be written as
+This function can also be written as
 \begin{equation}
@@
 \end{equation}
-if, the function $\symHelper(\symX)$ is defined as
+where the function $\symHelper(\symX)$ is defined as
 \begin{equation}
@@
 
-It is easy to see that the $\symPoissonRateEstimate$ remains equal or is larger after updating the HyperLogLog data structure as defined by \algoref{alg:insert}.
-An update potentially changes the cardinality vector $(\symCount_0,\ldots,\symCount_{\symRegRange+1})$ to $(\symCount_0,\ldots,\symCount_\symIndexI-1,\ldots,\symCount_\symIndexJ+1,\ldots,\symCount_{\symRegRange+1})$ where $\symIndexI < \symIndexJ$.
+It is easy to see that the estimate $\symPoissonRateEstimate$ remains equal or becomes larger, when inserting an element into the HyperLogLog sketch following \algoref{alg:insert}.
+An update potentially changes the multiplicity vector $(\symCount_0,\ldots,\symCount_{\symRegRange+1})$ to $(\symCount_0,\ldots,\symCount_\symIndexI-1,\ldots,\symCount_\symIndexJ+1,\ldots,\symCount_{\symRegRange+1})$ where $\symIndexI < \symIndexJ$.
 Writing \eqref{equ:func} as
@@
 Keeping $\symX$ fixed during an update decreases $\symFunc(\symX)$.
-As a consequence, since $\symFunc(\symX)$ is increasing the new root and hence the estimate must be larger than before the update.
+As a consequence, since $\symFunc(\symX)$ is increasing, the new root and hence the estimate must be larger than before the update.
 
@@
 \end{equation}
-Using the monotonicity of $\symHelper$ the lower bound
+Using the monotonicity of $\symHelper$, the lower bound
 \begin{equation}
@@
 Note, if the HyperLogLog sketch is in the intermediate range, where $\symCount_0=\symCount_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the \emph{raw} estimate \eqref{equ:intermediate_estimate}.
-Hence, the maximum likelihood method directly leads to the harmonic mean that is used for the \emph{raw} estimate.
+Hence, the maximum likelihood method leads directly to the harmonic mean that is used for the \emph{raw} estimate.
 If $\symCount_0>0$ or $\symCount_{\symRegRange+1}>0$ the bounds do no longer follow the harmonic mean and this is also the reason why the \emph{raw} estimate fails for small and large cardinalities.
@@
 Since $\symFunc$ is concave and increasing, both, Newton-Raphson iteration and the secant method, converge to the root, if the function is negative for the starting points.
-In the following the secant method is chosen to derive the cardinality estimation algorithm presented below.
-Even though it has the disadvantage of slower convergence, a single iteration is simpler to calculate as it does not require the evaluation of the first derivative as shown in \autoref{fig:original_estimate}.
+In the following we start from the secant method to derive the new cardinality estimation algorithm.
+Even though the secant method has the disadvantage of slower convergence, a single iteration is simpler to calculate as it does not require the evaluation of the first derivative.
 
@@
 \item The function $\symHelper(\symX)$ only needs to be evaluated at values $\lbrace\frac{\symX}{2^{\symRegVal'_\text{max}}},\frac{\symX}{2^{\symRegVal'_\text{max}-1}}\ldots,\frac{\symX}{2^{\symRegVal_\text{min}}}\rbrace$ where $\symRegVal'_\text{max} := \min(\symRegVal_\text{max}, \symRegRange)$.
-This series corresponds to geometric series with ratio 2.
+This series corresponds to a geometric series with ratio 2.
 A straightforward calculation using \eqref{equ:helper} is very expensive because of the exponential function.

### 2016-04-25 python/distribution_charts.py

@@ -52,9 +52,7 @@ def createFigure(p, q, dataSource, figureName):
     datap2s = numpy.divide(numpy.percentile(data, pp2s*100., axis=1), cardinalities)-1.
     datam1s = numpy.divide(numpy.percentile(data, pm1s*100., axis=1), cardinalities)-1.
     datap1s = numpy.divide(numpy.percentile(data, pp1s*100., axis=1), cardinalities)-1.
-    data05 = numpy.divide(numpy.percentile(data, 5, axis=1), cardinalities)-1.
     data50 = numpy.divide(numpy.percentile(data, 50, axis=1), cardinalities)-1.
-    data95 = numpy.divide(numpy.percentile(data, 95, axis=1), cardinalities)-1.
     ax1.plot(cardinalities, data50, color='black', linewidth=1.0, label='median')
     ax1.fill_between(cardinalities, datam3s, datap3s, facecolor='#dddddd', edgecolor='none')
     ax1.fill_between(cardinalities, datam2s, datap2s, facecolor='#bbbbbb', edgecolor='none')

### 2016-04-30 paper/paper.tex
minor changes in paper, output avg min and max register values

@@
 
-\autoref{fig:raw_estimate} shows the distribution of the relative error of the \emph{raw} estimate for $10000$ randomly generated HyperLogLog sketches over the number of distinct element insertions.
+\autoref{fig:raw_estimate} shows the distribution of the relative error of the \emph{raw} estimate for 10\,000 randomly generated HyperLogLog sketches over the number of distinct element insertions.
 Obviously, the \emph{raw} estimate is biased for small and large cardinalities and fails to return accurate estimates.
@@
 \caption{The distribution of the relative estimation error over the cardinality as obtained by the raw estimation method \cite{Flajolet2007}.
-10000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$ have been evaluated.}
+10\,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$ have been evaluated.}
 \label{fig:raw_estimate}
@@
 \caption{The distribution of the relative estimation error over the cardinality as obtained by the original estimation method \cite{Flajolet2007}.
-10000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$ have been evaluated.}
+10\,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$ have been evaluated.}
 \label{fig:original_estimate}
@@
 If we denote $\symRegVal_\text{min}:=\min\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$ and $\symRegVal_\text{max}:=\max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$,  it is sufficient to loop over all indices in the range $[\symRegVal_\text{min}, \symRegVal_\text{max}]$.
-\item It is sufficient to calculate the coefficient of the linear term $\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}$ once and reuse it for any further iterations of the secant method.
+\item The coefficient of the linear term $\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}$ can be precalculated and reused for all function evaluations.
 \item Many programming languages allow the efficient multiplication and division by any integral power of 2 using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
@@
 A straightforward calculation using \eqref{equ:helper} is very expensive because of the exponential function.
-However, if we know $\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
+However, if we already know $\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
 \begin{equation}
@@
 \subsection{Estimation error}
+In order to verify the new estimation algorithm, we generated 10\,000 HyperLogLog sketches and inserted up to 50 billion unique elements.
+Element hash values have been mocked by random numbers.
+For the following results we used the Mersenne Twister random number generator with 19937 bit state size from the C++ standard library.
+
+\autoref{fig:estimation_error_12_20} shows the distribution of the relative error of the estimated cardinality using \algoref{alg:estimate_ml} compared to the true cardinality for $\symPrecision=12$ and $\symRegRange=20$.
+As the median shows, the error is essentially unbiased over the entire cardinality range except for very small values.
+The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is an order of magnitude larger than the operating range upper bound of the \emph{raw} estimate (\autoref{fig:raw_estimate}) or the original method (\autoref{fig:original_estimate}).
+
+\begin{figure}
+\centering
+\includesvg[width=1\textwidth]{max_likelihood_estimate_12_20}
+\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
+\label{fig:estimation_error_12_20}
+\end{figure}
+
 \begin{figure}
@@
 \label{fig:estimation_error_8_24}
-\end{figure}
-\begin{figure}
-\centering
-\includesvg[width=1\textwidth]{max_likelihood_estimate_12_20}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
-\label{fig:estimation_error_12_20}
 \end{figure}
@@
 Given the corresponding register values $\vec{\symRegVal}_1$ and
-$\vec{\symRegVal}_2$ we attempt to estimate the cardinalities of the pair-wise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$.
+$\vec{\symRegVal}_2$, we attempt to estimate the cardinalities of the pair-wise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$.
 Motivated by the good results we have obtained for cardinality estimation of a single HyperLogLog sketch, we want to get these estimates using the maximum likelihood method applied to the joint probability distribution of $\vec{\symRegVal}_1$ and $\vec{\symRegVal}_2$.
@@
 Under the Poisson model the register values are independent and identically distributed.
-Therefore, we first consider a single register and derive the joint probability distribution for its values in both sketches.
+Therefore, we first derive the joint probability distribution for a single register that has value $\symRegValVariate_1$ in the first HyperLogLog sketch representing $\symSetS_1$ and value $\symRegValVariate_2$ in the second HyperLogLog sketch representing $\symSetS_2$.
 
@@
 
-Since under the Poisson model, the values for different registers are independent we are now able to write the joint probability mass function for the joint state of both HyperLogLog sketches.
+Since under the Poisson model, the values for different registers are independent we are now able to write the joint probability mass function for the joint state of both HyperLogLog sketches
 \begin{equation}
@@
 \symProbabilityMass(\symRegVal_1,\symRegVal_2)
-^{\symCount_{\symRegVal_1\symRegVal_2}}
+^{\symCount_{\symRegVal_1\symRegVal_2}}.
 \end{equation}
@@
 \label{equ:h2_relative_error}
-\symError_2 = \frac{\tilde{\symHelper}_2}{\symHelper_2}-1
-\end{equation}
-Putting \eqref{equ:h2} and \eqref{equ:h2_approx} into \eqref{equ:h2_relative_error} and using first-order approximations 
+\symError_2 = \frac{\tilde{\symHelper}_2}{\symHelper_2}-1.
+\end{equation}
+Putting \eqref{equ:h2} and \eqref{equ:h2_approx} into \eqref{equ:h2_relative_error} and using the first-order approximations 
 \begin{equation}
@@
 \frac{2\symHelper_1}{x+2\left(1-\symHelper_1\right)}
-+\symBigO(\symError_1^2)
++\symBigO(\symError_1^2),
 \end{equation}
@@
 \end{equation}
-By numerical means it can be easily shown that
+By numerical means it is easy to show that
 \begin{equation}

### 2016-04-30 c++/benchmark.cpp

@@ -157,11 +157,13 @@ void run(const int p, const int q, const string& resultsFileName) {
             int innerLoop1IterationsCount;
             int innerLoop2IterationsCount;
             int logEvaluationCount;
+            int kMin;
+            int kMax;
             
             
             auto start = std::chrono::system_clock::now();
             for(auto c : counts) {
-                sum += maxLikelihoodEstimate(c, outerLoopIterationsCount, innerLoop1IterationsCount, innerLoop2IterationsCount, logEvaluationCount);
+                sum += maxLikelihoodEstimate(c, outerLoopIterationsCount, innerLoop1IterationsCount, innerLoop2IterationsCount, logEvaluationCount, kMin, kMax);
                 outerLoopIterationsCountSum += outerLoopIterationsCount;
                 innerLoop1IterationsCountSum += innerLoop1IterationsCount;
                 innerLoop2IterationsCountSum += innerLoop2IterationsCount;

### 2016-04-30 c++/cardinalityestimation.hpp

@@ -30,7 +30,7 @@ int getPFromCounts(const std::vector<int>& c) {
     return getPFromNumberRegisters(m);
 }
 
-double maxLikelihoodEstimate(const std::vector<int>& c, int& outerLoopIterationsCount, int& innerLoop1IterationsCount, int& innerLoop2IterationsCount, int& logEvaluationCount) {
+double maxLikelihoodEstimate(const std::vector<int>& c, int& outerLoopIterationsCount, int& innerLoop1IterationsCount, int& innerLoop2IterationsCount, int& logEvaluationCount, int& kMin, int& kMax) {
 
     const double eps = 1e-2;
     
@@ -41,11 +41,9 @@ double maxLikelihoodEstimate(const std::vector<int>& c, int& outerLoopIterations
     
     int q = c.size() - 2;
 
-    int kMin;
     for(kMin=0; c[kMin]==0; ++kMin);
     int kMinPrime = std::max(1, kMin);
     
-    int kMax;
     for(kMax=q+1; c[kMax]==0; --kMax);
     int kMaxPrime = std::min(q, kMax);
 
@@ -127,8 +125,10 @@ double maxLikelihoodEstimate(const std::vector<int>& c) {
     int innerLoop1IterationsCount;
     int innerLoop2IterationsCount;
     int logEvaluationCount;
+    int kMin;
+    int kMax;
     
-    return maxLikelihoodEstimate(c, outerLoopIterationsCount, innerLoop1IterationsCount, innerLoop2IterationsCount, logEvaluationCount);
+    return maxLikelihoodEstimate(c, outerLoopIterationsCount, innerLoop1IterationsCount, innerLoop2IterationsCount, logEvaluationCount, kMin, kMax);
 }
 
 double flajoletSmallRangeEstimate(const std::vector<int>& c, int m) {

### 2016-04-30 c++/simulator.cpp

@@ -148,11 +148,16 @@ int main(int argc, char* argv[])
         vector<int> innerLoop1IterationsCount(resultBufferSize);
         vector<int> innerLoop2IterationsCount(resultBufferSize);
         vector<int> logEvaluationCount(resultBufferSize);
+        vector<int> kMin(resultBufferSize);
+        vector<int> kMax(resultBufferSize);
         
         vector<double> strongLowerBoundEstimates(resultBufferSize);
         vector<double> weakLowerBoundEstimates(resultBufferSize);
         vector<double> strongUpperBoundEstimates(resultBufferSize);
         vector<double> weakUpperBoundEstimates(resultBufferSize);
+        
+        vector<double> kMinSum(cardinalities.size(), 0.);
+        vector<double> kMaxSum(cardinalities.size(), 0.);
 
         #pragma omp parallel for
         for(size_t seedCounter = 0; seedCounter < seeds.size(); ++seedCounter) {
@@ -173,15 +178,26 @@ int main(int argc, char* argv[])
                 
                 size_t resultPos = cardinalityIdx*seeds.size()+seedCounter;
                 
+                int kMin;
+                int kMax;
+                
                 flajoletSmallRangeEstimates[resultPos] = flajoletSmallRangeEstimate(c);
                 flajoletMidRangeEstimates[resultPos] = flajoletRawEstimate(c);
                 flajoletEstimates[resultPos] = flajoletEstimate(c);
-                maxLikelihoodEstimates[resultPos] = maxLikelihoodEstimate(c, outerLoopIterationsCount[resultPos], innerLoop1IterationsCount[resultPos], innerLoop2IterationsCount[resultPos], logEvaluationCount[resultPos]);
+                maxLikelihoodEstimates[resultPos] = maxLikelihoodEstimate(c, outerLoopIterationsCount[resultPos], innerLoop1IterationsCount[resultPos], innerLoop2IterationsCount[resultPos], logEvaluationCount[resultPos], kMin, kMax);
                 weakLowerBoundEstimates[resultPos] = weakLowerBoundEstimate(c);
                 strongLowerBoundEstimates[resultPos] = strongLowerBoundEstimate(c);
                 weakUpperBoundEstimates[resultPos] = weakUpperBoundEstimate(c);
                 strongUpperBoundEstimates[resultPos] = strongUpperBoundEstimate(c);
+                
[…]

### 2016-05-01 paper/paper.tex
added some paragraphs to the result section

@@
 As the median shows, the error is essentially unbiased over the entire cardinality range except for very small values.
-The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is an order of magnitude larger than the operating range upper bound of the \emph{raw} estimate (\autoref{fig:raw_estimate}) or the original method (\autoref{fig:original_estimate}).
+The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is about an order of magnitude larger than the operating range upper bound of the \emph{raw} estimate (\autoref{fig:raw_estimate}) or the original method (\autoref{fig:original_estimate}).
 
@@
 
+The new estimation algorithm also works well for other HyperLogLog configurations.
+First we considered the original HyperLogLog algorithm that used a 32-bit hash function ($\symPrecision + \symRegRange = 32$).
+The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, $\symPrecision=20$ are shown in \autoref{fig:estimation_error_8_24}, \autoref{fig:estimation_error_16_16}, and \autoref{fig:estimation_error_20_12}, respectively.
+As expected, since  $\symPrecision + \symRegRange = 32$ is kept constant, the operating range remains more or less the same, while the relative error decreases with increasing precision.
+Again, the new algorithm gives inherently unbiased estimates.
+
 \begin{figure}
@@
 \end{figure}
+
+\begin{figure}
+\centering
+\includesvg[width=1\textwidth]{max_likelihood_estimate_16_16}
+\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
+\label{fig:estimation_error_16_16}
+\end{figure}
+
+\begin{figure}
+\centering
+\includesvg[width=1\textwidth]{max_likelihood_estimate_20_12}
+\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 20$ and $\symRegRange=12$.}
+\label{fig:estimation_error_20_12}
+\end{figure}
+
+As proposed in \cite{Heule2013}, the operating range can be extended by
+replacing the 32-bit hash function by a 64-bit hash function.
+\autoref{fig:estimation_error_12_52} shows the relative error for such a HyperLogLog configuration with precision  $\symPrecision=12$.
+In this case, in order to use all the 64 bits of the hash value, $\symRegRange$ must be chosen to be equal to $64 - \symPrecision = 52$.
+As a consequence, in order to represent the maximum possible register value $\symRegRange+1 = 53$, 6 bits are needed for each register.
+The doubled hash value size shifts the maximum trackable cardinality value towards $2^{64}$.
+As \autoref{fig:estimation_error_12_52} shows, when compared to the 32-bit hash value case given in \autoref{fig:estimation_error_12_20}, the estimation error remains constant over the entire simulated cardinality range up to 50 billions.
+
+\begin{figure}
+\centering
+\includesvg[width=1\textwidth]{max_likelihood_estimate_12_52}
+\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
+\label{fig:estimation_error_12_52}
+\end{figure}
+
+\autoref{fig:estimation_error_12_14}
+
 \begin{figure}
@@
 \end{figure}
-\begin{figure}
-\centering
-\includesvg[width=1\textwidth]{max_likelihood_estimate_16_16}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
-\label{fig:estimation_error_16_16}
-\end{figure}
-\begin{figure}
-\centering
-\includesvg[width=1\textwidth]{max_likelihood_estimate_20_12}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 20$ and $\symRegRange=12$.}
-\label{fig:estimation_error_20_12}
-\end{figure}
-\begin{figure}
-\centering
-\includesvg[width=1\textwidth]{max_likelihood_estimate_12_52}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
-\label{fig:estimation_error_12_52}
-\end{figure}
-
-\autoref{fig:estimation_error_8_24}
-\autoref{fig:estimation_error_12_14}
-\autoref{fig:estimation_error_12_20}
-\autoref{fig:estimation_error_12_52}
-\autoref{fig:estimation_error_16_16}
-\autoref{fig:estimation_error_20_12}
 

### 2016-05-03 paper/paper.tex
added a paragraph to the result section

@@
 
-\autoref{fig:estimation_error_12_14}
+We also evaluated the case $\symPrecision = 12$ and $\symRegRange=14$, which is interesting, because the register values are limited to the range $[0, 15]$.
+As a consequence, 4 bits are sufficient for representing a single register value.
+This allows two registers to share a single byte, which is beneficial from a performance perspective.
+Nevertheless, this configuration still allows the estimation of cardinalities up to 100 millions as shown in \autoref{fig:estimation_error_12_14}, which could be enough for many applications.
 

### 2016-05-03 paper/paper.tex
added performance results section

@@
 \subsection{Performance}
-\autoref{fig:avg_exec_time}
+In order to prove that the new algorithm is not only accurate, but also fast, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch.
+For different cardinalities we loaded the multiplicity vectors of 1000 precalculated and randomly generated HyperLogLog sketches into main memory.
+The average computation time was determined by cycling over these multiplicity vectors and passing them as input to the new estimation algorithm.
+For each evaluated cardinality value the average execution time was calculated after 100 cycles which corresponds to 100\,000 algorithm executions.
+The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \autoref{fig:avg_exec_time}.
+All these benchmarks where carried out on an Intel Core i5-2500K clocking at 3.3GHz.
+As can be seen, the average calculation time did never exceed 700\,ns.
+
+The numbers do not yet include the required processing time to calculate the multiplicity vector from the HyperLogLog sketch, which requires a complete scan over all registers and counting the different register values into an array.
+A theoretical lower bound for this processing time can be derived using the maximum memory bandwidth of the CPU, which is 21\,GB/s for an Intel Core i5-2500K.
+For precision $\symPrecision=12$ there are 4096 registers, each of them requires at least 5 bits.
+Hence, the total data size of the HyperLogLog sketch is 2.5\,kB minimum and transfer time from main memory to CPU will be at least 120\,ns.
+Having this value in mind, the presented processing time numbers are quite satisfying.
 

### 2016-05-03 python/benchmark-charts.py
added low resolution performance chart

@@ -58,6 +58,7 @@ ax_time.plot(cardinalities_12_52, avg_exec_time_nanos_12_52, color='blue', lines
 
 ax_time.legend(loc=2,prop={'size':12})
 
+fig_time.savefig('../paper/avg_exec_time.png', format='png', dpi=100)
 fig_time.savefig('../paper/avg_exec_time.svg', format='svg', dpi=600)
 plt.close(fig_time)
 

### 2016-05-21 python/power-series-function.py
added python script for the power series that is used for the raw estimate correction

@@ -1,0 +1,60 @@
+#############################
+# Copyright 2016 Otmar Ertl #
+#############################
+
+from numpy import ones, array
+from math import expm1, cosh, tanh, sinh, log, sqrt, exp, pi, sin
+import matplotlib.pyplot as plt
+from scipy.optimize import minimize_scalar
+
+numPoints = 100000
+xMin = 0.000
+xMax = 1
+
+def h2(x):
+    return exp(-pow(2,x))*pow(2,x)
+    
+def h(x):
+    
+    a = h2(x)
+    amin = 0.
+    k = 0
+    while True:
+        k -= 1
+        aminold = amin
+        amin += h2(k+x)
+        if (amin == aminold):
+            break
+    
+    amax = 0.;
+    k = 0
+    while True:
+        k += 1
+        amaxold = amax
+        amax += h2(k+x)
+        if (amax == amaxold):
+            break
+    
+    return log(2.)*(a + amin + amax)
+
+
+
+def f1(x):
+    return h(x) - 1.
+    
+def f2(x):
+    return 1. - h(x)
+
+iter = range(0,numPoints)
+xValues = list([xMin + i*((xMax - xMin)/(numPoints-1)) for i in iter])
+yValues = list([f1(xValues[i]) for i in iter])
+plt.plot(xValues, yValues)
+
+
+res1 = minimize_scalar(f1, bounds = (0,1), method='bounded')
+res2 = minimize_scalar(f2, bounds = (0,1), method='bounded')
+
+print(res1)
+print(res2)
+plt.show()
+

### 2016-05-21 paper/paper.tex
reorganized paper, new paper title, added new correction for raw estimate, added more simulation results

@@
 
-\documentclass[11pt]{article} % use larger type; default would be 10pt
-
-\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
+\documentclass[a4paper]{scrartcl}
+%\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
 \usepackage{amsmath}
 \usepackage{amssymb}
-%\usepackage{natbib}
-
-%%% Examples of Article customizations
-% These packages are optional, depending whether you want the features they provide.
-% See the LaTeX Companion or other references for full information.
-
-%%% PAGE DIMENSIONS
-\usepackage{geometry} % to change the page dimensions
-\geometry
-{a4paper} % or letterpaper (US) or a5paper or....
-% \geometry{margin=2in} % for example, change the margins to 2 inches all round
-% \geometry{landscape} % set up the page for landscape
-%   read geometry.pdf for detailed page layout information
-
 \usepackage{graphicx} % support the \includegraphics command and options
 \usepackage{svg}
-
 \usepackage{hyperref}
@@
 \usepackage{url}
-
-% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
-
-%%% PACKAGES
 \usepackage{booktabs} % for much better looking tables
@@
 
-%%% HEADERS & FOOTERS
-\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
-\pagestyle{fancy} % options: empty , plain , fancy
-\renewcommand{\headrulewidth}{0pt} % customise the layout...
-\lhead{}\chead{}\rhead{}
-\lfoot{}\cfoot{\thepage}\rfoot{}
-
-%%% SECTION TITLE APPEARANCE
-\usepackage{sectsty}
-\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
-% (This matches ConTeXt defaults)
-
-%%% ToC (table of contents) APPEARANCE
-\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
-\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
-\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
-\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!
+\renewcommand{\sffamily}{\rmfamily}
 
@@
 \symDefine{\symZ}{z}
+\symDefine{\symInteger}{z}
+\symDefine{\symIntegerVariate}{Z}
 \symDefine{\symCardinality}{n}
@@
 \symDefine{\symBigO}{\mathcal{O}}
+\symDefine{\symExpectation}{\mathbb{E}}
 \symDefine{\symProbability}{P}
@@
 \symDefine{\symCountMatrix}{\mathbf{c}}
+\symDefine{\symPowerSeriesFunc}{\xi}
+\symDefine{\symEpsPowerSeriesFunc}{\varepsilon_\symPowerSeriesFunc}
 
@@
 
-\title{Maximum likelihood estimation of cardinalities\\ from HyperLogLog sketches}
+\title{Cardinality estimation methods for HyperLogLog sketches}
+\subtitle{Draft version}
 \author{Otmar Ertl}
-%\date{} % Activate to display a given date or no date (if empty),
-         % otherwise the current date is printed
 
@@
 \begin{abstract}
-This paper presents a new method to estimate the cardinality of a multiset recorded by HyperLogLog sketches.
-The maximum likelihood method is applied to the probability mass function under the Poisson model.
-A numerical algorithm is presented that is able to calculate the estimate efficiently.
-Simulations finally show that this new estimation procedure works well and is inherently unbiased over a wide range of cardinalities.
+This paper presents new methods to estimate the cardinalities of multisets recorded by HyperLogLog sketches.
+A theoretically founded extension to the original estimate is presented that eliminates the bias for small and large cardinalities.
+Based on the maximum likelihood principle another unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate.
+The maximum-likelihood method is also appropriate to improve the cardinality estimates of set intersections when compared to the inclusion-exclusion principle.
+The new methods are demonstrated and verified by extensive simulations.
 \end{abstract}
@@
 
-\section{The HyperLogLog algorithm}
+\section{HyperLogLog data structure}
 The HyperLogLog algorithm uses a sketching data structure that consists of $\symNumReg$ registers.
@@
 
-\subsection{Original cardinality estimation method}
-The cardinality estimate is based on the idea that approximately $\symNumReg  2^{\symRegVal_\symIndexJ}$ distinct element insertions are needed until the value of register $\symIndexJ$ reaches $\symRegVal_\symIndexJ$.
+\subsection{Joint probability distribution of register values}
+
+Under the assumption of an ideal hash function, the probability mass function for the register values $\vec{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
+\begin{equation}
+\label{equ:multinomialProbabilityMass}
+\symProbabilityMass(\vec{\symRegVal}\vert\symCardinality)
+=
+\sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots\symCardinality_\symNumReg}
+\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexJ=1}^\symNumReg \symRegProbability_{\symCardinality_\symIndexJ\symRegVal_\symIndexJ},
+\end{equation}
+where $\symCardinality$ is the cardinality.
+The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities.
+$\symRegProbability_{\symNu\symKappa}$ is the probability that a register is equal to $\symKappa$ after inserting $\symNu$ distinct elements
+\begin{equation}
+\symRegProbability_{\symNu\symKappa} 
+:=
+\begin{cases}
+1 & \symNu=0 \wedge \symKappa = 0\\
+0 & \symNu=0 \wedge 1\leq\symKappa\leq\symRegRange+1\\
+0 & \symNu\geq1 \wedge \symKappa = 0\\
+\left(1-\frac{1}{2^\symKappa}\right)^\symNu - \left(1-\frac{1}{2^{\symKappa-1}}\right)^\symNu & \symNu \geq 1 \wedge 1\leq\symKappa\leq\symRegRange\\
+1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symNu & \symNu\geq 1 \wedge \symKappa = \symRegRange +1 \\
+\end{cases}
+\end{equation}
+Note that the order of register values $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ is not important for the estimation of the cardinality.
+More formally, the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$ is a sufficient statistic for $\symCardinality$.
+Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace = 0^{\symCount_0}1^{\symCount_1}\cdots\symRegRange^{\symCount_{\symRegRange}}(\symRegRange+1)^{\symCount_{\symRegRange+1}}$ where $\symCount_\symIndexJ$ is the multiplicity of value $\symIndexJ$.
+As a consequence, the multiplicity vector $\vec{\symCount} := (\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is also a sufficient statistic for the cardinality.
+In addition, this vector also contains all the information about the HyperLogLog sketch that is required for  cardinality estimation: $\symRegRange=\dim(\vec{\symCount})-2$ and  $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$.
+
+\subsection{Poisson approximation}
+Unfortunately, the probability mass function \eqref{equ:multinomialProbabilityMass} is difficult to analyze, because the register values are statistically dependent.
+Therefore, as proposed in \cite{Flajolet2007} a Poisson model can be used, which assumes that the cardinality itself is distributed according to a Poisson distribution
+\begin{equation}
+\symCardinality \sim \text{Poisson}(\symPoissonRate).
+\end{equation}
+Under the Poisson model the register values are distributed according to
+\begin{align}
+\symProbabilityMass(\vec{\symRegVal}\vert\symPoissonRate) 
+&= 
+\sum_{\symCardinality=0}^\infty \symProbabilityMass(\vec{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
+\nonumber\\
+&= 
+\sum_{\symCardinality_1=0}^\infty
+\cdots
+\sum_{\symCardinality_\symNumReg=0}^\infty
+\prod_{\symIndexJ=1}^\symNumReg
+\symRegProbability_
+{\symCardinality_\symIndexJ\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexJ}}{\symCardinality_\symIndexJ!\symNumReg^{\symCardinality_\symIndexJ}}
+\nonumber\\
+&= 
+\prod_{\symIndexJ=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability_
+{\symCardinality\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
+\nonumber\\
+&= 
+\prod_{\symRegVal=0}^{\symRegRange+1} \left[
+\sum_{\symCardinality=0}^\infty
+\symRegProbability_
+{\symCardinality\symRegVal}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
+\right]^{\symCount_\symRegVal}
+\nonumber\\
+&=
+e^{-\symCount_0\frac{\symPoissonRate}{\symNumReg}}
+\cdot
+\left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\symCount_\symRegVal}\right)
+\cdot
+\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\symCount_{\symRegRange+1}}
+\end{align}
+This factorization shows that all register values are now independent and identically distributed under the Poisson model.
+The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is given by
+\begin{equation}
+\label{equ:register_value_distribution}
+\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
+=
+\begin{cases}
+0 & \symRegVal < 0 \\
+e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}} & 0\leq \symRegVal \leq \symRegRange \\
+1 & \symRegVal > \symRegRange
+\end{cases}
+\end{equation}
+Due to the simpler probability mass function, it is easier to find an estimate  $\symPoissonRateEstimate$ for the Poisson rate $\symPoissonRate$ than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}.
+Since, the expectation of the cardinality under the Poisson model is equal to $\symPoissonRate$, it makes sense to take $\symPoissonRateEstimate$ direclty as estimate for the cardinality, hence, $\symCardinalityEstimate := \symPoissonRateEstimate$.
+The Poisson approximation is known to work well for large rates.
+However, as simulation results show later, the Poisson approximation gives high quality estimates for small cardinalities as well.
+
+\section{Original cardinality estimation method}
+\label{sec:cardinality_estimation}
+The original cardinality estimate \cite{Flajolet2007} is based on the idea that approximately $\symNumReg  2^{\symRegVal_\symIndexJ}$ distinct element insertions are needed until the value of register $\symIndexJ$ reaches $\symRegVal_\symIndexJ$.
 Given that, a cardinality estimate can be obtained by calculating the average over the values $\lbrace \symNumReg 2^{\symRegVal_1},\ldots,\symNumReg2^{\symRegVal_\symNumReg}\rbrace$.
@@
 \label{equ:intermediate_estimate}
-\symCardinalityOriginalEstimate = \frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
+\symCardinalityOriginalEstimate
+= 
+\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
+= 
+\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}
 \end{equation}
@@
 \end{equation}
-For $\symPrecision \geq 4$ theses values can be approximated as follows
-\begin{equation}
-\symAlpha_{16}\approx 0.673,
-\,
-\symAlpha_{32}\approx 0.697,
-\,
-\symAlpha_{64}\approx 0.709,
-\,
-\symAlpha_\symNumReg\approx 0.7213/(1+1.079/\symNumReg)
-\,\text{for}\,\symNumReg\geq 64 
-\end{equation}
+%=
+%\left(
+%\symNumReg
+%\log(2)
+%\int_0^1
+%v^\symNumReg
+%\frac{2^v}{\left(2^v-1\right)^2}
+%dv
+%\right)^{-1}
+Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}.
+These approximations are used in many HyperLogLog implementations.
+However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision\geq 20$.
+For HyperLogLog sketches that are used in practice with more than 256 registers ($\symPrecision\geq 8$), it is sufficient to use the limit $\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg$ as approximation
+\begin{equation}
+\symAlpha_\symNumReg \approx \symAlpha_\infty = \frac{1}{2\log 2} \approx 0.7213475,
+\end{equation}
+because the corresponding approximation error is already small compared to the estimation error.
 
@@
 However, in case $\symRegRange \geq 30$ 6-bit registers as used in \cite{Heule2013} are required to allow representing all values in the range $[0, \symRegRange+1]$.
-If 64-bit hash values are used as proposed in \cite{Heule2013} $(\symPrecision+\symRegRange=64)$, a correction for large cardinalities is not required.
+If 64-bit hash values are used as proposed in \cite{Heule2013} $(\symPrecision+\symRegRange=64)$, a correction for large cardinalities can be circumvented.
 Cardinalities in the order of $2^{64}$ for which saturated register become a problem never occur in practice.
@@
 
-\section{Cardinality estimation}
-\label{sec:cardinality_estimation}
-In the following we describe a new cardinality estimation method that does not require special handling for small or large cardinalities and which also does not require any empirically determined constants or corrections.
-The method is based on the maximum likelihood method.
+\subsection{Deriviation of the \emph{raw} estimate}
+Assume following integer distribution
+\begin{equation}
+\label{equ:assumed_register_val_distribution}
+\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate) = e^{-\frac{\symPoissonRate}{\symNumReg 2^{ \symRegVal}}}.
+\end{equation}
+Note that, this distribution differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$.
+The expectation of $2^{-\symRegVal}$ is given by
+\begin{equation}
+\symExpectation(2^{-\symRegVal})
+=
+\sum_{\symRegVal = -\infty}^\infty
+2^{-\symRegVal}
+e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}
+\left(1-e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}\right)
+=
+\frac{1}{2}
+\sum_{\symRegVal = -\infty}^\infty
+2^{\symRegVal}
+e^{-\frac{\symPoissonRate}{\symNumReg} 2^{\symRegVal}}
+=
+\frac{\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\left( \symPoissonRate/\symNumReg\right)\right)}{2 \log(2)\symPoissonRate}
+\end{equation}
+where the function 
+\begin{equation}
+\label{equ:power_series_function}
+\symPowerSeriesFunc(\symY):= \log(2) \sum_{\symRegVal = -\infty}^\infty
+2^{\symRegVal+\symY}
+e^{-2^{\symRegVal+\symY}}
+\end{equation}
+ is a smooth periodic function with period 1. Numerical evaluations show that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
+\leq \symPowerSeriesFunc(\symY) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=9.885\cdot 10^{-6}$.
+
+Let $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ be a sample taken from the distribution described by \eqref{equ:assumed_register_val_distribution}.
+For large sample sizes $\symNumReg\rightarrow \infty$ we have asymptotically
+\begin{equation}
+\label{equ:expectation_power_two}
+\symExpectation\left(
+\frac{1}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
+\right)
+\underset{\symNumReg\rightarrow \infty}{=}
+\frac{1}{\symExpectation
+\left(2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}\right)}
+=
+\frac{1}{\symNumReg \symExpectation\left(2^{-\symRegVal}\right)}.
+\end{equation}
+Together with \eqref{equ:expectation_power_two} we get
+\begin{equation}
+\symPoissonRate
+=
+\symExpectation\left(
+\frac{\symNumReg^2 \symPowerSeriesFunc\!\left(\log_2\left( \symPoissonRate/\symNumReg\right)\right)}{2\log(2)\left(2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}\right)}
+\right)
+\quad
+\text{for $\symNumReg\rightarrow \infty$} 
+\end{equation}
+and therefore, the estimator
+\begin{equation}
+\label{equ:rawestimator_2}
+\symPoissonRateEstimate 
+= 
+\frac{\symNumReg^2}{2\log(2)\left(2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}\right)}
+\end{equation}
+for the Poisson parameter, is asymptotically almost unbiased with maximum relative offset $\symEpsPowerSeriesFunc$.
+This estimator corresponds to the \emph{raw} estimate, already presented in \eqref{equ:intermediate_estimate} and can also be written as
+\begin{equation}
+\label{equ:raw_estimate_with_multiplicities}
+\symPoissonRateEstimate 
+= 
+\frac{\symNumReg^2}{2\log(2)\sum_{\symRegVal=-\infty}^\infty \symCount'_\symRegVal 2^{-\symRegVal}}
+\end{equation}
+where $\symCount'_\symRegVal$ is the multiplicity of value $\symRegVal$ in the sample $\left\lbrace \symRegVal_1,\ldots,\symRegVal_\symNumReg\right\rbrace$.
+By defintion, we have $\sum_{\symRegVal=-\infty}^\infty \symCount'_\symRegVal = \symNumReg$.
+
+\subsection{Limitations of the \emph{raw} estimate}
+As we have already seen in \autoref{fig:raw_estimate}, the \emph{raw} estimate does not work very well for small or large cardinalities.
+The reason for this is that the distribution of register values \eqref{equ:register_value_distribution} is different from \eqref{equ:assumed_register_val_distribution} for which the \emph{raw} estimate was derived.
+Since a random variable $\symRegValVariate'$ that obeys \eqref{equ:assumed_register_val_distribution} can be transformed into random variable $\symRegValVariate$ that follows \eqref{equ:register_value_distribution} using the transformation $\symRegValVariate = \min\left(\max\left(\symRegValVariate', 0\right), \symRegRange+1\right)$, the register values of a HyperLogLog sketch can be seen as the result after applying this transformation to a sample distributed according to \eqref{equ:assumed_register_val_distribution}.
+The corresponding multiplicities are related by
+\begin{equation}
+\label{equ:multiplicity_transformation}
+\begin{aligned}
+\symCount_0 &= \textstyle\sum_{\symRegVal = -\infty}^{0} \symCount'_\symRegVal, \\
+\symCount_\symRegVal &=  \symCount'_\symRegVal, \quad \text{for $\symRegVal \in [1, \symRegRange]$},\\
+\symCount_{\symRegRange+1} &= \textstyle\sum_{\symRegVal = \symRegRange + 1}^{\infty} \symCount'_\symRegVal.
+\end{aligned}
+\end{equation}
+Obviously, as long as $\symCount_0\approx\symCount'_0$ and  $\symCount'_{\symRegRange+1}\approx\symCount_{\symRegRange+1}$ the set of register values is similar to the originating sample.
+In case $\symCount_0=0$ and $\symCount_{\symRegRange+1}=0$, we can even be sure that the register values are identical to the originating sample.
+However, for small amd large cardinalities where $\symCount_0$ and $\symCount_{\symRegRange+1}$ approach $\symNumReg$, respectively, the \emph{raw} estimate needs to be corrected.
+
+\subsection{Correcting the \emph{raw} estimate}
+Given a HyperLogLog sketch with its multiplicity vector $\vec{\symCount}$, we try to find estimates $\hat{\symCount}'_\symRegVal$ for $\symCount'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and use them instead in the estimation formula \eqref{equ:raw_estimate_with_multiplicities}.
+For $\symRegVal\in[1,\symRegRange]$ where $\symCount_\symRegVal = \symCount'_\symRegVal$ we have the trivial estimators
+\begin{equation}
+\hat{\symCount}'_\symRegVal := \symCount_\symRegVal\quad \text{for $\symRegVal \in [1, \symRegRange]$}.
+\end{equation}
+To get estimators for all other $\symRegVal$, we consider their expectations
+\begin{equation}
+\symExpectation(\symCount'_\symRegVal)
+=
+\symNumReg e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}\right).
+\end{equation}
+From \eqref{equ:register_value_distribution} we know that $\symExpectation\left(\frac{\symCount_0}{\symNumReg}\right)=
+e^{-\frac{\symPoissonRate}{\symNumReg}}$ and $\symExpectation\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)=
+e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}$, and therefore, we can also write
+\begin{equation}
+\symExpectation(\symCount'_\symRegVal)
+=
+\symNumReg
+\symExpectation\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}
+\left(1-\symExpectation\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
+\end{equation}
+and
+\begin{equation}
+\symExpectation(\symCount'_\symRegVal)
+=
+\symNumReg
+\symExpectation\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}
+\left(1-\symExpectation\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}\right),
+\end{equation}
+which motivates us to use
+\begin{equation}
+\hat{\symCount}'_\symRegVal
+:=
+\symNumReg
+\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}
+\left(1-\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
+\end{equation}
+as estimator for $\symRegVal \leq 0$ and
+\begin{equation}
+\hat{\symCount}'_\symRegVal
+:=
+\symNumReg
+\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}
+\left(1-\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}\right)
+\end{equation}
+as estimator for $\symRegVal \geq \symRegRange+1$, respectively.
+
+Inserting all these estimators into \eqref{equ:raw_estimate_with_multiplicities} gives
+\begin{equation}
+\label{equ:correctedestimator}
+\symPoissonRateEstimate 
+= 
+\frac{\symNumReg^2}{2\log(2)\sum_{\symRegVal=-\infty}^\infty \hat{\symCount}'_\symRegVal 2^{-\symRegVal}}
+=
+\frac{\symNumReg^2}{2\log(2)\left(\symCount_0^{\text{corr}} + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symCount_{\symRegRange+1}^{\text{corr}}2^{-(\symRegRange+1)}\right)}
+\end{equation}
+where $\symCount_0^{\text{corr}}$ and $\symCount_{\symRegRange+1}^{\text{corr}}$ are corrections for $\symCount_0$ and $\symCount_{\symRegRange+1}$, respectively (compare \eqref{equ:intermediate_estimate}).
+They are given by
+\begin{equation}
+\symCount_0^{\text{corr}} := \symNumReg
+\cdot
+\left(
+\frac{\symCount_0}{\symNumReg}
++
+\frac{1}{2}\sum_{\symRegVal=1}^\infty
+\left(\frac{\symCount_0}{\symNumReg}\right)^{2^\symRegVal} 2^{\symRegVal}
+\right)
+\end{equation}
+and
+\begin{equation}
+\symCount_{\symRegRange+1}^{\text{corr}}
+:=
+\symNumReg
+\cdot
+\left(
+\sum_{\symRegVal=1}^\infty
+\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)	^{2^{-\symRegVal}}
+2^{-\symRegVal}
+-
+\left(
+1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}
+\right)
+\right).
+\end{equation}
+
+For the special case of linear probabilistic counting with $\symRegRange=0$, we get for the  estimator \eqref{equ:correctedestimator}
+\begin{equation}
+\symPoissonRateEstimate 
+= 
+\frac{\symNumReg}{\log(2)\sum_{\symRegVal=-\infty}^\infty
+\left(\frac{\symCount_0}{\symNumReg}\right)^{2^\symRegVal}2^\symRegVal}
+=
+\frac{\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right)}{\symPowerSeriesFunc\left(\log_2\left(\log\left(\frac{\symNumReg}{\symCount_0}\right)\right)\right)}
+\approx
+\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right).
+\end{equation}
+Here we used function $\symPowerSeriesFunc$ defined in \eqref{equ:power_series_function} which is known to be approximately equal to 1.
+Hence, for this special case we get as expected almost the same estimator as given in \eqref{equ:small_range_estimate}.
+
+\subsection{Estimation error}
+
+see \autoref{fig:raw_corrected_estimation_error_8_24}, \autoref{fig:raw_corrected_estimation_error_12_20},
+\autoref{fig:raw_corrected_estimation_error_16_16},
+\autoref{fig:raw_corrected_estimation_error_20_12},
+\autoref{fig:raw_corrected_estimation_error_22_42},
+
+\begin{figure}
+\centering
+\includesvg[width=1\textwidth]{raw_corrected_estimate_8_24}
+\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
+\label{fig:raw_corrected_estimation_error_8_24}
+\end{figure}
+
+\begin{figure}
+\centering
+\includesvg[width=1\textwidth]{raw_corrected_estimate_12_20}
+\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
+\label{fig:raw_corrected_estimation_error_12_20}
+\end{figure}
+
+\begin{figure}
+\centering
+\includesvg[width=1\textwidth]{raw_corrected_estimate_16_16}
+\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
+\label{fig:raw_corrected_estimation_error_16_16}
+\end{figure}
+
+\begin{figure}
+\centering
+\includesvg[width=1\textwidth]{raw_corrected_estimate_20_12}
+\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 20$ and $\symRegRange=12$.}
+\label{fig:raw_corrected_estimation_error_20_12}
+\end{figure}
+
+\begin{figure}
+\centering
+\includesvg[width=1\textwidth]{raw_corrected_estimate_22_42}
+\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=42$.}
+\label{fig:raw_corrected_estimation_error_22_42}
+\end{figure}
+
+\section{Maximum likelihood estimation}
 
@@
 In comparison, \algoref{alg:insert} uses only a single hash value that defines which register needs to be updated.
-As a consequence, the register values are statistically dependent, which makes the joint probability distribution of all register values more complex.
-
-In the following we use the Poisson approximation in order to simplify the model.
-Despite this approximation the maximum likelihood estimate will give us high quality estimates for the entire cardinality range as simulation results will show later.
-Furthermore, we will demonstrate that consequent application of the maximum likelihood method to the Poisson model directly reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values.
+As a consequence, the register values are statistically dependent, and the Poisson approximation is needed to make the register values statistically independent, which allows the factorization of the joint probability distribution of all register values.
+
+In the following we derive a new robust and efficient cardinality estimation algorithm that is based on the maximum likelihood method.
+Furthermore, we will demonstrate that consequent application of the maximum likelihood method reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values.
 The history of the HyperLogLog algorithm shows that the \emph{raw} estimate \eqref{alg:estimate_original} was first found after several attempts using the geometric mean \cite{Flajolet2007, Durand2003}.
 
-\subsection{Joint probability distribution of register values}
-
-Under the assumption of an ideal hash function, the probability mass function for the register values $\vec{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
-\begin{equation}
-\label{equ:multinomialProbabilityMass}
-\symProbabilityMass(\vec{\symRegVal}\vert\symCardinality)
-=
-\sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots\symCardinality_\symNumReg}
-\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexJ=1}^\symNumReg \symRegProbability_{\symCardinality_\symIndexJ\symRegVal_\symIndexJ},
-\end{equation}
-where $\symCardinality$ is the cardinality.
-The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities.
-$\symRegProbability_{\symNu\symKappa}$ is the probability that a register is equal to $\symKappa$ after inserting $\symNu$ distinct elements
-\begin{equation}
-\symRegProbability_{\symNu\symKappa} 
-:=
-\begin{cases}
-1 & \symNu=0 \wedge \symKappa = 0\\
-0 & \symNu=0 \wedge 1\leq\symKappa\leq\symRegRange+1\\
-0 & \symNu\geq1 \wedge \symKappa = 0\\
-\left(1-\frac{1}{2^\symKappa}\right)^\symNu - \left(1-\frac{1}{2^{\symKappa-1}}\right)^\symNu & \symNu \geq 1 \wedge 1\leq\symKappa\leq\symRegRange\\
-1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symNu & \symNu\geq 1 \wedge \symKappa = \symRegRange +1 \\
-\end{cases}
-\end{equation}
-Note that the order of register values $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ is not important for the estimation of the cardinality.
-More formally, the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$ is a sufficient statistic for $\symCardinality$.
-Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace = 0^{\symCount_0}1^{\symCount_1}\cdots\symRegRange^{\symCount_{\symRegRange}}(\symRegRange+1)^{\symCount_{\symRegRange+1}}$ where $\symCount_\symIndexJ$ is the multiplicity of value $\symIndexJ$.
-As a consequence, the multiplicity vector $\vec{\symCount} := (\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is also a sufficient statistic for the cardinality.
-In addition, this vector also contains all the information about the HyperLogLog sketch that is required for  cardinality estimation: $\symRegRange=\dim(\vec{\symCount})-2$ and  $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$.
-
-\subsection{Poisson approximation}
-Unfortunately, estimating the cardinality for given register values $\vec{\symRegVal}$ by maximizing the likelihood $\symLikelihood(n \vert \vec{\symRegVal}) = \symProbabilityMass(\vec{\symRegVal}\vert\symCardinality)$ is very expensive.
-Therefore, as proposed in \cite{Flajolet2007} a Poisson model is used instead, which assumes that the cardinality itself is distributed according to a Poisson distribution
-\begin{equation}
-\symCardinality \sim \text{Poisson}(\symPoissonRate).
-\end{equation}
-Under the Poisson model the register values are distributed according to
-\begin{align}
-\symProbabilityMass(\vec{\symRegVal}\vert\symPoissonRate) 
-&= 
-\sum_{\symCardinality=0}^\infty \symProbabilityMass(\vec{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
-\\
-&= 
-\sum_{\symCardinality_1=0}^\infty
-\cdots
-\sum_{\symCardinality_\symNumReg=0}^\infty
-\prod_{\symIndexJ=1}^\symNumReg
-\symRegProbability_
-{\symCardinality_\symIndexJ\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexJ}}{\symCardinality_\symIndexJ!\symNumReg^{\symCardinality_\symIndexJ}}
-\\
-&= 
-\prod_{\symIndexJ=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability_
-{\symCardinality\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
-\\
-&= 
-\prod_{\symRegVal=0}^{\symRegRange+1} \left[
-\sum_{\symCardinality=0}^\infty
-\symRegProbability_
-{\symCardinality\symRegVal}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
-\right]^{\symCount_\symRegVal}
-\\
-&=
-e^{-\symCount_0\frac{\symPoissonRate}{\symNumReg}}
-\cdot
-\left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\symCount_\symRegVal}\right)
-\cdot
-\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\symCount_{\symRegRange+1}}
-\end{align}
-This factorization shows that all register values are independent and identically distributed under the Poisson model.
-The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is
-\begin{equation}
-\label{equ:register_value_distribution}
-\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
-=
-\begin{cases}
-0 & \symRegVal < 0 \\
-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}} & 0\leq \symRegVal \leq \symRegRange \\
-1 & \symRegVal > \symRegRange
-\end{cases}
-\end{equation}
-
-\subsection{Maximum likelihood estimation}
-The key idea of this paper is to find an estimate $\symPoissonRateEstimate$ for the Poisson parameter $\symPoissonRate$ via maximum likelihood estimation and use it directly as the cardinality estimate
-\begin{equation}
-\symCardinalityEstimate_\text{ML} := \symPoissonRateEstimate
-\end{equation}
-
-As supported by simulations that will be presented later, this approximation works well for the entire range of cardinalities.
-In order to get an estimate for $\symPoissonRateEstimate$, maximum-likelihood estimation is used which works out quite well under the Poisson model.
-The log-likelihood and its derivative are given by
+\subsection{Log-likelihood function}
+
+Under the Poisson model the log-likelihood and its derivative are given by
 \begin{equation}
@@
 
-\section{Results}
-
 \subsection{Estimation error}
@@
 \end{figure}
+
+\begin{figure}
+\centering
+\includesvg[width=1\textwidth]{max_likelihood_estimate_22_42}
+\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=42$.}
+\label{fig:estimation_error_22_42}
+\end{figure}
+
+TODO \autoref{fig:estimation_error_22_42}
 
@@
 
+\subsection{Results}
+
+TODO
+
 \section{Conclusion and future work}
+
+TODO
 

### 2016-05-21 c++/benchmark.cpp

@@ -136,6 +136,9 @@ void run(const int p, const int q, const string& resultsFileName) {
         double maxLikelihoodWithoutRegisterScan = 0.;
         double flajoletWithRegisterScan = 0.;
         double flajoletWithoutRegisterScan = 0.;
+        double flajoletCorrectedWithoutRegisterScan = 0.;
+        long flajoletCorrectedSmallCorrectionIterationsSum = 0;
+        long flajoletCorrectedLargeCorrectionIterationsSum = 0;
         long outerLoopIterationsCountSum = 0;
         long innerLoop1IterationsCountSum = 0;
         long innerLoop2IterationsCountSum = 0;
@@ -173,6 +176,21 @@ void run(const int p, const int q, const string& resultsFileName) {
             auto elapsed = chrono::duration_cast<chrono::nanoseconds>(end - start);
             maxLikelihoodWithoutRegisterScan = elapsed.count()/(double)dataSize;
         }
+        
+        {
+            int smallCorrectionIterations;
+            int largeCorrectionIterations;
+            
+            auto start = std::chrono::system_clock::now();
+            for(auto c : counts) {
+                sum += flajoletRawEstimateCorrected(c, smallCorrectionIterations, largeCorrectionIterations);
+                flajoletCorrectedSmallCorrectionIterationsSum += smallCorrectionIterations;
+                flajoletCorrectedLargeCorrectionIterationsSum += largeCorrectionIterations;
+            }
+            auto end = chrono::system_clock::now();
+            auto elapsed = chrono::duration_cast<chrono::nanoseconds>(end - start);
+            flajoletCorrectedWithoutRegisterScan = elapsed.count()/(double)dataSize;
+        }
             
         /*{
             auto start = std::chrono::system_clock::now();
@@ -204,6 +222,9 @@ void run(const int p, const int q, const string& resultsFileName) {
         outputFile << " " << innerLoop1IterationsCountSum/(double)dataSize;
         outputFile << " " << innerLoop2IterationsCountSum/(double)dataSize;
         outputFile << " " << logEvaluationCountSum/(double)dataSize;
+        outputFile << " " << flajoletCorrectedWithoutRegisterScan;
+        outputFile << " " << flajoletCorrectedSmallCorrectionIterationsSum/(double)dataSize;
+        outputFile << " " << flajoletCorrectedLargeCorrectionIterationsSum/(double)dataSize;
         outputFile << endl;
     }
     

### 2016-05-21 c++/cardinalityestimation.hpp

@@ -30,6 +30,8 @@ int getPFromCounts(const std::vector<int>& c) {
     return getPFromNumberRegisters(m);
 }
 
+// TODO define classes with precalculated constants
+
 double maxLikelihoodEstimate(const std::vector<int>& c, int& outerLoopIterationsCount, int& innerLoop1IterationsCount, int& innerLoop2IterationsCount, int& logEvaluationCount, int& kMin, int& kMax) {
 
     const double eps = 1e-2;

### 2016-05-21 python/distribution_charts.py

@@ -75,9 +75,38 @@ def createFigure(p, q, dataSource, figureName):
 createFigure(12, 20, 'flajolet_estimates', 'original_estimate')
 createFigure(12, 20, 'flajolet_mid_range_estimates', 'raw_estimate')
 createFigure(12, 20, 'flajolet_small_range_estimates', 'small_range_estimate')
-createFigure(8, 24, 'max_likelihood_estimates', 'max_likelihood_estimate_8_24')
+
+#createFigure(4, 28,  'max_likelihood_estimates', 'max_likelihood_estimate_4_28')
+#createFigure(6, 26,  'max_likelihood_estimates', 'max_likelihood_estimate_6_26')
+createFigure(8, 24,  'max_likelihood_estimates', 'max_likelihood_estimate_8_24')
 createFigure(12, 20, 'max_likelihood_estimates', 'max_likelihood_estimate_12_20')
 createFigure(16, 16, 'max_likelihood_estimates', 'max_likelihood_estimate_16_16')
 createFigure(20, 12, 'max_likelihood_estimates', 'max_likelihood_estimate_20_12')
+createFigure(22, 42, 'max_likelihood_estimates', 'max_likelihood_estimate_22_42')
 createFigure(12, 52, 'max_likelihood_estimates', 'max_likelihood_estimate_12_52')
 createFigure(12, 14, 'max_likelihood_estimates', 'max_likelihood_estimate_12_14')
+createFigure(12,  0, 'max_likelihood_estimates', 'max_likelihood_estimate_12_0')
+
+
+#createFigure(4, 28,  'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_4_28')
+#createFigure(6, 26,  'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_6_26')
+createFigure(8, 24,  'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_8_24')
+createFigure(12, 20, 'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_12_20')
+createFigure(16, 16, 'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_16_16')
+createFigure(20, 12, 'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_20_12')
+createFigure(22, 42, 'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_22_42')
[…]

### 2016-05-23 paper/paper.tex
implemented estimation algorithm that uses precalculated corrections added new algorithm to paper

@@
 \symDefine{\symPowerSeriesFunc}{\xi}
+\symDefine{\symSmallCorrectionFunc}{\sigma}
+\symDefine{\symLargeCorrectionFunc}{\tau}
 \symDefine{\symEpsPowerSeriesFunc}{\varepsilon_\symPowerSeriesFunc}
@@
 =
-\frac{\symNumReg^2}{2\log(2)\left(\symCount_0^{\text{corr}} + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symCount_{\symRegRange+1}^{\text{corr}}2^{-(\symRegRange+1)}\right)}
-\end{equation}
-where $\symCount_0^{\text{corr}}$ and $\symCount_{\symRegRange+1}^{\text{corr}}$ are corrections for $\symCount_0$ and $\symCount_{\symRegRange+1}$, respectively (compare \eqref{equ:intermediate_estimate}).
-They are given by
-\begin{equation}
-\symCount_0^{\text{corr}} := \symNumReg
+\frac{\symNumReg^2}
+{2\log(2)
+\left(
+\symSmallCorrectionFunc(\symCount_0) + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symLargeCorrectionFunc(\symCount_{\symRegRange+1}) 2^{-(\symRegRange+1)}
+\right)}
+\end{equation}
+where $\symSmallCorrectionFunc(\symCount_0)$ and $\symLargeCorrectionFunc(\symCount_{\symRegRange+1})$ replace $\symCount_0$ and $\symCount_{\symRegRange+1}$ in the \emph{raw} estimate \eqref{equ:intermediate_estimate}, respectively.
+The functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are given by
+\begin{equation}
+\symSmallCorrectionFunc(\symCount) := \symNumReg
 \cdot
 \left(
-\frac{\symCount_0}{\symNumReg}
+\frac{\symCount}{\symNumReg}
 +
 \frac{1}{2}\sum_{\symRegVal=1}^\infty
-\left(\frac{\symCount_0}{\symNumReg}\right)^{2^\symRegVal} 2^{\symRegVal}
+\left(\frac{\symCount}{\symNumReg}\right)^{2^\symRegVal} 2^{\symRegVal}
 \right)
@@
 \begin{equation}
-\symCount_{\symRegRange+1}^{\text{corr}}
+\symLargeCorrectionFunc(\symCount)
 :=
@@
 \sum_{\symRegVal=1}^\infty
-\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)	^{2^{-\symRegVal}}
+\left(1-\frac{\symCount}{\symNumReg}\right)	^{2^{-\symRegVal}}
 2^{-\symRegVal}
@@
 \left(
-1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}
-\right)
-\right).
+1-\frac{\symCount}{\symNumReg}
+\right)
+\right)
+=
+\symNumReg
+\cdot
+\left(
+\frac{\symCount}{\symNumReg}
+-
+\sum_{\symRegVal=1}^\infty
+\left(1-\left(1-\frac{\symCount}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
+2^{-\symRegVal}
+\right)
 \end{equation}
@@
 Hence, for this special case we get as expected almost the same estimator as given in \eqref{equ:small_range_estimate}.
+\subsection{Estimation algorithm}
+\begin{algorithm}
+\caption{Cardinality estimation algorithm based on the corrected \emph{raw} estimate.}
+\label{alg:estimate_raw_corrected}
+\begin{algorithmic}
+\Function {EstimateCardinality}{$\vec{\symCount}$}
+\State $\symNumReg \gets \left\Vert\vec{\symCount}\right\Vert_1$
+\State $\symZ\gets 0.5\cdot \symLargeCorrectionFunc(\symCount_{\symRegRange+1})$
+\For{$\symRegVal\gets \symRegRange, 1$}
+\State $\symZ\gets\symZ + 0.5\cdot \symCount_\symRegVal$
+\EndFor
+\State $\symZ\gets\symZ+\symSmallCorrectionFunc(\symCount_0)$
+\State \Return$\symNumReg^2 / \left(2\log(2)\symZ\right)$
+\EndFunction
+\end{algorithmic}
+\end{algorithm}
 

### 2016-05-23 c++/cardinality_calculator.cpp

@@ -186,7 +186,8 @@ int main(int argc, char* argv[])
 
                 flajoletSmallRangeEstimates[resultPos] = flajoletSmallRangeEstimate(c);
                 flajoletMidRangeEstimates[resultPos] = flajoletRawEstimate(c);
-                flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate(c, numSmallCorrectionIterations[resultPos], numLargeCorrectionIterations[resultPos]);
+                //flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate(c, numSmallCorrectionIterations[resultPos], numLargeCorrectionIterations[resultPos]);
+                flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate2(c);
                 flajoletEstimates[resultPos] = flajoletEstimate(c);
                 maxLikelihoodEstimates[resultPos] = maxLikelihoodEstimator.estimate(c, outerLoopIterationsCount[resultPos], innerLoop1IterationsCount[resultPos], innerLoop2IterationsCount[resultPos], logEvaluationCount[resultPos], kMin, kMax);
                 weakLowerBoundEstimates[resultPos] = weakLowerBoundEstimate(c);

### 2016-05-23 c++/cardinalityestimation.hpp

@@ -179,59 +179,90 @@ double flajoletRawEstimate(const std::vector<int>& c, int m) {
 class CorrectedRawEstimator {
     const int p;
     const int q;
-    const int m;
+    const int m = 1 << p;
+    const double m2alpha = (m/(2.*std::log(2)))*m;
+    const std::vector<double> tauValues = initTau(m);
+    const std::vector<double> sigmaValues = initSigma(m);
+
+    static double sigma(int c, int m, int& numIterations) {
+        numIterations = 0;
+        if (c==m) return std::numeric_limits<double>::infinity();
+        double powerOfX = static_cast<double>(c)/static_cast<double>(m);
+        double result = powerOfX;
+        double previousResult;
+        double powerOf2 = 1;
+        do {
+            numIterations += 1;
+            powerOfX *= powerOfX;
+            previousResult = result;
+            result += powerOfX * powerOf2;
+            powerOf2 += powerOf2;
+        } while(previousResult < result);
+        return m*result;
+    }
+
+    static double tau(int c, int m, int& numIterations) {
+        numIterations = 0;
+        double result = static_cast<double>(c)/static_cast<double>(m);
+        double powerOfX  = 1 - result;
+        double previousResult;
+        double powerOf2 = 0.5;
+        do {
+            numIterations += 1;
+            powerOfX = std::sqrt(powerOfX);
+            previousResult = result;
+            result -= (1 - powerOfX)*powerOf2;
+            powerOf2 *= 0.5;
+        } while(previousResult > result);
+        return m*result;
+    }
+
+    static std::vector<double> initSigma(int m) {
+        int numIterations;
+        std::vector<double> result(m+2);
+        for (int c = 0; c < m+2; ++c) {
+            result.push_back(sigma(c, m, numIterations));
+        }
+        return result;
+    }
+
+    static std::vector<double> initTau(int m) {
+        int numIterations;
+        std::vector<double> result(m+2);
+        for (int c = 0; c < m+2; ++c) {
+            result.push_back(tau(c, m, numIterations));
+        }
+        return result;
+    }
+
 
 public:
 
-    CorrectedRawEstimator(const int p_, const int q_) :p(p_), q(q_), m(1 << p_) {}
+    CorrectedRawEstimator(const int p_, const int q_) :p(p_), q(q_) {}
 
[…]

### 2016-05-26 c++/benchmark.cpp
fixed issue in corrected raw estimate + other changes

@@ -2,7 +2,7 @@
 //# Copyright 2016 Otmar Ertl #
 //#############################
 
-#include "cardinalityestimation.hpp"
+#include "cardinality_estimation.hpp"
 #include "hyperloglog.hpp"
 
 #include <iostream>
@@ -121,8 +121,8 @@ void run(const int p, const int q, const string& resultsFileName) {
 
             vector<int> tmpC = readCounts(line);
             vector<int> c(q+2);
-            for (int i = 0; i < tmpC.size(); ++i) {
-                c[min(i, q+1)] += tmpC[i];
+            for (size_t i = 0; i < tmpC.size(); ++i) {
+                c[min(i, size_t(q+1))] += tmpC[i];
             }
             for (int i = 0; i < numLoops; ++i) {
                 counts[seedCounter + i *numData] = c;
@@ -140,6 +140,7 @@ void run(const int p, const int q, const string& resultsFileName) {
         double flajoletWithRegisterScan = 0.;
         double flajoletWithoutRegisterScan = 0.;
         double flajoletCorrectedWithoutRegisterScan = 0.;
+        double flajoletCorrectedWithoutRegisterScanPrecalculated = 0.;
         long flajoletCorrectedSmallCorrectionIterationsSum = 0;
         long flajoletCorrectedLargeCorrectionIterationsSum = 0;
         long outerLoopIterationsCountSum = 0;
@@ -194,6 +195,16 @@ void run(const int p, const int q, const string& resultsFileName) {
             auto elapsed = chrono::duration_cast<chrono::nanoseconds>(end - start);
             flajoletCorrectedWithoutRegisterScan = elapsed.count()/(double)dataSize;
         }
+        
+        {
+            auto start = std::chrono::system_clock::now();
+            for(auto c : counts) {
+                sum += correctedRawEstimator.estimate2(c);
+            }
+            auto end = chrono::system_clock::now();
+            auto elapsed = chrono::duration_cast<chrono::nanoseconds>(end - start);
+            flajoletCorrectedWithoutRegisterScanPrecalculated = elapsed.count()/(double)dataSize;
+        }
 
         /*{
             auto start = std::chrono::system_clock::now();
@@ -228,6 +239,7 @@ void run(const int p, const int q, const string& resultsFileName) {
         outputFile << " " << flajoletCorrectedWithoutRegisterScan;
[…]

### 2016-05-26 c++/cardinality_calculator.cpp

@@ -2,7 +2,7 @@
 //# Copyright 2016 Otmar Ertl #
 //#############################
 
-#include "cardinalityestimation.hpp"
+#include "cardinality_estimation.hpp"
 #include "hyperloglog.hpp"
 
 #include <iostream>
@@ -175,8 +175,8 @@ int main(int argc, char* argv[])
                 getline(dataFile, line);
                 vector<int> tmpC = readCounts(line);
                 vector<int> c(q+2);
-                for (int i = 0; i < tmpC.size(); ++i) {
-                    c[min(i, q+1)] += tmpC[i];
+                for (size_t i = 0; i < tmpC.size(); ++i) {
+                    c[min(i, size_t(q+1))] += tmpC[i];
                 }
 
                 size_t resultPos = cardinalityIdx*seeds.size()+seedCounter;
@@ -186,8 +186,8 @@ int main(int argc, char* argv[])
 
                 flajoletSmallRangeEstimates[resultPos] = flajoletSmallRangeEstimate(c);
                 flajoletMidRangeEstimates[resultPos] = flajoletRawEstimate(c);
-                //flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate(c, numSmallCorrectionIterations[resultPos], numLargeCorrectionIterations[resultPos]);
-                flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate2(c);
+                flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate(c, numSmallCorrectionIterations[resultPos], numLargeCorrectionIterations[resultPos]);
+                //flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate2(c);
                 flajoletEstimates[resultPos] = flajoletEstimate(c);
                 maxLikelihoodEstimates[resultPos] = maxLikelihoodEstimator.estimate(c, outerLoopIterationsCount[resultPos], innerLoop1IterationsCount[resultPos], innerLoop2IterationsCount[resultPos], logEvaluationCount[resultPos], kMin, kMax);
                 weakLowerBoundEstimates[resultPos] = weakLowerBoundEstimate(c);
@@ -216,7 +216,7 @@ int main(int argc, char* argv[])
         printToFile(filePrefix + "max_likelihood_estimates.dat", &maxLikelihoodEstimates[0], seeds.size(), cardinalities.size());
[…]

### 2016-05-26 c++/cardinality_estimation.hpp

@@ -21,7 +21,7 @@ int getPFromNumberRegisters(int m) {
     int p;
     std::frexp(m, &p);
     p -= 1;
-    assert(m == size_t(1) << p);
+    assert(m == 1 << p);
     return p;
 }
 
@@ -179,10 +179,10 @@ double flajoletRawEstimate(const std::vector<int>& c, int m) {
 class CorrectedRawEstimator {
     const int p;
     const int q;
-    const int m = 1 << p;
-    const double m2alpha = (m/(2.*std::log(2)))*m;
-    const std::vector<double> tauValues = initTau(m);
-    const std::vector<double> sigmaValues = initSigma(m);
+    const int m; // = 1 << p;
+    const double m2alpha; // = (m/(2.*std::log(2)))*m;
+    const std::vector<double> tauValues; // = initTau(m);
+    const std::vector<double> sigmaValues; // = initSigma(m);
 
     static double sigma(int c, int m, int& numIterations) {
         numIterations = 0;
@@ -203,34 +203,36 @@ class CorrectedRawEstimator {
 
     static double tau(int c, int m, int& numIterations) {
         numIterations = 0;
-        double result = static_cast<double>(c)/static_cast<double>(m);
-        double powerOfX  = 1 - result;
+
+        double powerOfX = static_cast<double>(m-c)/static_cast<double>(m);
+
+        double result = 0;
         double previousResult;
-        double powerOf2 = 0.5;
+        double powerOf2 = 1.0;
         do {
             numIterations += 1;
             powerOfX = std::sqrt(powerOfX);
             previousResult = result;
-            result -= (1 - powerOfX)*powerOf2;
+            result += (1 - powerOfX)*powerOfX*powerOf2;
             powerOf2 *= 0.5;
-        } while(previousResult > result);
+        } while(previousResult < result);
         return m*result;
     }
 
     static std::vector<double> initSigma(int m) {
         int numIterations;
-        std::vector<double> result(m+2);
-        for (int c = 0; c < m+2; ++c) {
-            result.push_back(sigma(c, m, numIterations));
+        std::vector<double> result(m+1);
+        for (int c = 0; c <= m; ++c) {
+            result[c] = sigma(c, m, numIterations);
         }
         return result;
     }
 
[…]

### 2016-05-26 c++/hyperloglog.hpp

@@ -117,9 +117,9 @@ public:
     //
     // the corresponding  count matrix is
     //
-    //                / 0 0 2 \
-    // count matrix = | 0 1 0 |
-    //                \ 1 0 0 /
+    //                ( 0 0 2 )
+    // count matrix = ( 0 1 0 )
+    //                ( 1 0 0 )
     //
     // or as array representation
     //

### 2016-05-26 c++/intersection_calculation.cpp

@@ -2,7 +2,7 @@
 //# Copyright 2016 Otmar Ertl #
 //#############################
 
-#include "cardinalityestimation.hpp"
+#include "cardinality_estimation.hpp"
 #include "hyperloglog.hpp"
 
 #include <iostream>
@@ -37,251 +37,251 @@ const string resultsFileName = intersectDataPathName + "results.dat";
 
 
 template <typename T> void printToFile(const string& fileName, const T* data, size_t numSketches, bool append) {
-	ofstream file(fileName, (append)?(ofstream::out | ofstream::app):ofstream::out);
-	for (size_t s = 0; s < numSketches; ++s) {
-		if (s>0) {
-			file << " ";
-		}
-		file << data[s];
-	}
-	file << endl;
+    ofstream file(fileName, (append)?(ofstream::out | ofstream::app):ofstream::out);
+    for (size_t s = 0; s < numSketches; ++s) {
+        if (s>0) {
+            file << " ";
+        }
+        file << data[s];
+    }
+    file << endl;
 }
 
 vector<int> readCounts(string s) {
-	stringstream stream(s);
-	int i;
-	vector<int> v;
-	while(stream >> i) {
-		v.push_back(i);
-	}
-	v.shrink_to_fit();
-	return v;
+    stringstream stream(s);
+    int i;
+    vector<int> v;
+    while(stream >> i) {
+        v.push_back(i);
+    }
+    v.shrink_to_fit();
+    return v;
 }
 
 int main(int argc, char* argv[])
 {
-	
-	vector<string> jointCardFileNames;
-	{
-		ifstream jointCardFileNamesFile(jointCardFileName);
-		
-		string line;
-		while (getline(jointCardFileNamesFile, line))
-		{
-			jointCardFileNames.push_back(line);
-		}
-	}
-	
-	cout << "jointCardFileNames size "  << jointCardFileNames.size() << endl;
-	
-	ofstream resultsFile(resultsFileName);
-	
-	resultsFile << "trueCardA;";
-	resultsFile << "trueCardB;";
-	resultsFile << "trueCardX;";
-	
-	resultsFile << "jaccardIndex;";
-	resultsFile << "logRatio;";
-	
[…]

### 2016-05-26 python/distribution_charts.py

@@ -90,14 +90,14 @@ createFigure(12,  0, 'max_likelihood_estimates', 'max_likelihood_estimate_12_0')
 
 #createFigure(4, 28,  'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_4_28')
 #createFigure(6, 26,  'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_6_26')
-createFigure(8, 24,  'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_8_24')
-createFigure(12, 20, 'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_12_20')
-createFigure(16, 16, 'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_16_16')
-createFigure(20, 12, 'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_20_12')
-createFigure(22, 42, 'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_22_42')
-createFigure(12, 52, 'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_12_52')
-createFigure(12, 14, 'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_12_14')
-createFigure(12, 0,  'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_12_0')
+createFigure(8, 24,  'corrected_raw_estimates', 'raw_corrected_estimate_8_24')
+createFigure(12, 20, 'corrected_raw_estimates', 'raw_corrected_estimate_12_20')
+createFigure(16, 16, 'corrected_raw_estimates', 'raw_corrected_estimate_16_16')
+createFigure(20, 12, 'corrected_raw_estimates', 'raw_corrected_estimate_20_12')
+createFigure(22, 42, 'corrected_raw_estimates', 'raw_corrected_estimate_22_42')
+createFigure(12, 52, 'corrected_raw_estimates', 'raw_corrected_estimate_12_52')
+createFigure(12, 14, 'corrected_raw_estimates', 'raw_corrected_estimate_12_14')
+createFigure(12, 0,  'corrected_raw_estimates', 'raw_corrected_estimate_12_0')
 
 #createFigure(4, 28,  'flajolet_mid_range_estimates', 'mid_range_estimates_4_28')
 #createFigure(6, 26,  'flajolet_mid_range_estimates', 'mid_range_estimates_6_26')

### 2016-05-26 c++/xxx.cpp
benchmark chart for corrected raw estimation algorithm

@@ -1,0 +1,78 @@
+#include <vector>
+#include <cassert>
+#include <cmath>
+#include <limits>
+#include <algorithm>
+#include <iostream>
+#include <cstdlib>
+
+using namespace std;
+
+
+double sigma(int c, int m, int& numIterations) {
+        numIterations = 0;
+        if (c==m) return std::numeric_limits<double>::infinity();
+        double powerOfX = static_cast<double>(c)/static_cast<double>(m);
+        double result = powerOfX;
+        double previousResult;
+        double powerOf2 = 1;
+        do {
+            numIterations += 1;
+            powerOfX *= powerOfX;
+            previousResult = result;
+            result += powerOfX * powerOf2;
+            powerOf2 += powerOf2;
+        } while(previousResult < result);
+        return m*result;
+    }
+double tau(int c, int m, int& numIterations) {
+	numIterations = 0;
+	double result = static_cast<double>(c)/static_cast<double>(m);
+	double powerOfX  = 1 - result;
+	double previousResult;
+	double powerOf2 = 0.5;
+	do {
+		numIterations += 1;
+		powerOfX = std::sqrt(powerOfX);
+		previousResult = result;
+		result -= (1 - powerOfX)*powerOf2;
+		powerOf2 *= 0.5;
+	} while(previousResult > result);
+	return m*result;
+}
+
+double tau2(int c, int m, int& numIterations) {
+	numIterations = 0;
+	
+	double powerOfX = static_cast<double>(m-c)/static_cast<double>(m);
+	
+	double result = 0;
+	double previousResult;
+	double powerOf2 = 1.0;
+	do {
+		numIterations += 1;
+		powerOfX = std::sqrt(powerOfX);
+		previousResult = result;
+		result += (1 - powerOfX)*powerOfX*powerOf2;
+		powerOf2 *= 0.5;
+	} while(previousResult < result);
+	return m*result;
+}
+
+int main(int argc, char* argv[]) {
+	int m = 4096;
+
+	for (int c =0; c <= m; ++c) {
+		
+		int numIterations;
+		int numIterations2;
+		
+		double t = sigma(c, m, numIterations);
+		double t2 = tau2(c, m, numIterations2);
+		
+		cout << c << " " << t << " " << t2 << " " << numIterations << " " << numIterations2 << endl;
+	}
+
+
+	return 0;
+}

### 2016-05-26 python/benchmark-charts.py

@@ -29,23 +29,31 @@ data_12_20 = readData(filename_12_20)
 data_12_52 = readData(filename_12_52)
 
 cardinalities_12_20 = data_12_20[:,0]
-avg_exec_time_nanos_12_20 = data_12_20[:,1]
-avg_outer_loop_count_12_20 = data_12_20[:,2]
-avg_inner_loop1_count_12_20 = data_12_20[:,3]
-avg_inner_loop2_count_12_20 = data_12_20[:,4]
-avg_log_eval_count_12_20 = data_12_20[:,5]
+max_likelihood_avg_exec_time_nanos_12_20 = data_12_20[:,1]
+max_likelihood_avg_outer_loop_count_12_20 = data_12_20[:,2]
+max_likelihood_avg_inner_loop1_count_12_20 = data_12_20[:,3]
+max_likelihood_avg_inner_loop2_count_12_20 = data_12_20[:,4]
+max_likelihood_avg_log_eval_count_12_20 = data_12_20[:,5]
+corrected_raw_1_avg_exec_time_nanos_12_20 = data_12_20[:,6]
+corrected_raw_small_iter_12_20 = data_12_20[:,7]
+corrected_raw_small_iter_12_20 = data_12_20[:,8]
+corrected_raw_2_avg_exec_time_nanos_12_20 = data_12_20[:,9]
 
 cardinalities_12_52 = data_12_52[:,0]
-avg_exec_time_nanos_12_52 = data_12_52[:,1]
-avg_outer_loop_count_12_52 = data_12_52[:,2]
-avg_inner_loop1_count_12_52 = data_12_52[:,3]
-avg_inner_loop2_count_12_52 = data_12_52[:,4]
-avg_log_eval_count_12_52 = data_12_52[:,5]
+max_likelihood_avg_exec_time_nanos_12_52 = data_12_52[:,1]
+max_likelihood_avg_outer_loop_count_12_52 = data_12_52[:,2]
+max_likelihood_avg_inner_loop1_count_12_52 = data_12_52[:,3]
+max_likelihood_avg_inner_loop2_count_12_52 = data_12_52[:,4]
+max_likelihood_avg_log_eval_count_12_52 = data_12_52[:,5]
+corrected_raw_1_avg_exec_time_nanos_12_52 = data_12_52[:,6]
+corrected_raw_small_iter_12_52 = data_12_52[:,7]
[…]

### 2016-05-26 paper/paper.tex
fixed issue in corrected raw estimate + other changes
benchmark chart for corrected raw estimation algorithm
* completed new corrected raw estimation algorithm * added corresponding benchmark chart * modified paper title

@@
 
-\title{Cardinality estimation methods for HyperLogLog sketches}
+\title{New cardinality estimation algorithms for HyperLogLog sketches}
 \subtitle{Draft version}
@@
 :=
-\symNumReg
-\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}
+\symNumReg\cdot
+\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}\cdot
 \left(1-\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
@@
 :=
-\symNumReg
+\symNumReg\cdot
 \left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}
+\cdot
 \left(1-\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}\right)
@@
 \left(
-\symSmallCorrectionFunc(\symCount_0) + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symLargeCorrectionFunc(\symCount_{\symRegRange+1}) 2^{-(\symRegRange+1)}
+\symSmallCorrectionFunc(\symCount_0,\symNumReg) + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symLargeCorrectionFunc(\symCount_{\symRegRange+1},\symNumReg) 2^{-(\symRegRange+1)}
 \right)}
@@
 \begin{equation}
-\symSmallCorrectionFunc(\symCount) := \symNumReg
+\symSmallCorrectionFunc(\symCount,\symNumReg) := \symNumReg
 \cdot
@@
 +
-\frac{1}{2}\sum_{\symRegVal=1}^\infty
-\left(\frac{\symCount}{\symNumReg}\right)^{2^\symRegVal} 2^{\symRegVal}
+\sum_{\symRegVal=1}^\infty
+\left(\frac{\symCount}{\symNumReg}\right)^{2^\symRegVal} 2^{\symRegVal-1}
 \right)
@@
 \begin{equation}
-\symLargeCorrectionFunc(\symCount)
+\symLargeCorrectionFunc(\symCount,\symNumReg)
 :=
@@
 \cdot
-\left(
 \sum_{\symRegVal=1}^\infty
-\left(1-\frac{\symCount}{\symNumReg}\right)	^{2^{-\symRegVal}}
-2^{-\symRegVal}
--
-\left(
-1-\frac{\symCount}{\symNumReg}
-\right)
-\right)
-=
-\symNumReg
+\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{-\symRegVal}}
 \cdot
-\left(
-\frac{\symCount}{\symNumReg}
--
-\sum_{\symRegVal=1}^\infty
-\left(1-\left(1-\frac{\symCount}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
-2^{-\symRegVal}
-\right)
+\left(1-\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
+\cdot
+2^{-(\symRegVal-1)}
 \end{equation}
@@
 \subsection{Estimation algorithm}
+see \algoref{alg:corrected_raw_estimation}
+
 \begin{algorithm}
 \caption{Cardinality estimation algorithm based on the corrected \emph{raw} estimate.}
-\label{alg:estimate_raw_corrected}
+\label{alg:corrected_raw_estimation}
 \begin{algorithmic}
@@
 \State $\symNumReg \gets \left\Vert\vec{\symCount}\right\Vert_1$
-\State $\symZ\gets 0.5\cdot \symLargeCorrectionFunc(\symCount_{\symRegRange+1})$
+\State $\symZ\gets 0.5\cdot \symLargeCorrectionFunc(\symCount_{\symRegRange+1},\symNumReg)$
 \For{$\symRegVal\gets \symRegRange, 1$}
-\State $\symZ\gets\symZ + 0.5\cdot \symCount_\symRegVal$
+\State $\symZ\gets0.5\cdot\left(\symZ + \symCount_\symRegVal\right)$
 \EndFor
-\State $\symZ\gets\symZ+\symSmallCorrectionFunc(\symCount_0)$
+\State $\symZ\gets\symZ+\symSmallCorrectionFunc(\symCount_0,\symNumReg)$
 \State \Return$\symNumReg^2 / \left(2\log(2)\symZ\right)$
+\EndFunction
+\\
+\Function {$\symSmallCorrectionFunc$}{$\symCount,\symNumReg$}
+\If{$\symCount=\symNumReg$}
+\State\Return$\infty$
+\EndIf
+\State $\symX\gets\symCount/\symNumReg$
+\State $\symY\gets1$
+\State $\symZ\gets\symX$
+\Repeat
+\State $\symX\gets\symX\cdot\symX$
+\State $\symZ'\gets\symZ$
+\State $\symZ\gets\symZ + \symX\cdot\symY$
+\State $\symY\gets2\cdot\symY$
+\Until{$\symZ=\symZ'$}
+\State\Return$\symNumReg\cdot\symZ$
+\EndFunction
+\\
+\Function {$\symLargeCorrectionFunc$}{$\symCount,\symNumReg$}
+\State $\symX\gets(\symNumReg-\symCount)/\symNumReg$
+\State $\symY\gets1$
+\State $\symZ\gets0$
+\Repeat
+\State $\symX\gets\sqrt{\symX}$
+\State $\symZ'\gets\symZ$
+\State $\symZ\gets\symZ + \left(1-\symX\right)\cdot\symX\cdot\symY$
+\State $\symY\gets0.5\cdot\symY$
+\Until{$\symZ=\symZ'$}
+\State\Return$\symNumReg\cdot\symZ$
 \EndFunction
@@
 \label{fig:raw_corrected_estimation_error_22_42}
+\end{figure}
+
+\subsection{Performance}
+see \autoref{fig:corrected_raw_avg_exec_time}
+
+\begin{figure}
+\centering
+\includesvg[width=1\textwidth]{corrected_raw_avg_exec_time}
+\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.
+Both cases, $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ precalculated or calculated on-demand have been considered.}
+\label{fig:corrected_raw_avg_exec_time}
 \end{figure}
@@
 \centering
-\includesvg[width=1\textwidth]{avg_exec_time}
+\includesvg[width=1\textwidth]{max_likelihood_avg_exec_time}
 \caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}

### 2016-05-26 c++/cardinality_estimation.hpp

@@ -187,10 +187,10 @@ class CorrectedRawEstimator {
     static double sigma(int c, int m, int& numIterations) {
         numIterations = 0;
         if (c==m) return std::numeric_limits<double>::infinity();
-        double powerOfX = static_cast<double>(c)/static_cast<double>(m);
-        double result = powerOfX;
         double previousResult;
+        double powerOfX = static_cast<double>(c)/static_cast<double>(m);
         double powerOf2 = 1;
+        double result = powerOfX;
         do {
             numIterations += 1;
             powerOfX *= powerOfX;
@@ -204,11 +204,10 @@ class CorrectedRawEstimator {
     static double tau(int c, int m, int& numIterations) {
         numIterations = 0;
 
-        double powerOfX = static_cast<double>(m-c)/static_cast<double>(m);
-
-        double result = 0;
         double previousResult;
+        double powerOfX = static_cast<double>(m-c)/static_cast<double>(m);
         double powerOf2 = 1.0;
+        double result = 0;
         do {
             numIterations += 1;
             powerOfX = std::sqrt(powerOfX);

### 2016-05-28 paper/paper.tex
changed definition of tau and sigma

@@
 \left(
-\symSmallCorrectionFunc(\symCount_0,\symNumReg) + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symLargeCorrectionFunc(\symCount_{\symRegRange+1},\symNumReg) 2^{-(\symRegRange+1)}
+\symNumReg\,\symSmallCorrectionFunc(\symCount_0/\symNumReg) + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symNumReg\, \symLargeCorrectionFunc(1-\symCount_{\symRegRange+1}/\symNumReg) 2^{-(\symRegRange+1)}
 \right)}
 \end{equation}
-where $\symSmallCorrectionFunc(\symCount_0)$ and $\symLargeCorrectionFunc(\symCount_{\symRegRange+1})$ replace $\symCount_0$ and $\symCount_{\symRegRange+1}$ in the \emph{raw} estimate \eqref{equ:intermediate_estimate}, respectively.
-The functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are given by
-\begin{equation}
-\symSmallCorrectionFunc(\symCount,\symNumReg) := \symNumReg
+where $\symNumReg\,\symSmallCorrectionFunc(\symCount_0/\symNumReg)$ and $\symNumReg\,\symLargeCorrectionFunc(1-\symCount_{\symRegRange+1}/\symNumReg)$ replace $\symCount_0$ and $\symCount_{\symRegRange+1}$ in the \emph{raw} estimate \eqref{equ:intermediate_estimate}, respectively.
+The functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are defined as
+\begin{equation}
+\symSmallCorrectionFunc(\symX) := 
+\symX
++
+\sum_{\symRegVal=1}^\infty
+\symX^{2^\symRegVal} 2^{\symRegVal-1}
+\end{equation}
+and
+\begin{equation}
+\symLargeCorrectionFunc(\symX)
+:=
+\sum_{\symRegVal=1}^\infty
+\symX^{2^{-\symRegVal}}
 \cdot
-\left(
-\frac{\symCount}{\symNumReg}
-+
-\sum_{\symRegVal=1}^\infty
-\left(\frac{\symCount}{\symNumReg}\right)^{2^\symRegVal} 2^{\symRegVal-1}
-\right)
-\end{equation}
-and
-\begin{equation}
-\symLargeCorrectionFunc(\symCount,\symNumReg)
-:=
-\symNumReg
+\left(1-\symX^{2^{-\symRegVal}}\right)
 \cdot
-\sum_{\symRegVal=1}^\infty
-\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{-\symRegVal}}
-\cdot
-\left(1-\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
-\cdot
-2^{-(\symRegVal-1)}
-\end{equation}
-
-For the special case of linear probabilistic counting with $\symRegRange=0$, we get for the  estimator \eqref{equ:correctedestimator}
+2^{-(\symRegVal-1)}.
+\end{equation}
+
+Note that, $2\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = -\symPowerSeriesFunc(\log_2(-\log(\symX)))/\left(\log(2)\log(\symX)\right)$.
+Therefore, we get for the special case of linear probabilistic counting with $\symRegRange=0$
 \begin{equation}
@@
 = 
-\frac{\symNumReg}{\log(2)\sum_{\symRegVal=-\infty}^\infty
-\left(\frac{\symCount_0}{\symNumReg}\right)^{2^\symRegVal}2^\symRegVal}
+\frac{\symNumReg}{\log(2)\left(
+2\symSmallCorrectionFunc(\symCount_0/\symNumReg)
++
+\symLargeCorrectionFunc(\symCount_0/\symNumReg)
+\right)}
 =
@@
 \end{equation}
-Here we used function $\symPowerSeriesFunc$ defined in \eqref{equ:power_series_function} which is known to be approximately equal to 1.
-Hence, for this special case we get as expected almost the same estimator as given in \eqref{equ:small_range_estimate}.
+almost the same estimator as given in \eqref{equ:small_range_estimate}. Here we used that $\symPowerSeriesFunc$ is known to be approximately equal to 1.
 \subsection{Estimation algorithm}

### 2016-05-28 c++/cardinality_estimation.hpp

@@ -179,16 +179,16 @@ double flajoletRawEstimate(const std::vector<int>& c, int m) {
 class CorrectedRawEstimator {
     const int p;
     const int q;
-    const int m; // = 1 << p;
-    const double m2alpha; // = (m/(2.*std::log(2)))*m;
-    const std::vector<double> tauValues; // = initTau(m);
-    const std::vector<double> sigmaValues; // = initSigma(m);
+    const int m = 1 << p;
+    const double m2alpha = (m/(2.*std::log(2)))*m;
+    const std::vector<double> tauValues = initTau(m);
+    const std::vector<double> sigmaValues = initSigma(m);
 
-    static double sigma(int c, int m, int& numIterations) {
+    static double sigma(double x, int& numIterations) {
         numIterations = 0;
-        if (c==m) return std::numeric_limits<double>::infinity();
+        if (x == 1.) return std::numeric_limits<double>::infinity();
         double previousResult;
-        double powerOfX = static_cast<double>(c)/static_cast<double>(m);
+        double powerOfX = x;
         double powerOf2 = 1;
         double result = powerOfX;
         do {
@@ -198,14 +198,13 @@ class CorrectedRawEstimator {
             result += powerOfX * powerOf2;
             powerOf2 += powerOf2;
         } while(previousResult < result);
-        return m*result;
+        return result;
     }
 
-    static double tau(int c, int m, int& numIterations) {
+    static double tau(double x, int& numIterations) {
         numIterations = 0;
-
         double previousResult;
-        double powerOfX = static_cast<double>(m-c)/static_cast<double>(m);
+        double powerOfX = x;
         double powerOf2 = 1.0;
         double result = 0;
         do {
@@ -215,14 +214,14 @@ class CorrectedRawEstimator {
             result += (1 - powerOfX)*powerOfX*powerOf2;
             powerOf2 *= 0.5;
         } while(previousResult < result);
-        return m*result;
+        return result;
     }
 
     static std::vector<double> initSigma(int m) {
         int numIterations;
         std::vector<double> result(m+1);
         for (int c = 0; c <= m; ++c) {
-            result[c] = sigma(c, m, numIterations);
+            result[c] = m * sigma(static_cast<double>(c)/static_cast<double>(m), numIterations);
         }
         return result;
     }
[…]

### 2016-05-28 c++/benchmark.cpp
fixed algorithm in paper

@@ -187,7 +187,7 @@ void run(const int p, const int q, const string& resultsFileName) {
 
             auto start = std::chrono::system_clock::now();
             for(auto c : counts) {
-                sum += correctedRawEstimator.estimate(c, smallCorrectionIterations, largeCorrectionIterations);
+                sum += correctedRawEstimator.estimate_on_demand(c, smallCorrectionIterations, largeCorrectionIterations);
                 flajoletCorrectedSmallCorrectionIterationsSum += smallCorrectionIterations;
                 flajoletCorrectedLargeCorrectionIterationsSum += largeCorrectionIterations;
             }
@@ -199,7 +199,7 @@ void run(const int p, const int q, const string& resultsFileName) {
         {
             auto start = std::chrono::system_clock::now();
             for(auto c : counts) {
-                sum += correctedRawEstimator.estimate2(c);
+                sum += correctedRawEstimator.estimate_precalculated(c);
             }
             auto end = chrono::system_clock::now();
             auto elapsed = chrono::duration_cast<chrono::nanoseconds>(end - start);

### 2016-05-28 c++/cardinality_calculator.cpp

@@ -186,7 +186,7 @@ int main(int argc, char* argv[])
 
                 flajoletSmallRangeEstimates[resultPos] = flajoletSmallRangeEstimate(c);
                 flajoletMidRangeEstimates[resultPos] = flajoletRawEstimate(c);
-                flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate(c, numSmallCorrectionIterations[resultPos], numLargeCorrectionIterations[resultPos]);
+                flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate_on_demand(c, numSmallCorrectionIterations[resultPos], numLargeCorrectionIterations[resultPos]);
                 //flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate2(c);
                 flajoletEstimates[resultPos] = flajoletEstimate(c);
                 maxLikelihoodEstimates[resultPos] = maxLikelihoodEstimator.estimate(c, outerLoopIterationsCount[resultPos], innerLoop1IterationsCount[resultPos], innerLoop2IterationsCount[resultPos], logEvaluationCount[resultPos], kMin, kMax);

### 2016-05-28 c++/cardinality_estimation.hpp

@@ -187,34 +187,32 @@ class CorrectedRawEstimator {
     static double sigma(double x, int& numIterations) {
         numIterations = 0;
         if (x == 1.) return std::numeric_limits<double>::infinity();
-        double previousResult;
-        double powerOfX = x;
-        double powerOf2 = 1;
-        double result = powerOfX;
+        double zPrime;
+        double y = 1;
+        double z = x;
         do {
             numIterations += 1;
-            powerOfX *= powerOfX;
-            previousResult = result;
-            result += powerOfX * powerOf2;
-            powerOf2 += powerOf2;
-        } while(previousResult < result);
-        return result;
+            x *= x;
+            zPrime = z;
+            z += x * y;
+            y += y;
+        } while(zPrime < z);
+        return z;
     }
 
     static double tau(double x, int& numIterations) {
         numIterations = 0;
-        double previousResult;
-        double powerOfX = x;
-        double powerOf2 = 1.0;
-        double result = 0;
+        double zPrime;
+        double y = 1.0;
+        double z = 0;
         do {
             numIterations += 1;
-            powerOfX = std::sqrt(powerOfX);
-            previousResult = result;
-            result += (1 - powerOfX)*powerOfX*powerOf2;
-            powerOf2 *= 0.5;
-        } while(previousResult < result);
-        return result;
+            x = std::sqrt(x);
+            zPrime = z;
+            z += (1 - x)*x*y;
+            y *= 0.5;
+        } while(zPrime < z);
+        return z;
     }
 
     static std::vector<double> initSigma(int m) {
@@ -230,7 +228,7 @@ class CorrectedRawEstimator {
         int numIterations;
         std::vector<double> result(m+1);
         for (int c = 0; c <= m; ++c) {
-            result[c] = m * tau(static_cast<double>(m-c)/static_cast<double>(m), numIterations);
+            result[c] = 0.5 * m * tau(static_cast<double>(m-c)/static_cast<double>(m), numIterations);
         }
         return result;
     }
@@ -240,7 +238,7 @@ public:
 
     CorrectedRawEstimator(const int p_, const int q_) : p(p_), q(q_) {}
 
-    double estimate(const std::vector<int>& c, int& numSmallCorrectionIterations, int& numLargeCorrectionIterations) const {
[…]

### 2016-05-29 paper/paper.tex
fixed algorithm in paper
use cleveref package, revised chapter 3

@@
 \usepackage{svg}
-\usepackage{hyperref}
 \usepackage{algpseudocode,algorithm,algorithmicx}
@@
 % These packages are all incorporated in the memoir class to one degree or another...
-
-\renewcommand{\sffamily}{\rmfamily}
-
-% ALGORITHMS
-\newcommand*{\algoref}[1]{\hyperref[#1]{Algorithm~\ref*{#1}}}
+\usepackage{hyperref}
+\usepackage[capitalise,nameinlink,noabbrev]{cleveref}
+
+% \renewcommand{\sffamily}{\rmfamily}
 
@@
 For performance reasons the number of registers is chosen to be a power of 2: $\symNumReg = 2^\symPrecision$.
-$\symPrecision$ is the precision that directly influences the relative error scaling like $1/\sqrt{\symNumReg}$.
-The number of bits per register defines the value range that can be represented.
-Initially all registers are set equal to 0.
+$\symPrecision$ is the precision that directly influences the relative error which scales like $1/\sqrt{\symNumReg}$.
+All registers start with the initial value 0.
+The HyperLogLog algorithm successively increases the register values up to a maximum value limited by the space per register which is usually between 5 and 8 bits.
 
 \subsection{Data element insertion}
+\label{sec:data_element_insertion}
 In order to insert a data element into a HyperLogLog data structure a hash value is calculated.
@@
 If this position exceeds the value of the selected register, the register value is replaced.
-\algoref{alg:insert} shows the update procedure for inserting a data element into the HyperLogLog sketch.
+\cref{alg:insert} shows the update procedure for inserting a data element into the HyperLogLog sketch.
 
@@
 In this case each register value is represented by a single bit.
-Hence, linear probabilistic counting can be regarded as a special case of the HyperLogLog algorithm, where $\symRegRange = 0$.
+Hence, linear probabilistic counting can be regarded as a special case of the HyperLogLog algorithm ($\symRegRange = 0$).
 
 \begin{algorithm}
-\caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers $\vec{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg$) initialized to 0 that can take all integers in $[0, \symRegRange+1]$.}
+\caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers $\vec{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg$) which are initialized to 0 and which can represent any integer in range $[0, \symRegRange+1]$.}
 \label{alg:insert}
@@
 As a consequence, the multiplicity vector $\vec{\symCount} := (\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is also a sufficient statistic for the cardinality.
-In addition, this vector also contains all the information about the HyperLogLog sketch that is required for  cardinality estimation: $\symRegRange=\dim(\vec{\symCount})-2$ and  $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$.
+In addition, this vector also contains all the information about the HyperLogLog sketch that is required for  cardinality estimation: $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$ and $\symRegRange=\dim(\vec{\symCount})-2$.
 
 \subsection{Poisson approximation}
+\label{sec:poisson_approximation}
 Unfortunately, the probability mass function \eqref{equ:multinomialProbabilityMass} is difficult to analyze, because the register values are statistically dependent.
@@
 \end{equation}
-Under the Poisson model the register values are distributed according to
+Under the Poisson model the distribution of the register values is
 \begin{align}
@@
 \cdot
-\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\symCount_{\symRegRange+1}}
+\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\symCount_{\symRegRange+1}}.
 \end{align}
-This factorization shows that all register values are now independent and identically distributed under the Poisson model.
+The final factorization shows that all register values are independent and identically distributed under the Poisson model.
 The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is given by
@@
 Due to the simpler probability mass function, it is easier to find an estimate  $\symPoissonRateEstimate$ for the Poisson rate $\symPoissonRate$ than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}.
-Since, the expectation of the cardinality under the Poisson model is equal to $\symPoissonRate$, it makes sense to take $\symPoissonRateEstimate$ direclty as estimate for the cardinality, hence, $\symCardinalityEstimate := \symPoissonRateEstimate$.
+Since, the expectation of the cardinality under the Poisson model is equal to $\symPoissonRate$, it makes sense to use $\symPoissonRateEstimate$ directly as estimate for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$.
 The Poisson approximation is known to work well for large rates.
@@
 Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers.
-It is used for the \emph{raw} estimate of the HyperLogLog algorithm and is given by
+The result is the so-called raw HyperLogLog estimate and given by
 \begin{equation}
@@
 du
-\right)^{-1}
-\end{equation}
-%=
-%\left(
-%\symNumReg
-%\log(2)
-%\int_0^1
-%v^\symNumReg
-%\frac{2^v}{\left(2^v-1\right)^2}
-%dv
-%\right)^{-1}
+\right)^{-1}.
+\end{equation}
 Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}.
@@
 However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision\geq 20$.
-For HyperLogLog sketches that are used in practice with more than 256 registers ($\symPrecision\geq 8$), it is sufficient to use the limit $\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg$ as approximation
-\begin{equation}
-\symAlpha_\symNumReg \approx \symAlpha_\infty = \frac{1}{2\log 2} \approx 0.7213475,
-\end{equation}
-because the corresponding approximation error is already small compared to the estimation error.
-
-\autoref{fig:raw_estimate} shows the distribution of the relative error of the \emph{raw} estimate for 10\,000 randomly generated HyperLogLog sketches over the number of distinct element insertions.
-Obviously, the \emph{raw} estimate is biased for small and large cardinalities and fails to return accurate estimates.
-As a consequence, to cover the entire range of cardinalities, a small and a large correction have been proposed, respectively.
-
-For small cardinalities the same estimator as used by the linear probabilistic counting algorithm \cite{Whang1990} which only incorporates the number of registers equal to 0 denoted by $\symCount_0$:
+For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use
+\begin{equation}
+\label{equ:alpha_inf}
+\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
+\end{equation}
+as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible when compared to the estimation error.
+
+\cref{fig:raw_estimate} shows the distribution of the relative error for the raw estimate as function of the cardinality.
+The chart is based on 10\,000 randomly generated HyperLogLog sketches.
+Obviously, the raw estimate is biased for small and large cardinalities and fails to return accurate estimates.
+Therefore, to cover the entire range of cardinalities, corrections for small and large cardinalities have been proposed, respectively.
+
+As mentioned in \cref{sec:data_element_insertion}, a HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be turned into a $(\symPrecision, 0)$-HyperLogLog sketch.
+Since $\symRegRange=0$ corresponds to linear probabilistic counting and the transformed HyperLogLog sketch corresponds to a bitset with $\symCount_0$ zeros, we can apply the corresponding cardinality estimator \cite{Whang1990}
 \begin{equation}
@@
 \frac{\symNumReg}{\symCount_0}
-\right)
-\end{equation}
-The corresponding relative estimation as depicted in \autoref{fig:small_range_estimate} shows that this estimator is convenient for small cardinalities.
-It was proposed to use the small-range estimator as long as $\symCardinalityOriginalEstimate \leq \frac{5}{2}\symNumReg$ where the factor $\frac{5}{2}$ was empirically determined.
-
-For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state meaning that they have reached the maximum possible value $\symRegRange + 1$, the \emph{raw} estimate underestimates the cardinalities.
+\right).
+\end{equation}
+The corresponding relative estimation error as depicted in \cref{fig:small_range_estimate} shows that this estimator is convenient for small cardinalities.
+It was proposed to use this small-range estimator as long as $\symCardinalityOriginalEstimate \leq \frac{5}{2}\symNumReg$ where the factor $\frac{5}{2}$ was empirically determined \cite{Flajolet2007}.
+
+For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state meaning that they have reached the maximum possible value $\symRegRange + 1$, the raw estimate underestimates the cardinalities.
 For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$ which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
@@
 
-The original estimation algorithm as presented in \cite{Flajolet2007} including small and large range corrections is summarized by \algoref{alg:estimate_original}. 
+The original estimation algorithm as presented in \cite{Flajolet2007} including small and large range corrections is summarized by \cref{alg:estimate_original}. 
 \begin{algorithm}
@@
 \State $\symCardinalityOriginalEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegVal_\symIndexJ}\right)^{-1}$
-\Comment \emph{raw} estimate \eqref{equ:intermediate_estimate}
+\Comment raw estimate \eqref{equ:intermediate_estimate}
 \If{$\symCardinalityOriginalEstimate\leq \frac{5}{2}\symNumReg$}
@@
 \end{algorithm}
-The relative estimation error for the original method is shown in \autoref{fig:original_estimate}.
+The relative estimation error for the original method is shown in \cref{fig:original_estimate}.
 Unfortunatley, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityOriginalEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap.
@@
 \includesvg[width=1\textwidth]{raw_estimate}
-\caption{The distribution of the relative estimation error over the cardinality as obtained by the raw estimation method \cite{Flajolet2007}.
+\caption{The distribution of the relative estimation error over the cardinality as for the raw estimator \cite{Flajolet2007}.
 10\,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$ have been evaluated.}
@@
 \includesvg[width=1\textwidth]{small_range_estimate}
-\caption{The distribution of the relative estimation error for the small range correction.}
+\caption{The distribution of the relative estimation error for the linear probabilistic counting estimator which can be used for small cardinalities.}
 \label{fig:small_range_estimate}
@@
 
-\subsection{Deriviation of the \emph{raw} estimate}
-Assume following integer distribution
+\subsection{Derivation of the raw estimator}
+In order to better understand why the raw estimator fails for small and large cardinalities we start with a brief derivation that does not require complex analysis.
+
+Consider following integer distribution
 \begin{equation}
@@
 \end{equation}
-Note that, this distribution differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$.
+Note that, this distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$.
 The expectation of $2^{-\symRegVal}$ is given by
-\begin{equation}
+\begin{multline}
+\label{equ:expectation_power_two}
 \symExpectation(2^{-\symRegVal})
 =
+\\
 \sum_{\symRegVal = -\infty}^\infty
@@
 =
-\frac{\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\left( \symPoissonRate/\symNumReg\right)\right)}{2 \log(2)\symPoissonRate}
-\end{equation}
+\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\left( \frac{\symPoissonRate}{\symNumReg}\right)\right)}{\symPoissonRate},
+\end{multline}
 where the function 
@@
 \label{equ:power_series_function}
-\symPowerSeriesFunc(\symY):= \log(2) \sum_{\symRegVal = -\infty}^\infty
-2^{\symRegVal+\symY}
-e^{-2^{\symRegVal+\symY}}
-\end{equation}
- is a smooth periodic function with period 1. Numerical evaluations show that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
-\leq \symPowerSeriesFunc(\symY) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=9.885\cdot 10^{-6}$.
-
-Let $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ be a sample taken from the distribution described by \eqref{equ:assumed_register_val_distribution}.
-For large sample sizes $\symNumReg\rightarrow \infty$ we have asymptotically
-\begin{equation}
-\label{equ:expectation_power_two}
+\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
+2^{\symRegVal+\symX}
+e^{-2^{\symRegVal+\symX}}
+\end{equation}
+ is a smooth periodic function with period 1. Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
+\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=9.885\cdot 10^{-6}$.
+
+Let $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ be a sample taken from \eqref{equ:assumed_register_val_distribution}. For large sample sizes $\symNumReg\rightarrow \infty$ we have asymptotically
+\begin{equation}
 \symExpectation\left(
@@
 \end{equation}
-Together with \eqref{equ:expectation_power_two} we get
+Together with \eqref{equ:expectation_power_two} we obtain
 \begin{equation}
@@
 \symExpectation\left(
-\frac{\symNumReg^2 \symPowerSeriesFunc\!\left(\log_2\left( \symPoissonRate/\symNumReg\right)\right)}{2\log(2)\left(2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}\right)}
+\frac{\symAlpha_\infty\,\symNumReg^2\,\symPowerSeriesFunc\!\left(\log_2\left( \symPoissonRate/\symNumReg\right)\right)}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
 \right)
 \quad
-\text{for $\symNumReg\rightarrow \infty$} 
-\end{equation}
-and therefore, the estimator
+\text{for $\symNumReg\rightarrow \infty$}.
+\end{equation}
+Therefore, 
 \begin{equation}
@@
 = 
-\frac{\symNumReg^2}{2\log(2)\left(2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}\right)}
-\end{equation}
-for the Poisson parameter, is asymptotically almost unbiased with maximum relative offset $\symEpsPowerSeriesFunc$.
-This estimator corresponds to the \emph{raw} estimate, already presented in \eqref{equ:intermediate_estimate} and can also be written as
+\frac{\symAlpha_\infty\,\symNumReg^2}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
+\end{equation}
+is an asymptotically almost unbiased estimator for the Poisson parameter.
+Its asymptotic relative bias is bounded by $\symEpsPowerSeriesFunc$.
+This estimator corresponds to the raw estimator \eqref{equ:intermediate_estimate}, if the Poisson parameter estimate is used as cardinality estimate (see \cref{sec:poisson_approximation}).
+
+The estimator \eqref{equ:rawestimator_2} can also be written as
 \begin{equation}
@@
 = 
-\frac{\symNumReg^2}{2\log(2)\sum_{\symRegVal=-\infty}^\infty \symCount'_\symRegVal 2^{-\symRegVal}}
-\end{equation}
-where $\symCount'_\symRegVal$ is the multiplicity of value $\symRegVal$ in the sample $\left\lbrace \symRegVal_1,\ldots,\symRegVal_\symNumReg\right\rbrace$.
-By defintion, we have $\sum_{\symRegVal=-\infty}^\infty \symCount'_\symRegVal = \symNumReg$.
-
-\subsection{Limitations of the \emph{raw} estimate}
-As we have already seen in \autoref{fig:raw_estimate}, the \emph{raw} estimate does not work very well for small or large cardinalities.
-The reason for this is that the distribution of register values \eqref{equ:register_value_distribution} is different from \eqref{equ:assumed_register_val_distribution} for which the \emph{raw} estimate was derived.
+\frac{\symAlpha_\infty\,\symNumReg^2}{\sum_{\symRegVal=-\infty}^\infty \symCount'_\symRegVal 2^{-\symRegVal}},
+\end{equation}
+if we introduce $\symCount'_\symRegVal$ as the multiplicity of value $\symRegVal$ in the sample $\left\lbrace \symRegVal_1,\ldots,\symRegVal_\symNumReg\right\rbrace$, 
+By definition, we have $\sum_{\symRegVal=-\infty}^\infty \symCount'_\symRegVal = \symNumReg$.
+
+\subsection{Limitations of the raw estimator}
+As we have already seen in \cref{fig:raw_estimate}, the raw estimate does not work very well for small or large cardinalities.
+The reason for this is that the distribution of register values \eqref{equ:register_value_distribution} is different from \eqref{equ:assumed_register_val_distribution} for which the raw estimator was derived.
 Since a random variable $\symRegValVariate'$ that obeys \eqref{equ:assumed_register_val_distribution} can be transformed into random variable $\symRegValVariate$ that follows \eqref{equ:register_value_distribution} using the transformation $\symRegValVariate = \min\left(\max\left(\symRegValVariate', 0\right), \symRegRange+1\right)$, the register values of a HyperLogLog sketch can be seen as the result after applying this transformation to a sample distributed according to \eqref{equ:assumed_register_val_distribution}.
@@
 \symCount_0 &= \textstyle\sum_{\symRegVal = -\infty}^{0} \symCount'_\symRegVal, \\
-\symCount_\symRegVal &=  \symCount'_\symRegVal, \quad \text{for $\symRegVal \in [1, \symRegRange]$},\\
+\symCount_\symRegVal &=  \symCount'_\symRegVal, \quad 1\leq\symRegVal\leq\symRegRange,\\
 \symCount_{\symRegRange+1} &= \textstyle\sum_{\symRegVal = \symRegRange + 1}^{\infty} \symCount'_\symRegVal.
@@
 In case $\symCount_0=0$ and $\symCount_{\symRegRange+1}=0$, we can even be sure that the register values are identical to the originating sample.
-However, for small amd large cardinalities where $\symCount_0$ and $\symCount_{\symRegRange+1}$ approach $\symNumReg$, respectively, the \emph{raw} estimate needs to be corrected.
-
-\subsection{Correcting the \emph{raw} estimate}
-Given a HyperLogLog sketch with its multiplicity vector $\vec{\symCount}$, we try to find estimates $\hat{\symCount}'_\symRegVal$ for $\symCount'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and use them instead in the estimation formula \eqref{equ:raw_estimate_with_multiplicities}.
+However, for small and large cardinalities where $\symCount_0$ and $\symCount_{\symRegRange+1}$ approach $\symNumReg$, respectively, the raw estimator  needs to be corrected.
+
+\subsection{Corrections to the raw estimator}
+Given a HyperLogLog sketch with multiplicity vector $\vec{\symCount}$, we try to find estimates $\hat{\symCount}'_\symRegVal$ for $\symCount'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and use them instead in the estimation formula \eqref{equ:raw_estimate_with_multiplicities}.
 For $\symRegVal\in[1,\symRegRange]$ where $\symCount_\symRegVal = \symCount'_\symRegVal$ we have the trivial estimators
 \begin{equation}
-\hat{\symCount}'_\symRegVal := \symCount_\symRegVal\quad \text{for $\symRegVal \in [1, \symRegRange]$}.
+\hat{\symCount}'_\symRegVal := \symCount_\symRegVal,\quad 1\leq\symRegVal\leq\symRegRange.
 \end{equation}
@@
 
-Inserting all these estimators into \eqref{equ:raw_estimate_with_multiplicities} gives
+Inserting all these estimators into \eqref{equ:raw_estimate_with_multiplicities} finally gives
 \begin{equation}
@@
 = 
-\frac{\symNumReg^2}{2\log(2)\sum_{\symRegVal=-\infty}^\infty \hat{\symCount}'_\symRegVal 2^{-\symRegVal}}
-=
-\frac{\symNumReg^2}
-{2\log(2)
+\frac{\symAlpha_\infty\symNumReg^2}{\sum_{\symRegVal=-\infty}^\infty \hat{\symCount}'_\symRegVal 2^{-\symRegVal}}
+=
+\frac{\symAlpha_\infty\symNumReg^2}
+{
 \left(
@@
 \end{equation}
-where $\symNumReg\,\symSmallCorrectionFunc(\symCount_0/\symNumReg)$ and $\symNumReg\,\symLargeCorrectionFunc(1-\symCount_{\symRegRange+1}/\symNumReg)$ replace $\symCount_0$ and $\symCount_{\symRegRange+1}$ in the \emph{raw} estimate \eqref{equ:intermediate_estimate}, respectively.
+which we call the corrected raw estimator.
+Here $\symNumReg\,\symSmallCorrectionFunc(\symCount_0/\symNumReg)$ and $\symNumReg\,\symLargeCorrectionFunc(1-\symCount_{\symRegRange+1}/\symNumReg)$ are replacements for  $\symCount_0$ and $\symCount_{\symRegRange+1}$ in the raw estimator \eqref{equ:intermediate_estimate}, respectively.
 The functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are defined as
@@
 \end{equation}
-almost the same estimator as given in \eqref{equ:small_range_estimate}. Here we used that $\symPowerSeriesFunc$ is known to be approximately equal to 1.
+almost the same estimator as given in \eqref{equ:small_range_estimate}. Here we used the fact that $\symPowerSeriesFunc$ can be well approximated by 1.
 \subsection{Estimation algorithm}
-see \algoref{alg:corrected_raw_estimation}
+The corrected raw estimator \eqref{equ:correctedestimator} leads to a new cardinality estimation algorithm for HyperLogLog sketches.
+\cref{alg:corrected_raw_estimation} demonstrates the numerical evaluation of the estimator.
+The values of functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ can be either calculated on-demand or pre-calculated.
+The possible value range for $\symCount_0$ and $\symCount_{\symRegRange+1}$ is $[0, \symNumReg]$.
+Therefore, if performance matters, it is possible to precalculate the function values for all possible arguments and keep them in lookup tables of size $\symNumReg+1$.
 
 \begin{algorithm}
-\caption{Cardinality estimation algorithm based on the corrected \emph{raw} estimate.}
+\caption{Cardinality estimation algorithm based on the corrected raw estimate.}
 \label{alg:corrected_raw_estimation}
@@
 \State $\symNumReg \gets \left\Vert\vec{\symCount}\right\Vert_1$
-\State $\symZ\gets 0.5\cdot \symLargeCorrectionFunc(\symCount_{\symRegRange+1},\symNumReg)$
+\State $\symZ\gets 0.5\cdot \symNumReg\cdot\symLargeCorrectionFunc(1 - \symCount_{\symRegRange+1}/\symNumReg)$
+\Comment{\parbox[t]{.5\linewidth}{alternatively, take $0.5\cdot \symNumReg\cdot\symLargeCorrectionFunc(1 - \symCount_{\symRegRange+1}/\symNumReg)$ from precalculated lookup table}}
 \For{$\symRegVal\gets \symRegRange, 1$}
@@
 \EndFor
-\State $\symZ\gets\symZ+\symSmallCorrectionFunc(\symCount_0,\symNumReg)$
+\State $\symZ\gets\symZ+\symNumReg\cdot\symSmallCorrectionFunc(\symCount_0/\symNumReg)$
+\Comment{\parbox[t]{.5\linewidth}{alternatively, take $\symNumReg\cdot\symSmallCorrectionFunc(\symCount_0/\symNumReg)$ from precalculated lookup table}}
 \State \Return$\symNumReg^2 / \left(2\log(2)\symZ\right)$
@@
 \\
-\Function {$\symSmallCorrectionFunc$}{$\symCount,\symNumReg$}
-\If{$\symCount=\symNumReg$}
+\Function {$\symSmallCorrectionFunc$}{$\symX$}
+\Comment $\symX\in[0,1]$
+\If{$\symX=1$}
 \State\Return$\infty$
 \EndIf
-\State $\symX\gets\symCount/\symNumReg$
 \State $\symY\gets1$
@@
 \Until{$\symZ=\symZ'$}
-\State\Return$\symNumReg\cdot\symZ$
+\State\Return$\symZ$
 \EndFunction
 \\
-\Function {$\symLargeCorrectionFunc$}{$\symCount,\symNumReg$}
-\State $\symX\gets(\symNumReg-\symCount)/\symNumReg$
+\Function {$\symLargeCorrectionFunc$}{$\symX$}
+\Comment $\symX\in[0,1]$
 \State $\symY\gets1$
@@
 \Until{$\symZ=\symZ'$}
-\State\Return$\symNumReg\cdot\symZ$
+\State\Return$\symZ$
 \EndFunction
@@
 
-see \autoref{fig:raw_corrected_estimation_error_8_24}, \autoref{fig:raw_corrected_estimation_error_12_20},
-\autoref{fig:raw_corrected_estimation_error_16_16},
-\autoref{fig:raw_corrected_estimation_error_20_12},
-\autoref{fig:raw_corrected_estimation_error_22_42},
+see \cref{fig:raw_corrected_estimation_error_8_24}, \cref{fig:raw_corrected_estimation_error_12_20},
+\cref{fig:raw_corrected_estimation_error_16_16},
+\cref{fig:raw_corrected_estimation_error_20_12},
+\cref{fig:raw_corrected_estimation_error_22_42},
 
@@
 \subsection{Performance}
-see \autoref{fig:corrected_raw_avg_exec_time}
+see \cref{fig:corrected_raw_avg_exec_time}
 
@@
 However, there, a different element insertion algorithm was assumed that works with time complexity $\symBigO(\symNumReg)$, because each register calculates its own hash value from the inserted element and updates its current value accordingly.
-In comparison, \algoref{alg:insert} uses only a single hash value that defines which register needs to be updated.
+In comparison, \cref{alg:insert} uses only a single hash value that defines which register needs to be updated.
 As a consequence, the register values are statistically dependent, and the Poisson approximation is needed to make the register values statistically independent, which allows the factorization of the joint probability distribution of all register values.
@@
 Furthermore, we will demonstrate that consequent application of the maximum likelihood method reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values.
-The history of the HyperLogLog algorithm shows that the \emph{raw} estimate \eqref{alg:estimate_original} was first found after several attempts using the geometric mean \cite{Flajolet2007, Durand2003}.
+The history of the HyperLogLog algorithm shows that the raw estimate \eqref{alg:estimate_original} was first found after several attempts using the geometric mean \cite{Flajolet2007, Durand2003}.
 
@@
 \end{equation}
-$\symHelper(\symX)$ is strictly increasing and concave as can be seen in \autoref{fig:helper_function}.
+$\symHelper(\symX)$ is strictly increasing and concave as can be seen in \cref{fig:helper_function}.
 For non-negative values $\symX$ the function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
@@
 
-It is easy to see that the estimate $\symPoissonRateEstimate$ remains equal or becomes larger, when inserting an element into the HyperLogLog sketch following \algoref{alg:insert}.
+It is easy to see that the estimate $\symPoissonRateEstimate$ remains equal or becomes larger, when inserting an element into the HyperLogLog sketch following \cref{alg:insert}.
 An update potentially changes the multiplicity vector $(\symCount_0,\ldots,\symCount_{\symRegRange+1})$ to $(\symCount_0,\ldots,\symCount_\symIndexI-1,\ldots,\symCount_\symIndexJ+1,\ldots,\symCount_{\symRegRange+1})$ where $\symIndexI < \symIndexJ$.
@@
 \end{equation}
-Note, if the HyperLogLog sketch is in the intermediate range, where $\symCount_0=\symCount_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the \emph{raw} estimate \eqref{equ:intermediate_estimate}.
-Hence, the maximum likelihood method leads directly to the harmonic mean that is used for the \emph{raw} estimate.
-If $\symCount_0>0$ or $\symCount_{\symRegRange+1}>0$ the bounds do no longer follow the harmonic mean and this is also the reason why the \emph{raw} estimate fails for small and large cardinalities.
+Note, if the HyperLogLog sketch is in the intermediate range, where $\symCount_0=\symCount_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimate \eqref{equ:intermediate_estimate}.
+Hence, the maximum likelihood method leads directly to the harmonic mean that is used for the raw estimate.
+If $\symCount_0>0$ or $\symCount_{\symRegRange+1}>0$ the bounds do no longer follow the harmonic mean and this is also the reason why the raw estimate fails for small and large cardinalities.
 
@@
 \end{equation}
-Note, this recursive formula is stable in a sense that the relative error of $\symHelper\left(2\symX\right)$ is smaller than that of $\symHelper\left(\symX\right)$ as shown in \autoref{app:helper_stable}.
+Note, this recursive formula is stable in a sense that the relative error of $\symHelper\left(2\symX\right)$ is smaller than that of $\symHelper\left(\symX\right)$ as shown in \cref{app:helper_stable}.
 
@@
 
-All these optimizations together finally give the new cardinality estimation algorithm presented in \algoref{alg:estimate_ml}.
+All these optimizations together finally give the new cardinality estimation algorithm presented in \cref{alg:estimate_ml}.
 
@@
 
-\autoref{fig:estimation_error_12_20} shows the distribution of the relative error of the estimated cardinality using \algoref{alg:estimate_ml} compared to the true cardinality for $\symPrecision=12$ and $\symRegRange=20$.
+\cref{fig:estimation_error_12_20} shows the distribution of the relative error of the estimated cardinality using \cref{alg:estimate_ml} compared to the true cardinality for $\symPrecision=12$ and $\symRegRange=20$.
 As the median shows, the error is essentially unbiased over the entire cardinality range except for very small values.
-The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is about an order of magnitude larger than the operating range upper bound of the \emph{raw} estimate (\autoref{fig:raw_estimate}) or the original method (\autoref{fig:original_estimate}).
+The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is about an order of magnitude larger than the operating range upper bound of the raw estimate (\cref{fig:raw_estimate}) or the original method (\cref{fig:original_estimate}).
 
@@
 First we considered the original HyperLogLog algorithm that used a 32-bit hash function ($\symPrecision + \symRegRange = 32$).
-The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, $\symPrecision=20$ are shown in \autoref{fig:estimation_error_8_24}, \autoref{fig:estimation_error_16_16}, and \autoref{fig:estimation_error_20_12}, respectively.
+The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, $\symPrecision=20$ are shown in \cref{fig:estimation_error_8_24}, \cref{fig:estimation_error_16_16}, and \cref{fig:estimation_error_20_12}, respectively.
 As expected, since  $\symPrecision + \symRegRange = 32$ is kept constant, the operating range remains more or less the same, while the relative error decreases with increasing precision.
@@
 replacing the 32-bit hash function by a 64-bit hash function.
-\autoref{fig:estimation_error_12_52} shows the relative error for such a HyperLogLog configuration with precision  $\symPrecision=12$.
+\cref{fig:estimation_error_12_52} shows the relative error for such a HyperLogLog configuration with precision  $\symPrecision=12$.
 In this case, in order to use all the 64 bits of the hash value, $\symRegRange$ must be chosen to be equal to $64 - \symPrecision = 52$.
@@
 The doubled hash value size shifts the maximum trackable cardinality value towards $2^{64}$.
-As \autoref{fig:estimation_error_12_52} shows, when compared to the 32-bit hash value case given in \autoref{fig:estimation_error_12_20}, the estimation error remains constant over the entire simulated cardinality range up to 50 billions.
+As \cref{fig:estimation_error_12_52} shows, when compared to the 32-bit hash value case given in \cref{fig:estimation_error_12_20}, the estimation error remains constant over the entire simulated cardinality range up to 50 billions.
 
@@
 This allows two registers to share a single byte, which is beneficial from a performance perspective.
-Nevertheless, this configuration still allows the estimation of cardinalities up to 100 millions as shown in \autoref{fig:estimation_error_12_14}, which could be enough for many applications.
+Nevertheless, this configuration still allows the estimation of cardinalities up to 100 millions as shown in \cref{fig:estimation_error_12_14}, which could be enough for many applications.
 
@@
 
-TODO \autoref{fig:estimation_error_22_42}
+TODO \cref{fig:estimation_error_22_42}
 
@@
 For each evaluated cardinality value the average execution time was calculated after 100 cycles which corresponds to 100\,000 algorithm executions.
-The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \autoref{fig:avg_exec_time}.
+The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:avg_exec_time}.
 All these benchmarks where carried out on an Intel Core i5-2500K clocking at 3.3GHz.
@@
 \end{equation}
-As in \autoref{sec:cardinality_estimation} where we found that the multiplicity vector $\vec{\symCount}$ of a sketch is a sufficient statistic for the cardinality, the multiplicity matrix of two sketches is a sufficient statistic for the cardinalities of $\symSetX$, $\symSetA$, and $\symSetB$.
+As in \cref{sec:cardinality_estimation} where we found that the multiplicity vector $\vec{\symCount}$ of a sketch is a sufficient statistic for the cardinality, the multiplicity matrix of two sketches is a sufficient statistic for the cardinalities of $\symSetX$, $\symSetA$, and $\symSetB$.
 

### 2016-05-31 python/power-series-function.py
new chart showing deviation of xi function from 1

@@ -8,14 +8,14 @@ import matplotlib.pyplot as plt
 from scipy.optimize import minimize_scalar
 
 numPoints = 100000
-xMin = 0.000
-xMax = 1
+xMin = -2
+xMax = 2
 
 def h2(x):
     return exp(-pow(2,x))*pow(2,x)
-    
+
 def h(x):
-    
+
     a = h2(x)
     amin = 0.
     k = 0
@@ -25,7 +25,7 @@ def h(x):
         amin += h2(k+x)
         if (amin == aminold):
             break
-    
+
     amax = 0.;
     k = 0
     while True:
@@ -34,14 +34,14 @@ def h(x):
         amax += h2(k+x)
         if (amax == amaxold):
             break
-    
+
     return log(2.)*(a + amin + amax)
 
 
 
 def f1(x):
     return h(x) - 1.
-    
+
 def f2(x):
     return 1. - h(x)
 
@@ -56,5 +56,13 @@ res2 = minimize_scalar(f2, bounds = (0,1), method='bounded')
 
 print(res1)
 print(res2)
-plt.show()
+
+fig1, ax1 = plt.subplots(subplot_kw=dict(xlim=[xMin,xMax], ylim=[-1.1e-5,1.1e-5]))
+fig1.set_size_inches(6, 4)
+ax1.plot(xValues, yValues, color='black')
+ax1.grid(b=True, which='major', color='black', linestyle='--')
+ax1.ticklabel_format(axis='y', style = 'sci', useOffset=False,  scilimits=(-1e-6,1e-6))
+ax1.set_xlabel(r"$x$")
+ax1.set_ylabel(r"$\xi(x)-1$")
+fig1.savefig('../paper/power-series-function-minus-1.svg', format='svg', dpi=600)
 

### 2016-06-02 paper/paper.tex
inserted new chart into paper
replaced some figures, eliminated results for p=20,q=10 and p=22,q=42, added results for p=22,q=10

@@
 $\symPrecision$ is the precision that directly influences the relative error which scales like $1/\sqrt{\symNumReg}$.
-All registers start with the initial value 0.
+All registers start with initial value 0.
 The HyperLogLog algorithm successively increases the register values up to a maximum value limited by the space per register which is usually between 5 and 8 bits.
@@
  is a smooth periodic function with period 1. Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
-\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=9.885\cdot 10^{-6}$.
+\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=9.885\cdot 10^{-6}$ (see \cref{fig:power_series_func}).
+
+\begin{figure}
+\centering
+\includesvg[width=0.6\textwidth]{power-series-function-minus-1}
+\caption{The deviation of $\symPowerSeriesFunc(\symX)$ from 1.}
+\label{fig:power_series_func}
+\end{figure}
 
@@
 \cref{fig:raw_corrected_estimation_error_16_16},
-\cref{fig:raw_corrected_estimation_error_20_12},
-\cref{fig:raw_corrected_estimation_error_22_42},
+\cref{fig:raw_corrected_estimation_error_22_10},
 
@@
 \centering
-\includesvg[width=1\textwidth]{raw_corrected_estimate_20_12}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 20$ and $\symRegRange=12$.}
-\label{fig:raw_corrected_estimation_error_20_12}
-\end{figure}
-
-\begin{figure}
-\centering
-\includesvg[width=1\textwidth]{raw_corrected_estimate_22_42}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=42$.}
-\label{fig:raw_corrected_estimation_error_22_42}
+\includesvg[width=1\textwidth]{raw_corrected_estimate_22_10}
+\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
+\label{fig:raw_corrected_estimation_error_22_10}
 \end{figure}
@@
 First we considered the original HyperLogLog algorithm that used a 32-bit hash function ($\symPrecision + \symRegRange = 32$).
-The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, $\symPrecision=20$ are shown in \cref{fig:estimation_error_8_24}, \cref{fig:estimation_error_16_16}, and \cref{fig:estimation_error_20_12}, respectively.
+The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, $\symPrecision=22$ are shown in \cref{fig:estimation_error_8_24}, \cref{fig:estimation_error_16_16}, and \cref{fig:estimation_error_22_10}, respectively.
 As expected, since  $\symPrecision + \symRegRange = 32$ is kept constant, the operating range remains more or less the same, while the relative error decreases with increasing precision.
@@
 \centering
-\includesvg[width=1\textwidth]{max_likelihood_estimate_20_12}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 20$ and $\symRegRange=12$.}
-\label{fig:estimation_error_20_12}
+\includesvg[width=1\textwidth]{max_likelihood_estimate_22_10}
+\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
+\label{fig:estimation_error_22_10}
 \end{figure}
@@
 \end{figure}
-
-\begin{figure}
-\centering
-\includesvg[width=1\textwidth]{max_likelihood_estimate_22_42}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=42$.}
-\label{fig:estimation_error_22_42}
-\end{figure}
-
-TODO \cref{fig:estimation_error_22_42}
 

### 2016-06-02 python/distribution_charts.py

@@ -76,37 +76,18 @@ createFigure(12, 20, 'flajolet_estimates', 'original_estimate')
 createFigure(12, 20, 'flajolet_mid_range_estimates', 'raw_estimate')
 createFigure(12, 20, 'flajolet_small_range_estimates', 'small_range_estimate')
 
-#createFigure(4, 28,  'max_likelihood_estimates', 'max_likelihood_estimate_4_28')
-#createFigure(6, 26,  'max_likelihood_estimates', 'max_likelihood_estimate_6_26')
 createFigure(8, 24,  'max_likelihood_estimates', 'max_likelihood_estimate_8_24')
 createFigure(12, 20, 'max_likelihood_estimates', 'max_likelihood_estimate_12_20')
 createFigure(16, 16, 'max_likelihood_estimates', 'max_likelihood_estimate_16_16')
-createFigure(20, 12, 'max_likelihood_estimates', 'max_likelihood_estimate_20_12')
-createFigure(22, 42, 'max_likelihood_estimates', 'max_likelihood_estimate_22_42')
+createFigure(22, 10, 'max_likelihood_estimates', 'max_likelihood_estimate_22_10')
 createFigure(12, 52, 'max_likelihood_estimates', 'max_likelihood_estimate_12_52')
 createFigure(12, 14, 'max_likelihood_estimates', 'max_likelihood_estimate_12_14')
 createFigure(12,  0, 'max_likelihood_estimates', 'max_likelihood_estimate_12_0')
 
-
-#createFigure(4, 28,  'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_4_28')
-#createFigure(6, 26,  'flajolet_mid_range_estimates_corrected', 'raw_corrected_estimate_6_26')
 createFigure(8, 24,  'corrected_raw_estimates', 'raw_corrected_estimate_8_24')
 createFigure(12, 20, 'corrected_raw_estimates', 'raw_corrected_estimate_12_20')
 createFigure(16, 16, 'corrected_raw_estimates', 'raw_corrected_estimate_16_16')
-createFigure(20, 12, 'corrected_raw_estimates', 'raw_corrected_estimate_20_12')
-createFigure(22, 42, 'corrected_raw_estimates', 'raw_corrected_estimate_22_42')
+createFigure(22, 10, 'corrected_raw_estimates', 'raw_corrected_estimate_22_10')
[…]

### 2016-06-07 paper/paper.tex
some changes in paper, added function to generate mean/variance charts

@@
 Based on the maximum likelihood principle another unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate.
-The maximum-likelihood method is also appropriate to improve the cardinality estimates of set intersections when compared to the inclusion-exclusion principle.
+The maximum-likelihood method is also appropriate to improve cardinality estimates of set intersections compared to the inclusion-exclusion principle.
 The new methods are demonstrated and verified by extensive simulations.
@@
 In order to insert a data element into a HyperLogLog data structure a hash value is calculated.
-The first $\symPrecision$ bits of the hash value are used to select one of the $2^\symPrecision$ registers.
-For another $\symRegRange$ bits, the position of the first 1-bit is determined.
-If this position exceeds the value of the selected register, the register value is replaced.
+The leading $\symPrecision$ bits of the hash value are used to select one of the $2^\symPrecision$ registers.
+Among the next $\symRegRange$ bits, the position of the first 1-bit is determined which is a value in the range $[1,\symRegRange+1]$.
+The value $\symRegRange+1$ is used, if all $\symRegRange$ bits are equal to 0.
+If the position of the first 1-bit exceeds the value of the selected register, the register value is replaced.
 \cref{alg:insert} shows the update procedure for inserting a data element into the HyperLogLog sketch.
@@
 \end{equation}
-Note that the order of register values $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ is not important for the estimation of the cardinality.
+
+The order of register values $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ is not important for the estimation of the cardinality.
 More formally, the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$ is a sufficient statistic for $\symCardinality$.
@@
 As a consequence, the multiplicity vector $\vec{\symCount} := (\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is also a sufficient statistic for the cardinality.
-In addition, this vector also contains all the information about the HyperLogLog sketch that is required for  cardinality estimation: $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$ and $\symRegRange=\dim(\vec{\symCount})-2$.
+In addition, this vector contains all the information about the HyperLogLog sketch that is required for cardinality estimation.
+The two HyperLogLog parameters can be obtained by $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$ and $\symRegRange=\dim(\vec{\symCount})-2$, respectively.
 
@@
 \sum_{\symCardinality=0}^\infty \symProbabilityMass(\vec{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
-\nonumber\\
+\label{equ:poissonization}
+\\
 &= 
@@
 \end{equation}
-Due to the simpler probability mass function, it is easier to find an estimate  $\symPoissonRateEstimate$ for the Poisson rate $\symPoissonRate$ than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}.
-Since, the expectation of the cardinality under the Poisson model is equal to $\symPoissonRate$, it makes sense to use $\symPoissonRateEstimate$ directly as estimate for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$.
-The Poisson approximation is known to work well for large rates.
-However, as simulation results show later, the Poisson approximation gives high quality estimates for small cardinalities as well.
+
+\subsection{Depoissonization}
+Due to the simpler probability mass function, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\vec{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ rather than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}.
+Here $\vec{\symRegValVariate}$ denotes the vector of all observed register values.
+$\vec{\symRegValVariate}$ is distributed according to \eqref{equ:multinomialProbabilityMass}.
+Depoissonization \cite{Jacquet1998} finally allows to translate the estimates back to the fixed-size model.
+Assume we have found an unbiased estimator for the Poisson rate
+\begin{equation}
+\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = \symPoissonRate
+\quad
+\text{for all $\symPoissonRate\geq 0$}.
+\end{equation}
+We know from \eqref{equ:poissonization} 
+\begin{equation}
+\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = 
+\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
+\end{equation}
+and therefore
+\begin{equation}
+\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
+=
+\symPoissonRate
+\end{equation}
+holds for all $\symPoissonRate\geq0$. The inverse Poisson transform gives
+\begin{equation}
+\symExpectation(\symPoissonRateEstimate\vert\symCardinality) = \symCardinality.
+\end{equation}
+Therefore, the unbiased estimator $\symPoissonRateEstimate$ conditioned on $\symCardinality$ is also an unbiased estimator for $\symCardinality$.
+Hence, it makes sense to use $\symPoissonRateEstimate$ directly as estimate for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$.
+As simulation results will show later, the Poisson approximation works well over the entire cardinality range, even for estimators that are not exactly unbiased.
 
@@
 \end{equation}
-as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible when compared to the estimation error.
+as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible if compared to the estimation error.
 
@@
 The chart is based on 10\,000 randomly generated HyperLogLog sketches.
-Obviously, the raw estimate is biased for small and large cardinalities and fails to return accurate estimates.
+Obviously, the raw estimator is biased for small and large cardinalities and fails to return accurate estimates.
 Therefore, to cover the entire range of cardinalities, corrections for small and large cardinalities have been proposed, respectively.
@@
 
-For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state meaning that they have reached the maximum possible value $\symRegRange + 1$, the raw estimate underestimates the cardinalities.
+For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state meaning that they have reached the maximum possible value $\symRegRange + 1$, the raw estimator underestimates the cardinalities.
 For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$ which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
@@
 If 64-bit hash values are used as proposed in \cite{Heule2013} $(\symPrecision+\symRegRange=64)$, a correction for large cardinalities can be circumvented.
-Cardinalities in the order of $2^{64}$ for which saturated register become a problem never occur in practice.
+Cardinalities in the order of $2^{64}$ for which saturated registers become a problem never occur in practice.
 
@@
 \subsection{Derivation of the raw estimator}
-In order to better understand why the raw estimator fails for small and large cardinalities we start with a brief derivation that does not require complex analysis.
+In order to better understand why the raw estimator fails for small and large cardinalities, we start with a brief derivation that does not require complex analysis.
 
@@
 \end{equation}
-Note that, this distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$.
+This distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$.
 The expectation of $2^{-\symRegVal}$ is given by
@@
 \end{equation}
-if we introduce $\symCount'_\symRegVal$ as the multiplicity of value $\symRegVal$ in the sample $\left\lbrace \symRegVal_1,\ldots,\symRegVal_\symNumReg\right\rbrace$, 
+if $\symCount'_\symRegVal$ is the multiplicity of value $\symRegVal$ in the sample $\left\lbrace \symRegVal_1,\ldots,\symRegVal_\symNumReg\right\rbrace$. 
 By definition, we have $\sum_{\symRegVal=-\infty}^\infty \symCount'_\symRegVal = \symNumReg$.
@@
 \subsection{Limitations of the raw estimator}
-As we have already seen in \cref{fig:raw_estimate}, the raw estimate does not work very well for small or large cardinalities.
+As we have already seen in \cref{fig:raw_estimate}, the raw estimator does not work very well for small or large cardinalities.
 The reason for this is that the distribution of register values \eqref{equ:register_value_distribution} is different from \eqref{equ:assumed_register_val_distribution} for which the raw estimator was derived.
-Since a random variable $\symRegValVariate'$ that obeys \eqref{equ:assumed_register_val_distribution} can be transformed into random variable $\symRegValVariate$ that follows \eqref{equ:register_value_distribution} using the transformation $\symRegValVariate = \min\left(\max\left(\symRegValVariate', 0\right), \symRegRange+1\right)$, the register values of a HyperLogLog sketch can be seen as the result after applying this transformation to a sample distributed according to \eqref{equ:assumed_register_val_distribution}.
-The corresponding multiplicities are related by
+Since a random variable $\symRegValVariate'$ that obeys \eqref{equ:assumed_register_val_distribution} can be transformed into random variable $\symRegValVariate$ that follows \eqref{equ:register_value_distribution} using the transformation $\symRegValVariate = \min\left(\max\left(\symRegValVariate', 0\right), \symRegRange+1\right)$, the register values of a HyperLogLog sketch can be seen as the result after applying this transformation to a sample that is distributed according to \eqref{equ:assumed_register_val_distribution}.
+The corresponding multiplicities have  following relationships
 \begin{equation}
@@
 Obviously, as long as $\symCount_0\approx\symCount'_0$ and  $\symCount'_{\symRegRange+1}\approx\symCount_{\symRegRange+1}$ the set of register values is similar to the originating sample.
-In case $\symCount_0=0$ and $\symCount_{\symRegRange+1}=0$, we can even be sure that the register values are identical to the originating sample.
-However, for small and large cardinalities where $\symCount_0$ and $\symCount_{\symRegRange+1}$ approach $\symNumReg$, respectively, the raw estimator  needs to be corrected.
+In case $\symCount_0=0$ and $\symCount_{\symRegRange+1}=0$, which implies $\symCount_\symRegVal=0$ for $\symRegVal\leq0$ and $\symRegVal\geq\symRegRange+1$, we can even be sure that the register values are identical to the originating sample.
+However, for small and large cardinalities $\symCount_0$ and $\symCount_{\symRegRange+1}$ approach $\symNumReg$, respectively, and the raw estimator needs to be corrected.
 
@@
 almost the same estimator as given in \eqref{equ:small_range_estimate}. Here we used the fact that $\symPowerSeriesFunc$ can be well approximated by 1.
-\subsection{Estimation algorithm}
+\subsection{Corrected raw estimation algorithm}
 The corrected raw estimator \eqref{equ:correctedestimator} leads to a new cardinality estimation algorithm for HyperLogLog sketches.
@@
 \caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.
-Both cases, $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ precalculated or calculated on-demand have been considered.}
+Both cases, $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ precalculated and calculated on-demand have been considered.}
 \label{fig:corrected_raw_avg_exec_time}
@@
 The function is non-positive at 0 since $\symFunc(0)=\symCount_0-\symNumReg\leq 0$ and, in case $\symCount_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing.
-$\symCount_{\symRegRange}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $(\symRegRange+1)$, for which the maximum likelihood estimate would be infinity.
+$\symCount_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $(\symRegRange+1)$, for which the maximum likelihood estimate would be infinity.
 
@@
 \symCount_1 \left(
-\symHelper\left(\frac{x}{2^1}\right)
+\symHelper\!\left(\frac{x}{2^1}\right)
 +\frac{\symX}{2^1}-1
@@
 \symCount_2 \left(
-\symHelper\left(\frac{x}{2^2}\right)
+\symHelper\!\left(\frac{x}{2^2}\right)
 +\frac{\symX}{2^2}-1
@@
 \left(
-\symHelper\left(\frac{\symX}{2^\symRegRange}\right)
+\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
 +\frac{\symX}{2^\symRegRange}-1
@@
 \left(
-\symHelper\left(\frac{\symX}{2^{\symRegRange}}\right)
+\symHelper\!\left(\frac{\symX}{2^{\symRegRange}}\right)
 -1
@@
 Note, if the HyperLogLog sketch is in the intermediate range, where $\symCount_0=\symCount_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimate \eqref{equ:intermediate_estimate}.
-Hence, the maximum likelihood method leads directly to the harmonic mean that is used for the raw estimate.
-If $\symCount_0>0$ or $\symCount_{\symRegRange+1}>0$ the bounds do no longer follow the harmonic mean and this is also the reason why the raw estimate fails for small and large cardinalities.
+Hence, the maximum likelihood method leads directly to the harmonic mean that is used by the raw estimator.
+If $\symCount_0>0$ or $\symCount_{\symRegRange+1}>0$ the bounds do no longer follow the harmonic mean and this is also the reason why the raw estimator fails for small and large cardinalities.
 
@@
 =\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
-\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\left(\frac{\symX}{2^\symRegVal}\right)
-+
-\symCount_{\symRegRange+1}\symHelper\left(\frac{\symX}{2^\symRegRange}\right)
+\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\!\left(\frac{\symX}{2^\symRegVal}\right)
++
+\symCount_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
 \end{equation}
@@
 
-\subsection{The new cardinality estimation algorithm}
+\subsection{Maximum-likelihood estimation algorithm}
 In order to get a fast cardinality estimation algorithm, it is crucial to minimize the evaluation costs for \eqref{equ:funcprime}.

### 2016-06-07 python/distribution_charts.py

@@ -18,9 +18,9 @@ def readData(filename):
         return numpy.array([[float(e) for e in r] for r in reader][1:])
 
 def createFigure(p, q, dataSource, figureName):
-    
+
     filename = str(p) + '_' + str(q)
-        
+
     m = pow(2, p)
 
     cardinalities = readData('cardinalities.dat').flatten()
@@ -68,7 +68,68 @@ def createFigure(p, q, dataSource, figureName):
     fig1.savefig('../paper/' + figureName + '.svg', format='svg', dpi=600)
     fig1.savefig('../paper/' + figureName + '.png', format='png', dpi=100)
     plt.close(fig1)
-    
+
+    numpy.savetxt(dataDir + figureName + ".min_error", datamin, delimiter=',')
+    numpy.savetxt(dataDir + figureName + ".max_error", datamax, delimiter=',')
+
+def createFigure2(p, q, dataSource, figureName):
+
+    filename = str(p) + '_' + str(q)
+
+    m = pow(2, p)
+
+    cardinalities = readData('cardinalities.dat').flatten()
+    data = readData(filename + '_' + dataSource + '.dat')
+
+    plt.rc('text', usetex=True)
+    plt.rc('font', family='serif')
+
+    yMax = 6./pow(2., p*0.5)
+    fig1, ax1 = plt.subplots(subplot_kw=dict(xlim=[cardinalities[0],cardinalities[-1]], ylim=[-yMax,yMax], xscale='log', yscale='linear'))
+
+    fig1.set_size_inches(10, 5)
+    ax1.yaxis.grid(b=True, which='major', color='gray', linestyle='--')
+    ax1.xaxis.grid(b=True, which='major', color='gray', linestyle='--')
+    ax1.set_xlabel("cardinality")
+    ax1.set_ylabel("relative error")
+
+    pm3s = (1.-erf(3./sqrt(2.)))/2.
+    pp3s = (1.+erf(3./sqrt(2.)))/2.
+    pm2s = (1.-erf(2./sqrt(2.)))/2.
+    pp2s = (1.+erf(2./sqrt(2.)))/2.
+    pm1s = (1.-erf(1./sqrt(2.)))/2.
+    pp1s = (1.+erf(1./sqrt(2.)))/2.
+
+    datamin = numpy.divide(data.min(axis=1), cardinalities)-1.
+    datam3s = numpy.divide(numpy.percentile(data, pm3s*100., axis=1), cardinalities)-1.
[…]

### 2016-06-11 paper/paper.tex
paper
added GitHub link to paper, estimation error section for corrected raw
estimator, replaced "linear probabilistic counting" by "linear
counting", some other minor changes

@@
 \title{New cardinality estimation algorithms for HyperLogLog sketches}
-\subtitle{Draft version}
+\subtitle{Draft version, \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}}
 \author{Otmar Ertl}
@@
 
-Note that a $(\symPrecision, 0)$-HyperLogLog sketch corresponds to a bit array as used by linear probabilistic counting \cite{Whang1990}.
+Note that a $(\symPrecision, 0)$-HyperLogLog sketch corresponds to a bit array as used by linear counting \cite{Whang1990}.
 In this case each register value is represented by a single bit.
-Hence, linear probabilistic counting can be regarded as a special case of the HyperLogLog algorithm ($\symRegRange = 0$).
+Hence, linear counting can be regarded as a special case of the HyperLogLog algorithm ($\symRegRange = 0$).
 
@@
 These approximations are used in many HyperLogLog implementations.
-However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision\geq 20$.
+However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions.
 For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use
@@
 As mentioned in \cref{sec:data_element_insertion}, a HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be turned into a $(\symPrecision, 0)$-HyperLogLog sketch.
-Since $\symRegRange=0$ corresponds to linear probabilistic counting and the transformed HyperLogLog sketch corresponds to a bitset with $\symCount_0$ zeros, we can apply the corresponding cardinality estimator \cite{Whang1990}
+Since $\symRegRange=0$ corresponds to linear counting and the transformed HyperLogLog sketch corresponds to a bitset with $\symCount_0$ zeros, the linear counting cardinality estimator \cite{Whang1990} can be applied here
 \begin{equation}
@@
 \includesvg[width=1\textwidth]{small_range_estimate}
-\caption{The distribution of the relative estimation error for the linear probabilistic counting estimator which can be used for small cardinalities.}
+\caption{The distribution of the relative estimation error for the linear counting estimator which can be used for small cardinalities.}
 \label{fig:small_range_estimate}
@@
 \subsection{Derivation of the raw estimator}
+\label{sec:derivation_raw_estimator}
 In order to better understand why the raw estimator fails for small and large cardinalities, we start with a brief derivation that does not require complex analysis.
@@
 
-Note that, $2\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = -\symPowerSeriesFunc(\log_2(-\log(\symX)))/\left(\log(2)\log(\symX)\right)$.
-Therefore, we get for the special case of linear probabilistic counting with $\symRegRange=0$
+Using the identity $2\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = -\symPowerSeriesFunc(\log_2(-\log(\symX)))/\left(\log(2)\log(\symX)\right)$, we get for the linear counting case with $\symRegRange=0$
 \begin{equation}
@@
 \approx
-\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right).
-\end{equation}
-almost the same estimator as given in \eqref{equ:small_range_estimate}. Here we used the fact that $\symPowerSeriesFunc$ can be well approximated by 1.
+\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right),
+\end{equation}
+which is almost identical to the linear counting estimator used for the small range correction \eqref{equ:small_range_estimate}.
+Here we used the fact that $\symPowerSeriesFunc$ can be well approximated by 1.
 \subsection{Corrected raw estimation algorithm}
@@
 \subsection{Estimation error}
-
-see \cref{fig:raw_corrected_estimation_error_8_24}, \cref{fig:raw_corrected_estimation_error_12_20},
-\cref{fig:raw_corrected_estimation_error_16_16},
-\cref{fig:raw_corrected_estimation_error_22_10},
+In order to verify the new estimation algorithm, we generated 10\,000 HyperLogLog sketches and inserted up to 50 billion unique elements.
+Element hash values have been mocked by random numbers.
+For the following results we used the Mersenne Twister random number generator with 19937 bit state size from the C++ standard library.
+
+\cref{fig:raw_corrected_estimation_error_12_20} shows the distribution of the relative error of the estimated cardinality using \cref{alg:corrected_raw_estimation} compared to the true cardinality for $\symPrecision=12$ and $\symRegRange=20$.
+As the mean shows, the error is unbiased over the entire cardinality range.
+The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is about an order of magnitude larger than the operating range upper bound of the raw estimator (\cref{fig:raw_estimate}) or the original method (\cref{fig:original_estimate}).
+
+\begin{figure}
+\centering
+\includesvg[width=1\textwidth]{raw_corrected_estimate_12_20}
+\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
+\label{fig:raw_corrected_estimation_error_12_20}
+\end{figure}
+
+The new estimation algorithm also works well for other HyperLogLog configurations.
+First we considered the original HyperLogLog algorithm that used a 32-bit hash function ($\symPrecision + \symRegRange = 32$).
+The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, $\symPrecision=22$ are shown in \cref{fig:raw_corrected_estimation_error_8_24}, \cref{fig:raw_corrected_estimation_error_16_16}, and \cref{fig:raw_corrected_estimation_error_22_10}, respectively.
+As expected, since  $\symPrecision + \symRegRange = 32$ is kept constant, the operating range remains more or less the same, while the relative error decreases with increasing precision.
+Again, the new algorithm gives essentially unbiased estimates.
+Only for very large precisions, an oscillating bias becomes apparent (compare \cref{fig:raw_corrected_estimation_error_22_10}), that is caused by approximating the periodic function $\symPowerSeriesFunc$ by a constant (see \cref{sec:derivation_raw_estimator}).
 
@@
 \centering
-\includesvg[width=1\textwidth]{raw_corrected_estimate_12_20}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
-\label{fig:raw_corrected_estimation_error_12_20}
-\end{figure}
-
-\begin{figure}
-\centering
 \includesvg[width=1\textwidth]{raw_corrected_estimate_16_16}
@@
 \label{fig:raw_corrected_estimation_error_22_10}
+\end{figure}
+
+As proposed in \cite{Heule2013}, the operating range can be extended by
+replacing the 32-bit hash function by a 64-bit hash function.
+\cref{fig:raw_corrected_estimation_error_12_52} shows the relative error for such a HyperLogLog configuration with precision  $\symPrecision=12$.
+In this case, in order to use all the 64 bits of the hash value, $\symRegRange$ must be chosen to be equal to $64 - \symPrecision = 52$.
+As a consequence, in order to represent the maximum possible register value $\symRegRange+1 = 53$, 6 bits are needed for each register.
+The doubled hash value size shifts the maximum trackable cardinality value towards $2^{64}$.
+As \cref{fig:raw_corrected_estimation_error_12_52} shows, when compared to the 32-bit hash value case given in \cref{fig:raw_corrected_estimation_error_12_20}, the estimation error remains constant over the entire simulated cardinality range up to 50 billions.
+
+\begin{figure}
+\centering
+\includesvg[width=1\textwidth]{raw_corrected_estimate_12_52}
+\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
+\label{fig:raw_corrected_estimation_error_12_52}
+\end{figure}
+
+We also evaluated the case $\symPrecision = 12$ and $\symRegRange=14$, which is interesting, because the register values are limited to the range $[0, 15]$.
+As a consequence, 4 bits are sufficient for representing a single register value.
+This allows two registers to share a single byte, which is beneficial from a performance perspective.
+Nevertheless, this configuration still allows the estimation of cardinalities up to 100 millions as shown in \cref{fig:raw_corrected_estimation_error_12_14}, which could be enough for many applications.
+
+\begin{figure}
+\centering
+\includesvg[width=1\textwidth]{raw_corrected_estimate_12_14}
+\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
+\label{fig:raw_corrected_estimation_error_12_14}
 \end{figure}
@@
 
-Note that \eqref{equ:funcdef} can be solved analytically for the special case $\symRegRange=0$ which corresponds to the already mentioned linear probabilistic counting algorithm.
+Note that \eqref{equ:funcdef} can be solved analytically for the special case $\symRegRange=0$ which corresponds to the already mentioned linear counting algorithm.
 In this case, the maximum likelihood method under the Poisson model directly leads to the estimator presented in \cite{Whang1990} and which was also used for small range estimation \eqref{equ:small_range_estimate}.
@@
 \subsection{Estimation error}
-In order to verify the new estimation algorithm, we generated 10\,000 HyperLogLog sketches and inserted up to 50 billion unique elements.
-Element hash values have been mocked by random numbers.
-For the following results we used the Mersenne Twister random number generator with 19937 bit state size from the C++ standard library.
-
-\cref{fig:estimation_error_12_20} shows the distribution of the relative error of the estimated cardinality using \cref{alg:estimate_ml} compared to the true cardinality for $\symPrecision=12$ and $\symRegRange=20$.
-As the median shows, the error is essentially unbiased over the entire cardinality range except for very small values.
-The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is about an order of magnitude larger than the operating range upper bound of the raw estimate (\cref{fig:raw_estimate}) or the original method (\cref{fig:original_estimate}).
-
+
+see \cref{fig:estimation_error_12_20}, \cref{fig:estimation_error_8_24}, \cref{fig:estimation_error_16_16},\cref{fig:estimation_error_22_10},\cref{fig:estimation_error_12_52},\cref{fig:estimation_error_12_14}
 \begin{figure}
@@
 
-The new estimation algorithm also works well for other HyperLogLog configurations.
-First we considered the original HyperLogLog algorithm that used a 32-bit hash function ($\symPrecision + \symRegRange = 32$).
-The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, $\symPrecision=22$ are shown in \cref{fig:estimation_error_8_24}, \cref{fig:estimation_error_16_16}, and \cref{fig:estimation_error_22_10}, respectively.
-As expected, since  $\symPrecision + \symRegRange = 32$ is kept constant, the operating range remains more or less the same, while the relative error decreases with increasing precision.
-Again, the new algorithm gives inherently unbiased estimates.
-
 \begin{figure}
@@
 
-As proposed in \cite{Heule2013}, the operating range can be extended by
-replacing the 32-bit hash function by a 64-bit hash function.
-\cref{fig:estimation_error_12_52} shows the relative error for such a HyperLogLog configuration with precision  $\symPrecision=12$.
-In this case, in order to use all the 64 bits of the hash value, $\symRegRange$ must be chosen to be equal to $64 - \symPrecision = 52$.
-As a consequence, in order to represent the maximum possible register value $\symRegRange+1 = 53$, 6 bits are needed for each register.
-The doubled hash value size shifts the maximum trackable cardinality value towards $2^{64}$.
-As \cref{fig:estimation_error_12_52} shows, when compared to the 32-bit hash value case given in \cref{fig:estimation_error_12_20}, the estimation error remains constant over the entire simulated cardinality range up to 50 billions.
-
 \begin{figure}
@@
 \end{figure}
-
-We also evaluated the case $\symPrecision = 12$ and $\symRegRange=14$, which is interesting, because the register values are limited to the range $[0, 15]$.
-As a consequence, 4 bits are sufficient for representing a single register value.
-This allows two registers to share a single byte, which is beneficial from a performance perspective.
-Nevertheless, this configuration still allows the estimation of cardinalities up to 100 millions as shown in \cref{fig:estimation_error_12_14}, which could be enough for many applications.
 
@@
 \end{equation}
-If $\symHelper$ is not calculated exactly but approximated by an approximation $\symHelperApprox$ with maximum relative error $\symError_\symHelper\ll 1$
+If $\symHelper$ is not calculated exactly but approximated by $\symHelperApprox$ with maximum relative error $\symError_\symHelper\ll 1$
 \begin{equation}

### 2016-06-13 python/power-series-function.py
minor changes, fixed font style in chart

@@ -57,6 +57,9 @@ res2 = minimize_scalar(f2, bounds = (0,1), method='bounded')
 print(res1)
 print(res2)
 
+plt.rc('text', usetex=True)
+plt.rc('font', family='serif')
+
 fig1, ax1 = plt.subplots(subplot_kw=dict(xlim=[xMin,xMax], ylim=[-1.1e-5,1.1e-5]))
 fig1.set_size_inches(6, 4)
 ax1.plot(xValues, yValues, color='black')

### 2016-06-14 paper/paper.tex
minor changes, fixed font style in chart
refactored bibliography
some minor changes

@@
 \documentclass[a4paper]{scrartcl}
-%\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
+\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
 \usepackage{amsmath}
@@
 Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers.
-The result is the so-called raw HyperLogLog estimate and given by
+The result is the so-called raw estimator and given by
 \begin{equation}
@@
 \label{equ:small_range_estimate}
-\symCardinalityEstimate_\text{small} = \symNumReg \log\left(
-\frac{\symNumReg}{\symCount_0}
-\right).
+\symCardinalityEstimate_\text{small} = \symNumReg \log(\symNumReg/\symCount_0).
 \end{equation}
@@
 =
--2^{32}\log\left(1-\frac{\symCardinalityOriginalEstimate}{2^{32}}\right).
+-2^{32}\log\!\left(1-\symCardinalityOriginalEstimate/2^{32}\right).
 \end{equation}
@@
 =
-\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\left( \frac{\symPoissonRate}{\symNumReg}\right)\right)}{\symPoissonRate},
+\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\!\left( \frac{\symPoissonRate}{\symNumReg}\right)\right)}{\symPoissonRate},
 \end{multline}
@@
 \symExpectation\left(
-\frac{\symAlpha_\infty\,\symNumReg^2\,\symPowerSeriesFunc\!\left(\log_2\left( \symPoissonRate/\symNumReg\right)\right)}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
+\frac{\symAlpha_\infty\,\symNumReg^2\,\symPowerSeriesFunc\!\left(\log_2\!\left( \symPoissonRate/\symNumReg\right)\right)}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
 \right)
@@
 =
-\frac{\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right)}{\symPowerSeriesFunc\left(\log_2\left(\log\left(\frac{\symNumReg}{\symCount_0}\right)\right)\right)}
+\frac{\symNumReg\log\!\left(\frac{\symNumReg}{\symCount_0}\right)}{\symPowerSeriesFunc\!\left(\log_2\!\left(\log\!\left(\frac{\symNumReg}{\symCount_0}\right)\right)\right)}
 \approx
-\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right),
+\symNumReg\log\!\left(\symNumReg/\symCount_0\right),
 \end{equation}
@@
 \includesvg[width=1\textwidth]{corrected_raw_avg_exec_time}
-\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.
+\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3\,GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.
 Both cases, $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ precalculated and calculated on-demand have been considered.}
@@
 Note that \eqref{equ:funcdef} can be solved analytically for the special case $\symRegRange=0$ which corresponds to the already mentioned linear counting algorithm.
-In this case, the maximum likelihood method under the Poisson model directly leads to the estimator presented in \cite{Whang1990} and which was also used for small range estimation \eqref{equ:small_range_estimate}.
+In this case, the maximum likelihood method under the Poisson model directly leads to the linear counting  estimator presented in \cite{Whang1990} and which was also used for small range estimation \eqref{equ:small_range_estimate}.
 Despite the assumption of a Poisson model, it is a very good approximation of the optimal martingale estimator presented in \cite{Ting2014}.
-Due to this fact we could expect that maximum likelihood estimation under the Poisson model also works very well for the more  general HyperLogLog case.
+Due to this fact we could expect that maximum likelihood estimation under the Poisson model also works well for the more general HyperLogLog case.
 
@@
 \left(\symNumReg-\symCount_0\right)\cdot
-\symHelper\left(
+\symHelper\!\left(
 \symX\cdot
@@
 }
-\log\left(
+\log\!\left(
 1
@@
 The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:avg_exec_time}.
-All these benchmarks where carried out on an Intel Core i5-2500K clocking at 3.3GHz.
+All these benchmarks where carried out on an Intel Core i5-2500K clocking at 3.3\,GHz.
 As can be seen, the average calculation time did never exceed 700\,ns.
@@
 \includesvg[width=1\textwidth]{max_likelihood_avg_exec_time}
-\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
+\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3\,GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
 \label{fig:avg_exec_time}
@@
 +
-\log\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_1}}}\right)
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_1}}}\right)
 \left[1\leq\symRegVal_1<\symRegVal_2\right]
@@
 +
-\log\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}\right)
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}\right)
 \left[\symRegVal_2<\symRegVal_1\right]
@@
 +
-\log\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_2}}}\right)
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_2}}}\right)
 \left[1\leq\symRegVal_2<\symRegVal_1\right]
@@
 +
-\log\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal_2, \symRegRange\right)}}}\right)
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal_2, \symRegRange\right)}}}\right)
 \left[\symRegVal_1<\symRegVal_2\right]
@@
 +
-\log\left(
+\log\!\left(
 1
@@
 \sum_{\symRegVal=1}^{\symRegRange+1}
-\log\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
 \symCount_\symRegVal^\rightarrow
 +
-\log\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
 \symCount_\symRegVal^\leftarrow
@@
 \sum_{\symRegVal=1}^{\symRegRange+1}
-\log\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
 \symCount_\symRegVal^\downarrow
 +
-\log\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
 \symCount_\symRegVal^\uparrow
@@
 \sum_{\symRegVal=1}^{\symRegRange+1}
-\log\left(
+\log\!\left(
 1

### 2016-06-15 paper/paper.tex
performance section for corrected raw estimation algorithm

@@
 \title{New cardinality estimation algorithms for HyperLogLog sketches}
-\subtitle{Draft version, \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}}
+\subtitle{Draft version \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}}
 \author{Otmar Ertl}
@@
 To reduce the estimation error for cardinalities close to this region, it was proposed to correct $\symCardinalityOriginalEstimate$ for bias.
-The empirically determined bias correction data can be either stored as set of interpolation points \cite{Heule2013}, as lookup table \cite{Rhodes2015}, or as best-fitting polynomial \cite{Sanfilippo2014}.
+Empirically collected bias correction data is either stored as set of interpolation points \cite{Heule2013}, as lookup table \cite{Rhodes2015}, or as best-fitting polynomial \cite{Sanfilippo2014}.
+However, all these empirical approaches treat the symptom and not the cause.
 
@@
 In order to verify the new estimation algorithm, we generated 10\,000 HyperLogLog sketches and inserted up to 50 billion unique elements.
-Element hash values have been mocked by random numbers.
+Assuming a strong universal hash function, element hash values can be mocked by random numbers.
 For the following results we used the Mersenne Twister random number generator with 19937 bit state size from the C++ standard library.
@@
 \subsection{Performance}
-see \cref{fig:corrected_raw_avg_exec_time}
+In order to evaluate the performance of the corrected raw estimation algorithm, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch.
+For different cardinalities we loaded the precalculated multiplicity vectors of 1000 randomly generated HyperLogLog sketches into main memory.
+The average computation time was determined by cycling over these multiplicity vectors and passing them as input to the algorithm.
+For each evaluated cardinality value the average execution time was calculated after 100 cycles which corresponds to 100\,000 algorithm executions.
+The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:corrected_raw_avg_exec_time}.
+Two variants of \cref{alg:corrected_raw_estimation} have been evaluated for which the functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ have been either calculated on-demand or taken from a lookup table.
+All these benchmarks where carried out on an Intel Core i5-2500K clocking at 3.3\,GHz.
+
+The results show that the execution times are nearly constant for $\symRegRange=52$.
+Using the lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed, because $\symCount_{53}$ is equal to zero with high probability for realistic cardinalities.
+
+For the case $\symRegRange=20$ with precalculated correction values the computation time is again -- as expected -- independent of the cardinality.
+The faster computation times compared to the $\symRegRange=52$ case can be explained by the much smaller dimension of the multiplicity vector which is equal to $\symRegRange+2$.
+If the functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are calculated on-demand, the execution times for larger cardinalities are doubled.
+This comes from the fact that the calculation of $\symLargeCorrectionFunc$ is much more expensive than that of $\symSmallCorrectionFunc$, because it requires more iterations and involves square root evaluations.
+However, we can imagine that a better numerical approximation of $\symLargeCorrectionFunc$ can be found than that given in \cref{alg:corrected_raw_estimation} and which allows on-demand evaluation without significant extra costs.
+
+The numbers do not yet include the required processing time to extract the multiplicity vector out from the HyperLogLog sketch, which requires a complete scan over all registers and counting the different register values into an array.
+A theoretical lower bound for this processing time can be derived using the maximum memory bandwidth of the CPU, which is 21\,GB/s for an Intel Core i5-2500K.
+For precision $\symPrecision=12$ there are 4096 registers, each of them requires at least 5 bits.
+Hence, the total data size of the HyperLogLog sketch is 2.5\,kB minimum and transfer time from main memory to CPU will be at least 120\,ns.
+Having this value in mind, the presented numbers for estimating the cardinality from the multiplicity vector are quite satisfying.
 
@@
 \subsection{Performance}
-In order to prove that the new algorithm is not only accurate, but also fast, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch.
+see \cref{fig:avg_exec_time}
+%In order to prove that the new algorithm is not only accurate, but also fast, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch.
 For different cardinalities we loaded the multiplicity vectors of 1000 precalculated and randomly generated HyperLogLog sketches into main memory.
@@
 As can be seen, the average calculation time did never exceed 700\,ns.
-
-The numbers do not yet include the required processing time to calculate the multiplicity vector from the HyperLogLog sketch, which requires a complete scan over all registers and counting the different register values into an array.
-A theoretical lower bound for this processing time can be derived using the maximum memory bandwidth of the CPU, which is 21\,GB/s for an Intel Core i5-2500K.
-For precision $\symPrecision=12$ there are 4096 registers, each of them requires at least 5 bits.
-Hence, the total data size of the HyperLogLog sketch is 2.5\,kB minimum and transfer time from main memory to CPU will be at least 120\,ns.
-Having this value in mind, the presented processing time numbers are quite satisfying.
 

### 2016-06-15 paper/paper.tex
minor changes

@@
 -\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCount_\symRegVal}{2^\symRegVal}+ 
-\sum_{\symRegVal=1}^{\symRegRange}\symCount_\symRegVal \log\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)
-+
-\symCount_{\symRegRange+1} \log\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegRange}}}\right)
+\sum_{\symRegVal=1}^{\symRegRange}\symCount_\symRegVal \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)
++
+\symCount_{\symRegRange+1} \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegRange}}}\right)
 \end{equation}
@@
 \symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
-\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\left(\frac{\symX}{2^\symRegVal}\right)
-+
-\symCount_{\symRegRange+1}\symHelper\left(\frac{\symX}{2^\symRegRange}\right)
+\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\!\left(\frac{\symX}{2^\symRegVal}\right)
++
+\symCount_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
 -
@@
 \symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
-\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
-+
-\symCount_{\symRegRange+1}\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
+\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
++
+\symCount_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
 -
@@
 \item Many programming languages allow the efficient multiplication and division by any integral power of 2 using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
-\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\lbrace\frac{\symX}{2^{\symRegVal'_\text{max}}},\frac{\symX}{2^{\symRegVal'_\text{max}-1}}\ldots,\frac{\symX}{2^{\symRegVal_\text{min}}}\rbrace$ where $\symRegVal'_\text{max} := \min(\symRegVal_\text{max}, \symRegRange)$.
+\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\left\lbrace\frac{\symX}{2^{\symRegVal'_\text{max}}},\frac{\symX}{2^{\symRegVal'_\text{max}-1}}\ldots,\frac{\symX}{2^{\symRegVal_\text{min}}}\right\rbrace$ where $\symRegVal'_\text{max} := \min(\symRegVal_\text{max}, \symRegRange)$.
 This series corresponds to a geometric series with ratio 2.
 A straightforward calculation using \eqref{equ:helper} is very expensive because of the exponential function.
-However, if we already know $\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
+However, if we already know $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
 \begin{equation}
 \label{equ:helper_recursion1}
-\symHelper\left(2\symX\right) = \frac{\symX+2\symHelper\left(\symX\right)\left(1-\symHelper\left(\symX\right)\right)}{\symX+2\left(1-\symHelper\left(\symX\right)\right)}
+\symHelper\!\left(2\symX\right) = \frac{\symX+2\symHelper\!\left(\symX\right)\left(1-\symHelper\!\left(\symX\right)\right)}{\symX+2\left(1-\symHelper\!\left(\symX\right)\right)}
 \end{equation}
@@
 \label{equ:helper_recursion2}
-\symHelper\left(\frac{\symX}{2^{\symRegVal}}\right) = \frac{\frac{\symX}{2^{\symRegVal+2}}+\symHelper\left(\frac{\symX}{2^{\symRegVal+1}}\right)\left(1-\symHelper\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}{\frac{\symX}{2^{\symRegVal+2}}+\left(1-\symHelper\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}
-\end{equation}
-Note, this recursive formula is stable in a sense that the relative error of $\symHelper\left(2\symX\right)$ is smaller than that of $\symHelper\left(\symX\right)$ as shown in \cref{app:helper_stable}.
-
-\item If $\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is smaller than 0.5, the function $\symHelper(\symX)$ can be well approximated by a Taylor series around $\symX=0$
+\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right) = \frac{\frac{\symX}{2^{\symRegVal+2}}+\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}{\frac{\symX}{2^{\symRegVal+2}}+\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}
+\end{equation}
+Note, this recursive formula is stable in a sense that the relative error of $\symHelper\!\left(2\symX\right)$ is smaller than that of $\symHelper\!\left(\symX\right)$ as shown in \cref{app:helper_stable}.
+
+\item If $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is smaller than 0.5, the function $\symHelper(\symX)$ can be well approximated by a Taylor series around $\symX=0$
 \begin{equation}
@@
 \end{equation}
-In fact, $\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)\leq 0.5$ is almost always fulfilled as long as registers are not saturated.
+In fact, $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)\leq 0.5$ is almost always fulfilled as long as registers are not saturated.
 Using \eqref{equ:strong_upper_bound} it is straightforward to see that $\frac{\symX}{2^{\symRegVal'_\text{max}}} \leq \log 2 \approx 0.693$, if $\symCount_{\symRegRange+1}=0$.
@@
 By definition, $\frac{\symX}{2^{\symE+1}}\leq 0.5$ which allows using the Taylor approximation.
-$\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is finally obtained after $\symE + 1 -\symRegVal'_\text{max}$ iterations using \eqref{equ:helper_recursion2}.
+$\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is finally obtained after $\symE + 1 -\symRegVal'_\text{max}$ iterations using \eqref{equ:helper_recursion2}.
 \end{itemize}
@@
 \symXEstimate\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
-\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\left(\frac{\symXEstimate}{2^\symRegVal}\right)
-+
-\symCount_{\symRegRange+1}\symHelper\left(\frac{\symXEstimate}{2^\symRegRange}\right)
+\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\!\left(\frac{\symXEstimate}{2^\symRegVal}\right)
++
+\symCount_{\symRegRange+1}\symHelper\!\left(\frac{\symXEstimate}{2^\symRegRange}\right)
 =
@@
 \symXEstimate\left(1+\symError_\symX\right)\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
-\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelperApprox\left(\frac{\symXEstimate\left(1+\symError_\symX\right)}{2^\symRegVal}\right)
+\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelperApprox\!\left(\frac{\symXEstimate\left(1+\symError_\symX\right)}{2^\symRegVal}\right)
 +
 \symCount_{\symRegRange+1}
-\symHelperApprox\left(
+\symHelperApprox\!\left(
 \frac{\symXEstimate\left(1+\symError_\symX\right)}{2^\symRegRange}\right)
@@
 \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal
-\symHelper\left(\frac{\symXEstimate\left(1+\symError_\symX\right)
+\symHelper\!\left(\frac{\symXEstimate\left(1+\symError_\symX\right)
 }{2^\symRegVal}\right)
@@
 \symCount_{\symRegRange+1}
-\symHelper\left(
+\symHelper\!\left(
 \frac{\symXEstimate\left(1+\symError_\symX\right)}{2^\symRegRange}\right)
@@
 \symCount_\symRegVal
-\symHelper\left(\frac{\symXEstimate
+\symHelper\!\left(\frac{\symXEstimate
 }{2^\symRegVal}\right)
@@
 \symCount_{\symRegRange+1}
-\symHelper\left(
+\symHelper\!\left(
 \frac{\symXEstimate}{2^\symRegRange}\right)

### 2016-06-15 paper/paper.tex
reformatted some equations

@@
 
-Using the identity $2\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = -\symPowerSeriesFunc(\log_2(-\log(\symX)))/\left(\log(2)\log(\symX)\right)$, we get for the linear counting case with $\symRegRange=0$
+Using the identity $2\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = \symPowerSeriesFunc(\log_2(\log(1/\symX)))/\left(\log(2)\log(1/\symX)\right)$, we get for the linear counting case with $\symRegRange=0$
 \begin{equation}
@@
 =
-\frac{\symNumReg\log\!\left(\frac{\symNumReg}{\symCount_0}\right)}{\symPowerSeriesFunc\!\left(\log_2\!\left(\log\!\left(\frac{\symNumReg}{\symCount_0}\right)\right)\right)}
+\frac{\symNumReg\log\!\left(\symNumReg/\symCount_0\right)}{\symPowerSeriesFunc\!\left(\log_2\!\left(\log\!\left(\symNumReg/\symCount_0\right)\right)\right)}
 \approx

### 2016-06-17 paper/paper.tex
modified paper

@@
 
-% !TEX TS-program = pdflatex
-% !TEX encoding = UTF-8 Unicode
-
-% This is a simple template for a LaTeX document using the "article" class.
-% See "book", "report", "letter" for other types of document.
-
 \documentclass[a4paper]{scrartcl}
-\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
+\usepackage[utf8]{inputenc}
 \usepackage{amsmath}
 \usepackage{amssymb}
-\usepackage{graphicx} % support the \includegraphics command and options
+\usepackage{graphicx}
 \usepackage{svg}
@@
 \usepackage{url}
-\usepackage{booktabs} % for much better looking tables
-\usepackage{array} % for better arrays (eg matrices) in maths
-\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
-\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
-\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
-% These packages are all incorporated in the memoir class to one degree or another...
+\usepackage{booktabs}
+\usepackage{array}
+\usepackage{paralist}
+\usepackage{verbatim}
+\usepackage{subfig}
 \usepackage{hyperref}
@@
 
-% \renewcommand{\sffamily}{\rmfamily}
-
 % SYMBOLS
 \newcommand*{\symDefine}[2]{\newcommand{{#1}}{{#2}}}
-
 \symDefine{\symPrecision}{p}
@@
 
-%%% END Article customizations
-
 \title{New cardinality estimation algorithms for HyperLogLog sketches}
@@
 All registers start with initial value 0.
-The HyperLogLog algorithm successively increases the register values up to a maximum value limited by the space per register which is usually between 5 and 8 bits.
+Each element insertion potentially increases the value of one of these registers.
+The maximum value a register can reach is a natural bound given either by the output size of the used hash algorithm or the space that is reserved for a single register.
+Common implementations allocate up to 8 bits per register.
 
@@
 
-The number of bits of the consumed hash value $\symPrecision+\symRegRange$ limit the maximum cardinality that can be tracked.
-Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent, which will reduce estimation accuracy drastically.
+For element insertion a hash value with bit size $\symPrecision+\symRegRange$ is required.
+The consumed number of bits also limits the maximum cardinality that can be tracked.
+Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent and the estimation accuracy is drastically reduced.
 
@@
 
-Note that a $(\symPrecision, 0)$-HyperLogLog sketch corresponds to a bit array as used by linear counting \cite{Whang1990}.
+\cref{alg:insert} has some properties which are especially useful in distributed environments.
+The insertion order of elements has no influence on the final HyperLogLog sketch.
+Furthermore, HyperLogLog sketches with same parameters $(\symPrecision, \symRegRange)$ can be easily merged.
+The resulting register values are simply the register-wise maximum values.
+The final HyperLogLog sketch corresponds to the union of both sets that have been represented by the input sketches.
+
+A $(\symPrecision, 0)$-HyperLogLog sketch corresponds to a bit array as used by linear counting \cite{Whang1990}.
 In this case each register value is represented by a single bit.
@@
 \subsection{Depoissonization}
+\label{sec:depoissonization}
 Due to the simpler probability mass function, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\vec{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ rather than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}.
@@
 \begin{algorithm}
-\caption{Cardinality estimation algorithm based on the corrected raw estimate.}
+\caption{Cardinality estimation algorithm based on the corrected raw estimator.}
 \label{alg:corrected_raw_estimation}
@@
 \section{Maximum likelihood estimation}
-
-The maximum likelihood method was already applied to HyperLogLog sketches in \cite{Clifford2012}.
-However, there, a different element insertion algorithm was assumed that works with time complexity $\symBigO(\symNumReg)$, because each register calculates its own hash value from the inserted element and updates its current value accordingly.
-In comparison, \cref{alg:insert} uses only a single hash value that defines which register needs to be updated.
-As a consequence, the register values are statistically dependent, and the Poisson approximation is needed to make the register values statistically independent, which allows the factorization of the joint probability distribution of all register values.
-
-In the following we derive a new robust and efficient cardinality estimation algorithm that is based on the maximum likelihood method.
+We know from \cref{sec:depoissonization} that any unbiased estimator for the Poisson parameter is also an unbiased estimator for the cardinality.
+We know that under suitable regularity conditions of the probability mass function the maximum likelihood estimator is asymptotically efficient \cite{Casella2002}.
+This means, if the number of registers $\symNumReg$ is large enough, the maximum likelihood method should give us an unbiased estimator for the cardinality.
+
+We are not the first applying the maximum likelihood method to HyperLogLog sketches \cite{Clifford2012}.
+However, there a different element insertion algorithm was considered, which is very inefficient in practice due to its $\symBigO(\symNumReg)$ runtime complexity.
+Each register has its own hash function which is applied to the inserted element.
+The result is used to update the corresponding register.
+The advantage is that the register values are statistically independent in this case, which allows factorization of the joint probability distribution of all register values.
+
+In contrast, \cref{alg:insert} uses what is also called stochastic averaging.
+Instead of iterating through all registers, a single hash value is used, to select a single register and to update its value.
+Since the register values are statistically dependent in this case, the maximum likelihood method is much more complicated to apply.
+However, with the help of the Poisson approximation we are finally able to derive a new robust and efficient cardinality estimation algorithm.
 Furthermore, we will demonstrate that consequent application of the maximum likelihood method reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values.
@@
 \tilde{\symHelper}_2 = 
-\frac{\symX+2\tilde{\symHelper}_1(1-\tilde{\symHelper}_1)}{\symX+2(1-\tilde{\symHelper_1})}
+\frac{\symX+2\tilde{\symHelper}_1(1-\tilde{\symHelper}_1)}{\symX+2(1-\tilde{\symHelper}_1)}
 \end{equation}

### 2016-06-18 paper/paper.tex
some improvements

@@
 Here $\vec{\symRegValVariate}$ denotes the vector of all observed register values.
-$\vec{\symRegValVariate}$ is distributed according to \eqref{equ:multinomialProbabilityMass}.
+Each component of $\vec{\symRegValVariate}$ is distributed according to \eqref{equ:multinomialProbabilityMass}.
 Depoissonization \cite{Jacquet1998} finally allows to translate the estimates back to the fixed-size model.
@@
 The original cardinality estimate \cite{Flajolet2007} is based on the idea that approximately $\symNumReg  2^{\symRegVal_\symIndexJ}$ distinct element insertions are needed until the value of register $\symIndexJ$ reaches $\symRegVal_\symIndexJ$.
-Given that, a cardinality estimate can be obtained by calculating the average over the values $\lbrace \symNumReg 2^{\symRegVal_1},\ldots,\symNumReg2^{\symRegVal_\symNumReg}\rbrace$.
+Given that, a rough cardinality estimate can be obtained by calculating the average over the values $\lbrace \symNumReg 2^{\symRegVal_1},\ldots,\symNumReg2^{\symRegVal_\symNumReg}\rbrace$.
 
@@
 These approximations are used in many HyperLogLog implementations.
-However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions.
+However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision$.
 For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use
@@
 \end{equation}
-as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible if compared to the estimation error.
-
-\cref{fig:raw_estimate} shows the distribution of the relative error for the raw estimate as function of the cardinality.
+as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible compared to the estimation error.
+
+\cref{fig:raw_estimate} shows the distribution of the relative error for the raw estimator as function of the cardinality.
 The chart is based on 10\,000 randomly generated HyperLogLog sketches.
@@
 \end{equation}
-
 The original estimation algorithm as presented in \cite{Flajolet2007} including small and large range corrections is summarized by \cref{alg:estimate_original}. 
 \begin{algorithm}
-\caption{Original procedure for estimating the cardinality from a  HyperLogLog data structure using 32-bit hash values ($\symPrecision+\symRegRange = 32$) for insertion of data items \cite{Flajolet2007}.}
+\caption{Original cardinality estimation algorithm for HyperLogLog sketches that use 32-bit hash values ($\symPrecision+\symRegRange = 32$) for insertion of data items \cite{Flajolet2007}.}
 \label{alg:estimate_original}
@@
 However, instead of underestimating the cardinalities, they are now overestimated.
+Another indication for the incorrectness of the proposed large range correction \eqref{equ:large_range_estimate} is the fact that it is not even defined for all possible states.
+For instance, consider a $(\symPrecision,\symRegRange)$-HyperLogLog sketch with $\symPrecision+\symRegRange=32$ for which all registers are equal to the maximum possible value $\symRegRange+1$.
+The raw estimate would be $\symCardinalityEstimate_\text{raw} = \symAlpha_\symNumReg 2^{33}$, which is greater than $2^{32}$ and outside of the domain of the large range correction.
+
 A simple approach to avoid the large range correction at all is the use of larger hash values $(\symPrecision+\symRegRange>32)$ which shift the biased region of $\symCardinalityOriginalEstimate$ to higher cardinalities.
@@
 As we have already seen in \cref{fig:raw_estimate}, the raw estimator does not work very well for small or large cardinalities.
-The reason for this is that the distribution of register values \eqref{equ:register_value_distribution} is different from \eqref{equ:assumed_register_val_distribution} for which the raw estimator was derived.
+The reason is that the distribution of register values \eqref{equ:register_value_distribution} is different from \eqref{equ:assumed_register_val_distribution} for which the raw estimator was derived.
 Since a random variable $\symRegValVariate'$ that obeys \eqref{equ:assumed_register_val_distribution} can be transformed into random variable $\symRegValVariate$ that follows \eqref{equ:register_value_distribution} using the transformation $\symRegValVariate = \min\left(\max\left(\symRegValVariate', 0\right), \symRegRange+1\right)$, the register values of a HyperLogLog sketch can be seen as the result after applying this transformation to a sample that is distributed according to \eqref{equ:assumed_register_val_distribution}.
@@
 \end{equation}
-From \eqref{equ:register_value_distribution} we know that $\symExpectation\left(\frac{\symCount_0}{\symNumReg}\right)=
-e^{-\frac{\symPoissonRate}{\symNumReg}}$ and $\symExpectation\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)=
+From \eqref{equ:register_value_distribution} we know that $\symExpectation\!\left(\symCount_0/\symNumReg\right)=
+e^{-\frac{\symPoissonRate}{\symNumReg}}$ and $\symExpectation\!\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)=
 e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}$, and therefore, we can also write
@@
 \symNumReg
-\symExpectation\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}
-\left(1-\symExpectation\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
+\left(\symExpectation\!\left(\symCount_0/\symNumReg\right)\right)^{2^{-\symRegVal}}
+\left(1-\left(\symExpectation\!\left(\symCount_0/\symNumReg\right)\right)^{2^{-\symRegVal}}\right)
 \end{equation}
@@
 \symNumReg
-\symExpectation\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}
-\left(1-\symExpectation\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}\right),
+\left(\symExpectation\!\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)\right)^{2^{\symRegRange-\symRegVal}}
+\left(1-\left(\symExpectation\!\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)\right)^{2^{\symRegRange-\symRegVal}}\right),
 \end{equation}
@@
 \symNumReg\cdot
-\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}\cdot
-\left(1-\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
+\left(\symCount_0/\symNumReg\right)^{2^{-\symRegVal}}\cdot
+\left(1-\left(\symCount_0/\symNumReg\right)^{2^{-\symRegVal}}\right)
 \end{equation}
@@
 \symNumReg\cdot
-\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}
+\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)^{2^{\symRegRange-\symRegVal}}
 \cdot
-\left(1-\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}\right)
+\left(1-\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)^{2^{\symRegRange-\symRegVal}}\right)
 \end{equation}
@@
 
-Inserting all these estimators into \eqref{equ:raw_estimate_with_multiplicities} finally gives
+Inserting all these estimators into \eqref{equ:raw_estimate_with_multiplicities} as replacements for $\symCount'_\symRegVal$ finally gives
 \begin{equation}
@@
 {
-\left(
-\symNumReg\,\symSmallCorrectionFunc(\symCount_0/\symNumReg) + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symNumReg\, \symLargeCorrectionFunc(1-\symCount_{\symRegRange+1}/\symNumReg) 2^{-(\symRegRange+1)}
-\right)}
+\symNumReg\,\symSmallCorrectionFunc(\symCount_0/\symNumReg) + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symNumReg\, \symLargeCorrectionFunc(1-\symCount_{\symRegRange+1}/\symNumReg) 2^{-\symRegRange}
+}
 \end{equation}
 which we call the corrected raw estimator.
-Here $\symNumReg\,\symSmallCorrectionFunc(\symCount_0/\symNumReg)$ and $\symNumReg\,\symLargeCorrectionFunc(1-\symCount_{\symRegRange+1}/\symNumReg)$ are replacements for  $\symCount_0$ and $\symCount_{\symRegRange+1}$ in the raw estimator \eqref{equ:intermediate_estimate}, respectively.
+Here $\symNumReg\,\symSmallCorrectionFunc(\symCount_0/\symNumReg)$ and $2\symNumReg\,\symLargeCorrectionFunc(1-\symCount_{\symRegRange+1}/\symNumReg)$ are replacements for  $\symCount_0$ and $\symCount_{\symRegRange+1}$ in the raw estimator \eqref{equ:intermediate_estimate}, respectively.
 The functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are defined as
@@
 \cdot
-2^{-(\symRegVal-1)}.
-\end{equation}
-
-Using the identity $2\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = \symPowerSeriesFunc(\log_2(\log(1/\symX)))/\left(\log(2)\log(1/\symX)\right)$, we get for the linear counting case with $\symRegRange=0$
+2^{-\symRegVal}.
+\end{equation}
+
+Using the identity $\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = \symAlpha_\infty\symPowerSeriesFunc(\log_2(\log(\symX^{-1})))/\log(\symX^{-1})$, we get for the linear counting case with $\symRegRange=0$
 \begin{equation}
@@
 = 
-\frac{\symNumReg}{\log(2)\left(
-2\symSmallCorrectionFunc(\symCount_0/\symNumReg)
+\frac{\symAlpha_\infty\symNumReg}{
+\symSmallCorrectionFunc(\symCount_0/\symNumReg)
 +
 \symLargeCorrectionFunc(\symCount_0/\symNumReg)
-\right)}
+}
 =
@@
 \State $\symNumReg \gets \left\Vert\vec{\symCount}\right\Vert_1$
-\State $\symZ\gets 0.5\cdot \symNumReg\cdot\symLargeCorrectionFunc(1 - \symCount_{\symRegRange+1}/\symNumReg)$
-\Comment{\parbox[t]{.5\linewidth}{alternatively, take $0.5\cdot \symNumReg\cdot\symLargeCorrectionFunc(1 - \symCount_{\symRegRange+1}/\symNumReg)$ from precalculated lookup table}}
+\State $\symZ\gets \cdot \symNumReg\cdot\symLargeCorrectionFunc(1 - \symCount_{\symRegRange+1}/\symNumReg)$
+\Comment{\parbox[t]{.5\linewidth}{alternatively, take $\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCount_{\symRegRange+1}/\symNumReg)$ from precalculated lookup table}}
 \For{$\symRegVal\gets \symRegRange, 1$}
@@
 \State $\symZ'\gets\symZ$
+\State $\symY\gets0.5\cdot\symY$
 \State $\symZ\gets\symZ + \left(1-\symX\right)\cdot\symX\cdot\symY$
-\State $\symY\gets0.5\cdot\symY$
 \Until{$\symZ=\symZ'$}

### 2016-06-18 c++/cardinality_calculator.cpp

@@ -187,7 +187,7 @@ int main(int argc, char* argv[])
                 flajoletSmallRangeEstimates[resultPos] = flajoletSmallRangeEstimate(c);
                 flajoletMidRangeEstimates[resultPos] = flajoletRawEstimate(c);
                 flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate_on_demand(c, numSmallCorrectionIterations[resultPos], numLargeCorrectionIterations[resultPos]);
-                //flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate2(c);
+                //flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate_precalculated(c);
                 flajoletEstimates[resultPos] = flajoletEstimate(c);
                 maxLikelihoodEstimates[resultPos] = maxLikelihoodEstimator.estimate(c, outerLoopIterationsCount[resultPos], innerLoop1IterationsCount[resultPos], innerLoop2IterationsCount[resultPos], logEvaluationCount[resultPos], kMin, kMax);
                 weakLowerBoundEstimates[resultPos] = weakLowerBoundEstimate(c);

### 2016-06-18 c++/cardinality_estimation.hpp

@@ -209,8 +209,8 @@ class CorrectedRawEstimator {
             numIterations += 1;
             x = std::sqrt(x);
             zPrime = z;
-            z += (1 - x)*x*y;
             y *= 0.5;
+            z += (1 - x)*x*y;
         } while(zPrime < z);
         return z;
     }
@@ -228,7 +228,7 @@ class CorrectedRawEstimator {
         int numIterations;
         std::vector<double> result(m+1);
         for (int c = 0; c <= m; ++c) {
-            result[c] = 0.5 * m * tau(static_cast<double>(m-c)/static_cast<double>(m), numIterations);
+            result[c] = m * tau(static_cast<double>(m-c)/static_cast<double>(m), numIterations);
         }
         return result;
     }
@@ -243,7 +243,7 @@ public:
         numSmallCorrectionIterations = 0;
         numLargeCorrectionIterations = 0;
 
-        double z = 0.5 * m * tau(static_cast<double>(m-c[q+1])/static_cast<double>(m), numLargeCorrectionIterations);
+        double z = m * tau(static_cast<double>(m-c[q+1])/static_cast<double>(m), numLargeCorrectionIterations);
         for (int k = q; k >= 1; --k) {
             z += c[k];
             z *= 0.5;

### 2016-06-19 paper/paper.tex
improvements

@@
 \symDefine{\symCardinalityEstimate}{\hat{\symCardinality}}
-\symDefine{\symCardinalityOriginalEstimate}{\symCardinalityEstimate_\text{raw}}
+\symDefine{\symCardinalityRawEstimate}{\symCardinalityEstimate_\text{raw}}
 \symDefine{\symError}{\varepsilon}
@@
 \label{equ:intermediate_estimate}
-\symCardinalityOriginalEstimate
+\symCardinalityRawEstimate
 = 
@@
 The corresponding relative estimation error as depicted in \cref{fig:small_range_estimate} shows that this estimator is convenient for small cardinalities.
-It was proposed to use this small-range estimator as long as $\symCardinalityOriginalEstimate \leq \frac{5}{2}\symNumReg$ where the factor $\frac{5}{2}$ was empirically determined \cite{Flajolet2007}.
+It was proposed to use this small-range estimator as long as $\symCardinalityRawEstimate \leq \frac{5}{2}\symNumReg$ where the factor $\frac{5}{2}$ was empirically determined \cite{Flajolet2007}.
 
@@
 =
--2^{32}\log\!\left(1-\symCardinalityOriginalEstimate/2^{32}\right).
+-2^{32}\log\!\left(1-\symCardinalityRawEstimate/2^{32}\right).
 \end{equation}
@@
 \State $\symNumReg \gets \dim \vec{\symRegVal}$
-\State $\symCardinalityOriginalEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegVal_\symIndexJ}\right)^{-1}$
+\State $\symCardinalityRawEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegVal_\symIndexJ}\right)^{-1}$
 \Comment raw estimate \eqref{equ:intermediate_estimate}
-\If{$\symCardinalityOriginalEstimate\leq \frac{5}{2}\symNumReg$}
+\If{$\symCardinalityRawEstimate\leq \frac{5}{2}\symNumReg$}
 \State $\symCount_0 = \left|\{\symIndexJ |\symRegVal_\symIndexJ=0\}\right|$
@@
 \Else
-\State \Return $\symCardinalityOriginalEstimate$
+\State \Return $\symCardinalityRawEstimate$
 \EndIf
-\ElsIf{$\symCardinalityOriginalEstimate\leq \frac{1}{30}2^{32}$}
-\State \Return $\symCardinalityOriginalEstimate$
+\ElsIf{$\symCardinalityRawEstimate\leq \frac{1}{30}2^{32}$}
+\State \Return $\symCardinalityRawEstimate$
 \Else
-\State\Return $-2^{32}\log(1-\symCardinalityOriginalEstimate/2^{32})$\Comment large range correction \eqref{equ:large_range_estimate}
+\State\Return $-2^{32}\log(1-\symCardinalityRawEstimate/2^{32})$\Comment large range correction \eqref{equ:large_range_estimate}
 \EndIf
@@
 The relative estimation error for the original method is shown in \cref{fig:original_estimate}.
-Unfortunatley, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityOriginalEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap.
+Unfortunatley, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityRawEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap.
 Therefore, the estimation error is much larger near the transition region.
-To reduce the estimation error for cardinalities close to this region, it was proposed to correct $\symCardinalityOriginalEstimate$ for bias.
+To reduce the estimation error for cardinalities close to this region, it was proposed to correct $\symCardinalityRawEstimate$ for bias.
 Empirically collected bias correction data is either stored as set of interpolation points \cite{Heule2013}, as lookup table \cite{Rhodes2015}, or as best-fitting polynomial \cite{Sanfilippo2014}.
@@
 For instance, consider a $(\symPrecision,\symRegRange)$-HyperLogLog sketch with $\symPrecision+\symRegRange=32$ for which all registers are equal to the maximum possible value $\symRegRange+1$.
-The raw estimate would be $\symCardinalityEstimate_\text{raw} = \symAlpha_\symNumReg 2^{33}$, which is greater than $2^{32}$ and outside of the domain of the large range correction.
-
-A simple approach to avoid the large range correction at all is the use of larger hash values $(\symPrecision+\symRegRange>32)$ which shift the biased region of $\symCardinalityOriginalEstimate$ to higher cardinalities.
-However, in case $\symRegRange \geq 30$ 6-bit registers as used in \cite{Heule2013} are required to allow representing all values in the range $[0, \symRegRange+1]$.
-If 64-bit hash values are used as proposed in \cite{Heule2013} $(\symPrecision+\symRegRange=64)$, a correction for large cardinalities can be circumvented.
-Cardinalities in the order of $2^{64}$ for which saturated registers become a problem never occur in practice.
+The raw estimate would be $\symCardinalityRawEstimate = \symAlpha_\symNumReg 2^{33}$, which is greater than $2^{32}$ and outside of the domain of the large range correction.
+
+A simple approach to avoid the need of any large range correction is to extend the operating range of the raw estimator to larger cardinalities.
+This can be accomplished by using hash values with more bits $(\symPrecision+\symRegRange>32)$.
+The drawback is that 5-bit registers are no longer sufficient to represent all possible values in the range $[0, \symRegRange+1]$, if $\symRegRange \geq 30$.
+For HyperLogLog sketches with 6-bit registers in combination with 64-bit hash values $(\symPrecision+\symRegRange=64)$ as proposed in \cite{Heule2013}, it is unrealistic in practice to encounter cardinalities of order $2^{64}$, which would require some correction.
 
@@
 =
-\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\!\left( \frac{\symPoissonRate}{\symNumReg}\right)\right)}{\symPoissonRate},
+\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\!\left(\symPoissonRate/\symNumReg\right)\right)}{\symPoissonRate},
 \end{multline}
@@
 \begin{equation}
-\symExpectation\left(
+\symExpectation\!\left(
 \frac{1}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
@@
 \frac{1}{\symExpectation
-\left(2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}\right)}
-=
-\frac{1}{\symNumReg \symExpectation\left(2^{-\symRegVal}\right)}.
+\!\left(2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}\right)}
+=
+\frac{1}{\symNumReg \symExpectation\!\left(2^{-\symRegVal}\right)}.
 \end{equation}
@@
 =
-\symExpectation\left(
+\symExpectation\!\left(
 \frac{\symAlpha_\infty\,\symNumReg^2\,\symPowerSeriesFunc\!\left(\log_2\!\left( \symPoissonRate/\symNumReg\right)\right)}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
@@
 \end{equation}
-Obviously, as long as $\symCount_0\approx\symCount'_0$ and  $\symCount'_{\symRegRange+1}\approx\symCount_{\symRegRange+1}$ the set of register values is similar to the originating sample.
-In case $\symCount_0=0$ and $\symCount_{\symRegRange+1}=0$, which implies $\symCount_\symRegVal=0$ for $\symRegVal\leq0$ and $\symRegVal\geq\symRegRange+1$, we can even be sure that the register values are identical to the originating sample.
-However, for small and large cardinalities $\symCount_0$ and $\symCount_{\symRegRange+1}$ approach $\symNumReg$, respectively, and the raw estimator needs to be corrected.
+Obviously, as long as $\symCount_0\ll\symNumReg$ and  $\symCount_{\symRegRange+1}\ll\symNumReg$ the set of register values is similar to the originating sample.
+In case $\symCount_0=0$ and $\symCount_{\symRegRange+1}=0$, which implies $\symCount'_\symRegVal=0$ for $\symRegVal\leq0$ and $\symRegVal\geq\symRegRange+1$, we can even be sure that the register values are identical to the originating sample.
+This explains why the raw estimator only works well in the intermediate cardinality range for which $\symCount_0$ and $\symCount_{\symRegRange+1}$ are negligible.
 
 \subsection{Corrections to the raw estimator}
-Given a HyperLogLog sketch with multiplicity vector $\vec{\symCount}$, we try to find estimates $\hat{\symCount}'_\symRegVal$ for $\symCount'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and use them instead in the estimation formula \eqref{equ:raw_estimate_with_multiplicities}.
-For $\symRegVal\in[1,\symRegRange]$ where $\symCount_\symRegVal = \symCount'_\symRegVal$ we have the trivial estimators
+Given a HyperLogLog sketch with multiplicity vector $\vec{\symCount}=\left(\symCount_0,\ldots,\symCount_{\symRegRange+1}\right)$, we try to find estimates $\hat{\symCount}'_\symRegVal$ for $\symCount'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and replace them in the estimation formula \eqref{equ:raw_estimate_with_multiplicities}.
+For $\symRegVal\in[1,\symRegRange]$ where $\symCount_\symRegVal = \symCount'_\symRegVal$ we can use the trivial estimators
 \begin{equation}
@@
 \subsection{Estimation error}
+\label{sec:corrected_raw_estimation_error}
 In order to verify the new estimation algorithm, we generated 10\,000 HyperLogLog sketches and inserted up to 50 billion unique elements.
@@
 The new estimation algorithm also works well for other HyperLogLog configurations.
-First we considered the original HyperLogLog algorithm that used a 32-bit hash function ($\symPrecision + \symRegRange = 32$).
+First we considered configurations using a 32-bit hash function ($\symPrecision + \symRegRange = 32$).
 The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, $\symPrecision=22$ are shown in \cref{fig:raw_corrected_estimation_error_8_24}, \cref{fig:raw_corrected_estimation_error_16_16}, and \cref{fig:raw_corrected_estimation_error_22_10}, respectively.
@@
 2^{\symRegVal'_\text{max}}
-\log\left(
+\log\!\left(
 1+
@@
 \end{equation}
-Note, if the HyperLogLog sketch is in the intermediate range, where $\symCount_0=\symCount_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimate \eqref{equ:intermediate_estimate}.
-Hence, the maximum likelihood method leads directly to the harmonic mean that is used by the raw estimator.
-If $\symCount_0>0$ or $\symCount_{\symRegRange+1}>0$ the bounds do no longer follow the harmonic mean and this is also the reason why the raw estimator fails for small and large cardinalities.
+If the HyperLogLog sketch is in the intermediate range, where $\symCount_0=\symCount_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimator \eqref{equ:intermediate_estimate}.
+Hence, the maximum likelihood method leads directly to the harmonic that is used by the raw estimator.
 
@@
 Since the expected statistical errror for the HyperLogLog data structure scales according to $\frac{1}{\sqrt{\symNumReg}}$ \cite{Flajolet2007}, it makes sense to choose $\symStopDelta = \frac{\symStopEpsilon}{\sqrt{\symNumReg}}$ with some constant $\symStopEpsilon$.
-For all following results we have used $\symStopEpsilon = 10^{-2}$.
+For all results that will follow in \cref{sec:maximum_likelihood_estimation_error} we have used $\symStopEpsilon = 10^{-2}$.
 
@@
 \begin{algorithm}
-\caption{Cardinality estimation}
+\caption{Maximum likelihood cardinality estimation}
 \ContinuedFloat
@@
 \begin{algorithm}
-\caption{Cardinality estimation (continued)}
+\caption{Maximum likelihood cardinality estimation (continued)}
 \begin{algorithmic}
@@
 \State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
-\Comment calculate $\symHelper\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion2}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
+\Comment calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion2}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
 \State $\symXNormalized \gets 2\symXNormalized$
@@
 \State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
-\Comment calculate $\symHelper\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion2}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
+\Comment calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion2}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
 \State $\symFuncPrime\gets \symFuncPrime + \symCount_\symRegVal\cdot\symHelper$
@@
 \subsection{Estimation error}
+\label{sec:maximum_likelihood_estimation_error}
+As in \cref{sec:corrected_raw_estimation_error}, where we have validated the corrected raw estimation algorithm, we ran again tests using the maximum likelihood estimation algorithm for different HyperLogLog parameters.
 
@@
 \begin{equation}
-\symRegValVariate_1 = \max\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right)
+\symRegValVariate_1 = \max\!\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right)
 ,\quad
-\symRegValVariate_2 = \max\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right)
+\symRegValVariate_2 = \max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right)
 \end{equation}
@@
 \begin{align}
-\symProbability\left(
+\symProbability\!\left(
 \symRegValVariate_1 \leq \symRegVal_1
@@
 &=
-\symProbability\left(
-\max\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_1
+\symProbability\!\left(
+\max\!\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_1
 \wedge
-\max\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_2
+\max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_2
 \right)
@@
 \symProbability
-\left(
+\!\left(
 \symRegValVariate_\symSetASuffix \leq \symRegVal_1
@@
 \wedge
-\symRegValVariate_\symSetXSuffix \leq \min\left(\symRegVal_1, \symRegVal_2\right)
+\symRegValVariate_\symSetXSuffix \leq \min\!\left(\symRegVal_1, \symRegVal_2\right)
 \right)
@@
 \symProbability
-\left(
+\!\left(
 \symRegValVariate_\symSetBSuffix \leq \symRegVal_2
@@
 \symProbability
-\left(
-\symRegValVariate_\symSetXSuffix \leq \min\left(\symRegVal_1, \symRegVal_2\right)
+\!\left(
+\symRegValVariate_\symSetXSuffix \leq \min\!\left(\symRegVal_1, \symRegVal_2\right)
 \right)

### 2016-06-19 paper/paper.tex
result sections for maximum likelihood approach

@@
 This can be accomplished by using hash values with more bits $(\symPrecision+\symRegRange>32)$.
-The drawback is that 5-bit registers are no longer sufficient to represent all possible values in the range $[0, \symRegRange+1]$, if $\symRegRange \geq 30$.
+The drawback is that 5-bit registers are no longer sufficient to represent all possible values in the range $[0, \symRegRange+1]$ in case $\symRegRange \geq 30$.
 For HyperLogLog sketches with 6-bit registers in combination with 64-bit hash values $(\symPrecision+\symRegRange=64)$ as proposed in \cite{Heule2013}, it is unrealistic in practice to encounter cardinalities of order $2^{64}$, which would require some correction.
@@
 \subsection{Performance}
+\label{sec:corrected_raw_estimation_algorithm}
 In order to evaluate the performance of the corrected raw estimation algorithm, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch.
@@
 \label{sec:maximum_likelihood_estimation_error}
-As in \cref{sec:corrected_raw_estimation_error}, where we have validated the corrected raw estimation algorithm, we ran again tests using the maximum likelihood estimation algorithm for different HyperLogLog parameters.
-
-see \cref{fig:estimation_error_12_20}, \cref{fig:estimation_error_8_24}, \cref{fig:estimation_error_16_16},\cref{fig:estimation_error_22_10},\cref{fig:estimation_error_12_52},\cref{fig:estimation_error_12_14}
+In order to investigate the estimation error for the maximum likelihood estimation algorithm, we investigated the same HyperLogLog configurations as in \cref{sec:corrected_raw_estimation_error} for the corrected raw estimation algorithm.
+\cref{fig:estimation_error_12_20}, \cref{fig:estimation_error_8_24}, \cref{fig:estimation_error_16_16},\cref{fig:estimation_error_22_10},\cref{fig:estimation_error_12_52}, and \cref{fig:estimation_error_12_14} show very similar results for various HyperLogLog parameters.
+
+What is different for the maximum likelihood estimation approach is a somewhat smaller median bias with less oscillations around zero for small cardinalities.
+Furthermore, contrary to the raw estimator which reveals a small oscillating bias for the mean (see \cref{fig:raw_corrected_estimation_error_22_10}), the maximum likelihood estimator seems to be completely unbiased (see \cref{fig:estimation_error_22_10}).
+
 \begin{figure}
@@
 \subsection{Performance}
-see \cref{fig:avg_exec_time}
-%In order to prove that the new algorithm is not only accurate, but also fast, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch.
-For different cardinalities we loaded the multiplicity vectors of 1000 precalculated and randomly generated HyperLogLog sketches into main memory.
-The average computation time was determined by cycling over these multiplicity vectors and passing them as input to the new estimation algorithm.
-For each evaluated cardinality value the average execution time was calculated after 100 cycles which corresponds to 100\,000 algorithm executions.
+We also measured the performance of \cref{alg:estimate_ml} using the same test setup as described in \cref{sec:corrected_raw_estimation_algorithm}.
 The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:avg_exec_time}.
-All these benchmarks where carried out on an Intel Core i5-2500K clocking at 3.3\,GHz.
-As can be seen, the average calculation time did never exceed 700\,ns.
+The average computation time shows a different behavior over the cardinality than those of the corrected raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}).
+As can be seen, the average calculation time is longer for most cardinalities, but nevertheless it never exceeds 700\,ns which is still fast enough for many applications.
 
@@
 \end{multline}
-For $\symX \geq 0$ $\symHelper'(\symX)\in[0,0.5]$. Hence, there exists a $\symBeta\in[0,0.5]$ for which
+Since $\symHelper'(\symX)\in[0,0.5]$ for $\symX \geq 0$, there exists a $\symBeta\in[0,0.5]$ for which
 \begin{multline}

### 2016-06-25 paper/paper.tex
some improvements

@@
 This paper presents new methods to estimate the cardinalities of multisets recorded by HyperLogLog sketches.
-A theoretically founded extension to the original estimate is presented that eliminates the bias for small and large cardinalities.
+A theoretically motivated extension to the original estimator is presented that eliminates the bias for small and large cardinalities.
 Based on the maximum likelihood principle another unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate.
-The maximum-likelihood method is also appropriate to improve cardinality estimates of set intersections compared to the inclusion-exclusion principle.
+The maximum likelihood method is also appropriate to improve cardinality estimates of set intersections compared to the inclusion-exclusion principle.
 The new methods are demonstrated and verified by extensive simulations.
@@
 
-Currently, the best algorithm that also works for distributed setups is the near-optimal HyperLogLog algorithm \cite{Flajolet2007} with space complexity $\Omega(\symError^{-2} \log\log\symCardinality +\log \symCardinality)$.
+Currently, the most memory efficient algorithm that also works for distributed setups is the near-optimal HyperLogLog algorithm \cite{Flajolet2007} with space complexity $\Omega(\symError^{-2} \log\log\symCardinality +\log \symCardinality)$.
 The originally proposed estimation method has some problems to guarantee the same estimation error over the entire range of cardinalities.
@@
 The HyperLogLog algorithm uses a sketching data structure that consists of $\symNumReg$ registers.
-For performance reasons the number of registers is chosen to be a power of 2: $\symNumReg = 2^\symPrecision$.
+For performance reasons the number of registers $\symNumReg$ is chosen to be a power of 2, $\symNumReg = 2^\symPrecision$.
 $\symPrecision$ is the precision that directly influences the relative error which scales like $1/\sqrt{\symNumReg}$.
-All registers start with initial value 0.
+All registers start with initial value zero.
 Each element insertion potentially increases the value of one of these registers.
@@
 
-For element insertion a hash value with bit size $\symPrecision+\symRegRange$ is required.
-The consumed number of bits also limits the maximum cardinality that can be tracked.
+The sum $\symPrecision+\symRegRange$ corresponds to the number of consumed hash value bits and defines the maximum cardinality that can be tracked.
 Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent and the estimation accuracy is drastically reduced.
@@
 Furthermore, HyperLogLog sketches with same parameters $(\symPrecision, \symRegRange)$ can be easily merged.
-The resulting register values are simply the register-wise maximum values.
-The final HyperLogLog sketch corresponds to the union of both sets that have been represented by the input sketches.
+The resulting register values are simply given by the register-wise maximum values.
+The final HyperLogLog sketch corresponds to the union of both sets that have been represented by the two input sketches.
 
@@
 In this case each register value is represented by a single bit.
-Hence, linear counting can be regarded as a special case of the HyperLogLog algorithm ($\symRegRange = 0$).
+Hence, linear counting can be regarded as a special case of the HyperLogLog algorithm for which $\symRegRange = 0$.
 
 \begin{algorithm}
-\caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers $\vec{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg$) which are initialized to 0 and which can represent any integer in range $[0, \symRegRange+1]$.}
+\caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers.
+All register values $\vec{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg$) are initially equal to 0 and remain in the range $[0, \symRegRange+1]$.}
 \label{alg:insert}
@@
 
-Under the assumption of an ideal hash function, the probability mass function for the register values $\vec{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
+Under the assumption of a strong universal hash function, the probability mass function for the register values $\vec{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
 \begin{equation}
@@
 In addition, this vector contains all the information about the HyperLogLog sketch that is required for cardinality estimation.
-The two HyperLogLog parameters can be obtained by $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$ and $\symRegRange=\dim(\vec{\symCount})-2$, respectively.
+The two HyperLogLog parameters can be obtained by $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$ and $\symRegRange=\dim\vec{\symCount}-2$, respectively.
 
@@
 The final factorization shows that all register values are independent and identically distributed under the Poisson model.
-The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is given by
+The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is defined by
 \begin{equation}
@@
 \end{equation}
-holds for all $\symPoissonRate\geq0$. The inverse Poisson transform gives
+holds for all $\symPoissonRate\geq0$. The unique solution of this equation is given by
 \begin{equation}
@@
 \end{equation}
-Therefore, the unbiased estimator $\symPoissonRateEstimate$ conditioned on $\symCardinality$ is also an unbiased estimator for $\symCardinality$.
-Hence, it makes sense to use $\symPoissonRateEstimate$ directly as estimate for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$.
+Consequently, the unbiased estimator $\symPoissonRateEstimate$ conditioned on $\symCardinality$ is also an unbiased estimator for $\symCardinality$ which motivates us to use $\symPoissonRateEstimate$ directly as estimator for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$.
 As simulation results will show later, the Poisson approximation works well over the entire cardinality range, even for estimators that are not exactly unbiased.
@@
 \label{sec:cardinality_estimation}
-The original cardinality estimate \cite{Flajolet2007} is based on the idea that approximately $\symNumReg  2^{\symRegVal_\symIndexJ}$ distinct element insertions are needed until the value of register $\symIndexJ$ reaches $\symRegVal_\symIndexJ$.
-Given that, a rough cardinality estimate can be obtained by calculating the average over the values $\lbrace \symNumReg 2^{\symRegVal_1},\ldots,\symNumReg2^{\symRegVal_\symNumReg}\rbrace$.
+The original cardinality estimator \cite{Flajolet2007} is based on the idea that approximately $\symNumReg  2^{\symRegVal-1}$ distinct element insertions are needed until the value of a register reaches $\symRegVal$.
+Given that, a rough cardinality estimate can be obtained by calculating the average over the values $\lbrace \symNumReg 2^{\symRegVal_1-1},\ldots,\symNumReg2^{\symRegVal_\symNumReg-1}\rbrace$.
 
@@
 =
+2\alpha_\symNumReg
+\operatorname{HM}
+\!\left(
+\symNumReg 2^{\symRegVal_1-1},\ldots,\symNumReg2^{\symRegVal_\symNumReg-1}
+\right)
+= 
 \frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
 = 
-\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}
-\end{equation}
-where $\symAlpha_\symNumReg$  is a correction factor that is fixed for a given number of registers $\symNumReg$ and which was derived to be \cite{Flajolet2007}
+\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}.
+\end{equation}
+Here $2\alpha_\symNumReg$ is a bias correction factor. $\alpha_\symNumReg$ for a given number of registers $\symNumReg$ was derived to be \cite{Flajolet2007}
 \begin{equation}
@@
 The chart is based on 10\,000 randomly generated HyperLogLog sketches.
+More details of the experimental setup will be explained later in \cref{sec:corrected_raw_estimation_error}.
 Obviously, the raw estimator is biased for small and large cardinalities and fails to return accurate estimates.
@@
 The drawback is that 5-bit registers are no longer sufficient to represent all possible values in the range $[0, \symRegRange+1]$ in case $\symRegRange \geq 30$.
-For HyperLogLog sketches with 6-bit registers in combination with 64-bit hash values $(\symPrecision+\symRegRange=64)$ as proposed in \cite{Heule2013}, it is unrealistic in practice to encounter cardinalities of order $2^{64}$, which would require some correction.
+For HyperLogLog sketches with 6-bit registers in combination with 64-bit hash values $(\symPrecision+\symRegRange=64)$ as proposed in \cite{Heule2013}, it is unrealistic in practice to encounter cardinalities of order $2^{64}$ for which the raw estimator would return incorrect results.
 
@@
 \end{equation}
-This distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$.
+For now we ignore that this distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$.
 The expectation of $2^{-\symRegVal}$ is given by
@@
 As we have already seen in \cref{fig:raw_estimate}, the raw estimator does not work very well for small or large cardinalities.
-The reason is that the distribution of register values \eqref{equ:register_value_distribution} is different from \eqref{equ:assumed_register_val_distribution} for which the raw estimator was derived.
+The reason is that the distribution of register values \eqref{equ:register_value_distribution} differs from \eqref{equ:assumed_register_val_distribution} for which the raw estimator was derived.
 Since a random variable $\symRegValVariate'$ that obeys \eqref{equ:assumed_register_val_distribution} can be transformed into random variable $\symRegValVariate$ that follows \eqref{equ:register_value_distribution} using the transformation $\symRegValVariate = \min\left(\max\left(\symRegValVariate', 0\right), \symRegRange+1\right)$, the register values of a HyperLogLog sketch can be seen as the result after applying this transformation to a sample that is distributed according to \eqref{equ:assumed_register_val_distribution}.
@@
 \end{equation}
-To get estimators for all other $\symRegVal$, we consider their expectations
+To get estimators for all other $\symRegVal\notin[1,\symRegRange]$, we consider their expectations
 \begin{equation}
@@
 :=
-\symNumReg\cdot
-\left(\symCount_0/\symNumReg\right)^{2^{-\symRegVal}}\cdot
+\symNumReg
+\left(\symCount_0/\symNumReg\right)^{2^{-\symRegVal}}
 \left(1-\left(\symCount_0/\symNumReg\right)^{2^{-\symRegVal}}\right)
@@
 :=
-\symNumReg\cdot
+\symNumReg
 \left(1-\symCount_{\symRegRange+1}/\symNumReg\right)^{2^{\symRegRange-\symRegVal}}
-\cdot
 \left(1-\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)^{2^{\symRegRange-\symRegVal}}\right)
@@
 \symX^{2^{-\symRegVal}}
-\cdot
 \left(1-\symX^{2^{-\symRegVal}}\right)
-\cdot
 2^{-\symRegVal}.
@@
 
-Using the identity $\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = \symAlpha_\infty\symPowerSeriesFunc(\log_2(\log(\symX^{-1})))/\log(\symX^{-1})$, we get for the linear counting case with $\symRegRange=0$
+Using the identity $\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = \symAlpha_\infty\symPowerSeriesFunc(\log_2(\log(1/\symX)))/\log(1/\symX)$, we get for the linear counting case with $\symRegRange=0$
 \begin{equation}
@@
 which is almost identical to the linear counting estimator used for the small range correction \eqref{equ:small_range_estimate}.
-Here we used the fact that $\symPowerSeriesFunc$ can be well approximated by 1.
+Here we used the fact that $\symPowerSeriesFunc$ can be well approximated by 1 (see \cref{sec:derivation_raw_estimator}).
 \subsection{Corrected raw estimation algorithm}
@@
 In order to evaluate the performance of the corrected raw estimation algorithm, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch.
-For different cardinalities we loaded the precalculated multiplicity vectors of 1000 randomly generated HyperLogLog sketches into main memory.
+For different cardinalities we loaded precalculated multiplicity vectors of 1000 randomly generated HyperLogLog sketches into main memory.
 The average computation time was determined by cycling over these multiplicity vectors and passing them as input to the algorithm.
-For each evaluated cardinality value the average execution time was calculated after 100 cycles which corresponds to 100\,000 algorithm executions.
+For each evaluated cardinality value the average execution time was calculated after 100 cycles which corresponds to 100\,000 algorithm executions for each cardinality value.
 The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:corrected_raw_avg_exec_time}.
@@
 The results show that the execution times are nearly constant for $\symRegRange=52$.
-Using the lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed, because $\symCount_{53}$ is equal to zero with high probability for realistic cardinalities.
+Using a lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed, because $\symCount_{53}$ is equal to zero with high probability for realistic cardinalities.
 
@@
 \section{Maximum likelihood estimation}
-We know from \cref{sec:depoissonization} that any unbiased estimator for the Poisson parameter is also an unbiased estimator for the cardinality.
+Moreover, we know from \cref{sec:depoissonization} that any unbiased estimator for the Poisson parameter is also an unbiased estimator for the cardinality.
 We know that under suitable regularity conditions of the probability mass function the maximum likelihood estimator is asymptotically efficient \cite{Casella2002}.
@@
 \end{equation}
-As a consequence, the maximum-likelihood estimate for the Poisson parameter is given by 
+As a consequence, the maximum likelihood estimate for the Poisson parameter is given by 
 \begin{equation}
@@
 
-\subsection{Maximum-likelihood estimation algorithm}
+\subsection{Maximum likelihood estimation algorithm}
 In order to get a fast cardinality estimation algorithm, it is crucial to minimize the evaluation costs for \eqref{equ:funcprime}.

### 2016-06-25 paper/paper.tex
some improvements

@@
 The results show that the execution times are nearly constant for $\symRegRange=52$.
-Using a lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed, because $\symCount_{53}$ is equal to zero with high probability for realistic cardinalities.
+Using a lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed due to the small probability of saturated registers ($\symCount_{53}>0$) for realistic cardinalities.
 
@@
 In case $\frac{\symX}{2^{\symRegVal'_\text{max}}} > 0.5$, the value $\frac{\symX}{2^{\symE+1}}$ is taken instead, where $\symE$ is the exponent of the floating point representation of $\symX$, $\symE = 1+\lfloor\log_2(\symX)\rfloor$.
-By definition, $\frac{\symX}{2^{\symE+1}}\leq 0.5$ which allows using the Taylor approximation.
+By definition, $\frac{\symX}{2^{\symE+1}}\leq 0.5$ which allows using the Taylor series approximation.
 $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is finally obtained after $\symE + 1 -\symRegVal'_\text{max}$ iterations using \eqref{equ:helper_recursion2}.
@@
 \begin{algorithm}
-\caption{Maximum likelihood cardinality estimation}
+\caption{Maximum likelihood cardinality estimation.}
 \ContinuedFloat
@@
 \begin{algorithm}
-\caption{Maximum likelihood cardinality estimation (continued)}
+\caption{Maximum likelihood cardinality estimation (continued).}
 \begin{algorithmic}
@@
 In order to investigate the estimation error for the maximum likelihood estimation algorithm, we investigated the same HyperLogLog configurations as in \cref{sec:corrected_raw_estimation_error} for the corrected raw estimation algorithm.
-\cref{fig:estimation_error_12_20}, \cref{fig:estimation_error_8_24}, \cref{fig:estimation_error_16_16},\cref{fig:estimation_error_22_10},\cref{fig:estimation_error_12_52}, and \cref{fig:estimation_error_12_14} show very similar results for various HyperLogLog parameters.
+\cref{fig:max_likelihood_estimation_error_12_20}, \cref{fig:max_likelihood_estimation_error_8_24}, \cref{fig:max_likelihood_estimation_error_16_16},\cref{fig:max_likelihood_estimation_error_22_10},\cref{fig:max_likelihood_estimation_error_12_52}, and \cref{fig:max_likelihood_estimation_error_12_14} show very similar results for various HyperLogLog parameters.
 
 What is different for the maximum likelihood estimation approach is a somewhat smaller median bias with less oscillations around zero for small cardinalities.
-Furthermore, contrary to the raw estimator which reveals a small oscillating bias for the mean (see \cref{fig:raw_corrected_estimation_error_22_10}), the maximum likelihood estimator seems to be completely unbiased (see \cref{fig:estimation_error_22_10}).
+Furthermore, contrary to the raw estimator which reveals a small oscillating bias for the mean (see \cref{fig:raw_corrected_estimation_error_22_10}), the maximum likelihood estimator seems to be completely unbiased (see \cref{fig:max_likelihood_estimation_error_22_10}).
 
@@
 \includesvg[width=1\textwidth]{max_likelihood_estimate_12_20}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
-\label{fig:estimation_error_12_20}
+\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
+\label{fig:max_likelihood_estimation_error_12_20}
 \end{figure}
@@
 \includesvg[width=1\textwidth]{max_likelihood_estimate_8_24}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
-\label{fig:estimation_error_8_24}
+\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
+\label{fig:max_likelihood_estimation_error_8_24}
 \end{figure}
@@
 \includesvg[width=1\textwidth]{max_likelihood_estimate_16_16}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
-\label{fig:estimation_error_16_16}
+\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
+\label{fig:max_likelihood_estimation_error_16_16}
 \end{figure}
@@
 \includesvg[width=1\textwidth]{max_likelihood_estimate_22_10}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
-\label{fig:estimation_error_22_10}
+\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
+\label{fig:max_likelihood_estimation_error_22_10}
 \end{figure}
@@
 \includesvg[width=1\textwidth]{max_likelihood_estimate_12_52}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
-\label{fig:estimation_error_12_52}
+\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
+\label{fig:max_likelihood_estimation_error_12_52}
 \end{figure}
@@
 \includesvg[width=1\textwidth]{max_likelihood_estimate_12_14}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
-\label{fig:estimation_error_12_14}
+\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
+\label{fig:max_likelihood_estimation_error_12_14}
 \end{figure}
@@
 The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:avg_exec_time}.
-The average computation time shows a different behavior over the cardinality than those of the corrected raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}).
-As can be seen, the average calculation time is longer for most cardinalities, but nevertheless it never exceeds 700\,ns which is still fast enough for many applications.
+The average computation time for the maximum likelihood algorithm shows a different behavior than for the corrected raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}).
+As can be seen, the average execution time is larger for most cardinalities, but nevertheless it never exceeds 700\,ns which is still fast enough for many applications.
 
@@
 \includesvg[width=1\textwidth]{max_likelihood_avg_exec_time}
-\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3\,GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
+\caption{Average execution time of the maximum likelihood estimation algorithm as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3\,GHz for HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
 \label{fig:avg_exec_time}

### 2016-06-25 paper/paper.tex
changed vector notation

@@
 \caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers.
-All register values $\vec{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg$) are initially equal to 0 and remain in the range $[0, \symRegRange+1]$.}
+All register values $\boldsymbol{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ are initially equal to 0 and remain in the range $[0, \symRegRange+1]$.}
 \label{alg:insert}
 \begin{algorithmic}
-\Procedure {InsertElement}{\symDataItem, $\vec{\symRegVal}$}
+\Procedure {InsertElement}{\symDataItem, $\boldsymbol{\symRegVal}$}
 \State $\langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision,\symBitRepB_1,\ldots,\symBitRepB_\symRegRange\rangle_2 \gets$ $(\symPrecision + \symRegRange)$-bit hash value of $\symDataItem$
@@
 
-Under the assumption of a strong universal hash function, the probability mass function for the register values $\vec{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
+Under the assumption of a strong universal hash function, the probability mass function for the register values $\boldsymbol{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
 \begin{equation}
 \label{equ:multinomialProbabilityMass}
-\symProbabilityMass(\vec{\symRegVal}\vert\symCardinality)
+\symProbabilityMass(\boldsymbol{\symRegVal}\vert\symCardinality)
 =
@@
 Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace = 0^{\symCount_0}1^{\symCount_1}\cdots\symRegRange^{\symCount_{\symRegRange}}(\symRegRange+1)^{\symCount_{\symRegRange+1}}$ where $\symCount_\symIndexJ$ is the multiplicity of value $\symIndexJ$.
-As a consequence, the multiplicity vector $\vec{\symCount} := (\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is also a sufficient statistic for the cardinality.
+As a consequence, the multiplicity vector $\boldsymbol{\symCount} := (\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is also a sufficient statistic for the cardinality.
 In addition, this vector contains all the information about the HyperLogLog sketch that is required for cardinality estimation.
-The two HyperLogLog parameters can be obtained by $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$ and $\symRegRange=\dim\vec{\symCount}-2$, respectively.
+The two HyperLogLog parameters can be obtained by $\symPrecision = \log_2 \left\Vert\boldsymbol{\symCount}\right\Vert_1$ and $\symRegRange=\dim\boldsymbol{\symCount}-2$, respectively.
 
@@
 \begin{align}
-\symProbabilityMass(\vec{\symRegVal}\vert\symPoissonRate) 
+\symProbabilityMass(\boldsymbol{\symRegVal}\vert\symPoissonRate) 
 &= 
-\sum_{\symCardinality=0}^\infty \symProbabilityMass(\vec{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
+\sum_{\symCardinality=0}^\infty \symProbabilityMass(\boldsymbol{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
 \label{equ:poissonization}
@@
 \label{sec:depoissonization}
-Due to the simpler probability mass function, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\vec{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ rather than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}.
-Here $\vec{\symRegValVariate}$ denotes the vector of all observed register values.
-Each component of $\vec{\symRegValVariate}$ is distributed according to \eqref{equ:multinomialProbabilityMass}.
+Due to the simpler probability mass function, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\boldsymbol{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ rather than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}.
+Here $\boldsymbol{\symRegValVariate}$ denotes the vector of all observed register values.
+Each component of $\boldsymbol{\symRegValVariate}$ is distributed according to \eqref{equ:multinomialProbabilityMass}.
 Depoissonization \cite{Jacquet1998} finally allows to translate the estimates back to the fixed-size model.
@@
 \begin{algorithmic}
-\Function {EstimateCardinality}{$\vec{\symRegVal}$}
-\State $\symNumReg \gets \dim \vec{\symRegVal}$
+\Function {EstimateCardinality}{$\boldsymbol{\symRegVal}$}
+\State $\symNumReg \gets \dim \boldsymbol{\symRegVal}$
 \State $\symCardinalityRawEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegVal_\symIndexJ}\right)^{-1}$
@@
 %\begin{algorithm}
-%\caption{Calculate the sufficient statistic $\vec{\symCount}=(\symCount_0, \ldots,\symCount_{\symRegRange+1})$ for given register values $\vec{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg)$.}
+%\caption{Calculate the sufficient statistic $\boldsymbol{\symCount}=(\symCount_0, \ldots,\symCount_{\symRegRange+1})$ for given register values $\boldsymbol{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg)$.}
 %\begin{algorithmic}
-%\Function {CountRegisterValues}{$\vec{\symRegVal}$}
-%\State $\vec{\symCount}\gets\vec{0}$
+%\Function {CountRegisterValues}{$\boldsymbol{\symRegVal}$}
+%\State $\boldsymbol{\symCount}\gets\boldsymbol{0}$
 %\ForAll{$\symRegVal'\in\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$}
@@
 %\EndFor
-%\State\Return $\vec{\symCount}$
+%\State\Return $\boldsymbol{\symCount}$
 %\EndFunction
@@
 \subsection{Corrections to the raw estimator}
-Given a HyperLogLog sketch with multiplicity vector $\vec{\symCount}=\left(\symCount_0,\ldots,\symCount_{\symRegRange+1}\right)$, we try to find estimates $\hat{\symCount}'_\symRegVal$ for $\symCount'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and replace them in the estimation formula \eqref{equ:raw_estimate_with_multiplicities}.
+Given a HyperLogLog sketch with multiplicity vector $\boldsymbol{\symCount}=\left(\symCount_0,\ldots,\symCount_{\symRegRange+1}\right)$, we try to find estimates $\hat{\symCount}'_\symRegVal$ for $\symCount'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and replace them in the estimation formula \eqref{equ:raw_estimate_with_multiplicities}.
 For $\symRegVal\in[1,\symRegRange]$ where $\symCount_\symRegVal = \symCount'_\symRegVal$ we can use the trivial estimators
@@
 \begin{algorithmic}
-\Function {EstimateCardinality}{$\vec{\symCount}$}
-\State $\symNumReg \gets \left\Vert\vec{\symCount}\right\Vert_1$
+\Function {EstimateCardinality}{$\boldsymbol{\symCount}$}
+\State $\symNumReg \gets \left\Vert\boldsymbol{\symCount}\right\Vert_1$
 \State $\symZ\gets \cdot \symNumReg\cdot\symLargeCorrectionFunc(1 - \symCount_{\symRegRange+1}/\symNumReg)$
@@
 \label{equ:log_likelihood_single}
-\log \mathcal{\symLikelihood}(\symPoissonRate\vert\vec{\symRegVal}) = 
+\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegVal}) = 
 -\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCount_\symRegVal}{2^\symRegVal}+ 
@@
 \begin{equation}
-\frac{d}{d\symPoissonRate}\log \mathcal{\symLikelihood}(\symPoissonRate\vert\vec{\symRegVal}) 
+\frac{d}{d\symPoissonRate}\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegVal}) 
 =
@@
 \begin{algorithmic}
-\Function {EstimateCardinality}{$\vec{\symCount}$}
-\State $\symRegRange\gets\dim(\vec{\symCount})-2$
+\Function {EstimateCardinality}{$\boldsymbol{\symCount}$}
+\State $\symRegRange\gets\dim(\boldsymbol{\symCount})-2$
 \State $\symRegVal_\text{min} \gets \min\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$
@@
 Assume two different sets $\symSetS_1$ and $\symSetS_2$ that have been recorded by two HyperLogLog sketches with same parameters $(\symPrecision,\symRegRange)$.
-Given the corresponding register values $\vec{\symRegVal}_1$ and
-$\vec{\symRegVal}_2$, we attempt to estimate the cardinalities of the pair-wise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$.
-Motivated by the good results we have obtained for cardinality estimation of a single HyperLogLog sketch, we want to get these estimates using the maximum likelihood method applied to the joint probability distribution of $\vec{\symRegVal}_1$ and $\vec{\symRegVal}_2$.
+Given the corresponding register values $\boldsymbol{\symRegVal}_1$ and
+$\boldsymbol{\symRegVal}_2$, we attempt to estimate the cardinalities of the pair-wise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$.
+Motivated by the good results we have obtained for cardinality estimation of a single HyperLogLog sketch, we want to get these estimates using the maximum likelihood method applied to the joint probability distribution of $\boldsymbol{\symRegVal}_1$ and $\boldsymbol{\symRegVal}_2$.
 
@@
 \begin{equation}
-\symProbabilityMass(\vec{\symRegVal}_1,\vec{\symRegVal}_2)
+\symProbabilityMass(\boldsymbol{\symRegVal}_1,\boldsymbol{\symRegVal}_2)
 =
@@
 \end{equation}
-As in \cref{sec:cardinality_estimation} where we found that the multiplicity vector $\vec{\symCount}$ of a sketch is a sufficient statistic for the cardinality, the multiplicity matrix of two sketches is a sufficient statistic for the cardinalities of $\symSetX$, $\symSetA$, and $\symSetB$.
+As in \cref{sec:cardinality_estimation} where we found that the multiplicity vector $\boldsymbol{\symCount}$ of a sketch is a sufficient statistic for the cardinality, the multiplicity matrix of two sketches is a sufficient statistic for the cardinalities of $\symSetX$, $\symSetA$, and $\symSetB$.
 
@@
 \vert
-\vec{\symRegVal}_1,
-\vec{\symRegVal}_2
+\boldsymbol{\symRegVal}_1,
+\boldsymbol{\symRegVal}_2
 )

### 2016-06-26 paper/paper.tex
consistent notation of random variates

@@
 \symDefine{\symCount}{c}
+\symDefine{\symCountVariate}{C}
 \symDefine{\symAlpha}{\alpha}
@@
 \symDefine{\symRegProbability}{\gamma}
-\symDefine{\symNu}{\nu}
-\symDefine{\symKappa}{\kappa}
 \symDefine{\symLikelihood}{\mathcal{L}}
@@
 \caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers.
-All register values $\boldsymbol{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ are initially equal to 0 and remain in the range $[0, \symRegRange+1]$.}
+All register values $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ are initially equal to 0 and remain in the range $[0, \symRegRange+1]$.}
 \label{alg:insert}
 \begin{algorithmic}
-\Procedure {InsertElement}{\symDataItem, $\boldsymbol{\symRegVal}$}
+\Procedure {InsertElement}{\symDataItem, $\boldsymbol{\symRegValVariate}$}
 \State $\langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision,\symBitRepB_1,\ldots,\symBitRepB_\symRegRange\rangle_2 \gets$ $(\symPrecision + \symRegRange)$-bit hash value of $\symDataItem$
 \State $\symIndexJ \gets 1+ \langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision\rangle_2$
-\State $\symRegVal' = \min(\{\symS\mid \symS \in [1, \symRegRange]  \wedge  \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
-\State $\symRegVal_\symIndexJ \gets\max(\symRegVal_\symIndexJ, \symRegVal')$
+\State $\symRegValVariate' = \min(\{\symS\mid \symS \in [1, \symRegRange]  \wedge  \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
+\State $\symRegValVariate_\symIndexJ \gets\max(\symRegValVariate_\symIndexJ, \symRegValVariate')$
 \EndProcedure
@@
 
-Under the assumption of a strong universal hash function, the probability mass function for the register values $\boldsymbol{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
+Under the assumption of a strong universal hash function, the probability that the register values $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ are equal to $\boldsymbol{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ is given by the corresponding probability mass function
 \begin{equation}
@@
 \sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots\symCardinality_\symNumReg}
-\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexJ=1}^\symNumReg \symRegProbability_{\symCardinality_\symIndexJ\symRegVal_\symIndexJ},
+\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexJ=1}^\symNumReg \symRegProbability(\symRegVal_\symIndexJ\vert\symCardinality_\symIndexJ)
 \end{equation}
@@
 The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities.
-$\symRegProbability_{\symNu\symKappa}$ is the probability that a register is equal to $\symKappa$ after inserting $\symNu$ distinct elements
-\begin{equation}
-\symRegProbability_{\symNu\symKappa} 
+$\symRegProbability(\symRegVal\vert\symCardinality)$ is the probability that a register has value $\symRegVal$ after inserting $\symCardinality$ distinct elements
+\begin{equation}
+\symRegProbability(\symRegVal\vert\symCardinality) 
 :=
 \begin{cases}
-1 & \symNu=0 \wedge \symKappa = 0\\
-0 & \symNu=0 \wedge 1\leq\symKappa\leq\symRegRange+1\\
-0 & \symNu\geq1 \wedge \symKappa = 0\\
-\left(1-\frac{1}{2^\symKappa}\right)^\symNu - \left(1-\frac{1}{2^{\symKappa-1}}\right)^\symNu & \symNu \geq 1 \wedge 1\leq\symKappa\leq\symRegRange\\
-1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symNu & \symNu\geq 1 \wedge \symKappa = \symRegRange +1 \\
+1 & \symCardinality=0 \wedge \symRegVal = 0\\
+0 & \symCardinality=0 \wedge 1\leq\symRegVal\leq\symRegRange+1\\
+0 & \symCardinality\geq1 \wedge \symRegVal = 0\\
+\left(1-\frac{1}{2^\symRegVal}\right)^\symCardinality - \left(1-\frac{1}{2^{\symRegVal-1}}\right)^\symCardinality & \symCardinality \geq 1 \wedge 1\leq\symRegVal\leq\symRegRange\\
+1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symCardinality & \symCardinality\geq 1 \wedge \symRegVal = \symRegRange +1 \\
 \end{cases}
@@
 
-The order of register values $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ is not important for the estimation of the cardinality.
-More formally, the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$ is a sufficient statistic for $\symCardinality$.
-Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace = 0^{\symCount_0}1^{\symCount_1}\cdots\symRegRange^{\symCount_{\symRegRange}}(\symRegRange+1)^{\symCount_{\symRegRange+1}}$ where $\symCount_\symIndexJ$ is the multiplicity of value $\symIndexJ$.
-As a consequence, the multiplicity vector $\boldsymbol{\symCount} := (\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is also a sufficient statistic for the cardinality.
+The order of register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ is not important for the estimation of the cardinality.
+More formally, the multiset $\lbrace\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg\rbrace$ is a sufficient statistic for $\symCardinality$.
+Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg\rbrace = 0^{\symCountVariate_0}1^{\symCountVariate_1}\cdots\symRegRange^{\symCountVariate_{\symRegRange}}(\symRegRange+1)^{\symCountVariate_{\symRegRange+1}}$ where $\symCountVariate_\symIndexJ$ is the multiplicity of value $\symIndexJ$.
+As a consequence, the multiplicity vector $\boldsymbol{\symCountVariate} := (\symCountVariate_0,\ldots,\symCountVariate_{\symRegRange+1})$ is also a sufficient statistic for the cardinality.
 In addition, this vector contains all the information about the HyperLogLog sketch that is required for cardinality estimation.
-The two HyperLogLog parameters can be obtained by $\symPrecision = \log_2 \left\Vert\boldsymbol{\symCount}\right\Vert_1$ and $\symRegRange=\dim\boldsymbol{\symCount}-2$, respectively.
+The two HyperLogLog parameters can be obtained by $\symPrecision = \log_2 \left\Vert\boldsymbol{\symCountVariate}\right\Vert_1$ and $\symRegRange=\dim\boldsymbol{\symCountVariate}-2$, respectively.
 
@@
 \prod_{\symIndexJ=1}^\symNumReg
-\symRegProbability_
-{\symCardinality_\symIndexJ\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexJ}}{\symCardinality_\symIndexJ!\symNumReg^{\symCardinality_\symIndexJ}}
+\symRegProbability(\symCardinality_\symIndexJ\vert\symRegVal_\symIndexJ)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexJ}}{\symCardinality_\symIndexJ!\symNumReg^{\symCardinality_\symIndexJ}}
 \nonumber\\
 &= 
-\prod_{\symIndexJ=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability_
-{\symCardinality\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
+\prod_{\symIndexJ=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability(\symCardinality_\symIndexJ\vert\symRegVal_\symIndexJ)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
 \nonumber\\
@@
 \sum_{\symCardinality=0}^\infty
-\symRegProbability_
-{\symCardinality\symRegVal}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
+\symRegProbability(\symCardinality_\symIndexJ\vert\symRegVal_\symIndexJ)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
 \right]^{\symCount_\symRegVal}
 \nonumber\\
+\label{equ:poisson_pmf}
 &=
 e^{-\symCount_0\frac{\symPoissonRate}{\symNumReg}}
-\cdot
 \left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\symCount_\symRegVal}\right)
-\cdot
 \left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\symCount_{\symRegRange+1}}.
 \end{align}
+Here $\boldsymbol{\symCount}=(\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is the multiplicity vector for $\boldsymbol{\symRegVal}$.
+$\symCount_\symRegVal$ denotes the multiplicity of value $\symRegVal$  in the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$.
 The final factorization shows that all register values are independent and identically distributed under the Poisson model.
@@
 1 & \symRegVal > \symRegRange
-\end{cases}
+\end{cases}.
 \end{equation}
@@
 The original cardinality estimator \cite{Flajolet2007} is based on the idea that approximately $\symNumReg  2^{\symRegVal-1}$ distinct element insertions are needed until the value of a register reaches $\symRegVal$.
-Given that, a rough cardinality estimate can be obtained by calculating the average over the values $\lbrace \symNumReg 2^{\symRegVal_1-1},\ldots,\symNumReg2^{\symRegVal_\symNumReg-1}\rbrace$.
+Given that, a rough cardinality estimate can be obtained by calculating the average over the values $\lbrace \symNumReg 2^{\symRegValVariate_1-1},\ldots,\symNumReg2^{\symRegValVariate_\symNumReg-1}\rbrace$.
 
@@
 \!\left(
-\symNumReg 2^{\symRegVal_1-1},\ldots,\symNumReg2^{\symRegVal_\symNumReg-1}
-\right)
-= 
-\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
-= 
-\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}.
+\symNumReg 2^{\symRegValVariate_1-1},\ldots,\symNumReg2^{\symRegValVariate_\symNumReg-1}
+\right)
+= 
+\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegValVariate_\symIndexJ}}
+= 
+\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCountVariate_\symRegVal 2^{-\symRegVal}}.
 \end{equation}
@@
 As mentioned in \cref{sec:data_element_insertion}, a HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be turned into a $(\symPrecision, 0)$-HyperLogLog sketch.
-Since $\symRegRange=0$ corresponds to linear counting and the transformed HyperLogLog sketch corresponds to a bitset with $\symCount_0$ zeros, the linear counting cardinality estimator \cite{Whang1990} can be applied here
+Since $\symRegRange=0$ corresponds to linear counting and the transformed HyperLogLog sketch corresponds to a bitset with $\symCountVariate_0$ zeros, the linear counting cardinality estimator \cite{Whang1990} can be applied here
 \begin{equation}
 \label{equ:small_range_estimate}
-\symCardinalityEstimate_\text{small} = \symNumReg \log(\symNumReg/\symCount_0).
+\symCardinalityEstimate_\text{small} = \symNumReg \log(\symNumReg/\symCountVariate_0).
 \end{equation}
@@
 \begin{algorithmic}
-\Function {EstimateCardinality}{$\boldsymbol{\symRegVal}$}
-\State $\symNumReg \gets \dim \boldsymbol{\symRegVal}$
-\State $\symCardinalityRawEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegVal_\symIndexJ}\right)^{-1}$
+\Function {EstimateCardinality}{$\boldsymbol{\symRegValVariate}$}
+\State $\symNumReg \gets \dim \boldsymbol{\symRegValVariate}$
+\State $\symCardinalityRawEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegValVariate_\symIndexJ}\right)^{-1}$
 \Comment raw estimate \eqref{equ:intermediate_estimate}
 \If{$\symCardinalityRawEstimate\leq \frac{5}{2}\symNumReg$}
-\State $\symCount_0 = \left|\{\symIndexJ |\symRegVal_\symIndexJ=0\}\right|$
-\If{$\symCount_0\neq0$}
-\State \Return $\symNumReg \log(\symNumReg / \symCount_0)$ \Comment small range correction \eqref{equ:small_range_estimate}
+\State $\symCountVariate_0 = \left|\{\symIndexJ |\symRegValVariate_\symIndexJ=0\}\right|$
+\If{$\symCountVariate_0\neq0$}
+\State \Return $\symNumReg \log(\symNumReg / \symCountVariate_0)$ \Comment small range correction \eqref{equ:small_range_estimate}
 \Else
@@
 For now we ignore that this distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$.
-The expectation of $2^{-\symRegVal}$ is given by
+The expectation of $2^{-\symRegValVariate}$ is given by
 \begin{multline}
 \label{equ:expectation_power_two}
-\symExpectation(2^{-\symRegVal})
+\symExpectation(2^{-\symRegValVariate})
 =
@@
 
-Let $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ be a sample taken from \eqref{equ:assumed_register_val_distribution}. For large sample sizes $\symNumReg\rightarrow \infty$ we have asymptotically
+Let $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ be a sample taken from \eqref{equ:assumed_register_val_distribution}.
+For large sample sizes $\symNumReg\rightarrow \infty$ we have asymptotically
 \begin{equation}
 \symExpectation\!\left(
-\frac{1}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
+\frac{1}{2^{-\symRegValVariate_1}+\ldots+2^{-\symRegValVariate_\symNumReg}}
 \right)
@@
 \frac{1}{\symExpectation
-\!\left(2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}\right)}
-=
-\frac{1}{\symNumReg \symExpectation\!\left(2^{-\symRegVal}\right)}.
+\!\left(2^{-\symRegValVariate_1}+\ldots+2^{-\symRegValVariate_\symNumReg}\right)}
+=
+\frac{1}{\symNumReg \symExpectation\!\left(2^{-\symRegValVariate}\right)}.
 \end{equation}
@@
 \symExpectation\!\left(
-\frac{\symAlpha_\infty\,\symNumReg^2\,\symPowerSeriesFunc\!\left(\log_2\!\left( \symPoissonRate/\symNumReg\right)\right)}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
+\frac{\symAlpha_\infty\,\symNumReg^2\,\symPowerSeriesFunc\!\left(\log_2\!\left( \symPoissonRate/\symNumReg\right)\right)}{2^{-\symRegValVariate_1}+\ldots+2^{-\symRegValVariate_\symNumReg}}
 \right)
@@
 = 
-\frac{\symAlpha_\infty\,\symNumReg^2}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
+\frac{\symAlpha_\infty\,\symNumReg^2}{2^{-\symRegValVariate_1}+\ldots+2^{-\symRegValVariate_\symNumReg}}
 \end{equation}
@@
 = 
-\frac{\symAlpha_\infty\,\symNumReg^2}{\sum_{\symRegVal=-\infty}^\infty \symCount'_\symRegVal 2^{-\symRegVal}},
-\end{equation}
-if $\symCount'_\symRegVal$ is the multiplicity of value $\symRegVal$ in the sample $\left\lbrace \symRegVal_1,\ldots,\symRegVal_\symNumReg\right\rbrace$. 
-By definition, we have $\sum_{\symRegVal=-\infty}^\infty \symCount'_\symRegVal = \symNumReg$.
+\frac{\symAlpha_\infty\,\symNumReg^2}{\sum_{\symRegVal=-\infty}^\infty \symCountVariate'_\symRegVal 2^{-\symRegVal}},
+\end{equation}
+if $\symCountVariate'_\symRegVal$ is the multiplicity of value $\symRegVal$ in the sample $\left\lbrace \symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg\right\rbrace$. 
+By definition, we have $\sum_{\symRegVal=-\infty}^\infty \symCountVariate'_\symRegVal = \symNumReg$.
 
@@
 \begin{aligned}
-\symCount_0 &= \textstyle\sum_{\symRegVal = -\infty}^{0} \symCount'_\symRegVal, \\
-\symCount_\symRegVal &=  \symCount'_\symRegVal, \quad 1\leq\symRegVal\leq\symRegRange,\\
-\symCount_{\symRegRange+1} &= \textstyle\sum_{\symRegVal = \symRegRange + 1}^{\infty} \symCount'_\symRegVal.
+\symCountVariate_0 &= \textstyle\sum_{\symRegVal = -\infty}^{0} \symCountVariate'_\symRegVal, \\
+\symCountVariate_\symRegVal &=  \symCountVariate'_\symRegVal, \quad 1\leq\symRegVal\leq\symRegRange,\\
+\symCountVariate_{\symRegRange+1} &= \textstyle\sum_{\symRegVal = \symRegRange + 1}^{\infty} \symCountVariate'_\symRegVal.
 \end{aligned}
 \end{equation}
-Obviously, as long as $\symCount_0\ll\symNumReg$ and  $\symCount_{\symRegRange+1}\ll\symNumReg$ the set of register values is similar to the originating sample.
-In case $\symCount_0=0$ and $\symCount_{\symRegRange+1}=0$, which implies $\symCount'_\symRegVal=0$ for $\symRegVal\leq0$ and $\symRegVal\geq\symRegRange+1$, we can even be sure that the register values are identical to the originating sample.
-This explains why the raw estimator only works well in the intermediate cardinality range for which $\symCount_0$ and $\symCount_{\symRegRange+1}$ are negligible.
+Obviously, as long as $\symCountVariate_0\ll\symNumReg$ and  $\symCountVariate_{\symRegRange+1}\ll\symNumReg$ the set of register values is similar to the originating sample.
+In case $\symCountVariate_0=0$ and $\symCountVariate_{\symRegRange+1}=0$, which implies $\symCountVariate'_\symRegVal=0$ for $\symRegVal\leq0$ and $\symRegVal\geq\symRegRange+1$, we can even be sure that the register values are identical to the originating sample.
+This explains why the raw estimator only works well in the intermediate cardinality range for which the number of registers with values equal to 0 or $\symRegRange+1$ is negligible.
 
 \subsection{Corrections to the raw estimator}
-Given a HyperLogLog sketch with multiplicity vector $\boldsymbol{\symCount}=\left(\symCount_0,\ldots,\symCount_{\symRegRange+1}\right)$, we try to find estimates $\hat{\symCount}'_\symRegVal$ for $\symCount'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and replace them in the estimation formula \eqref{equ:raw_estimate_with_multiplicities}.
-For $\symRegVal\in[1,\symRegRange]$ where $\symCount_\symRegVal = \symCount'_\symRegVal$ we can use the trivial estimators
-\begin{equation}
-\hat{\symCount}'_\symRegVal := \symCount_\symRegVal,\quad 1\leq\symRegVal\leq\symRegRange.
+Given a HyperLogLog sketch with multiplicity vector $\boldsymbol{\symCountVariate}=\left(\symCountVariate_0,\ldots,\symCountVariate_{\symRegRange+1}\right)$, we try to find estimates $\hat{\symCount}'_\symRegVal$ for $\symCountVariate'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and replace them in the estimation formula \eqref{equ:raw_estimate_with_multiplicities}.
+For $\symRegVal\in[1,\symRegRange]$ where $\symCountVariate_\symRegVal = \symCountVariate'_\symRegVal$ we can use the trivial estimators
+\begin{equation}
+\hat{\symCount}'_\symRegVal := \symCountVariate_\symRegVal,\quad 1\leq\symRegVal\leq\symRegRange.
 \end{equation}
@@
 \begin{equation}
-\symExpectation(\symCount'_\symRegVal)
+\symExpectation(\symCountVariate'_\symRegVal)
 =
@@
 \end{equation}
-From \eqref{equ:register_value_distribution} we know that $\symExpectation\!\left(\symCount_0/\symNumReg\right)=
-e^{-\frac{\symPoissonRate}{\symNumReg}}$ and $\symExpectation\!\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)=
+From \eqref{equ:register_value_distribution} we know that $\symExpectation\!\left(\symCountVariate_0/\symNumReg\right)=
+e^{-\frac{\symPoissonRate}{\symNumReg}}$ and $\symExpectation\!\left(1-\symCountVariate_{\symRegRange+1}/\symNumReg\right)=
 e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}$, and therefore, we can also write
 \begin{equation}
-\symExpectation(\symCount'_\symRegVal)
+\symExpectation(\symCountVariate'_\symRegVal)
 =
 \symNumReg
-\left(\symExpectation\!\left(\symCount_0/\symNumReg\right)\right)^{2^{-\symRegVal}}
-\left(1-\left(\symExpectation\!\left(\symCount_0/\symNumReg\right)\right)^{2^{-\symRegVal}}\right)
+\left(\symExpectation\!\left(\symCountVariate_0/\symNumReg\right)\right)^{2^{-\symRegVal}}
+\left(1-\left(\symExpectation\!\left(\symCountVariate_0/\symNumReg\right)\right)^{2^{-\symRegVal}}\right)
 \end{equation}
@@
 \begin{equation}
-\symExpectation(\symCount'_\symRegVal)
+\symExpectation(\symCountVariate'_\symRegVal)
 =
 \symNumReg
-\left(\symExpectation\!\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)\right)^{2^{\symRegRange-\symRegVal}}
-\left(1-\left(\symExpectation\!\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)\right)^{2^{\symRegRange-\symRegVal}}\right),
+\left(\symExpectation\!\left(1-\symCountVariate_{\symRegRange+1}/\symNumReg\right)\right)^{2^{\symRegRange-\symRegVal}}
+\left(1-\left(\symExpectation\!\left(1-\symCountVariate_{\symRegRange+1}/\symNumReg\right)\right)^{2^{\symRegRange-\symRegVal}}\right),
 \end{equation}
@@
 \symNumReg
-\left(\symCount_0/\symNumReg\right)^{2^{-\symRegVal}}
-\left(1-\left(\symCount_0/\symNumReg\right)^{2^{-\symRegVal}}\right)
+\left(\symCountVariate_0/\symNumReg\right)^{2^{-\symRegVal}}
+\left(1-\left(\symCountVariate_0/\symNumReg\right)^{2^{-\symRegVal}}\right)
 \end{equation}
@@
 \symNumReg
-\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)^{2^{\symRegRange-\symRegVal}}
-\left(1-\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)^{2^{\symRegRange-\symRegVal}}\right)
+\left(1-\symCountVariate_{\symRegRange+1}/\symNumReg\right)^{2^{\symRegRange-\symRegVal}}
+\left(1-\left(1-\symCountVariate_{\symRegRange+1}/\symNumReg\right)^{2^{\symRegRange-\symRegVal}}\right)
 \end{equation}
@@
 {
-\symNumReg\,\symSmallCorrectionFunc(\symCount_0/\symNumReg) + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symNumReg\, \symLargeCorrectionFunc(1-\symCount_{\symRegRange+1}/\symNumReg) 2^{-\symRegRange}
+\symNumReg\,\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg) + \sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal 2^{-\symRegVal} + \symNumReg\, \symLargeCorrectionFunc(1-\symCountVariate_{\symRegRange+1}/\symNumReg) 2^{-\symRegRange}
 }
@@
 which we call the corrected raw estimator.
-Here $\symNumReg\,\symSmallCorrectionFunc(\symCount_0/\symNumReg)$ and $2\symNumReg\,\symLargeCorrectionFunc(1-\symCount_{\symRegRange+1}/\symNumReg)$ are replacements for  $\symCount_0$ and $\symCount_{\symRegRange+1}$ in the raw estimator \eqref{equ:intermediate_estimate}, respectively.
+Here $\symNumReg\,\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$ and $2\symNumReg\,\symLargeCorrectionFunc(1-\symCountVariate_{\symRegRange+1}/\symNumReg)$ are replacements for  $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$ in the raw estimator \eqref{equ:intermediate_estimate}, respectively.
 The functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are defined as
@@
 \frac{\symAlpha_\infty\symNumReg}{
-\symSmallCorrectionFunc(\symCount_0/\symNumReg)
-+
-\symLargeCorrectionFunc(\symCount_0/\symNumReg)
-}
-=
-\frac{\symNumReg\log\!\left(\symNumReg/\symCount_0\right)}{\symPowerSeriesFunc\!\left(\log_2\!\left(\log\!\left(\symNumReg/\symCount_0\right)\right)\right)}
+\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)
++
+\symLargeCorrectionFunc(\symCountVariate_0/\symNumReg)
+}
+=
+\frac{\symNumReg\log\!\left(\symNumReg/\symCountVariate_0\right)}{\symPowerSeriesFunc\!\left(\log_2\!\left(\log\!\left(\symNumReg/\symCountVariate_0\right)\right)\right)}
 \approx
-\symNumReg\log\!\left(\symNumReg/\symCount_0\right),
+\symNumReg\log\!\left(\symNumReg/\symCountVariate_0\right),
 \end{equation}
@@
 The values of functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ can be either calculated on-demand or pre-calculated.
-The possible value range for $\symCount_0$ and $\symCount_{\symRegRange+1}$ is $[0, \symNumReg]$.
+The possible value range for $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$ is $[0, \symNumReg]$.
 Therefore, if performance matters, it is possible to precalculate the function values for all possible arguments and keep them in lookup tables of size $\symNumReg+1$.
@@
 \begin{algorithmic}
-\Function {EstimateCardinality}{$\boldsymbol{\symCount}$}
-\State $\symNumReg \gets \left\Vert\boldsymbol{\symCount}\right\Vert_1$
-\State $\symZ\gets \cdot \symNumReg\cdot\symLargeCorrectionFunc(1 - \symCount_{\symRegRange+1}/\symNumReg)$
-\Comment{\parbox[t]{.5\linewidth}{alternatively, take $\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCount_{\symRegRange+1}/\symNumReg)$ from precalculated lookup table}}
+\Function {EstimateCardinality}{$\boldsymbol{\symCountVariate}$}
+\State $\symNumReg \gets \left\Vert\boldsymbol{\symCountVariate}\right\Vert_1$
+\State $\symZ\gets \cdot \symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$
+\Comment{\parbox[t]{.5\linewidth}{alternatively, take $\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$ from precalculated lookup table}}
 \For{$\symRegVal\gets \symRegRange, 1$}
-\State $\symZ\gets0.5\cdot\left(\symZ + \symCount_\symRegVal\right)$
+\State $\symZ\gets0.5\cdot\left(\symZ + \symCountVariate_\symRegVal\right)$
 \EndFor
-\State $\symZ\gets\symZ+\symNumReg\cdot\symSmallCorrectionFunc(\symCount_0/\symNumReg)$
-\Comment{\parbox[t]{.5\linewidth}{alternatively, take $\symNumReg\cdot\symSmallCorrectionFunc(\symCount_0/\symNumReg)$ from precalculated lookup table}}
+\State $\symZ\gets\symZ+\symNumReg\cdot\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$
+\Comment{\parbox[t]{.5\linewidth}{alternatively, take $\symNumReg\cdot\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$ from precalculated lookup table}}
 \State \Return$\symNumReg^2 / \left(2\log(2)\symZ\right)$
@@
 The results show that the execution times are nearly constant for $\symRegRange=52$.
-Using a lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed due to the small probability of saturated registers ($\symCount_{53}>0$) for realistic cardinalities.
+Using a lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed due to the small probability of saturated registers ($\symCountVariate_{53}>0$) for realistic cardinalities.
 
@@
 
-Under the Poisson model the log-likelihood and its derivative are given by
+Using the probability mass function of the Poisson model \eqref{equ:poisson_pmf} the log-likelihood and its derivative are given by
 \begin{equation}
 \label{equ:log_likelihood_single}
-\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegVal}) = 
--\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCount_\symRegVal}{2^\symRegVal}+ 
-\sum_{\symRegVal=1}^{\symRegRange}\symCount_\symRegVal \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)
-+
-\symCount_{\symRegRange+1} \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegRange}}}\right)
+\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegValVariate}) = 
+-\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}+ 
+\sum_{\symRegVal=1}^{\symRegRange}\symCountVariate_\symRegVal \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)
++
+\symCountVariate_{\symRegRange+1} \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegRange}}}\right)
 \end{equation}
@@
 \begin{equation}
-\frac{d}{d\symPoissonRate}\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegVal}) 
+\frac{d}{d\symPoissonRate}\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegValVariate}) 
 =
 -\frac{1}{\symPoissonRate}\left(
-\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
-\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}}
-+\symCount_{\symRegRange+1}\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}}
+\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
+\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}}
++\symCountVariate_{\symRegRange+1}\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}}
 \right).
@@
 :=
-\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
-\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\frac{\frac{\symX}{2^\symRegVal}}{1-e^{\frac{\symX}{2^\symRegVal}}}
-+\symCount_{\symRegRange+1}\frac{\frac{\symX}{2^\symRegRange}}{1-e^{\frac{\symX}{2^\symRegRange}}}.
+\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
+\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\frac{\frac{\symX}{2^\symRegVal}}{1-e^{\frac{\symX}{2^\symRegVal}}}
++\symCountVariate_{\symRegRange+1}\frac{\frac{\symX}{2^\symRegRange}}{1-e^{\frac{\symX}{2^\symRegRange}}}.
 \end{equation}
@@
 :=
-\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
-\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\!\left(\frac{\symX}{2^\symRegVal}\right)
-+
-\symCount_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
--
-\left(\symNumReg-\symCount_0\right),
+\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
+\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symX}{2^\symRegVal}\right)
++
+\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
+-
+\left(\symNumReg-\symCountVariate_0\right),
 \end{equation}
@@
 Since the function $\symFunc(x)$ is also strictly increasing, it is obvious that there exists a unique root $\symXEstimate$ for which $\symFunc(\symXEstimate)=0$.
-The function is non-positive at 0 since $\symFunc(0)=\symCount_0-\symNumReg\leq 0$ and, in case $\symCount_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing.
-$\symCount_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $(\symRegRange+1)$, for which the maximum likelihood estimate would be infinity.
+The function is non-positive at 0 since $\symFunc(0)=\symCountVariate_0-\symNumReg\leq 0$ and, in case $\symCountVariate_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing.
+$\symCountVariate_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $(\symRegRange+1)$, for which the maximum likelihood estimate would be infinity.
 
 It is easy to see that the estimate $\symPoissonRateEstimate$ remains equal or becomes larger, when inserting an element into the HyperLogLog sketch following \cref{alg:insert}.
-An update potentially changes the multiplicity vector $(\symCount_0,\ldots,\symCount_{\symRegRange+1})$ to $(\symCount_0,\ldots,\symCount_\symIndexI-1,\ldots,\symCount_\symIndexJ+1,\ldots,\symCount_{\symRegRange+1})$ where $\symIndexI < \symIndexJ$.
+An update potentially changes the multiplicity vector $(\symCountVariate_0,\ldots,\symCountVariate_{\symRegRange+1})$ to $(\symCountVariate_0,\ldots,\symCountVariate_\symIndexI-1,\ldots,\symCountVariate_\symIndexJ+1,\ldots,\symCountVariate_{\symRegRange+1})$ where $\symIndexI < \symIndexJ$.
 Writing \eqref{equ:func} as
@@
 :=
-\symCount_0 \symX
-+
-\symCount_1 \left(
+\symCountVariate_0 \symX
++
+\symCountVariate_1 \left(
 \symHelper\!\left(\frac{x}{2^1}\right)
@@
 +
-\symCount_2 \left(
+\symCountVariate_2 \left(
 \symHelper\!\left(\frac{x}{2^2}\right)
@@
 +
-\symCount_\symRegRange
+\symCountVariate_\symRegRange
 \left(
@@
 +
-\symCount_{\symRegRange+1}
+\symCountVariate_{\symRegRange+1}
 \left(
@@
 \end{multline}
-shows that the coefficient of $\symCount_\symIndexI$ is larger than the coefficient of $\symCount_\symIndexJ$ in case $\symIndexI < \symIndexJ$.
+shows that the coefficient of $\symCountVariate_\symIndexI$ is larger than the coefficient of $\symCountVariate_\symIndexJ$ in case $\symIndexI < \symIndexJ$.
 Keeping $\symX$ fixed during an update decreases $\symFunc(\symX)$.
@@
 In this case, the maximum likelihood method under the Poisson model directly leads to the linear counting  estimator presented in \cite{Whang1990} and which was also used for small range estimation \eqref{equ:small_range_estimate}.
-Despite the assumption of a Poisson model, it is a very good approximation of the optimal martingale estimator presented in \cite{Ting2014}.
 Due to this fact we could expect that maximum likelihood estimation under the Poisson model also works well for the more general HyperLogLog case.
@@
 \symX
-\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
-+
-\left(\symNumReg-\symCount_0\right)\cdot
+\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
++
+\left(\symNumReg-\symCountVariate_0\right)\cdot
 \symHelper\!\left(
@@
 \sum_{\symRegVal=1}^{\symRegRange}
-\frac{\symCount_\symRegVal}{2^\symRegVal}
-+
-\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
+\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
++
+\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
 \right)\right)
 -
-\left(\symNumReg-\symCount_0\right).
+\left(\symNumReg-\symCountVariate_0\right).
 \end{equation}
@@
 \label{equ:strong_lower_bound}
-\symXEstimate\geq \frac{\symNumReg-\symCount_0}{\sum_{\symRegVal=1}^{\symRegRange}
-\frac{\symCount_\symRegVal}{2^\symRegVal}
-+
-\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
+\symXEstimate\geq \frac{\symNumReg-\symCountVariate_0}{\sum_{\symRegVal=1}^{\symRegRange}
+\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
++
+\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
 }
@@
 \frac{\sum_{\symRegVal=1}^{\symRegRange}
-\frac{\symCount_\symRegVal}{2^\symRegVal}
-+
-\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
+\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
++
+\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
 }
@@
 \sum_{\symRegVal=0}^{\symRegRange}
-\frac{\symCount_\symRegVal}{2^\symRegVal}
+\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
 }
@@
 \geq
-\frac{\symNumReg-\symCount_0}
-{\symCount_0+\frac{3}{2}\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCount_\symRegVal}{2^\symRegVal} + \frac{\symCount_{\symRegRange+1}}{2^\symRegRange}}.
+\frac{\symNumReg-\symCountVariate_0}
+{\symCountVariate_0+\frac{3}{2}\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}}.
 \end{equation}
@@
 \geq
-\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
-\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
-+
-\symCount_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
--
-\left(\symNumReg-\symCount_0\right),
-\end{equation}
-for $\symFunc$ can be found, where $\symRegVal'_\text{max} := \min(\symRegVal_\text{max}, \symRegRange)$ and
-$\symRegVal_\text{max} := \max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$.
+\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
+\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)
++
+\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)
+-
+\left(\symNumReg-\symCountVariate_0\right),
+\end{equation}
+for $\symFunc$ can be found, where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$ and
+$\symRegValVariate_\text{max} := \max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$.
 Again, inserting $\symXEstimate$ and transformation gives
@@
 \leq
-2^{\symRegVal'_\text{max}}
+2^{\symRegValVariate'_\text{max}}
 \log\!\left(
 1+
-\frac{\symNumReg-\symCount_0}
+\frac{\symNumReg-\symCountVariate_0}
 {
-2^{\symRegVal'_\text{max}}
+2^{\symRegValVariate'_\text{max}}
 \sum_{\symRegVal=0}^{\symRegRange}
-\frac{\symCount_\symRegVal}{2^\symRegVal}
+\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
 }
@@
 \leq
-\frac{\symNumReg-\symCount_0}
+\frac{\symNumReg-\symCountVariate_0}
 {\sum_{\symRegVal=0}^{\symRegRange}
-\frac{\symCount_\symRegVal}{2^\symRegVal}}.
-\end{equation}
-If the HyperLogLog sketch is in the intermediate range, where $\symCount_0=\symCount_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimator \eqref{equ:intermediate_estimate}.
-Hence, the maximum likelihood method leads directly to the harmonic that is used by the raw estimator.
+\frac{\symCountVariate_\symRegVal}{2^\symRegVal}}.
+\end{equation}
+If the HyperLogLog sketch is in the intermediate range, where $\symCountVariate_0=\symCountVariate_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimator \eqref{equ:intermediate_estimate}.
+Hence, the maximum likelihood method leads directly to the harmonic mean that is used by the raw estimator.
 
@@
 \end{equation}
-If $\symX_0 = 0$ with $\symFunc(\symX_0)=-\left(\symNumReg-\symCount_0\right)$, and $\symX_1$ is equal to one of the derived lower bounds \eqref{equ:strong_lower_bound} or \eqref{equ:weak_lower_bound}, the sequence $\lbrace\symX_\symIndexI\rbrace$ is montone increasing.
+If $\symX_0 = 0$ with $\symFunc(\symX_0)=-\left(\symNumReg-\symCountVariate_0\right)$, and $\symX_1$ is equal to one of the derived lower bounds \eqref{equ:strong_lower_bound} or \eqref{equ:weak_lower_bound}, the sequence $\lbrace\symX_\symIndexI\rbrace$ is montone increasing.
 Using the definitions
@@
 \label{equ:funcprime}
-\symFuncPrime(\symX):=\symFunc(\symX) + \left(\symNumReg-\symCount_0\right)
-=\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
-\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\!\left(\frac{\symX}{2^\symRegVal}\right)
-+
-\symCount_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
+\symFuncPrime(\symX):=\symFunc(\symX) + \left(\symNumReg-\symCountVariate_0\right)
+=\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
+\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symX}{2^\symRegVal}\right)
++
+\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
 \end{equation}
@@
 \Delta\symX_{\symIndexI} = \Delta\symX_{\symIndexI-1}
-\frac{\left(\symNumReg-\symCount_0\right)-\symFuncPrime(\symX_{\symIndexI-1})}{\symFuncPrime(\symX_{\symIndexI-1}) - \symFuncPrime(\symX_{\symIndexI-2})}\\
+\frac{\left(\symNumReg-\symCountVariate_0\right)-\symFuncPrime(\symX_{\symIndexI-1})}{\symFuncPrime(\symX_{\symIndexI-1}) - \symFuncPrime(\symX_{\symIndexI-2})}\\
 \symX_{\symIndexI} = \symX_{\symIndexI-1} + \Delta\symX_{\symIndexI}
@@
 \begin{itemize}
-\item Only a fraction of all count values $\symCount_\symRegVal$ is non-zero.
-If we denote $\symRegVal_\text{min}:=\min\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$ and $\symRegVal_\text{max}:=\max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$,  it is sufficient to loop over all indices in the range $[\symRegVal_\text{min}, \symRegVal_\text{max}]$.
-\item The coefficient of the linear term $\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}$ can be precalculated and reused for all function evaluations.
+\item Only a fraction of all count values $\symCountVariate_\symRegVal$ is non-zero.
+If we denote $\symRegValVariate_\text{min}:=\min\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$ and $\symRegValVariate_\text{max}:=\max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$,  it is sufficient to loop over all indices in the range $[\symRegValVariate_\text{min}, \symRegValVariate_\text{max}]$.
+\item The coefficient of the linear term $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}$ can be precalculated and reused for all function evaluations.
 \item Many programming languages allow the efficient multiplication and division by any integral power of 2 using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
-\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\left\lbrace\frac{\symX}{2^{\symRegVal'_\text{max}}},\frac{\symX}{2^{\symRegVal'_\text{max}-1}}\ldots,\frac{\symX}{2^{\symRegVal_\text{min}}}\right\rbrace$ where $\symRegVal'_\text{max} := \min(\symRegVal_\text{max}, \symRegRange)$.
+\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\left\lbrace\frac{\symX}{2^{\symRegValVariate'_\text{max}}},\frac{\symX}{2^{\symRegValVariate'_\text{max}-1}}\ldots,\frac{\symX}{2^{\symRegValVariate_\text{min}}}\right\rbrace$ where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$.
 This series corresponds to a geometric series with ratio 2.
 A straightforward calculation using \eqref{equ:helper} is very expensive because of the exponential function.
-However, if we already know $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
+However, if we already know $\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
 \begin{equation}
@@
 
-\item If $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is smaller than 0.5, the function $\symHelper(\symX)$ can be well approximated by a Taylor series around $\symX=0$
+\item If $\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ is smaller than 0.5, the function $\symHelper(\symX)$ can be well approximated by a Taylor series around $\symX=0$
 \begin{equation}
@@
 \end{equation}
-In fact, $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)\leq 0.5$ is almost always fulfilled as long as registers are not saturated.
-Using \eqref{equ:strong_upper_bound} it is straightforward to see that $\frac{\symX}{2^{\symRegVal'_\text{max}}} \leq \log 2 \approx 0.693$, if $\symCount_{\symRegRange+1}=0$.
-In case $\frac{\symX}{2^{\symRegVal'_\text{max}}} > 0.5$, the value $\frac{\symX}{2^{\symE+1}}$ is taken instead, where $\symE$ is the exponent of the floating point representation of $\symX$, $\symE = 1+\lfloor\log_2(\symX)\rfloor$.
+In fact, $\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)\leq 0.5$ is almost always fulfilled as long as registers are not saturated.
+Using \eqref{equ:strong_upper_bound} it is straightforward to see that $\frac{\symX}{2^{\symRegValVariate'_\text{max}}} \leq \log 2 \approx 0.693$, if $\symCountVariate_{\symRegRange+1}=0$.
+In case $\frac{\symX}{2^{\symRegValVariate'_\text{max}}} > 0.5$, the value $\frac{\symX}{2^{\symE+1}}$ is taken instead, where $\symE$ is the exponent of the floating point representation of $\symX$, $\symE = 1+\lfloor\log_2(\symX)\rfloor$.
 By definition, $\frac{\symX}{2^{\symE+1}}\leq 0.5$ which allows using the Taylor series approximation.
-$\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is finally obtained after $\symE + 1 -\symRegVal'_\text{max}$ iterations using \eqref{equ:helper_recursion2}.
+$\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ is finally obtained after $\symE + 1 -\symRegValVariate'_\text{max}$ iterations using \eqref{equ:helper_recursion2}.
 \end{itemize}
@@
 \begin{algorithmic}
-\Function {EstimateCardinality}{$\boldsymbol{\symCount}$}
-\State $\symRegRange\gets\dim(\boldsymbol{\symCount})-2$
-\State $\symRegVal_\text{min} \gets \min\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$
-\If{$\symRegVal_\text{min} > \symRegRange$}
+\Function {EstimateCardinality}{$\boldsymbol{\symCountVariate}$}
+\State $\symRegRange\gets\dim(\boldsymbol{\symCountVariate})-2$
+\State $\symRegValVariate_\text{min} \gets \min\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$
+\If{$\symRegValVariate_\text{min} > \symRegRange$}
 \State\Return $\infty$
 \EndIf
-\State $\symRegVal'_\text{min} \gets \max(\symRegVal_\text{min}, 1)$
-\State $\symRegVal_\text{max} \gets \max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$
-\State $\symRegVal'_\text{max} \gets \min(\symRegVal_\text{max}, \symRegRange)$
+\State $\symRegValVariate'_\text{min} \gets \max(\symRegValVariate_\text{min}, 1)$
+\State $\symRegValVariate_\text{max} \gets \max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$
+\State $\symRegValVariate'_\text{max} \gets \min(\symRegValVariate_\text{max}, \symRegRange)$
 \State $\symZ \gets 0$
-\State $\symNumReg'\gets \symCount_{\symRegRange+1}$
-\State $\symY \gets 2^{-\symRegVal'_\text{max}}$
-\For{$\symRegVal \gets \symRegVal'_\text{max},\symRegVal'_\text{min}$}
-\State $\symZ \gets \symZ + \symCount_\symRegVal\cdot\symY$
+\State $\symNumReg'\gets \symCountVariate_{\symRegRange+1}$
+\State $\symY \gets 2^{-\symRegValVariate'_\text{max}}$
+\For{$\symRegVal \gets \symRegValVariate'_\text{max},\symRegValVariate'_\text{min}$}
+\State $\symZ \gets \symZ + \symCountVariate_\symRegVal\cdot\symY$
 \Comment here $\symY = 2^{-\symRegVal}$
 \State $\symY \gets 2\symY$
-\State $\symNumReg'\gets\symNumReg' + \symCount_\symRegVal$
+\State $\symNumReg'\gets\symNumReg' + \symCountVariate_\symRegVal$
 \EndFor
-\State $\symNumReg \gets \symNumReg' + \symCount_0$ 
-\State $\symCount' \gets \symCount_{\symRegRange+1}$
+\State $\symNumReg \gets \symNumReg' + \symCountVariate_0$ 
+\State $\symCount \gets \symCountVariate_{\symRegRange+1}$
 \If{$\symRegRange\geq 1$}
-\State $\symCount' \gets \symCount' + \symCount_{\symRegVal'_\text{max}}$
+\State $\symCount \gets \symCount + \symCountVariate_{\symRegValVariate'_\text{max}}$
 \EndIf
 \State $\symFuncPrime_\text{prev}\gets 0$
-\State $\symA\gets \symZ + \symCount_0$
+\State $\symA\gets \symZ + \symCountVariate_0$
 \State $\symB\gets \symZ + 
-\symCount_{\symRegRange+1}\cdot 2^{-\symRegRange}$
+\symCountVariate_{\symRegRange+1}\cdot 2^{-\symRegRange}$
 \If{$\symB \leq 1.5\cdot\symA$}
@@
 \State $\symXNormalized \gets
-\symX \cdot 2^{-\max(\symRegVal'_\text{max}+1, \symE+2)}$
+\symX \cdot 2^{-\max(\symRegValVariate'_\text{max}+1, \symE+2)}$
 \Comment $\symXNormalized \in [0, 0.25]$
@@
 \Comment Taylor approximation \eqref{equ:taylor}
-\For{$\symRegVal\gets \symE,\symRegVal'_\text{max}$}
+\For{$\symRegVal\gets \symE,\symRegValVariate'_\text{max}$}
 \State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
@@
 \EndFor
-\State $\symFuncPrime \gets \symCount'\cdot\symHelper$
+\State $\symFuncPrime \gets \symCount\cdot\symHelper$
 \Comment compare \eqref{equ:funcprime}
-\For{$\symRegVal\gets(\symRegVal'_\text{max}-1),\symRegVal'_\text{min}$}
+\For{$\symRegVal\gets(\symRegValVariate'_\text{max}-1),\symRegValVariate'_\text{min}$}
 \State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
 \Comment calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion2}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
-\State $\symFuncPrime\gets \symFuncPrime + \symCount_\symRegVal\cdot\symHelper$
+\State $\symFuncPrime\gets \symFuncPrime + \symCountVariate_\symRegVal\cdot\symHelper$
 \State $\symXNormalized \gets 2\symXNormalized$
@@
 Assume two different sets $\symSetS_1$ and $\symSetS_2$ that have been recorded by two HyperLogLog sketches with same parameters $(\symPrecision,\symRegRange)$.
-Given the corresponding register values $\boldsymbol{\symRegVal}_1$ and
-$\boldsymbol{\symRegVal}_2$, we attempt to estimate the cardinalities of the pair-wise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$.
-Motivated by the good results we have obtained for cardinality estimation of a single HyperLogLog sketch, we want to get these estimates using the maximum likelihood method applied to the joint probability distribution of $\boldsymbol{\symRegVal}_1$ and $\boldsymbol{\symRegVal}_2$.
+Given the corresponding register values $\boldsymbol{\symRegValVariate}_1$ and
+$\boldsymbol{\symRegValVariate}_2$, we attempt to estimate the cardinalities of the pair-wise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$.
+Motivated by the good results we have obtained for cardinality estimation of a single HyperLogLog sketch, we want to get these estimates using the maximum likelihood method applied to the joint probability distribution of $\boldsymbol{\symRegValVariate}_1$ and $\boldsymbol{\symRegValVariate}_2$.
 

### 2016-07-12 paper/paper.tex
major revision

@@
 \symDefine{\symB}{b}
-\symDefine{\symE}{e}
+\symDefine{\symKappa}{\kappa}
 \symDefine{\symX}{x}
@@
 \title{New cardinality estimation algorithms for HyperLogLog sketches}
-\subtitle{Draft version \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}}
-\author{Otmar Ertl}
-
+\subtitle{Draft version}
+\author{Otmar Ertl \\ otmar.ertl@gmail.com}
 \begin{document}
 \maketitle
+\let\thefootnote\relax\footnotetext{This article together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}.}
 \begin{abstract}
@@
 $\symPrecision$ is the precision that directly influences the relative error which scales like $1/\sqrt{\symNumReg}$.
-All registers start with initial value zero.
+All registers start with zero initial value.
 Each element insertion potentially increases the value of one of these registers.
@@
 Among the next $\symRegRange$ bits, the position of the first 1-bit is determined which is a value in the range $[1,\symRegRange+1]$.
-The value $\symRegRange+1$ is used, if all $\symRegRange$ bits are equal to 0.
-If the position of the first 1-bit exceeds the value of the selected register, the register value is replaced.
+The value $\symRegRange+1$ is used, if all $\symRegRange$ bits are zeros.
+If the position of the first 1-bit exceeds the current value of the selected register, the register value is replaced.
 \cref{alg:insert} shows the update procedure for inserting a data element into the HyperLogLog sketch.
 
-A HyperLogLog sketch can be characterized by the parameter pair $(\symPrecision, \symRegRange)$.
-The first parameter controls the relative error while the second defines the possible value range of a register.
+A HyperLogLog sketch can be characterized by a parameter pair $(\symPrecision, \symRegRange)$.
+The first parameter $\symPrecision$ is the precision and controls the relative error while the second defines the possible value range of a register.
 A register can take all values starting from 0 to $\symRegRange+1$, inclusively.
-
 The sum $\symPrecision+\symRegRange$ corresponds to the number of consumed hash value bits and defines the maximum cardinality that can be tracked.
-Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent and the estimation accuracy is drastically reduced.
+Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent and the estimation accuracy will be drastically reduced.
 
@@
 \cref{alg:insert} has some properties which are especially useful in distributed environments.
-The insertion order of elements has no influence on the final HyperLogLog sketch.
-Furthermore, HyperLogLog sketches with same parameters $(\symPrecision, \symRegRange)$ can be easily merged.
-The resulting register values are simply given by the register-wise maximum values.
-The final HyperLogLog sketch corresponds to the union of both sets that have been represented by the two input sketches.
+First, the insertion order of elements has no influence on the final HyperLogLog sketch.
+Furthermore, any two HyperLogLog sketches with same parameters $(\symPrecision, \symRegRange)$ representing  sets $\symSetA$ and $\symSetB$ can be easily merged.
+The resulting register values are simply given by the register-wise maximum values
+\begin{equation}
+\label{equ:hll_union}
+\symRegValVariate^{\symSetA\cup\symSetB}_{\symIndexI} = 
+\max
+\!
+\left(
+\symRegValVariate^{\symSetA}_{\symIndexI},
+\symRegValVariate^{\symSetB}_{\symIndexI}
+\right)
+\quad
+\text{for $1\leq\symIndexI\leq\symNumReg$}.
+\end{equation}
+The final HyperLogLog sketch represents the union of $\symSetA$ and $\symSetB$.
 
 A $(\symPrecision, 0)$-HyperLogLog sketch corresponds to a bit array as used by linear counting \cite{Whang1990}.
-In this case each register value is represented by a single bit.
+Each register value can be stored by a single bit in this case.
 Hence, linear counting can be regarded as a special case of the HyperLogLog algorithm for which $\symRegRange = 0$.
@@
 \caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers.
-All register values $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ are initially equal to 0 and remain in the range $[0, \symRegRange+1]$.}
+All registers $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ have zero initial value and remain in the range $[0, \symRegRange+1]$.}
 \label{alg:insert}
@@
 \State $\langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision,\symBitRepB_1,\ldots,\symBitRepB_\symRegRange\rangle_2 \gets$ $(\symPrecision + \symRegRange)$-bit hash value of $\symDataItem$
-\State $\symIndexJ \gets 1+ \langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision\rangle_2$
-\State $\symRegValVariate' = \min(\{\symS\mid \symS \in [1, \symRegRange]  \wedge  \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
-\State $\symRegValVariate_\symIndexJ \gets\max(\symRegValVariate_\symIndexJ, \symRegValVariate')$
+\State $\symIndexI \gets 1+ \langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision\rangle_2$
+\Comment $\symIndexI\in[1,2^\symPrecision]$
+\State $\symRegVal = \min(\{\symS\mid \symS \in [1, \symRegRange]  \wedge  \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
+\Comment $\symRegVal\in[1,\symRegRange+1]$
+\State $\symRegValVariate_\symIndexI \gets\max(\symRegValVariate_\symIndexI, \symRegVal)$
 \EndProcedure
@@
 
-Under the assumption of a strong universal hash function, the probability that the register values $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ are equal to $\boldsymbol{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ is given by the corresponding probability mass function
+Under the assumption of a uniform hash function, the probability that the register values $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ are equal to $\boldsymbol{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ is given by the corresponding probability mass function
 \begin{equation}
@@
 =
-\sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots\symCardinality_\symNumReg}
-\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexJ=1}^\symNumReg \symRegProbability(\symRegVal_\symIndexJ\vert\symCardinality_\symIndexJ)
+\sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots,\symCardinality_\symNumReg}
+\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexI=1}^\symNumReg \symRegProbability(\symRegVal_\symIndexI\vert\symCardinality_\symIndexI)
 \end{equation}
@@
 The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities.
-$\symRegProbability(\symRegVal\vert\symCardinality)$ is the probability that a register has value $\symRegVal$ after inserting $\symCardinality$ distinct elements
+$\symRegProbability(\symRegVal\vert\symCardinality)$ is the probability that the smallest of $\symCardinality$ uniformly distributed random integers in range $[0, 2^\symRegRange)$ has a binary representation with its first 1-bit at position $\symRegVal$,
 \begin{equation}
@@
 \left(1-\frac{1}{2^\symRegVal}\right)^\symCardinality - \left(1-\frac{1}{2^{\symRegVal-1}}\right)^\symCardinality & \symCardinality \geq 1 \wedge 1\leq\symRegVal\leq\symRegRange\\
-1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symCardinality & \symCardinality\geq 1 \wedge \symRegVal = \symRegRange +1 \\
-\end{cases}
+1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symCardinality & \symCardinality\geq 1 \wedge \symRegVal = \symRegRange +1
+\end{cases}.
 \end{equation}
@@
 More formally, the multiset $\lbrace\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg\rbrace$ is a sufficient statistic for $\symCardinality$.
-Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg\rbrace = 0^{\symCountVariate_0}1^{\symCountVariate_1}\cdots\symRegRange^{\symCountVariate_{\symRegRange}}(\symRegRange+1)^{\symCountVariate_{\symRegRange+1}}$ where $\symCountVariate_\symIndexJ$ is the multiplicity of value $\symIndexJ$.
+Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg\rbrace = 0^{\symCountVariate_0}1^{\symCountVariate_1}\cdots\symRegRange^{\symCountVariate_{\symRegRange}}(\symRegRange+1)^{\symCountVariate_{\symRegRange+1}}$ where $\symCountVariate_\symRegVal$ is the multiplicity of value $\symRegVal$.
 As a consequence, the multiplicity vector $\boldsymbol{\symCountVariate} := (\symCountVariate_0,\ldots,\symCountVariate_{\symRegRange+1})$ is also a sufficient statistic for the cardinality.
@@
 \label{sec:poisson_approximation}
-Unfortunately, the probability mass function \eqref{equ:multinomialProbabilityMass} is difficult to analyze, because the register values are statistically dependent.
-Therefore, as proposed in \cite{Flajolet2007} a Poisson model can be used, which assumes that the cardinality itself is distributed according to a Poisson distribution
+Due to the statistical dependence of the register values, the probability mass function \eqref{equ:multinomialProbabilityMass} is difficult to analyze.
+Therefore, a Poisson model can be used \cite{Flajolet2007}, which assumes that the cardinality itself is distributed according to a Poisson distribution
 \begin{equation}
@@
 \sum_{\symCardinality_\symNumReg=0}^\infty
-\prod_{\symIndexJ=1}^\symNumReg
-\symRegProbability(\symCardinality_\symIndexJ\vert\symRegVal_\symIndexJ)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexJ}}{\symCardinality_\symIndexJ!\symNumReg^{\symCardinality_\symIndexJ}}
+\prod_{\symIndexI=1}^\symNumReg
+\symRegProbability(\symRegVal_\symIndexI\vert\symCardinality_\symIndexI)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexI}}{\symCardinality_\symIndexI!\symNumReg^{\symCardinality_\symIndexI}}
 \nonumber\\
 &= 
-\prod_{\symIndexJ=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability(\symCardinality_\symIndexJ\vert\symRegVal_\symIndexJ)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
+\prod_{\symIndexI=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability(\symRegVal_\symIndexI\vert\symCardinality)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
 \nonumber\\
 &= 
-\prod_{\symRegVal=0}^{\symRegRange+1} \left[
+\prod_{\symRegVal=0}^{\symRegRange+1} \left(
 \sum_{\symCardinality=0}^\infty
-\symRegProbability(\symCardinality_\symIndexJ\vert\symRegVal_\symIndexJ)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
-\right]^{\symCount_\symRegVal}
+\symRegProbability(\symRegVal\vert\symCardinality)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
+\right)^{\!\!\symCount_\symRegVal}
 \nonumber\\
@@
 e^{-\symCount_0\frac{\symPoissonRate}{\symNumReg}}
-\left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\symCount_\symRegVal}\right)
-\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\symCount_{\symRegRange+1}}.
+\left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\!\symCount_\symRegVal}\right)
+\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\!\symCount_{\symRegRange+1}}.
 \end{align}
-Here $\boldsymbol{\symCount}=(\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is the multiplicity vector for $\boldsymbol{\symRegVal}$.
-$\symCount_\symRegVal$ denotes the multiplicity of value $\symRegVal$  in the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$.
-The final factorization shows that all register values are independent and identically distributed under the Poisson model.
+Here $\symCount_\symRegVal$ denotes the multiplicity of value $\symRegVal$  in the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$.
+The final factorization shows that under the Poisson values the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ are independent and identically distributed.
 The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is defined by
@@
 Due to the simpler probability mass function, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\boldsymbol{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ rather than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}.
-Here $\boldsymbol{\symRegValVariate}$ denotes the vector of all observed register values.
-Each component of $\boldsymbol{\symRegValVariate}$ is distributed according to \eqref{equ:multinomialProbabilityMass}.
 Depoissonization \cite{Jacquet1998} finally allows to translate the estimates back to the fixed-size model.
@@
 \end{equation}
-Consequently, the unbiased estimator $\symPoissonRateEstimate$ conditioned on $\symCardinality$ is also an unbiased estimator for $\symCardinality$ which motivates us to use $\symPoissonRateEstimate$ directly as estimator for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$.
+Hence, the unbiased estimator $\symPoissonRateEstimate$ conditioned on $\symCardinality$ is also an unbiased estimator for $\symCardinality$ which motivates us to use $\symPoissonRateEstimate$ directly as estimator for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$.
 As simulation results will show later, the Poisson approximation works well over the entire cardinality range, even for estimators that are not exactly unbiased.
@@
 \label{sec:cardinality_estimation}
-The original cardinality estimator \cite{Flajolet2007} is based on the idea that approximately $\symNumReg  2^{\symRegVal-1}$ distinct element insertions are needed until the value of a register reaches $\symRegVal$.
-Given that, a rough cardinality estimate can be obtained by calculating the average over the values $\lbrace \symNumReg 2^{\symRegValVariate_1-1},\ldots,\symNumReg2^{\symRegValVariate_\symNumReg-1}\rbrace$.
+The original cardinality estimator \cite{Flajolet2007} is based on the idea that the number of distinct element insertions a register needs to reach the value $\symRegVal$ is proportional to $\symNumReg  2^{\symRegVal}$.
+Given that, a rough cardinality estimate can be obtained by calculating the average over the values $\lbrace \symNumReg 2^{\symRegValVariate_1},\ldots,\symNumReg2^{\symRegValVariate_\symNumReg}\rbrace$.
 
@@
 First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}.
-Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers.
-The result is the so-called raw estimator and given by
-\begin{equation}
-\label{equ:intermediate_estimate}
+Finally, the harmonic mean (HM) was found to give even better estimates as it is inherently less sensitive to outliers.
+The result is the so-called raw estimator which is given by
+\begin{equation}
+\label{equ:raw_estimator}
 \symCardinalityRawEstimate
@@
 = 
-\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegValVariate_\symIndexJ}}
+\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexI=1}^{\symNumReg}2^{-\symRegValVariate_\symIndexI}}
 = 
@@
 \end{equation}
-Here $2\alpha_\symNumReg$ is a bias correction factor. $\alpha_\symNumReg$ for a given number of registers $\symNumReg$ was derived to be \cite{Flajolet2007}
+Here $\alpha_\symNumReg$ is a bias correction factor which was derived for a given number of registers $\symNumReg$ to be \cite{Flajolet2007}
 \begin{equation}
@@
 \left(
-\log_2\left(
+\log_2\!\left(
 \frac{2+u}{1+u}
@@
 du
-\right)^{-1}.
+\right)^{\!\!-1}.
 \end{equation}
@@
 \end{equation}
-as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible compared to the estimation error.
+as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:raw_estimator}, because the additional bias is negligible compared to the estimation error.
 
@@
 More details of the experimental setup will be explained later in \cref{sec:corrected_raw_estimation_error}.
-Obviously, the raw estimator is biased for small and large cardinalities and fails to return accurate estimates.
-Therefore, to cover the entire range of cardinalities, corrections for small and large cardinalities have been proposed, respectively.
-
-As mentioned in \cref{sec:data_element_insertion}, a HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be turned into a $(\symPrecision, 0)$-HyperLogLog sketch.
-Since $\symRegRange=0$ corresponds to linear counting and the transformed HyperLogLog sketch corresponds to a bitset with $\symCountVariate_0$ zeros, the linear counting cardinality estimator \cite{Whang1990} can be applied here
-\begin{equation}
-\label{equ:small_range_estimate}
+Obviously, the raw estimator is biased for small and large cardinalities where it fails to return accurate estimates.
+In order to cover the entire range of cardinalities, corrections for small and large cardinalities have been proposed.
+
+As mentioned in \cref{sec:data_element_insertion}, a HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be mapped to a $(\symPrecision, 0)$-HyperLogLog sketch.
+Since $\symRegRange=0$ corresponds to linear counting and the reduced HyperLogLog sketch corresponds to a bitset with $\symCountVariate_0$ zeros, the linear counting cardinality estimator \cite{Whang1990} can be used
+\begin{equation}
+\label{equ:linear_counting_estimator}
 \symCardinalityEstimate_\text{small} = \symNumReg \log(\symNumReg/\symCountVariate_0).
@@
 The corresponding relative estimation error as depicted in \cref{fig:small_range_estimate} shows that this estimator is convenient for small cardinalities.
-It was proposed to use this small-range estimator as long as $\symCardinalityRawEstimate \leq \frac{5}{2}\symNumReg$ where the factor $\frac{5}{2}$ was empirically determined \cite{Flajolet2007}.
-
-For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state meaning that they have reached the maximum possible value $\symRegRange + 1$, the raw estimator underestimates the cardinalities.
-For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$ which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
+It was proposed to use this estimator for small cardinalities as long as $\symCardinalityRawEstimate \leq \frac{5}{2}\symNumReg$ where the factor $\frac{5}{2}$ was empirically determined \cite{Flajolet2007}.
+
+For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state, meaning that they have reached the maximum possible value $\symRegRange + 1$, the raw estimator underestimates the cardinalities.
+For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$, which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
 \begin{equation}
@@
 \end{equation}
-The original estimation algorithm as presented in \cite{Flajolet2007} including small and large range corrections is summarized by \cref{alg:estimate_original}. 
+The original estimation algorithm as presented in \cite{Flajolet2007} including corrections for small and large cardinalities is summarized by \cref{alg:estimate_original}. 
 \begin{algorithm}
@@
 \State $\symNumReg \gets \dim \boldsymbol{\symRegValVariate}$
-\State $\symCardinalityRawEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegValVariate_\symIndexJ}\right)^{-1}$
-\Comment raw estimate \eqref{equ:intermediate_estimate}
+\State $\symCardinalityRawEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexI=1}^\symNumReg 2^{-\symRegValVariate_\symIndexI}\right)^{-1}$
+\Comment raw estimate \eqref{equ:raw_estimator}
 \If{$\symCardinalityRawEstimate\leq \frac{5}{2}\symNumReg$}
-\State $\symCountVariate_0 = \left|\{\symIndexJ |\symRegValVariate_\symIndexJ=0\}\right|$
+\State $\symCountVariate_0 = \left|\{\symIndexI \vert\symRegValVariate_\symIndexI=0\}\right|$
 \If{$\symCountVariate_0\neq0$}
-\State \Return $\symNumReg \log(\symNumReg / \symCountVariate_0)$ \Comment small range correction \eqref{equ:small_range_estimate}
+\State \Return $\symNumReg \log(\symNumReg / \symCountVariate_0)$ \Comment small range correction \eqref{equ:linear_counting_estimator}
 \Else
@@
 \includesvg[width=1\textwidth]{raw_estimate}
-\caption{The distribution of the relative estimation error over the cardinality as for the raw estimator \cite{Flajolet2007}.
-10\,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$ have been evaluated.}
+\caption{The distribution of the relative estimation error over the cardinality for the raw estimator after evaluation of 10\,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
 \label{fig:raw_estimate}
@@
 \includesvg[width=1\textwidth]{small_range_estimate}
-\caption{The distribution of the relative estimation error for the linear counting estimator which can be used for small cardinalities.}
+\caption{The distribution of the relative estimation error for the linear counting estimator after evaluation of 10\,000 randomly generated bitmaps of size $2^{12}$ which correspond to HyperLogLog sketches with parameters $\symPrecision = 12$ and $\symRegRange=0$.}
 \label{fig:small_range_estimate}
@@
 \includesvg[width=1\textwidth]{original_estimate}
-\caption{The distribution of the relative estimation error over the cardinality as obtained by the original estimation method \cite{Flajolet2007}.
-10\,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$ have been evaluated.}
+\caption{The distribution of the relative estimation error over the cardinality for the original estimation algorithm after evaluation of 10\,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
 \label{fig:original_estimate}
@@
 
-The large range correction is not satisfying either as it does not reduce the estimation error.
-Quite the contrary, it even increases the estimation error.
+The large range correction formula \eqref{equ:large_range_estimate} is not satisfying either as it does not reduce the estimation error.
+Quite the contrary, it even makes the bias worse.
 However, instead of underestimating the cardinalities, they are now overestimated.
-Another indication for the incorrectness of the proposed large range correction \eqref{equ:large_range_estimate} is the fact that it is not even defined for all possible states.
+Another indication for the incorrectness of the proposed large range correction is the fact that it is not even defined for all possible states.
 For instance, consider a $(\symPrecision,\symRegRange)$-HyperLogLog sketch with $\symPrecision+\symRegRange=32$ for which all registers are equal to the maximum possible value $\symRegRange+1$.
-The raw estimate would be $\symCardinalityRawEstimate = \symAlpha_\symNumReg 2^{33}$, which is greater than $2^{32}$ and outside of the domain of the large range correction.
+The raw estimate would be $\symCardinalityRawEstimate = \symAlpha_\symNumReg 2^{33}$, which is greater than $2^{32}$ and outside of the domain of the large range correction formula.
 
 A simple approach to avoid the need of any large range correction is to extend the operating range of the raw estimator to larger cardinalities.
-This can be accomplished by using hash values with more bits $(\symPrecision+\symRegRange>32)$.
-The drawback is that 5-bit registers are no longer sufficient to represent all possible values in the range $[0, \symRegRange+1]$ in case $\symRegRange \geq 30$.
-For HyperLogLog sketches with 6-bit registers in combination with 64-bit hash values $(\symPrecision+\symRegRange=64)$ as proposed in \cite{Heule2013}, it is unrealistic in practice to encounter cardinalities of order $2^{64}$ for which the raw estimator would return incorrect results.
-
-%\begin{algorithm}
-%\caption{Calculate the sufficient statistic $\boldsymbol{\symCount}=(\symCount_0, \ldots,\symCount_{\symRegRange+1})$ for given register values $\boldsymbol{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg)$.}
-%\begin{algorithmic}
-%\Function {CountRegisterValues}{$\boldsymbol{\symRegVal}$}
-%\State $\boldsymbol{\symCount}\gets\boldsymbol{0}$
-%\ForAll{$\symRegVal'\in\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$}
-%\State $\symCount_{\symRegVal'} \gets \symCount_{\symRegVal'} + 1$
-%\EndFor
-%\State\Return $\boldsymbol{\symCount}$
-%\EndFunction
-%\end{algorithmic}
-%\end{algorithm}
-%
+This can be easily accomplished by using hash values with more bits $(\symPrecision+\symRegRange>32)$.
+If $\symRegRange \geq 30$, 5-bit registers are no longer sufficient to represent all possible values in the range $[0, \symRegRange+1]$.
+If, for example, 64-bit hash values $(\symPrecision+\symRegRange=64)$ are used, as proposed in \cite{Heule2013}, 6 bits per register are needed.
+Hash value sizes with more than 64 bits are useless in practice, because it is unrealistic to encounter cardinalities of order $2^{64}$ for which the raw estimator would be biased.
 
@@
 \label{sec:derivation_raw_estimator}
-In order to better understand why the raw estimator fails for small and large cardinalities, we start with a brief derivation that does not require complex analysis.
-
-Consider following integer distribution
+In order to better understand why the raw estimator fails for small and large cardinalities, we start with a brief and simple derivation without the restriction tolarge cardinalities ($\symCardinality\rightarrow\infty$) and without using complex analysis as in \cite{Flajolet2007}.
+
+Let us assume that the register values have following cumulative distribution function
 \begin{equation}
@@
 For now we ignore that this distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$.
-The expectation of $2^{-\symRegValVariate}$ is given by
+For a random variable $\symRegValVariate$ obeying \eqref{equ:assumed_register_val_distribution} the expectation of $2^{-\symRegValVariate}$ is given by
 \begin{multline}
@@
 2^{-\symRegVal}
-e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}
-\left(1-e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}\right)
+\left(
+e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}
+-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal-1}}}
+\right)
 =
@@
 \end{equation}
-Therefore, 
+Therefore, the asymptotic relative bias of 
 \begin{equation}
@@
 \end{equation}
-is an asymptotically almost unbiased estimator for the Poisson parameter.
-Its asymptotic relative bias is bounded by $\symEpsPowerSeriesFunc$.
-This estimator corresponds to the raw estimator \eqref{equ:intermediate_estimate}, if the Poisson parameter estimate is used as cardinality estimate (see \cref{sec:poisson_approximation}).
-
-The estimator \eqref{equ:rawestimator_2} can also be written as
+is bounded by $\symEpsPowerSeriesFunc$, which makes this statistic a good estimator for the Poisson parameter.
+It also corresponds to the raw estimator \eqref{equ:raw_estimator}, if the Poisson parameter estimate is used as cardinality estimate (see \cref{sec:depoissonization}).
+
+\subsection{Limitations of the raw estimator}
+The raw estimator is based on two prerequisites.
+First, the number of registers needs to be sufficiently large ($\symNumReg\rightarrow\infty$).
+And second, the distribution of register values can be described by \eqref{equ:assumed_register_val_distribution}.
+However, the latter is not true for small and large cardinalities, which is finally the reason for the bias of the raw estimator.
+
+A random variable $\symRegValVariate'$ with cumulative distribution function \eqref{equ:assumed_register_val_distribution} can be transformed into a random variable $\symRegValVariate$ with cumulative distribution function \eqref{equ:register_value_distribution} using
+\begin{equation}
+\label{equ:dist_transformation}
+\symRegValVariate = \min\!\left(\max\!\left(\symRegValVariate', 0\right), \symRegRange+1\right).
+\end{equation}
+Therefore, the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ can be seen as the result after applying this transformation to a sample $\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg$ of the distribution described by \eqref{equ:assumed_register_val_distribution}.
+If all registers values of the HyperLogLog sketch lie in the range $[1,\symRegRange]$, they are identical to the values $\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg$.
+In other words, the observed register values are also a plausible sample of the assumed distribution described by \eqref{equ:assumed_register_val_distribution} in this case.
+Hence, as long as all or at least most register values are in the range $[1,\symRegRange]$ the approximation of \eqref{equ:register_value_distribution} by \eqref{equ:assumed_register_val_distribution} is valid.
+This explains why the raw estimator works best for intermediate cardinalities.
+However, for small and large cardinalities many registers have values equal to 0 or $\symRegRange+1$, respectively, which cannot be described by \eqref{equ:assumed_register_val_distribution} and which finally leads to the observed bias.
+
+\subsection{Corrections to the raw estimator}
+If we knew the values $\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg$ for which the transformation \eqref{equ:dist_transformation} led to the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$, we would be able to estimate $\symPoissonRate$ using
+\begin{equation}
+\symPoissonRateEstimate 
+= 
+\frac{\symAlpha_\infty\,\symNumReg^2}{2^{-\symRegValVariate'_1}+\ldots+2^{-\symRegValVariate'_\symNumReg}}.
+\end{equation}
+As we have already shown, this estimator is approximately unbiased, because all $\symRegValVariate'_\symIndexI$ follow the assumed distribution.
+It would be even sufficient, if we know the multiplicity $\symCountVariate'_\symRegVal$
+of each $\symRegVal\in\mathbb{Z}$ in $\lbrace\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg\rbrace$, $\symCountVariate'_\symRegVal := \left|\lbrace \symIndexI\vert \symRegVal = \symRegValVariate'_\symIndexI\rbrace\right|$, because the raw estimator can be also written as
 \begin{equation}
@@
 = 
-\frac{\symAlpha_\infty\,\symNumReg^2}{\sum_{\symRegVal=-\infty}^\infty \symCountVariate'_\symRegVal 2^{-\symRegVal}},
-\end{equation}
-if $\symCountVariate'_\symRegVal$ is the multiplicity of value $\symRegVal$ in the sample $\left\lbrace \symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg\right\rbrace$. 
-By definition, we have $\sum_{\symRegVal=-\infty}^\infty \symCountVariate'_\symRegVal = \symNumReg$.
-
-\subsection{Limitations of the raw estimator}
-As we have already seen in \cref{fig:raw_estimate}, the raw estimator does not work very well for small or large cardinalities.
-The reason is that the distribution of register values \eqref{equ:register_value_distribution} differs from \eqref{equ:assumed_register_val_distribution} for which the raw estimator was derived.
-Since a random variable $\symRegValVariate'$ that obeys \eqref{equ:assumed_register_val_distribution} can be transformed into random variable $\symRegValVariate$ that follows \eqref{equ:register_value_distribution} using the transformation $\symRegValVariate = \min\left(\max\left(\symRegValVariate', 0\right), \symRegRange+1\right)$, the register values of a HyperLogLog sketch can be seen as the result after applying this transformation to a sample that is distributed according to \eqref{equ:assumed_register_val_distribution}.
-The corresponding multiplicities have  following relationships
+\frac{\symAlpha_\infty\,\symNumReg^2}{\sum_{\symRegVal=-\infty}^\infty \symCountVariate'_\symRegVal 2^{-\symRegVal}}.
+\end{equation}
+Due to \eqref{equ:dist_transformation}, the multiplicities $\symCountVariate'_\symRegVal$ and the multiplicities $\symCountVariate_\symRegVal$ for the register values have following relationships
 \begin{equation}
@@
 \end{equation}
-Obviously, as long as $\symCountVariate_0\ll\symNumReg$ and  $\symCountVariate_{\symRegRange+1}\ll\symNumReg$ the set of register values is similar to the originating sample.
-In case $\symCountVariate_0=0$ and $\symCountVariate_{\symRegRange+1}=0$, which implies $\symCountVariate'_\symRegVal=0$ for $\symRegVal\leq0$ and $\symRegVal\geq\symRegRange+1$, we can even be sure that the register values are identical to the originating sample.
-This explains why the raw estimator only works well in the intermediate cardinality range for which the number of registers with values equal to 0 or $\symRegRange+1$ is negligible.
-
-\subsection{Corrections to the raw estimator}
-Given a HyperLogLog sketch with multiplicity vector $\boldsymbol{\symCountVariate}=\left(\symCountVariate_0,\ldots,\symCountVariate_{\symRegRange+1}\right)$, we try to find estimates $\hat{\symCount}'_\symRegVal$ for $\symCountVariate'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and replace them in the estimation formula \eqref{equ:raw_estimate_with_multiplicities}.
+The idea is now to find estimates $\hat{\symCount}'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and use them as replacements for $\symCountVariate'_\symRegVal$ in \eqref{equ:raw_estimate_with_multiplicities}.
 For $\symRegVal\in[1,\symRegRange]$ where $\symCountVariate_\symRegVal = \symCountVariate'_\symRegVal$ we can use the trivial estimators
@@
 \end{equation}
-To get estimators for all other $\symRegVal\notin[1,\symRegRange]$, we consider their expectations
+To get estimators for $\symRegVal\leq 0$ and $\symRegVal\geq\symRegRange+1$, we consider the expectations
 \begin{equation}
@@
 \hat{\symCount}'_\symRegVal
-:=
+=
 \symNumReg
@@
 
-Inserting all these estimators into \eqref{equ:raw_estimate_with_multiplicities} as replacements for $\symCount'_\symRegVal$ finally gives
+Inserting all these estimators into \eqref{equ:raw_estimate_with_multiplicities} as replacements for $\symCountVariate'_\symRegVal$ finally gives
 \begin{equation}
@@
 which we call the corrected raw estimator.
-Here $\symNumReg\,\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$ and $2\symNumReg\,\symLargeCorrectionFunc(1-\symCountVariate_{\symRegRange+1}/\symNumReg)$ are replacements for  $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$ in the raw estimator \eqref{equ:intermediate_estimate}, respectively.
+Here $\symNumReg\,\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$ and $2\symNumReg\,\symLargeCorrectionFunc(1-\symCountVariate_{\symRegRange+1}/\symNumReg)$ are replacements for  $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$ in the raw estimator \eqref{equ:raw_estimator}, respectively.
 The functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are defined as
@@
 \end{equation}
-which is almost identical to the linear counting estimator used for the small range correction \eqref{equ:small_range_estimate}.
-Here we used the fact that $\symPowerSeriesFunc$ can be well approximated by 1 (see \cref{sec:derivation_raw_estimator}).
+which is almost identical to the linear counting estimator used for the small range correction \eqref{equ:linear_counting_estimator}.
+Here we used again the fact that the function $\symPowerSeriesFunc$ can be well approximated by 1 (see \cref{sec:derivation_raw_estimator}).
 \subsection{Corrected raw estimation algorithm}
@@
 The values of functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ can be either calculated on-demand or pre-calculated.
-The possible value range for $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$ is $[0, \symNumReg]$.
+The possible value range of $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$ is $[0, \symNumReg]$.
 Therefore, if performance matters, it is possible to precalculate the function values for all possible arguments and keep them in lookup tables of size $\symNumReg+1$.
@@
 \State $\symNumReg \gets \left\Vert\boldsymbol{\symCountVariate}\right\Vert_1$
-\State $\symZ\gets \cdot \symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$
+\State $\symZ\gets\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$
 \Comment{\parbox[t]{.5\linewidth}{alternatively, take $\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$ from precalculated lookup table}}
@@
 \Comment{\parbox[t]{.5\linewidth}{alternatively, take $\symNumReg\cdot\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$ from precalculated lookup table}}
-\State \Return$\symNumReg^2 / \left(2\log(2)\symZ\right)$
+\State \Return$\symAlpha_\infty\symNumReg^2 / \symZ$
 \EndFunction
@@
 In order to verify the new estimation algorithm, we generated 10\,000 HyperLogLog sketches and inserted up to 50 billion unique elements.
-Assuming a strong universal hash function, element hash values can be mocked by random numbers.
-For the following results we used the Mersenne Twister random number generator with 19937 bit state size from the C++ standard library.
+Assuming a uniform hash function, element hash values can be mocked by random numbers.
+For the following results we used the Mersenne Twister random number generator with 19\,937 bit state size from the C++ standard library.
 
@@
 \includesvg[width=1\textwidth]{raw_corrected_estimate_12_20}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
+\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
 \label{fig:raw_corrected_estimation_error_12_20}
@@
 \includesvg[width=1\textwidth]{raw_corrected_estimate_8_24}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
+\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
 \label{fig:raw_corrected_estimation_error_8_24}
@@
 \includesvg[width=1\textwidth]{raw_corrected_estimate_16_16}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
+\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
 \label{fig:raw_corrected_estimation_error_16_16}
@@
 \includesvg[width=1\textwidth]{raw_corrected_estimate_22_10}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
+\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
 \label{fig:raw_corrected_estimation_error_22_10}
@@
 replacing the 32-bit hash function by a 64-bit hash function.
-\cref{fig:raw_corrected_estimation_error_12_52} shows the relative error for such a HyperLogLog configuration with precision  $\symPrecision=12$.
-In this case, in order to use all the 64 bits of the hash value, $\symRegRange$ must be chosen to be equal to $64 - \symPrecision = 52$.
-As a consequence, in order to represent the maximum possible register value $\symRegRange+1 = 53$, 6 bits are needed for each register.
+\cref{fig:raw_corrected_estimation_error_12_52} shows the relative error for such a HyperLogLog configuration with parameters $\symPrecision=12$ and $\symRegRange=52$.
 The doubled hash value size shifts the maximum trackable cardinality value towards $2^{64}$.
@@
 \includesvg[width=1\textwidth]{raw_corrected_estimate_12_52}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
+\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
 \label{fig:raw_corrected_estimation_error_12_52}
@@
 \includesvg[width=1\textwidth]{raw_corrected_estimate_12_14}
-\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
+\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
 \label{fig:raw_corrected_estimation_error_12_14}
@@
 A theoretical lower bound for this processing time can be derived using the maximum memory bandwidth of the CPU, which is 21\,GB/s for an Intel Core i5-2500K.
-For precision $\symPrecision=12$ there are 4096 registers, each of them requires at least 5 bits.
-Hence, the total data size of the HyperLogLog sketch is 2.5\,kB minimum and transfer time from main memory to CPU will be at least 120\,ns.
+If we consider a HyperLogLog sketch with precision $\symPrecision=12$ which uses 5 bits per register, the total data size of the HyperLogLog sketch is 2.5\,kB minimum.
+Consequently, the transfer time from main memory to CPU will be at least 120\,ns.
 Having this value in mind, the presented numbers for estimating the cardinality from the multiplicity vector are quite satisfying.
@@
 Furthermore, we will demonstrate that consequent application of the maximum likelihood method reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values.
-The history of the HyperLogLog algorithm shows that the raw estimate \eqref{alg:estimate_original} was first found after several attempts using the geometric mean \cite{Flajolet2007, Durand2003}.
+The history of the HyperLogLog algorithm shows that the raw estimator \eqref{equ:raw_estimator} was first found after several attempts using the geometric mean \cite{Flajolet2007, Durand2003}.
 
@@
 The function is non-positive at 0 since $\symFunc(0)=\symCountVariate_0-\symNumReg\leq 0$ and, in case $\symCountVariate_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing.
-$\symCountVariate_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $(\symRegRange+1)$, for which the maximum likelihood estimate would be infinity.
+$\symCountVariate_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $\symRegRange+1$, for which the maximum likelihood estimate would be infinite.
 
@@
 Keeping $\symX$ fixed during an update decreases $\symFunc(\symX)$.
-As a consequence, since $\symFunc(\symX)$ is increasing, the new root and hence the estimate must be larger than before the update.
-
-Note that \eqref{equ:funcdef} can be solved analytically for the special case $\symRegRange=0$ which corresponds to the already mentioned linear counting algorithm.
-In this case, the maximum likelihood method under the Poisson model directly leads to the linear counting  estimator presented in \cite{Whang1990} and which was also used for small range estimation \eqref{equ:small_range_estimate}.
+As a consequence, since $\symFunc(\symX)$ is increasing, the new root and hence the estimate must be larger than prior the update.
+
+For the special case $\symRegRange=0$, which corresponds to the already mentioned linear counting algorithm, \eqref{equ:funcdef} can be solved analytically.
+In this case, the maximum likelihood method under the Poisson model leads directly to the linear counting estimator \eqref{equ:linear_counting_estimator}.
 Due to this fact we could expect that maximum likelihood estimation under the Poisson model also works well for the more general HyperLogLog case.
@@
 \symX\cdot
-\left(
+\frac{
 \sum_{\symRegVal=1}^{\symRegRange}
@@
 \frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
-\right)\right)
+}
+{\symNumReg-\symCountVariate_0}\right)
 -
@@
 -
-\left(\symNumReg-\symCountVariate_0\right),
-\end{equation}
-for $\symFunc$ can be found, where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$ and
+\left(\symNumReg-\symCountVariate_0\right)
+\end{equation}
+can be found, where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$ and
 $\symRegValVariate_\text{max} := \max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$.
@@
 \end{equation}
-If the HyperLogLog sketch is in the intermediate range, where $\symCountVariate_0=\symCountVariate_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimator \eqref{equ:intermediate_estimate}.
+If the HyperLogLog sketch is in the intermediate range, where $\symCountVariate_0=\symCountVariate_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimator \eqref{equ:raw_estimator}.
 Hence, the maximum likelihood method leads directly to the harmonic mean that is used by the raw estimator.
@@
 \subsection{Computation of the maximum likelihood estimate}
-Since $\symFunc$ is concave and increasing, both, Newton-Raphson iteration and the secant method, converge to the root, if the function is negative for the starting points.
+\label{sec:comp_ml_estimate}
+Since $\symFunc$ is concave and increasing, both, Newton-Raphson iteration and the secant method, will converge to the root, if the function is negative for the starting points.
 In the following we start from the secant method to derive the new cardinality estimation algorithm.
 Even though the secant method has the disadvantage of slower convergence, a single iteration is simpler to calculate as it does not require the evaluation of the first derivative.
-
 An iteration step of the secant method can be written as
@@
 \end{equation}
-If $\symX_0 = 0$ with $\symFunc(\symX_0)=-\left(\symNumReg-\symCountVariate_0\right)$, and $\symX_1$ is equal to one of the derived lower bounds \eqref{equ:strong_lower_bound} or \eqref{equ:weak_lower_bound}, the sequence $\lbrace\symX_\symIndexI\rbrace$ is montone increasing.
+If $\symX_0 = 0$ with $\symFunc(\symX_0)=-\left(\symNumReg-\symCountVariate_0\right)$, and $\symX_1$ is equal to one of the derived lower bounds \eqref{equ:strong_lower_bound} or \eqref{equ:weak_lower_bound}, the sequence $(\symX_0, \symX_1, \symX_2, \ldots)$ is montone increasing.
 Using the definitions
@@
 Since the expected statistical errror for the HyperLogLog data structure scales according to $\frac{1}{\sqrt{\symNumReg}}$ \cite{Flajolet2007}, it makes sense to choose $\symStopDelta = \frac{\symStopEpsilon}{\sqrt{\symNumReg}}$ with some constant $\symStopEpsilon$.
-For all results that will follow in \cref{sec:maximum_likelihood_estimation_error} we have used $\symStopEpsilon = 10^{-2}$.
+For all results that will be presented in \cref{sec:maximum_likelihood_estimation_error} we have used $\symStopEpsilon = 10^{-2}$.
 
@@
 If we denote $\symRegValVariate_\text{min}:=\min\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$ and $\symRegValVariate_\text{max}:=\max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$,  it is sufficient to loop over all indices in the range $[\symRegValVariate_\text{min}, \symRegValVariate_\text{max}]$.
-\item The coefficient of the linear term $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}$ can be precalculated and reused for all function evaluations.
+\item The sum $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}$ in \eqref{equ:funcprime} can be precalculated and reused for all function evaluations.
 \item Many programming languages allow the efficient multiplication and division by any integral power of 2 using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
@@
 A straightforward calculation using \eqref{equ:helper} is very expensive because of the exponential function.
-However, if we already know $\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
+However, if we know $\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
 \begin{equation}
 \label{equ:helper_recursion1}
-\symHelper\!\left(2\symX\right) = \frac{\symX+2\symHelper\!\left(\symX\right)\left(1-\symHelper\!\left(\symX\right)\right)}{\symX+2\left(1-\symHelper\!\left(\symX\right)\right)}
-\end{equation}
-or
-\begin{equation}
-\label{equ:helper_recursion2}
-\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right) = \frac{\frac{\symX}{2^{\symRegVal+2}}+\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}{\frac{\symX}{2^{\symRegVal+2}}+\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}
-\end{equation}
-Note, this recursive formula is stable in a sense that the relative error of $\symHelper\!\left(2\symX\right)$ is smaller than that of $\symHelper\!\left(\symX\right)$ as shown in \cref{app:helper_stable}.
-
-\item If $\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ is smaller than 0.5, the function $\symHelper(\symX)$ can be well approximated by a Taylor series around $\symX=0$
+\symHelper(4\symX) = \frac{\symX+\symHelper(2\symX)\left(1-\symHelper(2\symX)\right)}{\symX+\left(1-\symHelper(2\symX)\right)}.
+\end{equation}
+%or
+%\begin{equation}
+%\label{equ:helper_recursion2}
+%\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right) = \frac{\frac{\symX}{2^{\symRegVal+2}}+\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}{\frac{\symX}{2^{\symRegVal+2}}+\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}
+%\end{equation}
+This recursive formula is stable in a sense that the relative error of $\symHelper(4\symX)$ is smaller than that of $\symHelper(2\symX)$ as shown in \cref{app:helper_stable}.
+
+\item If $\symX$ is smaller than 0.5, the function $\symHelper(\symX)$ can be well approximated by a Taylor series around $\symX=0$
 \begin{equation}
@@
 \end{equation}
-In fact, $\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)\leq 0.5$ is almost always fulfilled as long as registers are not saturated.
-Using \eqref{equ:strong_upper_bound} it is straightforward to see that $\frac{\symX}{2^{\symRegValVariate'_\text{max}}} \leq \log 2 \approx 0.693$, if $\symCountVariate_{\symRegRange+1}=0$.
-In case $\frac{\symX}{2^{\symRegValVariate'_\text{max}}} > 0.5$, the value $\frac{\symX}{2^{\symE+1}}$ is taken instead, where $\symE$ is the exponent of the floating point representation of $\symX$, $\symE = 1+\lfloor\log_2(\symX)\rfloor$.
-By definition, $\frac{\symX}{2^{\symE+1}}\leq 0.5$ which allows using the Taylor series approximation.
-$\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ is finally obtained after $\symE + 1 -\symRegValVariate'_\text{max}$ iterations using \eqref{equ:helper_recursion2}.
+The smallest argument for which $\symHelper$ needs to be evaluated is $\frac{\symX}{2^{\symRegValVariate'_\text{max}}}$. If 
+$\symCountVariate_{\symRegRange+1}=0$, $\frac{\symX}{2^{\symRegValVariate'_\text{max}}} \leq \log 2 \approx 0.693$ holds, which is straightforward to see using \eqref{equ:strong_upper_bound}.
+In practice, $\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\leq0.5$ is satisfied most of the time as long as only a few registers are saturated ($\symCountVariate_{\symRegRange+1}\ll \symNumReg$).
+In case $\frac{\symX}{2^{\symRegValVariate'_\text{max}}} > 0.5$, $\symHelper\!\left(\frac{\symX}{2^{\symKappa}}\right)$ is calculated instead with $\symKappa = 2+\lfloor\log_2(\symX)\rfloor$.
+By definition, $\frac{\symX}{2^{\symKappa}}\leq 0.5$ which allows using the Taylor series approximation.
+$\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ is finally obtained after $\symKappa -\symRegValVariate'_\text{max}$ iterations using \eqref{equ:helper_recursion1}.
 \end{itemize}
@@
 \EndFor
+\State\Comment here $\symZ=\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$
 \State $\symNumReg \gets \symNumReg' + \symCountVariate_0$ 
@@
 \State $\symA\gets \symZ + \symCountVariate_0$
+\Comment $\symA = \sum_{\symRegVal=0}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$
 \State $\symB\gets \symZ + 
 \symCountVariate_{\symRegRange+1}\cdot 2^{-\symRegRange}$
+\Comment $\symB = \sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}$
 \If{$\symB \leq 1.5\cdot\symA$}
@@
 \State $\Delta\symX \gets \symX$
-\While{$\Delta\symX > \symX\cdot\symError$} \Comment secant method iteration, $\symStopEpsilon = 10^{-2}$
-\State $\symE \gets 1+\lfloor\log_2(\symX)\rfloor$
+\While{$\Delta\symX > \symX\cdot\symError$} \Comment secant method iteration, $\symStopEpsilon = 10^{-2}$ (see \cref{sec:comp_ml_estimate})
+\State $\symKappa \gets 2+\lfloor\log_2(\symX)\rfloor$
 \State $\symXNormalized \gets
-\symX \cdot 2^{-\max(\symRegValVariate'_\text{max}+1, \symE+2)}$
+\symX \cdot2^{-\max(\symRegValVariate'_\text{max}, \symKappa)-1}$
 \Comment $\symXNormalized \in [0, 0.25]$
@@
 \Comment Taylor approximation \eqref{equ:taylor}
-\For{$\symRegVal\gets \symE,\symRegValVariate'_\text{max}$}
+\For{$\symRegVal\gets (\symKappa - 1),\symRegValVariate'_\text{max}$}
 \State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
-\Comment calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion2}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
+\Comment calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
 \State $\symXNormalized \gets 2\symXNormalized$
@@
 \State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
-\Comment calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion2}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
+\Comment calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
 \State $\symFuncPrime\gets \symFuncPrime + \symCountVariate_\symRegVal\cdot\symHelper$
@@
 In order to investigate the estimation error for the maximum likelihood estimation algorithm, we investigated the same HyperLogLog configurations as in \cref{sec:corrected_raw_estimation_error} for the corrected raw estimation algorithm.
-\cref{fig:max_likelihood_estimation_error_12_20}, \cref{fig:max_likelihood_estimation_error_8_24}, \cref{fig:max_likelihood_estimation_error_16_16},\cref{fig:max_likelihood_estimation_error_22_10},\cref{fig:max_likelihood_estimation_error_12_52}, and \cref{fig:max_likelihood_estimation_error_12_14} show very similar results for various HyperLogLog parameters.
+\cref{fig:max_likelihood_estimation_error_12_20}, \cref{fig:max_likelihood_estimation_error_8_24},  \cref{fig:max_likelihood_estimation_error_16_16}, \cref{fig:max_likelihood_estimation_error_22_10}, \cref{fig:max_likelihood_estimation_error_12_52}, and \cref{fig:max_likelihood_estimation_error_12_14} show very similar results for various HyperLogLog parameters.
 
@@
 The average computation time for the maximum likelihood algorithm shows a different behavior than for the corrected raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}).
-As can be seen, the average execution time is larger for most cardinalities, but nevertheless it never exceeds 700\,ns which is still fast enough for many applications.
+As can be seen, the average execution time is larger for most cardinalities, but nevertheless, it never exceeds 700\,ns which is still fast enough for many applications.
 
@@
 \section{Cardinality estimation of set intersections and differences}
-Assume two different sets $\symSetS_1$ and $\symSetS_2$ that have been recorded by two HyperLogLog sketches with same parameters $(\symPrecision,\symRegRange)$.
-Given the corresponding register values $\boldsymbol{\symRegValVariate}_1$ and
-$\boldsymbol{\symRegValVariate}_2$, we attempt to estimate the cardinalities of the pair-wise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$.
-Motivated by the good results we have obtained for cardinality estimation of a single HyperLogLog sketch, we want to get these estimates using the maximum likelihood method applied to the joint probability distribution of $\boldsymbol{\symRegValVariate}_1$ and $\boldsymbol{\symRegValVariate}_2$.
-
+
+While the union of two sets that are represented by HyperLogLog sketches can be straightforwardly computed \eqref{equ:hll_union}, the computation of cardinalities of other set operations like intersections and differences is more challenging.
+A common approach to calculate the cardinality of intersections is the inclusion-exclusion principle
+\begin{equation}
+\label{equ:inclusion-exclusion}
+\left\vert\symSetS_1 \cap \symSetS_2\right\vert
+=  
+\left\vert\symSetS_1\right\vert  
++
+\left\vert\symSetS_2\right\vert
+-
+\left\vert\symSetS_1 \cup \symSetS_2\right\vert.
+\end{equation}
+Unfortunately, this approach can be very inaccurate. Due to estimation errors of HyperLogLog sketches the result of \eqref{equ:inclusion-exclusion} can be even negative.
+
+Motivated by the good results we have obtained for a single HyperLogLog sketch using the maximum likelihood method in combination with the Poisson approximation, we are tempted to also use it for the estimation of set operations.
+Assume two given  HyperLogLog sketches with register values $\boldsymbol{\symRegValVariate}_1$ and $\boldsymbol{\symRegValVariate}_2$ representing the sets $\symSetS_1$ and $\symSetS_2$, respectively.
+The goal is to find estimates for the cardinalities of the pair-wise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$.
+
+The Poisson approximation allows us to assume that the elements of $\symSetA$ are inserted into the HyperLogLog sketch representing $\symSetS_1$ at rate $\symPoissonRate_\symSetA$.
+Similarly, the elements of $\symSetB$ are inserted into the HyperLogLog sketch representing $\symSetS_2$ at rate $\symPoissonRate_\symSetB$.
+Furthermore, we assume that elements of $\symSetX$ are inserted into both HyperLogLog sketches simultaneously at rate $\symPoissonRate_\symSetX$.
+The goal is now to find estimates, $\symPoissonRateEstimate_\symSetA$, $\symPoissonRateEstimate_\symSetB$, and $\symPoissonRateEstimate_\symSetX$ for the rates which could also be good estimates for the wanted cardinalities of $\symSetA$, $\symSetB$, and $\symSetX$.
+
+\subsection{Joint log-likelihood funtion}
+In order to get maximum-likelihood estimators for $\symPoissonRateEstimate_\symSetA$, $\symPoissonRateEstimate_\symSetB$, and $\symPoissonRateEstimate_\symSetX$ we need to derive the joint probability distribution of two HyperLogLog sketches.
 Under the Poisson model the register values are independent and identically distributed.
@@
 
-The HyperLogLog sketch that represents $\symSetS_1$ could have also been obtained by constructing two HyperLogLog sketches from sets $\symSetA$ and $\symSetX$ and by merging both by taking for each register the maximum value of both sketches.
+The HyperLogLog sketch that represents $\symSetS_1$ can be thought to be constructed from two HyperLogLog sketches representing $\symSetA$ and $\symSetX$ and merging both according to \eqref{equ:hll_union}.
 Analogously, the HyperLogLog sketch for $\symSetS_2$ could have been obtained from sketches for $\symSetB$ and $\symSetX$.
-Let us consider the register values $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ at a certain position of the HyperLogLog sketches for $\symSetA$, $\symSetB$, and $\symSetX$, respectively.
+Let  $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ be the value of the considered register in the HyperLogLog sketch representing $\symSetA$, $\symSetB$, and $\symSetX$, respectively.
 The corresponding values in sketches for $\symSetS_1$ and $\symSetS_2$ are given by
@@
 ,\quad
-\symRegValVariate_2 = \max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right)
+\symRegValVariate_2 = \max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right).
 \end{equation}
@@
 \eqref{equ:register_value_distribution}.
-If we assume that elements are added to $\symSetA$, $\symSetB$, and $\symSetX$ at rates $\symPoissonRate_\symSetXSuffix$, $\symPoissonRate_\symSetASuffix$, and $\symPoissonRate_\symSetBSuffix$, respectively, the probability that a certain register has a value less than or equal to $\symRegVal_1$ in the first HyperLogLog sketch and simultaneously a value less than or equal to $\symRegVal_2$ in the second one can be written as
+If we take into account that elements are added to $\symSetA$, $\symSetB$, and $\symSetX$ at rates $\symPoissonRate_\symSetXSuffix$, $\symPoissonRate_\symSetASuffix$, and $\symPoissonRate_\symSetBSuffix$, respectively, the probability that a certain register has a value less than or equal to $\symRegVal_1$ in the first HyperLogLog sketch and simultaneously a value less than or equal to $\symRegVal_2$ in the second one can be written as
 \begin{equation}
@@
 &
-\symRegRange < \symRegVal_1=\symRegVal_2
+\symRegRange < \symRegVal_1 \wedge \symRegRange < \symRegVal_2
 \end{cases}
@@
 
-Since under the Poisson model, the values for different registers are independent we are now able to write the joint probability mass function for the joint state of both HyperLogLog sketches
+Since under the Poisson model, the values for different registers are independent we are now able to write the joint probability mass function for all registers in both HyperLogLog sketches
 \begin{equation}
@@
 =
-\prod_{\symRegVal_1 = 0}^{\symRegRange+1}
-\prod_{\symRegVal_2 = 0}^{\symRegRange+1}
-\symProbabilityMass(\symRegVal_1,\symRegVal_2)
-^{\symCount_{\symRegVal_1\symRegVal_2}}.
-\end{equation}
-Here we have used the multiplicity matrix $\symCountMatrix = \left(\symCount_{\symRegVal_1\symRegVal_2}\right)_{\symRegVal_1\symRegVal_2\in[0,\symRegRange+1]}$ defined by
-\begin{equation}
-\symCount_{\symRegVal_1\symRegVal_2}
-:=
-\left|\left\lbrace
-\left(\symIndexI,\symIndexJ\right) \vert
-\symRegVal_{\symIndexI,1} = \symRegVal_1
-\wedge
-\symRegVal_{\symIndexJ,2} = \symRegVal_2
-\right\rbrace\right|
-\end{equation}
-As in \cref{sec:cardinality_estimation} where we found that the multiplicity vector $\boldsymbol{\symCount}$ of a sketch is a sufficient statistic for the cardinality, the multiplicity matrix of two sketches is a sufficient statistic for the cardinalities of $\symSetX$, $\symSetA$, and $\symSetB$.
+\prod_{\symIndexI = 1}^{\symNumReg}
+\prod_{\symIndexJ = 1}^{\symNumReg}
+\symProbabilityMass(\symRegVal_{1\symIndexI},\symRegVal_{2\symIndexJ}).
+\end{equation}
 
@@
 \vert
-\boldsymbol{\symRegVal}_1,
-\boldsymbol{\symRegVal}_2
+\boldsymbol{\symRegValVariate}_1,
+\boldsymbol{\symRegValVariate}_2
 )
 =
-\sum_{\symRegVal_1 = 0}^{\symRegRange+1}
-\sum_{\symRegVal_2 = 0}^{\symRegRange+1}
-\symCount_{\symRegVal_1\symRegVal_2}
-\log(\symProbabilityMass(\symRegVal_1,\symRegVal_2))
-\end{equation}
-Insertion of \eqref{equ:joint_log_pmf_single_register} results in
+\sum_{\symIndexI = 1}^{\symNumReg}
+\sum_{\symIndexJ = 1}^{\symNumReg}
+\log(\symProbabilityMass(\symRegValVariate_{1\symIndexI},\symRegValVariate_{2\symIndexJ}))
+\end{equation}
+Insertion of \eqref{equ:joint_log_pmf_single_register} results in the two HyperLogLog case analog of \eqref{equ:log_likelihood_single}
 \begin{align}
@@
 \vert
-\boldsymbol{\symRegVal}_1,
-\boldsymbol{\symRegVal}_2
+\boldsymbol{\symRegValVariate}_1,
+\boldsymbol{\symRegValVariate}_2
 )
@@
 \frac{
-  \symCount^\leftarrow_\symRegVal+
-  \symCount_{\symRegVal\symRegVal}+
-  \symCount^\rightarrow_\symRegVal
+  \symCountVariate^{<}_{1\symRegVal}+
+  \symCountVariate^{=}_{\symRegVal}+
+  \symCountVariate^{>}_{1\symRegVal}
 }{2^{\symRegVal}}
@@
 \frac{
-  \symCount^\uparrow_\symRegVal+
-  \symCount_{\symRegVal\symRegVal}+
-  \symCount^\downarrow_\symRegVal
+  \symCountVariate^{<}_{2\symRegVal}+
+  \symCountVariate^{=}_{\symRegVal}+
+  \symCountVariate^{>}_{2\symRegVal}
 }{2^{\symRegVal}}
@@
 \frac{
-  \symCount^\rightarrow_\symRegVal+
-  \symCount_{\symRegVal\symRegVal}+
-  \symCount^\downarrow_\symRegVal
+  \symCountVariate^{<}_{1\symRegVal}+
+  \symCountVariate^{=}_{\symRegVal}+
+  \symCountVariate^{<}_{2\symRegVal}
 }{2^{\symRegVal}}
@@
 \log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
-\symCount_\symRegVal^\rightarrow
+\symCountVariate^{<}_{1\symRegVal}
 +
 \log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
-\symCount_\symRegVal^\leftarrow
+\symCountVariate^{>}_{1\symRegVal}
 \nonumber\\
@@
 \log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
-\symCount_\symRegVal^\downarrow
+\symCountVariate^{<}_{2\symRegVal}
 +
 \log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
-\symCount_\symRegVal^\uparrow
+\symCountVariate^{>}_{2\symRegVal}
 \nonumber\\
@@
 \right)
-\symCount_{\symRegVal\symRegVal}
+\symCountVariate^{=}_{\symRegVal}
 \end{align}
-which is the two HyperLogLog case analog of \eqref{equ:log_likelihood_single}.
-The constants $\symCount^\uparrow_\symRegVal:=\sum_{\symIndexI=0}^{\symRegVal-1} \symCount_{\symIndexI\symRegVal}$, $\symCount^\rightarrow_\symRegVal:=\sum_{\symIndexI=\symRegVal+1}^{\symRegRange+1} \symCount_{\symRegVal\symIndexI}$, $\symCount^\downarrow_\symRegVal:=\sum_{\symIndexI=\symRegVal+1}^{\symRegRange+1} \symCount_{\symIndexI\symRegVal}$, and $\symCount^\leftarrow_\symRegVal:=\sum_{\symIndexI=0}^{\symRegVal-1} \symCount_{\symRegVal\symIndexI}$ correspond to sums within the multiplicity matrix, which are obtained by aggregating all elements that are in up, right, down, and left directions relative to the $\symRegVal$-th diagonal entry  $\symCount_{\symRegVal\symRegVal}$, respectively.
+where the constant $\symCountVariate^{>}_{1\symRegVal}$,
+$\symCountVariate^{<}_{1\symRegVal}$,
+$\symCountVariate^{>}_{2\symRegVal}$,
+$\symCountVariate^{<}_{2\symRegVal}$,
+and $\symCountVariate^{=}_{\symRegVal}$ are defined as
+\begin{equation}
+\begin{aligned}
+\symCountVariate^{>}_{1\symRegVal}&:=\left|\lbrace\symIndexI\vert\symRegVal=\symRegValVariate_{1\symIndexI}>
+\symRegValVariate_{2\symIndexI}\rbrace\right|,
+\\
+\symCountVariate^{<}_{1\symRegVal}&:=\left|\lbrace\symIndexI\vert\symRegVal=\symRegValVariate_{1\symIndexI}<
+\symRegValVariate_{2\symIndexI}\rbrace\right|,
+\\
+\symCountVariate^{>}_{2\symRegVal}&:=\left|\lbrace\symIndexI\vert\symRegVal=\symRegValVariate_{2\symIndexI}>
+\symRegValVariate_{1\symIndexI}\rbrace\right|,
+\\
+\symCountVariate^{<}_{2\symRegVal}&:=\left|\lbrace\symIndexI\vert\symRegVal=\symRegValVariate_{2\symIndexI}<
+\symRegValVariate_{1\symIndexI}\rbrace\right|,
+\\
+\symCountVariate^{=}_{\symRegVal}&:=\left|\lbrace\symIndexI\vert\symRegVal=\symRegValVariate_{1\symIndexI}=
+\symRegValVariate_{2\symRegVal}\rbrace\right|.
+\end{aligned}
+\end{equation}
 
 The log-likelihood function \eqref{equ:log_likelihood_pair} does not always have a strict global maximum point.
-For example, in case $\symCountMatrix$ is a strict lower triangular matrix which corresponds to the case that each register of the first HyperLogLog sketch is larger than the corresponding value in the second HyperLogLog sketch, the function can be rewritten as sum of two functions, one dependent on $\symPoissonRate_\symSetASuffix$ and the other dependent on $(\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix)$.
+For example, if all register values of the first HyperLogLog sketch are larger than the corresponding values in the second HyperLogLog sketch, the function can be rewritten as sum of two functions, one dependent on $\symPoissonRate_\symSetASuffix$ and the other dependent on $(\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix)$.
 The maximum is obtained, if $\symPoissonRate_\symSetASuffix = \symPoissonRateEstimate_1$ and $\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix = \symPoissonRateEstimate_2$.
@@
 In this case \eqref{equ:log_likelihood_pair} is separable into the sum of two log-likelihood functions that depend on $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$, respectively, and that follow \eqref{equ:log_likelihood_single}.
-Hence, the joint maximum likelihood estimation of $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$ gives the same results as estimating them independently by maximizing \eqref{equ:log_likelihood_single}.
+Hence, the joint maximum likelihood estimation of $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$ gives the same results as estimating them independently by maximization of \eqref{equ:log_likelihood_single}.
 
@@
 \frac{1}{e^{\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}-1}
-\frac{\symCount^\rightarrow_\symRegVal}{2^{\min\left(\symRegVal,\symRegRange\right)}}
+\frac{\symCountVariate^{<}_{1\symRegVal}}{2^{\min\left(\symRegVal,\symRegRange\right)}}
 \nonumber\\
@@
 \frac{1}{e^{\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}-1}
-\frac{\symCount^\leftarrow_\symRegVal}{2^{\min\left(\symRegVal,\symRegRange\right)}}
+\frac{\symCountVariate^{>}_{1\symRegVal}}{2^{\min\left(\symRegVal,\symRegRange\right)}}
 \nonumber\\
@@
 }
-\frac{\symCount_{\symRegVal\symRegVal}
+\frac{\symCountVariate^{=}_{\symRegVal}
 }{2^{\min\left(\symRegVal,\symRegRange\right)}}
@@
 \frac{1}{e^{\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}-1}
-\frac{\symCount^\downarrow_\symRegVal}{2^{\min\left(\symRegVal,\symRegRange\right)}}
+\frac{\symCountVariate^{<}_{2\symRegVal}}{2^{\min\left(\symRegVal,\symRegRange\right)}}
 \nonumber\\
@@
 \label{app:helper_stable}
-In order to investigate the error propagation of a single recursion step using \eqref{equ:helper_recursion1} we define $\symHelper_1 := \symHelper(\symX)$ and $\symHelper_2 := \symHelper(2\symX)$.
+In order to investigate the error propagation of a single recursion step using \eqref{equ:helper_recursion1} we define $\symHelper_1 := \symHelper(2\symX)$ and $\symHelper_2 := \symHelper(4\symX)$.
 The recursion formula simplifies to
@@
 \label{equ:h2}
-\symHelper_2 = \frac{\symX+2\symHelper_1(1-\symHelper_1)}{\symX+2(1-\symHelper_1)}.
+\symHelper_2 = \frac{\symX+\symHelper_1(1-\symHelper_1)}{\symX+(1-\symHelper_1)}.
 \end{equation}
@@
 \tilde{\symHelper}_2 = 
-\frac{\symX+2\tilde{\symHelper}_1(1-\tilde{\symHelper}_1)}{\symX+2(1-\tilde{\symHelper}_1)}
+\frac{\symX+\tilde{\symHelper}_1(1-\tilde{\symHelper}_1)}{\symX+(1-\tilde{\symHelper}_1)}
 \end{equation}
@@
 \end{equation}
-Putting \eqref{equ:h2} and \eqref{equ:h2_approx} into \eqref{equ:h2_relative_error} and using the first-order approximations 
-\begin{equation}
-\frac{\symX+2\tilde{\symHelper}_1(1-\tilde{\symHelper}_1)}
-{\symX+2\symHelper_1(1-\symHelper_1)}
-=
-1
-+
+Combination of \eqref{equ:h2}, \eqref{equ:h2_approx}, and \eqref{equ:h2_relative_error} yields for its absolute value
+\begin{equation}
+\left|\symError_2\right|
+=
+\left|\symError_1\right|
+\frac{
+\left|
+\frac{\symHelper_1\left(1-2\symHelper_1\right)}{\symX+\symHelper_1\left(1-\symHelper_1\right)}
++
+\frac{\symHelper_1}{\symX+1-\symHelper_1}
+-
 \symError_1
-\frac{2\symHelper_1\left(1-2\symHelper_1\right)}
-{\symX + 2\symHelper_1\left(1-\symHelper_1\right)}
-+\symBigO(\symError_1^2)
-\end{equation}
+\frac{\symHelper_1^2}{\symX+\symHelper_1\left(1-\symHelper_1\right)}
+\right|
+}{
+\left|
+1-
+\symError_1
+\frac{\symHelper_1}{\symX+1-\symHelper_1}
+\right|
+}.
+\end{equation}
+The triangle inequality lead to
+\begin{equation}
+\left|\symError_2\right|
+\leq
+\left|\symError_1\right|
+\frac{
+\left|
+\frac{\symHelper_1\left(1-2\symHelper_1\right)}{\symX+\symHelper_1\left(1-\symHelper_1\right)}
++
+\frac{\symHelper_1}{\symX+1-\symHelper_1}
+\right|
++
+\left|
+\symError_1
+\right|
+\frac{\symHelper_1^2}{\symX+\symHelper_1\left(1-\symHelper_1\right)}
+}{
+\left|
+1-
+\symError_1
+\frac{\symHelper_1}{\symX+1-\symHelper_1}
+\right|
+}
+\end{equation}
+
+By numerical means it is easy to show that the inequalities $\left|
+\frac{\symHelper_1\left(1-2\symHelper_1\right)}{\symX+\symHelper_1\left(1-\symHelper_1\right)}
++
+\frac{\symHelper_1}{\symX+1-\symHelper_1}
+\right|\leq\input{max_errror_propagation_factor1.txt}
+$,
+$\frac{\symHelper_1^2}{\symX+\symHelper_1\left(1-\symHelper_1\right)}
+\leq\input{max_errror_propagation_factor2.txt}$,
 and
-\begin{equation}
-\frac{\symX+2(1-\symHelper_1)}
-{\symX+2(1-\tilde{\symHelper}_1)}
-=
-1
-+
-\symError_1
-\frac{2\symHelper_1}{x+2\left(1-\symHelper_1\right)}
-+\symBigO(\symError_1^2),
-\end{equation}
-we obtain
-\begin{equation}
-\symError_2
-=
-\symError_1
-\left(
-\frac{2\symHelper_1\left(1-2\symHelper_1\right)}
-{\symX + 2\symHelper_1\left(1-\symHelper_1\right)}
-+
-\frac{2\symHelper_1}{x+2\left(1-\symHelper_1\right)}
-\right)
-+\symBigO(\symError_1^2)
-\end{equation}
-By numerical means it is easy to show that
-\begin{equation}
-0\leq\frac{2\symHelper(\symX)\left(1-2\symHelper(\symX)\right)}
-{\symX + 2\symHelper(\symX)\left(1-\symHelper(\symX)\right)}
-+
-\frac{2\symHelper(\symX)}{x+2\left(1-\symHelper(\symX)\right)} \leq \input{max_errror_propagation_factor.txt}
-\end{equation}
-holds for all $\symX \geq 0$. Therefore,
-\begin{equation}
-|\symError_2|\leq \input{max_errror_propagation_factor.txt}\  |\symError_1| + \symBigO(\symError_1^2)
+$\frac{\symHelper_1}{\symX+1-\symHelper_1}\leq\input{max_errror_propagation_factor3.txt}$ hold for all $\symX\geq0$.
+Therefore, if we additionally assume, for example, $\left|\symError_1\right|\leq0.1$, we get
+\begin{equation}
+\left|\symError_2\right|
+\leq
+\left|\symError_1\right|
+\frac{\input{max_errror_propagation_factor1.txt} + 0.1\cdot\input{max_errror_propagation_factor2.txt}}{1-0.1\cdot\input{max_errror_propagation_factor3.txt}}
+\leq
+\left|\symError_1\right|
+\cdot
+\input{max_errror_propagation_factor4.txt}
 \end{equation}
@@
 \label{equ:max_likelihood_2}
-\symXEstimate\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
-\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\!\left(\frac{\symXEstimate}{2^\symRegVal}\right)
-+
-\symCount_{\symRegRange+1}\symHelper\!\left(\frac{\symXEstimate}{2^\symRegRange}\right)
-=
-\symNumReg-\symCount_0
+\symXEstimate\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
+\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symXEstimate}{2^\symRegVal}\right)
++
+\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symXEstimate}{2^\symRegRange}\right)
+=
+\symNumReg-\symCountVariate_0
 \end{equation}
@@
 \label{equ:ml_equation}
-\symXEstimate\left(1+\symError_\symX\right)\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
-\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelperApprox\!\left(\frac{\symXEstimate\left(1+\symError_\symX\right)}{2^\symRegVal}\right)
-+
-\symCount_{\symRegRange+1}
+\symXEstimate\left(1+\symError_\symX\right)\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
+\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelperApprox\!\left(\frac{\symXEstimate\left(1+\symError_\symX\right)}{2^\symRegVal}\right)
++
+\symCountVariate_{\symRegRange+1}
 \symHelperApprox\!\left(
@@
 =
-\symNumReg-\symCount_0
+\symNumReg-\symCountVariate_0
 \end{equation}
@@
 \begin{multline}
-\symXEstimate\left(1+\symError_\symX\right)\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
+\symXEstimate\left(1+\symError_\symX\right)\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
 \left(
@@
 \left(
-\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal
+\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal
 \symHelper\!\left(\frac{\symXEstimate\left(1+\symError_\symX\right)
@@
 +
-\symCount_{\symRegRange+1}
+\symCountVariate_{\symRegRange+1}
 \symHelper\!\left(
@@
 =
-\symNumReg-\symCount_0
+\symNumReg-\symCountVariate_0
 \end{multline}
@@
 \label{equ:appendix3}
-\symXEstimate\left(1+\symError_\symX\right)\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
+\symXEstimate\left(1+\symError_\symX\right)\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
 \left(
@@
 \sum_{\symRegVal=1}^\symRegRange 
-\symCount_\symRegVal
+\symCountVariate_\symRegVal
 \symHelper\!\left(\frac{\symXEstimate
@@
 +
-\frac{\symCount_\symRegVal}{2^\symRegVal}
+\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
 \symXEstimate\symError_\symX
@@
 \left(
-\symCount_{\symRegRange+1}
+\symCountVariate_{\symRegRange+1}
 \symHelper\!\left(
@@
 +
-\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
+\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
 \symXEstimate\symError_\symX\symBeta
@@
 \symXEstimate
-\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
--\left(\symNumReg-\symCount_0\right)
+\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+-\left(\symNumReg-\symCountVariate_0\right)
 }
@@
 \left(
-\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
+\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
 +
@@
 \sum_{\symRegVal=1}^\symRegRange 
-\frac{\symCount_\symRegVal
+\frac{\symCountVariate_\symRegVal
 }{2^\symRegVal}
 +
-\frac{\symCount_{\symRegRange+1}
+\frac{\symCountVariate_{\symRegRange+1}
 }{2^\symRegRange}
@@
 \frac{
-\left(\symNumReg-\symCount_0\right)-
+\left(\symNumReg-\symCountVariate_0\right)-
 \symXEstimate
-\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
+\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
 }
@@
 \symXEstimate
-\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
+\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
 }
@@
 \frac{1}{2}
-\sum_{\symRegVal=1}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
-+
-\frac{\symCount_{\symRegRange+1}
+\sum_{\symRegVal=1}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
++
+\frac{\symCountVariate_{\symRegRange+1}
 }{2^\symRegRange}
@@
 {
-\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
+\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
 }
@@
 \frac{
-\frac{\symCount_{\symRegRange+1}
+\frac{\symCountVariate_{\symRegRange+1}
 }{2^\symRegRange}
@@
 {
-\sum_{\symRegVal=1}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
+\sum_{\symRegVal=1}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
 }
@@
 \frac{
-\symCount_{\symRegRange+1}
+\symCountVariate_{\symRegRange+1}
 }
 {
-\symNumReg-\symCount_{\symRegRange+1}
-}
-\right)
-\end{equation}
-Hence, as long as most registers are not in the saturated state ($\symCount_{\symRegRange+1}\ll\symNumReg$), the relative error $\symError_\symX$ of the calculated estimate using the approximation $\symHelperApprox(\symX)$ for $\symHelper(\symX)$ has the same order of magnitude as $\symError_\symHelper$.
+\symNumReg-\symCountVariate_{\symRegRange+1}
+}
+\right)
+\end{equation}
+Hence, as long as most registers are not in the saturated state ($\symCountVariate_{\symRegRange+1}\ll\symNumReg$), the relative error $\symError_\symX$ of the calculated estimate using the approximation $\symHelperApprox(\symX)$ for $\symHelper(\symX)$ has the same order of magnitude as $\symError_\symHelper$.
 
@@
 %{
-%\symNumReg-\symCount_{\symRegRange+1}
+%\symNumReg-\symCountVariate_{\symRegRange+1}
 %}

### 2016-07-12 c++/cardinality_estimation.hpp

@@ -93,12 +93,12 @@ public:
 
         double deltaX = x;
         while(deltaX > x*relativeErrorLimit) {
-            int e;
-            frexp(x, &e);
-            double xPrime = ldexp(x, -std::max(kMaxPrime+1, e+2));
+            int kappaMinus1;
+            frexp(x, &kappaMinus1);
+            double xPrime = ldexp(x, -std::max(kMaxPrime+1, kappaMinus1+2));
             double xPrime2 = xPrime*xPrime;
             double h = xPrime - xPrime2/3 + (xPrime2*xPrime2)*(1./45. - xPrime2/472.5);
-            for (int k = e; k >= kMaxPrime; --k) {
+            for (int k = kappaMinus1; k >= kMaxPrime; --k) {
                 double hPrime = 1. - h;
                 h = (xPrime + h*hPrime)/(xPrime+hPrime);
                 xPrime += xPrime;
@@ -175,6 +175,32 @@ double flajoletRawEstimate(const std::vector<int>& c, int m) {
     double alpha = getAlpha(m);
     return (alpha*m)*m/z;
 }
+
+class OptimalLinearCountingEstimator {
+
+    const int p;
+    const int m = 1 << p;
+    const std::vector<double> results = initResults(m);
+
+    static std::vector<double> initResults(int m) {
+
+        std::vector<double> results(m+1);
+        results[m] = 0;
+        for (int c = m-1; c >= 0; --c) {
+            results[c] = results[c + 1] + m/(double)c; 
+        }
+        return results;
+    }
+
+public:
+
+    OptimalLinearCountingEstimator(const int p_) : p(p_) {}
+
+    double operator()(const std::vector<int>& c) const {
+        return results[c[0]];
+    }
+};
+
 
 class CorrectedRawEstimator {
     const int p;
@@ -423,9 +449,14 @@ void eval_joint_log_likelihood_function_and_derivatives(
 
     for (int k = q; k >= 0; --k) {
         double pow2k = std::ldexp(1., -k);
-        term1a += (jointStatistic[(q+2)*0+k]+jointStatistic[(q+2)*2+k]+jointStatistic[(q+2)*4+k]) * pow2k;
-        term1b += (jointStatistic[(q+2)*0+k]+jointStatistic[(q+2)*1+k]+jointStatistic[(q+2)*3+k]) * pow2k;
[…]

### 2016-07-12 python/benchmark-charts.py

@@ -61,8 +61,8 @@ ax_time.xaxis.grid(b=True, which='major', color='gray', linestyle='--')
 ax_time.set_xlabel("cardinality")
 ax_time.set_ylabel("time (ns)")
 
-ax_time.plot(cardinalities_12_20, max_likelihood_avg_exec_time_nanos_12_20, color='green', linestyle='-',linewidth=1.0, label='p=12 q=20')
-ax_time.plot(cardinalities_12_52, max_likelihood_avg_exec_time_nanos_12_52, color='blue', linestyle='-',linewidth=1.0, label='p=12 q=52')
+ax_time.plot(cardinalities_12_20, max_likelihood_avg_exec_time_nanos_12_20, color='green', linestyle='-',linewidth=1.0, label='p=12, q=20')
+ax_time.plot(cardinalities_12_52, max_likelihood_avg_exec_time_nanos_12_52, color='blue', linestyle='-',linewidth=1.0, label='p=12, q=52')
 
 ax_time.legend(loc=2,prop={'size':12})
 
@@ -78,10 +78,10 @@ ax_time.xaxis.grid(b=True, which='major', color='gray', linestyle='--')
 ax_time.set_xlabel("cardinality")
 ax_time.set_ylabel("time (ns)")
 
-ax_time.plot(cardinalities_12_20, corrected_raw_1_avg_exec_time_nanos_12_20, color='green', linestyle='-',linewidth=1.0, label='p=12 q=20 on-demand')
-ax_time.plot(cardinalities_12_52, corrected_raw_1_avg_exec_time_nanos_12_52, color='blue', linestyle='-',linewidth=1.0, label='p=12 q=52 on-demand')
-ax_time.plot(cardinalities_12_20, corrected_raw_2_avg_exec_time_nanos_12_20, color='red', linestyle='-',linewidth=1.0, label='p=12 q=20 precalculated')
-ax_time.plot(cardinalities_12_52, corrected_raw_2_avg_exec_time_nanos_12_52, color='black', linestyle='-',linewidth=1.0, label='p=12 q=52 precalculated')
+ax_time.plot(cardinalities_12_20, corrected_raw_1_avg_exec_time_nanos_12_20, color='green', linestyle='-',linewidth=1.0, label='p=12, q=20, on-demand')
[…]

### 2016-07-12 python/recursion-stability.py

@@ -5,18 +5,37 @@
 from math import expm1, ceil
 from scipy.optimize import minimize
 
+epsBound = 0.1
+
 def h(x):
     if x == 0:
         return 0
     else:
         return 1 - x/expm1(x)
-        
-def f(x):
-    return -((2*h(x)*(1-2*h(x)))/(x+2*h(x)*(1-h(x)))) - 2*h(x)/(x+2*(1-h(x)))
-    
-    
-res = minimize(f, 1)
 
-out = open('../paper/max_errror_propagation_factor.txt', 'w')
-out.write(str(ceil(-res.fun*1000)/1000.))
+def f1(x):
+    return -((h(2*x)*(1-2*h(2*x)))/(x+h(2*x)*(1-h(2*x)))) - h(2*x)/(x+1-h(2*x))
 
+def f2(x):
+    return - h(2*x)*h(2*x)/(x+h(2*x)*(1-h(2*x)))
+
+def f3(x):
+    return - h(2*x)/(x+1-h(2*x))
+
+res1 = minimize(f1, 1)
+res2 = minimize(f2, 1)
+res3 = minimize(f3, 1)
+
+factor1 = ceil(-res1.fun*1000)/1000.
+factor2 = ceil(-res2.fun*1000)/1000.
+factor3 = ceil(-res3.fun*1000)/1000.
+factor4 = ceil((factor1 + epsBound*factor2)/(1. - factor3*epsBound)*1000.)/1000.
+
+out1 = open('../paper/max_errror_propagation_factor1.txt', 'w')
+out1.write(str(factor1))
+out2 = open('../paper/max_errror_propagation_factor2.txt', 'w')
+out2.write(str(factor2))
+out3 = open('../paper/max_errror_propagation_factor3.txt', 'w')
+out3.write(str(factor3))
+out4 = open('../paper/max_errror_propagation_factor4.txt', 'w')
+out4.write(str(factor4))

### 2016-07-13 c++/cardinality_estimation.hpp
some minor changes

@@ -149,8 +149,8 @@ double flajoletSmallRangeEstimate(const std::vector<int>& c) {
 }
 
 double getAlpha(int m) {
-    return 1./(2.*std::log(2));
-    /*assert(m >= 16);
+    //return 1./(2.*std::log(2));
+    assert(m >= 16);
     double alpha;
     if (m == 16) {
         alpha = 0.673;
@@ -164,7 +164,7 @@ double getAlpha(int m) {
     else {
         alpha = 0.7213/(1. + 1.079/m);
     }
-    return alpha;*/
+    return alpha;
 }
 
 double flajoletRawEstimate(const std::vector<int>& c, int m) {
@@ -449,23 +449,23 @@ void eval_joint_log_likelihood_function_and_derivatives(
 
     for (int k = q; k >= 0; --k) {
         double pow2k = std::ldexp(1., -k);
-        const int cCenter = jointStatistic[(q+2)*0+k];
-        const int cUp     = jointStatistic[(q+2)*1+k];
-        const int cRight  = jointStatistic[(q+2)*2+k];
-        const int cDown   = jointStatistic[(q+2)*3+k];
-        const int cLeft   = jointStatistic[(q+2)*4+k];
-        term1a += (cCenter + cRight + cLeft) * pow2k;
-        term1b += (cCenter + cUp + cDown) * pow2k;
-        term1x += (cCenter + cRight + cDown) * pow2k;
+        const int cEqual    = jointStatistic[(q+2)*0+k];
+        const int cLarger2  = jointStatistic[(q+2)*1+k];
+        const int cSmaller1 = jointStatistic[(q+2)*2+k];
+        const int cSmaller2 = jointStatistic[(q+2)*3+k];
+        const int cLarger1  = jointStatistic[(q+2)*4+k];
+        term1a += (cEqual + cSmaller1 + cLarger1 ) * pow2k;
+        term1b += (cEqual + cLarger2  + cSmaller2) * pow2k;
+        term1x += (cEqual + cSmaller1 + cSmaller2) * pow2k;
     }
 
     for (int k = q+1; k >= 1; --k) {
 
-        const int cCenter = jointStatistic[(q+2)*0+k];
-        const int cUp     = jointStatistic[(q+2)*1+k];
-        const int cRight  = jointStatistic[(q+2)*2+k];
-        const int cDown   = jointStatistic[(q+2)*3+k];
[…]

### 2016-07-13 c++/hyperloglog.hpp

@@ -97,50 +97,20 @@ public:
             unsigned char val2 = hll2.registers[i];
             
             if (val1 < val2) {
-                data[stride*1 + val2] += 1; // up
-                data[stride*2 + val1] += 1; // right
+                data[stride*1 + val2] += 1; // C^(>)_2k
+                data[stride*2 + val1] += 1; // C^(<)_1k
             }
             else if (val1 > val2) {
-                data[stride*3 + val2] += 1; // down
-                data[stride*4 + val1] += 1; // left
+                data[stride*3 + val2] += 1; // C^(<)_2k
+                data[stride*4 + val1] += 1; // C^(>)_1k
             }
             else {
-                data[stride*0 + val1] += 1; // center
+                data[stride*0 + val1] += 1; // C^(=)_k
             }
         }
         return data;
     }
-    
-    // example with q = 1, p = 2
-    // register values of hll1 : {0, 1, 2, 0}
-    // register values of hll2 : {2, 1, 0, 2}
-    //
-    // the corresponding  count matrix is
-    //
-    //                ( 0 0 2 )
-    // count matrix = ( 0 1 0 )
-    //                ( 1 0 0 )
-    //
-    // or as array representation
-    //
-    // count matrix = [0, 0, 2, 0, 1, 0, 1, 0, 0]
-    static std::vector<int> getCountMatrix(const HyperLogLog& hll1, const HyperLogLog& hll2) {
-        assert(hll1.p == hll2.p);
-        assert(hll1.q == hll2.q);
-        
-        unsigned char q = hll1.q;
-        size_t stride = q+2;
-        
-        std::vector<int> countMatrix(stride*stride);
-        
-        for (size_t i = 0; i < hll1.registers.size(); ++i) {
-            countMatrix[hll1.registers[i] + stride*hll2.registers[i]] += 1;
-        }
-        
-        return countMatrix;
-    }
-
-        
+            
     void add(std::uint_fast64_t hashValue) {
         
         if ((registerValueFilter | hashValue) == (~uint_fast64_t(0))) {

### 2016-07-13 python/helper-function.py

@@ -7,7 +7,7 @@ from math import expm1, cosh, tanh, sinh, log
 import matplotlib.pyplot as plt
 
 numPoints = 100000
-xMin = 1e-3
+xMin = 0
 xMax = 110
 
 def helper(x):
@@ -31,7 +31,7 @@ def helperApprox(x):
 
 iter = range(0,numPoints)
 xValues = list([xMin + i*((xMax - xMin)/(numPoints-1)) for i in iter])
-yValues = list([helperApprox(xValues[i]) for i in iter])
+yValues = list([helper(xValues[i]) for i in iter])
 
 plt.rc('text', usetex=True)
 plt.rc('font', family='serif')

### 2016-07-17 paper/paper.tex
some minor changes
minor changes

@@
 \maketitle
-\let\thefootnote\relax\footnotetext{This article together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}.}
+\let\thefootnote\relax\footnotetext{This paper together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}.}
 \begin{abstract}
@@
 =
-2\alpha_\symNumReg
+\alpha_\symNumReg
 \operatorname{HM}
 \!\left(
-\symNumReg 2^{\symRegValVariate_1-1},\ldots,\symNumReg2^{\symRegValVariate_\symNumReg-1}
+\symNumReg 2^{\symRegValVariate_1},\ldots,\symNumReg2^{\symRegValVariate_\symNumReg}
 \right)
@@
 $\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ is finally obtained after $\symKappa -\symRegValVariate'_\text{max}$ iterations using \eqref{equ:helper_recursion1}.
+As shown in \cref{app:error_approx}, a small approximation error of $\symHelper$ does not have much impact on the error of the maximum likelihood estimate as long as most registers are not yet saturated.
 \end{itemize}
@@
 
-The HyperLogLog sketch that represents $\symSetS_1$ can be thought to be constructed from two HyperLogLog sketches representing $\symSetA$ and $\symSetX$ and merging both according to \eqref{equ:hll_union}.
+The HyperLogLog sketch that represents $\symSetS_1$ can be thought to be constructed from two HyperLogLog sketches representing $\symSetA$ and $\symSetX$ and merging both using \eqref{equ:hll_union}.
 Analogously, the HyperLogLog sketch for $\symSetS_2$ could have been obtained from sketches for $\symSetB$ and $\symSetX$.
@@
 \symRegValVariate_\symSetXSuffix \leq \min\!\left(\symRegVal_1, \symRegVal_2\right)
-\right)
+\right).
 \end{align}
@@
 \symRegValVariate_2 \leq \symRegVal_2-1
-)
+),
 \end{multline}
@@
 \right)
-\left[1\leq\symRegVal_1=\symRegVal_2\right]
+\left[1\leq\symRegVal_1=\symRegVal_2\right].
 \end{align}
 
-Since under the Poisson model, the values for different registers are independent we are now able to write the joint probability mass function for all registers in both HyperLogLog sketches
+Since the values for different registers are independent under the Poisson model, we are now able to write the joint probability mass function for all registers in both HyperLogLog sketches
 \begin{equation}
@@
 \right)
-\symCountVariate^{=}_{\symRegVal}
+\symCountVariate^{=}_{\symRegVal},
 \end{align}
-where the constant $\symCountVariate^{>}_{1\symRegVal}$,
+where the constants $\symCountVariate^{>}_{1\symRegVal}$,
 $\symCountVariate^{<}_{1\symRegVal}$,
@@
 \symCountVariate^{=}_{\symRegVal}&:=\left|\lbrace\symIndexI\vert\symRegVal=\symRegValVariate_{1\symIndexI}=
-\symRegValVariate_{2\symRegVal}\rbrace\right|.
+\symRegValVariate_{2\symIndexI}\rbrace\right|.
 \end{aligned}
 \end{equation}
+These $5\symNumReg$ values are sufficient for estimating $\symPoissonRate_\symSetA$, $\symPoissonRate_\symSetB$, and $\symPoissonRate_\symSetX$.
+Actually, the set of these constants can be further reduced, because $\symCountVariate^{>}_{10}=\symCountVariate^{>}_{20}=
+\symCountVariate^{<}_{1(\symRegRange+1)}=\symCountVariate^{<}_{2(\symRegRange+1)}=0$
+always holds.
 
 The log-likelihood function \eqref{equ:log_likelihood_pair} does not always have a strict global maximum point.
-For example, if all register values of the first HyperLogLog sketch are larger than the corresponding values in the second HyperLogLog sketch, the function can be rewritten as sum of two functions, one dependent on $\symPoissonRate_\symSetASuffix$ and the other dependent on $(\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix)$.
+For example, if all register values of the first HyperLogLog sketch are larger than the corresponding values in the second HyperLogLog sketch, $\symCountVariate_{1\symRegVal}^{<}
+=
+\symCountVariate_{\symRegVal}^{=}
+=
+\symCountVariate_{2\symRegVal}^{>}
+=
+0$ for all $\symRegVal$, the function can be rewritten as sum of two functions, one dependent on $\symPoissonRate_\symSetASuffix$ and the other dependent on $\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix$.
 The maximum is obtained, if $\symPoissonRate_\symSetASuffix = \symPoissonRateEstimate_1$ and $\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix = \symPoissonRateEstimate_2$.
-Here $\symPoissonRateEstimate_1$ and $\symPoissonRateEstimate_2$ are the cardinality estimates for the first and second HyperLogLog sketch, respectively.
-Similarly, if all register values of the first are smaller than those of the second HyperLogLog sketch, the maximum is obtained when
-$\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix = \symPoissonRateEstimate_1$ and $\symPoissonRate_\symSetBSuffix = \symPoissonRateEstimate_2$.
-
-If we know that the two HyperLogLog sketches have been filled by elements taken from disjoint sets we can assume $\symPoissonRate_\symSetXSuffix=0$.
+Here $\symPoissonRateEstimate_1$ and $\symPoissonRateEstimate_2$ are the single HyperLogLog estimates for the first and second HyperLogLog sketch, respectively, which maximize the corresponding log-likelihood functions \eqref{equ:log_likelihood_single}.
+This means that the maximum-likelihood makes no clear statement about the intersection size in this case.
+The estimate for $\symPoissonRate_\symX$ could be anything between 0 and $\symPoissonRateEstimate_2$.
+For comparison, the inclusion-exclusion approach gives $\symPoissonRateEstimate_2$ as estimate.
+This result is questionable, because there is no evidence at all, that the sets $\symSetS_1$ and $\symSetS_2$ really have common elements.
+
+If it is given that the two HyperLogLog sketches represent disjoint sets, we know that $\symPoissonRate_\symSetXSuffix=0$.
 In this case \eqref{equ:log_likelihood_pair} is separable into the sum of two log-likelihood functions that depend on $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$, respectively, and that follow \eqref{equ:log_likelihood_single}.
-Hence, the joint maximum likelihood estimation of $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$ gives the same results as estimating them independently by maximization of \eqref{equ:log_likelihood_single}.
-
+Hence, the joint maximum likelihood estimation of $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$ gives as expected the same results as estimating them independently by maximizing \eqref{equ:log_likelihood_single}.
+
+\subsection{Computation of the maximum likelihood estimates}
 The first derivatives are
@@
 \end{equation}
-The triangle inequality lead to
+The triangle inequality leads to
 \begin{equation}
@@
 \cdot
-\input{max_errror_propagation_factor4.txt}
+\input{max_errror_propagation_factor4.txt},
 \end{equation}
@@
 \section{Error caused by approximation of $\symHelper(\symX)$}
+\label{app:error_approx}
 According to \eqref{equ:func} the exact estimate $\symXEstimate$ fulfills 
@@
 {
-\sum_{\symRegVal=1}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
 }

### 2016-08-07 paper/paper.tex
various changes:
* refactored c++ source code, new class for joint statistic of 2 hyperloglogs
* improved intersection calculation
* some paper improvements

@@
 \symDefine{\symX}{x}
+\symDefine{\symPhi}{\varphi}
 \symDefine{\symXEstimate}{\hat{\symX}}
@@
 \symDefine{\symHelperApprox}{\tilde{\symHelper}}
+\symDefine{\symHelperW}{w}
+\symDefine{\symHelperV}{v}
+\symDefine{\symHelperU}{u}
 \symDefine{\symSetA}{A}
@@
 \cref{alg:insert} shows the update procedure for inserting a data element into the HyperLogLog sketch.
+
+The desribed element insertion algorithm makes use of what is known as stochastic averaging \cite{Flajolet1985}.
+Instead of updating each of all $\symNumReg$ registers using $\symNumReg$ independent hash values, which would be an $\symBigO(\symNumReg)$ operation, only one register is selected and updated, which requires only a single hash function and reduces the complexity to $\symBigO(1)$.
 
@@
 The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities.
-$\symRegProbability(\symRegVal\vert\symCardinality)$ is the probability that the smallest of $\symCardinality$ uniformly distributed random integers in range $[0, 2^\symRegRange)$ has a binary representation with its first 1-bit at position $\symRegVal$,
+$\symRegProbability(\symRegVal\vert\symCardinality)$ is the probability that the value of a register is equal to $\symRegVal$, after it was selected $\symCardinality$ times by the insertion algorithm
 \begin{equation}
@@
 First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}.
-Finally, the harmonic mean (HM) was found to give even better estimates as it is inherently less sensitive to outliers.
+Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers.
 The result is the so-called raw estimator which is given by
@@
 \alpha_\symNumReg
-\operatorname{HM}
-\!\left(
-\symNumReg 2^{\symRegValVariate_1},\ldots,\symNumReg2^{\symRegValVariate_\symNumReg}
-\right)
+\frac{\symNumReg}
+{\frac{1}{\symNumReg 2^{\symRegValVariate_1}}+\ldots+\frac{1}{\symNumReg2^{\symRegValVariate_\symNumReg}}}
 = 
@@
 \cref{fig:raw_estimate} shows the distribution of the relative error for the raw estimator as function of the cardinality.
-The chart is based on 10\,000 randomly generated HyperLogLog sketches.
+The chart is based on 10,000 randomly generated HyperLogLog sketches.
 More details of the experimental setup will be explained later in \cref{sec:corrected_raw_estimation_error}.
@@
 \includesvg[width=1\textwidth]{raw_estimate}
-\caption{The distribution of the relative estimation error over the cardinality for the raw estimator after evaluation of 10\,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
+\caption{The distribution of the relative estimation error over the cardinality for the raw estimator after evaluation of 10,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
 \label{fig:raw_estimate}
@@
 \includesvg[width=1\textwidth]{small_range_estimate}
-\caption{The distribution of the relative estimation error for the linear counting estimator after evaluation of 10\,000 randomly generated bitmaps of size $2^{12}$ which correspond to HyperLogLog sketches with parameters $\symPrecision = 12$ and $\symRegRange=0$.}
+\caption{The distribution of the relative estimation error for the linear counting estimator after evaluation of 10,000 randomly generated bitmaps of size $2^{12}$ which correspond to HyperLogLog sketches with parameters $\symPrecision = 12$ and $\symRegRange=0$.}
 \label{fig:small_range_estimate}
@@
 \includesvg[width=1\textwidth]{original_estimate}
-\caption{The distribution of the relative estimation error over the cardinality for the original estimation algorithm after evaluation of 10\,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
+\caption{The distribution of the relative estimation error over the cardinality for the original estimation algorithm after evaluation of 10,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
 \label{fig:original_estimate}
@@
 \label{sec:corrected_raw_estimation_error}
-In order to verify the new estimation algorithm, we generated 10\,000 HyperLogLog sketches and inserted up to 50 billion unique elements.
+In order to verify the new estimation algorithm, we generated 10,000 HyperLogLog sketches and inserted up to 50 billion unique elements.
 Assuming a uniform hash function, element hash values can be mocked by random numbers.
-For the following results we used the Mersenne Twister random number generator with 19\,937 bit state size from the C++ standard library.
+For the following results we used the Mersenne Twister random number generator with 19,937 bit state size from the C++ standard library.
 
@@
 
-We are not the first applying the maximum likelihood method to HyperLogLog sketches \cite{Clifford2012}.
-However, there a different element insertion algorithm was considered, which is very inefficient in practice due to its $\symBigO(\symNumReg)$ runtime complexity.
-Each register has its own hash function which is applied to the inserted element.
-The result is used to update the corresponding register.
-The advantage is that the register values are statistically independent in this case, which allows factorization of the joint probability distribution of all register values.
-
-In contrast, \cref{alg:insert} uses what is also called stochastic averaging.
-Instead of iterating through all registers, a single hash value is used, to select a single register and to update its value.
-Since the register values are statistically dependent in this case, the maximum likelihood method is much more complicated to apply.
-However, with the help of the Poisson approximation we are finally able to derive a new robust and efficient cardinality estimation algorithm.
+We remark that we are not the first applying the maximum likelihood method to HyperLogLog sketches \cite{Clifford2012}, where the case without stochastic averaging (compare \cref{sec:data_element_insertion}) was considered.
+The register values are statistically independent by nature in this case which
+allows factorization of the joint probability mass function and allows a straightforward application of the maximum likelihood method.
+
+The more relevant case with stochastic averaging, which is considered in this paper, is more complicated.
+However, with the help of the Poisson approximation, we are able to still apply the maximum likelihood method and to derive a new robust and efficient cardinality estimation algorithm.
 Furthermore, we will demonstrate that consequent application of the maximum likelihood method reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values.
@@
 \centering
-\includesvg[width=0.45\textwidth]{helper}
+\includesvg[width=0.6\textwidth]{helper}
 \caption{The function $\symHelper(\symX)$.}
@@
 \item Many programming languages allow the efficient multiplication and division by any integral power of 2 using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
-\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\left\lbrace\frac{\symX}{2^{\symRegValVariate'_\text{max}}},\frac{\symX}{2^{\symRegValVariate'_\text{max}-1}}\ldots,\frac{\symX}{2^{\symRegValVariate_\text{min}}}\right\rbrace$ where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$.
+\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\left\lbrace\frac{\symX}{2^{\symRegValVariate'_\text{max}}},\frac{\symX}{2^{\symRegValVariate'_\text{max}-1}},\ldots,\frac{\symX}{2^{\symRegValVariate_\text{min}}}\right\rbrace$ where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$.
 This series corresponds to a geometric series with ratio 2.
@@
 The smallest argument for which $\symHelper$ needs to be evaluated is $\frac{\symX}{2^{\symRegValVariate'_\text{max}}}$. If 
-$\symCountVariate_{\symRegRange+1}=0$, $\frac{\symX}{2^{\symRegValVariate'_\text{max}}} \leq \log 2 \approx 0.693$ holds, which is straightforward to see using \eqref{equ:strong_upper_bound}.
+$\symCountVariate_{\symRegRange+1}=0$, we can find an upper bound for the smallest argument using \eqref{equ:strong_upper_bound}
+\begin{equation}
+\frac{\symX}{2^{\symRegValVariate'_\text{max}}} 
+\leq
+\log\!\left(
+1+
+\frac{\sum_{\symRegVal=0}^{\symRegValVariate'_\text{max}}\symCountVariate_\symRegVal}
+{
+2^{\symRegValVariate'_\text{max}}
+\sum_{\symRegVal=0}^{\symRegValVariate'_\text{max}}
+\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+}
+\right)
+\leq \log 2 \approx 0.693.
+\end{equation}
 In practice, $\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\leq0.5$ is satisfied most of the time as long as only a few registers are saturated ($\symCountVariate_{\symRegRange+1}\ll \symNumReg$).
@@
 
-All these optimizations together finally give the new cardinality estimation algorithm presented in \cref{alg:estimate_ml}.
+Putting all these optimizations together finally gives the new cardinality estimation algorithm as presented in \cref{alg:estimate_ml}.
+The algorithm requires only elementary operations, except for large cardinalities, where it is worth to take the strong \eqref{equ:strong_lower_bound} instead of the weak lower bound \eqref{equ:weak_lower_bound} as initial value, which requires a single logarithm evaluation.
+The stronger bound is a much better approximation especially for large cardinalities and substantially reduces the number of required iteration cycles.
 
@@
 \end{equation}
-Insertion of \eqref{equ:joint_log_pmf_single_register} results in the two HyperLogLog case analog of \eqref{equ:log_likelihood_single}
+Insertion of \eqref{equ:joint_log_pmf_single_register} results in the generalization of \eqref{equ:log_likelihood_single} for two HyperLogLog sketches
 \begin{align}
@@
 +
+\sum_{\symRegVal=1}^{\symRegRange}
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal}}}\right)
+\symCountVariate^{<}_{1\symRegVal}
++
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal}}}\right)
+\symCountVariate^{<}_{2\symRegVal}
+\nonumber\\
+&
++
 \sum_{\symRegVal=1}^{\symRegRange+1}
-\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
-\symCountVariate^{<}_{1\symRegVal}
-+
 \log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
 \symCountVariate^{>}_{1\symRegVal}
-\nonumber\\
-&
-+
-\sum_{\symRegVal=1}^{\symRegRange+1}
-\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
-\symCountVariate^{<}_{2\symRegVal}
 +
@@
 The maximum is obtained, if $\symPoissonRate_\symSetASuffix = \symPoissonRateEstimate_1$ and $\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix = \symPoissonRateEstimate_2$.
-Here $\symPoissonRateEstimate_1$ and $\symPoissonRateEstimate_2$ are the single HyperLogLog estimates for the first and second HyperLogLog sketch, respectively, which maximize the corresponding log-likelihood functions \eqref{equ:log_likelihood_single}.
+Here $\symPoissonRateEstimate_1$ and $\symPoissonRateEstimate_2$ are the maximum likelihood cardinality estimates for the first and second HyperLogLog sketch, respectively.
 This means that the maximum-likelihood makes no clear statement about the intersection size in this case.
 The estimate for $\symPoissonRate_\symX$ could be anything between 0 and $\symPoissonRateEstimate_2$.
-For comparison, the inclusion-exclusion approach gives $\symPoissonRateEstimate_2$ as estimate.
+For comparison, the inclusion-exclusion approach would give $\symPoissonRateEstimate_2$ as estimate.
 This result is questionable, because there is no evidence at all, that the sets $\symSetS_1$ and $\symSetS_2$ really have common elements.
 
-If it is given that the two HyperLogLog sketches represent disjoint sets, we know that $\symPoissonRate_\symSetXSuffix=0$.
-In this case \eqref{equ:log_likelihood_pair} is separable into the sum of two log-likelihood functions that depend on $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$, respectively, and that follow \eqref{equ:log_likelihood_single}.
-Hence, the joint maximum likelihood estimation of $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$ gives as expected the same results as estimating them independently by maximizing \eqref{equ:log_likelihood_single}.
+If it is given that the two HyperLogLog sketches represent disjoint sets, we know in advance that $\symPoissonRate_\symSetXSuffix=0$.
+In this case \eqref{equ:log_likelihood_pair} can be splitted into the sum of two unary functions with parameters $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$, respectively.
+As expected, these two functions are equivalent to the individual log-likelihood functions \eqref{equ:log_likelihood_single} of both HyperLogLog sketches.
 
 \subsection{Computation of the maximum likelihood estimates}
-The first derivatives are
-\begin{align}
-\frac{\partial\log \symLikelihood}{\partial\symPoissonRate_\symSetASuffix}
-(
-\symPoissonRate_\symSetASuffix,
-\symPoissonRate_\symSetBSuffix,
-\symPoissonRate_\symSetXSuffix
+The maximum likelihood estimates can be obtained by maximizing \eqref{equ:log_likelihood_pair}.
+Since the three parameters are all non-negative, this is a constrained optimization problem.
+In order to get rid of these constraints, we use the transformation $\symPoissonRate=\symNumReg e^{\symPhi}$.
+This mapping has also the nice property that relative accuracy limits are translated into absolute ones, because $\Delta\symPhi = \Delta\symPoissonRate/\symPoissonRate$.
+Many optimization algorithm implementations allow the definition of absolute limits rather than relative ones.
+
+The transformed log-likelihood function can be written as
+\begin{multline}
+\symFunc(\symPhi_\symSetASuffix,\symPhi_\symSetBSuffix,\symPhi_\symSetXSuffix):=
+\log \symLikelihood(
+\symNumReg e^{\symPhi_\symSetASuffix},
+\symNumReg e^{\symPhi_\symSetBSuffix},
+\symNumReg e^{\symPhi_\symSetXSuffix}
 \vert
@@
 \boldsymbol{\symRegValVariate}_2
-)
-=
-&
--
-\frac{1}{\symNumReg}
+)=
+\\
+\begin{aligned}
+=&-
 \sum_{\symRegVal=0}^{\symRegRange}
-\frac{
+\left(
   \symCountVariate^{<}_{1\symRegVal}+
@@
   \symCountVariate^{>}_{1\symRegVal}
-}{2^{\symRegVal}}
-+
-\frac{1}{\symNumReg}
-\sum_{\symRegVal=1}^{\symRegRange+1}
-\frac{1}{e^{\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}-1}
-\frac{\symCountVariate^{<}_{1\symRegVal}}{2^{\min\left(\symRegVal,\symRegRange\right)}}
-\nonumber\\
-&
-+
-\frac{1}{\symNumReg}\sum_{\symRegVal=1}^{\symRegRange+1}
-\frac{1}{e^{\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}-1}
-\frac{\symCountVariate^{>}_{1\symRegVal}}{2^{\min\left(\symRegVal,\symRegRange\right)}}
-\nonumber\\
-&
-+
-\frac{1}{\symNumReg}
-\sum_{\symRegVal=1}^{\symRegRange+1}
-\frac{
-e^{\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
--
-1
-}{
-e^{\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
--
-e^{\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
--
-e^{\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
-+
-1
-}
-\frac{\symCountVariate^{=}_{\symRegVal}
-}{2^{\min\left(\symRegVal,\symRegRange\right)}}
-\end{align}
-
-\begin{align}
-\frac{\partial\log \symLikelihood}{\partial\symPoissonRate_\symSetXSuffix}
-(
-\symPoissonRate_\symSetASuffix,
-\symPoissonRate_\symSetBSuffix,
-\symPoissonRate_\symSetXSuffix
-\vert
-\boldsymbol{\symRegValVariate}_1,
-\boldsymbol{\symRegValVariate}_2
-)
-=
-&
--
-\frac{1}{\symNumReg}
-\sum_{\symRegVal=0}^{\symRegRange}
-\frac{
+\right)
+\frac{e^{\symPhi_\symSetASuffix}}{2^{\symRegVal}}
++
+\left(
+  \symCountVariate^{<}_{2\symRegVal}+
+  \symCountVariate^{=}_{\symRegVal}+
+  \symCountVariate^{>}_{2\symRegVal}
+\right)
+\frac{e^{\symPhi_\symSetBSuffix}}{2^{\symRegVal}}
++
+\left(
   \symCountVariate^{<}_{1\symRegVal}+
@@
   \symCountVariate^{<}_{2\symRegVal}
-}{2^{\symRegVal}}
-+
-\frac{1}{\symNumReg}
+\right)
+\frac{e^{\symPhi_\symSetXSuffix}}{2^{\symRegVal}}
+\\
+&+
+\sum_{\symRegVal=1}^{\symRegRange}
+\symCountVariate^{<}_{1\symRegVal}
+\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetASuffix}
++e^{\symPhi_\symSetXSuffix}
+}{2^{\symRegVal}}}\right)
++
+\symCountVariate^{<}_{2\symRegVal}
+\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetBSuffix}
++
+e^{\symPhi_\symSetXSuffix}
+}{2^{\symRegVal}}}\right)
+\\
+&+
 \sum_{\symRegVal=1}^{\symRegRange+1}
-\frac{1}{e^{\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}-1}
-\frac{\symCountVariate^{<}_{1\symRegVal}}{2^{\min\left(\symRegVal,\symRegRange\right)}}
+\symCountVariate^{>}_{1\symRegVal}
+\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetASuffix}
+}{2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
++
+\symCountVariate^{>}_{2\symRegVal}
+\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetBSuffix}
+}{2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
+\\
+&+
+\sum_{\symRegVal=1}^{\symRegRange+1}
+\symCountVariate^{=}_{\symRegVal}
+\log\!\left(
+1
+-e^{-\frac{e^{\symPhi_\symSetASuffix}
++e^{\symPhi_\symSetXSuffix}
+}{2^{\min\left(\symRegVal,\symRegRange\right)}}}
+-
+e^{-\frac{e^{\symPhi_\symSetBSuffix}
++e^{\symPhi_\symSetXSuffix}
+}{2^{\min\left(\symRegVal,\symRegRange\right)}}}
++
+e^{-\frac{e^{\symPhi_\symSetASuffix}
++e^{\symPhi_\symSetBSuffix}
++e^{\symPhi_\symSetXSuffix}
+}{2^{\min\left(\symRegVal,\symRegRange\right)}}}
+\right).
+\end{aligned}
+\end{multline}
+
+Quasi-Newton methods are commonly used for finding the maximum. They require the calculation of the first derivatives which are given by
+\begin{align}
+\frac{\partial\symFunc}{\partial\symPhi_\symSetASuffix}
+=
+&
+-
+\sum_{\symRegVal=0}^{\symRegRange}
+\left(
+  \symCountVariate^{<}_{1\symRegVal}+
+  \symCountVariate^{=}_{\symRegVal}+
+  \symCountVariate^{>}_{1\symRegVal}
+\right)
+\frac{e^{\symPhi_\symSetASuffix}}{2^{\symRegVal}}
++
+\sum_{\symRegVal=1}^{\symRegRange}
+\symCountVariate^{<}_{1\symRegVal}
+\symHelperV\!\left(
+\frac{e^{\symPhi_\symSetASuffix}}
+{2^{\symRegVal}}
+,
+\frac{e^{\symPhi_\symSetXSuffix}}
+{2^{\symRegVal}}
+\right)
++
+\sum_{\symRegVal=1}^{\symRegRange+1}
+\symCountVariate^{>}_{1\symRegVal}
+\symHelperV\!\left(
+\frac{e^{\symPhi_\symSetASuffix}}
+{2^{\min\left(\symRegVal,\symRegRange\right)}}
+,
+0
+\right)
 \nonumber\\
@@
 +
-\frac{1}{\symNumReg}\sum_{\symRegVal=1}^{\symRegRange+1}
-\frac{1}{e^{\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}-1}
-\frac{\symCountVariate^{<}_{2\symRegVal}}{2^{\min\left(\symRegVal,\symRegRange\right)}}
+\sum_{\symRegVal=1}^{\symRegRange+1}
+\symCountVariate^{=}_{\symRegVal}
+\symHelperW\!
+\left(
+\frac{e^{\symPhi_\symSetASuffix}}
+{2^{\min\left(\symRegVal,\symRegRange\right)}}
+,
+\frac{e^{\symPhi_\symSetBSuffix}}
+{2^{\min\left(\symRegVal,\symRegRange\right)}}
+,
+\frac{e^{\symPhi_\symSetXSuffix}}
+{2^{\min\left(\symRegVal,\symRegRange\right)}}
+\right)
+\end{align}
+
+\begin{align}
+\frac{\partial\symFunc}{\partial\symPhi_\symSetBSuffix}
+=
+&
+-
+\sum_{\symRegVal=0}^{\symRegRange}
+\left(
+  \symCountVariate^{<}_{2\symRegVal}+
+  \symCountVariate^{=}_{\symRegVal}+
+  \symCountVariate^{>}_{2\symRegVal}
+\right)
+\frac{e^{\symPhi_\symSetBSuffix}}{2^{\symRegVal}}
++
+\sum_{\symRegVal=1}^{\symRegRange}
+\symCountVariate^{<}_{2\symRegVal}
+\symHelperV\!\left(
+\frac{e^{\symPhi_\symSetBSuffix}}
+{2^{\symRegVal}}
+,
+\frac{e^{\symPhi_\symSetXSuffix}}
+{2^{\symRegVal}}
+\right)
++
+\sum_{\symRegVal=1}^{\symRegRange+1}
+\symCountVariate^{>}_{2\symRegVal}
+\symHelperV\!\left(
+\frac{e^{\symPhi_\symSetBSuffix}}
+{2^{\min\left(\symRegVal,\symRegRange\right)}}
+,
+0
+\right)
 \nonumber\\
@@
 +
-\frac{1}{\symNumReg}
 \sum_{\symRegVal=1}^{\symRegRange+1}
+\symCountVariate^{=}_{\symRegVal}
+\symHelperW\!
+\left(
+\frac{e^{\symPhi_\symSetBSuffix}}
+{2^{\min\left(\symRegVal,\symRegRange\right)}}
+,
+\frac{e^{\symPhi_\symSetASuffix}}
+{2^{\min\left(\symRegVal,\symRegRange\right)}}
+,
+\frac{e^{\symPhi_\symSetXSuffix}}
+{2^{\min\left(\symRegVal,\symRegRange\right)}}
+\right)
+\end{align}
+
+\begin{align}
+\frac{\partial\symFunc}{\partial\symPhi_\symSetXSuffix}
+=
+&
+-
+\sum_{\symRegVal=0}^{\symRegRange}
+\left(
+  \symCountVariate^{<}_{1\symRegVal}+
+  \symCountVariate^{=}_{\symRegVal}+
+  \symCountVariate^{<}_{2\symRegVal}
+\right)
+\frac{e^{\symPhi_\symSetBSuffix}}{2^{\symRegVal}}
++
+\sum_{\symRegVal=1}^{\symRegRange}
+\symCountVariate^{<}_{1\symRegVal}
+\symHelperV\!\left(
+\frac{e^{\symPhi_\symSetXSuffix}}
+{2^{\symRegVal}}
+,
+\frac{e^{\symPhi_\symSetASuffix}}
+{2^{\symRegVal}}
+\right)
++
+\sum_{\symRegVal=1}^{\symRegRange}
+\symCountVariate^{<}_{2\symRegVal}
+\symHelperV\!\left(
+\frac{e^{\symPhi_\symSetXSuffix}}
+{2^{\symRegVal}}
+,
+\frac{e^{\symPhi_\symSetBSuffix}}
+{2^{\symRegVal}}
+\right)
+\nonumber\\
+&
++
+\sum_{\symRegVal=1}^{\symRegRange+1}
+\symCountVariate^{=}_{\symRegVal}
+\symHelperU\!
+\left(
+\frac{e^{\symPhi_\symSetASuffix}}
+{2^{\min\left(\symRegVal,\symRegRange\right)}}
+,
+\frac{e^{\symPhi_\symSetBSuffix}}
+{2^{\min\left(\symRegVal,\symRegRange\right)}}
+,
+\frac{e^{\symPhi_\symSetXSuffix}}
+{2^{\min\left(\symRegVal,\symRegRange\right)}}
+\right)
+\end{align}
+
+\begin{equation}
+\symHelperV\!\left(\symX_1,\symX_2\right)
+:=
+\frac{\symX_1}{
+e^{\symX_1+\symX_2}-1
+}
+=
+\frac{\symX_1}{\symX_1 + \symX_2}
 \frac{
-e^{\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
-+
-e^{\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
--
-1
+\symX_1 + \symX_2
+}
+{
+e^{\symX_1+\symX_2}-1
+}
+\end{equation}
+\begin{equation}
+\symHelperW\!\left(\symX_1,\symX_2,\symX_3\right)
+:=
+\frac{\symX_1\left(e^{\symX_2}-1\right)
 }{
-e^{\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
--
-e^{\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
--
-e^{\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
-+
-1
-}
-\frac{\symCountVariate^{=}_{\symRegVal}
-}{2^{\min\left(\symRegVal,\symRegRange\right)}}
-\end{align}
+e^{\symX_1+\symX_2+\symX_3}-e^{\symX_1}-e^{\symX_2}+1
+}
+\end{equation}
+\begin{equation}
+\symHelperU\!\left(\symX_1,\symX_2,\symX_3\right)
+:=
+\frac{\left(e^{\symX_1}+e^{\symX_2}-1\right)\symX_3
+}{
+e^{\symX_1+\symX_2+\symX_3}-e^{\symX_1}-e^{\symX_2}+1
+}
+\end{equation}
 
@@
 
-% TODO is it possible to prove 
-%$\left|\symError_\symX\right| 
-%\leq
-%\left|\symError_\symHelper\right|
-%\frac{1}{2}
-%\frac{
-%\symNumReg
-%}
-%{
-%\symNumReg-\symCountVariate_{\symRegRange+1}
-%}
-%$}
-
 \bibliographystyle{unsrt}

### 2016-08-07 c++/benchmark.cpp

@@ -26,7 +26,7 @@ const int maxPplusQ = 64;
 
 const int numLoops = 100;
 
-const int numData = 1000;
+const size_t desiredNumData = 1000;
 
 const string seedFileName = "../data/hll/seed.dat";
 const string cardFileName = "../data/cardinalities.dat";
@@ -77,7 +77,9 @@ void run(const int p, const int q, const string& resultsFileName) {
 
     cout << "seedSize "  << seeds.size() << endl;
 
-    assert(numData <= seeds.size());
+    const size_t numData = min(desiredNumData, seeds.size());
+
+    cout << "numData "  << numData << endl;
 
     vector<long> cardinalities;
     {
@@ -195,7 +197,7 @@ void run(const int p, const int q, const string& resultsFileName) {
             auto elapsed = chrono::duration_cast<chrono::nanoseconds>(end - start);
             flajoletCorrectedWithoutRegisterScan = elapsed.count()/(double)dataSize;
         }
-        
+
         {
             auto start = std::chrono::system_clock::now();
             for(auto c : counts) {

### 2016-08-07 c++/cardinality_estimation.hpp

@@ -16,6 +16,7 @@
 #include <gsl/gsl_vector.h>
 #include <gsl/gsl_multimin.h>
 
+#include "two_hyperloglog_statistic.hpp"
 
 int getPFromNumberRegisters(int m) {
     int p;
@@ -27,12 +28,6 @@ int getPFromNumberRegisters(int m) {
 
 int getPFromCounts(const std::vector<int>& c) {
     int m = std::accumulate(c.begin(), c.end(), 0);
-    return getPFromNumberRegisters(m);
-}
-
-int getPFromJointStatistic(const std::vector<int>& jointStatistic) {
-    size_t l = jointStatistic.size()/5;
-    const int m = std::accumulate(&jointStatistic[l*0], &jointStatistic[l*1], 0) + std::accumulate(&jointStatistic[l*2], &jointStatistic[l*3], 0) + std::accumulate(&jointStatistic[l*3], &jointStatistic[l*4], 0);
     return getPFromNumberRegisters(m);
 }
 
@@ -187,7 +182,7 @@ class OptimalLinearCountingEstimator {
         std::vector<double> results(m+1);
         results[m] = 0;
         for (int c = m-1; c >= 0; --c) {
-            results[c] = results[c + 1] + m/(double)c; 
+            results[c] = results[c + 1] + m/(double)c;
         }
         return results;
     }
@@ -319,7 +314,7 @@ double flajoletEstimate(const std::vector<int>& c) {
         eStar = e;
     }
     else {
-        if (e <= ldexp(1., 32)) { // is not part of original estimate
+        if (e <= ldexp(1., 32)) { // is not part of original algorithm
             eStar = -std::pow(2., 32)*std::log1p(-ldexp(e, -32));
         }
         else {
@@ -329,7 +324,6 @@ double flajoletEstimate(const std::vector<int>& c) {
     return eStar;
 }
 
-// lower bound estimate obtained by applying Jensen's inequality
 double strongLowerBoundEstimate(const std::vector<int>& c) {
     int m = std::accumulate(c.begin(), c.end(), 0);
     int q = c.size()-2;
@@ -348,7 +342,6 @@ double strongLowerBoundEstimate(const std::vector<int>& c) {
     return m*x;
 }
 
-// using log(1+x) >= 2*x/(x+2)
 double weakLowerBoundEstimate(const std::vector<int>& c) {
[…]

### 2016-08-07 c++/hyperloglog.hpp

@@ -12,18 +12,20 @@
 #include <random>
 #include <iostream>
 
+#include "two_hyperloglog_statistic.hpp"
+
 class HyperLogLog {
-    
+
     unsigned char p;
     unsigned char q;
-    
+
     std::vector<unsigned char> registers;
-    
+
     std::vector<int> counts;
-    
+
     unsigned char minCount;
     uint_fast64_t registerValueFilter;
-    
+
     HyperLogLog(unsigned char p_, unsigned char q_, std::vector<unsigned char> registers_) : p(p_), q(q_), registers(registers_), counts(q_+2), minCount(0) {
         assert((size_t(1) << p_ ) == registers.size());
         for (unsigned char r : registers) {
@@ -34,16 +36,16 @@ class HyperLogLog {
             minCount += 1;
         }
         registerValueFilter = (~uint_fast64_t(0)) << minCount;
-    } 
+    }
 
 public:
-    
+
     HyperLogLog(unsigned char p_, unsigned char q_) : p(p_), q(q_), registers(std::size_t(1) << p_), counts(q_+2), minCount(0), registerValueFilter(~uint_fast64_t(0)) {
         counts[0] = registers.size();
     }
 
     static HyperLogLog createFromCounts(const std::vector<int>& counts, std::mt19937_64& rng) {
-        
+
         assert(counts.size() >= 2);
         unsigned char q = counts.size() - 2;
         size_t sum = 0;
@@ -56,7 +58,7 @@ public:
         assert(sum == size_t(1) << p);
 
         std::vector<unsigned char> registers(sum);
-        
+
         size_t idx = 0;
         for (size_t j = 0; j < counts.size(); ++j) {
             for (int k = 0; k < counts[j]; ++k) {
@@ -64,64 +66,69 @@ public:
                 idx += 1;
             }
         }
-        
+
         std::shuffle(registers.begin(), registers.end(), rng);
-        
+
         return HyperLogLog(p, q, registers);
     }
-    
+
     static HyperLogLog merge(const HyperLogLog& hll1, const HyperLogLog& hll2) {
         assert(hll1.p == hll2.p);
         assert(hll1.q == hll2.q);
-        
+
         unsigned char p = hll1.p;
         unsigned char q = hll1.q;
-        
+
         std::vector<unsigned char> registers(hll1.registers.size());
         for (size_t i = 0; i < registers.size(); ++i) {
             registers[i] = std::max(hll1.registers[i], hll2.registers[i]);
         }
-        
+
[…]

### 2016-08-07 c++/intersection_calculation.cpp

@@ -3,6 +3,7 @@
 //#############################
 
 #include "cardinality_estimation.hpp"
+#include "two_hyperloglog_statistic.hpp"
 #include "hyperloglog.hpp"
 
 #include <iostream>
@@ -47,15 +48,34 @@ template <typename T> void printToFile(const string& fileName, const T* data, si
     file << endl;
 }
 
-vector<int> readCounts(string s) {
-    stringstream stream(s);
-    int i;
-    vector<int> v;
-    while(stream >> i) {
-        v.push_back(i);
-    }
-    v.shrink_to_fit();
-    return v;
+void printHeaders(ofstream& os, string prefix, string postfix) {
+    os << prefix << "Min" << postfix << ";";
+    //os << prefix << "M3s" << postfix << ";";
+    //os << prefix << "M2s" << postfix << ";";
+    //os << prefix << "M1s" << postfix << ";";
+    //os << prefix << "Med" << postfix << ";";
+    //os << prefix << "P1s" << postfix << ";";
+    //os << prefix << "P2s" << postfix << ";";
+    //os << prefix << "P3s" << postfix << ";";
+    os << prefix << "Max" << postfix << ";";
+    os << prefix << "Mean" << postfix << ";";
+    os << prefix << "StdDev" << postfix << ";";
+}
+
+void calculateAndPrintStatistics(ofstream& os, std::vector<double> data) {
+    std::sort(data.begin(), data.end());
+    os << gsl_stats_min(&data[0], 1, data.size()) << ";";
+    //os << gsl_stats_quantile_from_sorted_data(&data[0], 1, data.size(), pm3s) << ";";
+    //os << gsl_stats_quantile_from_sorted_data(&data[0], 1, data.size(), pm2s) << ";";
+    //os << gsl_stats_quantile_from_sorted_data(&data[0], 1, data.size(), pm1s) << ";";
+    //os << gsl_stats_quantile_from_sorted_data(&data[0], 1, data.size(), 0.5) << ";";
+    //os << gsl_stats_quantile_from_sorted_data(&data[0], 1, data.size(), pp1s) << ";";
+    //os << gsl_stats_quantile_from_sorted_data(&data[0], 1, data.size(), pp2s) << ";";
+    //os << gsl_stats_quantile_from_sorted_data(&data[0], 1, data.size(), pp3s) << ";";
[…]

### 2016-08-07 c++/joint_hyperloglog_data_generator.cpp

@@ -2,7 +2,7 @@
 //# Copyright 2016 Otmar Ertl #
 //#############################
 
-#include "cardinalityestimation.hpp"
+#include "cardinality_estimation.hpp"
 #include "hyperloglog.hpp"
 
 #include <iostream>
@@ -29,17 +29,6 @@ const string dataPathName = "../data/hll/";
 const string intersectDataPathName = "../data/hll_joint/";
 const string jointCardFileName = intersectDataPathName + "joint_cardinalities.dat";
 
-template <typename T> void printToFile(const string& fileName, const T* data, size_t numSketches, bool append) {
-    ofstream file(fileName, (append)?(ofstream::out | ofstream::app):ofstream::out);
-    for (size_t s = 0; s < numSketches; ++s) {
-        if (s>0) {
-            file << " ";
-        }
-        file << data[s];
-    }
-    file << endl;
-}
-
 vector<int> readCounts(string s) {
     stringstream stream(s);
     int i;
@@ -60,8 +49,6 @@ string getFileName(const string& seed, int p) {
     return s.str();
 }
 
-
-
 string getOutputFileName(const long cardA, long cardB, long cardX, int p, int q) {
     stringstream s;
     s << setfill('0') << setw(11) << cardA;
@@ -77,79 +64,76 @@ string getOutputFileName(const long cardA, long cardB, long cardX, int p, int q)
     return s.str();
 }
 
-
-
-
 int main(int argc, char* argv[])
 {
     mt19937_64 rng(chrono::system_clock::now().time_since_epoch().count());
 
-    
+
     vector<string> seeds;
     {
         ifstream seedFile(seedFileName);
-        
+
         string line;
         while (getline(seedFile, line))
         {
             seeds.push_back(line);
         }
     }
-    
+
     cout << "seedSize "  << seeds.size() << endl;
-    
+
     vector<long> cardinalities;
     {
         ifstream cardFile(cardFileName);
-        
+
         string line;
         while (getline(cardFile, line))
         {
             cardinalities.push_back(stol(line));
         }
     }
-    
+
     cout << "cardinalitySize "  << cardinalities.size() << endl;
-        
+
     const unsigned char p = 20;
     const unsigned char q = 44;
-    
+
     #pragma omp parallel
     while(true) {
-        
+
         uint_fast64_t seed;
         #pragma omp critical
         {
             seed = rng();
         }
-        
+
         mt19937_64 rng2(seed);
[…]

### 2016-08-07 c++/two_hyperloglog_statistic.hpp

@@ -1,0 +1,158 @@
+//#############################
+//# Copyright 2016 Otmar Ertl #
+//#############################
+
+#ifndef _TWO_HYPERLOGLOG_STATISTIC_HPP_
+#define _TWO_HYPERLOGLOG_STATISTIC_HPP_
+
+#include <string>
+#include <sstream>
+
+class TwoHyperLogLogStatistic {
+
+    const std::vector<int> equal;
+    const std::vector<int> smaller1;
+    const std::vector<int> smaller2;
+    const std::vector<int> larger1;
+    const std::vector<int> larger2;
+
+    int getSize() const {
+        return static_cast<int>(equal.size());
+    }
+
+public:
+
+    TwoHyperLogLogStatistic(
+        const std::vector<int>& equal_,
+        const std::vector<int>& smaller1_,
+        const std::vector<int>& smaller2_,
+        const std::vector<int>& larger1_,
+        const std::vector<int>& larger2_) :
+
+        equal(equal_),
+        smaller1(smaller1_),
+        smaller2(smaller2_),
+        larger1(larger1_),
+        larger2(larger2_)
+    {
+
+        assert(smaller1_.size() == equal_.size());
+        assert(smaller2_.size() == equal_.size());
+        assert(larger1_.size() == equal_.size());
+        assert(larger2_.size() == equal_.size());
+
+    }
+
+    const std::vector<int>& getSmaller1Counts() const {
+        return smaller1;
+    }
+
+    const std::vector<int>& getSmaller2Counts() const {
+        return smaller2;
+    }
+
+    const std::vector<int>& getLarger1Counts() const {
+        return larger1;
+    }
+
+    const std::vector<int>& getLarger2Counts() const {
+        return larger2;
+    }
+
+    const std::vector<int>& getEqualCounts() const {
+        return equal;
+    }
+
+    const std::vector<int> get1Counts() const {
+        const int size = getSize();
+        std::vector<int> result(size);
+        for (int i = 0; i < size; ++i) {
+            result[i] = equal[i] + smaller1[i] + larger1[i];
+        }
+        return result;
+    }
+
+    const std::vector<int> get2Counts() const {
+        const int size = getSize();
+        std::vector<int> result(size);
+        for (int i = 0; i < size; ++i) {
+            result[i] = equal[i] + smaller2[i] + larger2[i];
+        }
+        return result;
+    }
+
[…]

### 2016-08-15 paper/paper.tex
added intersection estimation results

@@
 \usepackage{hyperref}
+\usepackage{csvsimple}
 \usepackage[capitalise,nameinlink,noabbrev]{cleveref}
@@
 \end{equation}
-Unfortunately, this approach can be very inaccurate. Due to estimation errors of HyperLogLog sketches the result of \eqref{equ:inclusion-exclusion} can be even negative.
-
+Unfortunately, this approach can be very inaccurate.
+Due to estimation errors of HyperLogLog sketches the result of \eqref{equ:inclusion-exclusion} can even be negative.
 Motivated by the good results we have obtained for a single HyperLogLog sketch using the maximum likelihood method in combination with the Poisson approximation, we are tempted to also use it for the estimation of set operations.
+Moreover, it was recently pointed out without special consideration of HyperLogLog sketches, that the application of the maximum likelihood method to the joint likelihood function of two probabilistic data structures which represent the intersection operands gives better intersection cardinality estimates \cite{Ting2016}.
+
 Assume two given  HyperLogLog sketches with register values $\boldsymbol{\symRegValVariate}_1$ and $\boldsymbol{\symRegValVariate}_2$ representing the sets $\symSetS_1$ and $\symSetS_2$, respectively.
 The goal is to find estimates for the cardinalities of the pair-wise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$.
-
 The Poisson approximation allows us to assume that the elements of $\symSetA$ are inserted into the HyperLogLog sketch representing $\symSetS_1$ at rate $\symPoissonRate_\symSetA$.
@@
 \begin{aligned}
-=&-
+&-
 \sum_{\symRegVal=0}^{\symRegRange}
@@
 
-TODO
+TODO see \cref{tbl:joint_estimation_cases}, \cref{tbl:joint_estimation_a}, \cref{tbl:joint_estimation_b}, \cref{tbl:joint_estimation_x}
+
+\begin{table}
+\centering
+\caption{The list of cases for which the cardinalities $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$ have been estimated given two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing $\symSetA\cup\symSetX$ and  $\symSetB\cup\symSetX$, respectively.}
+\label{tbl:joint_estimation_cases}
+\csvreader[tabular=r|r|r|r|r|r, table head=\# & $|\symSetA|$ & $|\symSetB|$ & $|\symSetX|$ & $\frac{|\symSetX|}{|\symSetA\cup\symSetB\cup\symSetX|}$ & $\log_{10}\!\left(\frac{|\symSetA\cup\symSetX|}{|\symSetB\cup\symSetX|}\right)$\\\hline]
+{intersection.csv}{trueCardA=\trueCardA,trueCardB=\trueCardB,trueCardX=\trueCardX,trueJaccardIdx=\trueJaccardIdx ,trueLogRatio=\trueLogRatio}{
+\thecsvrow & \trueCardA & \trueCardB & \trueCardX & \trueJaccardIdx & \trueLogRatio
+}
+\end{table}
+
+\begin{table}
+\centering
+\caption{The mean and the standard deviation for the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetA|$ using the inclusion-exlusion (IE) principle and the maximum likelihood method (ML), respectively.}
+\label{tbl:joint_estimation_a}
+\csvreader[tabular=r|r|r|r|r, table head=\# & IE mean & IE stdev & ML mean & ML stdev\\\hline]
+{intersection.csv}{inclExclMeanA=\inclExclMeanA,inclExclStdDevA=\inclExclStdDevA,
+maxLikeMeanA=\maxLikeMeanA,maxLikeStdDevA=\maxLikeStdDevA}{
+\thecsvrow & \inclExclMeanA & \inclExclStdDevA & \maxLikeMeanA & \maxLikeStdDevA
+}
+\end{table}
+
+\begin{table}
+\centering
+\caption{The mean and the standard deviation for the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetB|$ using the inclusion-exlusion (IE) principle and the maximum likelihood method (ML), respectively.}
+\label{tbl:joint_estimation_b}
+\csvreader[tabular=r|r|r|r|r, table head=\# & IE mean & IE stdev & ML mean & ML stdev\\\hline]
+{intersection.csv}{inclExclMeanB=\inclExclMeanB,inclExclStdDevB=\inclExclStdDevB,
+maxLikeMeanB=\maxLikeMeanB,maxLikeStdDevB=\maxLikeStdDevB}{
+\thecsvrow & \inclExclMeanB & \inclExclStdDevB & \maxLikeMeanB & \maxLikeStdDevB
+}
+\end{table}
+
+\begin{table}
+\centering
+\caption{The mean and the standard deviation for the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetX|$ using the inclusion-exlusion (IE) principle and the maximum likelihood method (ML), respectively.}
+\label{tbl:joint_estimation_x}
+\csvreader[tabular=r|r|r|r|r, table head=\# & IE mean & IE stdev & ML mean & ML stdev\\\hline]
+{intersection.csv}{inclExclMeanX=\inclExclMeanX,inclExclStdDevX=\inclExclStdDevX,
+maxLikeMeanX=\maxLikeMeanX,maxLikeStdDevX=\maxLikeStdDevX}{
+\thecsvrow & \inclExclMeanX & \inclExclStdDevX & \maxLikeMeanX & \maxLikeStdDevX
+}
+\end{table}
 

### 2016-08-15 c++/benchmark.cpp

@@ -230,18 +230,18 @@ void run(const int p, const int q, const string& resultsFileName) {
         }*/
 
         outputFile << cardinality;
-        //outputFile << " " << maxLikelihoodWithRegisterScan;
-        outputFile << " " << maxLikelihoodWithoutRegisterScan;
-        //outputFile << " " << flajoletWithRegisterScan;
-        //outputFile << " " << flajoletWithoutRegisterScan;
-        outputFile << " " << outerLoopIterationsCountSum/(double)dataSize;
-        outputFile << " " << innerLoop1IterationsCountSum/(double)dataSize;
-        outputFile << " " << innerLoop2IterationsCountSum/(double)dataSize;
-        outputFile << " " << logEvaluationCountSum/(double)dataSize;
-        outputFile << " " << flajoletCorrectedWithoutRegisterScan;
-        outputFile << " " << flajoletCorrectedSmallCorrectionIterationsSum/(double)dataSize;
-        outputFile << " " << flajoletCorrectedLargeCorrectionIterationsSum/(double)dataSize;
-        outputFile << " " << flajoletCorrectedWithoutRegisterScanPrecalculated;
+        //outputFile << "," << maxLikelihoodWithRegisterScan;
+        outputFile << "," << maxLikelihoodWithoutRegisterScan;
+        //outputFile << "," << flajoletWithRegisterScan;
+        //outputFile << "," << flajoletWithoutRegisterScan;
+        outputFile << "," << outerLoopIterationsCountSum/(double)dataSize;
+        outputFile << "," << innerLoop1IterationsCountSum/(double)dataSize;
+        outputFile << "," << innerLoop2IterationsCountSum/(double)dataSize;
+        outputFile << "," << logEvaluationCountSum/(double)dataSize;
+        outputFile << "," << flajoletCorrectedWithoutRegisterScan;
+        outputFile << "," << flajoletCorrectedSmallCorrectionIterationsSum/(double)dataSize;
+        outputFile << "," << flajoletCorrectedLargeCorrectionIterationsSum/(double)dataSize;
+        outputFile << "," << flajoletCorrectedWithoutRegisterScanPrecalculated;
         outputFile << endl;
     }
 

### 2016-08-15 c++/cardinality_calculator.cpp

@@ -30,7 +30,7 @@ template <typename T> void printToFile(const string& fileName, const T* data, si
     for (size_t c = 0; c < numCardinalities; ++c) {
         for (size_t s = 0; s < numSketches; ++s) {
             if (s>0) {
-                file << " ";
+                file << ",";
             }
             file << data[c*numSketches+s];
         }

### 2016-08-15 c++/cardinality_estimation.hpp

@@ -31,6 +31,50 @@ int getPFromCounts(const std::vector<int>& c) {
     return getPFromNumberRegisters(m);
 }
 
+
+double estimateCardinalityFromCdf(const std::vector<double> cdf, int m) {
+    int q = cdf.size() - 1;
+
+    if (cdf[0] >= 1.) std::numeric_limits<double>::infinity();
+
+    double lowerTail = 0;
+    {
+        double arg = cdf[0];
+        double pow2 = 1;
+        while(true) {
+            const double lowerTailBak = lowerTail;
+            lowerTail += pow2 * arg;
+            if (lowerTail == lowerTailBak) break;
+            pow2 += pow2;
+            arg *= arg;
+        }
+    }
+
+    double upperTail = 0;
+    {
+        double arg = std::sqrt(cdf[cdf.size()-1]);
+        double pow2 = 0.5;
+        while(true) {
+            const double upperTailBak = upperTail;
+            upperTail += pow2 * arg;
+            if (upperTail == upperTailBak) break;
+            pow2 *= 0.5;
+            arg = std::sqrt(arg);
+        }
+    }
+
+    double sum = upperTail;
+    for (int k = q; k >= 1; --k) {
+        sum += cdf[k];
+        sum *= 0.5;
+    }
+
+    sum += lowerTail;
+
+    return (m/std::log(2))/sum;
+
+}
+
 class MaxLikelihoodEstimator {
     const double eps = 1e-2;
     const int p;
@@ -254,7 +298,6 @@ class CorrectedRawEstimator {
         return result;
     }
 
-
 public:
 
     CorrectedRawEstimator(const int p_, const int q_) : p(p_), q(q_) {}
@@ -420,6 +463,74 @@ void eval_joint_log_likelihood_function_and_derivatives(
     const double expPhiB = std::exp(phiB);
     const double expPhiX = std::exp(phiX);
 
+    const double expPhiAdivAX = expPhiA / (expPhiA + expPhiX);   // TODO handle division by 0?
+    const double expPhiXdivAX = expPhiX / (expPhiA + expPhiX);
+    const double expPhiBdivBX = expPhiB / (expPhiB + expPhiX);
[…]

### 2016-08-15 c++/intersection_calculation.cpp

@@ -33,8 +33,9 @@ const double median = 0.5;
 
 
 const string intersectDataPathName = "../data/hll_joint/";
+const string paperPathName = "../paper/";
 const string jointCardFileName = intersectDataPathName + "joint_cardinalities.dat";
-const string resultsFileName = intersectDataPathName + "results.dat";
+const string resultsFileName = paperPathName + "intersection.csv";
 
 
 template <typename T> void printToFile(const string& fileName, const T* data, size_t numSketches, bool append) {
@@ -49,33 +50,33 @@ template <typename T> void printToFile(const string& fileName, const T* data, si
 }
 
 void printHeaders(ofstream& os, string prefix, string postfix) {
-    os << prefix << "Min" << postfix << ";";
-    //os << prefix << "M3s" << postfix << ";";
-    //os << prefix << "M2s" << postfix << ";";
-    //os << prefix << "M1s" << postfix << ";";
-    //os << prefix << "Med" << postfix << ";";
-    //os << prefix << "P1s" << postfix << ";";
-    //os << prefix << "P2s" << postfix << ";";
-    //os << prefix << "P3s" << postfix << ";";
-    os << prefix << "Max" << postfix << ";";
-    os << prefix << "Mean" << postfix << ";";
-    os << prefix << "StdDev" << postfix << ";";
+    os << prefix << "Min" << postfix << ",";
+    os << prefix << "M3s" << postfix << ",";
+    os << prefix << "M2s" << postfix << ",";
+    os << prefix << "M1s" << postfix << ",";
+    os << prefix << "Med" << postfix << ",";
+    os << prefix << "P1s" << postfix << ",";
+    os << prefix << "P2s" << postfix << ",";
+    os << prefix << "P3s" << postfix << ",";
+    os << prefix << "Max" << postfix << ",";
+    os << prefix << "Mean" << postfix << ",";
+    os << prefix << "StdDev" << postfix << ",";
 }
 
 void calculateAndPrintStatistics(ofstream& os, std::vector<double> data) {
     std::sort(data.begin(), data.end());
-    os << gsl_stats_min(&data[0], 1, data.size()) << ";";
-    //os << gsl_stats_quantile_from_sorted_data(&data[0], 1, data.size(), pm3s) << ";";
-    //os << gsl_stats_quantile_from_sorted_data(&data[0], 1, data.size(), pm2s) << ";";
[…]

### 2016-08-15 c++/joint_hyperloglog_data_generator.cpp

@@ -2,7 +2,7 @@
 //# Copyright 2016 Otmar Ertl #
 //#############################
 
-#include "cardinality_estimation.hpp"
+#include "two_hyperloglog_statistic.hpp"
 #include "hyperloglog.hpp"
 
 #include <iostream>

### 2016-08-15 c++/two_hyperloglog_statistic.hpp

@@ -7,6 +7,9 @@
 
 #include <string>
 #include <sstream>
+#include <vector>
+#include <algorithm>
+#include <cassert>
 
 class TwoHyperLogLogStatistic {
 

### 2016-08-15 python/benchmark-charts.py

@@ -17,9 +17,9 @@ filename_12_52 = '../data/benchmark_results_12_52.dat'
 def readData(filename):
     csv.field_size_limit(sys.maxsize)
     with open(filename , 'r') as file:
-        reader = csv.reader(file, skipinitialspace=True,delimiter=' ')
+        reader = csv.reader(file, skipinitialspace=True)
         return numpy.array([[float(e) for e in r] for r in reader])
-        
+
 
 plt.rc('text', usetex=True)
 plt.rc('font', family='serif')

### 2016-08-15 python/distribution_charts.py

@@ -14,7 +14,7 @@ dataDir = '../data/'
 def readData(filename):
     csv.field_size_limit(sys.maxsize)
     with open(dataDir + filename , 'r') as file:
-        reader = csv.reader(file, skipinitialspace=True,delimiter=' ')
+        reader = csv.reader(file, skipinitialspace=True)
         return numpy.array([[float(e) for e in r] for r in reader][1:])
 
 def createFigure(p, q, dataSource, figureName):

### 2016-08-21 c++/cardinality_estimation.hpp
removed obsolete class

@@ -215,32 +215,6 @@ double flajoletRawEstimate(const std::vector<int>& c, int m) {
     return (alpha*m)*m/z;
 }
 
-class OptimalLinearCountingEstimator {
-
-    const int p;
-    const int m = 1 << p;
-    const std::vector<double> results = initResults(m);
-
-    static std::vector<double> initResults(int m) {
-
-        std::vector<double> results(m+1);
-        results[m] = 0;
-        for (int c = m-1; c >= 0; --c) {
-            results[c] = results[c + 1] + m/(double)c;
-        }
-        return results;
-    }
-
-public:
-
-    OptimalLinearCountingEstimator(const int p_) : p(p_) {}
-
-    double operator()(const std::vector<int>& c) const {
-        return results[c[0]];
-    }
-};
-
-
 class CorrectedRawEstimator {
     const int p;
     const int q;

### 2016-08-21 paper/paper.tex
some minor changes, updated .gitignore

@@
 \begin{equation}
+\label{equ:sufficient_joint_statistic}
 \begin{aligned}
@@
 As expected, these two functions are equivalent to the individual log-likelihood functions \eqref{equ:log_likelihood_single} of both HyperLogLog sketches.
+
+The inclusion-exclusion method does not use all the information given by the sufficient statistic \eqref{equ:sufficient_joint_statistic}, because the estimates are functions of the three vectors $(\boldsymbol\symCountVariate^{<}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{1})$,
+$(\boldsymbol\symCountVariate^{<}_{2} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2})$, and 
+$(\boldsymbol\symCountVariate^{>}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2})$.
+In contrast, the maximum-likelihood method uses all values of the sufficient statistic \eqref{equ:sufficient_joint_statistic}.
 

### 2016-08-30 paper/paper.tex
fixed equation
added script for comparison chart, fixed some typos

@@
 \label{sec:poisson_approximation}
-Due to the statistical dependence of the register values, the probability mass function \eqref{equ:multinomialProbabilityMass} is difficult to analyze.
+Due to the statistical dependence of the register values, the probability mass function \eqref{equ:multinomialProbabilityMass} is difficult for further analysis.
 Therefore, a Poisson model can be used \cite{Flajolet2007}, which assumes that the cardinality itself is distributed according to a Poisson distribution
@@
 Here $\symCount_\symRegVal$ denotes the multiplicity of value $\symRegVal$  in the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$.
-The final factorization shows that under the Poisson values the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ are independent and identically distributed.
+The final factorization shows that under the Poisson model the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ are independent and identically distributed.
 The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is defined by
@@
 \label{sec:derivation_raw_estimator}
-In order to better understand why the raw estimator fails for small and large cardinalities, we start with a brief and simple derivation without the restriction tolarge cardinalities ($\symCardinality\rightarrow\infty$) and without using complex analysis as in \cite{Flajolet2007}.
+In order to better understand why the raw estimator fails for small and large cardinalities, we start with a brief and simple derivation without the restriction to large cardinalities ($\symCardinality\rightarrow\infty$) and without using complex analysis as in \cite{Flajolet2007}.
 
@@
 As we have already shown, this estimator is approximately unbiased, because all $\symRegValVariate'_\symIndexI$ follow the assumed distribution.
-It would be even sufficient, if we know the multiplicity $\symCountVariate'_\symRegVal$
+It would be even sufficient, if we knew the multiplicity $\symCountVariate'_\symRegVal$
 of each $\symRegVal\in\mathbb{Z}$ in $\lbrace\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg\rbrace$, $\symCountVariate'_\symRegVal := \left|\lbrace \symIndexI\vert \symRegVal = \symRegValVariate'_\symIndexI\rbrace\right|$, because the raw estimator can be also written as
@@
 The average computation time was determined by cycling over these multiplicity vectors and passing them as input to the algorithm.
-For each evaluated cardinality value the average execution time was calculated after 100 cycles which corresponds to 100\,000 algorithm executions for each cardinality value.
+For each evaluated cardinality value the average execution time was calculated after 100 cycles which corresponds to 100,000 algorithm executions for each cardinality value.
 The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:corrected_raw_avg_exec_time}.
@@
 \section{Maximum likelihood estimation}
-Moreover, we know from \cref{sec:depoissonization} that any unbiased estimator for the Poisson parameter is also an unbiased estimator for the cardinality.
-We know that under suitable regularity conditions of the probability mass function the maximum likelihood estimator is asymptotically efficient \cite{Casella2002}.
+We know from \cref{sec:depoissonization} that any unbiased estimator for the Poisson parameter is also an unbiased estimator for the cardinality.
+Moreover, we know that under suitable regularity conditions of the probability mass function the maximum likelihood estimator is asymptotically efficient \cite{Casella2002}.
 This means, if the number of registers $\symNumReg$ is large enough, the maximum likelihood method should give us an unbiased estimator for the cardinality.
@@
 \label{sec:comp_ml_estimate}
-Since $\symFunc$ is concave and increasing, both, Newton-Raphson iteration and the secant method, will converge to the root, if the function is negative for the starting points.
+Since $\symFunc$ is concave and increasing, both, Newton-Raphson iteration and the secant method, will converge to the root, provided that the function is negative for the starting points.
 In the following we start from the secant method to derive the new cardinality estimation algorithm.
@@
 \end{equation}
-If $\symX_0 = 0$ with $\symFunc(\symX_0)=-\left(\symNumReg-\symCountVariate_0\right)$, and $\symX_1$ is equal to one of the derived lower bounds \eqref{equ:strong_lower_bound} or \eqref{equ:weak_lower_bound}, the sequence $(\symX_0, \symX_1, \symX_2, \ldots)$ is montone increasing.
+If we set $\symX_0$ equal to 0, for which  $\symFunc(\symX_0)=-\left(\symNumReg-\symCountVariate_0\right)$, and $\symX_1$ equal to one of the derived lower bounds \eqref{equ:strong_lower_bound} or \eqref{equ:weak_lower_bound}, the sequence $(\symX_0, \symX_1, \symX_2, \ldots)$ is montone increasing.
 Using the definitions
@@
 \right)
-\frac{e^{\symPhi_\symSetBSuffix}}{2^{\symRegVal}}
+\frac{e^{\symPhi_\symSetXSuffix}}{2^{\symRegVal}}
 +

### 2016-08-30 python/comparison.py

@@ -1,0 +1,80 @@
+#############################
+# Copyright 2016 Otmar Ertl #
+#############################
+
+from numpy import ones, array
+from math import expm1, cosh, tanh, sinh, log, erf, sqrt
+import matplotlib.pyplot as plt
+import csv
+import sys
+import numpy
+
+dataDir = '../data/'
+
+def readData(filename):
+    csv.field_size_limit(sys.maxsize)
+    with open(dataDir + filename , 'r') as file:
+        reader = csv.reader(file, skipinitialspace=True)
+        return numpy.array([[float(e) for e in r] for r in reader][1:])
+
+config = [
+["12_20_max_likelihood_estimates.dat", "blue", False, "maximum likelihood estimator"],
+#["12_20_flajolet_mid_range_estimates.dat", "gray", False, "raw estimator"],
+["12_20_flajolet_mid_range_estimates.dat", "red", True, "raw estimator (bias corrected)"],
+["12_20_flajolet_small_range_estimates.dat", "black", False, "linear counting estimator"],
+["12_20_corrected_raw_estimates.dat", "green", False, "corrected raw estimator"]
+]
+
+cardinalities = readData('cardinalities.dat').flatten()
+
+def applyBiasCorrection(data, cardinalities):
+    datatmp = numpy.copy(data)
+    for i in range(0, 100):
+        data_mean = numpy.mean(datatmp, axis=1) - cardinalities
+        datatmp -= numpy.interp(datatmp, cardinalities, data_mean)
+    return datatmp
+
+def calculate(filename, doBiasCorrection):
+
+    data = readData(filename)
+    if (doBiasCorrection):
+        data = applyBiasCorrection(data, cardinalities)
+    relative_error = data/cardinalities[:,None]-1.
+    datamean = numpy.mean(relative_error, axis=1)
+    datastd = numpy.std(relative_error, axis=1)
+    return (datamean, datastd)
+
+fig1, ax1 = plt.subplots(subplot_kw=dict(xlim=[cardinalities[0],cardinalities[-1]], ylim=[0, 0.02], xscale='log', yscale='linear'))
+
+fig1.set_size_inches(10, 5)
+ax1.yaxis.grid(b=True, which='major', color='gray', linestyle='--',dash_capstyle='round')
[…]

### 2016-08-30 python/distribution_charts.py

@@ -60,7 +60,7 @@ def createFigure(p, q, dataSource, figureName):
     data50 = numpy.divide(numpy.percentile(data, 50, axis=1), cardinalities)-1.
 
     ax1.plot(cardinalities, datamean, color='black', linewidth=0.75, label='mean', dashes=[1, 2], dash_capstyle='round')
-    ax1.plot(cardinalities, datastdpos, color='black', linewidth=0.75, linestyle = 'dotted', label='stddev', dashes = [1, 4], dash_capstyle='round')
+    ax1.plot(cardinalities, datastdpos, color='black', linewidth=0.75, linestyle = 'dotted', label='stdev', dashes = [1, 4], dash_capstyle='round')
     ax1.plot(cardinalities, datastdneg, color='black', linewidth=0.75, linestyle = 'dotted', dashes = [1, 4], dash_capstyle='round')
     ax1.plot(cardinalities, data50, color='black', linewidth=0.75, linestyle = 'solid', label='median')
     ax1.fill_between(cardinalities, datam3s, datap3s, facecolor='#dddddd', edgecolor='none')

### 2016-09-01 python/comparison.py
improved stdev comaprison chart

@@ -8,6 +8,10 @@ import matplotlib.pyplot as plt
 import csv
 import sys
 import numpy
+from scipy import interpolate
+from mpl_toolkits.axes_grid1.inset_locator import mark_inset
+from mpl_toolkits.axes_grid1.inset_locator import zoomed_inset_axes
+from mpl_toolkits.axes_grid1.inset_locator import mark_inset
 
 dataDir = '../data/'
 
@@ -29,7 +33,7 @@ cardinalities = readData('cardinalities.dat').flatten()
 
 def applyBiasCorrection(data, cardinalities):
     datatmp = numpy.copy(data)
-    for i in range(0, 100):
+    for i in range(0, 50):
         data_mean = numpy.mean(datatmp, axis=1) - cardinalities
         datatmp -= numpy.interp(datatmp, cardinalities, data_mean)
     return datatmp
@@ -44,13 +48,16 @@ def calculate(filename, doBiasCorrection):
     datastd = numpy.std(relative_error, axis=1)
     return (datamean, datastd)
 
-fig1, ax1 = plt.subplots(subplot_kw=dict(xlim=[cardinalities[0],cardinalities[-1]], ylim=[0, 0.02], xscale='log', yscale='linear'))
+fig1, ax1 = plt.subplots(subplot_kw=dict(xlim=[cardinalities[0],cardinalities[-1]], ylim=[0.0, 0.025], xscale='log', yscale='linear'))
 
 fig1.set_size_inches(10, 5)
 ax1.yaxis.grid(b=True, which='major', color='gray', linestyle='--',dash_capstyle='round')
 ax1.xaxis.grid(b=True, which='major', color='gray', linestyle='--',dash_capstyle='round')
 ax1.set_xlabel("cardinality")
 ax1.set_ylabel("stdev")
+
+axins = zoomed_inset_axes(ax1, 2.5, loc=2) #loc='lower left', bbox_to_anchor=(100, 0.01))
+
 
 fig2, ax2 = plt.subplots(subplot_kw=dict(xlim=[cardinalities[0],cardinalities[-1]], ylim=[-0.02,0.02], xscale='log', yscale='linear'))
 
@@ -68,9 +75,17 @@ for c in config:
     (mean, stddev) = calculate(c[0], c[2])
     ax1.plot(cardinalities, stddev, color=c[1], linewidth=0.75, label=c[3])
[…]

### 2016-09-01 paper/paper.tex
added new figure to paper

@@
 
+The corrected raw estimator beats the precision of methods that use bias correction \cite{Heule2013,Rhodes2015,Sanfilippo2014}.
+\cref{fig:stdev_comparison} shows the standard deviation of the raw estimator after bias correction.
+At the transition region to small cardinalities, for which all previous methods switch over to the linear counting estimator at some point, the corrected raw estimator has smaller variance.
+Obviously, the previous approaches cannot do better than given by the minimum of both curves for linear counting and raw estimator.
+In practice, the standard deviation is even larger, because the choice between both estimators must be made based on an estimate and not on the true cardinality, for which the intersection point of both curves represents the ideal transition point.
+
 \begin{figure}
@@
 \label{fig:raw_corrected_estimation_error_12_20}
+\end{figure}
+
+\begin{figure}
+\centering
+\includesvg[width=1\textwidth]{stdev_comparison}
+\caption{Standard deviations of different cardinality estimators as functions of the real cardinality.}
+\label{fig:stdev_comparison}
 \end{figure}

### 2016-09-12 c++/cardinality_estimation.hpp
performance improvement and different initial values for joint cardinality estimation

@@ -420,6 +420,18 @@ void inclusionExclusionTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& j
     cardinalityX = std::max(0., cardinalityBX + cardinalityAX - cardinalityABX);
 }
 
+void calcExp(double x, double& a, double& b) {
+    const double ln2 = std::log(2);
+    if (x >= ln2) {
+        a = std::exp(-x);
+        b = 1 - a;
+    }
+    else {
+        b = -std::expm1(-x);
+        a = 1 - b;
+    }
+}
+
 // the minimum of this function gives the max likelihood estimate
 void eval_joint_log_likelihood_function_and_derivatives(
     const TwoHyperLogLogStatistic& jointStatistic,
@@ -483,25 +495,19 @@ void eval_joint_log_likelihood_function_and_derivatives(
         const int cLarger1  = jointStatistic.getLarger1Counts()[k];
 
         if (cLarger1 > 0 || cEqual > 0) {
-            expm1a[k] = -std::expm1(-expPhiApow2k[k]);
-            expa[k] = std::exp(-expPhiApow2k[k]);
+            calcExp(expPhiApow2k[k], expa[k], expm1a[k]);
         }
         if (cLarger2 > 0 || cEqual > 0) {
-            expm1b[k] = -std::expm1(-expPhiBpow2k[k]);
-            expb[k] = std::exp(-expPhiBpow2k[k]);
+            calcExp(expPhiBpow2k[k], expb[k], expm1b[k]);
         }
         if (cSmaller1 > 0 || cEqual > 0) {
-            expm1ax[k] = -std::expm1(-expPhiAXpow2k[k]);
-            expax[k] = std::exp(-expPhiAXpow2k[k]);
+            calcExp(expPhiAXpow2k[k], expax[k], expm1ax[k]);
         }
         if (cSmaller2 > 0 || cEqual > 0) {
-            expm1bx[k] = -std::expm1(-expPhiBXpow2k[k]);
-            expbx[k] = std::exp(-expPhiBXpow2k[k]);
+            calcExp(expPhiBXpow2k[k], expbx[k], expm1bx[k]);
         }
-
         if (cEqual > 0) {
-            expm1x[k] = -std::expm1(-expPhiXpow2k[k]);
-            expx[k] = std::exp(-expPhiXpow2k[k]);
+            calcExp(expPhiXpow2k[k], expx[k], expm1x[k]);
         }
     }
 
[…]

### 2016-09-12 paper/paper.tex
added more cases demonstrating  joint cardinality estimation

@@
 \usepackage{hyperref}
+\usepackage{siunitx}
 \usepackage{csvsimple}
@@
 
+\sisetup{
+fixed-exponent = 2,
+scientific-notation = fixed,
+}
+
 \title{New cardinality estimation algorithms for HyperLogLog sketches}
@@
 Therefore, the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ can be seen as the result after applying this transformation to a sample $\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg$ of the distribution described by \eqref{equ:assumed_register_val_distribution}.
-If all registers values of the HyperLogLog sketch lie in the range $[1,\symRegRange]$, they are identical to the values $\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg$.
+If all registers values of the HyperLogLog sketch fall into the range $[1,\symRegRange]$, they must  be identical to the values $\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg$.
 In other words, the observed register values are also a plausible sample of the assumed distribution described by \eqref{equ:assumed_register_val_distribution} in this case.
-Hence, as long as all or at least most register values are in the range $[1,\symRegRange]$ the approximation of \eqref{equ:register_value_distribution} by \eqref{equ:assumed_register_val_distribution} is valid.
+Hence, as long as all or at least most register values are in the range $[1,\symRegRange]$, the approximation of \eqref{equ:register_value_distribution} by \eqref{equ:assumed_register_val_distribution} is valid.
 This explains why the raw estimator works best for intermediate cardinalities.
-However, for small and large cardinalities many registers have values equal to 0 or $\symRegRange+1$, respectively, which cannot be described by \eqref{equ:assumed_register_val_distribution} and which finally leads to the observed bias.
+However, for small and large cardinalities many registers have values equal to 0 or $\symRegRange+1$, respectively, which cannot be described by \eqref{equ:assumed_register_val_distribution} and finally leads to the observed bias.
 
 \subsection{Corrections to the raw estimator}
-If we knew the values $\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg$ for which the transformation \eqref{equ:dist_transformation} led to the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$, we would be able to estimate $\symPoissonRate$ using
+If we knew the values $\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg$ for which the transformation \eqref{equ:dist_transformation} led to the observed register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$, we would be able to estimate $\symPoissonRate$ using
 \begin{equation}
@@
 \end{equation}
-Due to \eqref{equ:dist_transformation}, the multiplicities $\symCountVariate'_\symRegVal$ and the multiplicities $\symCountVariate_\symRegVal$ for the register values have following relationships
+Due to \eqref{equ:dist_transformation}, the multiplicities $\symCountVariate'_\symRegVal$ and the multiplicities $\symCountVariate_\symRegVal$ for the observed register values have following relationships
 \begin{equation}
@@
 \includesvg[width=1\textwidth]{stdev_comparison}
-\caption{Standard deviations of different cardinality estimators as functions of the real cardinality.}
+\caption{Standard deviations of different cardinality estimators as functions of the true cardinality.}
 \label{fig:stdev_comparison}
@@
 \end{equation}
-These $5\symNumReg$ values are sufficient for estimating $\symPoissonRate_\symSetA$, $\symPoissonRate_\symSetB$, and $\symPoissonRate_\symSetX$.
+These $5\symNumReg$ values are sufficient for estimating $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$.
 Actually, the set of these constants can be further reduced, because $\symCountVariate^{>}_{10}=\symCountVariate^{>}_{20}=
@@
 \label{tbl:joint_estimation_cases}
-\csvreader[tabular=r|r|r|r|r|r, table head=\# & $|\symSetA|$ & $|\symSetB|$ & $|\symSetX|$ & $\frac{|\symSetX|}{|\symSetA\cup\symSetB\cup\symSetX|}$ & $\log_{10}\!\left(\frac{|\symSetA\cup\symSetX|}{|\symSetB\cup\symSetX|}\right)$\\\hline]
+\csvreader[before reading=\footnotesize, tabular=r|r|r|r|r|r, table head=\# & $|\symSetA|$ & $|\symSetB|$ & $|\symSetX|$ & $\frac{|\symSetX|}{|\symSetA\cup\symSetB\cup\symSetX|}$ & $\log_{10}\!\left(\frac{|\symSetA\cup\symSetX|}{|\symSetB\cup\symSetX|}\right)$\\\hline]
 {intersection.csv}{trueCardA=\trueCardA,trueCardB=\trueCardB,trueCardX=\trueCardX,trueJaccardIdx=\trueJaccardIdx ,trueLogRatio=\trueLogRatio}{
@@
 \label{tbl:joint_estimation_a}
-\csvreader[tabular=r|r|r|r|r, table head=\# & IE mean & IE stdev & ML mean & ML stdev\\\hline]
+\csvreader[before reading=\footnotesize, tabular=r|r|r|r|r, table head=\# & IE mean & IE stdev & ML mean & ML stdev\\\hline]
 {intersection.csv}{inclExclMeanA=\inclExclMeanA,inclExclStdDevA=\inclExclStdDevA,
@@
 \label{tbl:joint_estimation_b}
-\csvreader[tabular=r|r|r|r|r, table head=\# & IE mean & IE stdev & ML mean & ML stdev\\\hline]
+\csvreader[before reading=\footnotesize, tabular=r|r|r|r|r, table head=\# & IE mean & IE stdev & ML mean & ML stdev\\\hline]
 {intersection.csv}{inclExclMeanB=\inclExclMeanB,inclExclStdDevB=\inclExclStdDevB,
@@
 \label{tbl:joint_estimation_x}
-\csvreader[tabular=r|r|r|r|r, table head=\# & IE mean & IE stdev & ML mean & ML stdev\\\hline]
+\csvreader[before reading=\footnotesize, tabular=r|r|r|r|r, table head=\# & IE mean & IE stdev & ML mean & ML stdev\\\hline]
 {intersection.csv}{inclExclMeanX=\inclExclMeanX,inclExclStdDevX=\inclExclStdDevX,

### 2016-09-12 c++/intersection_calculation.cpp

@@ -23,6 +23,8 @@ const int minP = 4;
 const int maxP = 22;
 const int maxPplusQ = 64;
 
+const int analyzeP = 20;
+
 const double pm3s = (1.-erf(3./sqrt(2.)))/2.;
 const double pp3s = (1.+erf(3./sqrt(2.)))/2.;
 const double pm2s = (1.-erf(2./sqrt(2.)))/2.;
@@ -118,6 +120,8 @@ int main(int argc, char* argv[])
     resultsFile << "maxNumIterations,";
     resultsFile << "iterationAbortedCount" << endl;
 
+    int evaluationCounter = 0;
+
     for(const string& fileName : jointCardFileNames) {
 
         const long trueCardA = stol(fileName.substr(0, 11));
@@ -129,7 +133,7 @@ int main(int argc, char* argv[])
         const double jaccardIndex = trueCardX/static_cast<double>(trueCardA+trueCardB+trueCardX);
         const double logRatio = std::log10(trueCardA) - std::log10(trueCardB);
 
-        if (jaccardIndex < 1e-3 || jaccardIndex > 0.1 || std::fabs(logRatio) > 1) continue; // TODO filter interesting cardinalities
+        if (analyzeP != p || jaccardIndex < 1e-3 || jaccardIndex > 0.1 || std::fabs(logRatio) > 2) continue; // TODO filter cardinality combinations for table
 
         cout << trueCardA << " " << trueCardB << " " << trueCardX << " " << p << " " << q << endl;
 
@@ -173,7 +177,6 @@ int main(int argc, char* argv[])
                 bool maxNumIterationsReached = false;
                 bool iterationAborted = false;
                 int numIterations = 0;
-                //analyticalJointHyperLogLogEstimator(jointStatistic, estCardA, estCardB, estCardX);
                 maxLikelihoodTwoHyperLogLogEstimation(jointStatistic, estCardA, estCardB, estCardX, maxNumIterationsReached, iterationAborted, numIterations);
                 maxLikeEstimatedCardA.push_back(estCardA/trueCardA-1.);
                 maxLikeEstimatedCardB.push_back(estCardB/trueCardB-1.);
@@ -183,9 +186,11 @@ int main(int argc, char* argv[])
                 if (iterationAborted) iterationAbortedCount += 1;
                 numIterationsTotal += numIterations;
                 maxNumIterations = std::max(maxNumIterations, numIterations);
+
             }
 
             size += 1;
[…]

### 2016-09-12 c++/cardinality_estimation.hpp
use inclusion-exclusion principle to calculate initial values for BFGS algorithm

@@ -618,9 +618,9 @@ double log_likelihood_function_value_for_gsl(const gsl_vector *phi, void *params
 
 void maxLikelihoodTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& jointStatistic, double& cardinalityA, double& cardinalityB, double& cardinalityX, bool& maxNumIterationsReached, bool& iterationAborted, int& numIterations) {
 
-    const double eps = 1e-5;
-
-    const int maxNumIterations = 100000;
+    const double eps = 1e-6;
+    const double initalStepFactor = 2;
+    const int maxNumIterations = 10000;
 
     maxNumIterationsReached = false;
     iterationAborted = false;
@@ -629,23 +629,24 @@ void maxLikelihoodTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& jointS
 
     const MaxLikelihoodEstimator estimator(jointStatistic.getP(), jointStatistic.getQ());
 
+    double cardinalityAX = estimator(jointStatistic.get1Counts());
+    double cardinalityBX = estimator(jointStatistic.get2Counts());
+
     // special handling, the sets A u X and B u X are disjoint, therefore X = O
     if (jointStatistic.getMinCounts()[0] == m) {
         cardinalityX = 0;
-        cardinalityA = estimator(jointStatistic.get1Counts());
-        cardinalityB = estimator(jointStatistic.get2Counts());
+        cardinalityA = cardinalityAX;
+        cardinalityB = cardinalityBX;
         return;
     }
-
     // for all remaining cases the maximum likelihood estimates for A,B,X will be positive
 
-    // set initial vector
+    double cardinalityABX = estimator(jointStatistic.getMaxCounts());
 
-    double initCardinality = estimator(jointStatistic.getMaxCounts());
-
-    double initCardinalityA = initCardinality; //estimator(jointStatistic.get1Counts());
-    double initCardinalityB = initCardinality; //estimator(jointStatistic.get2Counts());
-    double initCardinalityX = initCardinality; //std::min(initCardinalityA, initCardinalityB);
+    // set initial vector using inclusion exclusion principle
+    double initCardinalityA = std::max(1., cardinalityABX - cardinalityBX);
+    double initCardinalityB = std::max(1., cardinalityABX - cardinalityAX);
[…]

### 2016-09-12 paper/paper.tex
number formatting in tables

@@
 
-\sisetup{
-fixed-exponent = 2,
-scientific-notation = fixed,
-}
-
 \title{New cardinality estimation algorithms for HyperLogLog sketches}
@@
 {intersection.csv}{trueCardA=\trueCardA,trueCardB=\trueCardB,trueCardX=\trueCardX,trueJaccardIdx=\trueJaccardIdx ,trueLogRatio=\trueLogRatio}{
-\thecsvrow & \trueCardA & \trueCardB & \trueCardX & \trueJaccardIdx & \trueLogRatio
+\thecsvrow & 
+\num{\trueCardA} & 
+\num{\trueCardB} & 
+\num{\trueCardX} & 
+\num[scientific-notation = true,round-mode = places,round-precision = 3]{\trueJaccardIdx} & 
+\num[scientific-notation = fixed,round-mode = places,round-precision = 3]{\trueLogRatio}
 }
@@
 maxLikeMeanA=\maxLikeMeanA,maxLikeStdDevA=\maxLikeStdDevA}{
-\thecsvrow & \inclExclMeanA & \inclExclStdDevA & \maxLikeMeanA & \maxLikeStdDevA
+\thecsvrow & 
+\num[scientific-notation = true,round-mode = places,round-precision = 3]{\inclExclMeanA} & 
+\num[scientific-notation = true,round-mode = places,round-precision = 3]{\inclExclStdDevA} & 
+\num[scientific-notation = true,round-mode = places,round-precision = 3]{\maxLikeMeanA} & 
+\num[scientific-notation = true,round-mode = places,round-precision = 3]{\maxLikeStdDevA}
 }
@@
 \label{tbl:joint_estimation_b}
-\csvreader[before reading=\footnotesize, tabular=r|r|r|r|r, table head=\# & IE mean & IE stdev & ML mean & ML stdev\\\hline]
+\csvreader[before reading=\footnotesize, tabular=r|r|r|r|r, table head= {\#} & {IE mean} & {IE stdev} & {ML mean} & {ML stdev}\\\hline]
 {intersection.csv}{inclExclMeanB=\inclExclMeanB,inclExclStdDevB=\inclExclStdDevB,
 maxLikeMeanB=\maxLikeMeanB,maxLikeStdDevB=\maxLikeStdDevB}{
-\thecsvrow & \inclExclMeanB & \inclExclStdDevB & \maxLikeMeanB & \maxLikeStdDevB
-}
+\thecsvrow & 
+\num[scientific-notation = true,round-mode = places,round-precision = 3]{\inclExclMeanB} & 
+\num[scientific-notation = true,round-mode = places,round-precision = 3]{\inclExclStdDevB} & 
+\num[scientific-notation = true,round-mode = places,round-precision = 3]{\maxLikeMeanB} & 
+\num[scientific-notation = true,round-mode = places,round-precision = 3]{\maxLikeStdDevB}
+}
+
 \end{table}
@@
 maxLikeMeanX=\maxLikeMeanX,maxLikeStdDevX=\maxLikeStdDevX}{
-\thecsvrow & \inclExclMeanX & \inclExclStdDevX & \maxLikeMeanX & \maxLikeStdDevX
+\thecsvrow & 
+\num[scientific-notation = true,round-mode = places,round-precision = 3]{\inclExclMeanX} & 
+\num[scientific-notation = true,round-mode = places,round-precision = 3]{\inclExclStdDevX} & 
+\num[scientific-notation = true,round-mode = places,round-precision = 3]{\maxLikeMeanX} & 
+\num[scientific-notation = true,round-mode = places,round-precision = 3]{\maxLikeStdDevX}
 }

### 2016-09-16 c++/cardinality_estimation.hpp
consistent joint hyperloglog estimation formulas in paper and implementation

@@ -421,7 +421,7 @@ void inclusionExclusionTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& j
 }
 
 void calcExp(double x, double& a, double& b) {
-    const double ln2 = std::log(2);
+    static const double ln2 = std::log(2);
     if (x >= ln2) {
         a = std::exp(-x);
         b = 1 - a;
@@ -449,34 +449,28 @@ void eval_joint_log_likelihood_function_and_derivatives(
     const double expPhiB = std::exp(phiB);
     const double expPhiX = std::exp(phiX);
 
-    double expm1a[q+1];  // 1-exp(-exp(phiA/2^k))
-    double expm1b[q+1];  // 1-exp(-exp(phiB/2^k))
-    double expa[q+1];  // exp(-exp(phiA/2^k))
-    double expb[q+1];  // exp(-exp(phiB/2^k))
-    double expm1x[q+1];
-    double expx[q+1];
-    double expPhiApow2k[q+1]; // exp(phiA/2^k)
-    double expPhiBpow2k[q+1]; // exp(phiB/2^k)
-    double expPhiXpow2k[q+1]; // exp(phiX/2^k)
+    double xa[q+1], ya[q+1], za[q+1];
+    double xb[q+1], yb[q+1], zb[q+1];
+    double xx[q+1], yx[q+1], zx[q+1];
 
-    expPhiApow2k[0] = expPhiA;
-    expPhiBpow2k[0] = expPhiB;
-    expPhiXpow2k[0] = expPhiX;
+    xa[0] = expPhiA;
+    xb[0] = expPhiB;
+    xx[0] = expPhiX;
     for (int k = 1; k <= q; ++k) {
-        expPhiApow2k[k] = expPhiApow2k[k-1] * 0.5;
-        expPhiBpow2k[k] = expPhiBpow2k[k-1] * 0.5;
-        expPhiXpow2k[k] = expPhiXpow2k[k-1] * 0.5;
+        xa[k] = xa[k-1] * 0.5;
+        xb[k] = xb[k-1] * 0.5;
+        xx[k] = xx[k-1] * 0.5;
     }
 
     for (int k = q; k >= 1; --k) {
         if (jointStatistic.get1Count(k) > 0) {
-            calcExp(expPhiApow2k[k], expa[k], expm1a[k]);
+            calcExp(xa[k], ya[k], za[k]);
         }
         if (jointStatistic.get2Count(k) > 0) {
[…]

### 2016-09-16 c++/cardinality_estimation.hpp
define stop criterion in joint cardinality estimation algorithm in terms of sqrt(m)

@@ -580,7 +580,7 @@ double log_likelihood_function_value_for_gsl(const gsl_vector *phi, void *params
 
 void maxLikelihoodTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& jointStatistic, double& cardinalityA, double& cardinalityB, double& cardinalityX, bool& maxNumIterationsReached, bool& iterationAborted, int& numIterations) {
 
-    const double eps = 1e-6;
+    const double eps = 1e-2;
     const double initalStepFactor = 2;
     const int maxNumIterations = 10000;
 
@@ -588,6 +588,7 @@ void maxLikelihoodTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& jointS
     iterationAborted = false;
 
     const int m = jointStatistic.getNumRegisters();
+    const double relativeErrorLimit = eps/(sqrt(m));
 
     const MaxLikelihoodEstimator estimator(jointStatistic.getP(), jointStatistic.getQ());
 
@@ -650,9 +651,9 @@ void maxLikelihoodTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& jointS
         double currentPhiX = gsl_vector_get(currentPhi, 2);
 
         if (
-            std::fabs(currentPhiA - lastPhiA) <= eps &&
-            std::fabs(currentPhiB - lastPhiB) <= eps &&
-            std::fabs(currentPhiX - lastPhiX) <= eps) {
+            std::fabs(currentPhiA - lastPhiA) <= relativeErrorLimit &&
+            std::fabs(currentPhiB - lastPhiB) <= relativeErrorLimit &&
+            std::fabs(currentPhiX - lastPhiX) <= relativeErrorLimit) {
             break;
         }
 

### 2016-09-16 c++/intersection_calculation.cpp
ensure that cardinality of set A is not smaller than that of B for all examples demonstrating joint cardinality estimation

@@ -124,14 +124,15 @@ int main(int argc, char* argv[])
 
     for(const string& fileName : jointCardFileNames) {
 
-        const long trueCardA = stol(fileName.substr(0, 11));
-        const long trueCardB = stol(fileName.substr(12, 23));
+        long trueCardA = stol(fileName.substr(0, 11));
+        long trueCardB = stol(fileName.substr(12, 23));
+
         const long trueCardX = stol(fileName.substr(24, 35));
         int p = stoi(fileName.substr(36, 38));
         int q = stoi(fileName.substr(39, 41));
 
         const double jaccardIndex = trueCardX/static_cast<double>(trueCardA+trueCardB+trueCardX);
-        const double logRatio = std::log10(trueCardA) - std::log10(trueCardB);
+        double logRatio = std::log10(trueCardA) - std::log10(trueCardB);
 
         if (analyzeP != p || jaccardIndex < 1e-3 || jaccardIndex > 0.1 || std::fabs(logRatio) > 2) continue; // TODO filter cardinality combinations for table
 
@@ -192,6 +193,13 @@ int main(int argc, char* argv[])
             size += 1;
             evaluationCounter += 1;
 
+        }
+
+        if(logRatio < 0) {
+            swap(inExclEstimatedCardA, inExclEstimatedCardB);
+            swap(maxLikeEstimatedCardA, maxLikeEstimatedCardB);
+            swap(trueCardA, trueCardB);
+            logRatio = -logRatio;
         }
 
         resultsFile << trueCardA << ",";

### 2016-09-18 paper/paper.tex
consistent joint hyperloglog estimation formulas in paper and implementation
added rmse ration of inclusion-exclusion principle and maximum likelihood method for joint cardinality estimation

--- 2016-09-12 paper/paper.tex
@@
 \usepackage{csvsimple}
+\usepackage{multirow}
 \usepackage[capitalise,nameinlink,noabbrev]{cleveref}
@@
 \symDefine{\symX}{x}
+\symDefine{\symY}{y}
+\symDefine{\symZ}{z}
 \symDefine{\symPhi}{\varphi}
@@
 \symDefine{\symS}{s}
-\symDefine{\symY}{y}
-\symDefine{\symZ}{z}
 \symDefine{\symInteger}{z}
@@
 
-\sisetup{
-fixed-exponent = 2,
-scientific-notation = fixed,
-}
+\newcommand{\numformat}[1]{{\num[scientific-notation = true,round-mode = places,round-precision = 3, output-exponent-marker = \ensuremath{\mathrm{e}}]{#1}}}
+
+\newcommand{\numformattwo}[1]{{\num[scientific-notation = fixed,round-mode = places,round-precision = 3]{#1}}}
 
@@
 The transformed log-likelihood function can be written as
+%\begin{multline}
+%\symFunc(\symPhi_\symSetASuffix,\symPhi_\symSetBSuffix,\symPhi_\symSetXSuffix):=
+%\log \symLikelihood(
+%\symNumReg e^{\symPhi_\symSetASuffix},
+%\symNumReg e^{\symPhi_\symSetBSuffix},
+%\symNumReg e^{\symPhi_\symSetXSuffix}
+%\vert
+%\boldsymbol{\symRegValVariate}_1,
+%\boldsymbol{\symRegValVariate}_2
+%)=
+%\\
+%\begin{aligned}
+%&-
+%\sum_{\symRegVal=0}^{\symRegRange}
+%\left(
+%  \symCountVariate^{<}_{1\symRegVal}+
+%  \symCountVariate^{=}_{\symRegVal}+
+%  \symCountVariate^{>}_{1\symRegVal}
+%\right)
+%\frac{e^{\symPhi_\symSetASuffix}}{2^{\symRegVal}}
+%+
+%\left(
+%  \symCountVariate^{<}_{2\symRegVal}+
+%  \symCountVariate^{=}_{\symRegVal}+
+%  \symCountVariate^{>}_{2\symRegVal}
+%\right)
+%\frac{e^{\symPhi_\symSetBSuffix}}{2^{\symRegVal}}
+%+
+%\left(
+%  \symCountVariate^{<}_{1\symRegVal}+
+%  \symCountVariate^{=}_{\symRegVal}+
+%  \symCountVariate^{<}_{2\symRegVal}
+%\right)
+%\frac{e^{\symPhi_\symSetXSuffix}}{2^{\symRegVal}}
+%\\
+%&+
+%\sum_{\symRegVal=1}^{\symRegRange}
+%\symCountVariate^{<}_{1\symRegVal}
+%\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetASuffix}
+%+e^{\symPhi_\symSetXSuffix}
+%}{2^{\symRegVal}}}\right)
+%+
+%\symCountVariate^{<}_{2\symRegVal}
+%\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetBSuffix}
+%+
+%e^{\symPhi_\symSetXSuffix}
+%}{2^{\symRegVal}}}\right)
+%\\
+%&+
+%\sum_{\symRegVal=1}^{\symRegRange+1}
+%\symCountVariate^{>}_{1\symRegVal}
+%\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetASuffix}
+%}{2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
+%+
+%\symCountVariate^{>}_{2\symRegVal}
+%\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetBSuffix}
+%}{2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
+%\\
+%&+
+%\sum_{\symRegVal=1}^{\symRegRange+1}
+%\symCountVariate^{=}_{\symRegVal}
+%\log\!\left(
+%1
+%-e^{-\frac{e^{\symPhi_\symSetASuffix}
+%+e^{\symPhi_\symSetXSuffix}
+%}{2^{\min\left(\symRegVal,\symRegRange\right)}}}
+%-
+%e^{-\frac{e^{\symPhi_\symSetBSuffix}
+%+e^{\symPhi_\symSetXSuffix}
+%}{2^{\min\left(\symRegVal,\symRegRange\right)}}}
+%+
+%e^{-\frac{e^{\symPhi_\symSetASuffix}
+%+e^{\symPhi_\symSetBSuffix}
+%+e^{\symPhi_\symSetXSuffix}
+%}{2^{\min\left(\symRegVal,\symRegRange\right)}}}
+%\right).
+%\end{aligned}
+%\end{multline}
+
 \begin{multline}
@@
 \begin{aligned}
+&+
+\sum_{\symRegVal=1}^{\symRegRange}
+\symCountVariate^{<}_{1\symRegVal}
+\log\!\left(
+\symZ_{\symSetXSuffix\symRegVal}
++
+\symY_{\symSetXSuffix\symRegVal}
+\symZ_{\symSetASuffix\symRegVal}
+\right)
++
+\symCountVariate^{<}_{2\symRegVal}
+\log\!\left(
+\symZ_{\symSetXSuffix\symRegVal}
++
+\symY_{\symSetXSuffix\symRegVal}
+\symZ_{\symSetBSuffix\symRegVal}
+\right)
+\\
+&+
+\sum_{\symRegVal=1}^{\symRegRange}
+\symCountVariate^{>}_{1\symRegVal}
+\log\!\left(
+\symZ_{\symSetASuffix\symRegVal}
+\right)
++
+\symCountVariate^{>}_{2\symRegVal}
+\log\!\left(
+\symZ_{\symSetBSuffix\symRegVal}
+\right)
++
+\symCountVariate^{=}_{\symRegVal}
+\log\!\left(
+\symZ_{\symSetXSuffix\symRegVal}
++
+\symY_{\symSetXSuffix\symRegVal}
+\symZ_{\symSetASuffix\symRegVal}
+\symZ_{\symSetBSuffix\symRegVal}
+\right)
+\\
+&+
+\symCountVariate^{>}_{1,\symRegRange+1}
+\log\!\left(
+\symZ_{\symSetASuffix\symRegRange}
+\right)
++
+\symCountVariate^{>}_{2,\symRegRange+1}
+\log\!\left(
+\symZ_{\symSetBSuffix\symRegRange}
+\right)
++
+\symCountVariate^{=}_{\symRegRange+1}
+\log\!\left(
+\symZ_{\symSetXSuffix\symRegRange}
++
+\symY_{\symSetXSuffix\symRegRange}
+\symZ_{\symSetASuffix\symRegRange}
+\symZ_{\symSetBSuffix\symRegRange}
+\right)
+\\
 &-
@@
 \right)
-\frac{e^{\symPhi_\symSetASuffix}}{2^{\symRegVal}}
+\symX_{\symSetASuffix\symRegVal}
 +
@@
 \right)
-\frac{e^{\symPhi_\symSetBSuffix}}{2^{\symRegVal}}
+\symX_{\symSetBSuffix\symRegVal}
 +
@@
 \right)
-\frac{e^{\symPhi_\symSetXSuffix}}{2^{\symRegVal}}
-\\
-&+
+\symX_{\symSetXSuffix\symRegVal}
+.
+\end{aligned}
+\end{multline}
+Here we introduced the following expressions for simplification:
+\begin{equation}
+\begin{aligned}
+\symX_{\symSetASuffix\symRegVal} &:=\frac{e^{\symPhi_\symSetASuffix}}{2^{\symRegVal}},
+&
+\symY_{\symSetASuffix\symRegVal} &:= e^{-\symX_{\symSetASuffix\symRegVal}},
+&
+\symZ_{\symSetASuffix\symRegVal} &:= 1 - \symY_{\symSetASuffix\symRegVal},
+\\
+\symX_{\symSetBSuffix\symRegVal} &:=\frac{e^{\symPhi_\symSetBSuffix}}{2^{\symRegVal}},
+&
+\symY_{\symSetBSuffix\symRegVal} &:= e^{-\symX_{\symSetBSuffix\symRegVal}},
+&
+\symZ_{\symSetBSuffix\symRegVal} &:= 1 - \symY_{\symSetBSuffix\symRegVal},
+\\
+\symX_{\symSetXSuffix\symRegVal} &:=\frac{e^{\symPhi_\symSetXSuffix}}{2^{\symRegVal}},
+&
+\symY_{\symSetXSuffix\symRegVal} &:= e^{-\symX_{\symSetXSuffix\symRegVal}},
+&
+\symZ_{\symSetXSuffix\symRegVal} &:= 1 - \symY_{\symSetXSuffix\symRegVal}.
+\end{aligned}
+\end{equation}
+
+Quasi-Newton methods are commonly used for finding the maximum. They require the calculation of the first derivatives which are given by
+
+\begin{multline}
+\frac{\partial\symFunc}{\partial\symPhi_\symSetASuffix}
+=
 \sum_{\symRegVal=1}^{\symRegRange}
 \symCountVariate^{<}_{1\symRegVal}
-\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetASuffix}
-+e^{\symPhi_\symSetXSuffix}
-}{2^{\symRegVal}}}\right)
-+
-\symCountVariate^{<}_{2\symRegVal}
-\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetBSuffix}
-+
-e^{\symPhi_\symSetXSuffix}
-}{2^{\symRegVal}}}\right)
-\\
-&+
-\sum_{\symRegVal=1}^{\symRegRange+1}
+\frac{
+\symY_{\symSetXSuffix\symRegVal}
+\symX_{\symSetASuffix\symRegVal}
+\symY_{\symSetASuffix\symRegVal}
+}{
+\symZ_{\symSetXSuffix\symRegVal}
++
+\symY_{\symSetXSuffix\symRegVal}
+\symZ_{\symSetASuffix\symRegVal}
+}
++
+\symCountVariate^{=}_{\symRegVal}
+\frac{
+\symY_{\symSetXSuffix\symRegVal}
+\symX_{\symSetASuffix\symRegVal}
+\symY_{\symSetASuffix\symRegVal}
+\symZ_{\symSetBSuffix\symRegVal}
+}
+{
+\symZ_{\symSetXSuffix\symRegVal}
++
+\symY_{\symSetXSuffix\symRegVal}
+\symZ_{\symSetASuffix\symRegVal}
+\symZ_{\symSetBSuffix\symRegVal}
+}
++
 \symCountVariate^{>}_{1\symRegVal}
-\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetASuffix}
-}{2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
-+
-\symCountVariate^{>}_{2\symRegVal}
-\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetBSuffix}
-}{2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
-\\
-&+
-\sum_{\symRegVal=1}^{\symRegRange+1}
-\symCountVariate^{=}_{\symRegVal}
-\log\!\left(
-1
--e^{-\frac{e^{\symPhi_\symSetASuffix}
-+e^{\symPhi_\symSetXSuffix}
-}{2^{\min\left(\symRegVal,\symRegRange\right)}}}
--
-e^{-\frac{e^{\symPhi_\symSetBSuffix}
-+e^{\symPhi_\symSetXSuffix}
-}{2^{\min\left(\symRegVal,\symRegRange\right)}}}
-+
-e^{-\frac{e^{\symPhi_\symSetASuffix}
-+e^{\symPhi_\symSetBSuffix}
-+e^{\symPhi_\symSetXSuffix}
-}{2^{\min\left(\symRegVal,\symRegRange\right)}}}
-\right).
-\end{aligned}
-\end{multline}
-
-Quasi-Newton methods are commonly used for finding the maximum. They require the calculation of the first derivatives which are given by
-\begin{align}
-\frac{\partial\symFunc}{\partial\symPhi_\symSetASuffix}
-=
-&
+\frac{
+\symX_{\symSetASuffix\symRegVal}
+\symY_{\symSetASuffix\symRegVal}
+}{
+\symZ_{\symSetASuffix\symRegVal}
+}
+\\
++
+\symCountVariate^{=}_{\symRegRange+1}
+\frac{
+\symY_{\symSetXSuffix\symRegRange}
+\symX_{\symSetASuffix\symRegRange}
+\symY_{\symSetASuffix\symRegRange}
+\symZ_{\symSetBSuffix\symRegRange}
+}
+{
+\symZ_{\symSetXSuffix\symRegRange}
++
+\symY_{\symSetXSuffix\symRegRange}
+\symZ_{\symSetASuffix\symRegRange}
+\symZ_{\symSetBSuffix\symRegRange}
+}
++
+\symCountVariate^{>}_{1,\symRegRange+1}
+\frac{
+\symX_{\symSetASuffix\symRegRange}
+\symY_{\symSetASuffix\symRegRange}
+}{
+\symZ_{\symSetASuffix\symRegRange}
+}
 -
@@
 \right)
-\frac{e^{\symPhi_\symSetASuffix}}{2^{\symRegVal}}
-+
+\symX_{\symSetASuffix\symRegVal}
+\end{multline}
+
+\begin{multline}
+\frac{\partial\symFunc}{\partial\symPhi_\symSetBSuffix}
+=
 \sum_{\symRegVal=1}^{\symRegRange}
 \symCountVariate^{<}_{1\symRegVal}
-\symHelperV\!\left(
-\frac{e^{\symPhi_\symSetASuffix}}
-{2^{\symRegVal}}
-,
-\frac{e^{\symPhi_\symSetXSuffix}}
-{2^{\symRegVal}}
-\right)
-+
-\sum_{\symRegVal=1}^{\symRegRange+1}
+\frac{
+\symY_{\symSetXSuffix\symRegVal}
+\symX_{\symSetBSuffix\symRegVal}
+\symY_{\symSetBSuffix\symRegVal}
+}{
+\symZ_{\symSetXSuffix\symRegVal}
++
+\symY_{\symSetXSuffix\symRegVal}
+\symZ_{\symSetBSuffix\symRegVal}
+}
++
+\symCountVariate^{=}_{\symRegVal}
+\frac{
+\symY_{\symSetXSuffix\symRegVal}
+\symX_{\symSetBSuffix\symRegVal}
+\symY_{\symSetBSuffix\symRegVal}
+\symZ_{\symSetASuffix\symRegVal}
+}
+{
+\symZ_{\symSetXSuffix\symRegVal}
++
+\symY_{\symSetXSuffix\symRegVal}
+\symZ_{\symSetBSuffix\symRegVal}
+\symZ_{\symSetASuffix\symRegVal}
+}
++
 \symCountVariate^{>}_{1\symRegVal}
-\symHelperV\!\left(
-\frac{e^{\symPhi_\symSetASuffix}}
-{2^{\min\left(\symRegVal,\symRegRange\right)}}
-,
-0
-\right)
-\nonumber\\
-&
-+
-\sum_{\symRegVal=1}^{\symRegRange+1}
-\symCountVariate^{=}_{\symRegVal}
-\symHelperW\!
-\left(
-\frac{e^{\symPhi_\symSetASuffix}}
-{2^{\min\left(\symRegVal,\symRegRange\right)}}
-,
-\frac{e^{\symPhi_\symSetBSuffix}}
-{2^{\min\left(\symRegVal,\symRegRange\right)}}
-,
-\frac{e^{\symPhi_\symSetXSuffix}}
-{2^{\min\left(\symRegVal,\symRegRange\right)}}
-\right)
-\end{align}
-
-\begin{align}
-\frac{\partial\symFunc}{\partial\symPhi_\symSetBSuffix}
-=
-&
+\frac{
+\symX_{\symSetBSuffix\symRegVal}
+\symY_{\symSetBSuffix\symRegVal}
+}{
+\symZ_{\symSetBSuffix\symRegVal}
+}
+\\
++
+\symCountVariate^{=}_{\symRegRange+1}
+\frac{
+\symY_{\symSetXSuffix\symRegRange}
+\symX_{\symSetBSuffix\symRegRange}
+\symY_{\symSetBSuffix\symRegRange}
+\symZ_{\symSetASuffix\symRegRange}
+}
+{
+\symZ_{\symSetXSuffix\symRegRange}
++
+\symY_{\symSetXSuffix\symRegRange}
+\symZ_{\symSetBSuffix\symRegRange}
+\symZ_{\symSetASuffix\symRegRange}
+}
++
+\symCountVariate^{>}_{1,\symRegRange+1}
+\frac{
+\symX_{\symSetBSuffix\symRegRange}
+\symY_{\symSetBSuffix\symRegRange}
+}{
+\symZ_{\symSetBSuffix\symRegRange}
+}
 -
@@
 \left(
-  \symCountVariate^{<}_{2\symRegVal}+
+  \symCountVariate^{<}_{1\symRegVal}+
   \symCountVariate^{=}_{\symRegVal}+
-  \symCountVariate^{>}_{2\symRegVal}
-\right)
-\frac{e^{\symPhi_\symSetBSuffix}}{2^{\symRegVal}}
-+
+  \symCountVariate^{>}_{1\symRegVal}
+\right)
+\symX_{\symSetBSuffix\symRegVal}
+\end{multline}
+
+\begin{multline}
+\frac{\partial\symFunc}{\partial\symPhi_\symSetXSuffix}
+=
 \sum_{\symRegVal=1}^{\symRegRange}
+\symCountVariate^{<}_{1\symRegVal}
+\frac{
+\symY_{\symSetXSuffix\symRegVal}
+\symY_{\symSetASuffix\symRegVal}
+\symX_{\symSetXSuffix\symRegVal}
+}{
+\symZ_{\symSetXSuffix\symRegVal}
++
+\symY_{\symSetXSuffix\symRegVal}
+\symZ_{\symSetASuffix\symRegVal}
+}
++
+\symCountVariate^{=}_{\symRegVal}
+\frac{
+\symX_{\symSetXSuffix\symRegVal}
+\symY_{\symSetXSuffix\symRegVal}
+\left(
+\symZ_{\symSetASuffix\symRegVal}
+\symY_{\symSetBSuffix\symRegVal}
++
+\symY_{\symSetASuffix\symRegVal}
+\symZ_{\symSetBSuffix\symRegVal}
++
+\symY_{\symSetASuffix\symRegVal}
+\symY_{\symSetBSuffix\symRegVal}
+\right)
+}
+{
+\symZ_{\symSetXSuffix\symRegVal}
++
+\symY_{\symSetXSuffix\symRegVal}
+\symZ_{\symSetBSuffix\symRegVal}
+\symZ_{\symSetASuffix\symRegVal}
+}
++
 \symCountVariate^{<}_{2\symRegVal}
-\symHelperV\!\left(
-\frac{e^{\symPhi_\symSetBSuffix}}
-{2^{\symRegVal}}
-,
-\frac{e^{\symPhi_\symSetXSuffix}}
-{2^{\symRegVal}}
-\right)
-+
-\sum_{\symRegVal=1}^{\symRegRange+1}
-\symCountVariate^{>}_{2\symRegVal}
-\symHelperV\!\left(
-\frac{e^{\symPhi_\symSetBSuffix}}
-{2^{\min\left(\symRegVal,\symRegRange\right)}}
-,
-0
-\right)
-\nonumber\\
-&
-+
-\sum_{\symRegVal=1}^{\symRegRange+1}
-\symCountVariate^{=}_{\symRegVal}
-\symHelperW\!
-\left(
-\frac{e^{\symPhi_\symSetBSuffix}}
-{2^{\min\left(\symRegVal,\symRegRange\right)}}
-,
-\frac{e^{\symPhi_\symSetASuffix}}
-{2^{\min\left(\symRegVal,\symRegRange\right)}}
-,
-\frac{e^{\symPhi_\symSetXSuffix}}
-{2^{\min\left(\symRegVal,\symRegRange\right)}}
-\right)
-\end{align}
-
-\begin{align}
-\frac{\partial\symFunc}{\partial\symPhi_\symSetXSuffix}
-=
-&
+\frac{
+\symY_{\symSetXSuffix\symRegVal}
+\symY_{\symSetBSuffix\symRegVal}
+\symX_{\symSetXSuffix\symRegVal}
+}{
+\symZ_{\symSetXSuffix\symRegVal}
++
+\symY_{\symSetXSuffix\symRegVal}
+\symZ_{\symSetBSuffix\symRegVal}
+}
+\\
++
+\symCountVariate^{=}_{\symRegRange+1}
+\frac{
+\symX_{\symSetXSuffix\symRegRange}
+\symY_{\symSetXSuffix\symRegRange}
+\left(
+\symZ_{\symSetASuffix\symRegRange}
+\symY_{\symSetBSuffix\symRegRange}
++
+\symY_{\symSetASuffix\symRegRange}
+\symZ_{\symSetBSuffix\symRegRange}
++
+\symY_{\symSetASuffix\symRegRange}
+\symY_{\symSetBSuffix\symRegRange}
+\right)
+}
+{
+\symZ_{\symSetXSuffix\symRegRange}
++
+\symY_{\symSetXSuffix\symRegRange}
+\symZ_{\symSetBSuffix\symRegRange}
+\symZ_{\symSetASuffix\symRegRange}
+}
 -
@@
 \right)
-\frac{e^{\symPhi_\symSetXSuffix}}{2^{\symRegVal}}
-+
-\sum_{\symRegVal=1}^{\symRegRange}
-\symCountVariate^{<}_{1\symRegVal}
-\symHelperV\!\left(
-\frac{e^{\symPhi_\symSetXSuffix}}
-{2^{\symRegVal}}
-,
-\frac{e^{\symPhi_\symSetASuffix}}
-{2^{\symRegVal}}
-\right)
-+
-\sum_{\symRegVal=1}^{\symRegRange}
-\symCountVariate^{<}_{2\symRegVal}
-\symHelperV\!\left(
-\frac{e^{\symPhi_\symSetXSuffix}}
-{2^{\symRegVal}}
-,
-\frac{e^{\symPhi_\symSetBSuffix}}
-{2^{\symRegVal}}
-\right)
-\nonumber\\
-&
-+
-\sum_{\symRegVal=1}^{\symRegRange+1}
-\symCountVariate^{=}_{\symRegVal}
-\symHelperU\!
-\left(
-\frac{e^{\symPhi_\symSetASuffix}}
-{2^{\min\left(\symRegVal,\symRegRange\right)}}
-,
-\frac{e^{\symPhi_\symSetBSuffix}}
-{2^{\min\left(\symRegVal,\symRegRange\right)}}
-,
-\frac{e^{\symPhi_\symSetXSuffix}}
-{2^{\min\left(\symRegVal,\symRegRange\right)}}
-\right)
-\end{align}
-
-\begin{equation}
-\symHelperV\!\left(\symX_1,\symX_2\right)
-:=
-\frac{\symX_1}{
-e^{\symX_1+\symX_2}-1
-}
-=
-\frac{\symX_1}{\symX_1 + \symX_2}
-\frac{
-\symX_1 + \symX_2
-}
+\symX_{\symSetXSuffix\symRegVal}
+\end{multline}
+
+\begin{algorithm}
+\caption{Joint cardinality estimation.}
+\label{alg:joint_cardinality_estimation}
+\begin{algorithmic}
+\Function {EstimateCardinalities}
 {
-e^{\symX_1+\symX_2}-1
-}
-\end{equation}
-\begin{equation}
-\symHelperW\!\left(\symX_1,\symX_2,\symX_3\right)
-:=
-\frac{\symX_1\left(e^{\symX_2}-1\right)
-}{
-e^{\symX_1+\symX_2+\symX_3}-e^{\symX_1}-e^{\symX_2}+1
-}
-\end{equation}
-\begin{equation}
-\symHelperU\!\left(\symX_1,\symX_2,\symX_3\right)
-:=
-\frac{\left(e^{\symX_1}+e^{\symX_2}-1\right)\symX_3
-}{
-e^{\symX_1+\symX_2+\symX_3}-e^{\symX_1}-e^{\symX_2}+1
-}
-\end{equation}
+$
+\boldsymbol\symCountVariate^{<}_{1},
+\boldsymbol\symCountVariate^{<}_{2},
+\boldsymbol\symCountVariate^{=},
+\boldsymbol\symCountVariate^{>}_{1},
+\boldsymbol\symCountVariate^{>}_{2}
+$ 
+}
+\State $\symNumReg\gets
+\left\Vert\boldsymbol\symCountVariate^{<}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{1}\right\Vert_1$
+\State $\symPoissonRate_{\symSetASuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{<}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{1}$}
+\State $\symPoissonRate_{\symSetBSuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{<}_{2} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2}$}
+\If{$\symCountVariate^{<}_{10}+\symCountVariate^{=}_0+\symCountVariate^{<}_{20} = \symNumReg$}
+\State \textbf{return} $\left(\symPoissonRate_{\symSetASuffix\symSetXSuffix},\symPoissonRate_{\symSetBSuffix\symSetXSuffix},0\right)$
+\EndIf
+\State $\symPoissonRate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{>}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2}$}
+\State $\symPhi_\symSetASuffix = \log(\max(1, \symPoissonRate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} - \symPoissonRate_{\symSetBSuffix\symSetXSuffix})/\symNumReg)$
+\State $\symPhi_\symSetBSuffix = \log(\max(1, \symPoissonRate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} - \symPoissonRate_{\symSetASuffix\symSetXSuffix})/\symNumReg)$
+\State $\symPhi_\symSetXSuffix = \log(\max(1,
+\symPoissonRate_{\symSetASuffix\symSetXSuffix}+
+\symPoissonRate_{\symSetBSuffix\symSetXSuffix}
+-
+\symPoissonRate_{\symSetASuffix\symSetBSuffix\symSetXSuffix})/\symNumReg)$
+\Repeat
+\State $\symPhi_\symSetASuffix'\gets\symPhi_\symSetASuffix,\ \symPhi'_\symSetBSuffix\gets\symPhi_\symSetBSuffix,\  \symPhi'_\symSetXSuffix\gets\symPhi_\symSetXSuffix$
+\State do BFGS algorithm step 
+\Until{
+$
+\left|\symPhi_\symSetASuffix - \symPhi'_\symSetASuffix\right| < \symError
+\wedge
+\left|\symPhi_\symSetBSuffix - \symPhi'_\symSetBSuffix\right| < \symError
+\wedge
+\left|\symPhi_\symSetXSuffix - \symPhi'_\symSetXSuffix\right| < \symError
+$}
+\State \textbf{return} $\left(
+\symNumReg e^{\symPhi_\symSetASuffix},
+\symNumReg e^{\symPhi_\symSetBSuffix},
+\symNumReg e^{\symPhi_\symSetXSuffix}
+\right)$
+\EndFunction
+\end{algorithmic}
+\end{algorithm}
 
@@
 {intersection.csv}{trueCardA=\trueCardA,trueCardB=\trueCardB,trueCardX=\trueCardX,trueJaccardIdx=\trueJaccardIdx ,trueLogRatio=\trueLogRatio}{
-\thecsvrow & \trueCardA & \trueCardB & \trueCardX & \trueJaccardIdx & \trueLogRatio
+\thecsvrow & 
+\num{\trueCardA} & 
+\num{\trueCardB} & 
+\num{\trueCardX} & 
+\numformat{\trueJaccardIdx} & 
+\numformattwo{\trueLogRatio}
 }
@@
 \centering
-\caption{The mean and the standard deviation for the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetA|$ using the inclusion-exlusion (IE) principle and the maximum likelihood method (ML), respectively.}
+\caption{The mean and the standard deviation for the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetA|$ using the inclusion-exlusion principle and the maximum likelihood method, respectively.}
 \label{tbl:joint_estimation_a}
-\csvreader[before reading=\footnotesize, tabular=r|r|r|r|r, table head=\# & IE mean & IE stdev & ML mean & ML stdev\\\hline]
-{intersection.csv}{inclExclMeanA=\inclExclMeanA,inclExclStdDevA=\inclExclStdDevA,
-maxLikeMeanA=\maxLikeMeanA,maxLikeStdDevA=\maxLikeStdDevA}{
-\thecsvrow & \inclExclMeanA & \inclExclStdDevA & \maxLikeMeanA & \maxLikeStdDevA
+\csvreader[before reading=\footnotesize, tabular=r||r|r|r||r|r|r||r, table head=
+ & 
+\multicolumn{3}{c||}{inclusion-exclusion principle}
+&
+\multicolumn{3}{c||}{maximum likelihood method}
+&
+\multicolumn{1}{c}{improvement}
+\\
+\# & 
+\multicolumn{1}{c|}{mean} & 
+\multicolumn{1}{c|}{stdev} & 
+\multicolumn{1}{c||}{rmse} & 
+\multicolumn{1}{c|}{mean} & 
+\multicolumn{1}{c|}{stdev} & 
+\multicolumn{1}{c||}{rmse} &
+\multicolumn{1}{c}{rmse ratio}
+\\\hline]
+{intersection.csv}{
+inclExclMeanA=\inclExclMeanA,
+inclExclStdDevA=\inclExclStdDevA,
+inclExclRMSEA=\inclExclRMSEA,
+maxLikeMeanA=\maxLikeMeanA,
+maxLikeStdDevA=\maxLikeStdDevA,
+maxLikeRMSEA=\maxLikeRMSEA,
+improvementRmseA=\improvementRmseA
+}{
+\thecsvrow & 
+\numformat{\inclExclMeanA} & 
+\numformat{\inclExclStdDevA} & 
+\numformat{\inclExclRMSEA} & 
+\numformat{\maxLikeMeanA} & 
+\numformat{\maxLikeStdDevA} & 
+\numformat{\maxLikeRMSEA} &
+\numformattwo{\improvementRmseA}
 }
@@
 \centering
-\caption{The mean and the standard deviation for the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetB|$ using the inclusion-exlusion (IE) principle and the maximum likelihood method (ML), respectively.}
+\caption{The mean and the standard deviation for the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetB|$ using the inclusion-exlusion principle and the maximum likelihood method, respectively.}
 \label{tbl:joint_estimation_b}
-\csvreader[before reading=\footnotesize, tabular=r|r|r|r|r, table head=\# & IE mean & IE stdev & ML mean & ML stdev\\\hline]
-{intersection.csv}{inclExclMeanB=\inclExclMeanB,inclExclStdDevB=\inclExclStdDevB,
-maxLikeMeanB=\maxLikeMeanB,maxLikeStdDevB=\maxLikeStdDevB}{
-\thecsvrow & \inclExclMeanB & \inclExclStdDevB & \maxLikeMeanB & \maxLikeStdDevB
+\csvreader[before reading=\footnotesize, tabular=r||r|r|r||r|r|r||r, table head=
+ & 
+\multicolumn{3}{c||}{inclusion-exclusion principle}
+&
+\multicolumn{3}{c||}{maximum likelihood method}
+&
+\multicolumn{1}{c}{improvement}
+\\
+\# & 
+\multicolumn{1}{c|}{mean} & 
+\multicolumn{1}{c|}{stdev} & 
+\multicolumn{1}{c||}{rmse} & 
+\multicolumn{1}{c|}{mean} & 
+\multicolumn{1}{c|}{stdev} & 
+\multicolumn{1}{c||}{rmse} &
+\multicolumn{1}{c}{rmse ratio}
+\\\hline]
+{intersection.csv}{
+inclExclMeanB=\inclExclMeanB,
+inclExclStdDevB=\inclExclStdDevB,
+inclExclRMSEB=\inclExclRMSEB,
+maxLikeMeanB=\maxLikeMeanB,
+maxLikeStdDevB=\maxLikeStdDevB,
+maxLikeRMSEB=\maxLikeRMSEB,
+improvementRmseB=\improvementRmseB
+}{
+\thecsvrow & 
+\numformat{\inclExclMeanB} & 
+\numformat{\inclExclStdDevB} & 
+\numformat{\inclExclRMSEB} & 
+\numformat{\maxLikeMeanB} & 
+\numformat{\maxLikeStdDevB} & 
+\numformat{\maxLikeRMSEB} &
+\numformattwo{\improvementRmseB}
 }
@@
 \centering
-\caption{The mean and the standard deviation for the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetX|$ using the inclusion-exlusion (IE) principle and the maximum likelihood method (ML), respectively.}
+\caption{The mean and the standard deviation for the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetX|$ using the inclusion-exlusion principle and the maximum likelihood method, respectively.}
 \label{tbl:joint_estimation_x}
-\csvreader[before reading=\footnotesize, tabular=r|r|r|r|r, table head=\# & IE mean & IE stdev & ML mean & ML stdev\\\hline]
-{intersection.csv}{inclExclMeanX=\inclExclMeanX,inclExclStdDevX=\inclExclStdDevX,
-maxLikeMeanX=\maxLikeMeanX,maxLikeStdDevX=\maxLikeStdDevX}{
-\thecsvrow & \inclExclMeanX & \inclExclStdDevX & \maxLikeMeanX & \maxLikeStdDevX
+\csvreader[before reading=\footnotesize, tabular=r||r|r|r||r|r|r||r, table head=
+ & 
+\multicolumn{3}{c||}{inclusion-exclusion principle}
+&
+\multicolumn{3}{c||}{maximum likelihood method}
+&
+\multicolumn{1}{c}{improvement}
+\\
+\# & 
+\multicolumn{1}{c|}{mean} & 
+\multicolumn{1}{c|}{stdev} & 
+\multicolumn{1}{c||}{rmse} & 
+\multicolumn{1}{c|}{mean} & 
+\multicolumn{1}{c|}{stdev} & 
+\multicolumn{1}{c||}{rmse} &
+\multicolumn{1}{c}{rmse ratio}
+\\\hline]
+{intersection.csv}{
+inclExclMeanX=\inclExclMeanX,
+inclExclStdDevX=\inclExclStdDevX,
+inclExclRMSEX=\inclExclRMSEX,
+maxLikeMeanX=\maxLikeMeanX,
+maxLikeStdDevX=\maxLikeStdDevX,
+maxLikeRMSEX=\maxLikeRMSEX,
+improvementRmseX=\improvementRmseX
+}{
+\thecsvrow & 
+\numformat{\inclExclMeanX} & 
+\numformat{\inclExclStdDevX} & 
+\numformat{\inclExclRMSEX} & 
+\numformat{\maxLikeMeanX} & 
+\numformat{\maxLikeStdDevX} & 
+\numformat{\maxLikeRMSEX} &
+\numformattwo{\improvementRmseX}
 }

### 2016-09-18 c++/cardinality_estimation.hpp

@@ -629,13 +629,13 @@ void maxLikelihoodTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& jointS
     double initCardinalityB = std::max(1., cardinalityABX - cardinalityAX);
     double initCardinalityX = std::max(1., cardinalityAX + cardinalityBX - cardinalityABX);
 
-    double lastPhiA = std::log(initCardinalityA/m);
-    double lastPhiB = std::log(initCardinalityB/m);
-    double lastPhiX = std::log(initCardinalityX/m);
+    double phiA = std::log(initCardinalityA/m);
+    double phiB = std::log(initCardinalityB/m);
+    double phiX = std::log(initCardinalityX/m);
     gsl_vector *phi = gsl_vector_alloc(3);
-    gsl_vector_set(phi, 0, lastPhiA);
-    gsl_vector_set(phi, 1, lastPhiB);
-    gsl_vector_set(phi, 2, lastPhiX);
+    gsl_vector_set(phi, 0, phiA);
+    gsl_vector_set(phi, 1, phiB);
+    gsl_vector_set(phi, 2, phiX);
 
     gsl_multimin_function_fdf my_func;
 
@@ -650,23 +650,23 @@ void maxLikelihoodTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& jointS
 
     for(numIterations = 0; numIterations < maxNumIterations; numIterations++)
     {
+        double phiPrimeA = phiA;
+        double phiPrimeB = phiB;
+        double phiPrimeX = phiX;
+
         gsl_multimin_fdfminimizer_iterate(solver);
-        const gsl_vector* currentPhi = gsl_multimin_fdfminimizer_x(solver);
-        double currentPhiA = gsl_vector_get(currentPhi, 0);
-        double currentPhiB = gsl_vector_get(currentPhi, 1);
-        double currentPhiX = gsl_vector_get(currentPhi, 2);
+
+        const gsl_vector* phi = gsl_multimin_fdfminimizer_x(solver);
+        phiA = gsl_vector_get(phi, 0);
+        phiB = gsl_vector_get(phi, 1);
+        phiX = gsl_vector_get(phi, 2);
 
         if (
-            std::fabs(currentPhiA - lastPhiA) <= relativeErrorLimit &&
-            std::fabs(currentPhiB - lastPhiB) <= relativeErrorLimit &&
[…]

### 2016-09-18 c++/intersection_calculation.cpp

@@ -63,10 +63,10 @@ void printHeaders(ofstream& os, string prefix, string postfix) {
     os << prefix << "Max" << postfix << ",";
     os << prefix << "Mean" << postfix << ",";
     os << prefix << "StdDev" << postfix << ",";
-    os << prefix << "RootMeanSquareDev" << postfix << ",";
+    os << prefix << "RMSE" << postfix << ",";
 }
 
-void calculateAndPrintStatistics(ofstream& os, std::vector<double> data) {
+void calculateAndPrintStatistics(ofstream& os, std::vector<double> data, double& stdev, double& rmse) {
     std::sort(data.begin(), data.end());
     os << gsl_stats_min(&data[0], 1, data.size()) << ",";
     os << gsl_stats_quantile_from_sorted_data(&data[0], 1, data.size(), pm3s) << ",";
@@ -78,9 +78,11 @@ void calculateAndPrintStatistics(ofstream& os, std::vector<double> data) {
     os << gsl_stats_quantile_from_sorted_data(&data[0], 1, data.size(), pp3s) << ",";
     os << gsl_stats_max(&data[0], 1, data.size()) << ",";
     const double mean = gsl_stats_mean(&data[0], 1, data.size());
+    stdev = gsl_stats_sd_with_fixed_mean(&data[0], 1, data.size(), mean);
+    rmse = gsl_stats_sd_with_fixed_mean(&data[0], 1, data.size(), 0);
     os << mean << ",";
-    os << gsl_stats_sd_with_fixed_mean(&data[0], 1, data.size(), mean) << ",";
-    os << gsl_stats_sd_with_fixed_mean(&data[0], 1, data.size(), 0) << ",";
+    os << stdev << ",";
+    os << rmse << ",";
 }
 
 int main(int argc, char* argv[])
@@ -116,6 +118,13 @@ int main(int argc, char* argv[])
     printHeaders(resultsFile, "maxLike", "B");
     printHeaders(resultsFile, "maxLike", "X");
     printHeaders(resultsFile, "maxLike", "JaccardIdx");
+
+    resultsFile << "improvementStdevA,";
+    resultsFile << "improvementRmseA,";
+    resultsFile << "improvementStdevB,";
+    resultsFile << "improvementRmseB,";
+    resultsFile << "improvementStdevX,";
+    resultsFile << "improvementRmseX,";
 
     resultsFile << "avgNumIterations,";
[…]

### 2016-09-25 paper/paper.tex
some work on paper

@@
 Based on the maximum likelihood principle another unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate.
-The maximum likelihood method is also appropriate to improve cardinality estimates of set intersections compared to the inclusion-exclusion principle.
+The maximum likelihood method is also appropriate to improve cardinality estimates of set intersection sizes compared to the inclusion-exclusion principle.
 The new methods are demonstrated and verified by extensive simulations.
@@
 The results show that the execution times are nearly constant for $\symRegRange=52$.
-Using a lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed due to the small probability of saturated registers ($\symCountVariate_{53}>0$) for realistic cardinalities.
+Using a lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed due to the small probability of saturated registers, $\symProbability(\symCountVariate_{53}>0)\approx 0$, for realistic cardinalities.
 
@@
 \State $\Delta\symX \gets \symX$
-\While{$\Delta\symX > \symX\cdot\symError$} \Comment secant method iteration, $\symStopEpsilon = 10^{-2}$ (see \cref{sec:comp_ml_estimate})
+\State $\symStopDelta\gets\symStopEpsilon/\sqrt{\symNumReg}$
+\Comment $\symStopEpsilon = 10^{-2}$ (see \cref{sec:comp_ml_estimate})
+\While{$\Delta\symX > \symX\cdot\symStopDelta$} \Comment secant method iteration
 \State $\symKappa \gets 2+\lfloor\log_2(\symX)\rfloor$
@@
 \end{equation}
-Unfortunately, this approach can be very inaccurate.
-Due to estimation errors of HyperLogLog sketches the result of \eqref{equ:inclusion-exclusion} can even be negative.
-Motivated by the good results we have obtained for a single HyperLogLog sketch using the maximum likelihood method in combination with the Poisson approximation, we are tempted to also use it for the estimation of set operations.
-Moreover, it was recently pointed out without special consideration of HyperLogLog sketches, that the application of the maximum likelihood method to the joint likelihood function of two probabilistic data structures which represent the intersection operands gives better intersection cardinality estimates \cite{Ting2016}.
+Unfortunately, this approach can be very inaccurate. Due to estimation errors of HyperLogLog sketches the result of \eqref{equ:inclusion-exclusion} can even be negative.
+
+It was recently pointed out without special focus on HyperLogLog sketches, that the application of the maximum likelihood method to the joint likelihood function of two probabilistic data structures which represent the intersection operands gives better intersection size estimates \cite{Ting2016}.
+For HyperLogLog sketches recorded without stochastic averaging (compare \cref{sec:data_element_insertion}) this was also shown in \cite{Cohen2016}, where also an algorithm based on the maximum likelihood principle was outlined.
+However, in practice, HyperLogLog sketches are recorded using stochastic averaging because of the much cheaper element insertions as described by \cref{alg:insert}.
+Motivated by the good results we have obtained for a single HyperLogLog sketch using the maximum likelihood method in combination with the Poisson approximation, we are tempted to apply this approach also for the estimation of set operations.
 
@@
 The goal is to find estimates for the cardinalities of the pair-wise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$.
-The Poisson approximation allows us to assume that the elements of $\symSetA$ are inserted into the HyperLogLog sketch representing $\symSetS_1$ at rate $\symPoissonRate_\symSetA$.
-Similarly, the elements of $\symSetB$ are inserted into the HyperLogLog sketch representing $\symSetS_2$ at rate $\symPoissonRate_\symSetB$.
-Furthermore, we assume that elements of $\symSetX$ are inserted into both HyperLogLog sketches simultaneously at rate $\symPoissonRate_\symSetX$.
-The goal is now to find estimates, $\symPoissonRateEstimate_\symSetA$, $\symPoissonRateEstimate_\symSetB$, and $\symPoissonRateEstimate_\symSetX$ for the rates which could also be good estimates for the wanted cardinalities of $\symSetA$, $\symSetB$, and $\symSetX$.
+The Poisson approximation allows us to assume that the elements of $\symSetA$ are inserted into the HyperLogLog sketch representing $\symSetS_1$ at rate $\symPoissonRate_\symSetASuffix$.
+Similarly, the elements of $\symSetB$ are inserted into the HyperLogLog sketch representing $\symSetS_2$ at rate $\symPoissonRate_\symSetBSuffix$.
+Furthermore, we assume that elements of $\symSetX$ are inserted into both HyperLogLog sketches simultaneously at rate $\symPoissonRate_\symSetXSuffix$.
+The goal is now to find estimates, $\symPoissonRateEstimate_\symSetASuffix$, $\symPoissonRateEstimate_\symSetBSuffix$, and $\symPoissonRateEstimate_\symSetXSuffix$ for the rates which could also be good estimates for the wanted cardinalities of $\symSetA$, $\symSetB$, and $\symSetX$.
 
@@
 \begin{multline}
+\label{equ:max_likelihood_phi_func}
 \symFunc(\symPhi_\symSetASuffix,\symPhi_\symSetBSuffix,\symPhi_\symSetXSuffix):=
@@
 \begin{equation}
-\begin{aligned}
-\symX_{\symSetASuffix\symRegVal} &:=\frac{e^{\symPhi_\symSetASuffix}}{2^{\symRegVal}},
-&
-\symY_{\symSetASuffix\symRegVal} &:= e^{-\symX_{\symSetASuffix\symRegVal}},
-&
-\symZ_{\symSetASuffix\symRegVal} &:= 1 - \symY_{\symSetASuffix\symRegVal},
-\\
-\symX_{\symSetBSuffix\symRegVal} &:=\frac{e^{\symPhi_\symSetBSuffix}}{2^{\symRegVal}},
-&
-\symY_{\symSetBSuffix\symRegVal} &:= e^{-\symX_{\symSetBSuffix\symRegVal}},
-&
-\symZ_{\symSetBSuffix\symRegVal} &:= 1 - \symY_{\symSetBSuffix\symRegVal},
-\\
-\symX_{\symSetXSuffix\symRegVal} &:=\frac{e^{\symPhi_\symSetXSuffix}}{2^{\symRegVal}},
-&
-\symY_{\symSetXSuffix\symRegVal} &:= e^{-\symX_{\symSetXSuffix\symRegVal}},
-&
-\symZ_{\symSetXSuffix\symRegVal} &:= 1 - \symY_{\symSetXSuffix\symRegVal}.
-\end{aligned}
-\end{equation}
-
-Quasi-Newton methods are commonly used for finding the maximum. They require the calculation of the first derivatives which are given by
+\label{equ:definition_xyz}
+\symX_{\ast\symRegVal} :=\frac{e^{\symPhi_\ast}}{2^{\symRegVal}},\quad
+\symY_{\ast\symRegVal} := e^{-\symX_{\ast\symRegVal}},\quad
+\symZ_{\ast\symRegVal} := 1 - \symY_{\ast\symRegVal}.
+\end{equation}
+
+Quasi-Newton methods are commonly used for finding the maximum of such multi-dimensional functions. They require the calculation of the first derivatives which are given by
 
 \begin{multline}
+\label{equ:df_dphia}
 \frac{\partial\symFunc}{\partial\symPhi_\symSetASuffix}
@@
 \begin{multline}
+\label{equ:df_dphib}
 \frac{\partial\symFunc}{\partial\symPhi_\symSetBSuffix}
@@
 \begin{multline}
+\label{equ:df_dphix}
 \frac{\partial\symFunc}{\partial\symPhi_\symSetXSuffix}
@@
 \end{multline}
+
+The calculation of \eqref{equ:max_likelihood_phi_func} and its derivatives requires some care when calculating $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$.
+Since $\symX_{\ast\symRegVal}$ is nonnegative, the value range of both is $[0,1]$.
+In order to reduce the numerical error of $\symZ_{\ast\symRegVal}$ for small $\symX_{\ast\symRegVal}$, it is essential to use the function $\texttt{expm1}(\symX):=e^\symX-1$ that is available in most programming languages.
+If $\symX_{\ast\symRegVal}$ is smaller than $\log(2)$, we calculate $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$ via $\symZ_{\ast\symRegVal} =
+-\texttt{expm1}(-\symX_{\ast\symRegVal})$ and $\symY_{\ast\symRegVal} = 1 - \symZ_{\ast\symRegVal}$ and not as defined in \eqref{equ:definition_xyz}.
+In this way the numerical error of both, $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$, is minimized and only a single exponential function evaluation is needed.
+
+Apart from that, the numerical evaluation of \eqref{equ:max_likelihood_phi_func} is straightforward.
+The arguments of all logarithms are in the range $[0,1]$.
+The case that some argument is equal to zero which would cause the logarithm to be negative infinite, does not occur in practice.
+Consider for example the logarithm evaluation associated with $\symCountVariate^{<}_{1\symRegVal}$ which is only relevant if $\symCountVariate^{<}_{1\symRegVal} > 0$.
+In this case however, we can be certain that the cardinality of $\symSetA\cup\symSetX$ is at least 1.
+Therefore, it is expected that at least one of the two maximum likelihood estimates $\symPoissonRateEstimate_\symSetASuffix$ and $\symPoissonRateEstimate_\symSetXSuffix$ is in the order of 1.
+If the optimization algorithm starts with appropriate initial values, function evaluations for which $\min(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)\ll 1$ are not expected.
+Furthermore, the argument of the logarithm can be bounded by $\symZ_{\symSetXSuffix\symRegVal}
++
+\symY_{\symSetXSuffix\symRegVal}
+\symZ_{\symSetASuffix\symRegVal}
+\geq
+\min(\symZ_{\symSetXSuffix\symRegVal},\symZ_{\symSetASuffix\symRegVal})
+=
+1-e^{-\frac{\min(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)}{\symNumReg 2^\symRegVal}}
+\approx
+\frac{\min(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)}{\symNumReg 2^\symRegVal}
+$.
+The last approximation is valid if $\min(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)\ll\symNumReg 2^\symRegVal$ and shows that the argument of the logarithm is large enough to be accurately represented by the double-precision floating-point format.
+Similar argumentation holds for all other logarithmic terms and also the divisions that appear in \eqref{equ:df_dphia},\eqref{equ:df_dphib}, and \eqref{equ:df_dphix}.
 
@@
 \left\Vert\boldsymbol\symCountVariate^{<}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{1}\right\Vert_1$
-\State $\symPoissonRate_{\symSetASuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{<}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{1}$}
-\State $\symPoissonRate_{\symSetBSuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{<}_{2} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2}$}
+\State $\symPoissonRateEstimate_{\symSetASuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{<}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{1}$}
+\State $\symPoissonRateEstimate_{\symSetBSuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{<}_{2} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2}$}
 \If{$\symCountVariate^{<}_{10}+\symCountVariate^{=}_0+\symCountVariate^{<}_{20} = \symNumReg$}
-\State \textbf{return} $\left(\symPoissonRate_{\symSetASuffix\symSetXSuffix},\symPoissonRate_{\symSetBSuffix\symSetXSuffix},0\right)$
+\State \textbf{return} $(\symPoissonRateEstimate_{\symSetASuffix\symSetXSuffix},\symPoissonRateEstimate_{\symSetBSuffix\symSetXSuffix},0)$
 \EndIf
 \State $\symPoissonRate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{>}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2}$}
-\State $\symPhi_\symSetASuffix = \log(\max(1, \symPoissonRate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} - \symPoissonRate_{\symSetBSuffix\symSetXSuffix})/\symNumReg)$
-\State $\symPhi_\symSetBSuffix = \log(\max(1, \symPoissonRate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} - \symPoissonRate_{\symSetASuffix\symSetXSuffix})/\symNumReg)$
+\State $\symPhi_\symSetASuffix = \log(\max(1, \symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} - \symPoissonRateEstimate_{\symSetBSuffix\symSetXSuffix})/\symNumReg)$
+\State $\symPhi_\symSetBSuffix = \log(\max(1, \symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} - \symPoissonRateEstimate_{\symSetASuffix\symSetXSuffix})/\symNumReg)$
 \State $\symPhi_\symSetXSuffix = \log(\max(1,
-\symPoissonRate_{\symSetASuffix\symSetXSuffix}+
-\symPoissonRate_{\symSetBSuffix\symSetXSuffix}
+\symPoissonRateEstimate_{\symSetASuffix\symSetXSuffix}+
+\symPoissonRateEstimate_{\symSetBSuffix\symSetXSuffix}
 -
-\symPoissonRate_{\symSetASuffix\symSetBSuffix\symSetXSuffix})/\symNumReg)$
+\symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix})/\symNumReg)$
+\State $\symStopDelta\gets\symStopEpsilon/\sqrt{\symNumReg}$
+\Comment $\symStopEpsilon = 10^{-2}$
 \Repeat
-\State $\symPhi_\symSetASuffix'\gets\symPhi_\symSetASuffix,\ \symPhi'_\symSetBSuffix\gets\symPhi_\symSetBSuffix,\  \symPhi'_\symSetXSuffix\gets\symPhi_\symSetXSuffix$
-\State do BFGS algorithm step 
+\State 
+$
+\symPhi'_\symSetASuffix\gets\symPhi_\symSetASuffix,\ \symPhi'_\symSetBSuffix\gets\symPhi_\symSetBSuffix,\  \symPhi'_\symSetXSuffix\gets\symPhi_\symSetXSuffix$
+\State
+$
+\symPhi_\symSetASuffix,
+\symPhi_\symSetBSuffix,
+\symPhi_\symSetXSuffix
+\gets$
+\Call{OptimizationStep}{$\symPhi_\symSetASuffix,
+\symPhi_\symSetBSuffix,
+\symPhi_\symSetXSuffix;
+\symFunc,
+\frac{\partial\symFunc}{\partial\symPhi_\symSetASuffix},
+\frac{\partial\symFunc}{\partial\symPhi_\symSetBSuffix},
+\frac{\partial\symFunc}{\partial\symPhi_\symSetXSuffix}
+$}
 \Until{
-$
-\left|\symPhi_\symSetASuffix - \symPhi'_\symSetASuffix\right| < \symError
-\wedge
-\left|\symPhi_\symSetBSuffix - \symPhi'_\symSetBSuffix\right| < \symError
-\wedge
-\left|\symPhi_\symSetXSuffix - \symPhi'_\symSetXSuffix\right| < \symError
+$\max(
+\left|\symPhi_\symSetASuffix - \symPhi'_\symSetASuffix\right|,
+\left|\symPhi_\symSetBSuffix - \symPhi'_\symSetBSuffix\right|,
+\left|\symPhi_\symSetXSuffix - \symPhi'_\symSetXSuffix\right|) \leq \symStopDelta
 $}
@@
 \subsection{Results}
+We used the popular Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm to solve this optimization. We used the implementation provided by the GNU scientific library \cite{Galassi2016}.
 
@@
 \label{tbl:joint_estimation_cases}
-\csvreader[before reading=\footnotesize, tabular=r|r|r|r|r|r, table head=\# & $|\symSetA|$ & $|\symSetB|$ & $|\symSetX|$ & $\frac{|\symSetX|}{|\symSetA\cup\symSetB\cup\symSetX|}$ & $\log_{10}\!\left(\frac{|\symSetA\cup\symSetX|}{|\symSetB\cup\symSetX|}\right)$\\\hline]
+\csvreader[before reading=\footnotesize, tabular=r||r|r|r||r|r, table head=\# & $|\symSetA|$ & $|\symSetB|$ & $|\symSetX|$ & $\frac{|\symSetX|}{|\symSetA\cup\symSetB\cup\symSetX|}$ & $\log_{10}\!\left(\frac{|\symSetA\cup\symSetX|}{|\symSetB\cup\symSetX|}\right)$\\\hline]
 {intersection.csv}{trueCardA=\trueCardA,trueCardB=\trueCardB,trueCardX=\trueCardX,trueJaccardIdx=\trueJaccardIdx ,trueLogRatio=\trueLogRatio}{

### 2016-09-25 c++/cardinality_estimation.hpp

@@ -420,15 +420,15 @@ void inclusionExclusionTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& j
     cardinalityX = std::max(0., cardinalityBX + cardinalityAX - cardinalityABX);
 }
 
-void calcExp(double x, double& a, double& b) {
+void calcExp(double x, double& y, double& z) {
     static const double ln2 = std::log(2);
     if (x >= ln2) {
-        a = std::exp(-x);
-        b = 1 - a;
+        y = std::exp(-x);
+        z = 1 - y;
     }
     else {
-        b = -std::expm1(-x);
-        a = 1 - b;
+        z = -std::expm1(-x);
+        y = 1 - z;
     }
 }
 

### 2016-09-26 paper/paper.tex
fixed some minor issues, reformatted some equations

@@
 The logarithm of the joint probability mass function can be written using Iverson bracket notation ($\left[\text{true}\right]:=1$, $\left[\text{false}\right]:=0$) as
-\begin{align}
+\begin{multline}
 \label{equ:joint_log_pmf_single_register}
 \log(\symProbabilityMass(\symRegVal_1,\symRegVal_2))
-=&
--\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\symRegVal_1}}
+=\\
+\begin{aligned}
+&
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_1}}}\right)
+\left[1\leq\symRegVal_1<\symRegVal_2\right]
++
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}\right)
+\left[\symRegVal_2<\symRegVal_1\right]
+\\
++
+&
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_2}}}\right)
+\left[1\leq\symRegVal_2<\symRegVal_1\right]
++
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal_2, \symRegRange\right)}}}\right)
+\left[\symRegVal_1<\symRegVal_2\right]
+\\
++
+&
+\log\!\left(
+1
+-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
+-
+e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
++
+e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
+\right)
+\left[1\leq\symRegVal_1=\symRegVal_2\right]
+\\
+-
+&
+\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\symRegVal_1}}
 \left[\symRegVal_1\leq\symRegRange\right]
@@
 \left[\symRegVal_1\leq\symRegRange\vee\symRegVal_2\leq\symRegRange\right]
-\nonumber\\
-&
-+
-\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_1}}}\right)
-\left[1\leq\symRegVal_1<\symRegVal_2\right]
-\nonumber\\
-&
-+
-\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}\right)
-\left[\symRegVal_2<\symRegVal_1\right]
-\nonumber\\
-&
-+
-\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_2}}}\right)
-\left[1\leq\symRegVal_2<\symRegVal_1\right]
-\nonumber\\
-&
-+
-\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal_2, \symRegRange\right)}}}\right)
-\left[\symRegVal_1<\symRegVal_2\right]
-\nonumber\\
-&
-+
-\log\!\left(
-1
--e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
--
-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
-+
-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
-\right)
-\left[1\leq\symRegVal_1=\symRegVal_2\right].
-\end{align}
+\end{aligned}
+\end{multline}
 
@@
 Insertion of \eqref{equ:joint_log_pmf_single_register} results in the generalization of \eqref{equ:log_likelihood_single} for two HyperLogLog sketches
-\begin{align}
+\begin{multline}
 \label{equ:log_likelihood_pair}
@@
 =
-&
+\\
+\begin{aligned}
+&
+\sum_{\symRegVal=1}^{\symRegRange}
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal}}}\right)
+\symCountVariate^{<}_{1\symRegVal}
++
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal}}}\right)
+\symCountVariate^{<}_{2\symRegVal}
+\\
++
+&
+\sum_{\symRegVal=1}^{\symRegRange+1}
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
+\symCountVariate^{>}_{1\symRegVal}
++
+\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
+\symCountVariate^{>}_{2\symRegVal}
+\\
++
+&
+\sum_{\symRegVal=1}^{\symRegRange+1}
+\log\!\left(
+1
+-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
 -
+e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
++
+e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
+\right)
+\symCountVariate^{=}_{\symRegVal}
+\\
+-
+&
 \frac{\symPoissonRate_\symSetASuffix}{\symNumReg}
@@
 }{2^{\symRegVal}}
-\nonumber\\
-&
 -
@@
   \symCountVariate^{<}_{2\symRegVal}
-}{2^{\symRegVal}}
-\nonumber\\
-&
-+
-\sum_{\symRegVal=1}^{\symRegRange}
-\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal}}}\right)
-\symCountVariate^{<}_{1\symRegVal}
-+
-\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal}}}\right)
-\symCountVariate^{<}_{2\symRegVal}
-\nonumber\\
-&
-+
-\sum_{\symRegVal=1}^{\symRegRange+1}
-\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
-\symCountVariate^{>}_{1\symRegVal}
-+
-\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
-\symCountVariate^{>}_{2\symRegVal}
-\nonumber\\
-&
-+
-\sum_{\symRegVal=1}^{\symRegRange+1}
-\log\!\left(
-1
--e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
--
-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
-+
-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
-\right)
-\symCountVariate^{=}_{\symRegVal},
-\end{align}
+}{2^{\symRegVal}},
+\end{aligned}
+\end{multline}
 where the constants $\symCountVariate^{>}_{1\symRegVal}$,
@@
 The transformed log-likelihood function can be written as
-%\begin{multline}
-%\symFunc(\symPhi_\symSetASuffix,\symPhi_\symSetBSuffix,\symPhi_\symSetXSuffix):=
-%\log \symLikelihood(
-%\symNumReg e^{\symPhi_\symSetASuffix},
-%\symNumReg e^{\symPhi_\symSetBSuffix},
-%\symNumReg e^{\symPhi_\symSetXSuffix}
-%\vert
-%\boldsymbol{\symRegValVariate}_1,
-%\boldsymbol{\symRegValVariate}_2
-%)=
-%\\
-%\begin{aligned}
-%&-
-%\sum_{\symRegVal=0}^{\symRegRange}
-%\left(
-%  \symCountVariate^{<}_{1\symRegVal}+
-%  \symCountVariate^{=}_{\symRegVal}+
-%  \symCountVariate^{>}_{1\symRegVal}
-%\right)
-%\frac{e^{\symPhi_\symSetASuffix}}{2^{\symRegVal}}
-%+
-%\left(
-%  \symCountVariate^{<}_{2\symRegVal}+
-%  \symCountVariate^{=}_{\symRegVal}+
-%  \symCountVariate^{>}_{2\symRegVal}
-%\right)
-%\frac{e^{\symPhi_\symSetBSuffix}}{2^{\symRegVal}}
-%+
-%\left(
-%  \symCountVariate^{<}_{1\symRegVal}+
-%  \symCountVariate^{=}_{\symRegVal}+
-%  \symCountVariate^{<}_{2\symRegVal}
-%\right)
-%\frac{e^{\symPhi_\symSetXSuffix}}{2^{\symRegVal}}
-%\\
-%&+
-%\sum_{\symRegVal=1}^{\symRegRange}
-%\symCountVariate^{<}_{1\symRegVal}
-%\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetASuffix}
-%+e^{\symPhi_\symSetXSuffix}
-%}{2^{\symRegVal}}}\right)
-%+
-%\symCountVariate^{<}_{2\symRegVal}
-%\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetBSuffix}
-%+
-%e^{\symPhi_\symSetXSuffix}
-%}{2^{\symRegVal}}}\right)
-%\\
-%&+
-%\sum_{\symRegVal=1}^{\symRegRange+1}
-%\symCountVariate^{>}_{1\symRegVal}
-%\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetASuffix}
-%}{2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
-%+
-%\symCountVariate^{>}_{2\symRegVal}
-%\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetBSuffix}
-%}{2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
-%\\
-%&+
-%\sum_{\symRegVal=1}^{\symRegRange+1}
-%\symCountVariate^{=}_{\symRegVal}
-%\log\!\left(
-%1
-%-e^{-\frac{e^{\symPhi_\symSetASuffix}
-%+e^{\symPhi_\symSetXSuffix}
-%}{2^{\min\left(\symRegVal,\symRegRange\right)}}}
-%-
-%e^{-\frac{e^{\symPhi_\symSetBSuffix}
-%+e^{\symPhi_\symSetXSuffix}
-%}{2^{\min\left(\symRegVal,\symRegRange\right)}}}
-%+
-%e^{-\frac{e^{\symPhi_\symSetASuffix}
-%+e^{\symPhi_\symSetBSuffix}
-%+e^{\symPhi_\symSetXSuffix}
-%}{2^{\min\left(\symRegVal,\symRegRange\right)}}}
-%\right).
-%\end{aligned}
-%\end{multline}
-
 \begin{multline}
@@
 -\texttt{expm1}(-\symX_{\ast\symRegVal})$ and $\symY_{\ast\symRegVal} = 1 - \symZ_{\ast\symRegVal}$ and not as defined in \eqref{equ:definition_xyz}.
-In this way the numerical error of both, $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$, is minimized and only a single exponential function evaluation is needed.
+In this way the numerical error of both, $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$, is minimized and only a single exponential function needs to be evaluated.
 
@@
 \subsection{Results}
-We used the popular Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm to solve this optimization. We used the implementation provided by the GNU scientific library \cite{Galassi2016}.
-
-TODO see \cref{tbl:joint_estimation_cases}, \cref{tbl:joint_estimation_a}, \cref{tbl:joint_estimation_b}, \cref{tbl:joint_estimation_x}
+We used the popular Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm to solve this optimization problem. We used the implementation provided by the GNU scientific library \cite{Galassi2016}.
+
+TODO see results \cref{tbl:joint_estimation_cases}, \cref{tbl:joint_estimation_a}, \cref{tbl:joint_estimation_b}, \cref{tbl:joint_estimation_x}
 

### 2016-09-28 paper/paper.tex
fixed some issues

@@
 \usepackage{hyperref}
-\usepackage{siunitx}
+\usepackage[binary-units = true]{siunitx}
 \usepackage{csvsimple}
@@
 \end{equation}
-Hence, the unbiased estimator $\symPoissonRateEstimate$ conditioned on $\symCardinality$ is also an unbiased estimator for $\symCardinality$ which motivates us to use $\symPoissonRateEstimate$ directly as estimator for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$.
+Hence, the unbiased estimator $\symPoissonRateEstimate$ conditioned on $\symCardinality$ is also an unbiased estimator for $\symCardinality$, which motivates us to use $\symPoissonRateEstimate$ directly as estimator for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$.
 As simulation results will show later, the Poisson approximation works well over the entire cardinality range, even for estimators that are not exactly unbiased.
@@
 \end{equation}
-as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:raw_estimator}, because the additional bias is negligible compared to the estimation error.
+as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:raw_estimator}, because the additional bias is negligible compared to the overall estimation error.
 
 \cref{fig:raw_estimate} shows the distribution of the relative error for the raw estimator as function of the cardinality.
-The chart is based on 10,000 randomly generated HyperLogLog sketches.
+The chart is based on \num{10000} randomly generated HyperLogLog sketches.
 More details of the experimental setup will be explained later in \cref{sec:corrected_raw_estimation_error}.
@@
 \includesvg[width=1\textwidth]{raw_estimate}
-\caption{The distribution of the relative estimation error over the cardinality for the raw estimator after evaluation of 10,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
+\caption{The distribution of the relative estimation error over the cardinality for the raw estimator after evaluation of \num{10000} randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
 \label{fig:raw_estimate}
@@
 \includesvg[width=1\textwidth]{small_range_estimate}
-\caption{The distribution of the relative estimation error for the linear counting estimator after evaluation of 10,000 randomly generated bitmaps of size $2^{12}$ which correspond to HyperLogLog sketches with parameters $\symPrecision = 12$ and $\symRegRange=0$.}
+\caption{The distribution of the relative estimation error for the linear counting estimator after evaluation of \num{10000} randomly generated bitmaps of size $2^{12}$ which correspond to HyperLogLog sketches with parameters $\symPrecision = 12$ and $\symRegRange=0$.}
 \label{fig:small_range_estimate}
@@
 \includesvg[width=1\textwidth]{original_estimate}
-\caption{The distribution of the relative estimation error over the cardinality for the original estimation algorithm after evaluation of 10,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
+\caption{The distribution of the relative estimation error over the cardinality for the original estimation algorithm after evaluation of \num{10000} randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
 \label{fig:original_estimate}
@@
 \label{sec:corrected_raw_estimation_error}
-In order to verify the new estimation algorithm, we generated 10,000 HyperLogLog sketches and inserted up to 50 billion unique elements.
+In order to verify the new estimation algorithm, we generated \num{10000} HyperLogLog sketches and inserted up to 50 billion unique elements.
 Assuming a uniform hash function, element hash values can be mocked by random numbers.
-For the following results we used the Mersenne Twister random number generator with 19,937 bit state size from the C++ standard library.
+For the following results we used the Mersenne Twister random number generator with \num{19937} bit state size from the C++ standard library.
 
@@
 In order to evaluate the performance of the corrected raw estimation algorithm, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch.
-For different cardinalities we loaded precalculated multiplicity vectors of 1000 randomly generated HyperLogLog sketches into main memory.
+For different cardinalities we loaded precalculated multiplicity vectors of \num{1000} randomly generated HyperLogLog sketches into main memory.
 The average computation time was determined by cycling over these multiplicity vectors and passing them as input to the algorithm.
-For each evaluated cardinality value the average execution time was calculated after 100 cycles which corresponds to 100,000 algorithm executions for each cardinality value.
+For each evaluated cardinality value the average execution time was calculated after 100 cycles which corresponds to \num{100000} algorithm executions for each cardinality value.
 The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:corrected_raw_avg_exec_time}.
 Two variants of \cref{alg:corrected_raw_estimation} have been evaluated for which the functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ have been either calculated on-demand or taken from a lookup table.
-All these benchmarks where carried out on an Intel Core i5-2500K clocking at 3.3\,GHz.
+All these benchmarks where carried out on an Intel Core i5-2500K clocking at \SI{3.3}{\giga\hertz}.
 
@@
 The numbers do not yet include the required processing time to extract the multiplicity vector out from the HyperLogLog sketch, which requires a complete scan over all registers and counting the different register values into an array.
-A theoretical lower bound for this processing time can be derived using the maximum memory bandwidth of the CPU, which is 21\,GB/s for an Intel Core i5-2500K.
-If we consider a HyperLogLog sketch with precision $\symPrecision=12$ which uses 5 bits per register, the total data size of the HyperLogLog sketch is 2.5\,kB minimum.
-Consequently, the transfer time from main memory to CPU will be at least 120\,ns.
+A theoretical lower bound for this processing time can be derived using the maximum memory bandwidth of the CPU, which is \SI[per-mode=symbol]{21}{\giga\byte\per\second} for an Intel Core i5-2500K.
+If we consider a HyperLogLog sketch with precision $\symPrecision=12$ which uses 5 bits per register, the total data size of the HyperLogLog sketch is \SI{2.5}{\kilo\byte} minimum.
+Consequently, the transfer time from main memory to CPU will be at least \SI{120}{\nano\second}.
 Having this value in mind, the presented numbers for estimating the cardinality from the multiplicity vector are quite satisfying.
@@
 \includesvg[width=1\textwidth]{corrected_raw_avg_exec_time}
-\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3\,GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.
+\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at \SI{3.3}{\giga\hertz} when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.
 Both cases, $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ precalculated and calculated on-demand have been considered.}
@@
 \item The sum $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}$ in \eqref{equ:funcprime} can be precalculated and reused for all function evaluations.
-\item Many programming languages allow the efficient multiplication and division by any integral power of 2 using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
+\item Many programming languages allow the efficient multiplication and division by any integral power of two using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
 \item The function $\symHelper(\symX)$ only needs to be evaluated at values $\left\lbrace\frac{\symX}{2^{\symRegValVariate'_\text{max}}},\frac{\symX}{2^{\symRegValVariate'_\text{max}-1}},\ldots,\frac{\symX}{2^{\symRegValVariate_\text{min}}}\right\rbrace$ where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$.
-This series corresponds to a geometric series with ratio 2.
+This series corresponds to a geometric series with ratio two.
 A straightforward calculation using \eqref{equ:helper} is very expensive because of the exponential function.
@@
 =
-\frac{\symX}{2} - \frac{\symX^2}{12} +\frac{\symX^4}{720}-\frac{\symX^6}{30240} + \symBigO(\symX^{8})
+\frac{\symX}{2} - \frac{\symX^2}{12} +\frac{\symX^4}{720}-\frac{\symX^6}{30240} + \symBigO(\symX^{8}),
 \end{equation}
@@
 The average computation time for the maximum likelihood algorithm shows a different behavior than for the corrected raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}).
-As can be seen, the average execution time is larger for most cardinalities, but nevertheless, it never exceeds 700\,ns which is still fast enough for many applications.
+As can be seen, the average execution time is larger for most cardinalities, but nevertheless, it never exceeds \SI{700}{\nano\second} which is still fast enough for many applications.
 
@@
 \includesvg[width=1\textwidth]{max_likelihood_avg_exec_time}
-\caption{Average execution time of the maximum likelihood estimation algorithm as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3\,GHz for HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
+\caption{Average execution time of the maximum likelihood estimation algorithm as a function of the true cardinality with an Intel Core i5-2500K clocking at \SI{3.3}{\giga\hertz} for HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
 \label{fig:avg_exec_time}
@@
 
-It was recently pointed out without special focus on HyperLogLog sketches, that the application of the maximum likelihood method to the joint likelihood function of two probabilistic data structures which represent the intersection operands gives better intersection size estimates \cite{Ting2016}.
+It was recently pointed out without special focus on HyperLogLog sketches, that the application of the maximum likelihood method to the joint likelihood function of two probabilistic data structures, which represent the intersection operands, gives better intersection size estimates \cite{Ting2016}.
 For HyperLogLog sketches recorded without stochastic averaging (compare \cref{sec:data_element_insertion}) this was also shown in \cite{Cohen2016}, where also an algorithm based on the maximum likelihood principle was outlined.
 However, in practice, HyperLogLog sketches are recorded using stochastic averaging because of the much cheaper element insertions as described by \cref{alg:insert}.
-Motivated by the good results we have obtained for a single HyperLogLog sketch using the maximum likelihood method in combination with the Poisson approximation, we are tempted to apply this approach also for the estimation of set operations.
+Motivated by the good results we have obtained for a single HyperLogLog sketch using the maximum likelihood method in combination with the Poisson approximation, we are tempted to apply this approach also for the estimation of set operation result sizes.
 
@@
 \subsection{Joint log-likelihood funtion}
-In order to get maximum-likelihood estimators for $\symPoissonRateEstimate_\symSetA$, $\symPoissonRateEstimate_\symSetB$, and $\symPoissonRateEstimate_\symSetX$ we need to derive the joint probability distribution of two HyperLogLog sketches.
+In order to get maximum-likelihood estimators for $\symPoissonRateEstimate_\symSetASuffix$, $\symPoissonRateEstimate_\symSetBSuffix$, and $\symPoissonRateEstimate_\symSetXSuffix$ we need to derive the joint probability distribution of two HyperLogLog sketches.
 Under the Poisson model the register values are independent and identically distributed.
@@
 \end{multline}
-where the constants $\symCountVariate^{>}_{1\symRegVal}$,
+where the constants
 $\symCountVariate^{<}_{1\symRegVal}$,
+$\symCountVariate^{>}_{1\symRegVal}$,
+$\symCountVariate^{<}_{2\symRegVal}$,
 $\symCountVariate^{>}_{2\symRegVal}$,
-$\symCountVariate^{<}_{2\symRegVal}$,
 and $\symCountVariate^{=}_{\symRegVal}$ are defined as
@@
 \begin{aligned}
+\symCountVariate^{<}_{1\symRegVal}&:=\left|\lbrace\symIndexI\vert\symRegVal=\symRegValVariate_{1\symIndexI}<
+\symRegValVariate_{2\symIndexI}\rbrace\right|,
+\\
 \symCountVariate^{>}_{1\symRegVal}&:=\left|\lbrace\symIndexI\vert\symRegVal=\symRegValVariate_{1\symIndexI}>
@@
 \\
-\symCountVariate^{<}_{1\symRegVal}&:=\left|\lbrace\symIndexI\vert\symRegVal=\symRegValVariate_{1\symIndexI}<
-\symRegValVariate_{2\symIndexI}\rbrace\right|,
+\symCountVariate^{<}_{2\symRegVal}&:=\left|\lbrace\symIndexI\vert\symRegVal=\symRegValVariate_{2\symIndexI}<
+\symRegValVariate_{1\symIndexI}\rbrace\right|,
 \\
 \symCountVariate^{>}_{2\symRegVal}&:=\left|\lbrace\symIndexI\vert\symRegVal=\symRegValVariate_{2\symIndexI}>
-\symRegValVariate_{1\symIndexI}\rbrace\right|,
-\\
-\symCountVariate^{<}_{2\symRegVal}&:=\left|\lbrace\symIndexI\vert\symRegVal=\symRegValVariate_{2\symIndexI}<
 \symRegValVariate_{1\symIndexI}\rbrace\right|,
@@
 Actually, the set of these constants can be further reduced, because $\symCountVariate^{>}_{10}=\symCountVariate^{>}_{20}=
-\symCountVariate^{<}_{1(\symRegRange+1)}=\symCountVariate^{<}_{2(\symRegRange+1)}=0$
+\symCountVariate^{<}_{1,\symRegRange+1}=\symCountVariate^{<}_{2,\symRegRange+1}=0$
 always holds.
@@
 The log-likelihood function \eqref{equ:log_likelihood_pair} does not always have a strict global maximum point.
-For example, if all register values of the first HyperLogLog sketch are larger than the corresponding values in the second HyperLogLog sketch, $\symCountVariate_{1\symRegVal}^{<}
+For example, if all register values of the first HyperLogLog sketch are larger than the corresponding values in the second HyperLogLog sketch, that is $\symCountVariate_{1\symRegVal}^{<}
 =
@@
 For comparison, the inclusion-exclusion approach would give $\symPoissonRateEstimate_2$ as estimate.
-This result is questionable, because there is no evidence at all, that the sets $\symSetS_1$ and $\symSetS_2$ really have common elements.
+This result is questionable, because there it is not evident in this case, that the sets $\symSetS_1$ and $\symSetS_2$ really have common elements.
 
@@
 
-The inclusion-exclusion method does not use all the information given by the sufficient statistic \eqref{equ:sufficient_joint_statistic}, because the estimates are functions of the three vectors $(\boldsymbol\symCountVariate^{<}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{1})$,
+The inclusion-exclusion method does not use all the available information given by the sufficient statistic \eqref{equ:sufficient_joint_statistic}, because the estimator is a function of the three vectors $(\boldsymbol\symCountVariate^{<}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{1})$,
 $(\boldsymbol\symCountVariate^{<}_{2} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2})$, and 
 $(\boldsymbol\symCountVariate^{>}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2})$.
-In contrast, the maximum-likelihood method uses all values of the sufficient statistic \eqref{equ:sufficient_joint_statistic}.
+In contrast, the maximum-likelihood method incorporates all the values of the sufficient statistic \eqref{equ:sufficient_joint_statistic} individually.
 
@@
 \sum_{\symRegVal=1}^{\symRegRange}
-\symCountVariate^{<}_{1\symRegVal}
+\symCountVariate^{<}_{2\symRegVal}
 \frac{
@@
 \symY_{\symSetXSuffix\symRegVal}
+\symZ_{\symSetASuffix\symRegVal}
 \symX_{\symSetBSuffix\symRegVal}
 \symY_{\symSetBSuffix\symRegVal}
-\symZ_{\symSetASuffix\symRegVal}
 }
@@
 \symY_{\symSetXSuffix\symRegVal}
+\symZ_{\symSetASuffix\symRegVal}
 \symZ_{\symSetBSuffix\symRegVal}
-\symZ_{\symSetASuffix\symRegVal}
-}
-+
-\symCountVariate^{>}_{1\symRegVal}
+}
++
+\symCountVariate^{>}_{2\symRegVal}
 \frac{
@@
 \symY_{\symSetXSuffix\symRegRange}
+\symZ_{\symSetASuffix\symRegRange}
 \symX_{\symSetBSuffix\symRegRange}
 \symY_{\symSetBSuffix\symRegRange}
-\symZ_{\symSetASuffix\symRegRange}
 }
@@
 \symY_{\symSetXSuffix\symRegRange}
+\symZ_{\symSetASuffix\symRegRange}
 \symZ_{\symSetBSuffix\symRegRange}
-\symZ_{\symSetASuffix\symRegRange}
-}
-+
-\symCountVariate^{>}_{1,\symRegRange+1}
+}
++
+\symCountVariate^{>}_{2,\symRegRange+1}
 \frac{
@@
 \left(
-  \symCountVariate^{<}_{1\symRegVal}+
+  \symCountVariate^{<}_{2\symRegVal}+
   \symCountVariate^{=}_{\symRegVal}+
-  \symCountVariate^{>}_{1\symRegVal}
+  \symCountVariate^{>}_{2\symRegVal}
 \right)
@@
 \frac{
+\symX_{\symSetXSuffix\symRegVal}
 \symY_{\symSetXSuffix\symRegVal}
 \symY_{\symSetASuffix\symRegVal}
-\symX_{\symSetXSuffix\symRegVal}
 }{
@@
 \left(
+\symY_{\symSetASuffix\symRegVal}
++
 \symZ_{\symSetASuffix\symRegVal}
-\symY_{\symSetBSuffix\symRegVal}
-+
-\symY_{\symSetASuffix\symRegVal}
-\symZ_{\symSetBSuffix\symRegVal}
-+
-\symY_{\symSetASuffix\symRegVal}
 \symY_{\symSetBSuffix\symRegVal}
@@
 \frac{
+\symX_{\symSetXSuffix\symRegVal}
 \symY_{\symSetXSuffix\symRegVal}
 \symY_{\symSetBSuffix\symRegVal}
-\symX_{\symSetXSuffix\symRegVal}
 }{
@@
 \left(
+\symY_{\symSetASuffix\symRegRange}
++
 \symZ_{\symSetASuffix\symRegRange}
 \symY_{\symSetBSuffix\symRegRange}
-+
-\symY_{\symSetASuffix\symRegRange}
-\symZ_{\symSetBSuffix\symRegRange}
-+
-\symY_{\symSetASuffix\symRegRange}
-\symY_{\symSetBSuffix\symRegRange}
 \right)
@@
 \symY_{\symSetXSuffix\symRegRange}
+\symZ_{\symSetASuffix\symRegRange}
 \symZ_{\symSetBSuffix\symRegRange}
-\symZ_{\symSetASuffix\symRegRange}
 }
@@
 The calculation of \eqref{equ:max_likelihood_phi_func} and its derivatives requires some care when calculating $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$.
-Since $\symX_{\ast\symRegVal}$ is nonnegative, the value range of both is $[0,1]$.
+Since $\symX_{\ast\symRegVal}$ is nonnegative, we have $\symY_{\ast\symRegVal},\symZ_{\ast\symRegVal}\in[0,1]$.
 In order to reduce the numerical error of $\symZ_{\ast\symRegVal}$ for small $\symX_{\ast\symRegVal}$, it is essential to use the function $\texttt{expm1}(\symX):=e^\symX-1$ that is available in most programming languages.
@@
 -\texttt{expm1}(-\symX_{\ast\symRegVal})$ and $\symY_{\ast\symRegVal} = 1 - \symZ_{\ast\symRegVal}$ and not as defined in \eqref{equ:definition_xyz}.
-In this way the numerical error of both, $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$, is minimized and only a single exponential function needs to be evaluated.
+In this way the numerical error of both, $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$, is minimized and still only a single exponential function needs to be evaluated.
 
@@
 The arguments of all logarithms are in the range $[0,1]$.
-The case that some argument is equal to zero which would cause the logarithm to be negative infinite, does not occur in practice.
+The case that some argument vanishes, which would cause the logarithm to be negative infinite, does not occur in practice.
 Consider for example the logarithm evaluation associated with $\symCountVariate^{<}_{1\symRegVal}$ which is only relevant if $\symCountVariate^{<}_{1\symRegVal} > 0$.
 In this case however, we can be certain that the cardinality of $\symSetA\cup\symSetX$ is at least 1.
-Therefore, it is expected that at least one of the two maximum likelihood estimates $\symPoissonRateEstimate_\symSetASuffix$ and $\symPoissonRateEstimate_\symSetXSuffix$ is in the order of 1.
+Therefore, it is expected that at least one of the two maximum likelihood estimates $\symPoissonRateEstimate_\symSetASuffix$ and $\symPoissonRateEstimate_\symSetXSuffix$ is at least in the order of 1.
 If the optimization algorithm starts with appropriate initial values, function evaluations for which $\min(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)\ll 1$ are not expected.
@@
 $.
-The last approximation is valid if $\min(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)\ll\symNumReg 2^\symRegVal$ and shows that the argument of the logarithm is large enough to be accurately represented by the double-precision floating-point format.
-Similar argumentation holds for all other logarithmic terms and also the divisions that appear in \eqref{equ:df_dphia},\eqref{equ:df_dphib}, and \eqref{equ:df_dphix}.
+The last approximation is valid, if $\min(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)\ll\symNumReg 2^\symRegVal$, and shows that the argument of the logarithm is large enough to be accurately represented by double-precision floating-point numbers.
+Similar argumentation holds for all other logarithmic terms and also the divisions that appear in \eqref{equ:df_dphia}, \eqref{equ:df_dphib}, and \eqref{equ:df_dphix}.
 
@@
 {
-$
-\boldsymbol\symCountVariate^{<}_{1},
-\boldsymbol\symCountVariate^{<}_{2},
-\boldsymbol\symCountVariate^{=},
-\boldsymbol\symCountVariate^{>}_{1},
-\boldsymbol\symCountVariate^{>}_{2}
-$ 
+$\boldsymbol\symCountVariate^{>}_{1}$,
+$\boldsymbol\symCountVariate^{<}_{1}$,
+$\boldsymbol\symCountVariate^{>}_{2}$,
+$\boldsymbol\symCountVariate^{<}_{2}$,
+$\boldsymbol\symCountVariate^{=}$
 }

### 2016-09-28 c++/cardinality_estimation.hpp

@@ -525,7 +525,7 @@ void eval_joint_log_likelihood_function_and_derivatives(
                 double tmp = cEqual * yx / arg;
                 fa -= tmp * yazb * xa;
                 fb -= tmp * ybza * xb;
-                fx -= tmp * (yazb + ybza + ya * yb) * xx;
+                fx -= tmp * (ya + ybza) * xx;
             }
         }
         if (k <= q) {

### 2016-10-04 paper/paper.tex
some improvements

@@
 
+\newcommand{\comm}[2]{{\Comment{\parbox[t]{#1\linewidth}{{#2}}}}}
+
 \title{New cardinality estimation algorithms for HyperLogLog sketches}
@@
 \State $\symIndexI \gets 1+ \langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision\rangle_2$
-\Comment $\symIndexI\in[1,2^\symPrecision]$
+\comm{0.2}{$\symIndexI\in[1,2^\symPrecision]$}
 \State $\symRegVal = \min(\{\symS\mid \symS \in [1, \symRegRange]  \wedge  \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
-\Comment $\symRegVal\in[1,\symRegRange+1]$
+\comm{0.2}{$\symRegVal\in[1,\symRegRange+1]$}
 \State $\symRegValVariate_\symIndexI \gets\max(\symRegValVariate_\symIndexI, \symRegVal)$
@@
 \right)
-\right)^\symNumReg
+\right)^{\!\symNumReg}
 du
@@
 \State $\symNumReg \gets \dim \boldsymbol{\symRegValVariate}$
-\State $\symCardinalityRawEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexI=1}^\symNumReg 2^{-\symRegValVariate_\symIndexI}\right)^{-1}$
-\Comment raw estimate \eqref{equ:raw_estimator}
+\State $\symCardinalityRawEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexI=1}^\symNumReg 2^{-\symRegValVariate_\symIndexI}\right)^{\!-1}$
+\comm{.4}{raw estimate \eqref{equ:raw_estimator}}
 \If{$\symCardinalityRawEstimate\leq \frac{5}{2}\symNumReg$}
@@
 \If{$\symCountVariate_0\neq0$}
-\State \Return $\symNumReg \log(\symNumReg / \symCountVariate_0)$ \Comment small range correction \eqref{equ:linear_counting_estimator}
+\State \Return $\symNumReg \log(\symNumReg / \symCountVariate_0)$
+\comm{.4}{small range correction \eqref{equ:linear_counting_estimator}}
 \Else
@@
 \Else
-\State\Return $-2^{32}\log(1-\symCardinalityRawEstimate/2^{32})$\Comment large range correction \eqref{equ:large_range_estimate}
+\State\Return $-2^{32}\log(1-\symCardinalityRawEstimate/2^{32})$
+\comm{.4}{large range correction \eqref{equ:large_range_estimate}}
 \EndIf
@@
  is a smooth periodic function with period 1. Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
-\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=9.885\cdot 10^{-6}$ (see \cref{fig:power_series_func}).
+\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=\num{9.885e-6}$ (see \cref{fig:power_series_func}).
 
@@
 \label{equ:dist_transformation}
-\symRegValVariate = \min\!\left(\max\!\left(\symRegValVariate', 0\right), \symRegRange+1\right).
+\symRegValVariate = \min\!\left(\max\!\left(\symRegValVariate',0\right), \symRegRange+1\right).
 \end{equation}
@@
 \begin{aligned}
-\symCountVariate_0 &= \textstyle\sum_{\symRegVal = -\infty}^{0} \symCountVariate'_\symRegVal, \\
-\symCountVariate_\symRegVal &=  \symCountVariate'_\symRegVal, \quad 1\leq\symRegVal\leq\symRegRange,\\
-\symCountVariate_{\symRegRange+1} &= \textstyle\sum_{\symRegVal = \symRegRange + 1}^{\infty} \symCountVariate'_\symRegVal.
+\symCountVariate_0 &= \textstyle\sum_{\symRegVal = -\infty}^{0} \symCountVariate'_\symRegVal,& \\
+\symCountVariate_\symRegVal &=  \symCountVariate'_\symRegVal,& \quad 1\leq\symRegVal\leq\symRegRange,\\
+\symCountVariate_{\symRegRange+1} &= \textstyle\sum_{\symRegVal = \symRegRange + 1}^{\infty} \symCountVariate'_\symRegVal.&
 \end{aligned}
@@
 \end{equation}
-To get estimators for $\symRegVal\leq 0$ and $\symRegVal\geq\symRegRange+1$, we consider the expectations
+To get estimators for $\symRegVal\leq 0$ and $\symRegVal\geq\symRegRange+1$, we consider the expectation of $\symCountVariate'_\symRegVal$
 \begin{equation}
@@
 The possible value range of $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$ is $[0, \symNumReg]$.
-Therefore, if performance matters, it is possible to precalculate the function values for all possible arguments and keep them in lookup tables of size $\symNumReg+1$.
+Therefore, if performance matters, the function values can be precalculated for all possible arguments and kept in lookup tables of size $\symNumReg+1$.
 
@@
 \State $\symZ\gets\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$
-\Comment{\parbox[t]{.5\linewidth}{alternatively, take $\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$ from precalculated lookup table}}
+\comm{.5}{alternatively, take $\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$ from precalculated lookup table}
 \For{$\symRegVal\gets \symRegRange, 1$}
@@
 \State $\symZ\gets\symZ+\symNumReg\cdot\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$
-\Comment{\parbox[t]{.5\linewidth}{alternatively, take $\symNumReg\cdot\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$ from precalculated lookup table}}
+\comm{.5}{alternatively, take $\symNumReg\cdot\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$ from precalculated lookup table}
 \State \Return$\symAlpha_\infty\symNumReg^2 / \symZ$
+\comm{.5}{$\symAlpha_\infty := 1/(2\log(2))$}
 \EndFunction
@@
 \Function {$\symSmallCorrectionFunc$}{$\symX$}
-\Comment $\symX\in[0,1]$
+\comm{.5}{$\symX\in[0,1]$}
 \If{$\symX=1$}
@@
 \Function {$\symLargeCorrectionFunc$}{$\symX$}
-\Comment $\symX\in[0,1]$
+\comm{0.5}{$\symX\in[0,1]$}
 \State $\symY\gets1$
@@
 \includesvg[width=1\textwidth]{stdev_comparison}
-\caption{Standard deviations of different cardinality estimators as functions of the true cardinality.}
+\caption{Standard deviations of different cardinality estimators for a HyperLogLog sketch with paramters $\symPrecision = 12$ and $\symRegRange=20$ as functions of the true cardinality.}
 \label{fig:stdev_comparison}
@@
 \end{equation}
-In practice, $\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\leq0.5$ is satisfied most of the time as long as only a few registers are saturated ($\symCountVariate_{\symRegRange+1}\ll \symNumReg$).
+In practice, $\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\leq0.5$ is satisfied most of the time as long as only a few registers are saturated, that is $\symCountVariate_{\symRegRange+1}\ll \symNumReg$.
 In case $\frac{\symX}{2^{\symRegValVariate'_\text{max}}} > 0.5$, $\symHelper\!\left(\frac{\symX}{2^{\symKappa}}\right)$ is calculated instead with $\symKappa = 2+\lfloor\log_2(\symX)\rfloor$.
@@
 \State $\symZ \gets \symZ + \symCountVariate_\symRegVal\cdot\symY$
-\Comment here $\symY = 2^{-\symRegVal}$
+\comm{0.35}{here $\symY = 2^{-\symRegVal}$}
 \State $\symY \gets 2\symY$
@@
 \EndFor
-\State\Comment here $\symZ=\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$
+\State
+\comm{0.35}{here $\symZ=\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$}
 \State $\symNumReg \gets \symNumReg' + \symCountVariate_0$ 
@@
 \State $\symA\gets \symZ + \symCountVariate_0$
-\Comment $\symA = \sum_{\symRegVal=0}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$
+\comm{0.35}{$\symA = \sum_{\symRegVal=0}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$}
 \State $\symB\gets \symZ + 
 \symCountVariate_{\symRegRange+1}\cdot 2^{-\symRegRange}$
-\Comment $\symB = \sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}$
+\comm{0.35}{$\symB = \sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}$}
 \If{$\symB \leq 1.5\cdot\symA$}
-\State $\symX \gets \symNumReg'/(0.5\cdot \symB+\symA)$ \Comment weak lower bound \eqref{equ:weak_lower_bound}
+\State $\symX \gets \symNumReg'/(0.5\cdot \symB+\symA)$ \comm{0.35}{weak lower bound \eqref{equ:weak_lower_bound}}
 \Else
-\State $\symX \gets \symNumReg'/b \cdot \log(1+\symB/\symA)$ \Comment strong lower bound \eqref{equ:strong_lower_bound}
+\State $\symX \gets \symNumReg'/b \cdot \log(1+\symB/\symA)$ 
+\comm{0.35}{strong lower bound \eqref{equ:strong_lower_bound}}
 \EndIf
@@
 \State $\symStopDelta\gets\symStopEpsilon/\sqrt{\symNumReg}$
-\Comment $\symStopEpsilon = 10^{-2}$ (see \cref{sec:comp_ml_estimate})
-\While{$\Delta\symX > \symX\cdot\symStopDelta$} \Comment secant method iteration
+\comm{0.35}{$\symStopEpsilon = 10^{-2}$ (see \cref{sec:comp_ml_estimate})}
+\While{$\Delta\symX > \symX\cdot\symStopDelta$} 
+\comm{0.35}{secant method iteration}
 \State $\symKappa \gets 2+\lfloor\log_2(\symX)\rfloor$
@@
 \symX \cdot2^{-\max(\symRegValVariate'_\text{max}, \symKappa)-1}$
-\Comment $\symXNormalized \in [0, 0.25]$
+\comm{0.35}{$\symXNormalized \in [0, 0.25]$}
 \State $\symX''\gets \symXNormalized\cdot\symXNormalized$
@@
 \symX' - \symX''/3 + \left(\symX''\cdot \symX''\right)\cdot\left(1/45-\symX''/472.5\right)$
-\Comment Taylor approximation \eqref{equ:taylor}
+\comm{0.35}{Taylor approximation \eqref{equ:taylor}}
 \For{$\symRegVal\gets (\symKappa - 1),\symRegValVariate'_\text{max}$}
 \State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
-\Comment calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
+\comm{0.35}{calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$}
 \State $\symXNormalized \gets 2\symXNormalized$
@@
 \State $\symFuncPrime \gets \symCount\cdot\symHelper$
-\Comment compare \eqref{equ:funcprime}
+\comm{0.35}{compare \eqref{equ:funcprime}}
 \For{$\symRegVal\gets(\symRegValVariate'_\text{max}-1),\symRegValVariate'_\text{min}$}
 \State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
-\Comment calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$
+\comm{0.35}{calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$}
 \State $\symFuncPrime\gets \symFuncPrime + \symCountVariate_\symRegVal\cdot\symHelper$
@@
 \State $\Delta\symX \gets \Delta\symX \cdot \frac{\symNumReg' - \symFuncPrime}{\symFuncPrime - \symFuncPrime_\text{prev}}$
-\Comment see \eqref{equ:secant_delta}
+\comm{0.35}{see \eqref{equ:secant_delta}}
 \Else
@@
 Furthermore, we assume that elements of $\symSetX$ are inserted into both HyperLogLog sketches simultaneously at rate $\symPoissonRate_\symSetXSuffix$.
-The goal is now to find estimates, $\symPoissonRateEstimate_\symSetASuffix$, $\symPoissonRateEstimate_\symSetBSuffix$, and $\symPoissonRateEstimate_\symSetXSuffix$ for the rates which could also be good estimates for the wanted cardinalities of $\symSetA$, $\symSetB$, and $\symSetX$.
+We expect that good estimates $\symPoissonRateEstimate_\symSetASuffix$, $\symPoissonRateEstimate_\symSetBSuffix$, and $\symPoissonRateEstimate_\symSetXSuffix$ for the rates are also good estimates for the cardinalities of $\symSetA$, $\symSetB$, and $\symSetX$.
 
@@
 \frac{\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1,\symRegVal_2\right)}}
-\left[\symRegVal_1\leq\symRegRange\vee\symRegVal_2\leq\symRegRange\right]
+\left[\symRegVal_1\leq\symRegRange\vee\symRegVal_2\leq\symRegRange\right].
 \end{aligned}
@@
 \end{equation}
-Insertion of \eqref{equ:joint_log_pmf_single_register} results in the generalization of \eqref{equ:log_likelihood_single} for two HyperLogLog sketches
+Insertion of \eqref{equ:joint_log_pmf_single_register} results in
 \begin{multline}
@@
 
-If it is given that the two HyperLogLog sketches represent disjoint sets, we know in advance that $\symPoissonRate_\symSetXSuffix=0$.
-In this case \eqref{equ:log_likelihood_pair} can be splitted into the sum of two unary functions with parameters $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$, respectively.
-As expected, these two functions are equivalent to the individual log-likelihood functions \eqref{equ:log_likelihood_single} of both HyperLogLog sketches.
+Since \eqref{equ:log_likelihood_pair} is the generalization of \eqref{equ:log_likelihood_single} for two HyperLogLog sketches, the log-likelihood function for a single HyperLogLog sketch can be obtained by considering special cases.
+For example, assuming $\symPoissonRate_\symSetXSuffix=0$, which means that both sketches represent disjoint sets, \eqref{equ:log_likelihood_pair} can be splitted into the sum of two unary functions with parameters $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$, respectively.
+As expected, both functions correspond to \eqref{equ:log_likelihood_single}.
+Another extreme case is that both sketches represent identical sets.
+In this case
+$
+\symPoissonRate_\symSetASuffix
+=
+\symPoissonRate_\symSetBSuffix=0
+$ and all the register values must be pair-wise identical, which finally 
+yields \eqref{equ:log_likelihood_single} for $\symPoissonRate_\symSetXSuffix$.
 
@@
 $(\boldsymbol\symCountVariate^{>}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2})$.
-In contrast, the maximum-likelihood method incorporates all the values of the sufficient statistic \eqref{equ:sufficient_joint_statistic} individually.
+In contrast, the maximum-likelihood method incorporates each value of the sufficient statistic \eqref{equ:sufficient_joint_statistic} individually.
 
@@
 
-Quasi-Newton methods are commonly used for finding the maximum of such multi-dimensional functions. They require the calculation of the first derivatives which are given by
-
-\begin{multline}
-\label{equ:df_dphia}
+Quasi-Newton methods are commonly used for finding the maximum of such multi-dimensional functions.
+They require the calculation of the gradient $\nabla\symFunc = \left(\frac{\partial\symFunc}{\partial\symPhi_\symSetASuffix}, \frac{\partial\symFunc}{\partial\symPhi_\symSetBSuffix},
+\frac{\partial\symFunc}{\partial\symPhi_\symSetXSuffix}\right)$ which is, using 
+\begin{equation}
+\frac{\partial\symX_{\ast\symRegVal}}{\partial\symPhi_\ast} =
+\symX_{\ast\symRegVal},
+\quad
+\frac{\partial\symY_{\ast\symRegVal}}{\partial\symPhi_\ast}
+=
+-\symX_{\ast\symRegVal}\symY_{\ast\symRegVal},
+\quad
+\frac{\partial\symZ_{\ast\symRegVal}}{\partial\symPhi_\ast}
+=
+\symX_{\ast\symRegVal}\symY_{\ast\symRegVal},
+\end{equation}
+given by
+\begin{equation}
+\label{equ:df_dphi}
+\begin{aligned}
 \frac{\partial\symFunc}{\partial\symPhi_\symSetASuffix}
 =
+&
 \sum_{\symRegVal=1}^{\symRegRange}
@@
 \\
+&
 +
@@
 \right)
-\symX_{\symSetASuffix\symRegVal}
-\end{multline}
-
-\begin{multline}
-\label{equ:df_dphib}
+\symX_{\symSetASuffix\symRegVal},
+\\
 \frac{\partial\symFunc}{\partial\symPhi_\symSetBSuffix}
-=
+=&
 \sum_{\symRegVal=1}^{\symRegRange}
@@
 \\
-+
+&+
 \symCountVariate^{=}_{\symRegRange+1}
@@
 \right)
-\symX_{\symSetBSuffix\symRegVal}
-\end{multline}
-
-\begin{multline}
-\label{equ:df_dphix}
+\symX_{\symSetBSuffix\symRegVal},
+\\
 \frac{\partial\symFunc}{\partial\symPhi_\symSetXSuffix}
-=
+=&
 \sum_{\symRegVal=1}^{\symRegRange}
@@
 \right)
-\symX_{\symSetXSuffix\symRegVal}
-\end{multline}
+\symX_{\symSetXSuffix\symRegVal}.
+\end{aligned}
+\end{equation}
 
@@
 Apart from that, the numerical evaluation of \eqref{equ:max_likelihood_phi_func} is straightforward.
-The arguments of all logarithms are in the range $[0,1]$.
+The arguments of all logarithms are in range $[0,1]$.
 The case that some argument vanishes, which would cause the logarithm to be negative infinite, does not occur in practice.
@@
 The last approximation is valid, if $\min(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)\ll\symNumReg 2^\symRegVal$, and shows that the argument of the logarithm is large enough to be accurately represented by double-precision floating-point numbers.
-Similar argumentation holds for all other logarithmic terms and also the divisions that appear in \eqref{equ:df_dphia}, \eqref{equ:df_dphib}, and \eqref{equ:df_dphix}.
+Similar argumentation holds for all other logarithmic terms and also the divisions that appear in \eqref{equ:df_dphi}.
 
@@
 \State \textbf{return} $(\symPoissonRateEstimate_{\symSetASuffix\symSetXSuffix},\symPoissonRateEstimate_{\symSetBSuffix\symSetXSuffix},0)$
+\comm{0.3}{$\symSetA$ and $\symSetB$ are disjoint and $\symSetX$ is empty in this case}
 \EndIf
@@
 \State $\symStopDelta\gets\symStopEpsilon/\sqrt{\symNumReg}$
-\Comment $\symStopEpsilon = 10^{-2}$
+\comm{0.3}{$\symStopEpsilon = 10^{-2}$}
 \Repeat
@@
 \symFunc,
-\frac{\partial\symFunc}{\partial\symPhi_\symSetASuffix},
-\frac{\partial\symFunc}{\partial\symPhi_\symSetBSuffix},
-\frac{\partial\symFunc}{\partial\symPhi_\symSetXSuffix}
+\nabla\symFunc
 $}
@@
 \subsection{Results}
-We used the popular Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm to solve this optimization problem. We used the implementation provided by the GNU scientific library \cite{Galassi2016}.
+We used the popular Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{Press2007} to solve this optimization problem.
+We used the implementation provided by the GNU scientific library \cite{Galassi2016}.
 
@@
 \centering
-\caption{The list of cases for which the cardinalities $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$ have been estimated given two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing $\symSetA\cup\symSetX$ and  $\symSetB\cup\symSetX$, respectively.}
+\caption{List of true cardinalities of the pair-wise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetA\cup\symSetX$ and  $\symSetB\cup\symSetX$ was investigated.
+}
 \label{tbl:joint_estimation_cases}

### 2016-10-04 python/comparison.py

@@ -48,6 +48,10 @@ def calculate(filename, doBiasCorrection):
     datastd = numpy.std(relative_error, axis=1)
     return (datamean, datastd)
 
+plt.rc('text', usetex=True)
+plt.rc('font', family='serif')
+
+
 fig1, ax1 = plt.subplots(subplot_kw=dict(xlim=[cardinalities[0],cardinalities[-1]], ylim=[0.0, 0.025], xscale='log', yscale='linear'))
 
 fig1.set_size_inches(10, 5)
@@ -67,9 +71,6 @@ ax2.xaxis.grid(b=True, which='major', color='gray', linestyle='--',dash_capstyle
 ax2.set_xlabel("cardinality")
 ax2.set_ylabel("bias")
 
-
-plt.rc('text', usetex=True)
-plt.rc('font', family='serif')
 
 for c in config:
     (mean, stddev) = calculate(c[0], c[2])

### 2016-10-16 paper/paper.tex
some updates, improved tables showing joint estimation results

@@
 While the union of two sets that are represented by HyperLogLog sketches can be straightforwardly computed \eqref{equ:hll_union}, the computation of cardinalities of other set operations like intersections and differences is more challenging.
-A common approach to calculate the cardinality of intersections is the inclusion-exclusion principle
-\begin{equation}
-\label{equ:inclusion-exclusion}
+The conventional approach uses the relationships
+\begin{equation}
+\label{equ:conventional_approach}
+\begin{aligned}
+\left\vert\symSetS_1 \setminus \symSetS_2\right\vert
+&=
+\left\vert\symSetS_1 \cup \symSetS_2\right\vert
+-
+\left\vert\symSetS_2\right\vert,
+\\
+\left\vert\symSetS_2 \setminus \symSetS_1\right\vert
+&=
+\left\vert\symSetS_1 \cup \symSetS_2\right\vert
+-
+\left\vert\symSetS_1\right\vert,
+\\
 \left\vert\symSetS_1 \cap \symSetS_2\right\vert
-=  
+&=  
 \left\vert\symSetS_1\right\vert  
@@
 -
-\left\vert\symSetS_1 \cup \symSetS_2\right\vert.
-\end{equation}
-Unfortunately, this approach can be very inaccurate. Due to estimation errors of HyperLogLog sketches the result of \eqref{equ:inclusion-exclusion} can even be negative.
+\left\vert\symSetS_1 \cup \symSetS_2\right\vert,
+\end{aligned}
+\end{equation}
+and the fact that HyperLogLog sketches can be easily merged using \eqref{equ:hll_union}.
+The last equation is also known as inclusion-exclusion principle and can be used to calculate the intersection size.
+Unfortunately, the estimation error does not scale well for inclusion-exclusion principle.
+For small intersection sizes the estimation error can get very large \cite{Dasgupta2015}.
+In the worst case, the estimate can be even negative without additional restriction.
+Therefore, it was proposed to combine HyperLogLog sketches with MinHash sketches \cite{Pascoe2013, Cohen2016}, which improves the estimation error, even though at the expense of significant more space consumption.
 
@@
 Therefore, it is expected that at least one of the two maximum likelihood estimates $\symPoissonRateEstimate_\symSetASuffix$ and $\symPoissonRateEstimate_\symSetXSuffix$ is at least in the order of 1.
-If the optimization algorithm starts with appropriate initial values, function evaluations for which $\min(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)\ll 1$ are not expected.
+If the optimization algorithm starts with appropriate initial values, function evaluations for which $\max(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)\ll 1$ are not expected.
 Furthermore, the argument of the logarithm can be bounded by $\symZ_{\symSetXSuffix\symRegVal}
@@
 \geq
-\min(\symZ_{\symSetXSuffix\symRegVal},\symZ_{\symSetASuffix\symRegVal})
-=
-1-e^{-\frac{\min(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)}{\symNumReg 2^\symRegVal}}
+\frac{\symZ_{\symSetXSuffix\symRegVal}}{2}
++
+\frac{
+\symZ_{\symSetXSuffix\symRegVal}+
+\symY_{\symSetXSuffix\symRegVal}
+\symZ_{\symSetASuffix\symRegVal}
+}{2}
+\geq
+\frac{
+\symZ_{\symSetXSuffix\symRegVal}+
+\symZ_{\symSetASuffix\symRegVal}
+}{2}
+\geq
+\frac{
+\max(\symZ_{\symSetXSuffix\symRegVal},
+\symZ_{\symSetASuffix\symRegVal})
+}{2}
+=
+\frac{1}{2}
+\left(1-e^{-\frac{\max(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)}{\symNumReg 2^{\symRegVal}}}\right)
 \approx
-\frac{\min(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)}{\symNumReg 2^\symRegVal}
+\frac{\max(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)}{\symNumReg 2^{\symRegVal+1}}
 $.
-The last approximation is valid, if $\min(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)\ll\symNumReg 2^\symRegVal$, and shows that the argument of the logarithm is large enough to be accurately represented by double-precision floating-point numbers.
+The last approximation is valid, if $\max(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)\ll\symNumReg 2^{\symRegVal+1}$, and shows that the argument of the logarithm is large enough to be accurately represented by double-precision floating-point numbers.
 Similar argumentation holds for all other logarithmic terms and also the divisions that appear in \eqref{equ:df_dphi}.
+
+\cref{alg:joint_cardinality_estimation} demonstrates the calculation of the estimates given $\boldsymbol\symCountVariate^{>}_{1}$,
+$\boldsymbol\symCountVariate^{<}_{1}$,
+$\boldsymbol\symCountVariate^{>}_{2}$,
+$\boldsymbol\symCountVariate^{<}_{2}$, and
+$\boldsymbol\symCountVariate^{=}$. First, a case is distinguished, where
+all registers have a value equal to zero in at least one of both HyperLogLog sketches, that is $\symCountVariate^{<}_{10}+\symCountVariate^{=}_0+\symCountVariate^{<}_{20} = \symNumReg$.
+In this case it is certain that the HyperLogLog sketches represent disjoint sets and their corresponding  cardinality estimates can be  used for $\symPoissonRateEstimate_{\symSetASuffix}$ and $\symPoissonRateEstimate_{\symSetBSuffix}$, respectively.
+
+%TODO
+%In all other cases the cardinality of $\symSetA$ and $\symSetB$
+
+%The initial guess values are important to ensure fast convergence of the optimization algorithm.
+An obvious choice are the cardinality estimates obtained by using the conventional approach \eqref{equ:conventional_approach}.
+
+%As demonstrated by we In all other cases, the maximum-likelihood estimates for $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$ will be non-zero.
+We use the conventional approach to get start values for the estimates of $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$, respectively.
+However, in order to ensure that the values are positive we restrict them to values greater than or equal to 1.
+
+We used the popular Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{Press2007} to solve this optimization problem.
+We used the implementation provided by the GNU scientific library \cite{Galassi2016}.
 
@@
 \If{$\symCountVariate^{<}_{10}+\symCountVariate^{=}_0+\symCountVariate^{<}_{20} = \symNumReg$}
-\State \textbf{return} $(\symPoissonRateEstimate_{\symSetASuffix\symSetXSuffix},\symPoissonRateEstimate_{\symSetBSuffix\symSetXSuffix},0)$
-\comm{0.3}{$\symSetA$ and $\symSetB$ are disjoint and $\symSetX$ is empty in this case}
+\State $\symPoissonRateEstimate_{\symSetASuffix}\gets
+\symPoissonRateEstimate_{\symSetASuffix\symSetXSuffix},\ 
+\symPoissonRateEstimate_{\symSetBSuffix}\gets
+\symPoissonRateEstimate_{\symSetBSuffix\symSetXSuffix},\
+\symPoissonRateEstimate_{\symSetXSuffix}\gets0$
+\comm{0.27}{in this case $\symSetS_1\cap\symSetS_2 = \symSetX = \emptyset$ holds}
+\State \textbf{return} $(
+\symPoissonRateEstimate_{\symSetASuffix},
+\symPoissonRateEstimate_{\symSetBSuffix},
+\symPoissonRateEstimate_{\symSetXSuffix}
+)$
 \EndIf
-\State $\symPoissonRate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{>}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2}$}
-\State $\symPhi_\symSetASuffix = \log(\max(1, \symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} - \symPoissonRateEstimate_{\symSetBSuffix\symSetXSuffix})/\symNumReg)$
-\State $\symPhi_\symSetBSuffix = \log(\max(1, \symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} - \symPoissonRateEstimate_{\symSetASuffix\symSetXSuffix})/\symNumReg)$
-\State $\symPhi_\symSetXSuffix = \log(\max(1,
+\State $\symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{>}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2}$}
+\State $\symPhi_\symSetASuffix \gets \log(\max(1, \symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} - \symPoissonRateEstimate_{\symSetBSuffix\symSetXSuffix})/\symNumReg)$
+\State $\symPhi_\symSetBSuffix \gets \log(\max(1, \symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} - \symPoissonRateEstimate_{\symSetASuffix\symSetXSuffix})/\symNumReg)$
+\State $\symPhi_\symSetXSuffix \gets \log(\max(1,
 \symPoissonRateEstimate_{\symSetASuffix\symSetXSuffix}+
@@
 \State $\symStopDelta\gets\symStopEpsilon/\sqrt{\symNumReg}$
-\comm{0.3}{$\symStopEpsilon = 10^{-2}$}
+\comm{0.27}{$\symStopEpsilon = 10^{-2}$}
 \Repeat
@@
 $}
+\comm{0.27}{e.g. BFGS update step}
 \Until{
@@
 $}
-\State \textbf{return} $\left(
-\symNumReg e^{\symPhi_\symSetASuffix},
-\symNumReg e^{\symPhi_\symSetBSuffix},
-\symNumReg e^{\symPhi_\symSetXSuffix}
-\right)$
+\State 
+$
+\symPoissonRateEstimate_{\symSetASuffix}
+\gets \symNumReg e^{\symPhi_\symSetASuffix},\ \symPoissonRateEstimate_{\symSetBSuffix}
+\gets \symNumReg e^{\symPhi_\symSetBSuffix},\ 
+\symPoissonRateEstimate_{\symSetXSuffix}
+\gets \symNumReg e^{\symPhi_\symSetXSuffix}
+$
+\State \textbf{return} $(
+\symPoissonRateEstimate_{\symSetASuffix},
+\symPoissonRateEstimate_{\symSetBSuffix},
+\symPoissonRateEstimate_{\symSetXSuffix}
+)$
 \EndFunction
@@
 \subsection{Results}
-We used the popular Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{Press2007} to solve this optimization problem.
-We used the implementation provided by the GNU scientific library \cite{Galassi2016}.
-
-TODO see results \cref{tbl:joint_estimation_cases}, \cref{tbl:joint_estimation_a}, \cref{tbl:joint_estimation_b}, \cref{tbl:joint_estimation_x}
+
+TODO see results \cref{tbl:joint_estimation_cases}, \cref{tbl:joint_estimation_a}, \cref{tbl:joint_estimation_b}, \cref{tbl:joint_estimation_x},
+\cref{tbl:joint_estimation_abx}
 
@@
 \caption{List of true cardinalities of the pair-wise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetA\cup\symSetX$ and  $\symSetB\cup\symSetX$ was investigated.
+The average number of function \eqref{equ:max_likelihood_phi_func} and gradient evaluations \eqref{equ:df_dphi} are also given as needed by our implementation of \cref{alg:joint_cardinality_estimation} that uses the BFGS algorithm.
 }
 \label{tbl:joint_estimation_cases}
-\csvreader[before reading=\footnotesize, tabular=r||r|r|r||r|r, table head=\# & $|\symSetA|$ & $|\symSetB|$ & $|\symSetX|$ & $\frac{|\symSetX|}{|\symSetA\cup\symSetB\cup\symSetX|}$ & $\log_{10}\!\left(\frac{|\symSetA\cup\symSetX|}{|\symSetB\cup\symSetX|}\right)$\\\hline]
-{intersection.csv}{trueCardA=\trueCardA,trueCardB=\trueCardB,trueCardX=\trueCardX,trueJaccardIdx=\trueJaccardIdx ,trueLogRatio=\trueLogRatio}{
+\csvreader[before reading=\tiny, tabular=r|rrr|rr|rr, table head= & 
+\multicolumn{3}{c|}{true cardinalities} &
+\multicolumn{1}{c}{Jaccard index} &
+\multicolumn{1}{c|}{cardinality ratio} &
+\multicolumn{2}{c}{avg number evaluations}
+\\
+\# &  
+\multicolumn{1}{c}{$|\symSetA|$} & 
+\multicolumn{1}{c}{$|\symSetB|$} & 
+\multicolumn{1}{c|}{$|\symSetX|$} & 
+\multicolumn{1}{c}{$\frac{|\symSetX|}{|\symSetA\cup\symSetB\cup\symSetX|}$} & 
+\multicolumn{1}{c|}{$\frac{|\symSetA\cup\symSetX|}{|\symSetB\cup\symSetX|}$} &
+\multicolumn{1}{c}{$\symFunc$} &
+\multicolumn{1}{c}{$\nabla\symFunc$}
+\\\hline]
+{joint_cardinality_calculation.csv}{
+trueCardA=\trueCardA,
+trueCardB=\trueCardB,
+trueCardX=\trueCardX,
+trueJaccardIdx=\trueJaccardIdx,
+trueRatio=\trueRatio,
+avgNumFunctionEvaluations=\avgNumFunctionEvaluations,
+avgNumGradientEvaluations=\avgNumGradientEvaluations,
+}{
 \thecsvrow & 
@@
 \numformat{\trueJaccardIdx} & 
-\numformattwo{\trueLogRatio}
+\numformattwo{\trueRatio} &
+\numformattwo{\avgNumFunctionEvaluations} &
+\numformattwo{\avgNumGradientEvaluations}
 }
@@
 \centering
-\caption{The mean and the standard deviation for the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetA|$ using the inclusion-exlusion principle and the maximum likelihood method, respectively.}
+\caption{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetA|$ using the conventional approach and the maximum likelihood method, respectively.
+The root-mean-square error improvement factor of the maximum likelihood method is shown in the last column.}
 \label{tbl:joint_estimation_a}
-\csvreader[before reading=\footnotesize, tabular=r||r|r|r||r|r|r||r, table head=
- & 
-\multicolumn{3}{c||}{inclusion-exclusion principle}
-&
-\multicolumn{3}{c||}{maximum likelihood method}
+\csvreader[before reading=\tiny, tabular=r|rrr|rrr|r, table head=
+ & 
+\multicolumn{3}{c|}{conventional approach}
+&
+\multicolumn{3}{c|}{maximum likelihood method}
 &
@@
 \# & 
-\multicolumn{1}{c|}{mean} & 
-\multicolumn{1}{c|}{stdev} & 
-\multicolumn{1}{c||}{rmse} & 
-\multicolumn{1}{c|}{mean} & 
-\multicolumn{1}{c|}{stdev} & 
-\multicolumn{1}{c||}{rmse} &
+\multicolumn{1}{c}{mean} & 
+\multicolumn{1}{c}{stdev} & 
+\multicolumn{1}{c|}{rmse} & 
+\multicolumn{1}{c}{mean} & 
+\multicolumn{1}{c}{stdev} & 
+\multicolumn{1}{c|}{rmse} &
 \multicolumn{1}{c}{rmse ratio}
 \\\hline]
-{intersection.csv}{
+{joint_cardinality_calculation.csv}{
 inclExclMeanA=\inclExclMeanA,
@@
 \centering
-\caption{The mean and the standard deviation for the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetB|$ using the inclusion-exlusion principle and the maximum likelihood method, respectively.}
+\caption{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetB|$ using the conventional approach and the maximum likelihood method, respectively.
+The root-mean-square error improvement factor of the maximum likelihood method is shown in the last column.}
 \label{tbl:joint_estimation_b}
-\csvreader[before reading=\footnotesize, tabular=r||r|r|r||r|r|r||r, table head=
- & 
-\multicolumn{3}{c||}{inclusion-exclusion principle}
-&
-\multicolumn{3}{c||}{maximum likelihood method}
+\csvreader[before reading=\tiny, tabular=r|rrr|rrr|r, table head=
+ & 
+\multicolumn{3}{c|}{conventional approach}
+&
+\multicolumn{3}{c|}{maximum likelihood method}
 &
@@
 \# & 
-\multicolumn{1}{c|}{mean} & 
-\multicolumn{1}{c|}{stdev} & 
-\multicolumn{1}{c||}{rmse} & 
-\multicolumn{1}{c|}{mean} & 
-\multicolumn{1}{c|}{stdev} & 
-\multicolumn{1}{c||}{rmse} &
+\multicolumn{1}{c}{mean} & 
+\multicolumn{1}{c}{stdev} & 
+\multicolumn{1}{c|}{rmse} & 
+\multicolumn{1}{c}{mean} & 
+\multicolumn{1}{c}{stdev} & 
+\multicolumn{1}{c|}{rmse} &
 \multicolumn{1}{c}{rmse ratio}
 \\\hline]
-{intersection.csv}{
+{joint_cardinality_calculation.csv}{
 inclExclMeanB=\inclExclMeanB,
@@
 \centering
-\caption{The mean and the standard deviation for the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetX|$ using the inclusion-exlusion principle and the maximum likelihood method, respectively.}
+\caption{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetX|$ using the conventional approach and the maximum likelihood method, respectively.
+The root-mean-square error improvement factor of the maximum likelihood method is shown in the last column.}
 \label{tbl:joint_estimation_x}
-\csvreader[before reading=\footnotesize, tabular=r||r|r|r||r|r|r||r, table head=
- & 
-\multicolumn{3}{c||}{inclusion-exclusion principle}
-&
-\multicolumn{3}{c||}{maximum likelihood method}
+\csvreader[before reading=\tiny, tabular=r|rrr|rrr|r, table head=
+ & 
+\multicolumn{3}{c|}{conventional approach}
+&
+\multicolumn{3}{c|}{maximum likelihood method}
 &
@@
 \# & 
-\multicolumn{1}{c|}{mean} & 
-\multicolumn{1}{c|}{stdev} & 
-\multicolumn{1}{c||}{rmse} & 
-\multicolumn{1}{c|}{mean} & 
-\multicolumn{1}{c|}{stdev} & 
-\multicolumn{1}{c||}{rmse} &
+\multicolumn{1}{c}{mean} & 
+\multicolumn{1}{c}{stdev} & 
+\multicolumn{1}{c|}{rmse} & 
+\multicolumn{1}{c}{mean} & 
+\multicolumn{1}{c}{stdev} & 
+\multicolumn{1}{c|}{rmse} &
 \multicolumn{1}{c}{rmse ratio}
 \\\hline]
-{intersection.csv}{
+{joint_cardinality_calculation.csv}{
 inclExclMeanX=\inclExclMeanX,
@@
 
+\begin{table}
+\centering
+\caption{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetA\cup\symSetB\cup\symSetX|$ using the conventional approach and the maximum likelihood method, respectively.
+The root-mean-square error improvement factor of the maximum likelihood method is shown in the last column.}
+\label{tbl:joint_estimation_abx}
+\csvreader[before reading=\tiny, tabular=r|rrr|rrr|r, table head=
+ & 
+\multicolumn{3}{c|}{conventional approach}
+&
+\multicolumn{3}{c|}{maximum likelihood method}
+&
+\multicolumn{1}{c}{improvement}
+\\
+\# & 
+\multicolumn{1}{c}{mean} & 
+\multicolumn{1}{c}{stdev} & 
+\multicolumn{1}{c|}{rmse} & 
+\multicolumn{1}{c}{mean} & 
+\multicolumn{1}{c}{stdev} & 
+\multicolumn{1}{c|}{rmse} &
+\multicolumn{1}{c}{rmse ratio}
+\\\hline]
+{joint_cardinality_calculation.csv}{
+inclExclMeanABX=\inclExclMeanABX,
+inclExclStdDevABX=\inclExclStdDevABX,
+inclExclRMSEABX=\inclExclRMSEABX,
+maxLikeMeanABX=\maxLikeMeanABX,
+maxLikeStdDevABX=\maxLikeStdDevABX,
+maxLikeRMSEABX=\maxLikeRMSEABX,
+improvementRmseABX=\improvementRmseABX
+}{
+\thecsvrow & 
+\numformat{\inclExclMeanABX} & 
+\numformat{\inclExclStdDevABX} & 
+\numformat{\inclExclRMSEABX} & 
+\numformat{\maxLikeMeanABX} & 
+\numformat{\maxLikeStdDevABX} & 
+\numformat{\maxLikeRMSEABX} &
+\numformattwo{\improvementRmseABX}
+}
+\end{table}
+
 \section{Conclusion and future work}
@@
 TODO
+
+%We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches.
+All approaches are based on the Poisson approximation.
+As shown, an unbiased estimator for the rate in the Poisson model, is also unbiased for the estimation of the cardinality.
+We have observed that this approach also works well for Poisson rate estimators that are not perfectly unbiased.
+Therefore, it would be interesting to investigate which conditions on an approximately unbiased Poisson rate estimator are sufficient to guarantee approximate unbiasedness if used as cardinality estimator.
+
+%For the estimation from a single sketch, we have developed two fast algorithms that make use of the Poisson approximation.
+The first uses the original estimator extended by theoretically motivated correction terms.
+The second is based on the maximum-likelihood method and solves the corresponding optimization problem using the secant method.
+Both algorithms are inherently unbiased and can be implemented without empirical determined data.
+
+%Finally the maximum likelihood method was also applied to the estimation of set operation result sizes where the operands are represented by HyperLogLog sketches.
+The new approach significantly improves the cardinality estimates for set intersection, differences, and unions when compared to the inclusion-exclusion principle.
+An extension to more than two HyperLogLog sketches is possible.
+However, the effort scales like $\symBigO(2^n)$ if
 

### 2016-10-16 c++/cardinality_estimation.hpp

@@ -555,7 +555,9 @@ void eval_joint_log_likelihood_function_and_derivatives(
 void log_likelihood_function_value_and_derivatives_for_gsl(const gsl_vector *phi, void *params, double *f, gsl_vector *fGrad)
 {
 
-    const TwoHyperLogLogStatistic& jointStatistic = *((TwoHyperLogLogStatistic*)params);
+    const TwoHyperLogLogStatistic& jointStatistic = *((TwoHyperLogLogStatistic*)(((void**)(params))[0]));
+    int& numFunctionEvaluations = *((int*)(((void**)(params))[1]));
+    int& numGradientEvaluations = *((int*)(((void**)(params))[2]));
 
     double fa, fb, fx;
 
@@ -563,6 +565,9 @@ void log_likelihood_function_value_and_derivatives_for_gsl(const gsl_vector *phi
         jointStatistic,
         gsl_vector_get(phi, 0), gsl_vector_get(phi, 1), gsl_vector_get(phi, 2),
         *f, fa, fb, fx, true, true);
+
+    numFunctionEvaluations += 1;
+    numGradientEvaluations += 1;
 
     gsl_vector_set(fGrad, 0, fa);
     gsl_vector_set(fGrad, 1, fb);
@@ -572,7 +577,8 @@ void log_likelihood_function_value_and_derivatives_for_gsl(const gsl_vector *phi
 
 void log_likelihood_function_derivatives_for_gsl(const gsl_vector *phi, void *params, gsl_vector *fGrad)
 {
-    const TwoHyperLogLogStatistic& jointStatistic = *((TwoHyperLogLogStatistic*)params);
+    const TwoHyperLogLogStatistic& jointStatistic = *((TwoHyperLogLogStatistic*)(((void**)(params))[0]));
+    int& numGradientEvaluations = *((int*)(((void**)(params))[2]));
 
     double fa, fb, fx, f;
 
@@ -581,6 +587,8 @@ void log_likelihood_function_derivatives_for_gsl(const gsl_vector *phi, void *pa
         gsl_vector_get(phi, 0), gsl_vector_get(phi, 1), gsl_vector_get(phi, 2),
         f, fa, fb, fx, false, true);
 
+    numGradientEvaluations += 1;
+
     gsl_vector_set(fGrad, 0, fa);
     gsl_vector_set(fGrad, 1, fb);
     gsl_vector_set(fGrad, 2, fx);
[…]

### 2016-10-16 c++/joint_cardinality_calculation.cpp

@@ -37,7 +37,7 @@ const double median = 0.5;
 const string intersectDataPathName = "../data/hll_joint/";
 const string paperPathName = "../paper/";
 const string jointCardFileName = intersectDataPathName + "joint_cardinalities.dat";
-const string resultsFileName = paperPathName + "intersection.csv";
+const string resultsFileName = paperPathName + "joint_cardinality_calculation.csv";
 
 
 template <typename T> void printToFile(const string& fileName, const T* data, size_t numSketches, bool append) {
@@ -108,15 +108,21 @@ int main(int argc, char* argv[])
     resultsFile << "trueCardX,";
 
     resultsFile << "trueJaccardIdx,";
-    resultsFile << "trueLogRatio,";
+    resultsFile << "trueRatio,";
 
     printHeaders(resultsFile, "inclExcl", "A");
     printHeaders(resultsFile, "inclExcl", "B");
     printHeaders(resultsFile, "inclExcl", "X");
+    printHeaders(resultsFile, "inclExcl", "AX");
+    printHeaders(resultsFile, "inclExcl", "BX");
+    printHeaders(resultsFile, "inclExcl", "ABX");
     printHeaders(resultsFile, "inclExcl", "JaccardIdx");
     printHeaders(resultsFile, "maxLike", "A");
     printHeaders(resultsFile, "maxLike", "B");
     printHeaders(resultsFile, "maxLike", "X");
+    printHeaders(resultsFile, "maxLike", "AX");
+    printHeaders(resultsFile, "maxLike", "BX");
+    printHeaders(resultsFile, "maxLike", "ABX");
     printHeaders(resultsFile, "maxLike", "JaccardIdx");
 
     resultsFile << "improvementStdevA,";
@@ -125,7 +131,15 @@ int main(int argc, char* argv[])
     resultsFile << "improvementRmseB,";
     resultsFile << "improvementStdevX,";
     resultsFile << "improvementRmseX,";
+    resultsFile << "improvementStdevAX,";
+    resultsFile << "improvementRmseAX,";
+    resultsFile << "improvementStdevBX,";
+    resultsFile << "improvementRmseBX,";
+    resultsFile << "improvementStdevABX,";
+    resultsFile << "improvementRmseABX,";
 
+    resultsFile << "avgNumFunctionEvaluations,";
+    resultsFile << "avgNumGradientEvaluations,";
[…]

### 2016-10-17 paper/paper.tex
minor changes

@@
 \geq
-\frac{\symZ_{\symSetXSuffix\symRegVal}}{2}
-+
-\frac{
-\symZ_{\symSetXSuffix\symRegVal}+
-\symY_{\symSetXSuffix\symRegVal}
-\symZ_{\symSetASuffix\symRegVal}
-}{2}
+\max(\symZ_{\symSetXSuffix\symRegVal}, \symZ_{\symSetASuffix\symRegVal})
+=
+1-\exp\!\left(-\frac{\max(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)}{\symNumReg 2^{\symRegVal}}\right)
 \geq
-\frac{
-\symZ_{\symSetXSuffix\symRegVal}+
-\symZ_{\symSetASuffix\symRegVal}
-}{2}
-\geq
-\frac{
-\max(\symZ_{\symSetXSuffix\symRegVal},
-\symZ_{\symSetASuffix\symRegVal})
-}{2}
-=
-\frac{1}{2}
-\left(1-e^{-\frac{\max(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)}{\symNumReg 2^{\symRegVal}}}\right)
-\approx
-\frac{\max(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)}{\symNumReg 2^{\symRegVal+1}}
+\min\!\left(\frac{1}{2},\frac{\max(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)}{\symNumReg 2^{\symRegVal+1}}
+\right)
 $.
-The last approximation is valid, if $\max(\symPoissonRate_\symSetASuffix,\symPoissonRate_\symSetXSuffix)\ll\symNumReg 2^{\symRegVal+1}$, and shows that the argument of the logarithm is large enough to be accurately represented by double-precision floating-point numbers.
+For the last inequality we used $1-e^{-\symX}\geq\frac{1}{2}\min(1,\symX)$.
+The derived lower bound shows that the argument of the logarithm is large enough to be accurately represented by double-precision floating-point numbers, provided that $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetXSuffix$ are not both substantially smaller than 1.
 Similar argumentation holds for all other logarithmic terms and also the divisions that appear in \eqref{equ:df_dphi}.

### 2016-10-23 paper/paper.tex
some improvements

@@
 \maketitle
-\let\thefootnote\relax\footnotetext{This paper together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}.}
+\let\thefootnote\relax\footnotetext{This paper together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}.
+The first version of this paper was published on April 17, 2016.}
 \begin{abstract}
@@
 A theoretically motivated extension to the original estimator is presented that eliminates the bias for small and large cardinalities.
-Based on the maximum likelihood principle another unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate.
-The maximum likelihood method is also appropriate to improve cardinality estimates of set intersection sizes compared to the inclusion-exclusion principle.
-The new methods are demonstrated and verified by extensive simulations.
+Based on the maximum likelihood principle a second unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate.
+The maximum likelihood approach can also be applied to more than a single HyperLogLog sketch.
+In particular, it is shown that it gives more accurate cardinality estimates for the union, intersection, or relative complements of two sets that are represented by HyperLogLog sketches compared to conventional techniques like the inclusion-exclusion principle.
+All the new methods are demonstrated and verified by extensive simulations.
 \end{abstract}
@@
 Putting all these optimizations together finally gives the new cardinality estimation algorithm as presented in \cref{alg:estimate_ml}.
-The algorithm requires only elementary operations, except for large cardinalities, where it is worth to take the strong \eqref{equ:strong_lower_bound} instead of the weak lower bound \eqref{equ:weak_lower_bound} as initial value, which requires a single logarithm evaluation.
-The stronger bound is a much better approximation especially for large cardinalities and substantially reduces the number of required iteration cycles.
+The algorithm requires mainly only elementary operations.
+The square root that appears for the calculation of the stopping limit $\symStopDelta=\symStopEpsilon/\sqrt{\symNumReg}$ can be precomputed, because it only depends on the HyperLogLog precision parameter $\symPrecision$.
+For large cardinalities it is worth to take the strong \eqref{equ:strong_lower_bound} instead of the weak lower bound \eqref{equ:weak_lower_bound} as initial value, which requires a single logarithm evaluation.
+The stronger bound is a much better approximation especially for large cardinalities, where the extra logarithm evaluation is amortized by the savings of iteration cycles.
 
@@
 The last equation is also known as inclusion-exclusion principle and can be used to calculate the intersection size.
-Unfortunately, the estimation error does not scale well for inclusion-exclusion principle.
-For small intersection sizes the estimation error can get very large \cite{Dasgupta2015}.
-In the worst case, the estimate can be even negative without additional restriction.
+Unfortunately, the estimation error does not scale well for the inclusion-exclusion principle.
+For small intersection sizes the relative estimation error can become very large \cite{Dasgupta2015}.
+In the worst case, the estimate could be negative without artificial restriction to nonnegative values.
 Therefore, it was proposed to combine HyperLogLog sketches with MinHash sketches \cite{Pascoe2013, Cohen2016}, which improves the estimation error, even though at the expense of significant more space consumption.
@@
 \prod_{\symIndexI = 1}^{\symNumReg}
-\prod_{\symIndexJ = 1}^{\symNumReg}
-\symProbabilityMass(\symRegVal_{1\symIndexI},\symRegVal_{2\symIndexJ}).
+\symProbabilityMass(\symRegVal_{1\symIndexI},\symRegVal_{2\symIndexI}).
 \end{equation}
@@
 \sum_{\symIndexI = 1}^{\symNumReg}
-\sum_{\symIndexJ = 1}^{\symNumReg}
-\log(\symProbabilityMass(\symRegValVariate_{1\symIndexI},\symRegValVariate_{2\symIndexJ}))
+\log(\symProbabilityMass(\symRegValVariate_{1\symIndexI},\symRegValVariate_{2\symIndexI}))
 \end{equation}
@@
 
+Since \eqref{equ:log_likelihood_pair} is the generalization of \eqref{equ:log_likelihood_single} for two HyperLogLog sketches, the log-likelihood function for a single HyperLogLog sketch can be obtained by considering special cases.
+For example, assuming $\symPoissonRate_\symSetXSuffix=0$, which means that both sketches represent disjoint sets, \eqref{equ:log_likelihood_pair} can be splitted into the sum of two unary functions with parameters $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$ each of which correspond to \eqref{equ:log_likelihood_single} as expected.
+Or, consider two sketches representing identical sets.
+In this case all registers are equal, which means  that $\symCountVariate^{>}_{1\symRegVal}=\symCountVariate^{>}_{2\symRegVal}=
+\symCountVariate^{<}_{1\symRegVal}=\symCountVariate^{<}_{2\symRegVal}=0$ for all $\symRegVal$, and therefore the maximum likelihood method yields $\symPoissonRateEstimate_\symSetASuffix = \symPoissonRateEstimate_\symSetBSuffix = 0$ and the value for $\symPoissonRateEstimate_\symSetXSuffix$ will be equal to the single HyperLogLog maximum likelihood estimate \eqref{equ:log_likelihood_single}.
+
 The log-likelihood function \eqref{equ:log_likelihood_pair} does not always have a strict global maximum point.
@@
 For comparison, the inclusion-exclusion approach would give $\symPoissonRateEstimate_2$ as estimate.
-This result is questionable, because there it is not evident in this case, that the sets $\symSetS_1$ and $\symSetS_2$ really have common elements.
-
-Since \eqref{equ:log_likelihood_pair} is the generalization of \eqref{equ:log_likelihood_single} for two HyperLogLog sketches, the log-likelihood function for a single HyperLogLog sketch can be obtained by considering special cases.
+This result is questionable, because there is no evidence in this case, that the sets $\symSetS_1$ and $\symSetS_2$ really have common elements.
+
+%Since \eqref{equ:log_likelihood_pair} is the generalization of \eqref{equ:log_likelihood_single} for two HyperLogLog sketches, the log-likelihood function for a single HyperLogLog sketch can be obtained by considering special cases.
 For example, assuming $\symPoissonRate_\symSetXSuffix=0$, which means that both sketches represent disjoint sets, \eqref{equ:log_likelihood_pair} can be splitted into the sum of two unary functions with parameters $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$, respectively.
@@
 Another extreme case is that both sketches represent identical sets.
-In this case
-$
-\symPoissonRate_\symSetASuffix
-=
-\symPoissonRate_\symSetBSuffix=0
-$ and all the register values must be pair-wise identical, which finally 
-yields \eqref{equ:log_likelihood_single} for $\symPoissonRate_\symSetXSuffix$.
+In this case all registers are equal, which means  that $\symCountVariate^{>}_{1\symRegVal}=\symCountVariate^{>}_{2\symRegVal}=
+%\symCountVariate^{<}_{1\symRegVal}=\symCountVariate^{<}_{2\symRegVal}=0$ for all $\symRegVal$, and therefore the maximum likelihood method yields $\symPoissonRateEstimate_\symSetASuffix = \symPoissonRateEstimate_\symSetBSuffix = 0$ and the value for $\symPoissonRateEstimate_\symSetXSuffix$ will be equal to the individual maximum likelihood estimate of both identical sketches.
+
+%$
+%\symPoissonRate_\symSetASuffix
+%=
+%\symPoissonRate_\symSetBSuffix=0
+%$ and all the register values must be pair-wise identical, which finally 
+%yields \eqref{equ:log_likelihood_single} for $\symPoissonRate_\symSetXSuffix$.
 
@@
 \label{tbl:joint_estimation_cases}
-\csvreader[before reading=\tiny, tabular=r|rrr|rr|rr, table head= & 
+\csvreader[before reading=\scriptsize, tabular=r|rrr|rr|rr, table head= & 
 \multicolumn{3}{c|}{true cardinalities} &
@@
 \multicolumn{1}{c|}{cardinality ratio} &
-\multicolumn{2}{c}{avg number evaluations}
+\multicolumn{2}{c}{avg num evaluations}
 \\
@@
 \label{tbl:joint_estimation_a}
-\csvreader[before reading=\tiny, tabular=r|rrr|rrr|r, table head=
+\csvreader[before reading=\scriptsize, tabular=r|rrr|rrr|r, table head=
  & 
@@
 \label{tbl:joint_estimation_b}
-\csvreader[before reading=\tiny, tabular=r|rrr|rrr|r, table head=
+\csvreader[before reading=\scriptsize, tabular=r|rrr|rrr|r, table head=
  & 
@@
 \label{tbl:joint_estimation_x}
-\csvreader[before reading=\tiny, tabular=r|rrr|rrr|r, table head=
+\csvreader[before reading=\scriptsize, tabular=r|rrr|rrr|r, table head=
  & 
@@
 \label{tbl:joint_estimation_abx}
-\csvreader[before reading=\tiny, tabular=r|rrr|rrr|r, table head=
+\csvreader[before reading=\scriptsize, tabular=r|rrr|rrr|r, table head=
  & 

### 2016-10-23 paper/paper.tex
added conclusion and discussion of joint cardinality estimation algorithm results

@@
 \title{New cardinality estimation algorithms for HyperLogLog sketches}
-\subtitle{Draft version}
 \author{Otmar Ertl \\ otmar.ertl@gmail.com}
@@
 %TODO
-%In all other cases the cardinality of $\symSetA$ and $\symSetB$
 
@@
 
-We used the popular Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{Press2007} to solve this optimization problem.
-We used the implementation provided by the GNU scientific library \cite{Galassi2016}.
+For the general case the Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{Press2007} can be applied, which is a very popular algorithm for non-linear optimization problems.
+In particular, we used the implementation provided by the GNU scientific library \cite{Galassi2016}.
 
@@
 \subsection{Results}
-
-TODO see results \cref{tbl:joint_estimation_cases}, \cref{tbl:joint_estimation_a}, \cref{tbl:joint_estimation_b}, \cref{tbl:joint_estimation_x},
-\cref{tbl:joint_estimation_abx}
+In order to investigate the estimation error of the joint cardinality estimation algorithm \cref{alg:joint_cardinality_estimation} we constructed HyperLogLog sketch pairs with parameters $\symPrecision=20$ and $\symRegRange=44$ and with known cardinalities $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$.
+This is done by adding $|\symSetA|$ and $|\symSetB|$ random values to the first and second HyperLogLog sketch, respectively.
+Afterwards
+$|\symSetX|$ random values are added to both sketches.
+\cref{tbl:joint_estimation_cases} lists all the different cases we have evaluated.
+For each case \num{3333} different HyperLogLog sketch pairs were constructed and passed to the joint cardinality estimation algorithm.
+\cref{tbl:joint_estimation_cases} also shows the average number of function \eqref{equ:max_likelihood_phi_func} and gradient \eqref{equ:df_dphi}  evaluations, which the BFGS algorithm needed until convergence occured.
+
+\cref{tbl:joint_estimation_a}, \cref{tbl:joint_estimation_b}, and \cref{tbl:joint_estimation_x} compare the relative estimation error to the conventional approach using single sketch cardinality estimation together with \eqref{equ:conventional_approach}.
+The mean, the standard deviation, and the root-mean-square error are given.
+Furthermore, we calculated an improvement factor which represents the root-mean-square error ratio between both approaches.
+Since we only observed values larger than 1, the new maximum-likelihood estimation approach gives better estimates for all investigated cases.
+For some cases the improvement factor is clearly larger than 2.
+Due to square root error scaling law, this means that we would need 4 times more registers to get the same error when using the conventional approach.
+We also considered some cases with true cardinalities smaller  than the number of registers.
+As the results suggest and as expected the joint estimation algorithm works well over the entire cardinality range without special handling of small cardinalities.
+
+We also investigated, if the new approach could give better estimates for the union operation.
+The conventional approach merges both HyperLogLog sketches using \eqref{equ:hll_union} and estimates the union size using single sketch cardinality estimation.
+The joint cardinality estimation algorithm opens another way.
+The union can also be estimated by simply summing up the three estimates for the sets $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$.
+The corresponding results are shown in \cref{tbl:joint_estimation_abx}.
+As can be clearly seen the joint estimation algorithm is also able to improve the cardinality estimation of unions by an significant amount.
 
@@
 
-TODO
-
-%We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches.
+We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches.
 All approaches are based on the Poisson approximation.
@@
 
-%For the estimation from a single sketch, we have developed two fast algorithms that make use of the Poisson approximation.
+For the estimation from a single sketch, we have developed two fast algorithms that make use of the Poisson approximation.
 The first uses the original estimator extended by theoretically motivated correction terms.
 The second is based on the maximum-likelihood method and solves the corresponding optimization problem using the secant method.
-Both algorithms are inherently unbiased and can be implemented without empirical determined data.
-
-%Finally the maximum likelihood method was also applied to the estimation of set operation result sizes where the operands are represented by HyperLogLog sketches.
-The new approach significantly improves the cardinality estimates for set intersection, differences, and unions when compared to the inclusion-exclusion principle.
-An extension to more than two HyperLogLog sketches is possible.
-However, the effort scales like $\symBigO(2^n)$ if
+Both algorithms are inherently unbiased and can be implemented without dependence on empirically determined bias correction data.
+
+Finally the maximum likelihood method was also applied to the estimation of set operation result sizes where the operands are represented by HyperLogLog sketches.
+The new approach improves the cardinality estimates of set intersections, relative complements, as well as unions significantly when compared to conventional approaches such as the inclusion-exclusion principle.
+The presented maximum-likelihood method could also be used to estimate the cardinality of set operations between more than two HyperLogLog sketches.
+However, further research is necessary to determine, if this is feasible from a practical point of view.
+Similarly to the inclusion-exclusion principle, the effort would scale exponentially with the number of involved HyperLogLog sketches.
 

### 2016-10-26 paper/paper.tex
switched from gsl to dlib BFGS algorithm

@@
 For the general case the Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{Press2007} can be applied, which is a very popular algorithm for non-linear optimization problems.
-In particular, we used the implementation provided by the GNU scientific library \cite{Galassi2016}.
+In particular, we used the implementation provided by the Dlib C++ library \cite{King2009}.
 
@@
 The corresponding results are shown in \cref{tbl:joint_estimation_abx}.
-As can be clearly seen the joint estimation algorithm is also able to improve the cardinality estimation of unions by an significant amount.
+As can be clearly seen the joint estimation algorithm is also able to improve the cardinality estimation of unions by a significant amount.
 
@@
 \centering
-\caption{List of true cardinalities of the pair-wise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetA\cup\symSetX$ and  $\symSetB\cup\symSetX$ was investigated.
-The average number of function \eqref{equ:max_likelihood_phi_func} and gradient evaluations \eqref{equ:df_dphi} are also given as needed by our implementation of \cref{alg:joint_cardinality_estimation} that uses the BFGS algorithm.
+\caption{List of true cardinalities of the pair-wise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetS_1 = \symSetA\cup\symSetX$ and  $\symSetS_2 = \symSetB\cup\symSetX$ was investigated.
+The last column shows the average number of needed log-likelihood function
+ \eqref{equ:max_likelihood_phi_func} and gradient evaluations \eqref{equ:df_dphi} until convergence of the BFGS algorithm.
 }
@@
 \multicolumn{1}{c|}{cardinality ratio} &
-\multicolumn{2}{c}{avg num evaluations}
+\multicolumn{1}{c}{avg number}
 \\
@@
 \multicolumn{1}{c|}{$|\symSetX|$} & 
-\multicolumn{1}{c}{$\frac{|\symSetX|}{|\symSetA\cup\symSetB\cup\symSetX|}$} & 
-\multicolumn{1}{c|}{$\frac{|\symSetA\cup\symSetX|}{|\symSetB\cup\symSetX|}$} &
-\multicolumn{1}{c}{$\symFunc$} &
-\multicolumn{1}{c}{$\nabla\symFunc$}
+\multicolumn{1}{c}{$|\symSetS_1\cap\symSetS_2|/|\symSetS_1\cup\symSetS_2|$} & 
+\multicolumn{1}{c|}{$|\symSetS_1|/|\symSetS_2|$} &
+\multicolumn{1}{c}{iterations}
 \\\hline]
@@
 trueRatio=\trueRatio,
-avgNumFunctionEvaluations=\avgNumFunctionEvaluations,
-avgNumGradientEvaluations=\avgNumGradientEvaluations,
+avgNumEvaluations=\avgNumEvaluations
 }{
@@
 \numformattwo{\trueRatio} &
-\numformattwo{\avgNumFunctionEvaluations} &
-\numformattwo{\avgNumGradientEvaluations}
+\numformattwo{\avgNumEvaluations}
 }

### 2016-10-26 c++/cardinality_estimation.hpp

@@ -12,9 +12,12 @@
 #include <algorithm>
 #include <iostream>
 #include <cstdlib>
+#include <functional>
 
 #include <gsl/gsl_vector.h>
 #include <gsl/gsl_multimin.h>
+
+#include "dlib/optimization.h"
 
 #include "two_hyperloglog_statistic.hpp"
 
@@ -420,106 +423,108 @@ void inclusionExclusionTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& j
     cardinalityX = std::max(0., cardinalityBX + cardinalityAX - cardinalityABX);
 }
 
-void calcExp(double x, double& y, double& z) {
-    static const double ln2 = std::log(2);
-    if (x >= ln2) {
-        y = std::exp(-x);
-        z = 1 - y;
-    }
-    else {
-        z = -std::expm1(-x);
-        y = 1 - z;
-    }
-}
 
-// the minimum of this function gives the max likelihood estimate
-void eval_joint_log_likelihood_function_and_derivatives(
-    const TwoHyperLogLogStatistic& jointStatistic,
-    const double phiA,
-    const double phiB,
-    const double phiX,
-    double& f,
-    double& fa,
-    double& fb,
-    double& fx,
-    bool calcValue,
-    bool calcDerivative) {
 
+class LogLikelihoodFunctionForDlib {
+
+    const TwoHyperLogLogStatistic& jointStatistic;
+    int numEvaluations = 0;
+    double fa;
+    double fb;
+    double fx;
+    bool status = false;
     const int q = jointStatistic.getQ();
 
-    const double expPhiA = std::exp(phiA);
-    const double expPhiB = std::exp(phiB);
-    const double expPhiX = std::exp(phiX);
-
-    f = 0.;
-    fa = 0.;
-    fb = 0.;
-    fx = 0.;
-
-    double linTermA = 0;
-    double linTermB = 0;
-    double linTermX = 0;
-
-    double pow2k = ldexp(1., -q);
-    for (int k = q + 1; k >= 1; --k) {
-
-        int cSmaller1 = jointStatistic.getSmaller1Count(k);
-        int cLarger1 = jointStatistic.getLarger1Count(k);
-        int cSmaller2 = jointStatistic.getSmaller2Count(k);
-        int cLarger2 = jointStatistic.getLarger2Count(k);
-        int cEqual = jointStatistic.getEqualCount(k);
-
[…]

### 2016-10-26 c++/joint_cardinality_calculation.cpp

@@ -138,8 +138,7 @@ int main(int argc, char* argv[])
     resultsFile << "improvementStdevABX,";
     resultsFile << "improvementRmseABX,";
 
-    resultsFile << "avgNumFunctionEvaluations,";
-    resultsFile << "avgNumGradientEvaluations,";
+    resultsFile << "avgNumEvaluations,";
     resultsFile << "avgNumIterations,";
     resultsFile << "maxNumIterations" << endl;
 
@@ -180,8 +179,7 @@ int main(int argc, char* argv[])
         std::vector<double> maxLikeJaccardIdx;
         int maxNumIterations = 0;
         long numIterationsTotal = 0;
-        long numFunctionEvaluationsTotal = 0;
-        long numGradientEvaluationsTotal = 0;
+        long numEvaluationsTotal = 0;
         int size = 0;
 
         while (getline(statisticFile, line))
@@ -208,9 +206,8 @@ int main(int argc, char* argv[])
                 double estCardB = 0.;
                 double estCardX = 0.;
                 int numIterations = 0;
-                int numFunctionEvaluations = 0;
-                int numGradientEvaluations = 0;
-                maxLikelihoodTwoHyperLogLogEstimation(jointStatistic, estCardA, estCardB, estCardX, numIterations, numFunctionEvaluations, numGradientEvaluations);
+                int numEvaluations = 0;
+                maxLikelihoodTwoHyperLogLogEstimation(jointStatistic, estCardA, estCardB, estCardX, numIterations, numEvaluations);
                 maxLikeEstimatedCardA.push_back(estCardA/trueCardA-1.);
                 maxLikeEstimatedCardB.push_back(estCardB/trueCardB-1.);
                 maxLikeEstimatedCardX.push_back(estCardX/trueCardX-1.);
@@ -220,8 +217,7 @@ int main(int argc, char* argv[])
                 maxLikeEstimatedCardABX.push_back((estCardA+estCardB+estCardX)/(trueCardA+trueCardB+trueCardX)-1.);
                 numIterationsTotal += numIterations;
                 maxNumIterations = std::max(maxNumIterations, numIterations);
-                numFunctionEvaluationsTotal += numFunctionEvaluations;
-                numGradientEvaluationsTotal += numGradientEvaluations;
+                numEvaluationsTotal += numEvaluations;
             }
 
             size += 1;
@@ -280,8 +276,7 @@ int main(int argc, char* argv[])
[…]

### 2016-10-26 c++/cardinality_estimation.hpp
removed unused includes

@@ -14,9 +14,6 @@
 #include <cstdlib>
 #include <functional>
 
-#include <gsl/gsl_vector.h>
-#include <gsl/gsl_multimin.h>
-
 #include "dlib/optimization.h"
 
 #include "two_hyperloglog_statistic.hpp"

### 2016-11-12 paper/paper.tex
added section future work including idea for locality sensitive hashing, revised conclusion section

@@
 
-\section{Conclusion and future work}
-
-We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches.
-All approaches are based on the Poisson approximation.
-As shown, an unbiased estimator for the rate in the Poisson model, is also unbiased for the estimation of the cardinality.
-We have observed that this approach also works well for Poisson rate estimators that are not perfectly unbiased.
-Therefore, it would be interesting to investigate which conditions on an approximately unbiased Poisson rate estimator are sufficient to guarantee approximate unbiasedness if used as cardinality estimator.
-
+\section{Future work}
+As described in \cref{sec:depoissonization} an unbiased estimator for the rate in the Poisson model, is also unbiased for the estimation of the cardinality.
+We have shown that this approach works well for the corrected raw estimator as well as the maximum likelihood estimator even though both are only approximately unbiased.
+Therefore, it would be interesting to investigate, which conditions on an approximately unbiased Poisson rate estimator are sufficient to guarantee approximate unbiasedness, if used as cardinality estimator.
+
+Using the maximum-likelihood method we have been able to improve the cardinality estimates of intersections and relative complements of two HyperLogLog sketches.
+Unfortunately, finding the maximum using the BFGS algorithm requires many iterations which makes the joint cardinality estimation much more expensive than for a single HyperLogLog sketch.
+Since we have found the corrected raw estimator which is almost as accurate as the maximum-likelihood estimator for the single HyperLogLog case, we could imagine that there also exists a faster algorithm for the two HyperLogLog case.
+It is expected that such a new algorithm makes use of all the information given by the sufficient statistic.
+
+The presented maximum-likelihood method could also be used to estimate the cardinality of set operations between more than two HyperLogLog sketches.
+However, further research is necessary to determine, if this is feasible from a practical point of view.
+As for the inclusion-exclusion principle, the effort would scale exponentially with the number of involved HyperLogLog sketches.
+
+The maximum-likelihood method can also be used to estimate similarity measures such as the Jaccard index of two sets that are represented as HyperLogLog sketches.
+This directly leads to the question whether the HyperLogLog algorithm could be used for locality-sensitive hashing \cite{Leskovec2014, Wang2014}.
+Various locality-sensitive hashing algorithms have been proposed in the past.
+Among the most popular ones are the SimHash \cite{Charikar2002} and the min-wise hashing \cite{Broder1997, Li2011} algorithms whose hash collision probabilities are a function of the cosine and Jaccard similarities, respectively.
+
+The HyperLogLog algorithm itself can be regarded as hashing algorithm as it maps sets to register values.
+For sufficiently large cardinalities we can use the Poisson approximation and assume that the number of zero-valued HyperLogLog register can be ignored.
+Furthermore, if the HyperLogLog parameter $\symRegRange$ is chosen large enough, the number of saturated registers can be ignored as well.
+As a consequence, we can assume that the distribution of register values follows \eqref{equ:assumed_register_val_distribution} and the probability that a register has the same value for two different sets is (compare \eqref{equ:register_value_joint_pmf})
+\begin{equation}
+\symProbability(\symRegValVariate_1=\symRegValVariate_2)
+=
+\sum_{\symRegVal = -\infty}^{\infty}
+e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
+{\symNumReg 2^\symRegVal}
+}
+\left(
+1
+-
+e^{-\frac{\symPoissonRate_\symSetASuffix +  \symPoissonRate_\symSetXSuffix}
+{\symNumReg 2^\symRegVal}
+}
+-
+e^{-\frac{\symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
+{\symNumReg 2^\symRegVal}
+}
++
+e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
+{\symNumReg 2^\symRegVal}
+}
+\right).
+\end{equation}
+Using the approximation 
+$\sum_{\symRegVal=-\infty}^\infty
+e^{-\frac{\symX}{2^\symRegVal}} - e^{-\frac{\symY}{2^\symRegVal}}
+\approx
+2\symAlpha_\infty \left(\log(\symY)-\log(\symX)\right)$ we get
+\begin{equation}
+\symProbability(\symRegValVariate_1=\symRegValVariate_2)
+\approx
+1
++
+2\symAlpha_\infty
+\left(
+\log\left(
+1
+-
+\frac{1}{2}
+\frac{\symPoissonRate_\symSetASuffix}
+{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
+\symPoissonRate_\symSetXSuffix}
+\right)
++
+\log\left(
+1
+-
+\frac{1}{2}
+\frac{\symPoissonRate_\symSetBSuffix}
+{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
+\symPoissonRate_\symSetXSuffix}
+\right)
+\right).
+\end{equation}
+Further reduction by using the first order approximation $\log(1-\symX)\approx -\symX$ finally shows that the probability of a register having the same value for both sets is roughly linear to the Jaccard index $\frac{\symPoissonRate_\symSetXSuffix}
+{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
+\symPoissonRate_\symSetXSuffix}$
+\begin{equation}
+\symProbability(\symRegValVariate_1=\symRegValVariate_2)
+\approx
+\symAlpha_\infty
+\frac{\symPoissonRate_\symSetXSuffix}
+{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
+\symPoissonRate_\symSetXSuffix}
++
+(1-\symAlpha_\infty).
+\end{equation}
+This finding makes the HyperLogLog algorithm an interesting candidate for locality-sensitive hashing.
+What is also promising is the much simpler preprocessing because it only involves a single hash function (compare \cref{alg:insert}).
+In contrast, min-wise hashing is very costly, because many hash functions evaluations are necessary for all the needed permutations \cite{Li2011}.
+
+\section{Conclusion}
+
+We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches based on the Poisson approximation.
 For the estimation from a single sketch, we have developed two fast algorithms that make use of the Poisson approximation.
@@
 
-Finally the maximum likelihood method was also applied to the estimation of set operation result sizes where the operands are represented by HyperLogLog sketches.
+The maximum likelihood method was also applied to the estimation of set operation result sizes where the operands are represented by HyperLogLog sketches.
 The new approach improves the cardinality estimates of set intersections, relative complements, as well as unions significantly when compared to conventional approaches such as the inclusion-exclusion principle.
-The presented maximum-likelihood method could also be used to estimate the cardinality of set operations between more than two HyperLogLog sketches.
-However, further research is necessary to determine, if this is feasible from a practical point of view.
-Similarly to the inclusion-exclusion principle, the effort would scale exponentially with the number of involved HyperLogLog sketches.
-
 \appendix

### 2016-11-12 c++/cardinality_estimation.hpp
some minor changes

@@ -617,7 +617,7 @@ public:
 
 
 
-void maxLikelihoodTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& jointStatistic, double& cardinalityA, double& cardinalityB, double& cardinalityX, int& numIterations, int& numEvaluations) {
+void maxLikelihoodTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& jointStatistic, double& cardinalityA, double& cardinalityB, double& cardinalityX, int& numEvaluations) {
 
     const double eps = 1e-2;
 
@@ -636,7 +636,6 @@ void maxLikelihoodTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& jointS
         cardinalityX = 0;
         cardinalityA = cardinalityAX;
         cardinalityB = cardinalityBX;
-        numIterations = 0;
         return;
     }
     double cardinalityABX = estimator(jointStatistic.getMaxCounts());

### 2016-11-12 c++/joint_cardinality_calculation.cpp

@@ -137,10 +137,10 @@ int main(int argc, char* argv[])
     resultsFile << "improvementRmseBX,";
     resultsFile << "improvementStdevABX,";
     resultsFile << "improvementRmseABX,";
+    resultsFile << "improvementStdevJaccard,";
+    resultsFile << "improvementRmseJaccard,";
 
-    resultsFile << "avgNumEvaluations,";
-    resultsFile << "avgNumIterations,";
-    resultsFile << "maxNumIterations" << endl;
+    resultsFile << "avgNumEvaluations" << endl;
 
     int evaluationCounter = 0;
 
@@ -177,8 +177,6 @@ int main(int argc, char* argv[])
         std::vector<double> maxLikeEstimatedCardBX;
         std::vector<double> maxLikeEstimatedCardABX;
         std::vector<double> maxLikeJaccardIdx;
-        int maxNumIterations = 0;
-        long numIterationsTotal = 0;
         long numEvaluationsTotal = 0;
         int size = 0;
 
@@ -205,18 +203,15 @@ int main(int argc, char* argv[])
                 double estCardA = 0.;
                 double estCardB = 0.;
                 double estCardX = 0.;
-                int numIterations = 0;
                 int numEvaluations = 0;
-                maxLikelihoodTwoHyperLogLogEstimation(jointStatistic, estCardA, estCardB, estCardX, numIterations, numEvaluations);
+                maxLikelihoodTwoHyperLogLogEstimation(jointStatistic, estCardA, estCardB, estCardX, numEvaluations);
                 maxLikeEstimatedCardA.push_back(estCardA/trueCardA-1.);
                 maxLikeEstimatedCardB.push_back(estCardB/trueCardB-1.);
                 maxLikeEstimatedCardX.push_back(estCardX/trueCardX-1.);
-                maxLikeJaccardIdx.push_back((estCardX/(estCardA+estCardB+estCardX))/jaccardIndex-1.);
                 maxLikeEstimatedCardAX.push_back((estCardA+estCardX)/(trueCardA+trueCardX)-1.);
                 maxLikeEstimatedCardBX.push_back((estCardB+estCardX)/(trueCardB+trueCardX)-1.);
                 maxLikeEstimatedCardABX.push_back((estCardA+estCardB+estCardX)/(trueCardA+trueCardB+trueCardX)-1.);
-                numIterationsTotal += numIterations;
[…]

### 2016-11-19 paper/paper.tex
some minor changes
added algorithm for HyperLogLog compression

@@
 
-The desribed element insertion algorithm makes use of what is known as stochastic averaging \cite{Flajolet1985}.
-Instead of updating each of all $\symNumReg$ registers using $\symNumReg$ independent hash values, which would be an $\symBigO(\symNumReg)$ operation, only one register is selected and updated, which requires only a single hash function and reduces the complexity to $\symBigO(1)$.
-
-A HyperLogLog sketch can be characterized by a parameter pair $(\symPrecision, \symRegRange)$.
-The first parameter $\symPrecision$ is the precision and controls the relative error while the second defines the possible value range of a register.
-A register can take all values starting from 0 to $\symRegRange+1$, inclusively.
-The sum $\symPrecision+\symRegRange$ corresponds to the number of consumed hash value bits and defines the maximum cardinality that can be tracked.
-Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent and the estimation accuracy will be drastically reduced.
-
-At any time a $(\symPrecision, \symRegRange)$-HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be compressed into a $(\symPrecision', \symRegRange')$-HyperLogLog data structure, if $\symPrecision'\leq\symPrecision$ and $\symPrecision'+\symRegRange' \leq\symPrecision + \symRegRange$ is satisfied.
-This transformation is lossless in a sense that the resulting HyperLogLog sketch is the same as if all elements would have been recorded by a $(\symPrecision', \symRegRange')$-HyperLogLog sketch right from the beginning.
-
-\cref{alg:insert} has some properties which are especially useful in distributed environments.
-First, the insertion order of elements has no influence on the final HyperLogLog sketch.
-Furthermore, any two HyperLogLog sketches with same parameters $(\symPrecision, \symRegRange)$ representing  sets $\symSetA$ and $\symSetB$ can be easily merged.
-The resulting register values are simply given by the register-wise maximum values
-\begin{equation}
-\label{equ:hll_union}
-\symRegValVariate^{\symSetA\cup\symSetB}_{\symIndexI} = 
-\max
-\!
-\left(
-\symRegValVariate^{\symSetA}_{\symIndexI},
-\symRegValVariate^{\symSetB}_{\symIndexI}
-\right)
-\quad
-\text{for $1\leq\symIndexI\leq\symNumReg$}.
-\end{equation}
-The final HyperLogLog sketch represents the union of $\symSetA$ and $\symSetB$.
-
-A $(\symPrecision, 0)$-HyperLogLog sketch corresponds to a bit array as used by linear counting \cite{Whang1990}.
-Each register value can be stored by a single bit in this case.
-Hence, linear counting can be regarded as a special case of the HyperLogLog algorithm for which $\symRegRange = 0$.
-
 \begin{algorithm}
@@
 \comm{0.2}{$\symIndexI\in[1,2^\symPrecision]$}
-\State $\symRegVal = \min(\{\symS\mid \symS \in [1, \symRegRange]  \wedge  \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
+\State $\symRegVal \gets \min(\{\symS\mid \symS \in [1, \symRegRange]  \wedge  \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
 \comm{0.2}{$\symRegVal\in[1,\symRegRange+1]$}
@@
 \end{algorithm}
+
+The desribed element insertion algorithm makes use of what is known as stochastic averaging \cite{Flajolet1985}.
+Instead of updating each of all $\symNumReg$ registers using $\symNumReg$ independent hash values, which would be an $\symBigO(\symNumReg)$ operation, only one register is selected and updated, which requires only a single hash function and reduces the complexity to $\symBigO(1)$.
+
+A HyperLogLog sketch can be characterized by a parameter pair $(\symPrecision, \symRegRange)$.
+The first parameter $\symPrecision$ is the precision and controls the relative error while the second defines the possible value range of a register.
+A register can take all values starting from 0 to $\symRegRange+1$, inclusively.
+The sum $\symPrecision+\symRegRange$ corresponds to the number of consumed hash value bits and defines the maximum cardinality that can be tracked.
+Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent and the estimation accuracy will be drastically reduced.
+
+\cref{alg:insert} has some properties which are especially useful in distributed environments.
+First, the insertion order of elements has no influence on the final HyperLogLog sketch.
+Furthermore, any two HyperLogLog sketches with same parameters $(\symPrecision, \symRegRange)$ representing  sets $\symSetA$ and $\symSetB$ can be easily merged.
+The resulting register values are simply given by the register-wise maximum values
+\begin{equation}
+\label{equ:hll_union}
+\symRegValVariate^{\symSetA\cup\symSetB}_{\symIndexI} = 
+\max
+\!
+\left(
+\symRegValVariate^{\symSetA}_{\symIndexI},
+\symRegValVariate^{\symSetB}_{\symIndexI}
+\right)
+\quad
+\text{for $1\leq\symIndexI\leq\symNumReg$}.
+\end{equation}
+The final HyperLogLog sketch represents the union of $\symSetA$ and $\symSetB$.
+
+At any time a $(\symPrecision, \symRegRange)$-HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be compressed into a $(\symPrecision', \symRegRange')$-HyperLogLog data structure, if $\symPrecision'\leq\symPrecision$ and $\symPrecision'+\symRegRange' \leq\symPrecision + \symRegRange$ is satisfied (see \cref{alg:compress}).
+This transformation is lossless in a sense that the resulting HyperLogLog sketch is the same as if all elements would have been recorded by a $(\symPrecision', \symRegRange')$-HyperLogLog sketch right from the beginning.
+
+\begin{algorithm}
+\caption{Compression of a $(\symPrecision, \symRegRange)$-HyperLogLog sketch with register values $\boldsymbol{\symRegValVariate}$ into a $(\symPrecision', \symRegRange')$-HyperLogLog sketch with $\symPrecision'\leq\symPrecision$ and $\symPrecision'+\symRegRange' \leq\symPrecision + \symRegRange$.}
+\label{alg:compress}
+\begin{algorithmic}
+\Function {Compress}{\symDataItem, $\boldsymbol{\symRegValVariate}$}
+\State allocate registers for new HyperLogLog sketch $\boldsymbol{\symRegValVariate}' = (\symRegValVariate_1,\ldots,\symRegValVariate_{2^{\symPrecision'}})$
+\For{$\symIndexI\gets 1, 2^{\symPrecision'}$}
+\State $\symB \gets (\symIndexI-1)\cdot 2^{\symPrecision-\symPrecision'}$
+\State $\symIndexJ\gets 1$
+\While{$\symIndexJ \leq 2^{\symPrecision-\symPrecision'}
+\wedge
+\symRegValVariate_{\symB+\symIndexJ}=0
+$}
+\State $\symIndexJ\gets\symIndexJ+1$
+\EndWhile
+\If{$\symIndexJ = 1$}
+\State $\symRegValVariate'_\symIndexI \gets \min(\symRegValVariate_{\symB+\symIndexJ} + \symPrecision-\symPrecision',\symRegRange'+1)$
+\ElsIf{$\symIndexJ\leq2^{\symPrecision-\symPrecision'}$}
+\State $\langle \symBitRepA_1, \ldots, \symBitRepA_{\symPrecision-\symPrecision'}\rangle_2 \gets \symIndexJ-1$
+\State $\symRegValVariate'_\symIndexI \gets 
+\min(\{\symS\mid \symS \in [1, \symPrecision-\symPrecision']  \wedge  \symBitRepA_\symS = 1\} \cup \{\symRegRange'+1\})$
+\Else
+\State $\symRegValVariate'_\symIndexI \gets 0$
+\EndIf
+\EndFor
+\State \Return $\boldsymbol{\symRegValVariate}'$
+\EndFunction
+\end{algorithmic}
+\end{algorithm}
+
+A $(\symPrecision, 0)$-HyperLogLog sketch corresponds to a bit array as used by linear counting \cite{Whang1990}.
+Each register value can be stored by a single bit in this case.
+Hence, linear counting can be regarded as a special case of the HyperLogLog algorithm for which $\symRegRange = 0$.
 
@@
 \end{equation}
+
 The original estimation algorithm as presented in \cite{Flajolet2007} including corrections for small and large cardinalities is summarized by \cref{alg:estimate_original}. 
@@
 
+%\begin{table}
+%\centering
+%\caption{TODO}
+%\label{tbl:joint_estimation_jaccard}
+%\csvreader[before reading=\scriptsize, tabular=r|rrr|rrr|r, table head=
+% & 
+%\multicolumn{3}{c|}{conventional approach}
+%&
+%\multicolumn{3}{c|}{maximum likelihood method}
+%&
+%\multicolumn{1}{c}{improvement}
+%\\
+%\# & 
+%\multicolumn{1}{c}{mean} & 
+%\multicolumn{1}{c}{stdev} & 
+%\multicolumn{1}{c|}{rmse} & 
+%\multicolumn{1}{c}{mean} & 
+%\multicolumn{1}{c}{stdev} & 
+%\multicolumn{1}{c|}{rmse} &
+%\multicolumn{1}{c}{rmse ratio}
+%\\\hline]
+%{joint_cardinality_calculation.csv}{
+%inclExclMeanJaccardIdx=\inclExclMeanJaccardIdx,
+%inclExclStdDevJaccardIdx=\inclExclStdDevJaccardIdx,
+%inclExclRMSEJaccardIdx=\inclExclRMSEJaccardIdx,
+%maxLikeMeanJaccardIdx=\maxLikeMeanJaccardIdx,
+%maxLikeStdDevJaccardIdx=\maxLikeStdDevJaccardIdx,
+%maxLikeRMSEJaccardIdx=\maxLikeRMSEJaccardIdx,
+%improvementRmseJaccard=\improvementRmseJaccard
+%}{
+%\thecsvrow & 
+%\numformat{\inclExclMeanJaccardIdx} & 
+%\numformat{\inclExclStdDevJaccardIdx} & 
+%\numformat{\inclExclRMSEJaccardIdx} & 
+%\numformat{\maxLikeMeanJaccardIdx} & 
+%\numformat{\maxLikeStdDevJaccardIdx} & 
+%\numformat{\maxLikeRMSEJaccardIdx} &
+%\numformattwo{\improvementRmseJaccard}
+%}
+%\end{table}
+
 \section{Future work}

### 2016-11-19 paper/paper.tex
revised estimation error sections

@@
 \symDefine{\symIndexI}{i}
-\symDefine{\symIndexK}{k}
-\symDefine{\symIndexL}{l}
 \symDefine{\symIndexJ}{j}
@@
 \symDefine{\symEpsPowerSeriesFunc}{\varepsilon_\symPowerSeriesFunc}
+\symDefine{\symBiasCorrectionFunc}{w_{\text{corr}}}
 
@@
 
-The corrected raw estimator beats the precision of methods that use bias correction \cite{Heule2013,Rhodes2015,Sanfilippo2014}.
-\cref{fig:stdev_comparison} shows the standard deviation of the raw estimator after bias correction.
-At the transition region to small cardinalities, for which all previous methods switch over to the linear counting estimator at some point, the corrected raw estimator has smaller variance.
-Obviously, the previous approaches cannot do better than given by the minimum of both curves for linear counting and raw estimator.
-In practice, the standard deviation is even larger, because the choice between both estimators must be made based on an estimate and not on the true cardinality, for which the intersection point of both curves represents the ideal transition point.
-
 \begin{figure}
@@
 
+The corrected raw estimator beats the precision of methods that apply bias correction on the raw estimator \cite{Heule2013,Rhodes2015,Sanfilippo2014}.
+Based on the simulated data we have empirically determined the bias correction function $\symBiasCorrectionFunc$ for the raw estimator that satisfies $\symCardinality = \symExpectation(\symBiasCorrectionFunc(\symCardinalityRawEstimate)\vert\symCardinality)$ for all cardinalities.
+By definition,
+the estimator $\symCardinalityRawEstimate':=\symBiasCorrectionFunc(\symCardinalityRawEstimate)$ is unbiased and a function of the raw estimator.
+Its standard deviation can be compared with that of the corrected raw estimator in \cref{fig:stdev_comparison}.
+For cardinalities smaller than \num{10000} the empirical bias correction approach is not very precise.
+This is the reason why all previous approaches had to switch over to the linear counting estimator at some point.
+The standard deviation of the linear counting estimator is also shown in \cref{fig:stdev_comparison}.
+Obviously, the previous approaches cannot do better than given by the minimum of both curves for linear counting and raw estimator.
+In practice, the standard deviation is even larger, because the choice between both estimators must be made based on an estimate and not on the true cardinality, for which the intersection point of both curves represents the ideal transition point.
+In contrast, the corrected raw estimator performs well over the entire cardinality range.
+
 \begin{figure}
@@
 \includesvg[width=1\textwidth]{stdev_comparison}
-\caption{Standard deviations of different cardinality estimators for a HyperLogLog sketch with paramters $\symPrecision = 12$ and $\symRegRange=20$ as functions of the true cardinality.}
+\caption{Standard deviations of the relative error of different cardinality estimators for a HyperLogLog sketch with paramters $\symPrecision = 12$ and $\symRegRange=20$ over the true cardinality.}
 \label{fig:stdev_comparison}
@@
 What is different for the maximum likelihood estimation approach is a somewhat smaller median bias with less oscillations around zero for small cardinalities.
+The standard deviation of the relative error is also slightly better for the maximum likelihood estimator than for the corrected raw estimator as shown in \cref{fig:stdev_comparison}.
 Furthermore, contrary to the raw estimator which reveals a small oscillating bias for the mean (see \cref{fig:raw_corrected_estimation_error_22_10}), the maximum likelihood estimator seems to be completely unbiased (see \cref{fig:max_likelihood_estimation_error_22_10}).
@@
 This finding makes the HyperLogLog algorithm an interesting candidate for locality-sensitive hashing.
-What is also promising is the much simpler preprocessing because it only involves a single hash function (compare \cref{alg:insert}).
+What also looks promising is the much simpler preprocessing because it only involves a single hash function (compare \cref{alg:insert}).
 In contrast, min-wise hashing is very costly, because many hash functions evaluations are necessary for all the needed permutations \cite{Li2011}.

### 2016-11-19 python/comparison.py

@@ -24,7 +24,7 @@ def readData(filename):
 config = [
 ["12_20_max_likelihood_estimates.dat", "blue", False, "maximum likelihood estimator"],
 #["12_20_flajolet_mid_range_estimates.dat", "gray", False, "raw estimator"],
-["12_20_flajolet_mid_range_estimates.dat", "red", True, "raw estimator (bias corrected)"],
+["12_20_flajolet_mid_range_estimates.dat", "red", True, "subsequently bias corrected raw estimator"],
 ["12_20_flajolet_small_range_estimates.dat", "black", False, "linear counting estimator"],
 ["12_20_corrected_raw_estimates.dat", "green", False, "corrected raw estimator"]
 ]

### 2016-11-19 paper/paper.tex
some more imporvements

@@
 In case the data is not distributed and results do not need to be aggregated further, there are even more efficient estimation algorithms available.
-On the one hand there is the self-learning bitmap \cite{Chen2011, Chen2015}, and on the other hand there is the HyperLogLog algorithm extended by a historic inverse probability estimator that is continuously updated together with the HyperLogLog sketch \cite{Ting2014}.
+On the one hand there is the self-learning bitmap \cite{Chen2011}, and on the other hand there is the HyperLogLog algorithm extended by a historic inverse probability estimator that is continuously updated together with the HyperLogLog sketch \cite{Ting2014}.
 Both achieve the same estimation error using less space.
@@
 \caption{List of true cardinalities of the pair-wise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetS_1 = \symSetA\cup\symSetX$ and  $\symSetS_2 = \symSetB\cup\symSetX$ was investigated.
-The last column shows the average number of needed log-likelihood function
- \eqref{equ:max_likelihood_phi_func} and gradient evaluations \eqref{equ:df_dphi} until convergence of the BFGS algorithm.
+The last column shows the average number of log-likelihood function
+ \eqref{equ:max_likelihood_phi_func} and gradient \eqref{equ:df_dphi} evaluations until the stop criterion of the BFGS algorithm was satisfied.
 }
@@
 
-Using the maximum-likelihood method we have been able to improve the cardinality estimates of intersections and relative complements of two HyperLogLog sketches.
-Unfortunately, finding the maximum using the BFGS algorithm requires many iterations which makes the joint cardinality estimation much more expensive than for a single HyperLogLog sketch.
+Using the maximum-likelihood method we have been able to improve the cardinality estimates for the results of set operations between two HyperLogLog sketches.
+Unfortunately, joint cardinality estimation is much more expensive than for a single HyperLogLog sketch, because it requires maximization of a multi-dimensional function.
 Since we have found the corrected raw estimator which is almost as accurate as the maximum-likelihood estimator for the single HyperLogLog case, we could imagine that there also exists a faster algorithm for the two HyperLogLog case.
-It is expected that such a new algorithm makes use of all the information given by the sufficient statistic.
+It is expected that such a new algorithm makes use of all the information given by the sufficient statistic \eqref{equ:sufficient_joint_statistic}.
 
@@
 What also looks promising is the much simpler preprocessing because it only involves a single hash function (compare \cref{alg:insert}).
-In contrast, min-wise hashing is very costly, because many hash functions evaluations are necessary for all the needed permutations \cite{Li2011}.
+In contrast, min-wise hashing is very costly, because many hash function evaluations are necessary for all the needed permutations \cite{Li2011}.
 

### 2016-11-21 paper/paper.tex
minor changes

@@
 \symDefine{\symIndexJ}{j}
+\symDefine{\symIndexL}{l}
 \symDefine{\symCount}{c}
@@
 \end{equation}
- is a smooth periodic function with period 1. Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
+is a smooth periodic function with period 1. 
+%Its Fourier series is
+%\begin{equation}
+%\symPowerSeriesFunc(\symX) = 1 + \sum_{\symIndexL=1}^{\infty}
+%\symA_\symIndexL \cos(2\pi\symIndexL\symX) + \symB_\symIndexL \sin(2\pi\symIndexL\symX)
+%\end{equation}
+%with coefficients
+%\begin{equation}
+%\begin{aligned}
+%\symA_\symIndexL &= 2\log(2)\int_{-\infty}^\infty 2^\symX e^{-2^\symX}\cos(2\pi\symIndexL\symX)d\symX, \\
+%\symB_\symIndexL &= 2\log(2)\int_{-\infty}^\infty 2^\symX e^{-2^\symX}\sin(2\pi\symIndexL\symX)d\symX.
+%\end{aligned}
+%\end{equation}
+%Numerical evaluations give for the $\symA_1\approx \num{6.353e-6}$ and $\symB_1\approx \num{-7.572e-6}$.
+Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
 \leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=\num{9.885e-6}$ (see \cref{fig:power_series_func}).
@@
 Assume two given  HyperLogLog sketches with register values $\boldsymbol{\symRegValVariate}_1$ and $\boldsymbol{\symRegValVariate}_2$ representing the sets $\symSetS_1$ and $\symSetS_2$, respectively.
-The goal is to find estimates for the cardinalities of the pair-wise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$.
+The goal is to find estimates for the cardinalities of the pairwise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$.
 The Poisson approximation allows us to assume that the elements of $\symSetA$ are inserted into the HyperLogLog sketch representing $\symSetS_1$ at rate $\symPoissonRate_\symSetASuffix$.
@@
 \subsection{Joint log-likelihood funtion}
-In order to get maximum-likelihood estimators for $\symPoissonRateEstimate_\symSetASuffix$, $\symPoissonRateEstimate_\symSetBSuffix$, and $\symPoissonRateEstimate_\symSetXSuffix$ we need to derive the joint probability distribution of two HyperLogLog sketches.
+In order to get maximum likelihood estimators for $\symPoissonRateEstimate_\symSetASuffix$, $\symPoissonRateEstimate_\symSetBSuffix$, and $\symPoissonRateEstimate_\symSetXSuffix$ we need to derive the joint probability distribution of two HyperLogLog sketches.
 Under the Poisson model the register values are independent and identically distributed.
@@
 \end{align}
-Here the last transformation used the independence of $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$, because by definition, the sets $\symSetA$, $\symSetB$, and $\symSetX$ are disjoint.
+Here the last transformation used the independence of $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$, because by definition, the sets $\symSetA$, $\symSetB$, and $\symSetX$ are pairwise disjoint.
 Furthermore, under the Poisson model $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ obey
@@
 Here $\symPoissonRateEstimate_1$ and $\symPoissonRateEstimate_2$ are the maximum likelihood cardinality estimates for the first and second HyperLogLog sketch, respectively.
-This means that the maximum-likelihood makes no clear statement about the intersection size in this case.
+This means that the maximum likelihood method makes no clear statement about the intersection size in this case.
 The estimate for $\symPoissonRate_\symX$ could be anything between 0 and $\symPoissonRateEstimate_2$.
@@
 %\symPoissonRate_\symSetBSuffix=0
-%$ and all the register values must be pair-wise identical, which finally 
+%$ and all the register values must be pairwise identical, which finally 
 %yields \eqref{equ:log_likelihood_single} for $\symPoissonRate_\symSetXSuffix$.
@@
 $(\boldsymbol\symCountVariate^{>}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2})$.
-In contrast, the maximum-likelihood method incorporates each value of the sufficient statistic \eqref{equ:sufficient_joint_statistic} individually.
+In contrast, the maximum likelihood method incorporates each value of the sufficient statistic \eqref{equ:sufficient_joint_statistic} individually.
 
@@
 
-%As demonstrated by we In all other cases, the maximum-likelihood estimates for $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$ will be non-zero.
+%As demonstrated by we In all other cases, the maximum likelihood estimates for $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$ will be non-zero.
 We use the conventional approach to get start values for the estimates of $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$, respectively.
@@
 Furthermore, we calculated an improvement factor which represents the root-mean-square error ratio between both approaches.
-Since we only observed values larger than 1, the new maximum-likelihood estimation approach gives better estimates for all investigated cases.
+Since we only observed values larger than 1, the new maximum likelihood estimation approach gives better estimates for all investigated cases.
 For some cases the improvement factor is clearly larger than 2.
@@
 \centering
-\caption{List of true cardinalities of the pair-wise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetS_1 = \symSetA\cup\symSetX$ and  $\symSetS_2 = \symSetB\cup\symSetX$ was investigated.
+\caption{List of true cardinalities of the pairwise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetS_1 = \symSetA\cup\symSetX$ and  $\symSetS_2 = \symSetB\cup\symSetX$ was investigated.
 The last column shows the average number of log-likelihood function
@@
 
-Using the maximum-likelihood method we have been able to improve the cardinality estimates for the results of set operations between two HyperLogLog sketches.
+Using the maximum likelihood method we have been able to improve the cardinality estimates for the results of set operations between two HyperLogLog sketches.
 Unfortunately, joint cardinality estimation is much more expensive than for a single HyperLogLog sketch, because it requires maximization of a multi-dimensional function.
-Since we have found the corrected raw estimator which is almost as accurate as the maximum-likelihood estimator for the single HyperLogLog case, we could imagine that there also exists a faster algorithm for the two HyperLogLog case.
+Since we have found the corrected raw estimator which is almost as accurate as the maximum likelihood estimator for the single HyperLogLog case, we could imagine that there also exists a faster algorithm for the two HyperLogLog case.
 It is expected that such a new algorithm makes use of all the information given by the sufficient statistic \eqref{equ:sufficient_joint_statistic}.
 
-The presented maximum-likelihood method could also be used to estimate the cardinality of set operations between more than two HyperLogLog sketches.
+The presented maximum likelihood method could also be used to estimate the cardinality of set operations between more than two HyperLogLog sketches.
 However, further research is necessary to determine, if this is feasible from a practical point of view.
@@
 
-The maximum-likelihood method can also be used to estimate similarity measures such as the Jaccard index of two sets that are represented as HyperLogLog sketches.
+The maximum likelihood method can also be used to estimate similarity measures such as the Jaccard index of two sets that are represented as HyperLogLog sketches.
 This directly leads to the question whether the HyperLogLog algorithm could be used for locality-sensitive hashing \cite{Leskovec2014, Wang2014}.
@@
 The first uses the original estimator extended by theoretically motivated correction terms.
-The second is based on the maximum-likelihood method and solves the corresponding optimization problem using the secant method.
+The second is based on the maximum likelihood method and solves the corresponding optimization problem using the secant method.
 Both algorithms are inherently unbiased and can be implemented without dependence on empirically determined bias correction data.

### 2016-11-27 paper/paper.tex
added appendix that derives the deviation of function xi from constant 1

@@
 % SYMBOLS
+\renewcommand\Re{\operatorname{Re}}
+\renewcommand\Im{\operatorname{Im}}
+
 \newcommand*{\symDefine}[2]{\newcommand{{#1}}{{#2}}}
@@
 \symDefine{\symIndexI}{i}
+\symDefine{\symImaginary}{\rm{i}}
 \symDefine{\symIndexJ}{j}
@@
 is a smooth periodic function with period 1. 
-%Its Fourier series is
-%\begin{equation}
-%\symPowerSeriesFunc(\symX) = 1 + \sum_{\symIndexL=1}^{\infty}
-%\symA_\symIndexL \cos(2\pi\symIndexL\symX) + \symB_\symIndexL \sin(2\pi\symIndexL\symX)
-%\end{equation}
-%with coefficients
-%\begin{equation}
-%\begin{aligned}
-%\symA_\symIndexL &= 2\log(2)\int_{-\infty}^\infty 2^\symX e^{-2^\symX}\cos(2\pi\symIndexL\symX)d\symX, \\
-%\symB_\symIndexL &= 2\log(2)\int_{-\infty}^\infty 2^\symX e^{-2^\symX}\sin(2\pi\symIndexL\symX)d\symX.
-%\end{aligned}
-%\end{equation}
-%Numerical evaluations give for the $\symA_1\approx \num{6.353e-6}$ and $\symB_1\approx \num{-7.572e-6}$.
 Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
 \leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=\num{9.885e-6}$ (see \cref{fig:power_series_func}).
+This value can also be found using Fourier analysis as shown in in \cref{app:analysis_xi}.
 
@@
 The more relevant case with stochastic averaging, which is considered in this paper, is more complicated.
-However, with the help of the Poisson approximation, we are able to still apply the maximum likelihood method and to derive a new robust and efficient cardinality estimation algorithm.
+However, with the help of the Poisson approximation, it is still feasible to apply the maximum likelihood method and we are able to derive a new robust and efficient cardinality estimation algorithm.
 Furthermore, we will demonstrate that consequent application of the maximum likelihood method reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values.
@@
 \left(
-\log\left(
+\log\!\left(
 1
@@
 \appendix
+
+\section{Analysis of $\symPowerSeriesFunc(\symX)$}
+\label{app:analysis_xi}
+The Fourier series of the function $\symPowerSeriesFunc(\symX)$, which is defined by
+\begin{equation}
+\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
+2^{\symRegVal+\symX}
+e^{-2^{\symRegVal+\symX}}
+\end{equation}
+and which has period equal to 1, is
+\begin{equation}
+\symPowerSeriesFunc(\symX) = 1 + \Re\!\left(\sum_{\symIndexL=1}^{\infty}
+\symA_\symIndexL \exp(2\pi\symIndexL\symX\symImaginary)\right)
+\end{equation}
+with coefficients
+\begin{equation}
+\symA_\symIndexL = 2\int_0^1 \symPowerSeriesFunc(\symX) \exp(-2\pi\symIndexL\symX\symImaginary)d\symX
+=
+2\log(2) \int_{-\infty}^{\infty} 2^\symX
+e^{-2^\symX}\exp(-2\pi\symIndexL\symX\symImaginary)d\symX.
+\end{equation}
+The variable transformation $\symY = 2^\symX$ yields
+\begin{equation}
+\symA_\symIndexL
+=
+2\int_{0}^{\infty} e^{-\symY}
+\symY^{-\frac{2\pi\symIndexL\symImaginary}{\log(2)}}d\symY
+= 
+2\Gamma\!\left(1-\frac{2\pi\symIndexL\symImaginary}{\log(2)}\right)
+\end{equation}
+where $\Gamma$ denotes the gamma function. Using the identity $\left|\Gamma(1+\symImaginary\symX)\right| = 
+\sqrt{\frac{\pi\symX}{\sinh(\pi\symX)}}
+$
+we are able to write for the absolute values of the coefficients
+\begin{equation}
+|\symA_\symIndexL| = 
+2\sqrt{\frac{\symB\symIndexL}{\sinh(\symB\symIndexL)}}
+\quad
+\text{with}
+\
+\symB:=\frac{2\pi^2}{\log 2}.
+\end{equation}
+In particular, the amplitude of the first harmonic is 
+$|\symA_1|
+\approx
+\num{9.884e-6}$. Obviously, the maximum deviation of $\symPowerSeriesFunc(\symX)-1$ from the first harmonic must be smaller than $\sum_{\symIndexL=2}^{\infty} 
+|\symA_\symIndexL|$. The ratio of subsequent coefficients is given by
+\begin{equation}
+\frac{|\symA_{\symIndexL+1}|}{|\symA_{\symIndexL}|}
+=
+\sqrt{\frac{\symIndexL+1}{\symIndexL}}
+\sqrt{\frac{
+\sinh(\symB \symIndexL)
+}{
+\sinh(\symB (\symIndexL+1))
+}}
+=
+\sqrt{\frac{\symIndexL+1}{\symIndexL}}
+\sqrt{\frac{
+1
+}{
+\cosh(\symB)
++
+\frac{
+\sinh(\symB)
+}
+{
+\tanh(\symB \symIndexL)
+}
+}}.
+\end{equation}
+For $\symIndexL\geq 2$ we have $\sqrt{\frac{\symIndexL+1}{\symIndexL}}\leq \sqrt{\frac{3}{2}}$. Together with $\tanh(x)\leq 1$ we obtain
+\begin{equation}
+\frac{|\symA_{\symIndexL+1}|}{
+|\symA_{\symIndexL}|
+}
+\leq
+\sqrt{\frac{3}{2}}
+\sqrt{\frac{
+1
+}{
+\cosh(\symB)
++
+\sinh(\symB)
+}}
+\\
+=
+\sqrt{\frac{
+3
+}{
+2 e^\symB
+}}.
+\end{equation}
+Consequently, $|\symA_{\symIndexL}| \leq 
+|\symA_{2}|
+\left(\sqrt{\frac{
+3
+}{
+2 e^\symB
+}}\right)^{\symIndexL-2}$ for $\symIndexL\geq 2$ and 
+\begin{equation}
+\sum_{\symIndexL=2}^{\infty} 
+|\symA_\symIndexL|
+\leq
+|\symA_2|
+\sum_{\symIndexL=0}^{\infty} 
+\left(
+\sqrt{\frac{
+3
+}{
+2e^\symB
+}}
+\right)^\symIndexL
+=
+2\sqrt{
+\frac{2\symB}{\sinh(2\symB)}}
+\frac{1}{1-
+\sqrt{\frac{
+3
+}{
+2 e^\symB
+}}
+}
+\approx
+\num{9.154e-12}.
+\end{equation}
+For the maximum deviation from 1 we finally have
+\begin{equation}
+\max(|\symPowerSeriesFunc(\symX)-1|) \leq \sum_{\symIndexL=1}^{\infty} 
+|\symA_\symIndexL|
+\leq \num{9.885e6}.
+\end{equation}
 

### 2016-12-03 paper/paper.tex
some minor changes
added chart showing the probability of equal registers as a function of the Jaccard distance + some minor improvements

@@
 \symDefine{\symDataItem}{D}
+\symDefine{\symDistanceMeasure}{D}
 \symDefine{\symBitRepA}{a}
@@
 \leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=\num{9.885e-6}$ (see \cref{fig:power_series_func}).
-This value can also be found using Fourier analysis as shown in in \cref{app:analysis_xi}.
+This value can also be found using Fourier analysis as shown in \cref{app:analysis_xi}.
 
@@
 For large cardinalities it is worth to take the strong \eqref{equ:strong_lower_bound} instead of the weak lower bound \eqref{equ:weak_lower_bound} as initial value, which requires a single logarithm evaluation.
-The stronger bound is a much better approximation especially for large cardinalities, where the extra logarithm evaluation is amortized by the savings of iteration cycles.
+The stronger bound is a much better approximation especially for large cardinalities, where the extra logarithm evaluation is amortized by savings in the number of iteration cycles.
 
@@
 The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:avg_exec_time}.
-The average computation time for the maximum likelihood algorithm shows a different behavior than for the corrected raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}).
-As can be seen, the average execution time is larger for most cardinalities, but nevertheless, it never exceeds \SI{700}{\nano\second} which is still fast enough for many applications.
+The average computation time for the maximum likelihood algorithm shows a different behavior than for the corrected raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}).As can be seen, the average execution time is larger for most cardinalities, but nevertheless, it never exceeds \SI{700}{\nano\second} which is still fast enough for many applications.
+The steps in the chart can be explained by different numbers of iteration cycles until the secant method is stopped.
+For example, at a cardinality value around \num{5000} the average number of required cycles until the stop criterion is met increases abruptly from two to three.
+More than three iteration cycles have never been necessary for any cardinality estimate in this performance test.
 
@@
 \caption{List of true cardinalities of the pairwise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetS_1 = \symSetA\cup\symSetX$ and  $\symSetS_2 = \symSetB\cup\symSetX$ was investigated.
-The last column shows the average number of log-likelihood function
- \eqref{equ:max_likelihood_phi_func} and gradient \eqref{equ:df_dphi} evaluations until the stop criterion of the BFGS algorithm was satisfied.
+The last column shows the average number of iteration cycles until the stop criterion of the BFGS algorithm was satisfied.
 }
@@
 As described in \cref{sec:depoissonization} an unbiased estimator for the rate in the Poisson model, is also unbiased for the estimation of the cardinality.
-We have shown that this approach works well for the corrected raw estimator as well as the maximum likelihood estimator even though both are only approximately unbiased.
-Therefore, it would be interesting to investigate, which conditions on an approximately unbiased Poisson rate estimator are sufficient to guarantee approximate unbiasedness, if used as cardinality estimator.
+We have shown that this approach works well for the corrected raw estimator as well as for the maximum likelihood estimator even though both are only approximately unbiased.
+Therefore, it would be interesting what conditions on an approximately unbiased Poisson rate estimator are sufficient to guarantee approximate unbiasedness, if used as cardinality estimator.
 
@@
 The HyperLogLog algorithm itself can be regarded as hashing algorithm as it maps sets to register values.
-For sufficiently large cardinalities we can use the Poisson approximation and assume that the number of zero-valued HyperLogLog register can be ignored.
+For sufficiently large cardinalities we can use the Poisson approximation and assume that the number of zero-valued HyperLogLog registers can be ignored.
 Furthermore, if the HyperLogLog parameter $\symRegRange$ is chosen large enough, the number of saturated registers can be ignored as well.
@@
 2\symAlpha_\infty \left(\log(\symY)-\log(\symX)\right)$ we get
+
 \begin{equation}
@@
 2\symAlpha_\infty
-\left(
 \log\!\left(
@@
 \frac{1}{2}
-\frac{\symPoissonRate_\symSetASuffix}
+\symDistanceMeasure
++
+\frac{1}{4}
+\symDistanceMeasure^2
+\frac{
+\symPoissonRate_\symSetASuffix
+\symPoissonRate_\symSetBSuffix
+}{\left(
+\symPoissonRate_\symSetASuffix+
+\symPoissonRate_\symSetBSuffix
+\right)^2}
+\right)
+\end{equation}
+where $\symDistanceMeasure = \frac{\symPoissonRate_\symSetASuffix
++\symPoissonRate_\symSetBSuffix}
 {\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
-\symPoissonRate_\symSetXSuffix}
-\right)
-+
+\symPoissonRate_\symSetXSuffix}$
+ is the Jaccard distance. Since $
+\frac{
+\symPoissonRate_\symSetASuffix
+\symPoissonRate_\symSetBSuffix
+}{\left(
+\symPoissonRate_\symSetASuffix+
+\symPoissonRate_\symSetBSuffix
+\right)^2}
+$ is always in the range $[0,\frac{1}{4}]$, the probability for equal register values can be bounded by
+\begin{equation}
+1
++
+2\symAlpha_\infty
 \log\!\left(
@@
 \frac{1}{2}
-\frac{\symPoissonRate_\symSetBSuffix}
-{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
-\symPoissonRate_\symSetXSuffix}
-\right)
+\symDistanceMeasure
+\right)
+\lesssim
+\symProbability(\symRegValVariate_1=\symRegValVariate_2)
+\lesssim
+1
++
+2\symAlpha_\infty
+\log\!\left(
+1
+-
+\frac{1}{2}
+\symDistanceMeasure
++
+\frac{1}{16}
+\symDistanceMeasure^2
 \right).
 \end{equation}
-Further reduction by using the first order approximation $\log(1-\symX)\approx -\symX$ finally shows that the probability of a register having the same value for both sets is roughly linear to the Jaccard index $\frac{\symPoissonRate_\symSetXSuffix}
-{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
-\symPoissonRate_\symSetXSuffix}$
+As shown in \cref{fig:equal_register_probability} the bounds are very close, especially for small Jaccard distances where the probability can be well approximated by
 \begin{equation}
@@
 \approx
-\symAlpha_\infty
-\frac{\symPoissonRate_\symSetXSuffix}
-{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
-\symPoissonRate_\symSetXSuffix}
-+
-(1-\symAlpha_\infty).
-\end{equation}
+1-
+\symAlpha_\infty\symDistanceMeasure.
+\end{equation}
+\begin{figure}
+\centering
+\includesvg[width=0.6\textwidth]{equal-register-probability}
+\caption{The approximate probability range of equal register values as a function of the Jaccard distance.}
+\label{fig:equal_register_probability}
+\end{figure}
 This finding makes the HyperLogLog algorithm an interesting candidate for locality-sensitive hashing.
@@
 \symPowerSeriesFunc(\symX) = 1 + \Re\!\left(\sum_{\symIndexL=1}^{\infty}
-\symA_\symIndexL \exp(2\pi\symIndexL\symX\symImaginary)\right)
+\symA_\symIndexL e^{2\pi\symImaginary\symIndexL\symX}\right)
 \end{equation}
 with coefficients
-\begin{equation}
-\symA_\symIndexL = 2\int_0^1 \symPowerSeriesFunc(\symX) \exp(-2\pi\symIndexL\symX\symImaginary)d\symX
+\begin{multline}
+\symA_\symIndexL = 2\int_0^1 \symPowerSeriesFunc(\symX) e^{-2\pi\symImaginary\symIndexL\symX}d\symX
+=
+2\log(2)
+\int_0^1
+\sum_{\symRegVal = -\infty}^\infty
+2^{\symRegVal+\symX}
+e^{-2^{\symRegVal+\symX}}
+e^{-2\pi\symImaginary\symIndexL\symX}
+d\symX
+=
+\\
+=
+2\log(2)
+\sum_{\symRegVal = -\infty}^\infty
+\int_{\symRegVal}^{\symRegVal+1}
+2^{\symX}
+e^{-2^{\symX}}
+e^{-2\pi\symImaginary\symIndexL\symX}
+d\symX
 =
 2\log(2) \int_{-\infty}^{\infty} 2^\symX
-e^{-2^\symX}\exp(-2\pi\symIndexL\symX\symImaginary)d\symX.
-\end{equation}
+e^{-2^\symX}
+e^{-2\pi\symImaginary\symIndexL\symX}d\symX.
+\end{multline}
 The variable transformation $\symY = 2^\symX$ yields
@@
 2\int_{0}^{\infty} e^{-\symY}
-\symY^{-\frac{2\pi\symIndexL\symImaginary}{\log(2)}}d\symY
-= 
-2\Gamma\!\left(1-\frac{2\pi\symIndexL\symImaginary}{\log(2)}\right)
+\symY^{-\frac{2\pi\symImaginary\symIndexL}{\log(2)}}d\symY
+= 
+2\Gamma\!\left(1-\frac{2\pi\symImaginary\symIndexL}{\log(2)}\right)
 \end{equation}
@@
 |\symA_\symIndexL|
-\leq \num{9.885e6}.
+\leq \num{9.885e-6}.
 \end{equation}

### 2016-12-03 python/equal-register-probability.py

@@ -1,0 +1,39 @@
+#############################
+# Copyright 2016 Otmar Ertl #
+#############################
+
+from numpy import ones, array
+from math import expm1, cosh, tanh, sinh, log
+import matplotlib.pyplot as plt
+
+numPoints = 100000
+xMin = 0
+xMax = 1
+
+alphaInf = 1/(2.*log(2.))
+
+def upperBound(d):
+    return 1. + 2 * alphaInf * log(1 - d/2 + d*d/16)
+
+def lowerBound(d):
+    return 1. + 2 * alphaInf * log(1 - d/2)
+
+
+iter = range(0,numPoints)
+xValues = list([xMin + i*((xMax - xMin)/(numPoints-1)) for i in iter])
+yValuesUpperBound = list([upperBound(xValues[i]) for i in iter])
+yValuesLowerBound = list([lowerBound(xValues[i]) for i in iter])
+
+plt.rc('text', usetex=True)
+plt.rc('font', family='serif')
+
+fig1, ax1 = plt.subplots(subplot_kw=dict(xlim=[0,1], ylim=[0,1]))
+fig1.set_size_inches(6, 4)
+ax1.plot(xValues, yValuesUpperBound, color='black')
+ax1.plot(xValues, yValuesLowerBound, color='black')
+ax1.fill_between(xValues, yValuesLowerBound, yValuesUpperBound, facecolor='#dddddd', edgecolor='none')
+ax1.grid(b=True, which='major', color='black', linestyle='--')
+ax1.set_xlabel(r"Jaccard distance $D$")
+ax1.set_ylabel(r"probability of equal registers")
+fig1.savefig('../paper/equal-register-probability.svg', format='svg', dpi=600)
+

### 2016-12-04 paper/paper.tex
extended future work section + some other minor improvements

@@
 \symDefine{\symBitRepB}{b}
+\symDefine{\symNumBitsMinwiseHashing}{b}
 \symDefine{\symIndexI}{i}
-\symDefine{\symImaginary}{\rm{i}}
+\symDefine{\symImaginary}{\mathrm{i}}
 \symDefine{\symIndexJ}{j}
@@
 Therefore, if performance matters, the function values can be precalculated for all possible arguments and kept in lookup tables of size $\symNumReg+1$.
+The new estimation algorithm is very elegant, because it does neither contain magic numbers nor special cases as the original algorithm.
 
@@
 \section{Cardinality estimation of set intersections and differences}
+\label{sec:cardinality_estimation_set_intersections}
 
@@
 
-The maximum likelihood method can also be used to estimate similarity measures such as the Jaccard index of two sets that are represented as HyperLogLog sketches.
+The maximum likelihood method can also be used to estimate distance measures such as the Jaccard distance of two sets that are represented as HyperLogLog sketches.
 This directly leads to the question whether the HyperLogLog algorithm could be used for locality-sensitive hashing \cite{Leskovec2014, Wang2014}.
 Various locality-sensitive hashing algorithms have been proposed in the past.
-Among the most popular ones are the SimHash \cite{Charikar2002} and the min-wise hashing \cite{Broder1997, Li2011} algorithms whose hash collision probabilities are a function of the cosine and Jaccard similarities, respectively.
+Among the most popular ones are the SimHash \cite{Charikar2002} and the minwise hashing \cite{Broder1997} algorithms whose hash collision probabilities are a function of the angular and Jaccard distances, respectively.
+A generalization of the latter method is $\symNumBitsMinwiseHashing$-bit minwise hashing that improves memory efficiency by only storing the lowest $\symNumBitsMinwiseHashing$ bits of the minimum hash value \cite{Li2011}.
+The probability that two different sets are mapped to equal hash values $\symRegValVariate_1$ and $\symRegValVariate_2$ is roughly
+\begin{equation}
+\label{equ:minwise_hahsing_probability}
+\symProbability(\symRegValVariate_1=\symRegValVariate_2)
+\approx
+1-
+(1-\textstyle\frac{1}{2 ^\symNumBitsMinwiseHashing})\symDistanceMeasure.
+\end{equation}
 
@@
 \approx
-2\symAlpha_\infty \left(\log(\symY)-\log(\symX)\right)$ we get
+2\symAlpha_\infty \left(\log(\symY)-\log(\symX)\right)$ (compare \eqref{equ:approx_formula} in \cref{app:analysis_xi}) we get
 
@@
 -
-\frac{1}{2}
+\textstyle\frac{1}{2}
 \symDistanceMeasure
 +
-\frac{1}{4}
+\textstyle\frac{1}{4}
 \symDistanceMeasure^2
-\frac{
+\textstyle\frac{
 \symPoissonRate_\symSetASuffix
@@
 -
-\frac{1}{2}
+\textstyle\frac{1}{2}
 \symDistanceMeasure
@@
 -
-\frac{1}{2}
+\textstyle\frac{1}{2}
 \symDistanceMeasure
 +
-\frac{1}{16}
+\textstyle\frac{1}{16}
 \symDistanceMeasure^2
@@
 \end{figure}
-This finding makes the HyperLogLog algorithm an interesting candidate for locality-sensitive hashing.
-What also looks promising is the much simpler preprocessing because it only involves a single hash function (compare \cref{alg:insert}).
-In contrast, min-wise hashing is very costly, because many hash function evaluations are necessary for all the needed permutations \cite{Li2011}.
+This dependency on the Jaccard distance is very similar to that of minwise hashing \eqref{equ:minwise_hahsing_probability}, which makes the
+HyperLogLog algorithm an interesting candidate for locality-sensitive hashing with respect to the Jaccard similarity.
+The memory-efficiency of different hashing approaches can be measured by a storage factor that is the variance of the distance estimator multiplied by the number of bits used for storing the hash signature \cite{Li2011}.
+For a hash algorithm that maps two different sets to the same $\symNumBitsMinwiseHashing$-bit hash value with probability $1-\symA\symDistanceMeasure$ with some constant $\symA\in(0,1]$, the storage factor is given by $\symNumBitsMinwiseHashing\symDistanceMeasure\left(\frac{1}{\symA}-\symDistanceMeasure\right)$.
+Particularly, for the case $D=0.3$, this gives \num{6.72} for conventional minwise hashing with 32-bit hash values ($\symNumBitsMinwiseHashing=32$, $\symA\approx 1$), \num{0.51} for 1-bit minwise hashing ($\symNumBitsMinwiseHashing=1$, $\symA=\frac{1}{2}$), and  \num{1.63} for the HyperLogLog algorithm with 5-bit registers ($\symNumBitsMinwiseHashing=5$, $\symA=
+\symAlpha_\infty$). This means that the HyperLogLog algorithm is not as memory-efficient as 1-bit minwise hashing, but significantly better than 32-bit minwise hashing.
+
+It is worth mentioning that estimating the Jaccard distance by just counting equal registers does not use all the information contained by the HyperLogLog sketch.
+The method described in \cref{sec:cardinality_estimation_set_intersections} would allow a more accurate estimation of the Jaccard distance by using the estimates for intersection and union sizes.
+This could be used for additional more precise filtering when searching for similar items.
+
+Preprocessing is very costly for minwise hashing, because many hash function evaluations are necessary for all the needed permutations \cite{Li2011}.
+In contrast, the HyperLogLog algorithm requires only a single hash function evaluation which reduces preprocessing costs.
+In addition, HyperLogLog sketches can be easily merged, which is not possible for $\symNumBitsMinwiseHashing$-bit minwise hash signatures.
 
@@
 \begin{equation}
+\label{equ:def_xi}
 \symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
@@
 \begin{equation}
-\symPowerSeriesFunc(\symX) = 1 + \Re\!\left(\sum_{\symIndexL=1}^{\infty}
+\symPowerSeriesFunc(\symX) = \frac{\symA_0}{2} + \Re\!\left(\sum_{\symIndexL=1}^{\infty}
 \symA_\symIndexL e^{2\pi\symImaginary\symIndexL\symX}\right)
@@
 \end{equation}
-where $\Gamma$ denotes the gamma function. Using the identity $\left|\Gamma(1+\symImaginary\symX)\right| = 
+where $\Gamma$ denotes the gamma function. Obviously, the constant term in the Fourier series is equal to $1$, because $\symA_0=2$.
+
+In order to investigate the maximum deviation of $\symPowerSeriesFunc(\symX)$ from this value we consider all further coefficients $\symA_\symIndexL$ with $\symIndexL\geq 1$.
+Using the identity $\left|\Gamma(1+\symImaginary\symX)\right| =
 \sqrt{\frac{\pi\symX}{\sinh(\pi\symX)}}
@@
 \begin{equation}
-\max(|\symPowerSeriesFunc(\symX)-1|) \leq \sum_{\symIndexL=1}^{\infty} 
+\num{9.884e-6}
+\leq
+|\symA_1|
+-
+\sum_{\symIndexL=2}^{\infty} 
 |\symA_\symIndexL|
-\leq \num{9.885e-6}.
+\leq
+\max(|\symPowerSeriesFunc(\symX)-1|) 
+\leq 
+|\symA_1|
++
+\sum_{\symIndexL=2}^{\infty} 
+|\symA_\symIndexL|
+\leq 
+\num{9.885e-6}.
+\end{equation}
+
+An interesting approximation formula can be derived from $\symPowerSeriesFunc(\symX) \approx 1$ by integrating on both sides: 
+\begin{equation}
+\label{equ:approx_formula}
+\sum_{\symRegVal = -\infty}^\infty
+e^{-2^{\symRegVal+\symY}} - e^{-2^{\symRegVal+\symX}} \approx \symX - \symY.
 \end{equation}

### 2016-12-05 paper/paper.tex
improved future work section, discussion of one permutation hashing, added more references

@@
 In case the data is not distributed and results do not need to be aggregated further, there are even more efficient estimation algorithms available.
-On the one hand there is the self-learning bitmap \cite{Chen2011}, and on the other hand there is the HyperLogLog algorithm extended by a historic inverse probability estimator that is continuously updated together with the HyperLogLog sketch \cite{Ting2014}.
+On the one hand there is the self-learning bitmap \cite{Chen2011}, and on the other hand there is the HyperLogLog algorithm extended by a historic inverse probability estimator that is continuously updated together with the HyperLogLog sketch \cite{Ting2014, Cohen2014}.
 Both achieve the same estimation error using less space.
@@
 Particularly, for the case $D=0.3$, this gives \num{6.72} for conventional minwise hashing with 32-bit hash values ($\symNumBitsMinwiseHashing=32$, $\symA\approx 1$), \num{0.51} for 1-bit minwise hashing ($\symNumBitsMinwiseHashing=1$, $\symA=\frac{1}{2}$), and  \num{1.63} for the HyperLogLog algorithm with 5-bit registers ($\symNumBitsMinwiseHashing=5$, $\symA=
-\symAlpha_\infty$). This means that the HyperLogLog algorithm is not as memory-efficient as 1-bit minwise hashing, but significantly better than 32-bit minwise hashing.
-
-It is worth mentioning that estimating the Jaccard distance by just counting equal registers does not use all the information contained by the HyperLogLog sketch.
-The method described in \cref{sec:cardinality_estimation_set_intersections} would allow a more accurate estimation of the Jaccard distance by using the estimates for intersection and union sizes.
+\symAlpha_\infty$).
+This means that the HyperLogLog algorithm is not as memory-efficient as 1-bit minwise hashing, but significantly better than 32-bit minwise hashing, when estimating the Jaccard distance by just counting equal registers.
+However, in contrast to 1-bit minwise hashing the HyperLogLog sketch contains additional information which allows estimating the cardinality or merging two HyperLogLog hash signatures.
+Furthermore, the method described in \cref{sec:cardinality_estimation_set_intersections} would allow a more accurate estimation of the Jaccard distance by using the estimates for intersection and union sizes.
 This could be used for additional more precise filtering when searching for similar items.
 
-Preprocessing is very costly for minwise hashing, because many hash function evaluations are necessary for all the needed permutations \cite{Li2011}.
-In contrast, the HyperLogLog algorithm requires only a single hash function evaluation which reduces preprocessing costs.
-In addition, HyperLogLog sketches can be easily merged, which is not possible for $\symNumBitsMinwiseHashing$-bit minwise hash signatures.
+Since HyperLogLog sketches can be effectively constructed, because only a single hash function evaluation is needed for each item, the preprocessing step would be very fast.
+In contrast, preprocessing is very costly for minwise hashing, because of the many required permutations \cite{Li2011}.
+To overcome this problem, one permutation hashing was proposed \cite{Li2012} which keeps for a predefined number of bins the minimum of all values that are mapped to this bin.
+Actually, this is very similar to the HyperLogLog algorithm.
+The bins correspond to HyperLogLog registers which keep the first one-bit position of the minimum value instead of the minimum value itself.
+This close connection was also pointed out in \cite{Cohen2014}, where both methods have been considered as variants of $k$-partion minwise hashing.
+The only difference is that the HyperLogLog algorithm uses base-2 ranks while one permutation hashing uses full ranks.
 
 \section{Conclusion}
-
 We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches based on the Poisson approximation.

### 2016-12-09 paper/paper.tex
some minor improvements

@@
 For performance reasons the number of registers $\symNumReg$ is chosen to be a power of 2, $\symNumReg = 2^\symPrecision$.
-$\symPrecision$ is the precision that directly influences the relative error which scales like $1/\sqrt{\symNumReg}$.
+$\symPrecision$ is the precision parameter that directly controls the relative error which scales like $1/\sqrt{\symNumReg}$.
 All registers start with zero initial value.
@@
 The leading $\symPrecision$ bits of the hash value are used to select one of the $2^\symPrecision$ registers.
-Among the next $\symRegRange$ bits, the position of the first 1-bit is determined which is a value in the range $[1,\symRegRange+1]$.
+Among the next following $\symRegRange$ bits, the position of the first 1-bit is determined which is a value in the range $[1,\symRegRange+1]$.
 The value $\symRegRange+1$ is used, if all $\symRegRange$ bits are zeros.
@@
 \State $\langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision,\symBitRepB_1,\ldots,\symBitRepB_\symRegRange\rangle_2 \gets$ $(\symPrecision + \symRegRange)$-bit hash value of $\symDataItem$
+\comm{0.2}{$\symBitRepA_\symIndexI,\symBitRepB_\symIndexI\in\{0,1\}$}
 \State $\symIndexI \gets 1+ \langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision\rangle_2$
@@
 A HyperLogLog sketch can be characterized by a parameter pair $(\symPrecision, \symRegRange)$.
-The first parameter $\symPrecision$ is the precision and controls the relative error while the second defines the possible value range of a register.
+The precision parameter $\symPrecision$ controls the relative error while the second parameter defines the possible value range of a register.
 A register can take all values starting from 0 to $\symRegRange+1$, inclusively.
@@
 
-\cref{alg:insert} has some properties which are especially useful in distributed environments.
+\cref{alg:insert} has some properties which are especially useful for distributed data streams.
 First, the insertion order of elements has no influence on the final HyperLogLog sketch.
 Furthermore, any two HyperLogLog sketches with same parameters $(\symPrecision, \symRegRange)$ representing sets $\symSetA$ and $\symSetB$ can be easily merged.
-The resulting register values are simply given by the register-wise maximum values
+The HyperLogLog sketch that represents the union of both sets is simply constructed by taking the register-wise maximum values
 \begin{equation}
@@
 \end{equation}
-The final HyperLogLog sketch represents the union of $\symSetA$ and $\symSetB$.
-
-At any time a $(\symPrecision, \symRegRange)$-HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be compressed into a $(\symPrecision', \symRegRange')$-HyperLogLog data structure, if $\symPrecision'\leq\symPrecision$ and $\symPrecision'+\symRegRange' \leq\symPrecision + \symRegRange$ is satisfied (see \cref{alg:compress}).
+
+At any time a $(\symPrecision, \symRegRange)$-HyperLogLog sketch can be compressed into a $(\symPrecision', \symRegRange')$-HyperLogLog data structure, if $\symPrecision'\leq\symPrecision$ and $\symPrecision'+\symRegRange' \leq\symPrecision + \symRegRange$ is satisfied (see \cref{alg:compress}).
 This transformation is lossless in a sense that the resulting HyperLogLog sketch is the same as if all elements would have been recorded by a $(\symPrecision', \symRegRange')$-HyperLogLog sketch right from the beginning.
@@
 \begin{algorithm}
-\caption{Compression of a $(\symPrecision, \symRegRange)$-HyperLogLog sketch with register values $\boldsymbol{\symRegValVariate}$ into a $(\symPrecision', \symRegRange')$-HyperLogLog sketch with $\symPrecision'\leq\symPrecision$ and $\symPrecision'+\symRegRange' \leq\symPrecision + \symRegRange$.}
+\caption{Compression of a $(\symPrecision, \symRegRange)$-HyperLogLog sketch with register values $\boldsymbol{\symRegValVariate}=(\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ into a $(\symPrecision', \symRegRange')$-HyperLogLog sketch with $\symPrecision'\leq\symPrecision$ and $\symPrecision'+\symRegRange' \leq\symPrecision + \symRegRange$.}
 \label{alg:compress}
@@
 1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symCardinality & \symCardinality\geq 1 \wedge \symRegVal = \symRegRange +1
-\end{cases}.
+\end{cases}
 \end{equation}
@@
 \end{equation}
-Here $\alpha_\symNumReg$ is a bias correction factor which was derived for a given number of registers $\symNumReg$ to be \cite{Flajolet2007}
+Here $\alpha_\symNumReg$ is a bias correction factor which was derived for a given number of registers $\symNumReg$ as \cite{Flajolet2007}
 \begin{equation}
@@
 \end{equation}
-Therefore, the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ can be seen as the result after applying this transformation to a sample $\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg$ of the distribution described by \eqref{equ:assumed_register_val_distribution}.
-If all registers values of the HyperLogLog sketch fall into the range $[1,\symRegRange]$, they must  be identical to the values $\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg$.
+Therefore, register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ can be seen as the result after applying this transformation to a sample $\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg$ of the distribution described by \eqref{equ:assumed_register_val_distribution}.
+If all registers values fall into the range $[1,\symRegRange]$, they must  be identical to the values $\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg$.
 In other words, the observed register values are also a plausible sample of the assumed distribution described by \eqref{equ:assumed_register_val_distribution} in this case.
@@
 This explains why the raw estimator works best for intermediate cardinalities.
-However, for small and large cardinalities many registers have values equal to 0 or $\symRegRange+1$, respectively, which cannot be described by \eqref{equ:assumed_register_val_distribution} and finally leads to the observed bias.
+However, for small and large cardinalities many register values are equal to 0 or $\symRegRange+1$, respectively, which cannot be described by \eqref{equ:assumed_register_val_distribution} and finally leads to the observed bias.
 
 \subsection{Corrections to the raw estimator}
-If we knew the values $\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg$ for which the transformation \eqref{equ:dist_transformation} led to the observed register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$, we would be able to estimate $\symPoissonRate$ using
+If we knew the values $\symRegValVariate'_1,\ldots,\symRegValVariate'_\symNumReg$ for which transformation \eqref{equ:dist_transformation} led to the observed register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$, we would be able to estimate $\symPoissonRate$ using
 \begin{equation}
@@
 \symExpectation(\symCountVariate'_\symRegVal)
+=
+\symNumReg
+\left(
+\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
+-
+\symProbability(\symRegValVariate \leq \symRegVal-1\vert\symPoissonRate)
+\right)
 =
@@
 Assuming a uniform hash function, element hash values can be mocked by random numbers.
-For the following results we used the Mersenne Twister random number generator with \num{19937} bit state size from the C++ standard library.
+For the following results we used the Mersenne Twister random number generator with a state size  of \num{19937} bits from the C++ standard library.
 
@@
 For example, at a cardinality value around \num{5000} the average number of required cycles until the stop criterion is met increases abruptly from two to three.
-More than three iteration cycles have never been necessary for any cardinality estimate in this performance test.
+More than three iteration cycles have never been observed for any cardinality estimate in this performance test.
 
@@
 
-While the union of two sets that are represented by HyperLogLog sketches can be straightforwardly computed \eqref{equ:hll_union}, the computation of cardinalities of other set operations like intersections and differences is more challenging.
+While the union of two sets that are represented by HyperLogLog sketches can be straightforwardly computed using \eqref{equ:hll_union}, the computation of cardinalities of other set operations like intersections and differences is more challenging.
 The conventional approach uses the relationships
@@
 \section{Future work}
-As described in \cref{sec:depoissonization} an unbiased estimator for the rate in the Poisson model, is also unbiased for the estimation of the cardinality.
+As described in \cref{sec:depoissonization} an unbiased estimator for the rate in the Poisson model, is also unbiased estimator for the cardinality.
 We have shown that this approach works well for the corrected raw estimator as well as for the maximum likelihood estimator even though both are only approximately unbiased.
@@
 2 e^\symB
-}}.
-\end{equation}
-Consequently, $|\symA_{\symIndexL}| \leq 
+}}
+\end{equation}
+which leads to $|\symA_{\symIndexL}| \leq 
 |\symA_{2}|
@@
 2 e^\symB
-}}\right)^{\symIndexL-2}$ for $\symIndexL\geq 2$ and 
+}}\right)^{\symIndexL-2}$ for $\symIndexL\geq 2$ and further to
 \begin{equation}
@@
 \end{equation}
-For the maximum deviation from 1 we finally have
+As a consequence, the maximum deviation of $\symPowerSeriesFunc(\symX)$ from 1 is bounded by
 \begin{equation}

### 2016-12-11 paper/paper.tex
minor changes

@@
 \left(1-\frac{1}{2^\symRegVal}\right)^\symCardinality - \left(1-\frac{1}{2^{\symRegVal-1}}\right)^\symCardinality & \symCardinality \geq 1 \wedge 1\leq\symRegVal\leq\symRegRange\\
-1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symCardinality & \symCardinality\geq 1 \wedge \symRegVal = \symRegRange +1
+1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symCardinality & \symCardinality\geq 1 \wedge \symRegVal = \symRegRange +1.
 \end{cases}
@@
 e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}} & 0\leq \symRegVal \leq \symRegRange \\
-1 & \symRegVal > \symRegRange
-\end{cases}.
+1 & \symRegVal > \symRegRange.
+\end{cases}
 \end{equation}
@@
 
-Let $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ be a sample taken from \eqref{equ:assumed_register_val_distribution}.
+Let $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ be a sample distributed according to \eqref{equ:assumed_register_val_distribution}.
 For large sample sizes $\symNumReg\rightarrow \infty$ we have asymptotically
@@
 As the mean shows, the error is unbiased over the entire cardinality range.
-The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is about an order of magnitude larger than the operating range upper bound of the raw estimator (\cref{fig:raw_estimate}) or the original method (\cref{fig:original_estimate}).
+The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is about an order of magnitude larger than the operating range upper bound of the raw estimator (\cref{fig:raw_estimate}).
 
@@
 \includesvg[width=1\textwidth]{stdev_comparison}
-\caption{Standard deviations of the relative error of different cardinality estimators for a HyperLogLog sketch with paramters $\symPrecision = 12$ and $\symRegRange=20$ over the true cardinality.}
+\caption{Standard deviations of the relative error of different cardinality estimators over the true cardinality for a HyperLogLog sketch with paramters $\symPrecision = 12$ and $\symRegRange=20$ .}
 \label{fig:stdev_comparison}
@@
 \end{equation}
-if $\symXEstimate$ is the root of the function
+if $\symXEstimate$ denotes the root of the function
 \begin{equation}
@@
 The function is non-positive at 0 since $\symFunc(0)=\symCountVariate_0-\symNumReg\leq 0$ and, in case $\symCountVariate_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing.
-$\symCountVariate_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $\symRegRange+1$, for which the maximum likelihood estimate would be infinite.
+$\symCountVariate_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $\symRegRange+1$, for which the maximum likelihood estimate would be positive infinite.
 
@@
 If the HyperLogLog sketch is in the intermediate range, where $\symCountVariate_0=\symCountVariate_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimator \eqref{equ:raw_estimator}.
-Hence, the maximum likelihood method leads directly to the harmonic mean that is used by the raw estimator.
+Hence, consequent application of the maximum likelihood method leads directly to the harmonic mean that is used by the raw estimator.
 
@@
 \left(\symX_{\symIndexI-1}-\symX_{\symIndexI-2}\right)
-\frac{\symFunc(\symX_{\symIndexI-1})}{\symFunc(\symX_{\symIndexI-1}) - \symFunc(\symX_{\symIndexI-2})}
+\frac{\symFunc(\symX_{\symIndexI-1})}{\symFunc(\symX_{\symIndexI-1}) - \symFunc(\symX_{\symIndexI-2})}.
 \end{equation}
@@
 \Delta\symX_{\symIndexI} = \Delta\symX_{\symIndexI-1}
-\frac{\left(\symNumReg-\symCountVariate_0\right)-\symFuncPrime(\symX_{\symIndexI-1})}{\symFuncPrime(\symX_{\symIndexI-1}) - \symFuncPrime(\symX_{\symIndexI-2})}\\
-\symX_{\symIndexI} = \symX_{\symIndexI-1} + \Delta\symX_{\symIndexI}
+\frac{\left(\symNumReg-\symCountVariate_0\right)-\symFuncPrime(\symX_{\symIndexI-1})}{\symFuncPrime(\symX_{\symIndexI-1}) - \symFuncPrime(\symX_{\symIndexI-2})},
+\\
+\symX_{\symIndexI} = \symX_{\symIndexI-1} + \Delta\symX_{\symIndexI}.
 \end{gather}
@@
 \symX' - \symX''/3 + \left(\symX'' \symX''\right)\left(1/45-\symX''/472.5\right)
-+ \symBigO(\symX^{8})
++ \symBigO(\symX^{8}).
 \end{equation}
@@
 
-Putting all these optimizations together finally gives the new cardinality estimation algorithm as presented in \cref{alg:estimate_ml}.
+Putting all these optimizations together finally gives the new cardinality estimation algorithm presented in \cref{alg:estimate_ml}.
 The algorithm requires mainly only elementary operations.
@@
 &
-\symRegRange < \symRegVal_1 \wedge \symRegRange < \symRegVal_2
+\symRegRange < \symRegVal_1 \wedge \symRegRange < \symRegVal_2.
 \end{cases}
@@
 &
-\symRegVal_1 = \symRegVal_2 = \symRegRange + 1
+\symRegVal_1 = \symRegVal_2 = \symRegRange + 1.
 \end{cases}

### 2016-12-14 paper/paper.tex
some minor changes, removed params.json

@@
 
-Using the identity $\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = \symAlpha_\infty\symPowerSeriesFunc(\log_2(\log(1/\symX)))/\log(1/\symX)$, we get for the linear counting case with $\symRegRange=0$
+We can cross-check the new estimator for the linear counting case with $\symRegRange=0$.
+Using the identity $\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = \symAlpha_\infty\symPowerSeriesFunc(\log_2(\log(1/\symX)))/\log(1/\symX)$, we get
 \begin{equation}
@@
 \frac{\symNumReg\log\!\left(\symNumReg/\symCountVariate_0\right)}{\symPowerSeriesFunc\!\left(\log_2\!\left(\log\!\left(\symNumReg/\symCountVariate_0\right)\right)\right)}
-\approx
-\symNumReg\log\!\left(\symNumReg/\symCountVariate_0\right),
-\end{equation}
-which is almost identical to the linear counting estimator used for the small range correction \eqref{equ:linear_counting_estimator}.
-Here we used again the fact that the function $\symPowerSeriesFunc$ can be well approximated by 1 (see \cref{sec:derivation_raw_estimator}).
+\end{equation}
+which is as expected almost identical to the linear counting estimator \eqref{equ:linear_counting_estimator}, because $\symPowerSeriesFunc(\symX)\approx 1$ (see \cref{sec:derivation_raw_estimator}).
 \subsection{Corrected raw estimation algorithm}

### 2016-12-18 paper/paper.tex
get rid of svg package, replaced png figures by pdf figures
some minor improvements

@@
 \usepackage{graphicx}
-\usepackage{svg}
 \usepackage{algpseudocode,algorithm,algorithmicx}
@@
 \centering
-\includesvg[width=1\textwidth]{raw_estimate}
+\includegraphics[width=1\textwidth]{raw_estimate}
 \caption{The distribution of the relative estimation error over the cardinality for the raw estimator after evaluation of \num{10000} randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
@@
 \centering
-\includesvg[width=1\textwidth]{small_range_estimate}
+\includegraphics[width=1\textwidth]{small_range_estimate}
 \caption{The distribution of the relative estimation error for the linear counting estimator after evaluation of \num{10000} randomly generated bitmaps of size $2^{12}$ which correspond to HyperLogLog sketches with parameters $\symPrecision = 12$ and $\symRegRange=0$.}
@@
 \centering
-\includesvg[width=1\textwidth]{original_estimate}
+\includegraphics[width=1\textwidth]{original_estimate}
 \caption{The distribution of the relative estimation error over the cardinality for the original estimation algorithm after evaluation of \num{10000} randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
@@
 \centering
-\includesvg[width=0.6\textwidth]{power-series-function-minus-1}
+\includegraphics[width=0.6\textwidth]{power-series-function-minus-1}
 \caption{The deviation of $\symPowerSeriesFunc(\symX)$ from 1.}
@@
 \centering
-\includesvg[width=1\textwidth]{raw_corrected_estimate_12_20}
+\includegraphics[width=1\textwidth]{raw_corrected_estimate_12_20}
 \caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
@@
 \centering
-\includesvg[width=1\textwidth]{stdev_comparison}
+\includegraphics[width=1\textwidth]{stdev_comparison}
 \caption{Standard deviations of the relative error of different cardinality estimators over the true cardinality for a HyperLogLog sketch with paramters $\symPrecision = 12$ and $\symRegRange=20$ .}
@@
 \centering
-\includesvg[width=1\textwidth]{raw_corrected_estimate_8_24}
+\includegraphics[width=1\textwidth]{raw_corrected_estimate_8_24}
 \caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
@@
 \centering
-\includesvg[width=1\textwidth]{raw_corrected_estimate_16_16}
+\includegraphics[width=1\textwidth]{raw_corrected_estimate_16_16}
 \caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
@@
 \centering
-\includesvg[width=1\textwidth]{raw_corrected_estimate_22_10}
+\includegraphics[width=1\textwidth]{raw_corrected_estimate_22_10}
 \caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
@@
 \centering
-\includesvg[width=1\textwidth]{raw_corrected_estimate_12_52}
+\includegraphics[width=1\textwidth]{raw_corrected_estimate_12_52}
 \caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
@@
 \centering
-\includesvg[width=1\textwidth]{raw_corrected_estimate_12_14}
+\includegraphics[width=1\textwidth]{raw_corrected_estimate_12_14}
 \caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
@@
 \centering
-\includesvg[width=1\textwidth]{corrected_raw_avg_exec_time}
+\includegraphics[width=1\textwidth]{corrected_raw_avg_exec_time}
 \caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at \SI{3.3}{\giga\hertz} when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.
@@
 
-We remark that we are not the first applying the maximum likelihood method to HyperLogLog sketches \cite{Clifford2012}, where the case without stochastic averaging (compare \cref{sec:data_element_insertion}) was considered.
-The register values are statistically independent by nature in this case which
-allows factorization of the joint probability mass function and allows a straightforward application of the maximum likelihood method.
-
-The more relevant case with stochastic averaging, which is considered in this paper, is more complicated.
-However, with the help of the Poisson approximation, it is still feasible to apply the maximum likelihood method and we are able to derive a new robust and efficient cardinality estimation algorithm.
-Furthermore, we will demonstrate that consequent application of the maximum likelihood method reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values.
+For HyperLogLog sketches that have been obtained without stochastic averaging (see \cref{sec:data_element_insertion}) the maximum likelihood method was already previously considered for cardinality estimation \cite{Clifford2012}.
+However, in this case the register values are statistically independent by nature which
+allows factorization of the joint probability mass function and a straightforward calculation of the maximum likelihood estimate.
+The practically more relevant stochastic averaging case, which is considered in this paper, gives a more complicated likelihood function.
+However, the Poisson approximation makes the maximum likelihood method feasible again and we are finally able to derive another new robust and efficient cardinality estimation algorithm.
+Furthermore, in the course of the derivation we will demonstrate that consequent application of the maximum likelihood method reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values.
 The history of the HyperLogLog algorithm shows that the raw estimator \eqref{equ:raw_estimator} was first found after several attempts using the geometric mean \cite{Flajolet2007, Durand2003}.
@@
 \centering
-\includesvg[width=0.6\textwidth]{helper}
+\includegraphics[width=0.6\textwidth]{helper}
 \caption{The function $\symHelper(\symX)$.}
@@
 Since the expected statistical errror for the HyperLogLog data structure scales according to $\frac{1}{\sqrt{\symNumReg}}$ \cite{Flajolet2007}, it makes sense to choose $\symStopDelta = \frac{\symStopEpsilon}{\sqrt{\symNumReg}}$ with some constant $\symStopEpsilon$.
-For all results that will be presented in \cref{sec:maximum_likelihood_estimation_error} we have used $\symStopEpsilon = 10^{-2}$.
+For all results presented later in \cref{sec:maximum_likelihood_estimation_error} we have used $\symStopEpsilon = 10^{-2}$.
 
@@
 
-Putting all these optimizations together finally gives the new cardinality estimation algorithm presented in \cref{alg:estimate_ml}.
+Putting all these optimizations together finally gives the new cardinality estimation algorithm presented as \cref{alg:estimate_ml}.
 The algorithm requires mainly only elementary operations.
@@
 \centering
-\includesvg[width=1\textwidth]{max_likelihood_estimate_12_20}
+\includegraphics[width=1\textwidth]{max_likelihood_estimate_12_20}
 \caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
@@
 \centering
-\includesvg[width=1\textwidth]{max_likelihood_estimate_8_24}
+\includegraphics[width=1\textwidth]{max_likelihood_estimate_8_24}
 \caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
@@
 \centering
-\includesvg[width=1\textwidth]{max_likelihood_estimate_16_16}
+\includegraphics[width=1\textwidth]{max_likelihood_estimate_16_16}
 \caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
@@
 \centering
-\includesvg[width=1\textwidth]{max_likelihood_estimate_22_10}
+\includegraphics[width=1\textwidth]{max_likelihood_estimate_22_10}
 \caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
@@
 \centering
-\includesvg[width=1\textwidth]{max_likelihood_estimate_12_52}
+\includegraphics[width=1\textwidth]{max_likelihood_estimate_12_52}
 \caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
@@
 \centering
-\includesvg[width=1\textwidth]{max_likelihood_estimate_12_14}
+\includegraphics[width=1\textwidth]{max_likelihood_estimate_12_14}
 \caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
@@
 \centering
-\includesvg[width=1\textwidth]{max_likelihood_avg_exec_time}
+\includegraphics[width=1\textwidth]{max_likelihood_avg_exec_time}
 \caption{Average execution time of the maximum likelihood estimation algorithm as a function of the true cardinality with an Intel Core i5-2500K clocking at \SI{3.3}{\giga\hertz} for HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
@@
 \centering
-\includesvg[width=0.6\textwidth]{equal-register-probability}
+\includegraphics[width=0.6\textwidth]{equal-register-probability}
 \caption{The approximate probability range of equal register values as a function of the Jaccard distance.}
@@
 \right|
-}.
-\end{equation}
-The triangle inequality leads to
+}
+\end{equation}
+and the triangle inequality leads to
 \begin{equation}
@@
 \right|
-}
+}.
 \end{equation}

### 2016-12-20 paper/paper.tex
added task for generation of arXiv upload package, made cleveref options compatible with arXiv processing
minor changes

@@
 \usepackage{multirow}
-\usepackage[capitalise,nameinlink,noabbrev]{cleveref}
+\usepackage[capitalise]{cleveref}
 
@@
 \subsection{Maximum likelihood estimation algorithm}
-In order to get a fast cardinality estimation algorithm, it is crucial to minimize the evaluation costs for \eqref{equ:funcprime}.
-A couple of optimizations allow significant reduction of the computational effort:
+In order to get a fast cardinality estimation algorithm, it is crucial to minimize evaluation costs for \eqref{equ:funcprime}.
+A couple of optimizations allow significant reduction of computational effort:
 \begin{itemize}
@@
 \centering
-\caption{List of true cardinalities of the pairwise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetS_1 = \symSetA\cup\symSetX$ and  $\symSetS_2 = \symSetB\cup\symSetX$ was investigated.
+\captionabove{List of true cardinalities of the pairwise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetS_1 = \symSetA\cup\symSetX$ and  $\symSetS_2 = \symSetB\cup\symSetX$ was investigated.
 The last column shows the average number of iteration cycles until the stop criterion of the BFGS algorithm was satisfied.
@@
 \centering
-\caption{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetA|$ using the conventional approach and the maximum likelihood method, respectively.
+\captionabove{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetA|$ using the conventional approach and the maximum likelihood method, respectively.
 The root-mean-square error improvement factor of the maximum likelihood method is shown in the last column.}
@@
 \centering
-\caption{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetB|$ using the conventional approach and the maximum likelihood method, respectively.
+\captionabove{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetB|$ using the conventional approach and the maximum likelihood method, respectively.
 The root-mean-square error improvement factor of the maximum likelihood method is shown in the last column.}
@@
 \centering
-\caption{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetX|$ using the conventional approach and the maximum likelihood method, respectively.
+\captionabove{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetX|$ using the conventional approach and the maximum likelihood method, respectively.
 The root-mean-square error improvement factor of the maximum likelihood method is shown in the last column.}
@@
 \centering
-\caption{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetA\cup\symSetB\cup\symSetX|$ using the conventional approach and the maximum likelihood method, respectively.
+\captionabove{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetA\cup\symSetB\cup\symSetX|$ using the conventional approach and the maximum likelihood method, respectively.
 The root-mean-square error improvement factor of the maximum likelihood method is shown in the last column.}
@@
 
-%\begin{table}
-%\centering
-%\caption{TODO}
-%\label{tbl:joint_estimation_jaccard}
-%\csvreader[before reading=\scriptsize, tabular=r|rrr|rrr|r, table head=
-% & 
-%\multicolumn{3}{c|}{conventional approach}
-%&
-%\multicolumn{3}{c|}{maximum likelihood method}
-%&
-%\multicolumn{1}{c}{improvement}
-%\\
-%\# & 
-%\multicolumn{1}{c}{mean} & 
-%\multicolumn{1}{c}{stdev} & 
-%\multicolumn{1}{c|}{rmse} & 
-%\multicolumn{1}{c}{mean} & 
-%\multicolumn{1}{c}{stdev} & 
-%\multicolumn{1}{c|}{rmse} &
-%\multicolumn{1}{c}{rmse ratio}
-%\\\hline]
-%{joint_cardinality_calculation.csv}{
-%inclExclMeanJaccardIdx=\inclExclMeanJaccardIdx,
-%inclExclStdDevJaccardIdx=\inclExclStdDevJaccardIdx,
-%inclExclRMSEJaccardIdx=\inclExclRMSEJaccardIdx,
-%maxLikeMeanJaccardIdx=\maxLikeMeanJaccardIdx,
-%maxLikeStdDevJaccardIdx=\maxLikeStdDevJaccardIdx,
-%maxLikeRMSEJaccardIdx=\maxLikeRMSEJaccardIdx,
-%improvementRmseJaccard=\improvementRmseJaccard
-%}{
-%\thecsvrow & 
-%\numformat{\inclExclMeanJaccardIdx} & 
-%\numformat{\inclExclStdDevJaccardIdx} & 
-%\numformat{\inclExclRMSEJaccardIdx} & 
-%\numformat{\maxLikeMeanJaccardIdx} & 
-%\numformat{\maxLikeStdDevJaccardIdx} & 
-%\numformat{\maxLikeRMSEJaccardIdx} &
-%\numformattwo{\improvementRmseJaccard}
-%}
-%\end{table}
-
 \section{Future work}
@@
 \label{app:analysis_xi}
-The Fourier series of the function $\symPowerSeriesFunc(\symX)$, which is defined by
+The Fourier series of the periodic function
 \begin{equation}
@@
 2^{\symRegVal+\symX}
-e^{-2^{\symRegVal+\symX}}
-\end{equation}
-and which has period equal to 1, is
+e^{-2^{\symRegVal+\symX}},
+\end{equation}
+which has a period equal to 1, is
 \begin{equation}
@@
 \approx
-\num{9.884e-6}$. Obviously, the maximum deviation of $\symPowerSeriesFunc(\symX)-1$ from the first harmonic must be smaller than $\sum_{\symIndexL=2}^{\infty} 
+\num{9.884e-6}$. Clearly, the maximum deviation of $\symPowerSeriesFunc(\symX)-1$ from the first harmonic must be smaller than $\sum_{\symIndexL=2}^{\infty} 
 |\symA_\symIndexL|$. The ratio of subsequent coefficients is given by
@@
 \leq
-\max(|\symPowerSeriesFunc(\symX)-1|) 
+\max_{\symX}(|\symPowerSeriesFunc(\symX)-1|) 
 \leq 
@@
 \tilde{\symHelper}_2 = 
-\frac{\symX+\tilde{\symHelper}_1(1-\tilde{\symHelper}_1)}{\symX+(1-\tilde{\symHelper}_1)}
+\frac{\symX+\tilde{\symHelper}_1(1-\tilde{\symHelper}_1)}{\symX+(1-\tilde{\symHelper}_1)}.
 \end{equation}
@@
 =
-\symNumReg-\symCountVariate_0
+\symNumReg-\symCountVariate_0.
 \end{equation}
@@
 \end{equation}
-the solution of the equation will be off by a relative error $\symError_\symX$:
+the solution of the equation will be off by some relative error $\symError_\symX$:
 \begin{equation}
@@
 \begin{multline}
-\symXEstimate\left(1+\symError_\symX\right)\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
-\left(
-1+\symAlpha
-\right)
-\left(
+\symXEstimate\left(1+\symError_\symX\right)\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
++
+\left(1+\symAlpha\right)
 \sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal
@@
 +
+\\
++
+\left(1+\symAlpha\right)
 \symCountVariate_{\symRegRange+1}
@@
 \frac{\symXEstimate\left(1+\symError_\symX\right)}{2^\symRegRange}\right)
-\right)
-=
-\symNumReg-\symCountVariate_0
+=
+\symNumReg-\symCountVariate_0.
 \end{multline}
@@
 =
-\symNumReg-\symCountVariate_0
+\symNumReg-\symCountVariate_0.
 \end{multline}
@@
 \right)
-}
+}.
 \end{equation}
@@
 \sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
-}
+}.
 \end{equation}
@@
 }
-\right)
-\end{equation}
-Hence, as long as most registers are not in the saturated state ($\symCountVariate_{\symRegRange+1}\ll\symNumReg$), the relative error $\symError_\symX$ of the calculated estimate using the approximation $\symHelperApprox(\symX)$ for $\symHelper(\symX)$ has the same order of magnitude as $\symError_\symHelper$.
+\right).
+\end{equation}
+Hence, as long as most registers are not saturated ($\symCountVariate_{\symRegRange+1}\ll\symNumReg$), the relative error $\symError_\symX$ of the calculated estimate using the approximation $\symHelperApprox(\symX)$ for $\symHelper(\symX)$ has the same order of magnitude as $\symError_\symHelper$.
 

### 2016-12-21 paper/paper.tex
better description of starting point selection for secant method

@@
 The square root that appears for the calculation of the stopping limit $\symStopDelta=\symStopEpsilon/\sqrt{\symNumReg}$ can be precomputed, because it only depends on the HyperLogLog precision parameter $\symPrecision$.
-For large cardinalities it is worth to take the strong \eqref{equ:strong_lower_bound} instead of the weak lower bound \eqref{equ:weak_lower_bound} as initial value, which requires a single logarithm evaluation.
+
+For very large cardinalities it makes sense to use the strong \eqref{equ:strong_lower_bound} instead of the weak lower bound \eqref{equ:weak_lower_bound} as second starting point for the secant method.
 The stronger bound is a much better approximation especially for large cardinalities, where the extra logarithm evaluation is amortized by savings in the number of iteration cycles.
+Therefore, the presented algorithm switches over to the stronger bound, if
+\begin{equation}
+\frac{
+\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
+}{
+\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+}
+> 1.5
+\end{equation}
+is satisfied. The threshold value of $1.5$ was found to be a reasonable choice in order to reduce the computation time for large cardinalities significantly.
 

### 2016-12-21 paper/paper.tex
renamed "corrected raw estimator" -> "improved raw estimator"

@@
 \end{equation}
-which we call the corrected raw estimator.
+which we call the improved raw estimator.
 Here $\symNumReg\,\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$ and $2\symNumReg\,\symLargeCorrectionFunc(1-\symCountVariate_{\symRegRange+1}/\symNumReg)$ are replacements for  $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$ in the raw estimator \eqref{equ:raw_estimator}, respectively.
@@
 which is as expected almost identical to the linear counting estimator \eqref{equ:linear_counting_estimator}, because $\symPowerSeriesFunc(\symX)\approx 1$ (see \cref{sec:derivation_raw_estimator}).
-\subsection{Corrected raw estimation algorithm}
-The corrected raw estimator \eqref{equ:correctedestimator} leads to a new cardinality estimation algorithm for HyperLogLog sketches.
+\subsection{Improved raw estimation algorithm}
+The improved raw estimator \eqref{equ:correctedestimator} leads to a new cardinality estimation algorithm for HyperLogLog sketches.
 \cref{alg:corrected_raw_estimation} demonstrates the numerical evaluation of the estimator.
@@
 \begin{algorithm}
-\caption{Cardinality estimation algorithm based on the corrected raw estimator.}
+\caption{Cardinality estimation algorithm based on the improved raw estimator.}
 \label{alg:corrected_raw_estimation}
@@
 \includegraphics[width=1\textwidth]{raw_corrected_estimate_12_20}
-\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
+\caption{Relative error of the improved raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
 \label{fig:raw_corrected_estimation_error_12_20}
@@
 
-The corrected raw estimator beats the precision of methods that apply bias correction on the raw estimator \cite{Heule2013,Rhodes2015,Sanfilippo2014}.
+The improved raw estimator beats the precision of methods that apply bias correction on the raw estimator \cite{Heule2013,Rhodes2015,Sanfilippo2014}.
 Based on the simulated data we have empirically determined the bias correction function $\symBiasCorrectionFunc$ for the raw estimator that satisfies $\symCardinality = \symExpectation(\symBiasCorrectionFunc(\symCardinalityRawEstimate)\vert\symCardinality)$ for all cardinalities.
@@
 the estimator $\symCardinalityRawEstimate':=\symBiasCorrectionFunc(\symCardinalityRawEstimate)$ is unbiased and a function of the raw estimator.
-Its standard deviation can be compared with that of the corrected raw estimator in \cref{fig:stdev_comparison}.
+Its standard deviation can be compared with that of the improved raw estimator in \cref{fig:stdev_comparison}.
 For cardinalities smaller than \num{10000} the empirical bias correction approach is not very precise.
@@
 In practice, the standard deviation is even larger, because the choice between both estimators must be made based on an estimate and not on the true cardinality, for which the intersection point of both curves represents the ideal transition point.
-In contrast, the corrected raw estimator performs well over the entire cardinality range.
+In contrast, the improved raw estimator performs well over the entire cardinality range.
 
@@
 \includegraphics[width=1\textwidth]{raw_corrected_estimate_8_24}
-\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
+\caption{Relative error of the improved raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
 \label{fig:raw_corrected_estimation_error_8_24}
@@
 \includegraphics[width=1\textwidth]{raw_corrected_estimate_16_16}
-\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
+\caption{Relative error of the improved raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
 \label{fig:raw_corrected_estimation_error_16_16}
@@
 \includegraphics[width=1\textwidth]{raw_corrected_estimate_22_10}
-\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
+\caption{Relative error of the improved raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
 \label{fig:raw_corrected_estimation_error_22_10}
@@
 \includegraphics[width=1\textwidth]{raw_corrected_estimate_12_52}
-\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
+\caption{Relative error of the improved raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
 \label{fig:raw_corrected_estimation_error_12_52}
@@
 \includegraphics[width=1\textwidth]{raw_corrected_estimate_12_14}
-\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
+\caption{Relative error of the improved raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
 \label{fig:raw_corrected_estimation_error_12_14}
@@
 \label{sec:corrected_raw_estimation_algorithm}
-In order to evaluate the performance of the corrected raw estimation algorithm, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch.
+In order to evaluate the performance of the improved raw estimation algorithm, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch.
 For different cardinalities we loaded precalculated multiplicity vectors of \num{1000} randomly generated HyperLogLog sketches into main memory.
@@
 \label{sec:maximum_likelihood_estimation_error}
-In order to investigate the estimation error for the maximum likelihood estimation algorithm, we investigated the same HyperLogLog configurations as in \cref{sec:corrected_raw_estimation_error} for the corrected raw estimation algorithm.
+In order to investigate the estimation error for the maximum likelihood estimation algorithm, we investigated the same HyperLogLog configurations as in \cref{sec:corrected_raw_estimation_error} for the improved raw estimation algorithm.
 \cref{fig:max_likelihood_estimation_error_12_20}, \cref{fig:max_likelihood_estimation_error_8_24},  \cref{fig:max_likelihood_estimation_error_16_16}, \cref{fig:max_likelihood_estimation_error_22_10}, \cref{fig:max_likelihood_estimation_error_12_52}, and \cref{fig:max_likelihood_estimation_error_12_14} show very similar results for various HyperLogLog parameters.
@@
 What is different for the maximum likelihood estimation approach is a somewhat smaller median bias with less oscillations around zero for small cardinalities.
-The standard deviation of the relative error is also slightly better for the maximum likelihood estimator than for the corrected raw estimator as shown in \cref{fig:stdev_comparison}.
+The standard deviation of the relative error is also slightly better for the maximum likelihood estimator than for the improved raw estimator as shown in \cref{fig:stdev_comparison}.
 Furthermore, contrary to the raw estimator which reveals a small oscillating bias for the mean (see \cref{fig:raw_corrected_estimation_error_22_10}), the maximum likelihood estimator seems to be completely unbiased (see \cref{fig:max_likelihood_estimation_error_22_10}).
@@
 The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:avg_exec_time}.
-The average computation time for the maximum likelihood algorithm shows a different behavior than for the corrected raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}).As can be seen, the average execution time is larger for most cardinalities, but nevertheless, it never exceeds \SI{700}{\nano\second} which is still fast enough for many applications.
+The average computation time for the maximum likelihood algorithm shows a different behavior than for the improved raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}).As can be seen, the average execution time is larger for most cardinalities, but nevertheless, it never exceeds \SI{700}{\nano\second} which is still fast enough for many applications.
 The steps in the chart can be explained by different numbers of iteration cycles until the secant method is stopped.
@@
 As described in \cref{sec:depoissonization} an unbiased estimator for the rate in the Poisson model, is also unbiased estimator for the cardinality.
-We have shown that this approach works well for the corrected raw estimator as well as for the maximum likelihood estimator even though both are only approximately unbiased.
+We have shown that this approach works well for the improved raw estimator as well as for the maximum likelihood estimator even though both are only approximately unbiased.
 Therefore, it would be interesting what conditions on an approximately unbiased Poisson rate estimator are sufficient to guarantee approximate unbiasedness, if used as cardinality estimator.
@@
 Unfortunately, joint cardinality estimation is much more expensive than for a single HyperLogLog sketch, because it requires maximization of a multi-dimensional function.
-Since we have found the corrected raw estimator which is almost as accurate as the maximum likelihood estimator for the single HyperLogLog case, we could imagine that there also exists a faster algorithm for the two HyperLogLog case.
+Since we have found the improved raw estimator which is almost as accurate as the maximum likelihood estimator for the single HyperLogLog case, we could imagine that there also exists a faster algorithm for the two HyperLogLog case.
 It is expected that such a new algorithm makes use of all the information given by the sufficient statistic \eqref{equ:sufficient_joint_statistic}.

### 2016-12-21 python/comparison.py

@@ -22,11 +22,11 @@ def readData(filename):
         return numpy.array([[float(e) for e in r] for r in reader][1:])
 
 config = [
-["12_20_max_likelihood_estimates.dat", "blue", False, "maximum likelihood estimator"],
-#["12_20_flajolet_mid_range_estimates.dat", "gray", False, "raw estimator"],
-["12_20_flajolet_mid_range_estimates.dat", "red", True, "subsequently bias corrected raw estimator"],
 ["12_20_flajolet_small_range_estimates.dat", "black", False, "linear counting estimator"],
-["12_20_corrected_raw_estimates.dat", "green", False, "corrected raw estimator"]
+["12_20_flajolet_mid_range_estimates.dat", "red", True, "bias corrected raw estimator"],
+#["12_20_flajolet_mid_range_estimates.dat", "gray", False, "raw estimator"],
+["12_20_corrected_raw_estimates.dat", "green", False, "improved raw estimator"],
+["12_20_max_likelihood_estimates.dat", "blue", False, "maximum likelihood estimator"]
 ]
 
 cardinalities = readData('cardinalities.dat').flatten()
@@ -89,9 +89,7 @@ plt.setp(axins.get_xticklabels(), visible=False)
 mark_inset(ax1, axins, loc1=1, loc2=3, fc="none", ec="0.5")
 
 fig1.savefig('../paper/stdev_comparison.svg', format='svg', dpi=600)
-fig1.savefig('../paper/stdev_comparison.png', format='png', dpi=100)
 plt.close(fig1)
 fig2.savefig('../paper/mean_comparison.svg', format='svg', dpi=600)
-fig2.savefig('../paper/mean_comparison.png', format='png', dpi=100)
 plt.close(fig2)
 

### 2016-12-26 paper/paper.tex
minor improvements

@@
 The original cardinality estimator \cite{Flajolet2007} is based on the idea that the number of distinct element insertions a register needs to reach the value $\symRegVal$ is proportional to $\symNumReg  2^{\symRegVal}$.
-Given that, a rough cardinality estimate can be obtained by calculating the average over the values $\lbrace \symNumReg 2^{\symRegValVariate_1},\ldots,\symNumReg2^{\symRegValVariate_\symNumReg}\rbrace$.
+Given that, a rough cardinality estimate can be obtained by averaging the values $\lbrace \symNumReg 2^{\symRegValVariate_1},\ldots,\symNumReg2^{\symRegValVariate_\symNumReg}\rbrace$.
 
@@
 
-\section{Cardinality estimation of set intersections and differences}
+\section{Cardinality estimation of set intersections and complements}
 \label{sec:cardinality_estimation_set_intersections}
 
-While the union of two sets that are represented by HyperLogLog sketches can be straightforwardly computed using \eqref{equ:hll_union}, the computation of cardinalities of other set operations like intersections and differences is more challenging.
+While the union of two sets that are represented by HyperLogLog sketches can be straightforwardly computed using \eqref{equ:hll_union}, the computation of cardinalities of other set operations like intersections and complements is more challenging.
 The conventional approach uses the relationships
@@
 Unfortunately, the estimation error does not scale well for the inclusion-exclusion principle.
-For small intersection sizes the relative estimation error can become very large \cite{Dasgupta2015}.
+Especially for small Jaccard indices, the relative estimation error can become very large \cite{Dasgupta2015}.
 In the worst case, the estimate could be negative without artificial restriction to nonnegative values.
-Therefore, it was proposed to combine HyperLogLog sketches with MinHash sketches \cite{Pascoe2013, Cohen2016}, which improves the estimation error, even though at the expense of significant more space consumption.
+Therefore, it was proposed to combine HyperLogLog sketches with minwise hashing \cite{Pascoe2013, Cohen2016}, which improves the estimation error, even though at the expense of significant more space consumption.
 
 It was recently pointed out without special focus on HyperLogLog sketches, that the application of the maximum likelihood method to the joint likelihood function of two probabilistic data structures, which represent the intersection operands, gives better intersection size estimates \cite{Ting2016}.
-For HyperLogLog sketches recorded without stochastic averaging (compare \cref{sec:data_element_insertion}) this was also shown in \cite{Cohen2016}, where also an algorithm based on the maximum likelihood principle was outlined.
+For HyperLogLog sketches recorded without stochastic averaging (compare \cref{sec:data_element_insertion}) this was shown in \cite{Cohen2016}, where also an algorithm based on the maximum likelihood principle was outlined.
 However, in practice, HyperLogLog sketches are recorded using stochastic averaging because of the much cheaper element insertions as described by \cref{alg:insert}.
@@
 The goal is to find estimates for the cardinalities of the pairwise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$.
-The Poisson approximation allows us to assume that the elements of $\symSetA$ are inserted into the HyperLogLog sketch representing $\symSetS_1$ at rate $\symPoissonRate_\symSetASuffix$.
-Similarly, the elements of $\symSetB$ are inserted into the HyperLogLog sketch representing $\symSetS_2$ at rate $\symPoissonRate_\symSetBSuffix$.
-Furthermore, we assume that elements of $\symSetX$ are inserted into both HyperLogLog sketches simultaneously at rate $\symPoissonRate_\symSetXSuffix$.
+The Poisson approximation allows us to assume that pairwise distinct elements are inserted into the HyperLogLog sketches representing $\symSetS_1$ and $\symSetS_2$ at rates $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$, respectively.
+Furthermore, we assume that further unique elements are inserted into both HyperLogLog sketches simultaneously at rate $\symPoissonRate_\symSetXSuffix$.
 We expect that good estimates $\symPoissonRateEstimate_\symSetASuffix$, $\symPoissonRateEstimate_\symSetBSuffix$, and $\symPoissonRateEstimate_\symSetXSuffix$ for the rates are also good estimates for the cardinalities of $\symSetA$, $\symSetB$, and $\symSetX$.
@@
 \eqref{equ:register_value_distribution}.
-If we take into account that elements are added to $\symSetA$, $\symSetB$, and $\symSetX$ at rates $\symPoissonRate_\symSetXSuffix$, $\symPoissonRate_\symSetASuffix$, and $\symPoissonRate_\symSetBSuffix$, respectively, the probability that a certain register has a value less than or equal to $\symRegVal_1$ in the first HyperLogLog sketch and simultaneously a value less than or equal to $\symRegVal_2$ in the second one can be written as
+If we take into account that pairwise distinct elements are added to $\symSetA$, $\symSetB$, and $\symSetX$ at rates $\symPoissonRate_\symSetXSuffix$, $\symPoissonRate_\symSetASuffix$, and $\symPoissonRate_\symSetBSuffix$, respectively, the probability that a certain register has a value less than or equal to $\symRegVal_1$ in the first HyperLogLog sketch and simultaneously a value less than or equal to $\symRegVal_2$ in the second one can be written as
 \begin{equation}
@@
 \sum_{\symIndexI = 1}^{\symNumReg}
-\log(\symProbabilityMass(\symRegValVariate_{1\symIndexI},\symRegValVariate_{2\symIndexI}))
+\log(\symProbabilityMass(\symRegValVariate_{1\symIndexI},\symRegValVariate_{2\symIndexI})).
 \end{equation}
@@
 For each case \num{3333} different HyperLogLog sketch pairs were constructed and passed to the joint cardinality estimation algorithm.
-\cref{tbl:joint_estimation_cases} also shows the average number of function \eqref{equ:max_likelihood_phi_func} and gradient \eqref{equ:df_dphi}  evaluations, which the BFGS algorithm needed until convergence occured.
+\cref{tbl:joint_estimation_cases} also shows the average number of iterations of the BFGS algorithm until the stop criterion was satisfied.
+Each iteration step involved a function \eqref{equ:max_likelihood_phi_func} and a gradient \eqref{equ:df_dphi} evaluation.
 
@@
 The conventional approach merges both HyperLogLog sketches using \eqref{equ:hll_union} and estimates the union size using single sketch cardinality estimation.
-The joint cardinality estimation algorithm opens another way.
+The joint cardinality estimation algorithm provides another opportunity.
 The union can also be estimated by simply summing up the three estimates for the sets $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$.
@@
 However, further research is necessary to determine, if this is feasible from a practical point of view.
-As for the inclusion-exclusion principle, the effort would scale exponentially with the number of involved HyperLogLog sketches.
+As for the inclusion-exclusion principle, the effort would scale at least exponentially with the number of involved HyperLogLog sketches.
 
@@
 \end{equation}
-As shown in \cref{fig:equal_register_probability} the bounds are very close, especially for small Jaccard distances where the probability can be well approximated by
+As shown in \cref{fig:equal_register_probability} the bounds are very close, especially for small Jaccard distances, where the probability can be well approximated by
 \begin{equation}

### 2016-12-27 paper/paper.tex
added outline to introduction section

@@
 \let\thefootnote\relax\footnotetext{This paper together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}.
-The first version of this paper was published on April 17, 2016.}
+A first version of this paper was published on April 17, 2016.}
 \begin{abstract}
@@
 
-%In contrast, HyperLogLog sketches (without historic inverse probability estimator) built from different sets of elements can be merged and the result is identical to a HyperLogLog sketch obtained by inserting all elements of the union set.
-Therefore, the HyperLogLog sketch inherently supports cardinality estimation of unions.
-However, there are also applications which require the cardinality estimation of intersections.
+\subsection{Outline}
+In \cref{sec:hyperloglog_data_structure} we give an introduction to HyperLogLog sketches and the corresponding update algorithm.
+We also discuss how the resulting state can be statistically described and approximated by a Poisson model.
+
+In \cref{sec:cardinality_estimation} we describe the original cardinality estimation algorithm, its shortcomings at low and large cardinalities, and previous approaches to overcome them.
+We present a simple derivation of the original raw estimator and analyze the root cause for its limited operating range.
+We derive an improved version of the raw estimator, which leads to a new fast cardinality estimation algorithm that works over the full cardinality range as demonstrated by extensive simulations.
+
+In \cref{sec:max_likelihood_estimation} we derive a second even more precise cardinality estimation algorithm based on the maximum likelihood principle which is again verified by simulations.
+As presented in \cref{sec:cardinality_estimation_set_intersections} the same approach can be generalized to two HyperLogLog sketches, which allows result cardinality estimation of set operations like intersections or complements.
+The simulation results show that the estimates are significantly better than those obtained by the inclusion-exclusion principle.
+
+Finally, in \cref{sec:future_work} we discuss open problems and ideas for future work including HyperLogLog's potential use as locality-sensitve hashing algorithm before we conclude in \cref{sec:conclusion}.
 
 \section{HyperLogLog data structure}
+\label{sec:hyperloglog_data_structure}
 The HyperLogLog algorithm uses a sketching data structure that consists of $\symNumReg$ registers.
@@
 \section{Maximum likelihood estimation}
+\label{sec:max_likelihood_estimation}
 We know from \cref{sec:depoissonization} that any unbiased estimator for the Poisson parameter is also an unbiased estimator for the cardinality.
@@
 \section{Future work}
+\label{sec:future_work}
 As described in \cref{sec:depoissonization} an unbiased estimator for the rate in the Poisson model, is also unbiased estimator for the cardinality.
@@
 \section{Conclusion}
+\label{sec:conclusion}
 We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches based on the Poisson approximation.

### 2016-12-28 paper/paper.tex
some minor improvements

@@
 \begin{algorithmic}
-\Function {Compress}{\symDataItem, $\boldsymbol{\symRegValVariate}$}
+\Function {Compress}{$\boldsymbol{\symRegValVariate}$}
+\comm{.4}{$\boldsymbol{\symRegValVariate}\in\lbrace 0,1,\ldots,\symRegRange+1\rbrace^\symNumReg$, $\symNumReg = 2^\symPrecision$}
 \State allocate registers for new HyperLogLog sketch $\boldsymbol{\symRegValVariate}' = (\symRegValVariate_1,\ldots,\symRegValVariate_{2^{\symPrecision'}})$
@@
 \State \Return $\boldsymbol{\symRegValVariate}'$
+\comm{.4}{$\boldsymbol{\symRegValVariate}'\in\lbrace 0,1,\ldots,\symRegRange'+1\rbrace^{\symNumReg'}$, $\symNumReg' = 2^{\symPrecision'}$}
 \EndFunction
@@
 \Function {EstimateCardinality}{$\boldsymbol{\symRegValVariate}$}
-\State $\symNumReg \gets \dim \boldsymbol{\symRegValVariate}$
+\comm{.4}{$\boldsymbol{\symRegValVariate}\in\lbrace 0,1,\ldots,\symRegRange+1\rbrace^\symNumReg$, $\symNumReg = 2^\symPrecision$}
 \State $\symCardinalityRawEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexI=1}^\symNumReg 2^{-\symRegValVariate_\symIndexI}\right)^{\!-1}$
@@
 The values of functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ can be either calculated on-demand or pre-calculated.
-The possible value range of $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$ is $[0, \symNumReg]$.
+Possible values of $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$ are $\{0,1,\ldots, \symNumReg\}$.
 Therefore, if performance matters, the function values can be precalculated for all possible arguments and kept in lookup tables of size $\symNumReg+1$.
@@
 \begin{algorithm}
-\caption{Cardinality estimation algorithm based on the improved raw estimator.}
+\caption{Cardinality estimation based on the improved raw estimator.}
 \label{alg:corrected_raw_estimation}
@@
 \Function {EstimateCardinality}{$\boldsymbol{\symCountVariate}$}
-\State $\symNumReg \gets \left\Vert\boldsymbol{\symCountVariate}\right\Vert_1$
+\comm{.45}{$\dim \boldsymbol{\symCountVariate} = \symRegRange + 2$, $\left\Vert\boldsymbol{\symCountVariate}\right\Vert_1 = \symNumReg$}
 \State $\symZ\gets\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$
-\comm{.5}{alternatively, take $\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$ from precalculated lookup table}
+\comm{.45}{alternatively, take $\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$ from precalculated lookup table}
 \For{$\symRegVal\gets \symRegRange, 1$}
@@
 \State $\symZ\gets\symZ+\symNumReg\cdot\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$
-\comm{.5}{alternatively, take $\symNumReg\cdot\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$ from precalculated lookup table}
+\comm{.45}{alternatively, take $\symNumReg\cdot\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$ from precalculated lookup table}
 \State \Return$\symAlpha_\infty\symNumReg^2 / \symZ$
-\comm{.5}{$\symAlpha_\infty := 1/(2\log(2))$}
+\comm{.45}{$\symAlpha_\infty := 1/(2\log(2))$}
 \EndFunction
@@
 \Function {$\symSmallCorrectionFunc$}{$\symX$}
-\comm{.5}{$\symX\in[0,1]$}
+\comm{.45}{$\symX\in[0,1]$}
 \If{$\symX=1$}
@@
 \Function {$\symLargeCorrectionFunc$}{$\symX$}
-\comm{0.5}{$\symX\in[0,1]$}
+\comm{0.45}{$\symX\in[0,1]$}
 \State $\symY\gets1$
@@
 \begin{algorithm}
-\caption{Maximum likelihood cardinality estimation.}
+\caption{Cardinality estimation based on the maximum likelihood principle.}
 \ContinuedFloat
@@
 \Function {EstimateCardinality}{$\boldsymbol{\symCountVariate}$}
-\State $\symRegRange\gets\dim(\boldsymbol{\symCountVariate})-2$
+\comm{.35}{$\dim \boldsymbol{\symCountVariate} = \symRegRange + 2$, $\left\Vert\boldsymbol{\symCountVariate}\right\Vert_1 = \symNumReg$}
 \State $\symRegValVariate_\text{min} \gets \min\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$
@@
 \State $\symZ \gets 0$
-\State $\symNumReg'\gets \symCountVariate_{\symRegRange+1}$
 \State $\symY \gets 2^{-\symRegValVariate'_\text{max}}$
@@
 \State $\symY \gets 2\symY$
-\State $\symNumReg'\gets\symNumReg' + \symCountVariate_\symRegVal$
 \EndFor
@@
 \comm{0.35}{here $\symZ=\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$}
-\State $\symNumReg \gets \symNumReg' + \symCountVariate_0$ 
 \State $\symCount \gets \symCountVariate_{\symRegRange+1}$
@@
 \comm{0.35}{$\symB = \sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}$}
+\State $\symNumReg' \gets \symNumReg - \symCountVariate_0$
 \If{$\symB \leq 1.5\cdot\symA$}
@@
 \Else
-\State $\symX \gets \symNumReg'/b \cdot \log(1+\symB/\symA)$ 
+\State $\symX \gets \symNumReg'/\symB \cdot \log(1+\symB/\symA)$ 
 \comm{0.35}{strong lower bound \eqref{equ:strong_lower_bound}}
@@
 \begin{algorithm}
-\caption{Maximum likelihood cardinality estimation (continued).}
+\caption{Cardinality estimation based on the maximum likelihood principle (continued).}
 \begin{algorithmic}

### 2016-12-28 c++/benchmark.cpp

@@ -108,7 +108,7 @@ void run(const int p, const int q, const string& resultsFileName) {
     ofstream outputFile(resultsFileName);
 
     MaxLikelihoodEstimator maxLikelihoodEstimator(p,q);
-    CorrectedRawEstimator correctedRawEstimator(p,q);
+    ImprovedRawEstimator improvedRawEstimator(p,q);
 
     for(long cardinality : cardinalities) {
 
@@ -189,7 +189,7 @@ void run(const int p, const int q, const string& resultsFileName) {
 
             auto start = std::chrono::system_clock::now();
             for(auto c : counts) {
-                sum += correctedRawEstimator.estimate_on_demand(c, smallCorrectionIterations, largeCorrectionIterations);
+                sum += improvedRawEstimator.estimate_on_demand(c, smallCorrectionIterations, largeCorrectionIterations);
                 flajoletCorrectedSmallCorrectionIterationsSum += smallCorrectionIterations;
                 flajoletCorrectedLargeCorrectionIterationsSum += largeCorrectionIterations;
             }
@@ -201,7 +201,7 @@ void run(const int p, const int q, const string& resultsFileName) {
         {
             auto start = std::chrono::system_clock::now();
             for(auto c : counts) {
-                sum += correctedRawEstimator.estimate_precalculated(c);
+                sum += improvedRawEstimator.estimate_precalculated(c);
             }
             auto end = chrono::system_clock::now();
             auto elapsed = chrono::duration_cast<chrono::nanoseconds>(end - start);

### 2016-12-28 c++/cardinality_calculator.cpp

@@ -160,7 +160,7 @@ int main(int argc, char* argv[])
         vector<int> kMaxSum(cardinalities.size());
 
         const MaxLikelihoodEstimator maxLikelihoodEstimator(p,q);
-        const CorrectedRawEstimator correctedRawEstimator(p,q);
+        const ImprovedRawEstimator improvedRawEstimator(p,q);
 
         #pragma omp parallel for
         for(size_t seedCounter = 0; seedCounter < seeds.size(); ++seedCounter) {
@@ -186,7 +186,7 @@ int main(int argc, char* argv[])
 
                 flajoletSmallRangeEstimates[resultPos] = flajoletSmallRangeEstimate(c);
                 flajoletMidRangeEstimates[resultPos] = flajoletRawEstimate(c);
-                flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate_on_demand(c, numSmallCorrectionIterations[resultPos], numLargeCorrectionIterations[resultPos]);
+                flajoletMidRangeEstimatesCorrected[resultPos] = improvedRawEstimator.estimate_on_demand(c, numSmallCorrectionIterations[resultPos], numLargeCorrectionIterations[resultPos]);
                 //flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate_precalculated(c);
                 flajoletEstimates[resultPos] = flajoletEstimate(c);
                 maxLikelihoodEstimates[resultPos] = maxLikelihoodEstimator.estimate(c, outerLoopIterationsCount[resultPos], innerLoop1IterationsCount[resultPos], innerLoop2IterationsCount[resultPos], logEvaluationCount[resultPos], kMin, kMax);

### 2016-12-28 c++/cardinality_estimation.hpp

@@ -102,12 +102,10 @@ public:
         if (kMin > q) return std::numeric_limits<double>::infinity();
 
         double z = 0.;
-        int mPrime = c[q+1];
         double y = ldexp(1., -kMaxPrime);
         for (int k = kMaxPrime; k >= kMinPrime; --k) {
             z += y*c[k];
             y += y;
-            mPrime += c[k];
         }
 
         int cPrime = c[q+1];
@@ -119,6 +117,7 @@ public:
 
         double x;
         double a = z + c[0];
+        int mPrime = m - c[0];
         {
             double b = z + ldexp(c[q+1], -q);
             if (b <= 1.5*a) {
@@ -215,7 +214,7 @@ double flajoletRawEstimate(const std::vector<int>& c, int m) {
     return (alpha*m)*m/z;
 }
 
-class CorrectedRawEstimator {
+class ImprovedRawEstimator {
     const int p;
     const int q;
     const int m = 1 << p;
@@ -274,7 +273,7 @@ class CorrectedRawEstimator {
 
 public:
 
-    CorrectedRawEstimator(const int p_, const int q_) : p(p_), q(q_) {}
+    ImprovedRawEstimator(const int p_, const int q_) : p(p_), q(q_) {}
 
     double estimate_on_demand(const std::vector<int>& c, int& numSmallCorrectionIterations, int& numLargeCorrectionIterations) const {
 

### 2016-12-30 paper/paper.tex
mention monotonicity of improved raw estimator

@@
 Therefore, if performance matters, the function values can be precalculated for all possible arguments and kept in lookup tables of size $\symNumReg+1$.
+
 The new estimation algorithm is very elegant, because it does neither contain magic numbers nor special cases as the original algorithm.
+Moreover, the algorithm guarantees monotonicity of the cardinality estimate.
+The estimate will never become smaller when adding new elements.
+Although this sounds self-evident, previous approaches that combine estimators for different cardinality ranges or make use of empirical collected data have problems to satisfy this property throughout all cardinalities.
 

### 2017-01-01 paper/paper.tex
added algorithm for multiplicity vector extraction

@@
 \begin{algorithmic}
-\Procedure {InsertElement}{\symDataItem, $\boldsymbol{\symRegValVariate}$}
+\Procedure {InsertElement}{\symDataItem}
 \State $\langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision,\symBitRepB_1,\ldots,\symBitRepB_\symRegRange\rangle_2 \gets$ $(\symPrecision + \symRegRange)$-bit hash value of $\symDataItem$
@@
 The two HyperLogLog parameters can be obtained by $\symPrecision = \log_2 \left\Vert\boldsymbol{\symCountVariate}\right\Vert_1$ and $\symRegRange=\dim\boldsymbol{\symCountVariate}-2$, respectively.
+As \cref{alg:sufficient_statistic} shows the calculation of the multiplicity vector $\boldsymbol{\symCountVariate}$ for a given HyperLogLog sketch with register values $\boldsymbol{\symRegValVariate}$ is straightforward.
+
+\begin{algorithm}
+\caption{Multiplicity vector extraction from a $(\symPrecision, \symRegRange)$-HyperLogLog sketch with $\symNumReg=2^\symPrecision$ registers having values $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$.}
+\label{alg:sufficient_statistic}
+\begin{algorithmic}
+\Function {ExtractMultiplicities}{$\boldsymbol{\symRegValVariate}$}
+\comm{.4}{$\boldsymbol{\symRegValVariate}\in\lbrace 0,1,\ldots,\symRegRange+1\rbrace^\symNumReg$}
+\State $\boldsymbol{\symCountVariate}\gets \left(0,0,\ldots, 0\right)$
+\comm{.4}{$\boldsymbol{\symCountVariate} = (\symCountVariate_0,\ldots,\symCountVariate_{\symRegRange+1})$}
+\For{$\symIndexI\gets 1, \symNumReg$}
+\State $\symCountVariate_{\symRegValVariate_\symIndexI} \gets \symCountVariate_{\symRegValVariate_\symIndexI} + 1$
+\EndFor
+\State \Return $\boldsymbol{\symCountVariate}$
+\EndFunction
+\end{algorithmic}
+\end{algorithm}
 
@@
 \begin{algorithm}
-\caption{Cardinality estimation based on the improved raw estimator.}
+\caption{Cardinality estimation based on the improved raw estimator.
+The input is the multiplicity vector $\boldsymbol{\symCountVariate} = (\symCountVariate_0,\ldots,\symCountVariate_{\symRegRange+1})$ as obtained by \cref{alg:sufficient_statistic}.}
 \label{alg:corrected_raw_estimation}
@@
 \begin{algorithm}
-\caption{Cardinality estimation based on the maximum likelihood principle.}
+\caption{Cardinality estimation based on the maximum likelihood principle.
+The input is the multiplicity vector $\boldsymbol{\symCountVariate} = (\symCountVariate_0,\ldots,\symCountVariate_{\symRegRange+1})$ as obtained by \cref{alg:sufficient_statistic}.}
 \ContinuedFloat
@@
 \begin{algorithm}
-\caption{Cardinality estimation based on the maximum likelihood principle (continued).}
+\caption{(continued)}
 \begin{algorithmic}

### 2017-01-01 paper/paper.tex
changed format of special function names

@@
 \item The sum $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}$ in \eqref{equ:funcprime} can be precalculated and reused for all function evaluations.
-\item Many programming languages allow the efficient multiplication and division by any integral power of two using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
+\item Many programming languages allow the efficient multiplication and division by any integral power of two using special functions, such as $\operatorname{ldexp}$ in C/C++ or $\operatorname{scalb}$ in Java.
 \item The function $\symHelper(\symX)$ only needs to be evaluated at values $\left\lbrace\frac{\symX}{2^{\symRegValVariate'_\text{max}}},\frac{\symX}{2^{\symRegValVariate'_\text{max}-1}},\ldots,\frac{\symX}{2^{\symRegValVariate_\text{min}}}\right\rbrace$ where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$.
@@
 Since $\symX_{\ast\symRegVal}$ is nonnegative, we have $\symY_{\ast\symRegVal},\symZ_{\ast\symRegVal}\in[0,1]$.
-In order to reduce the numerical error of $\symZ_{\ast\symRegVal}$ for small $\symX_{\ast\symRegVal}$, it is essential to use the function $\texttt{expm1}(\symX):=e^\symX-1$ that is available in most programming languages.
+In order to reduce the numerical error of $\symZ_{\ast\symRegVal}$ for small $\symX_{\ast\symRegVal}$, it is essential to use the function $\operatorname{expm1}(\symX):=e^\symX-1$ that is available in most programming languages.
 If $\symX_{\ast\symRegVal}$ is smaller than $\log(2)$, we calculate $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$ via $\symZ_{\ast\symRegVal} =
--\texttt{expm1}(-\symX_{\ast\symRegVal})$ and $\symY_{\ast\symRegVal} = 1 - \symZ_{\ast\symRegVal}$ and not as defined in \eqref{equ:definition_xyz}.
+-\operatorname{expm1}(-\symX_{\ast\symRegVal})$ and $\symY_{\ast\symRegVal} = 1 - \symZ_{\ast\symRegVal}$ and not as defined in \eqref{equ:definition_xyz}.
 In this way the numerical error of both, $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$, is minimized and still only a single exponential function needs to be evaluated.

### 2017-01-04 paper/paper.tex
refactored joint cardinality estimation code, added alternative implementation using ceres

@@
 \begin{algorithmic}
-\Function {ExtractMultiplicities}{$\boldsymbol{\symRegValVariate}$}
+\Function {ExtractCounts}{$\boldsymbol{\symRegValVariate}$}
 \comm{.4}{$\boldsymbol{\symRegValVariate}\in\lbrace 0,1,\ldots,\symRegRange+1\rbrace^\symNumReg$}
@@
 trueRatio=\trueRatio,
-avgNumEvaluations=\avgNumEvaluations
+avgNumFunctionEvaluations=\avgNumFunctionEvaluations
 }{
@@
 \numformattwo{\trueRatio} &
-\numformattwo{\avgNumEvaluations}
+\numformattwo{\avgNumFunctionEvaluations}
 }

### 2017-01-04 c++/cardinality_estimation.hpp

@@ -14,7 +14,13 @@
 #include <cstdlib>
 #include <functional>
 
+#ifdef CARDINALITY_ESTIMATION_USE_CERES
+#include "ceres/ceres.h"
+#endif
+
+#ifdef CARDINALITY_ESTIMATION_USE_DLIB
 #include "dlib/optimization.h"
+#endif
 
 #include "two_hyperloglog_statistic.hpp"
 
@@ -29,50 +35,6 @@ int getPFromNumberRegisters(int m) {
 int getPFromCounts(const std::vector<int>& c) {
     int m = std::accumulate(c.begin(), c.end(), 0);
     return getPFromNumberRegisters(m);
-}
-
-
-double estimateCardinalityFromCdf(const std::vector<double> cdf, int m) {
-    int q = cdf.size() - 1;
-
-    if (cdf[0] >= 1.) std::numeric_limits<double>::infinity();
-
-    double lowerTail = 0;
-    {
-        double arg = cdf[0];
-        double pow2 = 1;
-        while(true) {
-            const double lowerTailBak = lowerTail;
-            lowerTail += pow2 * arg;
-            if (lowerTail == lowerTailBak) break;
-            pow2 += pow2;
-            arg *= arg;
-        }
-    }
-
-    double upperTail = 0;
-    {
-        double arg = std::sqrt(cdf[cdf.size()-1]);
-        double pow2 = 0.5;
-        while(true) {
-            const double upperTailBak = upperTail;
-            upperTail += pow2 * arg;
-            if (upperTail == upperTailBak) break;
-            pow2 *= 0.5;
-            arg = std::sqrt(arg);
-        }
-    }
-
-    double sum = upperTail;
-    for (int k = q; k >= 1; --k) {
-        sum += cdf[k];
-        sum *= 0.5;
-    }
-
-    sum += lowerTail;
-
-    return (m/std::log(2))/sum;
-
 }
 
 class MaxLikelihoodEstimator {
@@ -420,22 +382,17 @@ void inclusionExclusionTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& j
 }
 
 
-
-class LogLikelihoodFunctionForDlib {
+class JointLogLikelihoodFunction {
 
     const TwoHyperLogLogStatistic& jointStatistic;
-    int numEvaluations = 0;
-    double fa;
-    double fb;
-    double fx;
-    bool status = false;
     const int q;
+    std::vector<double> pow2k;
[…]

### 2017-01-04 c++/joint_cardinality_calculation.cpp

@@ -2,6 +2,8 @@
 //# Copyright 2016 Otmar Ertl #
 //#############################
 
+#define CARDINALITY_ESTIMATION_USE_DLIB
+
 #include "cardinality_estimation.hpp"
 #include "two_hyperloglog_statistic.hpp"
 #include "hyperloglog.hpp"
@@ -65,32 +67,40 @@ void printHeaders(ofstream& os, string prefix, string postfix) {
     os << prefix << "RMSE" << postfix << ",";
 }
 
-void calculateAndPrintStatistics(ofstream& os, std::vector<double> data, double& stdev, double& rmse) {
+void calculateAndPrintStatistics(ofstream& os, double trueValue, std::vector<double> data, double& relativeStdev, double& relativeRmse) {
 
     accumulator_set<double, stats<tag::min, tag::max, tag::sum_kahan> > acc;
-    accumulator_set<double, stats<tag::sum_kahan> > acc2;
-
-    acc = std::for_each(data.begin(), data.end(), acc);
 
     for (auto d : data) {
         acc(d);
-        acc2(d*d);
     }
 
-    os << extract_result<tag::min>(acc) << ",";
-    os << extract_result<tag::max>(acc) << ",";
-    const double mom0 = data.size();
-    const double mom1 = extract_result<tag::sum_kahan>(acc);
-    const double mean = mom1/mom0;
-    const double mom2 = extract_result<tag::sum_kahan>(acc2);
-    double t = mom0*mom2 - mom1 * mom1;
-    cout << mom0*mom2 << " " << mom1*mom1 << endl;
-    assert(t >= 0);
-    stdev = std::sqrt(mom0*mom2 - mom1 * mom1) / mom0;
-    rmse = std::sqrt(mom2/mom0);
-    os << mean << ",";
-    os << stdev << ",";
-    os << rmse << ",";
+    const double mean = extract_result<tag::sum_kahan>(acc)/data.size();
+    const double min = extract_result<tag::min>(acc);
+    const double max = extract_result<tag::max>(acc);
+
+    accumulator_set<double, stats<tag::sum_kahan> > acc_rmse;
+    accumulator_set<double, stats<tag::sum_kahan> > acc_stdev;
+
+    for (auto d : data) {
+        acc_rmse(std::pow(d - trueValue, 2));
+        acc_stdev(std::pow(d - mean, 2));
+    }
+
+    const double rmse = std::sqrt(extract_result<tag::sum_kahan>(acc_rmse)/data.size());
[…]

### 2017-01-06 paper/paper.tex
some minor improvements

@@
 \end{equation}
-as estimator for $\symRegVal \geq \symRegRange+1$, respectively.
-
-Inserting all these estimators into \eqref{equ:raw_estimate_with_multiplicities} as replacements for $\symCountVariate'_\symRegVal$ finally gives
+as estimator for $\symRegVal \geq \symRegRange+1$, respectively. This choice of estimators also conserves the mass of zero-valued and saturated registers, because  
+\eqref{equ:multiplicity_transformation} is satisfied,
+if $\symCountVariate'_\symRegVal$ is replaced by $\hat{\symCount}'_\symRegVal$.
+
+Inserting all the registers into \eqref{equ:raw_estimate_with_multiplicities} as replacements for $\symCountVariate'_\symRegVal$ finally gives
 \begin{equation}
@@
 In order to evaluate the performance of the improved raw estimation algorithm, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch.
-For different cardinalities we loaded precalculated multiplicity vectors of \num{1000} randomly generated HyperLogLog sketches into main memory.
+For different cardinalities we took \num{1000} randomly generated HyperLogLog sketches, calculated their
+multiplicity vectors using \cref{alg:sufficient_statistic} and kept all of them in main memory.
 The average computation time was determined by cycling over these multiplicity vectors and passing them as input to the algorithm.
@@
 However, we can imagine that a better numerical approximation of $\symLargeCorrectionFunc$ can be found than that given in \cref{alg:corrected_raw_estimation} and which allows on-demand evaluation without significant extra costs.
-
-The numbers do not yet include the required processing time to extract the multiplicity vector out from the HyperLogLog sketch, which requires a complete scan over all registers and counting the different register values into an array.
+If $2^{\symPrecision+\symRegRange}$ is greater than the maximum expected cardinality value, the number of saturated registers is negligible ($\symCountVariate_{\symRegRange+1}\ll \symNumReg$) and the calculation of $\symLargeCorrectionFunc$ can be omitted at all, because $\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg) \approx 0$ is always true.
+
+The numbers presented in \cref{fig:corrected_raw_avg_exec_time} do not include the processing time to extract the multiplicity vector out of the HyperLogLog sketch, which requires a complete scan over all registers and counting the different register values into an array (compare \cref{alg:sufficient_statistic}).
 A theoretical lower bound for this processing time can be derived using the maximum memory bandwidth of the CPU, which is \SI[per-mode=symbol]{21}{\giga\byte\per\second} for an Intel Core i5-2500K.
@@
 \end{equation}
-These $5\symNumReg$ values are sufficient for estimating $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$.
-Actually, the set of these constants can be further reduced, because $\symCountVariate^{>}_{10}=\symCountVariate^{>}_{20}=
-\symCountVariate^{<}_{1,\symRegRange+1}=\symCountVariate^{<}_{2,\symRegRange+1}=0$
-always holds.
+These $5\symNumReg$ values represent a sufficient statistic for estimating $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$.
+Actually, the number of values could be further reduced, because of the invariants $\symCountVariate^{>}_{10}=\symCountVariate^{>}_{20}=
+\symCountVariate^{<}_{1,\symRegRange+1}=\symCountVariate^{<}_{2,\symRegRange+1}=0$, $\sum_{\symRegVal=0}^{\symRegRange+1}{\symCountVariate^{<}_{1\symRegVal}} = \sum_{\symRegVal=0}^{\symRegRange+1}{\symCountVariate^{>}_{2\symRegVal}}$,
+$\sum_{\symRegVal=0}^{\symRegRange+1}{\symCountVariate^{<}_{2\symRegVal}} = \sum_{\symRegVal=0}^{\symRegRange+1}{\symCountVariate^{>}_{1\symRegVal}}$,
+and 
+$\sum_{\symRegVal=0}^{\symRegRange+1}{\symCountVariate^{<}_{1\symRegVal}}
++
+{\symCountVariate^{=}_{\symRegVal}}
++
+{\symCountVariate^{<}_{2\symRegVal}}=\symNumReg$.
 

### 2017-01-06 paper/paper.tex
improved description of joint estimation algorithm, redefined phi

@@
 Since the three parameters are all non-negative, this is a constrained optimization problem.
-In order to get rid of these constraints, we use the transformation $\symPoissonRate=\symNumReg e^{\symPhi}$.
+In order to get rid of these constraints, we use the transformation $\symPoissonRate=e^{\symPhi}$.
 This mapping has also the nice property that relative accuracy limits are translated into absolute ones, because $\Delta\symPhi = \Delta\symPoissonRate/\symPoissonRate$.
@@
 \log \symLikelihood(
-\symNumReg e^{\symPhi_\symSetASuffix},
-\symNumReg e^{\symPhi_\symSetBSuffix},
-\symNumReg e^{\symPhi_\symSetXSuffix}
+e^{\symPhi_\symSetASuffix},
+e^{\symPhi_\symSetBSuffix},
+e^{\symPhi_\symSetXSuffix}
 \vert
@@
 \label{equ:definition_xyz}
-\symX_{\ast\symRegVal} :=\frac{e^{\symPhi_\ast}}{2^{\symRegVal}},\quad
+\symX_{\ast\symRegVal} :=\frac{e^{\symPhi_\ast}}{\symNumReg 2^{\symRegVal}},\quad
 \symY_{\ast\symRegVal} := e^{-\symX_{\ast\symRegVal}},\quad
@@
 
-\cref{alg:joint_cardinality_estimation} demonstrates the calculation of the estimates given $\boldsymbol\symCountVariate^{>}_{1}$,
+\cref{alg:joint_cardinality_estimation} demonstrates the calculation of the estimates given the register values $\boldsymbol\symRegValVariate_{1}$ and 
+$\boldsymbol\symRegValVariate_{2}$ of both HyperLogLog sketches. First, the sufficient statistic consisting of $\boldsymbol\symCountVariate^{>}_{1}$,
 $\boldsymbol\symCountVariate^{<}_{1}$,
@@
 $\boldsymbol\symCountVariate^{<}_{2}$, and
-$\boldsymbol\symCountVariate^{=}$. First, a case is distinguished, where
+$\boldsymbol\symCountVariate^{=}$ is extracted. Next, a case is distinguished, where
 all registers have a value equal to zero in at least one of both HyperLogLog sketches, that is $\symCountVariate^{<}_{10}+\symCountVariate^{=}_0+\symCountVariate^{<}_{20} = \symNumReg$.
@@
 
-%TODO
-
-%The initial guess values are important to ensure fast convergence of the optimization algorithm.
-An obvious choice are the cardinality estimates obtained by using the conventional approach \eqref{equ:conventional_approach}.
-
-%As demonstrated by we In all other cases, the maximum likelihood estimates for $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$ will be non-zero.
-We use the conventional approach to get start values for the estimates of $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$, respectively.
-However, in order to ensure that the values are positive we restrict them to values greater than or equal to 1.
-
-For the general case the Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{Press2007} can be applied, which is a very popular algorithm for non-linear optimization problems.
+For the general case, the three-dimensional function $\symFunc(\symPhi_\symSetASuffix,\symPhi_\symSetBSuffix,\symPhi_\symSetXSuffix)$ needs to be optimized numerically.
+A very popular algorithm for such non-linear optimization problems is the Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{Press2007}.
 In particular, we used the implementation provided by the Dlib C++ library \cite{King2009}.
+Good initial guess values are important to ensure fast convergence for the optimization algorithm.
+An obvious choice are the cardinality estimates obtained by the conventional approach \eqref{equ:conventional_approach}.
+However, in order to ensure that their logarithms are defined, we require that the initial values are not smaller than 1.
+Within the optimization loop the function $\symFunc$ and its gradient $\nabla\symFunc$ need to evaluated using \eqref{equ:max_likelihood_phi_func} and \eqref{equ:df_dphi} as needed by the optimization algorithm.
+The optimization loop is continued as long as the absolute change of at least one parameter is larger than a predefined threshold.
+The threshold is again set proportional to $1/\sqrt{\symNumReg}$ (compare \cref{sec:comp_ml_estimate}).
 
@@
 {
-$\boldsymbol\symCountVariate^{>}_{1}$,
-$\boldsymbol\symCountVariate^{<}_{1}$,
-$\boldsymbol\symCountVariate^{>}_{2}$,
-$\boldsymbol\symCountVariate^{<}_{2}$,
-$\boldsymbol\symCountVariate^{=}$
-}
-\State $\symNumReg\gets
-\left\Vert\boldsymbol\symCountVariate^{<}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{1}\right\Vert_1$
+$\boldsymbol\symRegValVariate_{1}$,
+$\boldsymbol\symRegValVariate_{2}$
+}
+\comm{.32}{$\boldsymbol{\symRegValVariate}_1,\boldsymbol{\symRegValVariate}_2\in\lbrace 0,1,\ldots,\symRegRange+1\rbrace^\symNumReg$}
+\State 
+$\boldsymbol\symCountVariate^{>}_{1}\gets 0$,
+$\boldsymbol\symCountVariate^{<}_{1}\gets 0$,
+$\boldsymbol\symCountVariate^{>}_{2}\gets 0$,
+$\boldsymbol\symCountVariate^{<}_{2}\gets 0$,
+$\boldsymbol\symCountVariate^{=}\gets 0$
+\comm{.32}{$\boldsymbol\symCountVariate^{>}_{1},
+\boldsymbol\symCountVariate^{<}_{1},
+\boldsymbol\symCountVariate^{>}_{2},
+\boldsymbol\symCountVariate^{<}_{2},
+\boldsymbol\symCountVariate^{=}\\\in\lbrace 0, 1, \ldots,\symNumReg\rbrace^{\symRegRange+1}$}
+\For{$\symIndexI\gets 1, \symNumReg$}
+\If{$\symRegValVariate_{1\symIndexI}
+<
+\symRegValVariate_{2\symIndexI}$}
+\State $\symCountVariate^{<}_{1\symRegValVariate_{1\symIndexI}}
+\gets
+\symCountVariate^{<}_{1\symRegValVariate_{1\symIndexI}}+1$
+\State
+$\symCountVariate^{>}_{2\symRegValVariate_{2\symIndexI}}
+\gets
+\symCountVariate^{>}_{2\symRegValVariate_{2\symIndexI}}+1$
+\ElsIf{$\symRegValVariate_{1\symIndexI}
+>
+\symRegValVariate_{2\symIndexI}$}
+\State $\symCountVariate^{>}_{1\symRegValVariate_{1\symIndexI}}
+\gets
+\symCountVariate^{>}_{1\symRegValVariate_{1\symIndexI}}+1$
+\State
+$\symCountVariate^{<}_{2\symRegValVariate_{2\symIndexI}}
+\gets
+\symCountVariate^{<}_{2\symRegValVariate_{2\symIndexI}}+1$
+\Else
+\State $\symCountVariate^{=}_{\symRegValVariate_{1\symIndexI}}
+\gets
+\symCountVariate^{=}_{1\symRegValVariate_{1\symIndexI}}+1$
+\EndIf
+\EndFor
 \State $\symPoissonRateEstimate_{\symSetASuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{<}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{1}$}
+\comm{0.32}{use \cref{alg:corrected_raw_estimation} or \ref{alg:estimate_ml}}
 \State $\symPoissonRateEstimate_{\symSetBSuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{<}_{2} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2}$}
 \If{$\symCountVariate^{<}_{10}+\symCountVariate^{=}_0+\symCountVariate^{<}_{20} = \symNumReg$}
+\comm{0.32}{$\Leftrightarrow\min(\symRegValVariate_{1\symIndexI},\symRegValVariate_{2\symIndexI})=0\ \forall\symIndexI$}
 \State $\symPoissonRateEstimate_{\symSetASuffix}\gets
@@
 \symPoissonRateEstimate_{\symSetXSuffix}\gets0$
-\comm{0.27}{in this case $\symSetS_1\cap\symSetS_2 = \symSetX = \emptyset$ holds}
 \State \textbf{return} $(
@@
 \EndIf
+
+%\If{$\boldsymbol\symCountVariate^{>}_{1} = 0$}
+%\comm{0.27}{$\symRegValVariate_{1\symIndexI}\leq
+%\symRegValVariate_{2\symIndexI}\ \forall\symIndexI$}
+%\EndIf
+
+%\If{$\boldsymbol\symCountVariate^{>}_{2} = 0$}
+%\comm{0.27}{$\symRegValVariate_{2\symIndexI}\leq
+%\symRegValVariate_{1\symIndexI}\ \forall\symIndexI$}
+%\EndIf
+
 \State $\symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{>}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2}$}
-\State $\symPhi_\symSetASuffix \gets \log(\max(1, \symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} - \symPoissonRateEstimate_{\symSetBSuffix\symSetXSuffix})/\symNumReg)$
-\State $\symPhi_\symSetBSuffix \gets \log(\max(1, \symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} - \symPoissonRateEstimate_{\symSetASuffix\symSetXSuffix})/\symNumReg)$
+\State $\symPhi_\symSetASuffix \gets \log(\max(1, \symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} - \symPoissonRateEstimate_{\symSetBSuffix\symSetXSuffix}))$
+\State $\symPhi_\symSetBSuffix \gets \log(\max(1, \symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} - \symPoissonRateEstimate_{\symSetASuffix\symSetXSuffix}))$
 \State $\symPhi_\symSetXSuffix \gets \log(\max(1,
@@
 -
-\symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix})/\symNumReg)$
+\symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix}))$
 \State $\symStopDelta\gets\symStopEpsilon/\sqrt{\symNumReg}$
-\comm{0.27}{$\symStopEpsilon = 10^{-2}$}
+\comm{0.32}{$\symStopEpsilon = 10^{-2}$}
 \Repeat
+\comm{0.32}{start optimization}
 \State 
@@
 \symPhi_\symSetBSuffix,
-\symPhi_\symSetXSuffix;
-\symFunc,
-\nabla\symFunc
+\symPhi_\symSetXSuffix
 $}
-\comm{0.27}{e.g. BFGS update step}
+\comm{0.32}{calculate $\symFunc$ and
+$\nabla\symFunc$ as needed by optimizer using \eqref{equ:max_likelihood_phi_func} and \eqref{equ:df_dphi},
+respectively
+}
 \Until{
@@
 $}
+\comm{0.32}{stop criterion}
 \State 
@@
 \symPoissonRateEstimate_{\symSetASuffix}
-\gets \symNumReg e^{\symPhi_\symSetASuffix},\ \symPoissonRateEstimate_{\symSetBSuffix}
-\gets \symNumReg e^{\symPhi_\symSetBSuffix},\ 
+\gets e^{\symPhi_\symSetASuffix},\ \symPoissonRateEstimate_{\symSetBSuffix}
+\gets e^{\symPhi_\symSetBSuffix},\ 
 \symPoissonRateEstimate_{\symSetXSuffix}
-\gets \symNumReg e^{\symPhi_\symSetXSuffix}
+\gets e^{\symPhi_\symSetXSuffix}
 $

### 2017-01-06 c++/cardinality_estimation.hpp

@@ -523,7 +523,7 @@ public:
         numFunctionEvaluations(new std::size_t(0)),
         numGradientEvaluations(new std::size_t(0)) {
 
-        pow2k[q] = ldexp(1., -q);
+        pow2k[q] = ldexp(1., -q-jointStatistic.getP());
         pow2k[q+1] = pow2k[q];
         for (int k = q; k >= 1; --k) {
             pow2k[k-1] = pow2k[k] + pow2k[k];
@@ -548,6 +548,9 @@ public:
         linTermA += jointStatistic.get1Count(0);
         linTermB += jointStatistic.get2Count(0);
         linTermX += jointStatistic.getMinCount(0);
+        linTermA /= jointStatistic.getNumRegisters();
+        linTermB /= jointStatistic.getNumRegisters();
+        linTermX /= jointStatistic.getNumRegisters();
     }
 
     ~JointLogLikelihoodFunction() {
@@ -581,13 +584,16 @@ void estimateInitialCardinalities(const TwoHyperLogLogStatistic& jointStatistic,
         isOptimum = true;
         return;
     }
-    if(jointStatistic.getMinCounts() == jointStatistic.get1Counts()) {
+
+    const std::vector<int>& larger1Counts = jointStatistic.getLarger1Counts();
+    if(std::all_of(larger1Counts.begin(), larger1Counts.end(), [](int i) { return i==0; })) {
         cardinalityA = 0;
         cardinalityB = std::max(0., cardinalityBX - cardinalityAX);
         cardinalityX = cardinalityAX;
         return;
     }
-    if(jointStatistic.getMinCounts() == jointStatistic.get2Counts()) {
+    const std::vector<int>& larger2Counts = jointStatistic.getLarger2Counts();
+    if(std::all_of(larger2Counts.begin(), larger2Counts.end(), [](int i) { return i==0; })) {
         cardinalityB = 0;
         cardinalityA = std::max(0., cardinalityAX - cardinalityBX);
         cardinalityX = cardinalityBX;
@@ -665,9 +671,9 @@ void maxLikelihoodTwoHyperLogLogEstimation(const TwoHyperLogLogStatistic& jointS
         return;
     }
 
-    double phiA = (cardinalityA > 0)?std::log(cardinalityA/m):zeroPhi;
-    double phiB = (cardinalityB > 0)?std::log(cardinalityB/m):zeroPhi;
[…]

### 2017-01-07 paper/paper.tex
some more improvements

@@
 
-\newcommand{\numformat}[1]{{\num[scientific-notation = true,round-mode = places,round-precision = 3, output-exponent-marker = \ensuremath{\mathrm{e}}]{#1}}}
+\newcommand{\numformat}[1]{{\num[scientific-notation = true,round-mode = places,round-precision = 3, output-exponent-marker = \ensuremath{\mathrm{E}}]{#1}}}
 
@@
 In other words, the observed register values are also a plausible sample of the assumed distribution described by \eqref{equ:assumed_register_val_distribution} in this case.
-Hence, as long as all or at least most register values are in the range $[1,\symRegRange]$, the approximation of \eqref{equ:register_value_distribution} by \eqref{equ:assumed_register_val_distribution} is valid.
+Hence, as long as all or at least most register values are in the range $[1,\symRegRange]$, which is the case if $2^\symPrecision \ll \symPoissonRate \ll 2^{\symPrecision
++\symRegRange}$, the approximation of \eqref{equ:register_value_distribution} by \eqref{equ:assumed_register_val_distribution} is valid.
 This explains why the raw estimator works best for intermediate cardinalities.
@@
 $(\boldsymbol\symCountVariate^{>}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2})$.
-In contrast, the maximum likelihood method incorporates each value of the sufficient statistic \eqref{equ:sufficient_joint_statistic} individually.
+In contrast, the maximum likelihood method uses more information as it incorporates all the individual values of the sufficient statistic.
 
@@
 We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches based on the Poisson approximation.
-For the estimation from a single sketch, we have developed two fast algorithms that make use of the Poisson approximation.
+For the estimation from a single sketch, we have developed two different algorithms that are inherently unbiased over the full cardinality range without dependence on empirically determined bias correction data.
 The first uses the original estimator extended by theoretically motivated correction terms.
+Due to its simplicity we believe that it has the potential to become the standard textbook cardinality estimation algorithm for HyperLogLog sketches.
 The second is based on the maximum likelihood method and solves the corresponding optimization problem using the secant method.
-Both algorithms are inherently unbiased and can be implemented without dependence on empirically determined bias correction data.
-
-The maximum likelihood method was also applied to the estimation of set operation result sizes where the operands are represented by HyperLogLog sketches.
-The new approach improves the cardinality estimates of set intersections, relative complements, as well as unions significantly when compared to conventional approaches such as the inclusion-exclusion principle.
+The maximum likelihood method was also successfully applied to the estimation of set operation result sizes where the operands are represented by HyperLogLog sketches.
+The new approach improves the cardinality estimates of set intersections, relative complements, as well as unions significantly when compared to the conventional approach using the inclusion-exclusion principle.
 \appendix

### 2017-01-13 paper/paper.tex
added new insertion algorithm that keeps track of multiplicity vector, added description of simulations

@@
 
-\newcommand{\numformattwo}[1]{{\num[scientific-notation = fixed,round-mode = places,round-precision = 3]{#1}}}
+\newcommand{\numformattwo}[1]{{\num[scientific-notation = fixed,round-mode = places,round-precision = 3,round-integer-to-decimal=true]{#1}}}
 
@@
 We also discuss how the resulting state can be statistically described and approximated by a Poisson model.
-
 In \cref{sec:cardinality_estimation} we describe the original cardinality estimation algorithm, its shortcomings at low and large cardinalities, and previous approaches to overcome them.
@@
 We derive an improved version of the raw estimator, which leads to a new fast cardinality estimation algorithm that works over the full cardinality range as demonstrated by extensive simulations.
-
 In \cref{sec:max_likelihood_estimation} we derive a second even more precise cardinality estimation algorithm based on the maximum likelihood principle which is again verified by simulations.
@@
 The simulation results show that the estimates are significantly better than those obtained by the inclusion-exclusion principle.
-
 Finally, in \cref{sec:future_work} we discuss open problems and ideas for future work including HyperLogLog's potential use as locality-sensitve hashing algorithm before we conclude in \cref{sec:conclusion}.
@@
 \caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers.
-All registers $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ have zero initial value and remain in the range $[0, \symRegRange+1]$.}
+All registers $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ start from zero.}
 \label{alg:insert}
@@
 \State $\langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision,\symBitRepB_1,\ldots,\symBitRepB_\symRegRange\rangle_2 \gets$ $(\symPrecision + \symRegRange)$-bit hash value of $\symDataItem$
-\comm{0.2}{$\symBitRepA_\symIndexI,\symBitRepB_\symIndexI\in\{0,1\}$}
+\comm{0.25}{$\symBitRepA_\symIndexI,\symBitRepB_\symIndexI\in\{0,1\}$}
+\State $\symRegVal \gets \min(\{\symS\mid \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
+\comm{0.25}{$\symRegVal\in\lbrace 1,2,\ldots,\symRegRange+1\rbrace$}
 \State $\symIndexI \gets 1+ \langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision\rangle_2$
-\comm{0.2}{$\symIndexI\in[1,2^\symPrecision]$}
-\State $\symRegVal \gets \min(\{\symS\mid \symS \in [1, \symRegRange]  \wedge  \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
-\comm{0.2}{$\symRegVal\in[1,\symRegRange+1]$}
-\State $\symRegValVariate_\symIndexI \gets\max(\symRegValVariate_\symIndexI, \symRegVal)$
+\comm{0.25}{$\symIndexI\in\lbrace 1,2,\ldots,2^\symPrecision\rbrace$}
+\If{$\symRegVal>\symRegValVariate_\symIndexI$}
+\State $\symRegValVariate_\symIndexI \gets\symRegVal$
+\EndIf
 \EndProcedure
@@
 
-The desribed element insertion algorithm makes use of what is known as stochastic averaging \cite{Flajolet1985}.
+The described element insertion algorithm makes use of what is known as stochastic averaging \cite{Flajolet1985}.
 Instead of updating each of all $\symNumReg$ registers using $\symNumReg$ independent hash values, which would be an $\symBigO(\symNumReg)$ operation, only one register is selected and updated, which requires only a single hash function and reduces the complexity to $\symBigO(1)$.
@@
 \comm{.4}{$\boldsymbol{\symRegValVariate}\in\lbrace 0,1,\ldots,\symRegRange+1\rbrace^\symNumReg$, $\symNumReg = 2^\symPrecision$}
-\State allocate registers for new HyperLogLog sketch $\boldsymbol{\symRegValVariate}' = (\symRegValVariate_1,\ldots,\symRegValVariate_{2^{\symPrecision'}})$
+\State allocate $\boldsymbol{\symRegValVariate}' = (\symRegValVariate_1,\ldots,\symRegValVariate_{2^{\symPrecision'}})$
+\comm{.4}{$\boldsymbol{\symRegValVariate}'\in\lbrace 0,1,\ldots,\symRegRange'+1\rbrace^{\symNumReg'}$, $\symNumReg' = 2^{\symPrecision'}$}
 \For{$\symIndexI\gets 1, 2^{\symPrecision'}$}
@@
 \State $\symRegValVariate'_\symIndexI \gets 
-\min(\{\symS\mid \symS \in [1, \symPrecision-\symPrecision']  \wedge  \symBitRepA_\symS = 1\} \cup \{\symRegRange'+1\})$
+\min(\{\symS\mid \symBitRepA_\symS = 1\} \cup \{\symRegRange'+1\})$
 \Else
@@
 \State \Return $\boldsymbol{\symRegValVariate}'$
-\comm{.4}{$\boldsymbol{\symRegValVariate}'\in\lbrace 0,1,\ldots,\symRegRange'+1\rbrace^{\symNumReg'}$, $\symNumReg' = 2^{\symPrecision'}$}
 \EndFunction
@@
 \comm{.4}{$\boldsymbol{\symRegValVariate}\in\lbrace 0,1,\ldots,\symRegRange+1\rbrace^\symNumReg$}
+\State allocate $\boldsymbol{\symCountVariate} = (\symCountVariate_0,\ldots,\symCountVariate_{\symRegRange+1})$
 \State $\boldsymbol{\symCountVariate}\gets \left(0,0,\ldots, 0\right)$
-\comm{.4}{$\boldsymbol{\symCountVariate} = (\symCountVariate_0,\ldots,\symCountVariate_{\symRegRange+1})$}
 \For{$\symIndexI\gets 1, \symNumReg$}
@@
 \log_2\!\left(
-\frac{2+u}{1+u}
+\frac{2+\symX}{1+\symX}
 \right)
 \right)^{\!\symNumReg}
-du
+d\symX
 \right)^{\!\!-1}.
@@
 \label{equ:alpha_inf}
-\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
+\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} 
+= 0.72134752\ldots,
 \end{equation}
@@
 \comm{.4}{$\boldsymbol{\symRegValVariate}\in\lbrace 0,1,\ldots,\symRegRange+1\rbrace^\symNumReg$, $\symNumReg = 2^\symPrecision$}
-\State $\symCardinalityRawEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexI=1}^\symNumReg 2^{-\symRegValVariate_\symIndexI}\right)^{\!-1}$
+\State $\symCardinalityRawEstimate
+\gets
+\symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexI=1}^\symNumReg 2^{-\symRegValVariate_\symIndexI}\right)^{\!-1}$
 \comm{.4}{raw estimate \eqref{equ:raw_estimator}}
 \If{$\symCardinalityRawEstimate\leq \frac{5}{2}\symNumReg$}
-\State $\symCountVariate_0 = \left|\{\symIndexI \vert\symRegValVariate_\symIndexI=0\}\right|$
+\State $\symCountVariate_0\gets\left|\{\symIndexI \vert\symRegValVariate_\symIndexI=0\}\right|$
 \If{$\symCountVariate_0\neq0$}
@@
 \label{sec:corrected_raw_estimation_error}
-In order to verify the new estimation algorithm, we generated \num{10000} HyperLogLog sketches and inserted up to 50 billion unique elements.
-Assuming a uniform hash function, element hash values can be mocked by random numbers.
-For the following results we used the Mersenne Twister random number generator with a state size of \num{19937} bits from the C++ standard library.
+In order to verify the new estimation algorithm, we generated \num{10000} different HyperLogLog sketches and filled each of them with 50 billion unique elements.
+During element insertion, we took snapshots of the sketch state by storing the multiplicity vector at predefined cardinality values, which were roughly chosen according to a geometric series with ratio \num{1.01}.
+In order to achieve that in reasonable time for different HyperLogLog parameter combinations  $(\symPrecision,\symRegRange)$ we have been interested in, we applied a couple of optimizations.
+
+First, we assumed a uniform hash function.
+This allows us to simply generate a random number instead of getting a unique element and calculating its hash value.
+We used the Mersenne Twister random number generator with a state size of \num{19937} bits from the C++ standard library.
+Second, we used \cref{alg:insert_fast} for insertions, which does basically the same as \cref{alg:insert}, but additionally keeps track of the multiplicity vector and the minimum register value.
+In this way the expensive register scan needed to obtain the multiplicity vector as described by \cref{alg:sufficient_statistic} can be avoided.
+Furthermore, the knowledge of the minimum register value can be used to abort the insertion procedure early, if the hash value is not able to increment any register value.
+Especially at large cardinalities this saves a great number of register accesses and also reduces the number of cache misses.
+For example, if $\symRegValVariate_\text{min}=1$ only half of all insertions will access a register.
+If $\symRegValVariate_\text{min}=2$ it is already just a quarter, and so forth.
+Finally, instead of repeating the simulation for different HyperLogLog parametes, we did that only once using the quite accurate setting $\symPrecision=22$ and $\symRegRange=42$.
+Whenever taking a snapshot, we reduced the sketch state exploiting \cref{alg:compress} for desired parameters before calculating the multiplicity vector using \cref{alg:sufficient_statistic}.
+
+After all, we got the multiplicity vectors at predefined cardinalities of \num{10000} different simulation runs for various HyperLogLog parameters.
+All this data constitutes the basis of the presented results that follow.
+The generated data allows a quick and simple evaluation of cardinality estimation algorithms by just loading the persisted multiplicity vectors from hard disk and passing them as input to the algorithm.
 
@@
 The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is about an order of magnitude larger than the operating range upper bound of the raw estimator (\cref{fig:raw_estimate}).
+
+\begin{algorithm}
+\caption{Extension of \cref{alg:insert} that keeps track of the multiplicity vector $\boldsymbol{\symCountVariate} = (\symCountVariate_0,\ldots,\symCountVariate_{\symRegRange+1})$ and the minimum register value $\symRegValVariate_\text{min}$ during insertion.
+Initially, $\symRegValVariate_\text{min}=0$ and $\boldsymbol{\symCountVariate} = (2^\symPrecision, 0, 0, \dots, 0)$.}
+\label{alg:insert_fast}
+\begin{algorithmic}
+\Procedure {InsertElement}{\symDataItem}
+\State $\langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision,\symBitRepB_1,\ldots,\symBitRepB_\symRegRange\rangle_2 \gets$ $(\symPrecision + \symRegRange)$-bit hash value of $\symDataItem$
+\comm{0.25}{$\symBitRepA_\symIndexI,\symBitRepB_\symIndexI\in\{0,1\}$}
+
+\State $\symRegVal \gets \min(\{\symS\mid \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
+\comm{0.25}{$\symRegVal\in\lbrace 1,2,\ldots,\symRegRange+1\rbrace$}
+\If{$\symRegVal>\symRegValVariate_\text{min}$}
+\State $\symIndexI \gets 1+ \langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision\rangle_2$
+\comm{0.25}{$\symIndexI\in\lbrace 1,2,\ldots,2^\symPrecision\rbrace$}
+\If{$\symRegVal>\symRegValVariate_\symIndexI$}
+\State $\symCountVariate_{\symRegValVariate_\symIndexI} \gets \symCountVariate_{\symRegValVariate_\symIndexI} - 1$
+\State $\symCountVariate_{\symRegVal} \gets \symCountVariate_{\symRegVal} + 1$
+\If{$\symRegValVariate_\symIndexI
+=
+\symRegValVariate_\text{min}$}
+\While{$\symCountVariate_{\symRegValVariate_\text{min}}=0$}
+\State $\symRegValVariate_\text{min}\gets
+\symRegValVariate_\text{min} + 1$
+\EndWhile
+\EndIf
+\State $\symRegValVariate_\symIndexI \gets\symRegVal$
+
+\EndIf
+\EndIf
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
 
@@
 While the union of two sets that are represented by HyperLogLog sketches can be straightforwardly computed using \eqref{equ:hll_union}, the computation of cardinalities of other set operations like intersections and complements is more challenging.
-The conventional approach uses the relationships
+The conventional approach uses the inclusion-exclusion principle
 \begin{equation}
@@
 and the fact that HyperLogLog sketches can be easily merged using \eqref{equ:hll_union}.
-The last equation is also known as inclusion-exclusion principle and can be used to calculate the intersection size.
-Unfortunately, the estimation error does not scale well for the inclusion-exclusion principle.
+Unfortunately, the estimation error does not scale well for this approach.
 Especially for small Jaccard indices, the relative estimation error can become very large \cite{Dasgupta2015}.
@@
 Since \eqref{equ:log_likelihood_pair} is the generalization of \eqref{equ:log_likelihood_single} for two HyperLogLog sketches, the log-likelihood function for a single HyperLogLog sketch can be obtained by considering special cases.
-For example, assuming $\symPoissonRate_\symSetXSuffix=0$, which means that both sketches represent disjoint sets, \eqref{equ:log_likelihood_pair} can be splitted into the sum of two unary functions with parameters $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$ each of which correspond to \eqref{equ:log_likelihood_single} as expected.
+For example, assume $\symPoissonRate_\symSetXSuffix=0$, which means that both sketches represent disjoint sets, \eqref{equ:log_likelihood_pair} can be splitted into the sum of two unary functions with parameters $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$ each of which correspond to \eqref{equ:log_likelihood_single} as expected.
 Or, consider two sketches representing identical sets.

### 2017-01-14 paper/paper.tex
improved estimation algorithms, revised paper

@@
 \let\thefootnote\relax\footnotetext{This paper together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}.
-A first version of this paper was published on April 17, 2016.}
+A first version of this paper was published there on April 17, 2016.}
 \begin{abstract}
@@
 A simple approach to avoid the need of any large range correction is to extend the operating range of the raw estimator to larger cardinalities.
-This can be easily accomplished by using hash values with more bits $(\symPrecision+\symRegRange>32)$.
-If $\symRegRange \geq 30$, 5-bit registers are no longer sufficient to represent all possible values in the range $[0, \symRegRange+1]$.
-If, for example, 64-bit hash values $(\symPrecision+\symRegRange=64)$ are used, as proposed in \cite{Heule2013}, 6 bits per register are needed.
-Hash value sizes with more than 64 bits are useless in practice, because it is unrealistic to encounter cardinalities of order $2^{64}$ for which the raw estimator would be biased.
+This can be easily accomplished by increasing $\symPrecision+\symRegRange$, which corresponds to using hash values with more bits.
+Each additional bit doubles the operating range which scales like $2^{\symPrecision+\symRegRange}$.
+However, in case $\symRegRange \geq 31$ the number of possible register values, which are $\lbrace 0, 1, \ldots, \symRegRange+1\rbrace$, is greater than 32 which is no longer representable by 5-bit registers.
+Therefore, it was proposed to use 6 bits per register in combination with 64-bit hash values \cite{Heule2013}.
+Even larger hash values are needless in practice, because it is unrealistic to encounter cardinalities of order $2^{64}$.
 
@@
 \comm{0.45}{$\symX\in[0,1]$}
+\If{$\symX=0\vee \symX=1$}
+\State\Return$0$
+\EndIf
 \State $\symY\gets1$
@@
 However, we can imagine that a better numerical approximation of $\symLargeCorrectionFunc$ can be found than that given in \cref{alg:corrected_raw_estimation} and which allows on-demand evaluation without significant extra costs.
-If $2^{\symPrecision+\symRegRange}$ is greater than the maximum expected cardinality value, the number of saturated registers is negligible ($\symCountVariate_{\symRegRange+1}\ll \symNumReg$) and the calculation of $\symLargeCorrectionFunc$ can be omitted at all, because $\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg) \approx 0$ is always true.
-
-The numbers presented in \cref{fig:corrected_raw_avg_exec_time} do not include the processing time to extract the multiplicity vector out of the HyperLogLog sketch, which requires a complete scan over all registers and counting the different register values into an array (compare \cref{alg:sufficient_statistic}).
+If $2^{\symPrecision+\symRegRange}$ is greater than the maximum expected cardinality value, the number of saturated registers is negligible ($\symCountVariate_{\symRegRange+1}\ll \symNumReg$) and the calculation of $\symLargeCorrectionFunc$ can be omitted at all, because $\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg) \approx 0$ vanishes in this case.
+
+The numbers presented in \cref{fig:corrected_raw_avg_exec_time} do not include the processing time to extract the multiplicity vector out of the HyperLogLog sketch, which requires a complete scan over all registers and counting the different register values into an array as demonstrated by \cref{alg:sufficient_statistic}.
 A theoretical lower bound for this processing time can be derived using the maximum memory bandwidth of the CPU, which is \SI[per-mode=symbol]{21}{\giga\byte\per\second} for an Intel Core i5-2500K.
@@
 $\symHelper(\symX)$ is strictly increasing and concave as can be seen in \cref{fig:helper_function}.
-For non-negative values $\symX$ the function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
+For non-negative values this function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
 \begin{figure}
@@
 can be found, where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$ and
-$\symRegValVariate_\text{max} := \max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$.
+$\symRegValVariate_\text{max} := \max(\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace)$.
 Again, inserting $\symXEstimate$ and transformation gives
@@
 Since the expected statistical errror for the HyperLogLog data structure scales according to $\frac{1}{\sqrt{\symNumReg}}$ \cite{Flajolet2007}, it makes sense to choose $\symStopDelta = \frac{\symStopEpsilon}{\sqrt{\symNumReg}}$ with some constant $\symStopEpsilon$.
-For all results presented later in \cref{sec:maximum_likelihood_estimation_error} we have used $\symStopEpsilon = 10^{-2}$.
+For all results presented later in \cref{sec:maximum_likelihood_estimation_error} we used $\symStopEpsilon = 10^{-2}$.
 
@@
 \item Only a fraction of all count values $\symCountVariate_\symRegVal$ is non-zero.
-If we denote $\symRegValVariate_\text{min}:=\min\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$ and $\symRegValVariate_\text{max}:=\max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$,  it is sufficient to loop over all indices in the range $[\symRegValVariate_\text{min}, \symRegValVariate_\text{max}]$.
+If we denote $\symRegValVariate_\text{min}:=\min(\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace)$ and $\symRegValVariate_\text{max}:=\max(\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace)$,  it is sufficient to loop over all indices in the range $[\symRegValVariate_\text{min}, \symRegValVariate_\text{max}]$.
 \item The sum $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}$ in \eqref{equ:funcprime} can be precalculated and reused for all function evaluations.
@@
 \frac{\symX}{2^{\symRegValVariate'_\text{max}}} 
+\leq
+\frac{\symXEstimate}{2^{\symRegValVariate'_\text{max}}} 
 \leq
@@
 $\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ is finally obtained after $\symKappa -\symRegValVariate'_\text{max}$ iterations using \eqref{equ:helper_recursion1}.
-As shown in \cref{app:error_approx}, a small approximation error of $\symHelper$ does not have much impact on the error of the maximum likelihood estimate as long as most registers are not yet saturated.
+As shown in \cref{app:error_approx}, a small approximation error of $\symHelper$ does not have much impact on the error of the maximum likelihood estimate as long as most registers are not saturated.
 \end{itemize}
@@
 The square root that appears for the calculation of the stopping limit $\symStopDelta=\symStopEpsilon/\sqrt{\symNumReg}$ can be precomputed, because it only depends on the HyperLogLog precision parameter $\symPrecision$.
-
 For very large cardinalities it makes sense to use the strong \eqref{equ:strong_lower_bound} instead of the weak lower bound \eqref{equ:weak_lower_bound} as second starting point for the secant method.
@@
 \comm{.35}{$\dim \boldsymbol{\symCountVariate} = \symRegRange + 2$, $\left\Vert\boldsymbol{\symCountVariate}\right\Vert_1 = \symNumReg$}
-\State $\symRegValVariate_\text{min} \gets \min\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$
-\If{$\symRegValVariate_\text{min} > \symRegRange$}
+\If{$\symCountVariate_{\symRegRange+1} = \symNumReg$}
 \State\Return $\infty$
 \EndIf
+\State $\symRegValVariate_\text{min} \gets \min(\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace)$
 \State $\symRegValVariate'_\text{min} \gets \max(\symRegValVariate_\text{min}, 1)$
-\State $\symRegValVariate_\text{max} \gets \max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$
+\State $\symRegValVariate_\text{max} \gets \max(\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace)$
 \State $\symRegValVariate'_\text{max} \gets \min(\symRegValVariate_\text{max}, \symRegRange)$
 \State $\symZ \gets 0$
-\State $\symY \gets 2^{-\symRegValVariate'_\text{max}}$
 \For{$\symRegVal \gets \symRegValVariate'_\text{max},\symRegValVariate'_\text{min}$}
-\State $\symZ \gets \symZ + \symCountVariate_\symRegVal\cdot\symY$
-\comm{0.35}{here $\symY = 2^{-\symRegVal}$}
-\State $\symY \gets 2\symY$
+\State $\symZ \gets 0.5\cdot\symZ + \symCountVariate_\symRegVal$
 \EndFor
-\State
+\State $\symZ \gets \symZ\cdot 2^{-\symRegValVariate'_\text{min}}$
 \comm{0.35}{here $\symZ=\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$}
@@
 The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:avg_exec_time}.
-The average computation time for the maximum likelihood algorithm shows a different behavior than for the improved raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}).As can be seen, the average execution time is larger for most cardinalities, but nevertheless, it never exceeds \SI{700}{\nano\second} which is still fast enough for many applications.
+The average computation time for the maximum likelihood algorithm shows a different behavior than for the improved raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}).
+As can be seen, the average execution time is larger for most cardinalities, but nevertheless, it never exceeds \SI{700}{\nano\second} which is still fast enough for many applications.
 The steps in the chart can be explained by different numbers of iteration cycles until the secant method is stopped.
-For example, at a cardinality value around \num{5000} the average number of required cycles until the stop criterion is met increases abruptly from two to three.
+For example, at a cardinality value around \num{5000} the average number of required cycles until the stop criterion is satisfied increases abruptly from two to three.
 More than three iteration cycles have never been observed for any cardinality estimate in this performance test.
@@
 \end{multline}
-where the constants
+where
 $\symCountVariate^{<}_{1\symRegVal}$,
@@
 For a hash algorithm that maps two different sets to the same $\symNumBitsMinwiseHashing$-bit hash value with probability $1-\symA\symDistanceMeasure$ with some constant $\symA\in(0,1]$, the storage factor is given by $\symNumBitsMinwiseHashing\symDistanceMeasure\left(\frac{1}{\symA}-\symDistanceMeasure\right)$.
-Particularly, for the case $D=0.3$, this gives \num{6.72} for conventional minwise hashing with 32-bit hash values ($\symNumBitsMinwiseHashing=32$, $\symA\approx 1$), \num{0.51} for 1-bit minwise hashing ($\symNumBitsMinwiseHashing=1$, $\symA=\frac{1}{2}$), and  \num{1.63} for the HyperLogLog algorithm with 5-bit registers ($\symNumBitsMinwiseHashing=5$, $\symA=
-\symAlpha_\infty$).
+Particularly, for the case $D=0.3$, this gives \num{6.72} for conventional minwise hashing with 32-bit hash values ($\symNumBitsMinwiseHashing=32$, $\symA\approx 1$), \num{0.51} for 1-bit minwise hashing ($\symNumBitsMinwiseHashing=1$, $\symA=\frac{1}{2}$), and  \num{1.96} for the HyperLogLog algorithm with 6-bit registers ($\symNumBitsMinwiseHashing=6$, $\symA=\symAlpha_\infty$).
 This means that the HyperLogLog algorithm is not as memory-efficient as 1-bit minwise hashing, but significantly better than 32-bit minwise hashing, when estimating the Jaccard distance by just counting equal registers.

### 2017-01-14 c++/benchmark.cpp

@@ -166,13 +166,10 @@ void run(const int p, const int q, const string& resultsFileName) {
             int innerLoop1IterationsCount;
             int innerLoop2IterationsCount;
             int logEvaluationCount;
-            int kMin;
-            int kMax;
-
 
             auto start = std::chrono::system_clock::now();
             for(auto c : counts) {
-                sum += maxLikelihoodEstimator.estimate(c, outerLoopIterationsCount, innerLoop1IterationsCount, innerLoop2IterationsCount, logEvaluationCount, kMin, kMax);
+                sum += maxLikelihoodEstimator.estimate(c, outerLoopIterationsCount, innerLoop1IterationsCount, innerLoop2IterationsCount, logEvaluationCount);
                 outerLoopIterationsCountSum += outerLoopIterationsCount;
                 innerLoop1IterationsCountSum += innerLoop1IterationsCount;
                 innerLoop2IterationsCountSum += innerLoop2IterationsCount;

### 2017-01-14 c++/cardinality_calculator.cpp

@@ -156,9 +156,6 @@ int main(int argc, char* argv[])
         vector<double> strongUpperBoundEstimates(resultBufferSize);
         vector<double> weakUpperBoundEstimates(resultBufferSize);
 
-        vector<int> kMinSum(cardinalities.size());
-        vector<int> kMaxSum(cardinalities.size());
-
         const MaxLikelihoodEstimator maxLikelihoodEstimator(p,q);
         const ImprovedRawEstimator improvedRawEstimator(p,q);
 
@@ -181,36 +178,18 @@ int main(int argc, char* argv[])
 
                 size_t resultPos = cardinalityIdx*seeds.size()+seedCounter;
 
-                int kMin;
-                int kMax;
-
                 flajoletSmallRangeEstimates[resultPos] = flajoletSmallRangeEstimate(c);
                 flajoletMidRangeEstimates[resultPos] = flajoletRawEstimate(c);
                 flajoletMidRangeEstimatesCorrected[resultPos] = improvedRawEstimator.estimate_on_demand(c, numSmallCorrectionIterations[resultPos], numLargeCorrectionIterations[resultPos]);
                 //flajoletMidRangeEstimatesCorrected[resultPos] = correctedRawEstimator.estimate_precalculated(c);
                 flajoletEstimates[resultPos] = flajoletEstimate(c);
-                maxLikelihoodEstimates[resultPos] = maxLikelihoodEstimator.estimate(c, outerLoopIterationsCount[resultPos], innerLoop1IterationsCount[resultPos], innerLoop2IterationsCount[resultPos], logEvaluationCount[resultPos], kMin, kMax);
+                maxLikelihoodEstimates[resultPos] = maxLikelihoodEstimator.estimate(c, outerLoopIterationsCount[resultPos], innerLoop1IterationsCount[resultPos], innerLoop2IterationsCount[resultPos], logEvaluationCount[resultPos]);
                 weakLowerBoundEstimates[resultPos] = weakLowerBoundEstimate(c);
                 strongLowerBoundEstimates[resultPos] = strongLowerBoundEstimate(c);
                 weakUpperBoundEstimates[resultPos] = weakUpperBoundEstimate(c);
                 strongUpperBoundEstimates[resultPos] = strongUpperBoundEstimate(c);
-
-                #pragma omp atomic
-                kMinSum[cardinalityIdx] += kMin;
-
-                #pragma omp atomic
[…]

### 2017-01-14 c++/cardinality_estimation.hpp

@@ -48,12 +48,16 @@ public:
 
     MaxLikelihoodEstimator(const int p_, const int q_) :p(p_), q(q_), m(1 << p_), relativeErrorLimit(eps/(sqrt(m))) {}
 
-    double estimate(const std::vector<int>& c, int& outerLoopIterationsCount, int& innerLoop1IterationsCount, int& innerLoop2IterationsCount, int& logEvaluationCount, int& kMin, int& kMax) const {
+    double estimate(const std::vector<int>& c, int& outerLoopIterationsCount, int& innerLoop1IterationsCount, int& innerLoop2IterationsCount, int& logEvaluationCount) const {
 
         outerLoopIterationsCount = 0;
         innerLoop1IterationsCount = 0;
         innerLoop2IterationsCount = 0;
         logEvaluationCount = 0;
+
+        if (c[q+1] == m) return std::numeric_limits<double>::infinity();
+
+        int kMin, kMax;
 
         for(kMin=0; c[kMin]==0; ++kMin);
         int kMinPrime = std::max(1, kMin);
@@ -61,14 +65,11 @@ public:
         for(kMax=q+1; c[kMax]==0; --kMax);
         int kMaxPrime = std::min(q, kMax);
 
-        if (kMin > q) return std::numeric_limits<double>::infinity();
-
         double z = 0.;
-        double y = ldexp(1., -kMaxPrime);
         for (int k = kMaxPrime; k >= kMinPrime; --k) {
-            z += y*c[k];
-            y += y;
+            z = 0.5*z + c[k];
         }
+        z = ldexp(z, -kMinPrime);
 
         int cPrime = c[q+1];
         if (q >= 1) {
@@ -132,10 +133,8 @@ public:
         int innerLoop1IterationsCount;
         int innerLoop2IterationsCount;
         int logEvaluationCount;
-        int kMin;
-        int kMax;
 
-        return estimate(c, outerLoopIterationsCount, innerLoop1IterationsCount, innerLoop2IterationsCount, logEvaluationCount, kMin, kMax);
+        return estimate(c, outerLoopIterationsCount, innerLoop1IterationsCount, innerLoop2IterationsCount, logEvaluationCount);
     }
 };
 
@@ -196,12 +195,13 @@ class ImprovedRawEstimator {
             zPrime = z;
             z += x * y;
             y += y;
-        } while(zPrime < z);
[…]

### 2017-01-15 paper/paper.tex
some minor improvements

@@
 \end{equation}
-If the HyperLogLog sketch is in the intermediate range, where $\symCountVariate_0=\symCountVariate_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimator \eqref{equ:raw_estimator}.
-Hence, consequent application of the maximum likelihood method leads directly to the harmonic mean that is used by the raw estimator.
+If the HyperLogLog sketch is in the intermediate range, where $\symCountVariate_0=\symCountVariate_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor and both are proportional to the harmonic mean of $2^{\symRegValVariate_1},\ldots,2^{\symRegValVariate_\symNumReg}$.
+Hence, consequent application of the maximum likelihood method would have directly suggested to use a cardinality estimator that is proportional to the harmonic mean without knowing the raw estimator \eqref{equ:raw_estimator} in advance.
 
@@
 The estimate for $\symPoissonRate_\symX$ could be anything between 0 and $\symPoissonRateEstimate_2$.
-For comparison, the inclusion-exclusion approach would give $\symPoissonRateEstimate_2$ as estimate.
-This result is questionable, because there is no evidence in this case, that the sets $\symSetS_1$ and $\symSetS_2$ really have common elements.
-
-%Since \eqref{equ:log_likelihood_pair} is the generalization of \eqref{equ:log_likelihood_single} for two HyperLogLog sketches, the log-likelihood function for a single HyperLogLog sketch can be obtained by considering special cases.
-For example, assuming $\symPoissonRate_\symSetXSuffix=0$, which means that both sketches represent disjoint sets, \eqref{equ:log_likelihood_pair} can be splitted into the sum of two unary functions with parameters $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$, respectively.
-As expected, both functions correspond to \eqref{equ:log_likelihood_single}.
-Another extreme case is that both sketches represent identical sets.
-In this case all registers are equal, which means  that $\symCountVariate^{>}_{1\symRegVal}=\symCountVariate^{>}_{2\symRegVal}=
-%\symCountVariate^{<}_{1\symRegVal}=\symCountVariate^{<}_{2\symRegVal}=0$ for all $\symRegVal$, and therefore the maximum likelihood method yields $\symPoissonRateEstimate_\symSetASuffix = \symPoissonRateEstimate_\symSetBSuffix = 0$ and the value for $\symPoissonRateEstimate_\symSetXSuffix$ will be equal to the individual maximum likelihood estimate of both identical sketches.
-
-%$
-%\symPoissonRate_\symSetASuffix
-%=
-%\symPoissonRate_\symSetBSuffix=0
-%$ and all the register values must be pairwise identical, which finally 
-%yields \eqref{equ:log_likelihood_single} for $\symPoissonRate_\symSetXSuffix$.
+For comparison, the inclusion-exclusion approach would give $\symPoissonRateEstimate_2$ as estimate for $\symPoissonRate_\symX$.
+This result is questionable, because there is no evidence that sets $\symSetS_1$ and $\symSetS_2$ really have common elements in this case.
 
@@
 In order to reduce the numerical error of $\symZ_{\ast\symRegVal}$ for small $\symX_{\ast\symRegVal}$, it is essential to use the function $\operatorname{expm1}(\symX):=e^\symX-1$ that is available in most programming languages.
-If $\symX_{\ast\symRegVal}$ is smaller than $\log(2)$, we calculate $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$ via $\symZ_{\ast\symRegVal} =
--\operatorname{expm1}(-\symX_{\ast\symRegVal})$ and $\symY_{\ast\symRegVal} = 1 - \symZ_{\ast\symRegVal}$ and not as defined in \eqref{equ:definition_xyz}.
+If $\symX_{\ast\symRegVal}$ is smaller than $\log(2)$, we calculate $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$ using $\symZ_{\ast\symRegVal} =
+-\operatorname{expm1}(-\symX_{\ast\symRegVal})$ and $\symY_{\ast\symRegVal} = 1 - \symZ_{\ast\symRegVal}$, repsectively, and not as defined in \eqref{equ:definition_xyz}.
 In this way the numerical error of both, $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$, is minimized and still only a single exponential function needs to be evaluated.
-
 Apart from that, the numerical evaluation of \eqref{equ:max_likelihood_phi_func} is straightforward.
@@
 Good initial guess values are important to ensure fast convergence for the optimization algorithm.
-An obvious choice are the cardinality estimates obtained by the conventional approach \eqref{equ:conventional_approach}.
-However, in order to ensure that their logarithms are defined, we require that the initial values are not smaller than 1.
+An obvious choice are the cardinality estimates obtained by application of the inclusion-exclusion principle \eqref{equ:conventional_approach}.
+However, in order to ensure that their logarithms are all defined, we require that the initial values are not smaller than 1.
 Within the optimization loop the function $\symFunc$ and its gradient $\nabla\symFunc$ need to evaluated using \eqref{equ:max_likelihood_phi_func} and \eqref{equ:df_dphi} as needed by the optimization algorithm.
@@
 \comm{.32}{$\boldsymbol{\symRegValVariate}_1,\boldsymbol{\symRegValVariate}_2\in\lbrace 0,1,\ldots,\symRegRange+1\rbrace^\symNumReg$}
-\State 
-$\boldsymbol\symCountVariate^{>}_{1}\gets 0$,
-$\boldsymbol\symCountVariate^{<}_{1}\gets 0$,
-$\boldsymbol\symCountVariate^{>}_{2}\gets 0$,
-$\boldsymbol\symCountVariate^{<}_{2}\gets 0$,
-$\boldsymbol\symCountVariate^{=}\gets 0$
-\comm{.32}{$\boldsymbol\symCountVariate^{>}_{1},
-\boldsymbol\symCountVariate^{<}_{1},
-\boldsymbol\symCountVariate^{>}_{2},
-\boldsymbol\symCountVariate^{<}_{2},
-\boldsymbol\symCountVariate^{=}\\\in\lbrace 0, 1, \ldots,\symNumReg\rbrace^{\symRegRange+1}$}
+
+\State $\boldsymbol\symCountVariate^{>}_{1}\gets (0,\ldots,0)$
+\comm{.32}{$\boldsymbol{\symCountVariate}^{>}_{1} = (\symCountVariate^{>}_{10},\ldots,\symCountVariate^{>}_{1,\symRegRange+1})$}
+\State $\boldsymbol\symCountVariate^{<}_{1}\gets (0,\ldots,0)$
+\comm{.32}{$\boldsymbol{\symCountVariate}^{<}_{1} = (\symCountVariate^{<}_{10},\ldots,\symCountVariate^{<}_{1,\symRegRange+1})$}
+\State $\boldsymbol\symCountVariate^{>}_{2}\gets (0,\ldots,0)$
+\comm{.32}{$\boldsymbol{\symCountVariate}^{>}_{2} = (\symCountVariate^{>}_{20},\ldots,\symCountVariate^{>}_{2,\symRegRange+1})$}
+\State $\boldsymbol\symCountVariate^{<}_{2}\gets (0,\ldots,0)$
+\comm{.32}{$\boldsymbol{\symCountVariate}^{<}_{2} = (\symCountVariate^{<}_{20},\ldots,\symCountVariate^{<}_{2,\symRegRange+1})$}
+\State $\boldsymbol\symCountVariate^{=}\gets (0,\ldots,0)$
+\comm{.32}{$\boldsymbol{\symCountVariate}^{=} = (\symCountVariate^{=}_{0},\ldots,\symCountVariate^{=}_{\symRegRange+1})$}
+
 \For{$\symIndexI\gets 1, \symNumReg$}
@@
 \EndIf
-
-%\If{$\boldsymbol\symCountVariate^{>}_{1} = 0$}
-%\comm{0.27}{$\symRegValVariate_{1\symIndexI}\leq
-%\symRegValVariate_{2\symIndexI}\ \forall\symIndexI$}
-%\EndIf
-
-%\If{$\boldsymbol\symCountVariate^{>}_{2} = 0$}
-%\comm{0.27}{$\symRegValVariate_{2\symIndexI}\leq
-%\symRegValVariate_{1\symIndexI}\ \forall\symIndexI$}
-%\EndIf
-
 \State $\symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{>}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2}$}
@@
 For some cases the improvement factor is clearly larger than 2.
-Due to square root error scaling law, this means that we would need 4 times more registers to get the same error when using the conventional approach.
+Due to the  square root scaling of the error, this means that we would need 4 times more registers to get the same error when using the conventional approach.
 We also considered some cases with true cardinalities smaller  than the number of registers.

### 2017-01-22 paper/paper.tex
revised introduction and some other changes

@@
 \symDefine{\symSetB}{B}
+\symDefine{\symSetC}{C}
 \symDefine{\symSetS}{S}
@@
 \symDefine{\symSetBSuffix}{b}
+\symDefine{\symSetCSuffix}{c}
 \symDefine{\symSetXSuffix}{x}
@@
 The maximum likelihood approach can also be applied to more than a single HyperLogLog sketch.
-In particular, it is shown that it gives more accurate cardinality estimates for the union, intersection, or relative complements of two sets that are represented by HyperLogLog sketches compared to conventional techniques like the inclusion-exclusion principle.
+In particular, it is shown that it gives more accurate cardinality estimates for the union, intersection, or relative complements of two sets that are both represented by HyperLogLog sketches compared to the conventional technique using the inclusion-exclusion principle.
 All the new methods are demonstrated and verified by extensive simulations.
@@
 Counting the number of distinct elements in a data stream or large datasets is a common problem in big data processing.
-Often there are parallel streams  or data is spread over a cluster, which makes this task even more challenging.
 In principle, finding the number of distinct elements $\symCardinality$ with a maximum relative error $\symError$  in a data stream requires $\Omega(\symCardinality)$ space \cite{Alon1999}.
 However, probabilistic algorithms that achieve the requested accuracy only with high probability are able to drastically reduce space requirements.
-Many different probabilistic algorithms have been developed over the past two decades \cite{Metwally2008,Ting2014}.
-An algorithm with an optimal space complexity of $\Omega(\symError^{-2}+\log \symCardinality)$ \cite{Alon1999, Indyk2003} was finally presented \cite{Kane2010}.
-This algorithm, however, is not very efficient in practice \cite{Ting2014}.
-
-Currently, the most memory efficient algorithm that also works for distributed setups is the near-optimal HyperLogLog algorithm \cite{Flajolet2007} with space complexity $\Omega(\symError^{-2} \log\log\symCardinality +\log \symCardinality)$.
-The originally proposed estimation method has some problems to guarantee the same estimation error over the entire range of cardinalities.
-It was proposed to correct the estimate by empirical means \cite{Heule2013,Rhodes2015,Sanfilippo2014}.
-
-In case the data is not distributed and results do not need to be aggregated further, there are even more efficient estimation algorithms available.
-On the one hand there is the self-learning bitmap \cite{Chen2011}, and on the other hand there is the HyperLogLog algorithm extended by a historic inverse probability estimator that is continuously updated together with the HyperLogLog sketch \cite{Ting2014, Cohen2014}.
-Both achieve the same estimation error using less space.
-However, the estimated cardinality depends on the insertion order of elements and hence cannot be used in a distributed environment.
+Many different probabilistic algorithms have been developed over the past two decades \cite{Metwally2008,Ting2014} until a theoretically optimal algorithm was finally found \cite{Kane2010}.
+Although this algorithm achieves the optimal space complexity of $\Omega(\symError^{-2}+\log \symCardinality)$ \cite{Alon1999, Indyk2003}, it is not very efficient in practice \cite{Ting2014}.
+
+More practicable and already widely used for many applications is the HyperLogLog algorithm \cite{Flajolet2007} with a near-optimal space complexity $\Omega(\symError^{-2} \log\log\symCardinality +\log \symCardinality)$.
+It has the nice property that partial results can be easily merged, which is a necessity for distributed environments.
+Unfortunately, the originally proposed estimation method has some problems to guarantee the same accuracy over the full cardinality range.
+Therefore, a couple of variants have been developed to correct the original estimate by empirical means \cite{Heule2013,Rhodes2015,Sanfilippo2014}.
+
+A more theoretically profound estimator for HyperLogLog sketches that does not depend on empirical data and significantly improves the estimation error is the historic inverse probability estimator \cite{Ting2014, Cohen2014}.
+It trades memory-efficiency for mergeability.
+The estimator needs to be continuously updated while inserting elements and the estimate depends on the element insertion order.
+Moreover, the estimator cannot be further used after merging two sketches, which limits its application to single data streams.
+If this restriction is okay, the self-learning bitmap \cite{Chen2011}
+could also be used, which provides a similar trade-off and is also 
+more memory-efficient than the original HyperLogLog method.
+
+Sometimes not only the number of distinct elements but also a sample of them is needed, in order to allow later filtering according to some predicate and estimating the cardinalities of corresponding subsets.
+In this case the k-minimum values algorithm \cite{Beyer2007} is the method of choice which also allows set manipulations like construction of intersections, disjoint sets, or unions \cite{Dasgupta2015}.
+The latter operation is the only one that is natively supported by HyperLogLog sketches.
+A sketch that represents the set operation result is not always needed.
+One approach to estimate the corresponding result cardinality directly is based on the inclusion-exclusion principle, which however can become quite inaccurate, especially for small Jaccard indices of the input sets \cite{Dasgupta2015}.
+Alternatively, the HyperLogLog sketches can be combined with some additional data structure that allows estimating the Jaccard index.
+Together with the HyperLogLog cardinality estimates the error of the intersection size estimate can be reduced \cite{Pascoe2013, Cohen2016}, however, at the expense of a significantly larger memory footprint due to the additional data structure.
 
 \subsection{Outline}
-In \cref{sec:hyperloglog_data_structure} we give an introduction to HyperLogLog sketches and the corresponding update algorithm.
+This paper presents new algorithms for cardinality estimation from HyperLogLog sketches.
+In \cref{sec:hyperloglog_data_structure} we start with an introduction to HyperLogLog sketches and the corresponding update algorithm.
 We also discuss how the resulting state can be statistically described and approximated by a Poisson model.
@@
 The sum $\symPrecision+\symRegRange$ corresponds to the number of consumed hash value bits and defines the maximum cardinality that can be tracked.
-Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent and the estimation accuracy will be drastically reduced.
+Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent and the estimation error will increase drastically.
 
@@
 Furthermore, any two HyperLogLog sketches with same parameters $(\symPrecision, \symRegRange)$ representing sets $\symSetA$ and $\symSetB$ can be easily merged.
-The HyperLogLog sketch that represents the union of both sets is simply constructed by taking the register-wise maximum values
-\begin{equation}
-\label{equ:hll_union}
-\symRegValVariate^{\symSetA\cup\symSetB}_{\symIndexI} = 
-\max
-\!
-\left(
-\symRegValVariate^{\symSetA}_{\symIndexI},
-\symRegValVariate^{\symSetB}_{\symIndexI}
-\right)
-\quad
-\text{for $1\leq\symIndexI\leq\symNumReg$}.
-\end{equation}
+The HyperLogLog sketch that represents the union of both sets is simply constructed by taking the register-wise maximum values as demonstrated by \cref{alg:union}.
+
+\begin{algorithm}
+\caption{Merge operation for two HyperLogLog sketches with register values $\boldsymbol{\symRegValVariate}_1 = (\symRegValVariate_{11},\ldots,\symRegValVariate_{1\symNumReg})$ and
+$\boldsymbol{\symRegValVariate}_2 = (\symRegValVariate_{21},\ldots,\symRegValVariate_{2\symNumReg})$ representing sets $\symSetS_1$ and $\symSetS_2$, respectively, to obtain the register values $\boldsymbol{\symRegValVariate}' = (\symRegValVariate'_{1},\ldots,\symRegValVariate'_{\symNumReg})$ of a HyperLogLog sketch representing $\symSetS_1\cup\symSetS_2$.}
+\label{alg:union}
+\begin{algorithmic}
+\Function {Merge}{$\boldsymbol{\symRegValVariate}_1$,$\boldsymbol{\symRegValVariate}_2$}
+\comm{.4}{$\boldsymbol{\symRegValVariate}_1,\boldsymbol{\symRegValVariate}_2\in\lbrace 0,1,\ldots,\symRegRange+1\rbrace^\symNumReg$}
+\State allocate $\boldsymbol{\symRegValVariate'} = (\symRegValVariate'_{1},\ldots,\symRegValVariate'_{\symNumReg})$
+\comm{.4}{$\boldsymbol{\symRegValVariate}'\in\lbrace 0,1,\ldots,\symRegRange+1\rbrace^\symNumReg$}
+\For{$\symIndexI\gets 1, \symNumReg$}
+\State $\symRegValVariate'_{\symIndexI}
+\gets\max(\symRegValVariate_{1\symIndexI},\symRegValVariate_{2\symIndexI})$
+\EndFor
+\State \Return $\boldsymbol{\symRegValVariate'}$
+\EndFunction
+\end{algorithmic}
+\end{algorithm}
 
@@
 \comm{.4}{$\boldsymbol{\symRegValVariate}\in\lbrace 0,1,\ldots,\symRegRange+1\rbrace^\symNumReg$, $\symNumReg = 2^\symPrecision$}
-\State allocate $\boldsymbol{\symRegValVariate}' = (\symRegValVariate_1,\ldots,\symRegValVariate_{2^{\symPrecision'}})$
+\State allocate $\boldsymbol{\symRegValVariate}' = (\symRegValVariate'_1,\ldots,\symRegValVariate'_{2^{\symPrecision'}})$
 \comm{.4}{$\boldsymbol{\symRegValVariate}'\in\lbrace 0,1,\ldots,\symRegRange'+1\rbrace^{\symNumReg'}$, $\symNumReg' = 2^{\symPrecision'}$}
@@
 In addition, this vector contains all the information about the HyperLogLog sketch that is required for cardinality estimation.
-The two HyperLogLog parameters can be obtained by $\symPrecision = \log_2 \left\Vert\boldsymbol{\symCountVariate}\right\Vert_1$ and $\symRegRange=\dim\boldsymbol{\symCountVariate}-2$, respectively.
+The two HyperLogLog parameters can be obtained by $\symPrecision = \log_2 (\sum_{\symRegVal=0}^{\symRegRange+1}\symCountVariate_\symRegVal
+)$ and $\symRegRange=\dim\boldsymbol{\symCountVariate}-2$, respectively.
 As \cref{alg:sufficient_statistic} shows the calculation of the multiplicity vector $\boldsymbol{\symCountVariate}$ for a given HyperLogLog sketch with register values $\boldsymbol{\symRegValVariate}$ is straightforward.
@@
 Given that, a rough cardinality estimate can be obtained by averaging the values $\lbrace \symNumReg 2^{\symRegValVariate_1},\ldots,\symNumReg2^{\symRegValVariate_\symNumReg}\rbrace$.
-
 In the history of the HyperLogLog algorithm different averaging techniques have been proposed.
@@
 if $\symCountVariate'_\symRegVal$ is replaced by $\hat{\symCount}'_\symRegVal$. 
-
-Inserting all the registers into \eqref{equ:raw_estimate_with_multiplicities} as replacements for $\symCountVariate'_\symRegVal$ finally gives
+Plugging all these estimators into \eqref{equ:raw_estimate_with_multiplicities} as replacements for $\symCountVariate'_\symRegVal$ finally gives
 \begin{equation}
@@
 \Function {EstimateCardinality}{$\boldsymbol{\symCountVariate}$}
-\comm{.45}{$\dim \boldsymbol{\symCountVariate} = \symRegRange + 2$, $\left\Vert\boldsymbol{\symCountVariate}\right\Vert_1 = \symNumReg$}
+\comm{.45}{$\sum_{\symRegVal=0}^{\symRegRange+1}\symCountVariate_\symRegVal = \symNumReg$}
 \State $\symZ\gets\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$
@@
 \Function {EstimateCardinality}{$\boldsymbol{\symCountVariate}$}
-\comm{.35}{$\dim \boldsymbol{\symCountVariate} = \symRegRange + 2$, $\left\Vert\boldsymbol{\symCountVariate}\right\Vert_1 = \symNumReg$}
+\comm{.31}{$\sum_{\symRegVal=0}^{\symRegRange+1}\symCountVariate_\symRegVal = \symNumReg$}
 \If{$\symCountVariate_{\symRegRange+1} = \symNumReg$}
@@
 \State $\symZ \gets \symZ\cdot 2^{-\symRegValVariate'_\text{min}}$
-\comm{0.35}{here $\symZ=\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$}
+\comm{0.31}{here $\symZ=\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$}
 \State $\symCount \gets \symCountVariate_{\symRegRange+1}$
@@
 \State $\symA\gets \symZ + \symCountVariate_0$
-\comm{0.35}{$\symA = \sum_{\symRegVal=0}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$}
+\comm{0.31}{$\symA = \sum_{\symRegVal=0}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$}
 \State $\symB\gets \symZ + 
 \symCountVariate_{\symRegRange+1}\cdot 2^{-\symRegRange}$
-\comm{0.35}{$\symB = \sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}$}
+\comm{0.31}{$\symB = \sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}$}
 \State $\symNumReg' \gets \symNumReg - \symCountVariate_0$
 \If{$\symB \leq 1.5\cdot\symA$}
-\State $\symX \gets \symNumReg'/(0.5\cdot \symB+\symA)$ \comm{0.35}{weak lower bound \eqref{equ:weak_lower_bound}}
+\State $\symX \gets \symNumReg'/(0.5\cdot \symB+\symA)$ \comm{0.31}{weak lower bound \eqref{equ:weak_lower_bound}}
 \Else
 \State $\symX \gets \symNumReg'/\symB \cdot \log(1+\symB/\symA)$ 
-\comm{0.35}{strong lower bound \eqref{equ:strong_lower_bound}}
+\comm{0.31}{strong lower bound \eqref{equ:strong_lower_bound}}
 \EndIf
@@
 \State $\symStopDelta\gets\symStopEpsilon/\sqrt{\symNumReg}$
-\comm{0.35}{$\symStopEpsilon = 10^{-2}$ (see \cref{sec:comp_ml_estimate})}
+\comm{0.31}{$\symStopEpsilon = 10^{-2}$ (see \cref{sec:comp_ml_estimate})}
 \While{$\Delta\symX > \symX\cdot\symStopDelta$} 
-\comm{0.35}{secant method iteration}
+\comm{0.31}{secant method iteration}
 \State $\symKappa \gets 2+\lfloor\log_2(\symX)\rfloor$
@@
 \symX \cdot2^{-\max(\symRegValVariate'_\text{max}, \symKappa)-1}$
-\comm{0.35}{$\symXNormalized \in [0, 0.25]$}
+\comm{0.31}{$\symXNormalized \in [0, 0.25]$}
 \State $\symX''\gets \symXNormalized\cdot\symXNormalized$
@@
 \symX' - \symX''/3 + \left(\symX''\cdot \symX''\right)\cdot\left(1/45-\symX''/472.5\right)$
-\comm{0.35}{Taylor approximation \eqref{equ:taylor}}
+\comm{0.31}{Taylor approximation \eqref{equ:taylor}}
 \For{$\symRegVal\gets (\symKappa - 1),\symRegValVariate'_\text{max}$}
 \State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
-\comm{0.35}{calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$}
+\comm{0.31}{calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$}
 \State $\symXNormalized \gets 2\symXNormalized$
@@
 \State $\symFuncPrime \gets \symCount\cdot\symHelper$
-\comm{0.35}{compare \eqref{equ:funcprime}}
+\comm{0.31}{compare \eqref{equ:funcprime}}
 \For{$\symRegVal\gets(\symRegValVariate'_\text{max}-1),\symRegValVariate'_\text{min}$}
 \State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
-\comm{0.35}{calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$}
+\comm{0.31}{calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$}
 \State $\symFuncPrime\gets \symFuncPrime + \symCountVariate_\symRegVal\cdot\symHelper$
@@
 \State $\Delta\symX \gets \Delta\symX \cdot \frac{\symNumReg' - \symFuncPrime}{\symFuncPrime - \symFuncPrime_\text{prev}}$
-\comm{0.35}{see \eqref{equ:secant_delta}}
+\comm{0.31}{see \eqref{equ:secant_delta}}
 \Else
@@
 
-While the union of two sets that are represented by HyperLogLog sketches can be straightforwardly computed using \eqref{equ:hll_union}, the computation of cardinalities of other set operations like intersections and complements is more challenging.
+While the union of two sets that are represented by HyperLogLog sketches can be straightforwardly computed using \cref{alg:union}, the computation of cardinalities of other set operations like intersections and complements is more challenging.
 The conventional approach uses the inclusion-exclusion principle
@@
 \end{equation}
-and the fact that HyperLogLog sketches can be easily merged using \eqref{equ:hll_union}.
+and the fact that HyperLogLog sketches can be easily merged using \cref{alg:union}.
 Unfortunately, the estimation error does not scale well for this approach.
@@
 
-The HyperLogLog sketch that represents $\symSetS_1$ can be thought to be constructed from two HyperLogLog sketches representing $\symSetA$ and $\symSetX$ and merging both using \eqref{equ:hll_union}.
+The HyperLogLog sketch that represents $\symSetS_1$ can be thought to be constructed from two HyperLogLog sketches representing $\symSetA$ and $\symSetX$ and merging both using \cref{alg:union}.
 Analogously, the HyperLogLog sketch for $\symSetS_2$ could have been obtained from sketches for $\symSetB$ and $\symSetX$.
@@
 If $\symX_{\ast\symRegVal}$ is smaller than $\log(2)$, we calculate $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$ using $\symZ_{\ast\symRegVal} =
--\operatorname{expm1}(-\symX_{\ast\symRegVal})$ and $\symY_{\ast\symRegVal} = 1 - \symZ_{\ast\symRegVal}$, repsectively, and not as defined in \eqref{equ:definition_xyz}.
+-\operatorname{expm1}(-\symX_{\ast\symRegVal})$ and $\symY_{\ast\symRegVal} = 1 - \symZ_{\ast\symRegVal}$, respectively, and not as defined in \eqref{equ:definition_xyz}.
 In this way the numerical error of both, $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$, is minimized and still only a single exponential function needs to be evaluated.
 Apart from that, the numerical evaluation of \eqref{equ:max_likelihood_phi_func} is straightforward.
-The arguments of all logarithms are in range $[0,1]$.
+Apparently, the arguments of all logarithms are in range $[0,1]$.
 The case that some argument vanishes, which would cause the logarithm to be negative infinite, does not occur in practice.
@@
 \gets
-\symCountVariate^{=}_{1\symRegValVariate_{1\symIndexI}}+1$
+\symCountVariate^{=}_{\symRegValVariate_{1\symIndexI}}+1$
 \EndIf
@@
 We also investigated, if the new approach could give better estimates for the union operation.
-The conventional approach merges both HyperLogLog sketches using \eqref{equ:hll_union} and estimates the union size using single sketch cardinality estimation.
+The conventional approach merges both HyperLogLog sketches using \cref{alg:union} and estimates the union size using single sketch cardinality estimation.
 The joint cardinality estimation algorithm provides another opportunity.

### 2017-01-22 paper/paper.tex
minor change

@@
 First, the insertion order of elements has no influence on the final HyperLogLog sketch.
-Furthermore, any two HyperLogLog sketches with same parameters $(\symPrecision, \symRegRange)$ representing sets $\symSetA$ and $\symSetB$ can be easily merged.
-The HyperLogLog sketch that represents the union of both sets is simply constructed by taking the register-wise maximum values as demonstrated by \cref{alg:union}.
+Furthermore, any two HyperLogLog sketches with same parameters $(\symPrecision, \symRegRange)$ representing two different sets can be easily merged.
+The HyperLogLog sketch that represents the union of both sets can be easily constructed by taking the register-wise maximum values as demonstrated by \cref{alg:union}.
 

### 2017-01-25 c++/cardinality_estimation.hpp
better algorithm for calculation of function tau with faster convergence

@@ -204,15 +204,15 @@ class ImprovedRawEstimator {
         if (x == 0. || x == 1.) return 0.;
         double zPrime;
         double y = 1.0;
-        double z = 0;
+        double z = 1 - x;
         do {
             numIterations += 1;
             x = std::sqrt(x);
             zPrime = z;
             y *= 0.5;
-            z += (1 - x)*x*y;
+            z -= std::pow(1 - x, 2)*y;
         } while(zPrime != z);
-        return z;
+        return z / 3;
     }
 
     static std::vector<double> initSigma(int m) {

### 2017-02-02 paper/paper.tex
better algorithm for calculation of function tau with faster convergence
updated paper, some major changes

@@
 The maximum likelihood approach can also be applied to more than a single HyperLogLog sketch.
-In particular, it is shown that it gives more accurate cardinality estimates for the union, intersection, or relative complements of two sets that are both represented by HyperLogLog sketches compared to the conventional technique using the inclusion-exclusion principle.
+In particular, it is shown that it gives more precise cardinality estimates for union, intersection, or relative complements of two sets that are both represented by HyperLogLog sketches compared to the conventional technique using the inclusion-exclusion principle.
 All the new methods are demonstrated and verified by extensive simulations.
@@
 Moreover, the estimator cannot be further used after merging two sketches, which limits its application to single data streams.
-If this restriction is okay, the self-learning bitmap \cite{Chen2011}
-could also be used, which provides a similar trade-off and is also 
-more memory-efficient than the original HyperLogLog method.
-
-Sometimes not only the number of distinct elements but also a sample of them is needed, in order to allow later filtering according to some predicate and estimating the cardinalities of corresponding subsets.
-In this case the k-minimum values algorithm \cite{Beyer2007} is the method of choice which also allows set manipulations like construction of intersections, disjoint sets, or unions \cite{Dasgupta2015}.
+If this restriction is acceptable, the self-learning bitmap \cite{Chen2011}
+could also be used, which provides a similar trade-off and also needs less space than the original HyperLogLog method to achieve the same precision.
+
+Sometimes not only the number of distinct elements but also a sample of them is needed in order to allow later filtering according to some predicate and estimating the cardinalities of corresponding subsets.
+In this case the k-minimum values algorithm \cite{Beyer2007} is the method of choice which also allows set manipulations like the construction of intersections, disjoint sets, or unions \cite{Dasgupta2015}.
 The latter operation is the only one that is natively supported by HyperLogLog sketches.
@@
 As presented in \cref{sec:cardinality_estimation_set_intersections} the same approach can be generalized to two HyperLogLog sketches, which allows result cardinality estimation of set operations like intersections or complements.
-The simulation results show that the estimates are significantly better than those obtained by the inclusion-exclusion principle.
-Finally, in \cref{sec:future_work} we discuss open problems and ideas for future work including HyperLogLog's potential use as locality-sensitve hashing algorithm before we conclude in \cref{sec:conclusion}.
+The simulation results show that the estimates are significantly better than those obtained using the inclusion-exclusion principle.
+Finally, in \cref{sec:future_work} we discuss open problems and ideas for future work including HyperLogLog's potential application as locality-sensitve hashing algorithm before we conclude in \cref{sec:conclusion}.
 
@@
 For performance reasons the number of registers $\symNumReg$ is chosen to be a power of 2, $\symNumReg = 2^\symPrecision$.
-$\symPrecision$ is the precision parameter that directly controls the relative error which scales like $1/\sqrt{\symNumReg}$.
+$\symPrecision$ is the precision parameter that directly controls the relative error which scales like $1/\sqrt{\symNumReg}$ \cite{Flajolet2007}.
 All registers start with zero initial value.
@@
 \caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers.
-All registers $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ start from zero.}
+All registers $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ start from zero.
+$\langle\ldots\rangle_2$ denotes the binary representation of an integer.}
 \label{alg:insert}
@@
 \State $\symIndexI \gets 1+ \langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision\rangle_2$
-\comm{0.25}{$\symIndexI\in\lbrace 1,2,\ldots,2^\symPrecision\rbrace$}
+\comm{0.25}{$\symIndexI\in\lbrace 1,2,\ldots,\symNumReg\rbrace$}
 \If{$\symRegVal>\symRegValVariate_\symIndexI$}
@@
 )$ and $\symRegRange=\dim\boldsymbol{\symCountVariate}-2$, respectively.
-As \cref{alg:sufficient_statistic} shows the calculation of the multiplicity vector $\boldsymbol{\symCountVariate}$ for a given HyperLogLog sketch with register values $\boldsymbol{\symRegValVariate}$ is straightforward.
+\cref{alg:sufficient_statistic} shows the calculation of the multiplicity vector $\boldsymbol{\symCountVariate}$ for a given HyperLogLog sketch with register values $\boldsymbol{\symRegValVariate}$.
 
@@
 \label{sec:poisson_approximation}
-Due to the statistical dependence of the register values, the probability mass function \eqref{equ:multinomialProbabilityMass} is difficult for further analysis.
+Due to the statistical dependence of the register values, the probability mass function \eqref{equ:multinomialProbabilityMass} makes further analysis difficult.
 Therefore, a Poisson model can be used \cite{Flajolet2007}, which assumes that the cardinality itself is distributed according to a Poisson distribution
@@
 The final factorization shows that under the Poisson model the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ are independent and identically distributed.
-The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is defined by
+The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is given by
 \begin{equation}
@@
 \label{sec:depoissonization}
-Due to the simpler probability mass function, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\boldsymbol{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ rather than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}.
+Due to the simpler probability mass function, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\boldsymbol{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ than for the cardinality $\symCardinality$ under the fixed-size model \eqref{equ:multinomialProbabilityMass}.
 Depoissonization \cite{Jacquet1998} finally allows to translate the estimates back to the fixed-size model.
@@
 The chart is based on \num{10000} randomly generated HyperLogLog sketches.
-More details of the experimental setup will be explained later in \cref{sec:corrected_raw_estimation_error}.
+More details of the experimental setup will be explained later in \cref{sec:experimental_setup}.
 Obviously, the raw estimator is biased for small and large cardinalities where it fails to return accurate estimates.
-In order to cover the entire range of cardinalities, corrections for small and large cardinalities have been proposed.
+In order to cover the entire cardinality range, corrections for small and large cardinalities have been proposed.
 
@@
 \begin{equation}
+\label{equ:tau}
 \symSmallCorrectionFunc(\symX) := 
@@
 and
-\begin{equation}
+%\begin{equation}
+%\symLargeCorrectionFunc(\symX)
+%:=
+%\sum_{\symRegVal=1}^\infty
+%\symX^{2^{-\symRegVal}}
+%\left(1-\symX^{2^{-\symRegVal}}\right)
+%2^{-\symRegVal}.
+%\end{equation}
+\begin{equation}
+\label{equ:sigma}
 \symLargeCorrectionFunc(\symX)
 :=
+\frac{1}{2}
+\left(
+-
+\symX
++
 \sum_{\symRegVal=1}^\infty
 \symX^{2^{-\symRegVal}}
-\left(1-\symX^{2^{-\symRegVal}}\right)
-2^{-\symRegVal}.
-\end{equation}
-
-We can cross-check the new estimator for the linear counting case with $\symRegRange=0$.
-Using the identity $\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = \symAlpha_\infty\symPowerSeriesFunc(\log_2(\log(1/\symX)))/\log(1/\symX)$, we get
+2^{-\symRegVal}
+\right)
+.
+\end{equation}
+%=
+%\frac{1}{3}
+%\left(
+%1
+%-
+%\symX
+%-
+%\sum_{\symRegVal=1}^\infty
+%\left(
+%1-
+%\symX^{2^{-\symRegVal}}
+%\right)^2
+%2^{-\symRegVal}
+%\right)
+
+We can cross-check the new estimator for the linear counting case with $\symRegRange=0$. Using the identity 
+\begin{equation}
+\label{equ:sigma_tau_relationship}
+\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = \symAlpha_\infty\symPowerSeriesFunc(\log_2(\log(1/\symX)))/\log(1/\symX),
+\end{equation}
+ we get
 \begin{equation}
@@
 \subsection{Improved raw estimation algorithm}
-The improved raw estimator \eqref{equ:correctedestimator} leads to a new cardinality estimation algorithm for HyperLogLog sketches.
-\cref{alg:corrected_raw_estimation} demonstrates the numerical evaluation of the estimator.
-The values of functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ can be either calculated on-demand or pre-calculated.
-Possible values of $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$ are $\{0,1,\ldots, \symNumReg\}$.
-Therefore, if performance matters, the function values can be precalculated for all possible arguments and kept in lookup tables of size $\symNumReg+1$.
+The improved raw estimator \eqref{equ:correctedestimator} can be directly translated into a new cardinality estimation algorithm for HyperLogLog sketches as shown in \cref{alg:corrected_raw_estimation}.
+Since the series \eqref{equ:sigma} converges quadratically for all $\symX\in[0,1)$ and its terms can be recursively calculated using elementary operations, the function $\symSmallCorrectionFunc$ can be quickly calculated to machine precision.
+The case $\symX=1$ must be handled separately, because the series diverges and causes an infinite denominator in \eqref{equ:correctedestimator} and therefore a vanishing cardinality estimate.
+As this case only occurs if all register values are zero ($\symCountVariate_0=\symNumReg$), this is exactly what is expected.
+Among the remaining possible arguments $\lbrace0,\frac{1}{\symNumReg},\frac{2}{\symNumReg},\ldots,\frac{\symNumReg-1}{\symNumReg}\rbrace$ we have slowest convergence for $\symX=\frac{\symNumReg-1}{\symNumReg}$.
+However, even in this case we have a manageable amount of iteration cycles.
+For example, if double-precision floating-point arithmetic is used, the routine for calculating  $\symSmallCorrectionFunc$ converges after \num{18} and \num{26} iteration cycles for $\symPrecision=12$ and $\symPrecision=20$, respectively.
+
+%
+%The case $\symX=1$, for which the series does not converge, occurs only if all register values are zero ($\symCountVariate_0=\symNumReg$).
+However, the resulting positive infinite denominator in \eqref{equ:correctedestimator} yields zero for the cardinality estimate as expected.
+In order to estimate the number of iteration cycles needed to determine $\symSmallCorrectionFunc$ up to some precision $\symStopDelta$, we consider the case $\symX = 1-\frac{1}{\symNumReg}$ for which we have slowest convergence speed among all relevant argument values $\lbrace0,\frac{1}{\symNumReg},\frac{2}{\symNumReg},\ldots,1-\frac{1}{\symNumReg}\rbrace$.
+We can stop summation, if two subsequent terms differ by a factor more than $\symStopDelta$.
+This is the case, if $\frac{\symX^{2^{\symRegVal+1}} 2^{\symRegVal}}{\symX^{2^\symRegVal} 2^{\symRegVal-1}} < \symStopDelta$ which is equivalent to $
+%\symRegVal > \frac{
+%\log(\log(2/\symStopDelta))
+%-
+%\log(-\log(1-\frac{1}{\symNumReg}))
+%}
+%{
+%\log(2)
+%}
+%$. Due to the inequality $-\log(1-\frac{1}{\symNumReg})\geq \frac{1}{\symNumReg}$ this is also satisfied, if $
+%\symRegVal > \frac{
+%\log(\log(2/\symStopDelta))
+%}
+%{
+%\log(2)
+%}
+%+
+%\symPrecision
+%$
+%holds. Therefore, not more than $\symPrecision + 5$ iterations are necessary to achieve double-precision $\symStopDelta = 2^{-53}$.
+
+The calculation of $\symLargeCorrectionFunc$ is more expensive, because it involves square root evaluations. $\symLargeCorrectionFunc$ can also be written in a numerically more favorable way as 
+\begin{equation}
+\label{equ:tau2}
+\symLargeCorrectionFunc(\symX)
+:=
+\frac{1}{3}
+\left(
+1-\symX
+-
+\sum_{\symRegVal=1}^\infty
+\left(
+1-
+\symX^{2^{-\symRegVal}}
+\right)^{\!2}
+2^{-\symRegVal}
+\right).
+\end{equation}
+This representation shows faster convergence for $\symX>0$, because $1-\symX^{2^{-\symRegVal}}\leq -\log(\symX) 2^{-\symRegVal}$.
+Therefore, the convergence speed is comparable to a geometric series with ratio $1/8$.
+Disregarding the trivial case $\symLargeCorrectionFunc(0)=0$, $\symX=\frac{1}{\symNumReg}$ shows slowest convergence speed among all other possible parameters.
+For this case and if double-precision floating-point numbers are used, the function $\symLargeCorrectionFunc$ presented in \cref{alg:corrected_raw_estimation} converges after \num{21} and \num{22} iteration cycles for considered precisions $\symPrecision=12$ and $\symPrecision=20$, respectively.
+%
+%Since $\symLargeCorrectionFunc(0)$ and $\symLargeCorrectionFunc(1)$ are both equal to zero, we focus on the remaining relevant arguments $\symX\in[\frac{1}{\symNumReg},1-\frac{1}{\symNumReg}]$.
+Obviously, we have slowest convergence for $\symX = \frac{1}{\symNumReg}$.
+Since the left factor in \eqref{equ:tau2} is monotone increasing we can find
+%\begin{equation}
+%1-
+%\left(1-\frac{1}{\symNumReg}\right)
+%\sum_{\symRegVal=1}^\infty
+%\prod_{\symIndexI=1}^\symRegVal
+%\frac{1}
+%{
+%2
+%\left(
+%1+
+%{\frac{1}{\symNumReg}}^{2^{-\symRegVal}}
+%\right)^2
+%}
+%\geq
+%1-
+%\frac{1-\frac{1}{\symNumReg}}
+%{
+%\left(
+%1+
+%{\frac{1}{\sqrt{\symNumReg}}}
+%\right)^2
+%}
+%\sum_{\symRegVal=1}^\infty
+%\frac{1}
+%{
+%2^\symRegVal
+%}
+%=
+%\frac{2}{\sqrt{\symNumReg}+1}
+%\end{equation}
+%as lower bound for $\symX\geq\frac{1}{\symNumReg}$. 
+%
+%
+%
+% for which the smallest value is obtained for $\symX = 1-\frac{1}{\symNumReg}$. 
+%It can be shown that $\symLargeCorrectionFunc(1-\frac{1}{\symNumReg})\geq \frac{1}{4\symNumReg}$. Hence, summation in \eqref{equ:sigma} can be stopped, if $\left(
+%1-
+%\symX^{2^{-\symRegVal}}
+%\right)^2
+%2^{-\symRegVal}
+%< \frac{\symStopDelta}{4\symNumReg}
+%$. Since we have slowest convergence for $\symX=\frac{1}{\symNumReg}$, $\left(
+%1-
+%\symNumReg^{-2^{-\symRegVal}}
+%\right)^2
+%2^{-\symRegVal}
+%< \frac{\symStopDelta}{4\symNumReg}
+%$ is also a sufficient condition to achieve the desired precision $\symStopDelta$. Furthermore, $1-
+%\symNumReg^{-2^{-\symRegVal}}
+%=
+%1-e^{-\log(\symNumReg)2^{-\symRegVal}}
+%\leq
+%\log(\symNumReg)2^{-\symRegVal}
+%$ which allows to write 
+%$(\log(\symNumReg))^2 8^{-\symRegVal}
+%< \frac{\symStopDelta}{4\symNumReg}
+%$ and finally $\symRegVal > \frac{\symPrecision}{3} + \frac{2\log(\symPrecision)}{\log(8)} + 
+%\frac{\log(\log(16)/\symStopDelta)}{\log(8)}$.
+%
+%
+%
+%
+
+If performance matters $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$
+can also be precalculated for all possible values of $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$.
+As their value range is $\{0,1,\ldots, \symNumReg\}$, the function values can be kept in lookup tables of size $\symNumReg+1$.
+In this way a complete branch-free cardinality estimation can be realized.
+It is also thinkable to calculate $\symLargeCorrectionFunc$ using the approximation $\symLargeCorrectionFunc(\symX)\approx -\symAlpha_\infty/\log(\symX)-\symSmallCorrectionFunc(\symX)$ which can be obtained from \eqref{equ:sigma_tau_relationship}.
+The advantage is that the calculation of $\symSmallCorrectionFunc$ and the additional logarithm is slightly faster than the calculation of $\symSmallCorrectionFunc(\symX)$.
+However, for arguments close to 1, where $1/\log(\symX)$ and $\symSmallCorrectionFunc(\symX)$ are both very large while their difference is very small, special care is needed to avoid numerical cancellation.
 
@@
 \State $\symY\gets1$
-\State $\symZ\gets0$
+\State $\symZ\gets 1 - \symX$
 \Repeat
@@
 \State $\symY\gets0.5\cdot\symY$
-\State $\symZ\gets\symZ + \left(1-\symX\right)\cdot\symX\cdot\symY$
+\State $\symZ\gets\symZ - \left(1-\symX\right)^2\cdot\symY$
 \Until{$\symZ=\symZ'$}
-\State\Return$\symZ$
+\State\Return$\symZ/3$
 \EndFunction
@@
 
-\subsection{Estimation error}
-\label{sec:corrected_raw_estimation_error}
+\subsection{Experimental setup}
+\label{sec:experimental_setup}
 In order to verify the new estimation algorithm, we generated \num{10000} different HyperLogLog sketches and filled each of them with 50 billion unique elements.
@@
 If $\symRegValVariate_\text{min}=2$ it is already just a quarter, and so forth.
-Finally, instead of repeating the simulation for different HyperLogLog parametes, we did that only once using the quite accurate setting $\symPrecision=22$ and $\symRegRange=42$.
+Finally, instead of repeating the simulation for different HyperLogLog parameters, we did that only once using the quite accurate setting $\symPrecision=22$ and $\symRegRange=42$.
 Whenever taking a snapshot, we reduced the sketch state exploiting \cref{alg:compress} for desired parameters before calculating the multiplicity vector using \cref{alg:sufficient_statistic}.
 
-After all, we got the multiplicity vectors at predefined cardinalities of \num{10000} different simulation runs for various HyperLogLog parameters.
+After all, we got the multiplicity vectors at predefined cardinalities of \num{10000} different simulation runs for various HyperLogLog parameter settings.
 All this data constitutes the basis of the presented results that follow.
@@
 
-\cref{fig:raw_corrected_estimation_error_12_20} shows the distribution of the relative error of the estimated cardinality using \cref{alg:corrected_raw_estimation} compared to the true cardinality for $\symPrecision=12$ and $\symRegRange=20$.
-As the mean shows, the error is unbiased over the entire cardinality range.
-The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is about an order of magnitude larger than the operating range upper bound of the raw estimator (\cref{fig:raw_estimate}).
-
 \begin{algorithm}
 \caption{Extension of \cref{alg:insert} that keeps track of the multiplicity vector $\boldsymbol{\symCountVariate} = (\symCountVariate_0,\ldots,\symCountVariate_{\symRegRange+1})$ and the minimum register value $\symRegValVariate_\text{min}$ during insertion.
-Initially, $\symRegValVariate_\text{min}=0$ and $\boldsymbol{\symCountVariate} = (2^\symPrecision, 0, 0, \dots, 0)$.}
+Initially, $\symRegValVariate_\text{min}=0$ and $\boldsymbol{\symCountVariate} = (\symNumReg, 0, 0, \dots, 0)$.}
 \label{alg:insert_fast}
@@
 
+\subsection{Estimation error}
+\label{sec:corrected_raw_estimation_error}
+\cref{fig:raw_corrected_estimation_error_12_20} shows the distribution of the relative error of the estimated cardinality using \cref{alg:corrected_raw_estimation} compared to the true cardinality for $\symPrecision=12$ and $\symRegRange=20$.
+As the mean shows, the error is unbiased over the entire cardinality range.
+The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is about an order of magnitude larger than the operating range upper bound of the raw estimator (\cref{fig:raw_estimate}).
+
 \begin{figure}
@@
 The improved raw estimator beats the precision of methods that apply bias correction on the raw estimator \cite{Heule2013,Rhodes2015,Sanfilippo2014}.
-Based on the simulated data we have empirically determined the bias correction function $\symBiasCorrectionFunc$ for the raw estimator that satisfies $\symCardinality = \symExpectation(\symBiasCorrectionFunc(\symCardinalityRawEstimate)\vert\symCardinality)$ for all cardinalities.
+Based on the simulated data we have empirically determined the bias correction function $\symBiasCorrectionFunc$ for the raw estimator \eqref{equ:raw_estimator} that satisfies $\symCardinality = \symExpectation(\symBiasCorrectionFunc(\symCardinalityRawEstimate)\vert\symCardinality)$ for all cardinalities.
 By definition,
@@
 Again, the new algorithm gives essentially unbiased estimates.
-Only for very large precisions, an oscillating bias becomes apparent (compare \cref{fig:raw_corrected_estimation_error_22_10}), that is caused by approximating the periodic function $\symPowerSeriesFunc$ by a constant (see \cref{sec:derivation_raw_estimator}).
+Only for very high precisions, an oscillating bias becomes apparent (compare \cref{fig:raw_corrected_estimation_error_22_10}), that is caused by approximating the periodic function $\symPowerSeriesFunc$ by a constant (see \cref{sec:derivation_raw_estimator}).
 
@@
 \label{sec:corrected_raw_estimation_algorithm}
-In order to evaluate the performance of the improved raw estimation algorithm, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch.
-For different cardinalities we took \num{1000} randomly generated HyperLogLog sketches, calculated their
-multiplicity vectors using \cref{alg:sufficient_statistic} and kept all of them in main memory.
+To evaluate the performance of the improved raw estimation algorithm, we investigated the average computation time to obtain the cardinality from a given multiplicity vector.
+For different cardinality values we loaded the precalculated multiplicity vectors (see \cref{sec:experimental_setup}) of \num{1000} randomly generated HyperLogLog sketches into main memory.
 The average computation time was determined by cycling over these multiplicity vectors and passing them as input to the algorithm.
@@
 The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:corrected_raw_avg_exec_time}.
-Two variants of \cref{alg:corrected_raw_estimation} have been evaluated for which the functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ have been either calculated on-demand or taken from a lookup table.
+Two variants of \cref{alg:corrected_raw_estimation} have been evaluated for which the functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ have been either calculated on demand or taken from a lookup table.
 All these benchmarks where carried out on an Intel Core i5-2500K clocking at \SI{3.3}{\giga\hertz}.
@@
 The results show that the execution times are nearly constant for $\symRegRange=52$.
-Using a lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed due to the small probability of saturated registers, $\symProbability(\symCountVariate_{53}>0)\approx 0$, for realistic cardinalities.
+Using a lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed due to the small probability of saturated registers, $\symCountVariate_{53}=0$ usually holds for realistic cardinalities.
 
@@
 The faster computation times compared to the $\symRegRange=52$ case can be explained by the much smaller dimension of the multiplicity vector which is equal to $\symRegRange+2$.
-If the functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are calculated on-demand, the execution times for larger cardinalities are doubled.
+If the functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are calculated on demand, the execution times for larger cardinalities are doubled.
 This comes from the fact that the calculation of $\symLargeCorrectionFunc$ is much more expensive than that of $\symSmallCorrectionFunc$, because it requires more iterations and involves square root evaluations.
-However, we can imagine that a better numerical approximation of $\symLargeCorrectionFunc$ can be found than that given in \cref{alg:corrected_raw_estimation} and which allows on-demand evaluation without significant extra costs.
 If $2^{\symPrecision+\symRegRange}$ is greater than the maximum expected cardinality value, the number of saturated registers is negligible ($\symCountVariate_{\symRegRange+1}\ll \symNumReg$) and the calculation of $\symLargeCorrectionFunc$ can be omitted at all, because $\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg) \approx 0$ vanishes in this case.
@@
 \caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at \SI{3.3}{\giga\hertz} when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.
-Both cases, $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ precalculated and calculated on-demand have been considered.}
+Both cases, $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ precalculated and calculated on demand have been considered.}
 \label{fig:corrected_raw_avg_exec_time}
@@
 allows factorization of the joint probability mass function and a straightforward calculation of the maximum likelihood estimate.
-The practically more relevant stochastic averaging case, which is considered in this paper, gives a more complicated likelihood function.
+The practically more relevant stochastic averaging case, which is considered in this paper, would lead to a more complicated likelihood function.
 However, the Poisson approximation makes the maximum likelihood method feasible again and we are finally able to derive another new robust and efficient cardinality estimation algorithm.
@@
 \subsection{Inequalities for the maximum likelihood estimate}
-In the following lower and upper bounds for $\symXEstimate$ are derived.
+In the following we derive lower and upper bounds for $\symXEstimate$.
 Applying Jensen's inequality on $\symHelper$ in \eqref{equ:func} gives an upper bound for $\symFunc(\symX)$:
@@
 The algorithm requires mainly only elementary operations.
-The square root that appears for the calculation of the stopping limit $\symStopDelta=\symStopEpsilon/\sqrt{\symNumReg}$ can be precomputed, because it only depends on the HyperLogLog precision parameter $\symPrecision$.
 For very large cardinalities it makes sense to use the strong \eqref{equ:strong_lower_bound} instead of the weak lower bound \eqref{equ:weak_lower_bound} as second starting point for the secant method.
@@
 \State $\Delta\symX \gets \symX$
-\State $\symStopDelta\gets\symStopEpsilon/\sqrt{\symNumReg}$
-\comm{0.31}{$\symStopEpsilon = 10^{-2}$ (see \cref{sec:comp_ml_estimate})}
 \While{$\Delta\symX > \symX\cdot\symStopDelta$} 
-\comm{0.31}{secant method iteration}
+\comm{0.31}{secant method iteration\\
+$\symStopDelta=\symStopEpsilon/\sqrt{\symNumReg}$, $\symStopEpsilon = 10^{-2}$}
 \State $\symKappa \gets 2+\lfloor\log_2(\symX)\rfloor$
@@
 \label{sec:maximum_likelihood_estimation_error}
-In order to investigate the estimation error for the maximum likelihood estimation algorithm, we investigated the same HyperLogLog configurations as in \cref{sec:corrected_raw_estimation_error} for the improved raw estimation algorithm.
-\cref{fig:max_likelihood_estimation_error_12_20}, \cref{fig:max_likelihood_estimation_error_8_24},  \cref{fig:max_likelihood_estimation_error_16_16}, \cref{fig:max_likelihood_estimation_error_22_10}, \cref{fig:max_likelihood_estimation_error_12_52}, and \cref{fig:max_likelihood_estimation_error_12_14} show very similar results for various HyperLogLog parameters.
-
+We have investigated the estimation error of the maximum likelihood estimation algorithm for the same HyperLogLog configurations as for the improved raw estimation algorithm in \cref{sec:corrected_raw_estimation_error}.
+\cref{fig:max_likelihood_estimation_error_12_20}, \cref{fig:max_likelihood_estimation_error_8_24},  \cref{fig:max_likelihood_estimation_error_16_16}, \cref{fig:max_likelihood_estimation_error_22_10}, \cref{fig:max_likelihood_estimation_error_12_52}, and \cref{fig:max_likelihood_estimation_error_12_14} show very similar results for same HyperLogLog parameters.
 What is different for the maximum likelihood estimation approach is a somewhat smaller median bias with less oscillations around zero for small cardinalities.
@@
 The average computation time for the maximum likelihood algorithm shows a different behavior than for the improved raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}).
-As can be seen, the average execution time is larger for most cardinalities, but nevertheless, it never exceeds \SI{700}{\nano\second} which is still fast enough for many applications.
+The average execution time is larger for most cardinalities, but nevertheless, it never exceeds \SI{700}{\nano\second} which is still fast enough for many applications.
 The steps in the chart can be explained by different numbers of iteration cycles until the secant method is stopped.
@@
 For HyperLogLog sketches recorded without stochastic averaging (compare \cref{sec:data_element_insertion}) this was shown in \cite{Cohen2016}, where also an algorithm based on the maximum likelihood principle was outlined.
-However, in practice, HyperLogLog sketches are recorded using stochastic averaging because of the much cheaper element insertions as described by \cref{alg:insert}.
+However, in practice, HyperLogLog sketches are recorded using stochastic averaging because of the much cheaper element insertions.
 Motivated by the good results we have obtained for a single HyperLogLog sketch using the maximum likelihood method in combination with the Poisson approximation, we are tempted to apply this approach also for the estimation of set operation result sizes.
@@
 
-In order to get the maximum likelihood estimates $\symPoissonRateEstimate_\symSetASuffix$,
- $\symPoissonRateEstimate_\symSetBSuffix$, and  $\symPoissonRateEstimate_\symSetXSuffix$ we need to maximize the log-likelihood function given by
+The maximum likelihood estimates $\symPoissonRateEstimate_\symSetASuffix$,
+ $\symPoissonRateEstimate_\symSetBSuffix$, and  $\symPoissonRateEstimate_\symSetXSuffix$ are obtained by maximization of the log-likelihood function given by
 \begin{equation}
@@
 The derived lower bound shows that the argument of the logarithm is large enough to be accurately represented by double-precision floating-point numbers, provided that $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetXSuffix$ are not both substantially smaller than 1.
-Similar argumentation holds for all other logarithmic terms and also the divisions that appear in \eqref{equ:df_dphi}.
+Similar argumentation holds for all other logarithmic terms and also all divisions that appear in \eqref{equ:df_dphi}.
 
@@
 Within the optimization loop the function $\symFunc$ and its gradient $\nabla\symFunc$ need to evaluated using \eqref{equ:max_likelihood_phi_func} and \eqref{equ:df_dphi} as needed by the optimization algorithm.
-The optimization loop is continued as long as the absolute change of at least one parameter is larger than a predefined threshold.
+The optimization loop is continued as long as the absolute change of at least one parameter is larger than a predefined threshold $\symStopDelta$.
 The threshold is again set proportional to $1/\sqrt{\symNumReg}$ (compare \cref{sec:comp_ml_estimate}).
+For the results presented below we have used $\symStopDelta=\symStopEpsilon/\sqrt{\symNumReg}$ with $\symStopEpsilon = 10^{-2}$.
 
@@
 \State $\symPoissonRateEstimate_{\symSetASuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{<}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{1}$}
-\comm{0.32}{use \cref{alg:corrected_raw_estimation} or \ref{alg:estimate_ml}}
+\comm{0.32}{use \cref{alg:estimate_ml}}
 \State $\symPoissonRateEstimate_{\symSetBSuffix\symSetXSuffix} \gets$ \Call{EstimateCardinality}{$\boldsymbol\symCountVariate^{<}_{2} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2}$}
@@
 \symPoissonRateEstimate_{\symSetASuffix\symSetBSuffix\symSetXSuffix}))$
-\State $\symStopDelta\gets\symStopEpsilon/\sqrt{\symNumReg}$
-\comm{0.32}{$\symStopEpsilon = 10^{-2}$}
 \Repeat
@@
 $}
-\comm{0.32}{stop criterion}
+\comm{0.32}{stop criterion\\
+$\symStopDelta=\symStopEpsilon/\sqrt{\symNumReg}$, $\symStopEpsilon = 10^{-2}$}
 \State 
@@
 \subsection{Results}
-In order to investigate the estimation error of the joint cardinality estimation algorithm \cref{alg:joint_cardinality_estimation} we constructed HyperLogLog sketch pairs with parameters $\symPrecision=20$ and $\symRegRange=44$ and with known cardinalities $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$.
-This is done by adding $|\symSetA|$ and $|\symSetB|$ random values to the first and second HyperLogLog sketch, respectively.
-Afterwards
-$|\symSetX|$ random values are added to both sketches.
-\cref{tbl:joint_estimation_cases} lists all the different cases we have evaluated.
-For each case \num{3333} different HyperLogLog sketch pairs were constructed and passed to the joint cardinality estimation algorithm.
+To evaluate the estimation error of the new joint cardinality estimation approach we applied \cref{alg:joint_cardinality_estimation} to HyperLogLog sketch pairs with parameters $\symPrecision=20$ and $\symRegRange=44$ and with known cardinalities $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$.
+HyperLogLog sketch with a predefined cardinality can be quickly generated by taking a precalculated multiplicity vector with known cardinality, setting the register values accordingly, and shuffling them.
+Using this approach we generated three different HyperLogLog sketches with known cardinalities $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$.
+Then we merged the HyperLogLog sketches representing the disjoint sets $\symSetA$ and $\symSetX$ using \cref{alg:union} in order to get a sketch for $\symSetS_1 = \symSetA\cup\symSetX$.
+A second sketch was built analogously for $\symSetS_2=\symSetA\cup\symSetX$.
+Since three independent multiplicity vectors are needed to construct an example and we had simulated \num{10000} different HyperLogLog sketches previously as described in \cref{sec:experimental_setup}, we were able to construct \num{3333} different independent HyperLogLog pairs for a given set of known true cardinalities $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$.
+
+\cref{tbl:joint_estimation_cases} lists all the different cardinality configurations for which we have evaluated the joint cardinality estimation  algorithm.
 \cref{tbl:joint_estimation_cases} also shows the average number of iterations of the BFGS algorithm until the stop criterion was satisfied.
@@
 The bins correspond to HyperLogLog registers which keep the first one-bit position of the minimum value instead of the minimum value itself.
-This close connection was also pointed out in \cite{Cohen2014}, where both methods have been considered as variants of $k$-partion minwise hashing.
+This close connection was also pointed out in \cite{Cohen2014}, where both methods have been considered as variants of $k$-partition minwise hashing.
 The only difference is that the HyperLogLog algorithm uses base-2 ranks while one permutation hashing uses full ranks.

### 2017-02-02 python/convergence-speed-tau-sigma.py

@@ -1,0 +1,40 @@
+#############################
+# Copyright 2016 Otmar Ertl #
+#############################
+
+from math import exp, sqrt, log
+
+def tau(x, eps):
+    numIterations = 0
+    if x == 0. or x == 1.:
+        return (0., numIterations)
+    y = 1
+    z = 1 - x
+    while True:
+        numIterations += 1
+        x = sqrt(x)
+        zPrime = z
+        y *= 0.5
+        z -= pow(1-x,2)*y
+        if (zPrime - zPrime * eps <= z):
+            return (z, numIterations)
+
+def sigma(x, eps):
+    numIterations = 0
+    if (x == 1.):
+        return (float('inf'), numIterations)
+    y = 1
+    z = x
+    while True:
+        numIterations += 1
+        x *= x
+        zPrime = z
+        z += x * y
+        y += y
+        if (zPrime + zPrime * eps >= z):
+            return (z, numIterations)
+
+for p in range(1, 31):
+    s  = sigma(1. - pow(2.,-p), 0)
+    t  = tau(pow(2.,-p), 0)
+    print(str(p) + " " + str(s) + " " + str(t))

### 2017-02-02 python/distribution_charts.py

@@ -88,6 +88,7 @@ createFigure(8, 24,  'max_likelihood_estimates', 'max_likelihood_estimate_8_24')
 createFigure(12, 20, 'max_likelihood_estimates', 'max_likelihood_estimate_12_20')
 createFigure(16, 16, 'max_likelihood_estimates', 'max_likelihood_estimate_16_16')
 createFigure(22, 10, 'max_likelihood_estimates', 'max_likelihood_estimate_22_10')
+createFigure(22, 42, 'max_likelihood_estimates', 'max_likelihood_estimate_22_42')
 createFigure(12, 52, 'max_likelihood_estimates', 'max_likelihood_estimate_12_52')
 createFigure(12, 14, 'max_likelihood_estimates', 'max_likelihood_estimate_12_14')
 createFigure(12,  0, 'max_likelihood_estimates', 'max_likelihood_estimate_12_0')
@@ -96,6 +97,7 @@ createFigure(8, 24,  'corrected_raw_estimates', 'raw_corrected_estimate_8_24')
 createFigure(12, 20, 'corrected_raw_estimates', 'raw_corrected_estimate_12_20')
 createFigure(16, 16, 'corrected_raw_estimates', 'raw_corrected_estimate_16_16')
 createFigure(22, 10, 'corrected_raw_estimates', 'raw_corrected_estimate_22_10')
+createFigure(22, 42, 'corrected_raw_estimates', 'raw_corrected_estimate_22_42')
 createFigure(12, 52, 'corrected_raw_estimates', 'raw_corrected_estimate_12_52')
 createFigure(12, 14, 'corrected_raw_estimates', 'raw_corrected_estimate_12_14')
 createFigure(12, 0,  'corrected_raw_estimates', 'raw_corrected_estimate_12_0')

### 2017-02-03 paper/paper.tex
minor changes

@@
 Counting the number of distinct elements in a data stream or large datasets is a common problem in big data processing.
-In principle, finding the number of distinct elements $\symCardinality$ with a maximum relative error $\symError$  in a data stream requires $\Omega(\symCardinality)$ space \cite{Alon1999}.
+In principle, finding the number of distinct elements $\symCardinality$ with a maximum relative error $\symError$  in a data stream requires $\symBigO(\symCardinality)$ space \cite{Alon1999}.
 However, probabilistic algorithms that achieve the requested accuracy only with high probability are able to drastically reduce space requirements.
 Many different probabilistic algorithms have been developed over the past two decades \cite{Metwally2008,Ting2014} until a theoretically optimal algorithm was finally found \cite{Kane2010}.
-Although this algorithm achieves the optimal space complexity of $\Omega(\symError^{-2}+\log \symCardinality)$ \cite{Alon1999, Indyk2003}, it is not very efficient in practice \cite{Ting2014}.
-
-More practicable and already widely used for many applications is the HyperLogLog algorithm \cite{Flajolet2007} with a near-optimal space complexity $\Omega(\symError^{-2} \log\log\symCardinality +\log \symCardinality)$.
+Although this algorithm achieves the optimal space complexity of $\symBigO(\symError^{-2}+\log \symCardinality)$ \cite{Alon1999, Indyk2003}, it is not very efficient in practice \cite{Ting2014}.
+
+More practicable and already widely used for many applications is the HyperLogLog algorithm \cite{Flajolet2007} with a near-optimal space complexity $\symBigO(\symError^{-2} \log\log\symCardinality +\log \symCardinality)$.
 It has the nice property that partial results can be easily merged, which is a necessity for distributed environments.
@@
 \subsection{Outline}
-This paper presents new algorithms for cardinality estimation from HyperLogLog sketches.
+This paper presents new algorithms to extract cardinality information from HyperLogLog sketches.
 In \cref{sec:hyperloglog_data_structure} we start with an introduction to HyperLogLog sketches and the corresponding update algorithm.
@@
 As presented in \cref{sec:cardinality_estimation_set_intersections} the same approach can be generalized to two HyperLogLog sketches, which allows result cardinality estimation of set operations like intersections or complements.
-The simulation results show that the estimates are significantly better than those obtained using the inclusion-exclusion principle.
+The simulation results show that the estimates are significantly better than those of the conventional approach using the inclusion-exclusion principle.
 Finally, in \cref{sec:future_work} we discuss open problems and ideas for future work including HyperLogLog's potential application as locality-sensitve hashing algorithm before we conclude in \cref{sec:conclusion}.
@@
 \label{sec:hyperloglog_data_structure}
-The HyperLogLog algorithm uses a sketching data structure that consists of $\symNumReg$ registers.
-For performance reasons the number of registers $\symNumReg$ is chosen to be a power of 2, $\symNumReg = 2^\symPrecision$.
-$\symPrecision$ is the precision parameter that directly controls the relative error which scales like $1/\sqrt{\symNumReg}$ \cite{Flajolet2007}.
+The HyperLogLog algorithm collects information of incoming elements into a 
+very compact sketching data structure, that finally allows the estimation of the number of distinct elements.
+The data structure consists of $\symNumReg = 2^\symPrecision$ registers whose number is chosen to be a power of two for performance reasons.
+$\symPrecision$ is the precision parameter that directly controls the relative estimation error which scales like $1/\sqrt{\symNumReg}$ \cite{Flajolet2007}.
 All registers start with zero initial value.
@@
 \label{sec:data_element_insertion}
-In order to insert a data element into a HyperLogLog data structure a hash value is calculated.
+The insertion of a data element into a HyperLogLog data structure requires the calculation of a $(\symPrecision+\symRegRange)$-bit hash value.
 The leading $\symPrecision$ bits of the hash value are used to select one of the $2^\symPrecision$ registers.
@@
 Here $\symCount_\symRegVal$ denotes the multiplicity of value $\symRegVal$  in the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$.
-The final factorization shows that under the Poisson model the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ are independent and identically distributed.
+The final factorization shows that the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ are independent and identically distributed under the Poisson model.
 The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is given by
@@
 \label{sec:depoissonization}
-Due to the simpler probability mass function, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\boldsymbol{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ than for the cardinality $\symCardinality$ under the fixed-size model \eqref{equ:multinomialProbabilityMass}.
+Due to the simpler probability mass function \eqref{equ:poisson_pmf}, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\boldsymbol{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ than for the cardinality $\symCardinality$ under the fixed-size model \eqref{equ:multinomialProbabilityMass}.
 Depoissonization \cite{Jacquet1998} finally allows to translate the estimates back to the fixed-size model.
@@
 \symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} 
-= 0.72134752\ldots,
+= 0.72134752\ldots
 \end{equation}
@@
 \begin{equation}
-\label{equ:tau}
+\label{equ:sigma}
 \symSmallCorrectionFunc(\symX) := 
@@
 \begin{equation}
-\label{equ:sigma}
+\label{equ:tau}
 \symLargeCorrectionFunc(\symX)
@@
 In this way a complete branch-free cardinality estimation can be realized.
-It is also thinkable to calculate $\symLargeCorrectionFunc$ using the approximation $\symLargeCorrectionFunc(\symX)\approx -\symAlpha_\infty/\log(\symX)-\symSmallCorrectionFunc(\symX)$ which can be obtained from \eqref{equ:sigma_tau_relationship}.
-The advantage is that the calculation of $\symSmallCorrectionFunc$ and the additional logarithm is slightly faster than the calculation of $\symSmallCorrectionFunc(\symX)$.
-However, for arguments close to 1, where $1/\log(\symX)$ and $\symSmallCorrectionFunc(\symX)$ are both very large while their difference is very small, special care is needed to avoid numerical cancellation.
+It is also thinkable to calculate $\symLargeCorrectionFunc$ using the approximation $\symLargeCorrectionFunc(\symX)\approx \symAlpha_\infty/\log(1/\symX)-\symSmallCorrectionFunc(\symX)$ which can be obtained from \eqref{equ:sigma_tau_relationship}.
+The advantage is that the calculation of $\symSmallCorrectionFunc$ and the additional logarithm is slightly faster than the calculation of $\symLargeCorrectionFunc$.
+However, for arguments close to 1, where $\symAlpha_\infty/\log(1/\symX)$ and $\symSmallCorrectionFunc(\symX)$ are both very large while their difference is very small, special care is needed to avoid numerical cancellation.
 
@@
 \label{sec:experimental_setup}
-In order to verify the new estimation algorithm, we generated \num{10000} different HyperLogLog sketches and filled each of them with 50 billion unique elements.
+To verify the new estimation algorithm, we generated \num{10000} different HyperLogLog sketches and filled each of them with 50 billion unique elements.
 During element insertion, we took snapshots of the sketch state by storing the multiplicity vector at predefined cardinality values, which were roughly chosen according to a geometric series with ratio \num{1.01}.
-In order to achieve that in reasonable time for different HyperLogLog parameter combinations  $(\symPrecision,\symRegRange)$ we have been interested in, we applied a couple of optimizations.
+In order to achieve that in reasonable time for all the different HyperLogLog parameter combinations  $(\symPrecision,\symRegRange)$ we have been interested in, we applied a couple of optimizations.
 
 First, we assumed a uniform hash function.
-This allows us to simply generate a random number instead of getting a unique element and calculating its hash value.
+This allows us to simply generate random numbers instead of creating unique elements and calculating their hash values.
 We used the Mersenne Twister random number generator with a state size of \num{19937} bits from the C++ standard library.
@@
 The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, $\symPrecision=22$ are shown in \cref{fig:raw_corrected_estimation_error_8_24}, \cref{fig:raw_corrected_estimation_error_16_16}, and \cref{fig:raw_corrected_estimation_error_22_10}, respectively.
-As expected, since  $\symPrecision + \symRegRange = 32$ is kept constant, the operating range remains more or less the same, while the relative error decreases with increasing precision.
+As expected, since  $\symPrecision + \symRegRange = 32$ is kept constant, the operating range remains more or less the same, while the relative error decreases with increasing precision parameter $\symPrecision$.
 Again, the new algorithm gives essentially unbiased estimates.
@@
 Using a lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed due to the small probability of saturated registers, $\symCountVariate_{53}=0$ usually holds for realistic cardinalities.
-
-For the case $\symRegRange=20$ with precalculated correction values the computation time is again -- as expected -- independent of the cardinality.
-The faster computation times compared to the $\symRegRange=52$ case can be explained by the much smaller dimension of the multiplicity vector which is equal to $\symRegRange+2$.
-If the functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are calculated on demand, the execution times for larger cardinalities are doubled.
-This comes from the fact that the calculation of $\symLargeCorrectionFunc$ is much more expensive than that of $\symSmallCorrectionFunc$, because it requires more iterations and involves square root evaluations.
-If $2^{\symPrecision+\symRegRange}$ is greater than the maximum expected cardinality value, the number of saturated registers is negligible ($\symCountVariate_{\symRegRange+1}\ll \symNumReg$) and the calculation of $\symLargeCorrectionFunc$ can be omitted at all, because $\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg) \approx 0$ vanishes in this case.
+If lookup tables are used, the computation time is as expected independent of the cardinality also for the case $\symRegRange=20$.
+The faster computation times for $\symRegRange=20$ compared to the $\symRegRange=52$ case can be explained by the much smaller dimension of the multiplicity vector which is equal to $\symRegRange+2$.
+In contrast, if functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are calculated on demand, executions take significantly more time for larger cardinalities.
+The reason is that beginning at cardinality values in the order of \num{100000} the probability of saturated registers increases.
+This makes the calculation of $\symLargeCorrectionFunc$ necessary, which is much more expensive than that of $\symSmallCorrectionFunc$, because it requires more iterations and involves square root evaluations.
+However, the calculation of
+$\symLargeCorrectionFunc$ could be avoided at all, if the HyperLogLog parameters are appropriately chosen.
+If $2^{\symPrecision+\symRegRange}$ is much larger than the maximum expected cardinality value, the number of saturated registers will be negligible, $\symCountVariate_{\symRegRange+1}\ll \symNumReg$.
+The calculation of $\symLargeCorrectionFunc$ could be omitted in this case, because $\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg) \approx 0$.
 

### 2017-02-03 python/benchmark-charts.py

@@ -58,11 +58,11 @@ fig_time, ax_time = plt.subplots(subplot_kw=dict(xlim=[xMin,xMax], ylim=[0,800],
 fig_time.set_size_inches(10, 5)
 ax_time.yaxis.grid(b=True, which='major', color='gray', linestyle='--')
 ax_time.xaxis.grid(b=True, which='major', color='gray', linestyle='--')
-ax_time.set_xlabel("cardinality")
-ax_time.set_ylabel("time (ns)")
+ax_time.set_xlabel(r"cardinality")
+ax_time.set_ylabel(r"time (ns)")
 
-ax_time.plot(cardinalities_12_20, max_likelihood_avg_exec_time_nanos_12_20, color='green', linestyle='-',linewidth=1.0, label='p=12, q=20')
-ax_time.plot(cardinalities_12_52, max_likelihood_avg_exec_time_nanos_12_52, color='blue', linestyle='-',linewidth=1.0, label='p=12, q=52')
+ax_time.plot(cardinalities_12_20, max_likelihood_avg_exec_time_nanos_12_20, color='green', linestyle='-',linewidth=1.0, label=r"$p=12$, $q=20$")
+ax_time.plot(cardinalities_12_52, max_likelihood_avg_exec_time_nanos_12_52, color='blue', linestyle='-',linewidth=1.0, label=r"$p=12$, $q=52$")
 
 ax_time.legend(loc=2,prop={'size':12})
 
@@ -78,10 +78,10 @@ ax_time.xaxis.grid(b=True, which='major', color='gray', linestyle='--')
 ax_time.set_xlabel("cardinality")
 ax_time.set_ylabel("time (ns)")
 
-ax_time.plot(cardinalities_12_20, corrected_raw_1_avg_exec_time_nanos_12_20, color='green', linestyle='-',linewidth=1.0, label='p=12, q=20, on-demand')
-ax_time.plot(cardinalities_12_52, corrected_raw_1_avg_exec_time_nanos_12_52, color='blue', linestyle='-',linewidth=1.0, label='p=12, q=52, on-demand')
-ax_time.plot(cardinalities_12_20, corrected_raw_2_avg_exec_time_nanos_12_20, color='red', linestyle='-',linewidth=1.0, label='p=12, q=20, precalculated')
[…]

### 2017-02-04 paper/paper.tex
some minor changes

@@
 \symDefine{\symNumBitsMinwiseHashing}{b}
+\symDefine{\symImaginary}{\mathrm{i}}
 \symDefine{\symIndexI}{i}
-\symDefine{\symImaginary}{\mathrm{i}}
 \symDefine{\symIndexJ}{j}
@@
 \symDefine{\symB}{b}
-\symDefine{\symKappa}{\kappa}
+\symDefine{\symS}{s}
 \symDefine{\symX}{x}
@@
 \symDefine{\symZ}{z}
+\symDefine{\symKappa}{\kappa}
 \symDefine{\symPhi}{\varphi}
 \symDefine{\symXEstimate}{\hat{\symX}}
-\symDefine{\symXNormalized}{\symX'}
-\symDefine{\symN}{n}
-\symDefine{\symS}{s}
-\symDefine{\symInteger}{z}
-\symDefine{\symIntegerVariate}{Z}
 \symDefine{\symCardinality}{n}
@@
 \symDefine{\symHelperApprox}{\tilde{\symHelper}}
-\symDefine{\symHelperW}{w}
-\symDefine{\symHelperV}{v}
-\symDefine{\symHelperU}{u}
 \symDefine{\symSetA}{A}
 \symDefine{\symSetB}{B}
-\symDefine{\symSetC}{C}
 \symDefine{\symSetS}{S}
@@
 \symDefine{\symSetBSuffix}{b}
-\symDefine{\symSetCSuffix}{c}
 \symDefine{\symSetXSuffix}{x}
-\symDefine{\symCountMatrix}{\mathbf{c}}
 \symDefine{\symPowerSeriesFunc}{\xi}
@@
 
-At any time a $(\symPrecision, \symRegRange)$-HyperLogLog sketch can be compressed into a $(\symPrecision', \symRegRange')$-HyperLogLog data structure, if $\symPrecision'\leq\symPrecision$ and $\symPrecision'+\symRegRange' \leq\symPrecision + \symRegRange$ is satisfied (see \cref{alg:compress}).
+At any time a $(\symPrecision, \symRegRange)$-HyperLogLog sketch can be reduced to a $(\symPrecision', \symRegRange')$-HyperLogLog data structure, if $\symPrecision'\leq\symPrecision$ and $\symPrecision'+\symRegRange' \leq\symPrecision + \symRegRange$ is satisfied (see \cref{alg:compress}).
 This transformation is lossless in a sense that the resulting HyperLogLog sketch is the same as if all elements would have been recorded by a $(\symPrecision', \symRegRange')$-HyperLogLog sketch right from the beginning.
@@
 and
-%\begin{equation}
-%\symLargeCorrectionFunc(\symX)
-%:=
-%\sum_{\symRegVal=1}^\infty
-%\symX^{2^{-\symRegVal}}
-%\left(1-\symX^{2^{-\symRegVal}}\right)
-%2^{-\symRegVal}.
-%\end{equation}
 \begin{equation}
@@
 \end{equation}
-%=
-%\frac{1}{3}
-%\left(
-%1
-%-
-%\symX
-%-
-%\sum_{\symRegVal=1}^\infty
-%\left(
-%1-
-%\symX^{2^{-\symRegVal}}
-%\right)^2
-%2^{-\symRegVal}
-%\right)
 
@@
 
-%
-%The case $\symX=1$, for which the series does not converge, occurs only if all register values are zero ($\symCountVariate_0=\symNumReg$).
-However, the resulting positive infinite denominator in \eqref{equ:correctedestimator} yields zero for the cardinality estimate as expected.
-In order to estimate the number of iteration cycles needed to determine $\symSmallCorrectionFunc$ up to some precision $\symStopDelta$, we consider the case $\symX = 1-\frac{1}{\symNumReg}$ for which we have slowest convergence speed among all relevant argument values $\lbrace0,\frac{1}{\symNumReg},\frac{2}{\symNumReg},\ldots,1-\frac{1}{\symNumReg}\rbrace$.
-We can stop summation, if two subsequent terms differ by a factor more than $\symStopDelta$.
-This is the case, if $\frac{\symX^{2^{\symRegVal+1}} 2^{\symRegVal}}{\symX^{2^\symRegVal} 2^{\symRegVal-1}} < \symStopDelta$ which is equivalent to $
-%\symRegVal > \frac{
-%\log(\log(2/\symStopDelta))
-%-
-%\log(-\log(1-\frac{1}{\symNumReg}))
-%}
-%{
-%\log(2)
-%}
-%$. Due to the inequality $-\log(1-\frac{1}{\symNumReg})\geq \frac{1}{\symNumReg}$ this is also satisfied, if $
-%\symRegVal > \frac{
-%\log(\log(2/\symStopDelta))
-%}
-%{
-%\log(2)
-%}
-%+
-%\symPrecision
-%$
-%holds. Therefore, not more than $\symPrecision + 5$ iterations are necessary to achieve double-precision $\symStopDelta = 2^{-53}$.
-
 The calculation of $\symLargeCorrectionFunc$ is more expensive, because it involves square root evaluations. $\symLargeCorrectionFunc$ can also be written in a numerically more favorable way as 
@@
 For this case and if double-precision floating-point numbers are used, the function $\symLargeCorrectionFunc$ presented in \cref{alg:corrected_raw_estimation} converges after \num{21} and \num{22} iteration cycles for considered precisions $\symPrecision=12$ and $\symPrecision=20$, respectively.
-%
-%Since $\symLargeCorrectionFunc(0)$ and $\symLargeCorrectionFunc(1)$ are both equal to zero, we focus on the remaining relevant arguments $\symX\in[\frac{1}{\symNumReg},1-\frac{1}{\symNumReg}]$.
-Obviously, we have slowest convergence for $\symX = \frac{1}{\symNumReg}$.
-Since the left factor in \eqref{equ:tau2} is monotone increasing we can find
-%\begin{equation}
-%1-
-%\left(1-\frac{1}{\symNumReg}\right)
-%\sum_{\symRegVal=1}^\infty
-%\prod_{\symIndexI=1}^\symRegVal
-%\frac{1}
-%{
-%2
-%\left(
-%1+
-%{\frac{1}{\symNumReg}}^{2^{-\symRegVal}}
-%\right)^2
-%}
-%\geq
-%1-
-%\frac{1-\frac{1}{\symNumReg}}
-%{
-%\left(
-%1+
-%{\frac{1}{\sqrt{\symNumReg}}}
-%\right)^2
-%}
-%\sum_{\symRegVal=1}^\infty
-%\frac{1}
-%{
-%2^\symRegVal
-%}
-%=
-%\frac{2}{\sqrt{\symNumReg}+1}
-%\end{equation}
-%as lower bound for $\symX\geq\frac{1}{\symNumReg}$. 
-%
-%
-%
-% for which the smallest value is obtained for $\symX = 1-\frac{1}{\symNumReg}$. 
-%It can be shown that $\symLargeCorrectionFunc(1-\frac{1}{\symNumReg})\geq \frac{1}{4\symNumReg}$. Hence, summation in \eqref{equ:sigma} can be stopped, if $\left(
-%1-
-%\symX^{2^{-\symRegVal}}
-%\right)^2
-%2^{-\symRegVal}
-%< \frac{\symStopDelta}{4\symNumReg}
-%$. Since we have slowest convergence for $\symX=\frac{1}{\symNumReg}$, $\left(
-%1-
-%\symNumReg^{-2^{-\symRegVal}}
-%\right)^2
-%2^{-\symRegVal}
-%< \frac{\symStopDelta}{4\symNumReg}
-%$ is also a sufficient condition to achieve the desired precision $\symStopDelta$. Furthermore, $1-
-%\symNumReg^{-2^{-\symRegVal}}
-%=
-%1-e^{-\log(\symNumReg)2^{-\symRegVal}}
-%\leq
-%\log(\symNumReg)2^{-\symRegVal}
-%$ which allows to write 
-%$(\log(\symNumReg))^2 8^{-\symRegVal}
-%< \frac{\symStopDelta}{4\symNumReg}
-%$ and finally $\symRegVal > \frac{\symPrecision}{3} + \frac{2\log(\symPrecision)}{\log(8)} + 
-%\frac{\log(\log(16)/\symStopDelta)}{\log(8)}$.
-%
-%
-%
-%
 
@@
 First we considered configurations using a 32-bit hash function ($\symPrecision + \symRegRange = 32$).
-The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, $\symPrecision=22$ are shown in \cref{fig:raw_corrected_estimation_error_8_24}, \cref{fig:raw_corrected_estimation_error_16_16}, and \cref{fig:raw_corrected_estimation_error_22_10}, respectively.
+The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, and $\symPrecision=22$ are shown in \cref{fig:raw_corrected_estimation_error_8_24,fig:raw_corrected_estimation_error_16_16,fig:raw_corrected_estimation_error_22_10}, respectively.
 As expected, since  $\symPrecision + \symRegRange = 32$ is kept constant, the operating range remains more or less the same, while the relative error decreases with increasing precision parameter $\symPrecision$.
@@
 
-For HyperLogLog sketches that have been obtained without stochastic averaging (see \cref{sec:data_element_insertion}) the maximum likelihood method was already previously considered for cardinality estimation \cite{Clifford2012}.
-However, in this case the register values are statistically independent by nature which
+For HyperLogLog sketches that have been obtained without stochastic averaging (see \cref{sec:data_element_insertion}) the maximum likelihood method was already previously considered \cite{Clifford2012}.
+The register values are statistically independent by nature in this case, which
 allows factorization of the joint probability mass function and a straightforward calculation of the maximum likelihood estimate.
-The practically more relevant stochastic averaging case, which is considered in this paper, would lead to a more complicated likelihood function.
+The practically more relevant case which uses stochastic averaging and which is considered in this paper, would lead to a more complicated likelihood function (compare \eqref{equ:multinomialProbabilityMass}).
 However, the Poisson approximation makes the maximum likelihood method feasible again and we are finally able to derive another new robust and efficient cardinality estimation algorithm.
@@
 $\symHelper(\symX)$ is strictly increasing and concave as can be seen in \cref{fig:helper_function}.
-For non-negative values this function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
+For nonnegative values this function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
 \begin{figure}
@@
 Since the function $\symFunc(x)$ is also strictly increasing, it is obvious that there exists a unique root $\symXEstimate$ for which $\symFunc(\symXEstimate)=0$.
-The function is non-positive at 0 since $\symFunc(0)=\symCountVariate_0-\symNumReg\leq 0$ and, in case $\symCountVariate_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing.
+The function is nonpositive at 0 since $\symFunc(0)=\symCountVariate_0-\symNumReg\leq 0$ and, in case $\symCountVariate_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing.
 $\symCountVariate_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $\symRegRange+1$, for which the maximum likelihood estimate would be positive infinite.
@@
 \label{sec:comp_ml_estimate}
-Since $\symFunc$ is concave and increasing, both, Newton-Raphson iteration and the secant method, will converge to the root, provided that the function is negative for the starting points.
-In the following we start from the secant method to derive the new cardinality estimation algorithm.
+Since $\symFunc$ is concave and increasing, both, Newton-Raphson iteration and the secant method, will converge to the root, provided that the function is negative for the chosen starting points.
+We will use the secant method to derive the new cardinality estimation algorithm.
 Even though the secant method has the disadvantage of slower convergence, a single iteration is simpler to calculate as it does not require the evaluation of the first derivative.
@@
 \subsection{Maximum likelihood estimation algorithm}
-In order to get a fast cardinality estimation algorithm, it is crucial to minimize evaluation costs for \eqref{equ:funcprime}.
-A couple of optimizations allow significant reduction of computational effort:
+To get a fast cardinality estimation algorithm, it is crucial to minimize evaluation costs for \eqref{equ:funcprime}. A couple of optimizations allow significant reduction of computational effort:
 \begin{itemize}
-\item Only a fraction of all count values $\symCountVariate_\symRegVal$ is non-zero.
+\item Only a fraction of all count values $\symCountVariate_\symRegVal$ is nonzero.
 If we denote $\symRegValVariate_\text{min}:=\min(\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace)$ and $\symRegValVariate_\text{max}:=\max(\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace)$,  it is sufficient to loop over all indices in the range $[\symRegValVariate_\text{min}, \symRegValVariate_\text{max}]$.
@@
 \item Many programming languages allow the efficient multiplication and division by any integral power of two using special functions, such as $\operatorname{ldexp}$ in C/C++ or $\operatorname{scalb}$ in Java.
-\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\left\lbrace\frac{\symX}{2^{\symRegValVariate'_\text{max}}},\frac{\symX}{2^{\symRegValVariate'_\text{max}-1}},\ldots,\frac{\symX}{2^{\symRegValVariate_\text{min}}}\right\rbrace$ where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$.
+\item The function $\symHelper(\symX)$ only needs to be evaluated at points $\left\lbrace\frac{\symX}{2^{\symRegValVariate'_\text{max}}},\frac{\symX}{2^{\symRegValVariate'_\text{max}-1}},\ldots,\frac{\symX}{2^{\symRegValVariate_\text{min}}}\right\rbrace$ where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$.
 This series corresponds to a geometric series with ratio two.
@@
 \end{equation}
-%or
-%\begin{equation}
-%\label{equ:helper_recursion2}
-%\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right) = \frac{\frac{\symX}{2^{\symRegVal+2}}+\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}{\frac{\symX}{2^{\symRegVal+2}}+\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}
-%\end{equation}
 This recursive formula is stable in a sense that the relative error of $\symHelper(4\symX)$ is smaller than that of $\symHelper(2\symX)$ as shown in \cref{app:helper_stable}.
@@
 \State $\symKappa \gets 2+\lfloor\log_2(\symX)\rfloor$
-\State $\symXNormalized \gets
+\State $\symX' \gets
 \symX \cdot2^{-\max(\symRegValVariate'_\text{max}, \symKappa)-1}$
-\comm{0.31}{$\symXNormalized \in [0, 0.25]$}
-\State $\symX''\gets \symXNormalized\cdot\symXNormalized$
+\comm{0.31}{$\symX' \in [0, 0.25]$}
+\State $\symX''\gets \symX'\cdot\symX'$
 \State $\symHelper \gets
@@
 \For{$\symRegVal\gets (\symKappa - 1),\symRegValVariate'_\text{max}$}
-\State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
-\comm{0.31}{calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$}
-\State $\symXNormalized \gets 2\symXNormalized$
+\State $\symHelper \gets \frac{\symX'+\symHelper\cdot(1-\symHelper)}{\symX'+(1-\symHelper)}$
+\comm{0.31}{calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symX' = \frac{\symX}{2^{\symRegVal+2}}$}
+\State $\symX' \gets 2\symX'$
 \EndFor
@@
 \For{$\symRegVal\gets(\symRegValVariate'_\text{max}-1),\symRegValVariate'_\text{min}$}
-\State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
-\comm{0.31}{calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$}
+\State $\symHelper \gets \frac{\symX'+\symHelper\cdot(1-\symHelper)}{\symX'+(1-\symHelper)}$
+\comm{0.31}{calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symX' = \frac{\symX}{2^{\symRegVal+2}}$}
 \State $\symFuncPrime\gets \symFuncPrime + \symCountVariate_\symRegVal\cdot\symHelper$
-\State $\symXNormalized \gets 2\symXNormalized$
+\State $\symX' \gets 2\symX'$
 \EndFor
@@
 We have investigated the estimation error of the maximum likelihood estimation algorithm for the same HyperLogLog configurations as for the improved raw estimation algorithm in \cref{sec:corrected_raw_estimation_error}.
-\cref{fig:max_likelihood_estimation_error_12_20}, \cref{fig:max_likelihood_estimation_error_8_24},  \cref{fig:max_likelihood_estimation_error_16_16}, \cref{fig:max_likelihood_estimation_error_22_10}, \cref{fig:max_likelihood_estimation_error_12_52}, and \cref{fig:max_likelihood_estimation_error_12_14} show very similar results for same HyperLogLog parameters.
+\cref{fig:max_likelihood_estimation_error_12_20,fig:max_likelihood_estimation_error_8_24,fig:max_likelihood_estimation_error_16_16,fig:max_likelihood_estimation_error_22_10,fig:max_likelihood_estimation_error_12_52,fig:max_likelihood_estimation_error_12_14} show very similar results for same HyperLogLog parameters.
 What is different for the maximum likelihood estimation approach is a somewhat smaller median bias with less oscillations around zero for small cardinalities.
@@
 Here the last transformation used the independence of $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$, because by definition, the sets $\symSetA$, $\symSetB$, and $\symSetX$ are pairwise disjoint.
-Furthermore, under the Poisson model $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ obey
-\eqref{equ:register_value_distribution}.
+Furthermore, under the Poisson model $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ can be described by \eqref{equ:register_value_distribution}.
 If we take into account that pairwise distinct elements are added to $\symSetA$, $\symSetB$, and $\symSetX$ at rates $\symPoissonRate_\symSetXSuffix$, $\symPoissonRate_\symSetASuffix$, and $\symPoissonRate_\symSetBSuffix$, respectively, the probability that a certain register has a value less than or equal to $\symRegVal_1$ in the first HyperLogLog sketch and simultaneously a value less than or equal to $\symRegVal_2$ in the second one can be written as
@@
 
-Since \eqref{equ:log_likelihood_pair} is the generalization of \eqref{equ:log_likelihood_single} for two HyperLogLog sketches, the log-likelihood function for a single HyperLogLog sketch can be obtained by considering special cases.
+Since \eqref{equ:log_likelihood_pair} is a generalization of \eqref{equ:log_likelihood_single} for two HyperLogLog sketches, the log-likelihood function for a single HyperLogLog sketch can be obtained by considering special cases.
 For example, assume $\symPoissonRate_\symSetXSuffix=0$, which means that both sketches represent disjoint sets, \eqref{equ:log_likelihood_pair} can be splitted into the sum of two unary functions with parameters $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$ each of which correspond to \eqref{equ:log_likelihood_single} as expected.
@@
 The maximum likelihood estimates can be obtained by maximizing \eqref{equ:log_likelihood_pair}.
-Since the three parameters are all non-negative, this is a constrained optimization problem.
+Since the three parameters are all nonnegative, this is a constrained optimization problem.
 In order to get rid of these constraints, we use the transformation $\symPoissonRate=e^{\symPhi}$.
@@
 
-Quasi-Newton methods are commonly used for finding the maximum of such multi-dimensional functions.
+Quasi-Newton methods are commonly used to find the maximum of multi-dimensional functions.
 They require the calculation of the gradient $\nabla\symFunc = \left(\frac{\partial\symFunc}{\partial\symPhi_\symSetASuffix}, \frac{\partial\symFunc}{\partial\symPhi_\symSetBSuffix},
@@
 For the general case, the three-dimensional function $\symFunc(\symPhi_\symSetASuffix,\symPhi_\symSetBSuffix,\symPhi_\symSetXSuffix)$ needs to be optimized numerically.
-A very popular algorithm for such non-linear optimization problems is the Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{Press2007}.
+A very popular algorithm for such nonlinear optimization problems is the Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{Press2007}.
 In particular, we used the implementation provided by the Dlib C++ library \cite{King2009}.
@@
 \comm{0.32}{calculate $\symFunc$ and
-$\nabla\symFunc$ as needed by optimizer using \eqref{equ:max_likelihood_phi_func} and \eqref{equ:df_dphi},
-respectively
+$\nabla\symFunc$ using \eqref{equ:max_likelihood_phi_func} and \eqref{equ:df_dphi} as needed by optimizer
 }
@@
 To evaluate the estimation error of the new joint cardinality estimation approach we applied \cref{alg:joint_cardinality_estimation} to HyperLogLog sketch pairs with parameters $\symPrecision=20$ and $\symRegRange=44$ and with known cardinalities $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$.
-HyperLogLog sketch with a predefined cardinality can be quickly generated by taking a precalculated multiplicity vector with known cardinality, setting the register values accordingly, and shuffling them.
+HyperLogLog sketches with predefined cardinalities can be quickly generated by taking a precalculated multiplicity vector with known cardinality, setting the register values accordingly, and shuffling them.
 Using this approach we generated three different HyperLogLog sketches with known cardinalities $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$.
 Then we merged the HyperLogLog sketches representing the disjoint sets $\symSetA$ and $\symSetX$ using \cref{alg:union} in order to get a sketch for $\symSetS_1 = \symSetA\cup\symSetX$.
-A second sketch was built analogously for $\symSetS_2=\symSetA\cup\symSetX$.
-Since three independent multiplicity vectors are needed to construct an example and we had simulated \num{10000} different HyperLogLog sketches previously as described in \cref{sec:experimental_setup}, we were able to construct \num{3333} different independent HyperLogLog pairs for a given set of known true cardinalities $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$.
+A second sketch for $\symSetS_2=\symSetA\cup\symSetX$ was built analogously.
+Since three independent multiplicity vectors are needed to construct a single example and we had simulated \num{10000} different HyperLogLog sketches previously as described in \cref{sec:experimental_setup}, we were able to construct \num{3333} different independent HyperLogLog pairs for a given set of known true cardinalities $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$.
 
 \cref{tbl:joint_estimation_cases} lists all the different cardinality configurations for which we have evaluated the joint cardinality estimation  algorithm.
-\cref{tbl:joint_estimation_cases} also shows the average number of iterations of the BFGS algorithm until the stop criterion was satisfied.
+Among the considered cases there are also cardinalities that are small compared to the number of registers in order to prove that the new approach also covers the small cardinality range where many register values are equal to zero.
+The last column of \cref{tbl:joint_estimation_cases} shows the average number of iterations of the BFGS algorithm until the stop criterion was satisfied.
 Each iteration step involved a function \eqref{equ:max_likelihood_phi_func} and a gradient \eqref{equ:df_dphi} evaluation.
 
-\cref{tbl:joint_estimation_a}, \cref{tbl:joint_estimation_b}, and \cref{tbl:joint_estimation_x} compare the relative estimation error to the conventional approach using single sketch cardinality estimation together with \eqref{equ:conventional_approach}.
-The mean, the standard deviation, and the root-mean-square error are given.
+\cref{tbl:joint_estimation_a,tbl:joint_estimation_b,tbl:joint_estimation_x} compare the relative estimation error for cardinalities $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$ to that of the conventional approach using single sketch cardinality estimation together with \eqref{equ:conventional_approach}.
+The mean, the standard deviation, and the root-mean-square error have been calculated from the cardinality estimates of \num{3333} examples.
 Furthermore, we calculated an improvement factor which represents the root-mean-square error ratio between both approaches.
-Since we only observed values larger than 1, the new maximum likelihood estimation approach gives better estimates for all investigated cases.
-For some cases the improvement factor is clearly larger than 2.
-Due to the  square root scaling of the error, this means that we would need 4 times more registers to get the same error when using the conventional approach.
-We also considered some cases with true cardinalities smaller  than the number of registers.
-As the results suggest and as expected the joint estimation algorithm works well over the entire cardinality range without special handling of small cardinalities.
+Since we only observed values greater than one, the new maximum likelihood estimation approach gives better estimates for all investigated cases.
+For some cases the improvement factor is even clearly greater than two.
+Due to the  square root scaling of the error, this means that we would need four times more registers to get the same error when using the conventional approach.
+As the results suggest the joint estimation algorithm works well over the entire cardinality range without the need of special handling of small cardinalities.
 
@@
 The joint cardinality estimation algorithm provides another opportunity.
-The union can also be estimated by simply summing up the three estimates for the sets $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$.
+The union can also be estimated by simply summing up the three estimates for sets $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$.
 The corresponding results are shown in \cref{tbl:joint_estimation_abx}.
-As can be clearly seen the joint estimation algorithm is also able to improve the cardinality estimation of unions by a significant amount.
+As can be clearly seen, the joint estimation algorithm is also able to improve the cardinality estimation of unions by a significant amount.
 
@@
 \centering
-\captionabove{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetA|$ using the conventional approach and the maximum likelihood method, respectively.
-The root-mean-square error improvement factor of the maximum likelihood method is shown in the last column.}
+\captionabove{The mean, the standard deviation, and the root-mean-square error of the relative estimation error for all cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetA|$ using the conventional approach and the maximum likelihood method, respectively.}
 \label{tbl:joint_estimation_a}
@@
 \centering
-\captionabove{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetB|$ using the conventional approach and the maximum likelihood method, respectively.
-The root-mean-square error improvement factor of the maximum likelihood method is shown in the last column.}
+\captionabove{The mean, the standard deviation, and the root-mean-square error of the relative estimation error for all cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetB|$ using the conventional approach and the maximum likelihood method, respectively.}
 \label{tbl:joint_estimation_b}
@@
 \centering
-\captionabove{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetX|$ using the conventional approach and the maximum likelihood method, respectively.
-The root-mean-square error improvement factor of the maximum likelihood method is shown in the last column.}
+\captionabove{The mean, the standard deviation, and the root-mean-square error of the relative estimation error for all cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetX|$ using the conventional approach and the maximum likelihood method, respectively.}
 \label{tbl:joint_estimation_x}
@@
 \centering
-\captionabove{The mean, the standard deviation, and the root-mean-square error of the relative estimation error are given for all the cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetA\cup\symSetB\cup\symSetX|$ using the conventional approach and the maximum likelihood method, respectively.
-The root-mean-square error improvement factor of the maximum likelihood method is shown in the last column.}
+\captionabove{The mean, the standard deviation, and the root-mean-square error of the relative estimation error for all cases given in \cref{tbl:joint_estimation_cases} when estimating $|\symSetA\cup\symSetB\cup\symSetX|$ using the conventional approach and the maximum likelihood method, respectively.}
 \label{tbl:joint_estimation_abx}
@@
 Unfortunately, joint cardinality estimation is much more expensive than for a single HyperLogLog sketch, because it requires maximization of a multi-dimensional function.
-Since we have found the improved raw estimator which is almost as accurate as the maximum likelihood estimator for the single HyperLogLog case, we could imagine that there also exists a faster algorithm for the two HyperLogLog case.
+Since we have found the improved raw estimator which is almost as precise as the maximum likelihood estimator for the single HyperLogLog case, we could imagine that there also exists a faster algorithm for the two HyperLogLog case.
 It is expected that such a new algorithm makes use of all the information given by the sufficient statistic \eqref{equ:sufficient_joint_statistic}.
@@
 
-In order to investigate the maximum deviation of $\symPowerSeriesFunc(\symX)$ from this value we consider all further coefficients $\symA_\symIndexL$ with $\symIndexL\geq 1$.
+To investigate the maximum deviation of $\symPowerSeriesFunc(\symX)$ from this value we consider all further coefficients $\symA_\symIndexL$ with $\symIndexL\geq 1$.
 Using the identity $\left|\Gamma(1+\symImaginary\symX)\right| =

### 2017-02-09 paper/paper.tex
fixed some typos

@@
 The simulation results show that the estimates are significantly better than those of the conventional approach using the inclusion-exclusion principle.
-Finally, in \cref{sec:future_work} we discuss open problems and ideas for future work including HyperLogLog's potential application as locality-sensitve hashing algorithm before we conclude in \cref{sec:conclusion}.
+Finally, in \cref{sec:future_work} we discuss open problems and ideas for future work including HyperLogLog's potential application as locality-sensitive hashing algorithm before we conclude in \cref{sec:conclusion}.
 
@@
 The relative estimation error for the original method is shown in \cref{fig:original_estimate}.
-Unfortunatley, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityRawEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap.
+Unfortunately, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityRawEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap.
 Therefore, the estimation error is much larger near the transition region.
@@
 \includegraphics[width=1\textwidth]{stdev_comparison}
-\caption{Standard deviations of the relative error of different cardinality estimators over the true cardinality for a HyperLogLog sketch with paramters $\symPrecision = 12$ and $\symRegRange=20$ .}
+\caption{Standard deviations of the relative error of different cardinality estimators over the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$ .}
 \label{fig:stdev_comparison}
@@
 \end{equation}
-If we set $\symX_0$ equal to 0, for which  $\symFunc(\symX_0)=-\left(\symNumReg-\symCountVariate_0\right)$, and $\symX_1$ equal to one of the derived lower bounds \eqref{equ:strong_lower_bound} or \eqref{equ:weak_lower_bound}, the sequence $(\symX_0, \symX_1, \symX_2, \ldots)$ is montone increasing.
+If we set $\symX_0$ equal to 0, for which  $\symFunc(\symX_0)=-\left(\symNumReg-\symCountVariate_0\right)$, and $\symX_1$ equal to one of the derived lower bounds \eqref{equ:strong_lower_bound} or \eqref{equ:weak_lower_bound}, the sequence $(\symX_0, \symX_1, \symX_2, \ldots)$ is monotone increasing.
 Using the definitions
@@
 The iteration can be stopped, if $\Delta\symX_{\symIndexI} \leq \symStopDelta\cdot \symX_{\symIndexI}$.
-Since the expected statistical errror for the HyperLogLog data structure scales according to $\frac{1}{\sqrt{\symNumReg}}$ \cite{Flajolet2007}, it makes sense to choose $\symStopDelta = \frac{\symStopEpsilon}{\sqrt{\symNumReg}}$ with some constant $\symStopEpsilon$.
+Since the expected statistical error for the HyperLogLog data structure scales according to $\frac{1}{\sqrt{\symNumReg}}$ \cite{Flajolet2007}, it makes sense to choose $\symStopDelta = \frac{\symStopEpsilon}{\sqrt{\symNumReg}}$ with some constant $\symStopEpsilon$.
 For all results presented later in \cref{sec:maximum_likelihood_estimation_error} we used $\symStopEpsilon = 10^{-2}$.
@@
 
-\subsection{Joint log-likelihood funtion}
+\subsection{Joint log-likelihood function}
 In order to get maximum likelihood estimators for $\symPoissonRateEstimate_\symSetASuffix$, $\symPoissonRateEstimate_\symSetBSuffix$, and $\symPoissonRateEstimate_\symSetXSuffix$ we need to derive the joint probability distribution of two HyperLogLog sketches.
@@
 \end{equation}
-These $5\symNumReg$ values represent a sufficient statistic for estimating $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$.
+These $5(\symRegRange+2)$ values represent a sufficient statistic for estimating $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$.
 Actually, the number of values could be further reduced, because of the invariants $\symCountVariate^{>}_{10}=\symCountVariate^{>}_{20}=
@@
 Since \eqref{equ:log_likelihood_pair} is a generalization of \eqref{equ:log_likelihood_single} for two HyperLogLog sketches, the log-likelihood function for a single HyperLogLog sketch can be obtained by considering special cases.
-For example, assume $\symPoissonRate_\symSetXSuffix=0$, which means that both sketches represent disjoint sets, \eqref{equ:log_likelihood_pair} can be splitted into the sum of two unary functions with parameters $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$ each of which correspond to \eqref{equ:log_likelihood_single} as expected.
+For example, assume $\symPoissonRate_\symSetXSuffix=0$, which means that both sketches represent disjoint sets, \eqref{equ:log_likelihood_pair} can be split into the sum of two unary functions with parameters $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$ each of which correspond to \eqref{equ:log_likelihood_single} as expected.
 Or, consider two sketches representing identical sets.
@@
 
-Since HyperLogLog sketches can be effectively constructed, because only a single hash function evaluation is needed for each item, the preprocessing step would be very fast.
+Since HyperLogLog sketches can be efficiently constructed, because only a single hash function evaluation is needed for each item, the preprocessing step would be very fast.
 In contrast, preprocessing is very costly for minwise hashing, because of the many required permutations \cite{Li2011}.

### 2017-02-15 paper/paper.tex
fixed some more typos

@@
 \begin{abstract}
-This paper presents new methods to estimate the cardinalities of multisets recorded by HyperLogLog sketches.
+This paper presents new methods to estimate the cardinalities of data sets recorded by HyperLogLog sketches.
 A theoretically motivated extension to the original estimator is presented that eliminates the bias for small and large cardinalities.
@@
 A more theoretically profound estimator for HyperLogLog sketches that does not depend on empirical data and significantly improves the estimation error is the historic inverse probability estimator \cite{Ting2014, Cohen2014}.
-It trades memory-efficiency for mergeability.
+It trades memory efficiency for mergeability.
 The estimator needs to be continuously updated while inserting elements and the estimate depends on the element insertion order.
@@
 \frac{\symNumReg-\symCountVariate_0}
-{\symCountVariate_0+\frac{3}{2}\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}}.
+{\symCountVariate_0+\frac{3}{2}\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^{\symRegRange+1}}}.
 \end{equation}
@@
 The standard deviation of the relative error is also slightly better for the maximum likelihood estimator than for the improved raw estimator as shown in \cref{fig:stdev_comparison}.
-Furthermore, contrary to the raw estimator which reveals a small oscillating bias for the mean (see \cref{fig:raw_corrected_estimation_error_22_10}), the maximum likelihood estimator seems to be completely unbiased (see \cref{fig:max_likelihood_estimation_error_22_10}).
+Furthermore, contrary to the improved raw estimator which reveals a small oscillating bias for the mean (see \cref{fig:raw_corrected_estimation_error_22_10}), the maximum likelihood estimator seems to be completely unbiased (see \cref{fig:max_likelihood_estimation_error_22_10}).
 
@@
 Furthermore, under the Poisson model $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ can be described by \eqref{equ:register_value_distribution}.
-If we take into account that pairwise distinct elements are added to $\symSetA$, $\symSetB$, and $\symSetX$ at rates $\symPoissonRate_\symSetXSuffix$, $\symPoissonRate_\symSetASuffix$, and $\symPoissonRate_\symSetBSuffix$, respectively, the probability that a certain register has a value less than or equal to $\symRegVal_1$ in the first HyperLogLog sketch and simultaneously a value less than or equal to $\symRegVal_2$ in the second one can be written as
+If we take into account that pairwise distinct elements are added to $\symSetA$, $\symSetB$, and $\symSetX$ at rates $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$, respectively, the probability that a certain register has a value less than or equal to $\symRegVal_1$ in the first HyperLogLog sketch and simultaneously a value less than or equal to $\symRegVal_2$ in the second one can be written as
 \begin{equation}
@@
 HyperLogLog algorithm an interesting candidate for locality-sensitive hashing with respect to the Jaccard similarity.
-The memory-efficiency of different hashing approaches can be measured by a storage factor that is the variance of the distance estimator multiplied by the number of bits used for storing the hash signature \cite{Li2011}.
+The memory efficiency of different hashing approaches can be measured by a storage factor that is the variance of the distance estimator multiplied by the number of bits used for storing the hash signature \cite{Li2011}.
 For a hash algorithm that maps two different sets to the same $\symNumBitsMinwiseHashing$-bit hash value with probability $1-\symA\symDistanceMeasure$ with some constant $\symA\in(0,1]$, the storage factor is given by $\symNumBitsMinwiseHashing\symDistanceMeasure\left(\frac{1}{\symA}-\symDistanceMeasure\right)$.
@@
 \leq
-\left|\symError_\symHelper\right|
+\frac{\left|\symError_\symHelper\right|}{2}
 \frac{
-\frac{1}{2}
 \sum_{\symRegVal=1}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
@@
 \leq
-\left|\symError_\symHelper\right|
-\left(
-\frac{1}{2}
+\frac{\left|\symError_\symHelper\right|}{2}
+\left(
+1
 +

### 2017-02-19 paper/paper.tex
fixed some typos

@@
 
-More practicable and already widely used for many applications is the HyperLogLog algorithm \cite{Flajolet2007} with a near-optimal space complexity $\symBigO(\symError^{-2} \log\log\symCardinality +\log \symCardinality)$.
+More practicable and already widely used in many applications is the HyperLogLog algorithm \cite{Flajolet2007} with a near-optimal space complexity $\symBigO(\symError^{-2} \log\log\symCardinality +\log \symCardinality)$.
 It has the nice property that partial results can be easily merged, which is a necessity for distributed environments.
@@
 Sometimes not only the number of distinct elements but also a sample of them is needed in order to allow later filtering according to some predicate and estimating the cardinalities of corresponding subsets.
-In this case the k-minimum values algorithm \cite{Beyer2007} is the method of choice which also allows set manipulations like the construction of intersections, disjoint sets, or unions \cite{Dasgupta2015}.
+In this case the k-minimum values algorithm \cite{Beyer2007} is the method of choice which also allows set manipulations like the construction of intersections, relative complements, or unions \cite{Dasgupta2015}.
 The latter operation is the only one that is natively supported by HyperLogLog sketches.
@@
 \left(
-\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
--
-\symProbability(\symRegValVariate \leq \symRegVal-1\vert\symPoissonRate)
+\symProbability(\symRegValVariate' \leq \symRegVal\vert\symPoissonRate)
+-
+\symProbability(\symRegValVariate' \leq \symRegVal-1\vert\symPoissonRate)
 \right)
